---
title: "Study design and data"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  metadata_file: "data/metadata/mehreen_metadata_corrected_45.txt"
  counts_file: "output/rmd_runs/transcript-exon-primary-corrected45_20260809-153146/host_counts_no_rrna.tsv"
  reference_gff: "data/reference/GCF_023897955.1_iqSchGreg1.2_genomic.gff"
  output_tag: "study-design-data"
---

## Purpose

::: {.walkthrough-grid}
::: {.walkthrough-step}
**Step 1 - Inspect inputs**

Check that metadata, transcript-plus-exon counts, and annotation files are present.
:::
::: {.walkthrough-step}
**Step 2 - Match samples**

Confirm every count column has one diet and infection label.
:::
::: {.walkthrough-step}
**Step 3 - Confirm design**

Review group sizes and assigned-count depth before modeling.
:::
:::

This page records the input files, sample labels, experimental groups, and
basic count-table checks used by downstream analyses. It is meant to be
knitted before running DESeq2 pages so mismatched sample names or incomplete
metadata are caught early.

The corrected design contains 45 libraries. Sample `1024` is assigned to the
infected diet-33 group, and the recovered library `1044` is assigned to the
infected diet-83 group. These assignments are checked explicitly below before
any expression model is fitted.

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

# Keep all paths relative to the repository root used by workflowR.
project_dir <- normalizePath(".", mustWork = TRUE)

# Every run writes into a new folder, preserving older results for comparison.
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

metadata_file <- file.path(project_dir, params$metadata_file)
counts_file <- file.path(project_dir, params$counts_file)
reference_gff <- file.path(project_dir, params$reference_gff)
```

## Input files

```{r input-files}
input_files <- data.frame(
  role = c("sample metadata", "gene count matrix", "reference annotation"),
  path = c(params$metadata_file, params$counts_file, params$reference_gff),
  exists = file.exists(c(metadata_file, counts_file, reference_gff))
)

knitr::kable(input_files)

stopifnot(all(input_files$exists))
```

## Metadata

```{r metadata-summary}
metadata <- read.delim(metadata_file, check.names = FALSE)
metadata$label <- as.character(metadata$label)
metadata$Diet <- factor(metadata$Diet, levels = c("33", "50", "83"))
metadata$Treatment <- factor(metadata$Treatment, levels = c("Control", "Infected"))

stopifnot(metadata$Diet[metadata$label == "1024"] == "33")
stopifnot(metadata$Treatment[metadata$label == "1024"] == "Infected")
stopifnot(metadata$Diet[metadata$label == "1044"] == "83")
stopifnot(metadata$Treatment[metadata$label == "1044"] == "Infected")

# This table shows whether the design is balanced across diet and infection.
design_table <- with(metadata, table(Diet, Treatment))
knitr::kable(as.data.frame.matrix(design_table))

summary_table <- data.frame(
  n_samples = nrow(metadata),
  diets = paste(levels(droplevels(metadata$Diet)), collapse = ", "),
  treatments = paste(levels(droplevels(metadata$Treatment)), collapse = ", ")
)
knitr::kable(summary_table)
```

## Count matrix

```{r counts-check}
counts <- read.delim(counts_file, check.names = FALSE)
gene_ids <- counts[[1]]
count_matrix <- counts[, -1, drop = FALSE]
rownames(count_matrix) <- gene_ids

# Count columns include file prefixes; convert them back to metadata labels.
count_labels <- sub("_MERGE$", "", sub("^mehreen_", "", colnames(count_matrix)))
metadata_labels <- metadata$label

sample_match <- data.frame(
  count_column = colnames(count_matrix),
  inferred_label = count_labels,
  found_in_metadata = count_labels %in% metadata_labels
)

knitr::kable(head(sample_match, 12))

stopifnot(all(sample_match$found_in_metadata))
stopifnot(length(metadata_labels) == ncol(count_matrix))
```

## Count depth

```{r count-depth}
library(ggplot2)

depth_df <- data.frame(
  label = count_labels,
  total_counts = colSums(count_matrix)
)
depth_df <- merge(depth_df, metadata, by = "label", all.x = TRUE)

ggplot(depth_df, aes(x = Diet, y = total_counts, color = Treatment)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.12, height = 0, size = 2) +
  scale_y_continuous(labels = scales::comma) +
  theme_classic() +
  labs(
    title = "Library size by diet and infection",
    x = "Diet",
    y = "Total assigned counts",
    color = "Treatment"
  )

write.csv(depth_df, file.path(out_dir, "library_size_by_sample.csv"),
          row.names = FALSE)
```

## Reproduction notes

```{r session-info}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
