---
title: "Workflow and reproducibility"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  metadata_file: "data/metadata/mehreen_metadata_fixed.txt"
  counts_file: "data/raw_read_counts/master_counts.csv"
  output_tag: "workflow-reproducibility"
---

## Purpose

This page is the reproducibility map for the project. It explains where inputs
live, where new results are written, and how to rerun the workflowR pages
without relying on hidden helper scripts.

## Repository layout

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r repo-layout}
layout <- data.frame(
  folder = c("analysis/", "data/metadata/", "data/raw_read_counts/",
             "data/reference/", "docs/", "output/rmd_runs/", "scripts/"),
  role = c(
    "workflowR R Markdown pages; new analysis code should live here",
    "sample metadata used to align counts with biological groups",
    "raw and merged gene count tables",
    "reference annotation files",
    "rendered workflowR website and committed visual assets",
    "fresh output folders created by each Rmd render",
    "legacy standalone scripts retained for provenance"
  )
)

knitr::kable(layout)
```

## Rerun commands

Run these commands from the repository root. Each rendered analysis page makes a
new timestamped folder under `output/rmd_runs/`.

```{bash, eval=FALSE}
# Render one page after editing its parameters.
Rscript -e 'workflowr::wflow_build("analysis/03-count-qc-exploration.Rmd")'

# Render the main manuscript-oriented pages.
Rscript -e 'workflowr::wflow_build(c(
  "analysis/01-study-design-data.Rmd",
  "analysis/02-workflow-reproducibility.Rmd",
  "analysis/03-count-qc-exploration.Rmd",
  "analysis/04-deseq2-interaction-model.Rmd",
  "analysis/05-diet-pairwise-deg.Rmd",
  "analysis/06-infection-within-diet-deg.Rmd",
  "analysis/07-deg-overlap-prioritization.Rmd",
  "analysis/08-publication-figures.Rmd"
))'

# Preview the workflowR status before publishing.
Rscript -e 'workflowr::wflow_status()'
```

## Packages used by the analysis pages

```{r package-check}
packages <- c(
  "workflowr", "knitr", "rmarkdown", "ggplot2", "scales",
  "DESeq2", "pheatmap", "ggrepel", "EnhancedVolcano", "UpSetR"
)

package_status <- data.frame(
  package = packages,
  installed = vapply(packages, requireNamespace, logical(1), quietly = TRUE)
)

knitr::kable(package_status)

write.csv(package_status, file.path(out_dir, "package_status.csv"),
          row.names = FALSE)
```

## Analysis contract

The current analysis pages follow this pattern:

1. Put editable file paths, thresholds, and model choices in the page `params`.
2. Load metadata and counts directly in the Rmd.
3. Check that count columns match metadata before modeling.
4. Write derived tables and figures to a new folder under `output/rmd_runs/`.
5. Save `sessionInfo()` in the same folder as the results.

This keeps the manuscript figures reproducible while leaving the older scripts
available for comparison.

```{r session-info}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
