---
title: "Workflow and reproducibility"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  metadata_file: "data/metadata/mehreen_metadata_analysis_44_excluding_1044.txt"
  counts_file: "data/derived/counts/transcript_exon_primary_host_counts_no_rrna.tsv.gz"
  output_tag: "workflow-reproducibility"
---

## Purpose

::: {.walkthrough-grid}
::: {.walkthrough-step}
**Step 1 - Process reads**

Follow fastp, STAR, and competitive host-pathogen mapping on the HPC.
:::
::: {.walkthrough-step}
**Step 2 - Count genes twice**

Use `-t transcript,exon -g gene_id` for the primary analysis and
`-t exon -g gene_id` for the matched sensitivity analysis.
:::
::: {.walkthrough-step}
**Step 3 - Reproduce reports**

Run the visible R Markdown code into new timestamped output folders.
:::
:::

This page is the reproducibility map for the project. It explains where inputs
live, where new results are written, and how to rerun the workflowR pages
without relying on hidden helper scripts.

Both count definitions use the same frozen 44-library analysis cohort and
corrected metadata. Sample `1024` is infected diet 33; library `1044` was
processed and retained in the QC audit but excluded from biological models.
The two branches differ only in which annotation feature types are counted, so
their DEG results can be compared directly.

## Bundled data and generated-output boundary

The GitHub repository contains the R Markdown code, corrected metadata, rRNA
exclusion list, count matrices, DESeq2 results, reference and eggNOG
annotations, phase-reference data, taxonomy summaries, scaffold-audit evidence,
and rendered web figures. Inputs promoted from authoritative analysis runs are
stored under `data/derived/` with stable names so the reports do not depend on
historical timestamped output folders.

The NCBI GFF is committed in compressed form. Expand its ignored working copy
once after cloning, then validate all bundled files:

```{bash, eval=FALSE}
bash scripts/setup_reference_data.sh
shasum -a 256 -c data/SHA256SUMS
```

Raw sequencing reads, BAM files, complete Kraken classifications, and full HPC
run trees remain outside Git. Newly generated report outputs are written to
ignored timestamped directories under `output/rmd_runs/`; publication-facing
HTML and figure assets are committed under `docs/`.

## Repository layout

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r repo-layout}
layout <- data.frame(
  folder = c("analysis/", "data/metadata/", "data/excluded_loci/",
             "data/raw_read_counts/", "data/derived/", "data/reference/",
             "docs/", "output/rmd_runs/", "scripts/"),
  role = c(
    "workflowR R Markdown pages; new analysis code should live here",
    "sample metadata used to align counts with biological groups",
    "rRNA identifiers removed before normalization and modeling",
    "versioned raw, merged, and frozen analysis count tables",
    "stable inputs promoted from authoritative completed analyses",
    "versioned assembly reports and compressed reference annotation",
    "rendered workflowR website and committed visual assets",
    "ignored fresh output folders created by each Rmd render",
    "reproducible workflow and reference-setup scripts"
  )
)

knitr::kable(layout)
```

## RNA-seq workflow overview

This project starts from gene-level count files that were produced upstream on
the HPC. The diagram below separates the HPC preprocessing steps from the
workflowR analyses in this repository, so it is clear which part maps reads to
the *Schistocerca gregaria* reference and which part tests diet, infection,
and diet-by-infection expression patterns.

```{r hpc-workflow-figure, fig.width=12, fig.height=6}
library(ggplot2)
library(grid)

workflow_nodes <- data.frame(
  id = c(
    "libraries", "fastq_qc", "star_index", "star_align", "align_qc",
    "gene_counts", "merge_counts", "remove_rrna", "count_qc",
    "deseq2", "annotation", "figures"
  ),
  lane = c(rep("HPC preprocessing", 6), rep("workflowR analysis", 6)),
  x = c(1, 2.35, 3.7, 5.05, 6.4, 7.75, 1.6, 3.0, 4.4, 5.8, 7.2, 8.6),
  y = c(rep(2.35, 6), rep(0.85, 6)),
  title = c(
    "RNA libraries", "FASTQ QC", "STAR index", "STAR alignment",
    "Alignment QC", "Gene counts", "Merge counts", "Remove rRNA loci",
    "Count QC", "DESeq2 models", "eggNOG enrichment", "Publication figures"
  ),
  detail = c(
    "fat body; RiboZero", "read quality reports",
    "genome + GFF", "reads to genome",
    "mapping summaries", "transcript + exon primary; exon sensitivity",
    "master count matrix", "gregaria rRNA list",
    "depth, VST, PCA", "diet and infection",
    "GO and KEGG themes", "curated panels"
  ),
  group = c(
    "wet_lab", "hpc", "hpc", "hpc", "hpc", "hpc",
    "data", "data", "analysis", "analysis", "annotation", "figures"
  ),
  stringsAsFactors = FALSE
)

workflow_nodes$label <- paste0(workflow_nodes$title, "\n", workflow_nodes$detail)

workflow_edges <- data.frame(
  from = c(
    "libraries", "fastq_qc", "star_index", "star_align", "align_qc",
    "gene_counts", "merge_counts", "remove_rrna", "count_qc",
    "deseq2", "annotation"
  ),
  to = c(
    "fastq_qc", "star_index", "star_align", "align_qc", "gene_counts",
    "merge_counts", "remove_rrna", "count_qc", "deseq2",
    "annotation", "figures"
  ),
  stringsAsFactors = FALSE
)

edge_df <- merge(workflow_edges, workflow_nodes[, c("id", "x", "y")],
                 by.x = "from", by.y = "id")
edge_df <- merge(edge_df, workflow_nodes[, c("id", "x", "y")],
                 by.x = "to", by.y = "id", suffixes = c("_from", "_to"))

node_width <- 1.06
node_height <- 0.58

p_workflow <- ggplot() +
  annotate("rect", xmin = 0.28, xmax = 8.48, ymin = 1.88, ymax = 2.82,
           fill = "#F5F0FA", color = NA) +
  annotate("rect", xmin = 0.28, xmax = 9.32, ymin = 0.38, ymax = 1.32,
           fill = "#F3F7F2", color = NA) +
  annotate("text", x = 0.35, y = 2.74, label = "HPC preprocessing",
           hjust = 0, size = 4.2, fontface = "bold", color = "#4B3A5A") +
  annotate("text", x = 0.35, y = 1.24, label = "workflowR analysis",
           hjust = 0, size = 4.2, fontface = "bold", color = "#2F4C39") +
  geom_segment(
    data = edge_df,
    aes(x = x_from + node_width / 2, y = y_from,
        xend = x_to - node_width / 2, yend = y_to),
    arrow = arrow(length = unit(0.18, "cm"), type = "closed"),
    linewidth = 0.55,
    lineend = "round",
    color = "#3D4148"
  ) +
  geom_rect(
    data = workflow_nodes,
    aes(xmin = x - node_width / 2, xmax = x + node_width / 2,
        ymin = y - node_height / 2, ymax = y + node_height / 2,
        fill = group),
    color = "white",
    linewidth = 1
  ) +
  geom_text(
    data = workflow_nodes,
    aes(x = x, y = y, label = label),
    size = 3.15,
    lineheight = 0.9,
    color = "#1F2528"
  ) +
  scale_fill_manual(values = c(
    wet_lab = "#F6B26B",
    hpc = "#B998D9",
    data = "#8BBE8B",
    analysis = "#77A7C8",
    annotation = "#E59BB1",
    figures = "#F0C36A"
  )) +
  coord_cartesian(xlim = c(0.15, 9.55), ylim = c(0.2, 3.0), clip = "off") +
  theme_void() +
  theme(legend.position = "none",
        plot.margin = margin(10, 20, 10, 20))

p_workflow
ggsave(file.path(out_dir, "hpc_to_workflowr_workflow.png"),
       p_workflow, width = 12, height = 6, dpi = 300)
```

The important handoff for this repository is the count matrix. STAR produces
the alignments and featureCounts assigns paired fragments with
`-t transcript,exon -g gene_id`. The workflowR pages then merge and check those
counts, remove the loci listed in
`gregaria_rrna_list.txt`, and use the rRNA-filtered matrix for QC,
differential expression, annotation enrichment, and publication figures.

## Rerun commands

Run these commands from the repository root after expanding the bundled GFF.
Each rendered analysis page makes a new timestamped local folder under
`output/rmd_runs/`.

```{bash, eval=FALSE}
# Expand the committed reference annotation after a fresh clone.
bash scripts/setup_reference_data.sh

# Render one page after editing its parameters.
Rscript -e 'workflowr::wflow_build("analysis/03-count-qc-exploration.Rmd")'

# Render the manuscript-oriented pages in dependency order.
Rscript -e 'workflowr::wflow_build(c(
  "analysis/01-study-design-data.Rmd",
  "analysis/02-workflow-reproducibility.Rmd",
  "analysis/03-count-qc-exploration.Rmd",
  "analysis/47-paper-statistical-reference.Rmd",
  "analysis/28-placed-unplaced-scaffold-analysis.Rmd",
  "analysis/16-all-samples-infection-within-diet-deg.Rmd",
  "analysis/07-deg-overlap-prioritization.Rmd",
  "analysis/27-scaffold-origin-audit.Rmd",
  "analysis/10-go-term-enrichment.Rmd",
  "analysis/26-circled-cluster-enrichment.Rmd",
  "analysis/13-physiology-candidate-genes.Rmd",
  "analysis/30-curated-target-gene-figure.Rmd",
  "analysis/29-metarhizium-read-burden.Rmd",
  "analysis/11-outlier-sensitivity.Rmd",
  "analysis/35-count-definition-sensitivity.Rmd",
  "analysis/36-exon-only-deseq2-analysis.Rmd",
  "analysis/42-exon-only-go-term-enrichment.Rmd",
  "analysis/44-exon-only-expression-clusters.Rmd",
  "analysis/45-exon-only-curated-target-genes.Rmd",
  "analysis/46-exon-only-publication-figures.Rmd",
  "analysis/24-all-deg-catalogue.Rmd",
  "analysis/08-publication-figures.Rmd"
))'

# Preview the workflowR status before publishing.
Rscript -e 'workflowr::wflow_status()'
```

## Packages used by the analysis pages

```{r package-check}
packages <- c(
  "workflowr", "knitr", "rmarkdown", "ggplot2", "scales",
  "DESeq2", "pheatmap", "ggrepel", "EnhancedVolcano", "UpSetR"
)

package_status <- data.frame(
  package = packages,
  installed = vapply(packages, requireNamespace, logical(1), quietly = TRUE)
)

knitr::kable(package_status)

write.csv(package_status, file.path(out_dir, "package_status.csv"),
          row.names = FALSE)
```

## Analysis contract

The current analysis pages follow this pattern:

1. Put editable file paths, thresholds, and model choices in the page `params`.
2. Load metadata and counts directly in the Rmd.
3. Check that count columns match metadata before modeling.
4. Write derived tables and figures to a new folder under `output/rmd_runs/`.
5. Save `sessionInfo()` in the same folder as the results.

This keeps the manuscript figures reproducible while separating code and
web-ready figures from large deposited inputs and full result tables.

```{r session-info}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
