---
title: "Count QC and exploration"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  metadata_file: "data/metadata/mehreen_metadata_analysis_44_excluding_1044.txt"
  metadata_full_file: "data/metadata/mehreen_metadata_corrected_45.txt"
  counts_file: "output/rmd_runs/transcript-exon-primary-excluding1044_20260809-233420/host_counts_no_rrna.tsv"
  main_competitive_star_dir: "output/runs/host_pathogen_dual_20260727_000952/02-competitive-star"
  sample_1044_competitive_star_dir: "output/runs/host_pathogen_dual_1044_20260808_110441/02-competitive-star"
  min_total_count: 10
  top_variable_genes: 500
  output_tag: "count-qc-exploration"
---

## Purpose

::: {.walkthrough-grid}
::: {.walkthrough-step}
**Step 1 - Define the cohort**

Retain all valid libraries except 1044, whose alignment profile is documented
below as a supplementary QC exclusion.
:::
::: {.walkthrough-step}
**Step 2 - Explore samples**

Use library depth, PCA, and sample distances to describe variation.
:::
::: {.walkthrough-step}
**Step 3 - Interpret QC**

Identify unusual libraries without excluding them from the main analysis.
:::
:::

This page checks count depth, low-count filtering, sample clustering, and PCA
before differential expression models are interpreted. These checks are not a
separate biological result; they are the audit trail that tells us whether the
diet and infection contrasts are being estimated from comparable samples or
whether a small number of unusual libraries could dominate the model.

## Setup and data loading

The count matrix and metadata are matched by sample label before any plotting.
This small bookkeeping step is important because RNA-seq models assume that the
columns of the count matrix and the rows of the metadata describe the same
samples in the same order.

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c("DESeq2", "ggplot2", "patchwork", "pheatmap", "scales")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

metadata <- read.delim(file.path(project_dir, params$metadata_file),
                       check.names = FALSE)
metadata$label <- as.character(metadata$label)
metadata$Diet <- factor(metadata$Diet, levels = c("33", "50", "83"))
metadata$Treatment <- factor(metadata$Treatment, levels = c("Control", "Infected"))

mapping_metadata <- read.delim(
  file.path(project_dir, params$metadata_full_file),
  check.names = FALSE
)
mapping_metadata$label <- as.character(mapping_metadata$label)
mapping_metadata$Diet <- factor(mapping_metadata$Diet, levels = c("33", "50", "83"))
mapping_metadata$Treatment <- factor(
  mapping_metadata$Treatment,
  levels = c("Control", "Infected")
)

counts_raw <- read.delim(file.path(project_dir, params$counts_file),
                         check.names = FALSE)
counts <- as.matrix(counts_raw[, -1])
rownames(counts) <- counts_raw[[1]]
storage.mode(counts) <- "integer"

count_labels <- sub("_MERGE$", "", sub("^mehreen_", "", colnames(counts)))
metadata <- metadata[match(count_labels, metadata$label), ]
stopifnot(identical(metadata$label, count_labels))
rownames(metadata) <- metadata$label
colnames(counts) <- count_labels
```

## Library size

Library size is the first check because very low-depth samples can have noisy
gene estimates even when their biological label is correct. Here we compare the
total assigned counts across diet and infection groups, looking for samples that
fall far outside the rest of their condition.

```{r library-size}
library(ggplot2)

library_size <- data.frame(
  label = metadata$label,
  total_counts = colSums(counts),
  Diet = metadata$Diet,
  Treatment = metadata$Treatment
)

ggplot(library_size, aes(x = interaction(Diet, Treatment),
                         y = total_counts,
                         color = Treatment)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.12, height = 0, size = 2) +
  scale_y_continuous(labels = scales::comma) +
  theme_classic() +
  labs(
    title = "Total assigned counts per sample",
    x = "Diet and treatment",
    y = "Total counts",
    color = "Treatment"
  )

write.csv(library_size, file.path(out_dir, "library_size.csv"),
          row.names = FALSE)
```

## Mapping quality across diet and infection

The competitive STAR alignment provides a second library-level check that is
independent of the host gene-count totals. Here, "unique mapping" means that a
read mapped to one position in the combined locust--*Metarhizium* reference. It
does not mean that the read mapped specifically to the fungus. STAR's "too
short" category means that too little of the read aligned with sufficient
score; it does not necessarily mean that the original FASTQ read was physically
short.

The first figure compares two definitions in vertically stacked panels. Panel A uses only reads
with one accepted genomic location. Panel B adds reads that STAR aligned to 2--20
locations. Reads assigned to more than 20 loci remain excluded from Panel B.
The following four-panel graph separates the main STAR outcomes. Every library
is shown in this supplementary QC record, and 1044 is labeled because its
mapping profile is visibly different from the remaining samples. Sample 1044 is
excluded from the expression analyses that follow.

```{r mapping-quality-by-design, fig.width=11, fig.height=14}
star_dirs <- file.path(
  project_dir,
  c(params$main_competitive_star_dir,
    params$sample_1044_competitive_star_dir)
)
star_logs <- unlist(lapply(
  star_dirs,
  list.files,
  pattern = "_Log.final.out$",
  full.names = TRUE
))

read_star_metric <- function(path, metric) {
  lines <- readLines(path, warn = FALSE)
  matched <- lines[grepl(metric, lines, fixed = TRUE)]
  if (length(matched) != 1) {
    stop("Expected one STAR metric '", metric, "' in ", path)
  }
  value <- trimws(strsplit(matched, "|", fixed = TRUE)[[1]][2])
  as.numeric(gsub("%", "", value, fixed = TRUE))
}

mapping_qc <- data.frame(
  label = sub("_Log.final.out$", "", basename(star_logs)),
  input_reads = vapply(
    star_logs, read_star_metric, numeric(1), metric = "Number of input reads"
  ),
  unique_pct = vapply(
    star_logs, read_star_metric, numeric(1), metric = "Uniquely mapped reads %"
  ),
  multimapped_pct = vapply(
    star_logs, read_star_metric, numeric(1),
    metric = "% of reads mapped to multiple loci"
  ),
  unmapped_short_pct = vapply(
    star_logs, read_star_metric, numeric(1),
    metric = "% of reads unmapped: too short"
  ),
  unmapped_other_pct = vapply(
    star_logs, read_star_metric, numeric(1),
    metric = "% of reads unmapped: other"
  ),
  stringsAsFactors = FALSE
)

stopifnot(setequal(mapping_qc$label, mapping_metadata$label))
mapping_qc <- mapping_qc[match(mapping_metadata$label, mapping_qc$label), ]
mapping_qc$Diet <- mapping_metadata$Diet
mapping_qc$Treatment <- mapping_metadata$Treatment
mapping_qc$unique_plus_multimapped_pct <-
  mapping_qc$unique_pct + mapping_qc$multimapped_pct

test_mapping_design <- function(data, metric, mapping_definition, scenario) {
  data$mapping_logit <- qlogis(
    pmin(pmax(data[[metric]] / 100, 1e-6), 1 - 1e-6)
  )

  interaction_model <- stats::lm(mapping_logit ~ Diet * Treatment, data = data)
  interaction_test <- stats::drop1(
    interaction_model,
    scope = ~ Diet:Treatment,
    test = "F"
  )
  additive_model <- stats::lm(mapping_logit ~ Diet + Treatment, data = data)
  main_tests <- stats::drop1(additive_model, test = "F")

  data.frame(
    mapping_definition = mapping_definition,
    scenario = scenario,
    samples = nrow(data),
    interaction_p = interaction_test["Diet:Treatment", "Pr(>F)"],
    diet_p = main_tests["Diet", "Pr(>F)"],
    infection_p = main_tests["Treatment", "Pr(>F)"],
    stringsAsFactors = FALSE
  )
}

format_test_line <- function(test_row) {
  paste0(
    "Factorial models: diet x infection p = ",
    format.pval(test_row$interaction_p, digits = 3, eps = 0.001),
    "; diet p = ", format.pval(test_row$diet_p, digits = 3, eps = 0.001),
    "; infection p = ", format.pval(test_row$infection_p, digits = 3, eps = 0.001)
  )
}

mapping_figure_tests <- rbind(
  test_mapping_design(mapping_qc, "unique_pct", "Unique mapping", "All 45 libraries"),
  test_mapping_design(
    mapping_qc,
    "unique_plus_multimapped_pct",
    "Unique + multimapping",
    "All 45 libraries"
  )
)

treatment_colors <- c(Control = "#FAEBD7", Infected = "#8B0000")
point_position <- ggplot2::position_jitterdodge(
  jitter.width = 0.10,
  dodge.width = 0.72,
  seed = 33
)

make_mapping_design_plot <- function(data, metric, title, y_label, test_row) {
  plot_data <- data
  plot_data$mapping_percentage <- plot_data[[metric]]

  ggplot2::ggplot(
    plot_data,
    ggplot2::aes(x = Diet, y = mapping_percentage, fill = Treatment)
  ) +
    ggplot2::geom_boxplot(
      position = ggplot2::position_dodge(width = 0.72),
      width = 0.62,
      outlier.shape = NA,
      color = "black"
    ) +
    ggplot2::geom_point(
      shape = 21,
      size = 3.2,
      position = point_position,
      color = "black"
    ) +
    ggplot2::geom_text(
      data = plot_data[plot_data$label == "1044", ],
      ggplot2::aes(label = label),
      position = ggplot2::position_nudge(x = 0.20, y = 0.8),
      size = 4.2,
      fontface = "bold",
      inherit.aes = TRUE
    ) +
    ggplot2::scale_fill_manual(values = treatment_colors) +
    ggplot2::scale_y_continuous(
      limits = c(0, NA),
      expand = ggplot2::expansion(mult = c(0, 0.08))
    ) +
    ggplot2::theme_classic(base_size = 15) +
    ggplot2::theme(
      legend.position = "bottom",
      plot.title = ggplot2::element_text(face = "bold")
    ) +
    ggplot2::labs(
      title = title,
      subtitle = format_test_line(test_row),
      x = "Diet",
      y = y_label,
      fill = "Treatment"
    )
}

p_mapping_unique <- make_mapping_design_plot(
  mapping_qc,
  "unique_pct",
  "Unique mapping",
  "Uniquely mapped reads (%)",
  mapping_figure_tests[1, ]
)

p_mapping_unique_plus_multi <- make_mapping_design_plot(
  mapping_qc,
  "unique_plus_multimapped_pct",
  "Unique + multimapping",
  "Reads mapped to 1--20 loci (%)",
  mapping_figure_tests[2, ]
)

p_mapping_comparison <-
  p_mapping_unique / p_mapping_unique_plus_multi +
  patchwork::plot_layout(guides = "collect") +
  patchwork::plot_annotation(
    tag_levels = "A",
    title = "Competitive STAR mapping by diet and infection",
    subtitle = "All 45 libraries shown for QC; 1044 is excluded from downstream expression analyses",
    theme = ggplot2::theme(
      plot.title = ggplot2::element_text(size = 20, face = "bold"),
      plot.subtitle = ggplot2::element_text(size = 15)
    )
  ) &
  ggplot2::theme(
    legend.position = "bottom",
    plot.tag = ggplot2::element_text(size = 19, face = "bold")
  )

print(p_mapping_comparison)

mapping_plot_exports <- list(
  competitive_unique_mapping_by_design = p_mapping_unique,
  competitive_unique_plus_multimapping_by_design = p_mapping_unique_plus_multi,
  competitive_unique_vs_multimapping_by_design = p_mapping_comparison,
  figureS3_mapping_quality_exclusion_1044 = p_mapping_comparison
)
mapping_plot_dimensions <- list(
  competitive_unique_mapping_by_design = c(11, 6),
  competitive_unique_plus_multimapping_by_design = c(11, 6),
  competitive_unique_vs_multimapping_by_design = c(11, 14),
  figureS3_mapping_quality_exclusion_1044 = c(11, 14)
)

for (plot_name in names(mapping_plot_exports)) {
  dimensions <- mapping_plot_dimensions[[plot_name]]
  for (extension in c("png", "pdf", "svg")) {
  ggplot2::ggsave(
      file.path(out_dir, paste0(plot_name, ".", extension)),
      mapping_plot_exports[[plot_name]],
      width = dimensions[1],
      height = dimensions[2],
      dpi = 320
  )
  }
}
```

```{r mapping-outcomes-by-design, fig.width=12, fig.height=9}
mapping_metrics <- c(
  unique_pct = "Unique mapping",
  multimapped_pct = "Multiple loci",
  unmapped_short_pct = "Unmapped: too short",
  unmapped_other_pct = "Unmapped: other"
)

mapping_long <- do.call(rbind, lapply(names(mapping_metrics), function(metric) {
  data.frame(
    label = mapping_qc$label,
    Diet = mapping_qc$Diet,
    Treatment = mapping_qc$Treatment,
    metric = mapping_metrics[[metric]],
    percentage = mapping_qc[[metric]],
    stringsAsFactors = FALSE
  )
}))
mapping_long$metric <- factor(mapping_long$metric,
                              levels = unname(mapping_metrics))

p_mapping_outcomes <- ggplot2::ggplot(
  mapping_long,
  ggplot2::aes(x = Diet, y = percentage, fill = Treatment)
) +
  ggplot2::geom_boxplot(
    position = ggplot2::position_dodge(width = 0.72),
    width = 0.62,
    outlier.shape = NA,
    color = "black"
  ) +
  ggplot2::geom_point(
    shape = 21,
    size = 2.7,
    position = point_position,
    color = "black"
  ) +
  ggplot2::geom_text(
    data = mapping_long[mapping_long$label == "1044", ],
    ggplot2::aes(label = label),
    position = ggplot2::position_nudge(x = 0.20, y = 1.2),
    size = 3.6,
    fontface = "bold"
  ) +
  ggplot2::facet_wrap(~ metric, ncol = 2, scales = "free_y") +
  ggplot2::scale_fill_manual(values = treatment_colors) +
  ggplot2::theme_classic(base_size = 14) +
  ggplot2::theme(
    legend.position = "bottom",
    strip.background = ggplot2::element_rect(fill = "grey94", color = "black"),
    strip.text = ggplot2::element_text(face = "bold", size = 13)
  ) +
  ggplot2::labs(
    title = "STAR mapping outcomes across the experimental design",
    x = "Diet",
    y = "Reads (%)",
    fill = "Treatment"
  )

print(p_mapping_outcomes)

for (extension in c("png", "pdf", "svg")) {
  ggplot2::ggsave(
    file.path(out_dir, paste0("competitive_mapping_outcomes_by_design.", extension)),
    p_mapping_outcomes,
    width = 12,
    height = 9,
    dpi = 320
  )
}

write.csv(mapping_qc,
          file.path(out_dir, "competitive_mapping_quality_all_45_samples.csv"),
          row.names = FALSE)
```

### Does mapping quality differ with diet or infection?

Because percentages are bounded between 0 and 100, each mapping proportion is
logit-transformed before fitting the factorial model. Unique mapping and unique
plus multimapping are tested separately. We first test the diet-by-infection
interaction. Since neither interaction is supported, the diet and infection
terms are then tested in additive models. The all-library rows provide the
formal tests printed on the supplementary figure. The rows without 1044
describe the final biological analysis cohort.

```{r mapping-quality-model}
mapping_definitions <- c(
  unique_pct = "Unique mapping",
  unique_plus_multimapped_pct = "Unique + multimapping"
)

mapping_model_results <- do.call(rbind, lapply(
  names(mapping_definitions),
  function(metric) {
    rbind(
      test_mapping_design(
        mapping_qc,
        metric,
        mapping_definitions[[metric]],
        "All 45 libraries shown for QC"
      ),
      test_mapping_design(
        mapping_qc[mapping_qc$label != "1044", ],
        metric,
        mapping_definitions[[metric]],
        "Final 44-sample analysis cohort"
      )
    )
  }
))

mapping_summary_long <- do.call(rbind, lapply(
  names(mapping_definitions),
  function(metric) {
    data.frame(
      mapping_definition = mapping_definitions[[metric]],
      Diet = mapping_qc$Diet,
      Treatment = mapping_qc$Treatment,
      percentage = mapping_qc[[metric]]
    )
  }
))

mapping_group_summary <- aggregate(
  percentage ~ mapping_definition + Diet + Treatment,
  data = mapping_summary_long,
  FUN = function(x) c(mean = mean(x), sd = stats::sd(x), n = length(x))
)
mapping_group_summary <- data.frame(
  mapping_definition = mapping_group_summary$mapping_definition,
  Diet = mapping_group_summary$Diet,
  Treatment = mapping_group_summary$Treatment,
  mean_mapping_pct = mapping_group_summary$percentage[, "mean"],
  sd_mapping_pct = mapping_group_summary$percentage[, "sd"],
  samples = mapping_group_summary$percentage[, "n"]
)

knitr::kable(mapping_model_results, digits = 4,
             caption = "Factorial tests for both competitive mapping definitions")
knitr::kable(mapping_group_summary, digits = 2,
             caption = "Mapping percentage by experimental condition")

write.csv(mapping_model_results,
          file.path(out_dir, "competitive_mapping_design_tests.csv"),
          row.names = FALSE)
write.csv(mapping_group_summary,
          file.path(out_dir, "competitive_mapping_group_summary.csv"),
          row.names = FALSE)
```

For unique mapping alone, there is no supported diet-by-infection interaction
(`r sprintf("p = %.3f", mapping_model_results$interaction_p[1])`). Diet is
borderline (`r sprintf("p = %.3f", mapping_model_results$diet_p[1])`), whereas
infected libraries have a lower unique-mapping rate overall
(`r sprintf("p = %.3f", mapping_model_results$infection_p[1])`). When accepted
multimapping reads are added, neither the interaction
(`r sprintf("p = %.3f", mapping_model_results$interaction_p[3])`), diet
(`r sprintf("p = %.3f", mapping_model_results$diet_p[3])`), nor infection
(`r sprintf("p = %.3f", mapping_model_results$infection_p[3])`) crosses the
0.05 threshold in the all-sample model. This is a QC association and should not
be interpreted as proof that infection caused poor mapping: genome duplication,
fungal burden, library composition, and other technical differences can all
alter the proportion of reads accepted by the reference alignment.

## Low-count filtering

Genes with almost no reads across the whole experiment add multiple-testing
burden but usually cannot be estimated reliably. The filtering threshold keeps
genes with at least `r params$min_total_count` total counts, which is a light
screen intended to remove uninformative rows while preserving diet- and
infection-responsive genes for DESeq2.

```{r filtering}
keep <- rowSums(counts) >= params$min_total_count
filter_summary <- data.frame(
  min_total_count = params$min_total_count,
  genes_before = nrow(counts),
  genes_after = sum(keep),
  genes_removed = sum(!keep)
)

knitr::kable(filter_summary)
write.csv(filter_summary, file.path(out_dir, "filter_summary.csv"),
          row.names = FALSE)
```

## PCA

The PCA is calculated after a variance-stabilizing transformation (VST). VST
puts low-count and high-count genes onto a more comparable scale, so the plot is
useful for sample-level structure rather than being driven mostly by the most
abundant genes. We expect samples to show some organization by diet and/or
infection, but we also use this plot to identify samples that sit apart from
their expected group.

```{r pca}
dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = counts[keep, ],
  colData = metadata,
  design = ~ Diet + Treatment
)

vsd <- DESeq2::vst(dds, blind = TRUE)
pca_df <- DESeq2::plotPCA(vsd, intgroup = c("Diet", "Treatment"),
                          returnData = TRUE)
percent_var <- round(100 * attr(pca_df, "percentVar"))

p_pca <- ggplot(pca_df, aes(PC1, PC2, color = Diet, shape = Treatment)) +
  geom_point(size = 3) +
  theme_classic() +
  labs(
    title = "Variance-stabilized PCA",
    x = paste0("PC1: ", percent_var[1], "% variance"),
    y = paste0("PC2: ", percent_var[2], "% variance")
  )

p_pca
ggsave(file.path(out_dir, "pca_vst.png"), p_pca, width = 7, height = 5, dpi = 300)
```

## Sample distance heatmap

The sample distance heatmap asks the same QC question as the PCA in a different
format: which libraries are most similar after VST normalization? Samples from
the same diet and treatment should usually be closer to one another than to
unrelated groups. A sample that clusters away from its condition is a candidate
outlier and should be checked before it is allowed to drive differential
expression calls.

```{r sample-distance}
sample_dist <- dist(t(SummarizedExperiment::assay(vsd)))
sample_dist_matrix <- as.matrix(sample_dist)
rownames(sample_dist_matrix) <- paste(metadata$Diet, metadata$Treatment,
                                      metadata$label, sep = "_")
colnames(sample_dist_matrix) <- rownames(sample_dist_matrix)

png(file.path(out_dir, "sample_distance_heatmap.png"),
    width = 1600, height = 1400, res = 180)
pheatmap::pheatmap(sample_dist_matrix,
                   main = "Sample-to-sample distance")
dev.off()

pheatmap::pheatmap(sample_dist_matrix,
                   main = "Sample-to-sample distance")
```

## Most variable genes

This heatmap shows the `r params$top_variable_genes` genes with the strongest
sample-to-sample variation after VST transformation. It is not a DEG test. It is
a compact view of whether the largest expression patterns line up with the
experimental design and whether any sample carries a distinct expression profile
that might explain an outlier flag.

```{r variable-genes}
vst_mat <- SummarizedExperiment::assay(vsd)
gene_variance <- apply(vst_mat, 1, var)
top_n <- min(params$top_variable_genes, length(gene_variance))
top_genes <- names(sort(gene_variance, decreasing = TRUE))[seq_len(top_n)]

annotation_col <- data.frame(Diet = metadata$Diet,
                             Treatment = metadata$Treatment)
rownames(annotation_col) <- colnames(vst_mat)

png(file.path(out_dir, "top_variable_gene_heatmap.png"),
    width = 1600, height = 1600, res = 180)
pheatmap::pheatmap(vst_mat[top_genes, ],
                   show_rownames = FALSE,
                   annotation_col = annotation_col,
                   main = paste("Top", top_n, "variable genes"))
dev.off()

pheatmap::pheatmap(vst_mat[top_genes, ],
                   show_rownames = FALSE,
                   annotation_col = annotation_col,
                   main = paste("Top", top_n, "variable genes"))
```

```{r session-info}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
