---
title: "Count QC and exploration"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  metadata_file: "data/metadata/mehreen_metadata_analysis_44_excluding_1044.txt"
  metadata_full_file: "data/metadata/mehreen_metadata_corrected_45.txt"
  mass_file: "data/metadata/mehreen_predissection_mass_45.tsv"
  counts_file: "output/rmd_runs/transcript-exon-primary-excluding1044_20260809-233420/host_counts_no_rrna.tsv"
  main_competitive_star_dir: "output/runs/host_pathogen_dual_20260727_000952/02-competitive-star"
  sample_1044_competitive_star_dir: "output/runs/host_pathogen_dual_1044_20260808_110441/02-competitive-star"
  min_total_count: 10
  top_variable_genes: 500
  output_tag: "count-qc-exploration"
---

## Purpose

::: {.walkthrough-grid}
::: {.walkthrough-step}
**Step 1 - Define the cohort**

Retain all valid libraries except 1044, whose alignment profile is documented
below as a supplementary QC exclusion.
:::
::: {.walkthrough-step}
**Step 2 - Explore samples**

Use library depth, PCA, and sample distances to describe variation.
:::
::: {.walkthrough-step}
**Step 3 - Interpret QC**

Identify unusual libraries without excluding them from the main analysis.
:::
:::

This page checks count depth, low-count filtering, sample clustering, and PCA
before differential expression models are interpreted. These checks are not a
separate biological result; they are the audit trail that tells us whether the
diet and infection contrasts are being estimated from comparable samples or
whether a small number of unusual libraries could dominate the model.

## Setup and data loading

The count matrix and metadata are matched by sample label before any plotting.
This small bookkeeping step is important because RNA-seq models assume that the
columns of the count matrix and the rows of the metadata describe the same
samples in the same order.

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c("DESeq2", "ggplot2", "ggrepel", "patchwork", "pheatmap", "scales")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

metadata <- read.delim(file.path(project_dir, params$metadata_file),
                       check.names = FALSE)
metadata$label <- as.character(metadata$label)
metadata$Diet <- factor(metadata$Diet, levels = c("33", "50", "83"))
metadata$Treatment <- factor(metadata$Treatment, levels = c("Control", "Infected"))

mapping_metadata <- read.delim(
  file.path(project_dir, params$metadata_full_file),
  check.names = FALSE
)
mapping_metadata$label <- as.character(mapping_metadata$label)
mapping_metadata$Diet <- factor(mapping_metadata$Diet, levels = c("33", "50", "83"))
mapping_metadata$Treatment <- factor(
  mapping_metadata$Treatment,
  levels = c("Control", "Infected")
)

mass_metadata <- read.delim(
  file.path(project_dir, params$mass_file),
  check.names = FALSE,
  colClasses = c(label = "character", mass_g = "numeric")
)
stopifnot(
  nrow(mass_metadata) == 45L,
  !anyDuplicated(mass_metadata$label),
  !anyNA(mass_metadata$mass_g),
  setequal(mass_metadata$label, mapping_metadata$label)
)

counts_raw <- read.delim(file.path(project_dir, params$counts_file),
                         check.names = FALSE)
counts <- as.matrix(counts_raw[, -1])
rownames(counts) <- counts_raw[[1]]
storage.mode(counts) <- "integer"

count_labels <- sub("_MERGE$", "", sub("^mehreen_", "", colnames(counts)))
metadata <- metadata[match(count_labels, metadata$label), ]
stopifnot(identical(metadata$label, count_labels))
rownames(metadata) <- metadata$label
colnames(counts) <- count_labels
```

## Library size

Library size is the first check because very low-depth samples can have noisy
gene estimates even when their biological label is correct. Here we compare the
total assigned counts across diet and infection groups, looking for samples that
fall far outside the rest of their condition.

```{r library-size}
library(ggplot2)

library_size <- data.frame(
  label = metadata$label,
  total_counts = colSums(counts),
  Diet = metadata$Diet,
  Treatment = metadata$Treatment
)

ggplot(library_size, aes(x = interaction(Diet, Treatment),
                         y = total_counts,
                         color = Treatment)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.12, height = 0, size = 2) +
  scale_y_continuous(labels = scales::comma) +
  theme_classic() +
  labs(
    title = "Total assigned counts per sample",
    x = "Diet and treatment",
    y = "Total counts",
    color = "Treatment"
  )

write.csv(library_size, file.path(out_dir, "library_size.csv"),
          row.names = FALSE)
```

## Mapping quality across diet and infection

The competitive STAR alignment provides a second library-level check that is
independent of the host gene-count totals. Here, "unique mapping" means that a
read mapped to one position in the combined locust--*Metarhizium* reference. It
does not mean that the read mapped specifically to the fungus. STAR's "too
short" category means that too little of the read aligned with sufficient
score; it does not necessarily mean that the original FASTQ read was physically
short.

The first figure compares two definitions in vertically stacked panels. Panel A uses only reads
with one accepted genomic location. Panel B adds reads that STAR aligned to 2--20
locations. Reads assigned to more than 20 loci remain excluded from Panel B.
The following four-panel graph separates the main STAR outcomes. Every library
is shown in this supplementary QC record, and 1044 is labeled because its
mapping profile is visibly different from the remaining samples. Sample 1044 is
excluded from the expression analyses that follow.

```{r mapping-quality-by-design, fig.width=11, fig.height=14}
star_dirs <- file.path(
  project_dir,
  c(params$main_competitive_star_dir,
    params$sample_1044_competitive_star_dir)
)
star_logs <- unlist(lapply(
  star_dirs,
  list.files,
  pattern = "_Log.final.out$",
  full.names = TRUE
))

read_star_metric <- function(path, metric) {
  lines <- readLines(path, warn = FALSE)
  matched <- lines[grepl(metric, lines, fixed = TRUE)]
  if (length(matched) != 1) {
    stop("Expected one STAR metric '", metric, "' in ", path)
  }
  value <- trimws(strsplit(matched, "|", fixed = TRUE)[[1]][2])
  as.numeric(gsub("%", "", value, fixed = TRUE))
}

mapping_qc <- data.frame(
  label = sub("_Log.final.out$", "", basename(star_logs)),
  input_reads = vapply(
    star_logs, read_star_metric, numeric(1), metric = "Number of input reads"
  ),
  unique_pct = vapply(
    star_logs, read_star_metric, numeric(1), metric = "Uniquely mapped reads %"
  ),
  multimapped_pct = vapply(
    star_logs, read_star_metric, numeric(1),
    metric = "% of reads mapped to multiple loci"
  ),
  unmapped_short_pct = vapply(
    star_logs, read_star_metric, numeric(1),
    metric = "% of reads unmapped: too short"
  ),
  unmapped_other_pct = vapply(
    star_logs, read_star_metric, numeric(1),
    metric = "% of reads unmapped: other"
  ),
  stringsAsFactors = FALSE
)

stopifnot(setequal(mapping_qc$label, mapping_metadata$label))
mapping_qc <- mapping_qc[match(mapping_metadata$label, mapping_qc$label), ]
mapping_qc$Diet <- mapping_metadata$Diet
mapping_qc$Treatment <- mapping_metadata$Treatment
mapping_qc$unique_plus_multimapped_pct <-
  mapping_qc$unique_pct + mapping_qc$multimapped_pct

test_mapping_design <- function(data, metric, mapping_definition, scenario) {
  data$mapping_logit <- qlogis(
    pmin(pmax(data[[metric]] / 100, 1e-6), 1 - 1e-6)
  )

  interaction_model <- stats::lm(mapping_logit ~ Diet * Treatment, data = data)
  interaction_test <- stats::drop1(
    interaction_model,
    scope = ~ Diet:Treatment,
    test = "F"
  )
  additive_model <- stats::lm(mapping_logit ~ Diet + Treatment, data = data)
  main_tests <- stats::drop1(additive_model, test = "F")

  data.frame(
    mapping_definition = mapping_definition,
    scenario = scenario,
    samples = nrow(data),
    interaction_p = interaction_test["Diet:Treatment", "Pr(>F)"],
    diet_p = main_tests["Diet", "Pr(>F)"],
    infection_p = main_tests["Treatment", "Pr(>F)"],
    stringsAsFactors = FALSE
  )
}

format_test_line <- function(test_row) {
  paste0(
    "Factorial models: diet x infection p = ",
    format.pval(test_row$interaction_p, digits = 3, eps = 0.001),
    "; diet p = ", format.pval(test_row$diet_p, digits = 3, eps = 0.001),
    "; infection p = ", format.pval(test_row$infection_p, digits = 3, eps = 0.001)
  )
}

mapping_figure_tests <- rbind(
  test_mapping_design(mapping_qc, "unique_pct", "Unique mapping", "All 45 libraries"),
  test_mapping_design(
    mapping_qc,
    "unique_plus_multimapped_pct",
    "Unique + multimapping",
    "All 45 libraries"
  )
)

treatment_colors <- c(Control = "#FAEBD7", Infected = "#8B0000")
point_position <- ggplot2::position_jitterdodge(
  jitter.width = 0.10,
  dodge.width = 0.72,
  seed = 33
)

make_mapping_design_plot <- function(data, metric, title, y_label, test_row) {
  plot_data <- data
  plot_data$mapping_percentage <- plot_data[[metric]]

  ggplot2::ggplot(
    plot_data,
    ggplot2::aes(x = Diet, y = mapping_percentage, fill = Treatment)
  ) +
    ggplot2::geom_boxplot(
      position = ggplot2::position_dodge(width = 0.72),
      width = 0.62,
      outlier.shape = NA,
      color = "black"
    ) +
    ggplot2::geom_point(
      shape = 21,
      size = 3.2,
      position = point_position,
      color = "black"
    ) +
    ggplot2::geom_text(
      data = plot_data[plot_data$label == "1044", ],
      ggplot2::aes(label = label),
      position = ggplot2::position_nudge(x = 0.20, y = 0.8),
      size = 4.2,
      fontface = "bold",
      inherit.aes = TRUE
    ) +
    ggplot2::scale_fill_manual(values = treatment_colors) +
    ggplot2::scale_y_continuous(
      limits = c(0, NA),
      expand = ggplot2::expansion(mult = c(0, 0.08))
    ) +
    ggplot2::theme_classic(base_size = 15) +
    ggplot2::theme(
      legend.position = "bottom",
      plot.title = ggplot2::element_text(face = "bold")
    ) +
    ggplot2::labs(
      title = title,
      subtitle = format_test_line(test_row),
      x = "Diet",
      y = y_label,
      fill = "Treatment"
    )
}

p_mapping_unique <- make_mapping_design_plot(
  mapping_qc,
  "unique_pct",
  "Unique mapping",
  "Uniquely mapped reads (%)",
  mapping_figure_tests[1, ]
)

p_mapping_unique_plus_multi <- make_mapping_design_plot(
  mapping_qc,
  "unique_plus_multimapped_pct",
  "Unique + multimapping",
  "Reads mapped to 1--20 loci (%)",
  mapping_figure_tests[2, ]
)

p_mapping_comparison <-
  p_mapping_unique / p_mapping_unique_plus_multi +
  patchwork::plot_layout(guides = "collect") +
  patchwork::plot_annotation(
    tag_levels = "A",
    title = "Competitive STAR mapping by diet and infection",
    subtitle = "All 45 libraries shown for QC; 1044 is excluded from downstream expression analyses",
    theme = ggplot2::theme(
      plot.title = ggplot2::element_text(size = 20, face = "bold"),
      plot.subtitle = ggplot2::element_text(size = 15)
    )
  ) &
  ggplot2::theme(
    legend.position = "bottom",
    plot.tag = ggplot2::element_text(size = 19, face = "bold")
  )

print(p_mapping_comparison)

mapping_plot_exports <- list(
  competitive_unique_mapping_by_design = p_mapping_unique,
  competitive_unique_plus_multimapping_by_design = p_mapping_unique_plus_multi,
  competitive_unique_vs_multimapping_by_design = p_mapping_comparison,
  figureS3_mapping_quality_exclusion_1044 = p_mapping_comparison
)
mapping_plot_dimensions <- list(
  competitive_unique_mapping_by_design = c(11, 6),
  competitive_unique_plus_multimapping_by_design = c(11, 6),
  competitive_unique_vs_multimapping_by_design = c(11, 14),
  figureS3_mapping_quality_exclusion_1044 = c(11, 14)
)

for (plot_name in names(mapping_plot_exports)) {
  dimensions <- mapping_plot_dimensions[[plot_name]]
  for (extension in c("png", "pdf", "svg")) {
  ggplot2::ggsave(
      file.path(out_dir, paste0(plot_name, ".", extension)),
      mapping_plot_exports[[plot_name]],
      width = dimensions[1],
      height = dimensions[2],
      dpi = 320
  )
  }
}
```

```{r mapping-outcomes-by-design, fig.width=12, fig.height=9}
mapping_metrics <- c(
  unique_pct = "Unique mapping",
  multimapped_pct = "Multiple loci",
  unmapped_short_pct = "Unmapped: too short",
  unmapped_other_pct = "Unmapped: other"
)

mapping_long <- do.call(rbind, lapply(names(mapping_metrics), function(metric) {
  data.frame(
    label = mapping_qc$label,
    Diet = mapping_qc$Diet,
    Treatment = mapping_qc$Treatment,
    metric = mapping_metrics[[metric]],
    percentage = mapping_qc[[metric]],
    stringsAsFactors = FALSE
  )
}))
mapping_long$metric <- factor(mapping_long$metric,
                              levels = unname(mapping_metrics))

p_mapping_outcomes <- ggplot2::ggplot(
  mapping_long,
  ggplot2::aes(x = Diet, y = percentage, fill = Treatment)
) +
  ggplot2::geom_boxplot(
    position = ggplot2::position_dodge(width = 0.72),
    width = 0.62,
    outlier.shape = NA,
    color = "black"
  ) +
  ggplot2::geom_point(
    shape = 21,
    size = 2.7,
    position = point_position,
    color = "black"
  ) +
  ggplot2::geom_text(
    data = mapping_long[mapping_long$label == "1044", ],
    ggplot2::aes(label = label),
    position = ggplot2::position_nudge(x = 0.20, y = 1.2),
    size = 3.6,
    fontface = "bold"
  ) +
  ggplot2::facet_wrap(~ metric, ncol = 2, scales = "free_y") +
  ggplot2::scale_fill_manual(values = treatment_colors) +
  ggplot2::theme_classic(base_size = 14) +
  ggplot2::theme(
    legend.position = "bottom",
    strip.background = ggplot2::element_rect(fill = "grey94", color = "black"),
    strip.text = ggplot2::element_text(face = "bold", size = 13)
  ) +
  ggplot2::labs(
    title = "STAR mapping outcomes across the experimental design",
    x = "Diet",
    y = "Reads (%)",
    fill = "Treatment"
  )

print(p_mapping_outcomes)

for (extension in c("png", "pdf", "svg")) {
  ggplot2::ggsave(
    file.path(out_dir, paste0("competitive_mapping_outcomes_by_design.", extension)),
    p_mapping_outcomes,
    width = 12,
    height = 9,
    dpi = 320
  )
}

write.csv(mapping_qc,
          file.path(out_dir, "competitive_mapping_quality_all_45_samples.csv"),
          row.names = FALSE)
```

### Does mapping quality differ with diet or infection?

Because percentages are bounded between 0 and 100, each mapping proportion is
logit-transformed before fitting the factorial model. Unique mapping and unique
plus multimapping are tested separately. We first test the diet-by-infection
interaction. Since neither interaction is supported, the diet and infection
terms are then tested in additive models. The all-library rows provide the
formal tests printed on the supplementary figure. The rows without 1044
describe the final biological analysis cohort.

```{r mapping-quality-model}
mapping_definitions <- c(
  unique_pct = "Unique mapping",
  unique_plus_multimapped_pct = "Unique + multimapping"
)

mapping_model_results <- do.call(rbind, lapply(
  names(mapping_definitions),
  function(metric) {
    rbind(
      test_mapping_design(
        mapping_qc,
        metric,
        mapping_definitions[[metric]],
        "All 45 libraries shown for QC"
      ),
      test_mapping_design(
        mapping_qc[mapping_qc$label != "1044", ],
        metric,
        mapping_definitions[[metric]],
        "Final 44-sample analysis cohort"
      )
    )
  }
))

mapping_summary_long <- do.call(rbind, lapply(
  names(mapping_definitions),
  function(metric) {
    data.frame(
      mapping_definition = mapping_definitions[[metric]],
      Diet = mapping_qc$Diet,
      Treatment = mapping_qc$Treatment,
      percentage = mapping_qc[[metric]]
    )
  }
))

mapping_group_summary <- aggregate(
  percentage ~ mapping_definition + Diet + Treatment,
  data = mapping_summary_long,
  FUN = function(x) c(mean = mean(x), sd = stats::sd(x), n = length(x))
)
mapping_group_summary <- data.frame(
  mapping_definition = mapping_group_summary$mapping_definition,
  Diet = mapping_group_summary$Diet,
  Treatment = mapping_group_summary$Treatment,
  mean_mapping_pct = mapping_group_summary$percentage[, "mean"],
  sd_mapping_pct = mapping_group_summary$percentage[, "sd"],
  samples = mapping_group_summary$percentage[, "n"]
)

knitr::kable(mapping_model_results, digits = 4,
             caption = "Factorial tests for both competitive mapping definitions")
knitr::kable(mapping_group_summary, digits = 2,
             caption = "Mapping percentage by experimental condition")

write.csv(mapping_model_results,
          file.path(out_dir, "competitive_mapping_design_tests.csv"),
          row.names = FALSE)
write.csv(mapping_group_summary,
          file.path(out_dir, "competitive_mapping_group_summary.csv"),
          row.names = FALSE)
```

For unique mapping alone, there is no supported diet-by-infection interaction
(`r sprintf("p = %.3f", mapping_model_results$interaction_p[1])`). Diet is
borderline (`r sprintf("p = %.3f", mapping_model_results$diet_p[1])`), whereas
infected libraries have a lower unique-mapping rate overall
(`r sprintf("p = %.3f", mapping_model_results$infection_p[1])`). When accepted
multimapping reads are added, neither the interaction
(`r sprintf("p = %.3f", mapping_model_results$interaction_p[3])`), diet
(`r sprintf("p = %.3f", mapping_model_results$diet_p[3])`), nor infection
(`r sprintf("p = %.3f", mapping_model_results$infection_p[3])`) crosses the
0.05 threshold in the all-sample model. This is a QC association and should not
be interpreted as proof that infection caused poor mapping: genome duplication,
fungal burden, library composition, and other technical differences can all
alter the proportion of reads accepted by the reference alignment.

## Exploratory mapping-quality and mass association

The source datasheet records one mass measurement for every sequenced
individual under the column `Mass before extraction`. We use that measurement
as the pre-dissection individual-mass covariate requested here. This exploratory analysis asks
whether either competitive mapping definition varies systematically with mass.
Each point is one library and every sample identifier is shown. The three diet
panels each contain a separate regression for controls (dashed) and infected
locusts (solid). All six fits use the final 44-library cohort; library 1044 is
shown for quality-control context but does not influence a fitted line.
Spearman correlations describe the overall unadjusted monotonic association,
while a logit-scale linear model tests mass after accounting for diet and
infection. The within-group regressions are exploratory because the six groups
contain only 5--10 individuals; their slopes, R-squared values and P values are
exported with the figure.

```{r mapping-quality-by-mass, fig.width=11, fig.height=15}
mapping_mass <- merge(
  mapping_qc,
  mass_metadata,
  by = "label",
  all.x = TRUE,
  sort = FALSE
)
mapping_mass <- mapping_mass[match(mapping_qc$label, mapping_mass$label), ]
stopifnot(
  identical(mapping_mass$label, mapping_qc$label),
  !anyNA(mapping_mass$mass_g)
)

test_mapping_mass <- function(data, metric, mapping_definition, scenario) {
  correlation <- suppressWarnings(stats::cor.test(
    data$mass_g,
    data[[metric]],
    method = "spearman",
    exact = FALSE
  ))
  data$mapping_logit <- qlogis(
    pmin(pmax(data[[metric]] / 100, 1e-6), 1 - 1e-6)
  )
  adjusted_model <- stats::lm(
    mapping_logit ~ mass_g + Diet + Treatment,
    data = data
  )
  adjusted_test <- stats::drop1(adjusted_model, test = "F")

  data.frame(
    mapping_definition = mapping_definition,
    scenario = scenario,
    samples = nrow(data),
    spearman_rho = unname(correlation$estimate),
    spearman_p = correlation$p.value,
    adjusted_mass_p = adjusted_test["mass_g", "Pr(>F)"],
    stringsAsFactors = FALSE
  )
}

mapping_mass_tests <- do.call(rbind, lapply(
  names(mapping_definitions),
  function(metric) {
    rbind(
      test_mapping_mass(
        mapping_mass,
        metric,
        mapping_definitions[[metric]],
        "All 45 libraries shown for QC"
      ),
      test_mapping_mass(
        mapping_mass[mapping_mass$label != "1044", ],
        metric,
        mapping_definitions[[metric]],
        "Final 44-sample analysis cohort"
      )
    )
  }
))

format_mass_test <- function(test_rows) {
  paste0(
    "All 45: Spearman rho = ",
    sprintf("%.2f", test_rows$spearman_rho[1]),
    ", p = ", format.pval(test_rows$spearman_p[1], digits = 3, eps = 0.001),
    "; adjusted mass p = ",
    format.pval(test_rows$adjusted_mass_p[1], digits = 3, eps = 0.001),
    "\nWithout 1044: rho = ", sprintf("%.2f", test_rows$spearman_rho[2]),
    ", p = ", format.pval(test_rows$spearman_p[2], digits = 3, eps = 0.001),
    "; adjusted mass p = ",
    format.pval(test_rows$adjusted_mass_p[2], digits = 3, eps = 0.001)
  )
}

diet_colors <- c("33" = "#E67E22", "50" = "#2E8B57", "83" = "#7B4AA8")
treatment_shapes <- c(Control = 21, Infected = 24)
treatment_colors <- c(Control = "#667784", Infected = "#8B0000")
treatment_fills <- c(Control = "#FAEBD7", Infected = "#8B0000")
treatment_linetypes <- c(Control = "dashed", Infected = "solid")

test_mapping_mass_by_group <- function(data, metric, mapping_definition) {
  group_data <- data[data$label != "1044", , drop = FALSE]
  group_data$mapping_percentage <- group_data[[metric]]

  group_rows <- split(
    group_data,
    interaction(group_data$Diet, group_data$Treatment, drop = TRUE)
  )

  do.call(rbind, lapply(group_rows, function(group_df) {
    fit <- stats::lm(mapping_percentage ~ mass_g, data = group_df)
    fit_summary <- summary(fit)
    slope_row <- coef(fit_summary)["mass_g", ]
    data.frame(
      mapping_definition = mapping_definition,
      Diet = as.character(group_df$Diet[1]),
      Treatment = as.character(group_df$Treatment[1]),
      samples = nrow(group_df),
      slope_percentage_points_per_g = unname(slope_row["Estimate"]),
      slope_standard_error = unname(slope_row["Std. Error"]),
      slope_p = unname(slope_row["Pr(>|t|)"]),
      r_squared = unname(fit_summary$r.squared),
      stringsAsFactors = FALSE
    )
  }))
}

mapping_mass_group_tests <- do.call(rbind, lapply(
  names(mapping_definitions),
  function(metric) {
    test_mapping_mass_by_group(
      mapping_mass,
      metric,
      mapping_definitions[[metric]]
    )
  }
))

make_mapping_mass_plot <- function(data, metric, title, y_label, test_rows) {
  plot_data <- data
  plot_data$mapping_percentage <- plot_data[[metric]]
  fit_data <- plot_data[plot_data$label != "1044", , drop = FALSE]

  ggplot2::ggplot(
    plot_data,
    ggplot2::aes(
      x = mass_g,
      y = mapping_percentage,
      color = Treatment,
      fill = Treatment,
      shape = Treatment
    )
  ) +
    ggplot2::geom_smooth(
      data = fit_data,
      ggplot2::aes(
        group = Treatment,
        color = Treatment,
        fill = Treatment,
        linetype = Treatment
      ),
      method = "lm",
      formula = y ~ x,
      se = TRUE,
      alpha = 0.12,
      linewidth = 1.05
    ) +
    ggplot2::geom_point(size = 3.4, stroke = 1.15) +
    ggrepel::geom_text_repel(
      data = plot_data[plot_data$label != "1044", ],
      ggplot2::aes(label = label),
      color = "#252525",
      size = 3.1,
      box.padding = 0.28,
      point.padding = 0.15,
      min.segment.length = 0,
      max.overlaps = Inf,
      seed = 83,
      show.legend = FALSE
    ) +
    ggrepel::geom_label_repel(
      data = plot_data[plot_data$label == "1044", ],
      ggplot2::aes(label = label),
      size = 4.0,
      fontface = "bold",
      fill = "#FFF3CD",
      color = "#8B0000",
      box.padding = 0.45,
      point.padding = 0.25,
      min.segment.length = 0,
      seed = 1044,
      show.legend = FALSE
    ) +
    ggplot2::facet_wrap(
      ~ Diet,
      nrow = 1,
      labeller = ggplot2::labeller(Diet = function(x) paste("Diet", x))
    ) +
    ggplot2::scale_color_manual(values = treatment_colors) +
    ggplot2::scale_fill_manual(values = treatment_fills) +
    ggplot2::scale_shape_manual(values = treatment_shapes) +
    ggplot2::scale_linetype_manual(values = treatment_linetypes) +
    ggplot2::scale_x_continuous(
      breaks = scales::breaks_pretty(n = 6),
      expand = ggplot2::expansion(mult = c(0.08, 0.08))
    ) +
    ggplot2::labs(
      title = title,
      subtitle = paste0(
        format_mass_test(test_rows),
        "\nWithin each diet: dashed = control; solid = infected; fits exclude 1044"
      ),
      x = "Recorded individual mass (g)",
      y = y_label,
      color = "Treatment",
      fill = "Treatment",
      shape = "Treatment",
      linetype = "Treatment"
    ) +
    ggplot2::theme_classic(base_size = 15) +
    ggplot2::theme(
      legend.position = "bottom",
      plot.title = ggplot2::element_text(face = "bold", size = 18),
      plot.subtitle = ggplot2::element_text(size = 12.5, lineheight = 1.05),
      strip.background = ggplot2::element_rect(
        fill = "#F2F4F5", color = "#5F6B73", linewidth = 0.6
      ),
      strip.text = ggplot2::element_text(face = "bold", size = 14)
    )
}

p_mass_unique <- make_mapping_mass_plot(
  mapping_mass,
  "unique_pct",
  "Unique mapping and recorded individual mass",
  "Uniquely mapped reads (%)",
  mapping_mass_tests[mapping_mass_tests$mapping_definition == "Unique mapping", ]
)
p_mass_unique_plus_multi <- make_mapping_mass_plot(
  mapping_mass,
  "unique_plus_multimapped_pct",
  "Unique plus multimapping and recorded individual mass",
  "Reads mapped to 1--20 loci (%)",
  mapping_mass_tests[
    mapping_mass_tests$mapping_definition == "Unique + multimapping",
  ]
)

p_mapping_mass <-
  p_mass_unique / p_mass_unique_plus_multi +
  patchwork::plot_layout(guides = "collect") +
  patchwork::plot_annotation(
    tag_levels = "A",
    title = "Competitive STAR mapping in relation to individual mass",
    subtitle = paste(
      "All 45 libraries are shown for QC; 1044 remains excluded from",
      "downstream expression analyses"
    ),
    theme = ggplot2::theme(
      plot.title = ggplot2::element_text(size = 20, face = "bold"),
      plot.subtitle = ggplot2::element_text(size = 14)
    )
  ) &
  ggplot2::theme(
    legend.position = "bottom",
    plot.tag = ggplot2::element_text(size = 19, face = "bold")
  )

print(p_mapping_mass)

for (extension in c("png", "pdf", "svg")) {
  ggplot2::ggsave(
    file.path(
      out_dir,
      paste0("mapping_quality_by_individual_mass.", extension)
    ),
    p_mapping_mass,
    width = 16,
    height = 15,
    dpi = 320
  )
}

write.csv(
  mapping_mass,
  file.path(out_dir, "mapping_quality_and_individual_mass_45_samples.csv"),
  row.names = FALSE
)
write.csv(
  mapping_mass_tests,
  file.path(out_dir, "mapping_quality_mass_association_tests.csv"),
  row.names = FALSE
)
write.csv(
  mapping_mass_group_tests,
  file.path(out_dir, "mapping_quality_mass_group_regression_tests.csv"),
  row.names = FALSE
)

knitr::kable(
  mapping_mass_tests,
  digits = 4,
  caption = "Association between recorded individual mass and mapping quality"
)

knitr::kable(
  mapping_mass_group_tests,
  digits = 4,
  caption = paste(
    "Exploratory within-diet regressions of mapping quality on mass",
    "for control and infected locusts"
  )
)
```

Among the 44 retained libraries, factorial models found no supported
diet-by-infection interaction for unique mapping
(`r sprintf("P = %.3f", mapping_model_results$interaction_p[mapping_model_results$mapping_definition == "Unique mapping" & mapping_model_results$scenario == "Final 44-sample analysis cohort"])`)
and no diet main effect
(`r sprintf("P = %.3f", mapping_model_results$diet_p[mapping_model_results$mapping_definition == "Unique mapping" & mapping_model_results$scenario == "Final 44-sample analysis cohort"])`).
Infection was associated with lower unique mapping (control mean
`r sprintf("%.2f%%", mean(mapping_mass$unique_pct[mapping_mass$label != "1044" & mapping_mass$Treatment == "Control"]))`;
infected mean
`r sprintf("%.2f%%", mean(mapping_mass$unique_pct[mapping_mass$label != "1044" & mapping_mass$Treatment == "Infected"]))`;
`r sprintf("P = %.3f", mapping_model_results$infection_p[mapping_model_results$mapping_definition == "Unique mapping" & mapping_model_results$scenario == "Final 44-sample analysis cohort"])`).
The same pattern was observed after accepted multimapping reads were included:
the interaction (`r sprintf("P = %.3f", mapping_model_results$interaction_p[mapping_model_results$mapping_definition == "Unique + multimapping" & mapping_model_results$scenario == "Final 44-sample analysis cohort"])`)
and diet (`r sprintf("P = %.3f", mapping_model_results$diet_p[mapping_model_results$mapping_definition == "Unique + multimapping" & mapping_model_results$scenario == "Final 44-sample analysis cohort"])`)
were unsupported, whereas infected libraries had lower combined mapping yield
(control mean
`r sprintf("%.2f%%", mean(mapping_mass$unique_plus_multimapped_pct[mapping_mass$label != "1044" & mapping_mass$Treatment == "Control"]))`;
infected mean
`r sprintf("%.2f%%", mean(mapping_mass$unique_plus_multimapped_pct[mapping_mass$label != "1044" & mapping_mass$Treatment == "Infected"]))`;
`r sprintf("P = %.3f", mapping_model_results$infection_p[mapping_model_results$mapping_definition == "Unique + multimapping" & mapping_model_results$scenario == "Final 44-sample analysis cohort"])`).

Recorded mass did not explain mapping yield within this cohort. For unique
mapping, the Spearman association was
`r sprintf("rho = %.3f, P = %.3f", mapping_mass_tests$spearman_rho[mapping_mass_tests$mapping_definition == "Unique mapping" & mapping_mass_tests$scenario == "Final 44-sample analysis cohort"], mapping_mass_tests$spearman_p[mapping_mass_tests$mapping_definition == "Unique mapping" & mapping_mass_tests$scenario == "Final 44-sample analysis cohort"])`,
and the mass effect adjusted for diet and infection was
`r sprintf("P = %.3f", mapping_mass_tests$adjusted_mass_p[mapping_mass_tests$mapping_definition == "Unique mapping" & mapping_mass_tests$scenario == "Final 44-sample analysis cohort"])`.
After accepted multimapping reads were included, the corresponding values were
`r sprintf("rho = %.3f, P = %.3f", mapping_mass_tests$spearman_rho[mapping_mass_tests$mapping_definition == "Unique + multimapping" & mapping_mass_tests$scenario == "Final 44-sample analysis cohort"], mapping_mass_tests$spearman_p[mapping_mass_tests$mapping_definition == "Unique + multimapping" & mapping_mass_tests$scenario == "Final 44-sample analysis cohort"])`
and adjusted `r sprintf("P = %.3f", mapping_mass_tests$adjusted_mass_p[mapping_mass_tests$mapping_definition == "Unique + multimapping" & mapping_mass_tests$scenario == "Final 44-sample analysis cohort"])`.

## Body mass by diet and infection

For the experiment-level mass test, recorded individual mass is the response.
We first test the `Diet x Infection` interaction in the 44-library analysis
cohort. Because the interaction is not supported, diet and infection are then
tested in an additive model. Sample 1044 remains excluded from this biological
analysis because it was removed on independent mapping-quality grounds.

```{r individual-mass-by-design, fig.width=10, fig.height=7.5}
mass_design <- mapping_mass[mapping_mass$label != "1044", , drop = FALSE]
mass_design$Diet <- factor(mass_design$Diet, levels = c("33", "50", "83"))
mass_design$Treatment <- factor(
  mass_design$Treatment,
  levels = c("Control", "Infected")
)

mass_interaction_model <- stats::lm(
  mass_g ~ Diet * Treatment,
  data = mass_design
)
mass_interaction_test <- stats::drop1(mass_interaction_model, test = "F")
mass_interaction_p <- mass_interaction_test[
  "Diet:Treatment", "Pr(>F)"
]

mass_additive_model <- stats::lm(
  mass_g ~ Diet + Treatment,
  data = mass_design
)
mass_additive_test <- stats::drop1(mass_additive_model, test = "F")
mass_diet_p <- mass_additive_test["Diet", "Pr(>F)"]
mass_infection_p <- mass_additive_test["Treatment", "Pr(>F)"]

mass_test_table <- data.frame(
  term = c("Diet x Infection", "Diet", "Infection"),
  model = c(
    "Full factorial model",
    "Additive model after unsupported interaction",
    "Additive model after unsupported interaction"
  ),
  numerator_df = c(
    mass_interaction_test["Diet:Treatment", "Df"],
    mass_additive_test["Diet", "Df"],
    mass_additive_test["Treatment", "Df"]
  ),
  F = c(
    mass_interaction_test["Diet:Treatment", "F value"],
    mass_additive_test["Diet", "F value"],
    mass_additive_test["Treatment", "F value"]
  ),
  p_value = c(mass_interaction_p, mass_diet_p, mass_infection_p),
  stringsAsFactors = FALSE
)

mass_summary <- mass_design |>
  dplyr::group_by(Diet, Treatment) |>
  dplyr::summarise(
    samples = dplyr::n(),
    mean_mass_g = mean(mass_g),
    sd_mass_g = stats::sd(mass_g),
    se_mass_g = sd_mass_g / sqrt(samples),
    .groups = "drop"
  )

infection_coefficient <- stats::coef(mass_additive_model)["TreatmentInfected"]
infection_ci <- stats::confint(
  mass_additive_model,
  "TreatmentInfected",
  level = 0.95
)
mass_effect_table <- data.frame(
  comparison = "Infected - Control, adjusted for diet",
  estimate_g = unname(infection_coefficient),
  confidence_level = 0.95,
  lower_ci_g = unname(infection_ci[1]),
  upper_ci_g = unname(infection_ci[2]),
  p_value = mass_infection_p
)

mass_design_plot <- ggplot2::ggplot(
  mass_design,
  ggplot2::aes(x = Diet, y = mass_g, fill = Treatment, shape = Treatment)
) +
  ggplot2::geom_boxplot(
    ggplot2::aes(group = interaction(Diet, Treatment)),
    width = 0.62,
    position = ggplot2::position_dodge(width = 0.74),
    alpha = 0.72,
    outlier.shape = NA,
    color = "#202020",
    linewidth = 0.65
  ) +
  ggplot2::geom_point(
    position = ggplot2::position_jitterdodge(
      jitter.width = 0.12,
      jitter.height = 0,
      dodge.width = 0.74,
      seed = 20260811
    ),
    size = 3.4,
    stroke = 0.85,
    color = "#202020",
    alpha = 0.9
  ) +
  ggplot2::stat_summary(
    ggplot2::aes(group = Treatment),
    fun = mean,
    geom = "point",
    position = ggplot2::position_dodge(width = 0.74),
    shape = 23,
    size = 4.2,
    stroke = 1,
    fill = "white",
    color = "#111111"
  ) +
  ggplot2::scale_fill_manual(
    values = c(Control = "#FAEBD7", Infected = "#8B0000")
  ) +
  ggplot2::scale_shape_manual(values = c(Control = 21, Infected = 24)) +
  ggplot2::scale_y_continuous(
    labels = scales::label_number(accuracy = 0.01),
    expand = ggplot2::expansion(mult = c(0.04, 0.11))
  ) +
  ggplot2::labs(
    title = "Recorded individual mass by diet and infection treatment",
    subtitle = paste0(
      "Diet x infection: P = ", sprintf("%.3f", mass_interaction_p),
      "; diet: P = ", sprintf("%.3f", mass_diet_p),
      "; infection: P = ", sprintf("%.3f", mass_infection_p),
      "\nAdjusted infected-control difference = ",
      sprintf("%.3f", infection_coefficient), " g (95% CI ",
      sprintf("%.3f", infection_ci[1]), " to ",
      sprintf("%.3f", infection_ci[2]), ")"
    ),
    x = "Diet (protein:carbohydrate ratio)",
    y = "Recorded individual mass (g)",
    fill = "Treatment",
    shape = "Treatment"
  ) +
  ggplot2::theme_classic(base_size = 16) +
  ggplot2::theme(
    legend.position = "bottom",
    plot.title = ggplot2::element_text(face = "bold", size = 19),
    plot.subtitle = ggplot2::element_text(size = 13, lineheight = 1.08),
    axis.title = ggplot2::element_text(size = 16),
    axis.text = ggplot2::element_text(size = 14)
  )

print(mass_design_plot)

for (extension in c("png", "pdf", "svg")) {
  ggplot2::ggsave(
    file.path(
      out_dir,
      paste0("individual_mass_by_diet_infection.", extension)
    ),
    mass_design_plot,
    width = 10,
    height = 7.5,
    dpi = 320
  )
}

write.csv(
  mass_design,
  file.path(out_dir, "individual_mass_analysis_44_samples.csv"),
  row.names = FALSE
)
write.csv(
  mass_summary,
  file.path(out_dir, "individual_mass_group_summary.csv"),
  row.names = FALSE
)
write.csv(
  mass_test_table,
  file.path(out_dir, "individual_mass_diet_infection_tests.csv"),
  row.names = FALSE
)
write.csv(
  mass_effect_table,
  file.path(out_dir, "individual_mass_infection_effect.csv"),
  row.names = FALSE
)

knitr::kable(
  mass_test_table,
  digits = 4,
  caption = "Diet, infection, and diet-by-infection tests for recorded mass"
)
knitr::kable(
  mass_summary,
  digits = 4,
  caption = "Recorded mass summary for each diet-by-treatment group"
)
```

There is no supported diet-by-infection interaction
(`r sprintf("P = %.3f", mass_interaction_p)`) and no diet main effect
(`r sprintf("P = %.3f", mass_diet_p)`). Infected animals are lighter overall
after accounting for diet (`r sprintf("P = %.3f", mass_infection_p)`), with an
adjusted infected-minus-control difference of
`r sprintf("%.3f g", infection_coefficient)` (95% CI
`r sprintf("%.3f to %.3f g", infection_ci[1], infection_ci[2])`).

## Low-count filtering

Genes with almost no reads across the whole experiment add multiple-testing
burden but usually cannot be estimated reliably. The filtering threshold keeps
genes with at least `r params$min_total_count` total counts, which is a light
screen intended to remove uninformative rows while preserving diet- and
infection-responsive genes for DESeq2.

```{r filtering}
keep <- rowSums(counts) >= params$min_total_count
filter_summary <- data.frame(
  min_total_count = params$min_total_count,
  genes_before = nrow(counts),
  genes_after = sum(keep),
  genes_removed = sum(!keep)
)

knitr::kable(filter_summary)
write.csv(filter_summary, file.path(out_dir, "filter_summary.csv"),
          row.names = FALSE)
```

## PCA

The PCA is calculated after a variance-stabilizing transformation (VST). VST
puts low-count and high-count genes onto a more comparable scale, so the plot is
useful for sample-level structure rather than being driven mostly by the most
abundant genes. We expect samples to show some organization by diet and/or
infection, but we also use this plot to identify samples that sit apart from
their expected group.

```{r pca}
dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = counts[keep, ],
  colData = metadata,
  design = ~ Diet + Treatment
)

vsd <- DESeq2::vst(dds, blind = TRUE)
pca_df <- DESeq2::plotPCA(vsd, intgroup = c("Diet", "Treatment"),
                          returnData = TRUE)
percent_var <- round(100 * attr(pca_df, "percentVar"))

p_pca <- ggplot(pca_df, aes(PC1, PC2, color = Diet, shape = Treatment)) +
  geom_point(size = 3) +
  theme_classic() +
  labs(
    title = "Variance-stabilized PCA",
    x = paste0("PC1: ", percent_var[1], "% variance"),
    y = paste0("PC2: ", percent_var[2], "% variance")
  )

p_pca
ggsave(file.path(out_dir, "pca_vst.png"), p_pca, width = 7, height = 5, dpi = 300)
```

## Sample distance heatmap

The sample distance heatmap asks the same QC question as the PCA in a different
format: which libraries are most similar after VST normalization? Samples from
the same diet and treatment should usually be closer to one another than to
unrelated groups. A sample that clusters away from its condition is a candidate
outlier and should be checked before it is allowed to drive differential
expression calls.

```{r sample-distance}
sample_dist <- dist(t(SummarizedExperiment::assay(vsd)))
sample_dist_matrix <- as.matrix(sample_dist)
rownames(sample_dist_matrix) <- paste(metadata$Diet, metadata$Treatment,
                                      metadata$label, sep = "_")
colnames(sample_dist_matrix) <- rownames(sample_dist_matrix)

png(file.path(out_dir, "sample_distance_heatmap.png"),
    width = 1600, height = 1400, res = 180)
pheatmap::pheatmap(sample_dist_matrix,
                   main = "Sample-to-sample distance")
dev.off()

pheatmap::pheatmap(sample_dist_matrix,
                   main = "Sample-to-sample distance")
```

## Most variable genes

This heatmap shows the `r params$top_variable_genes` genes with the strongest
sample-to-sample variation after VST transformation. It is not a DEG test. It is
a compact view of whether the largest expression patterns line up with the
experimental design and whether any sample carries a distinct expression profile
that might explain an outlier flag.

```{r variable-genes}
vst_mat <- SummarizedExperiment::assay(vsd)
gene_variance <- apply(vst_mat, 1, var)
top_n <- min(params$top_variable_genes, length(gene_variance))
top_genes <- names(sort(gene_variance, decreasing = TRUE))[seq_len(top_n)]

annotation_col <- data.frame(Diet = metadata$Diet,
                             Treatment = metadata$Treatment)
rownames(annotation_col) <- colnames(vst_mat)

png(file.path(out_dir, "top_variable_gene_heatmap.png"),
    width = 1600, height = 1600, res = 180)
pheatmap::pheatmap(vst_mat[top_genes, ],
                   show_rownames = FALSE,
                   annotation_col = annotation_col,
                   main = paste("Top", top_n, "variable genes"))
dev.off()

pheatmap::pheatmap(vst_mat[top_genes, ],
                   show_rownames = FALSE,
                   annotation_col = annotation_col,
                   main = paste("Top", top_n, "variable genes"))
```

```{r session-info}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
