---
title: "Count QC and exploration"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  metadata_file: "data/metadata/mehreen_metadata_corrected_45.txt"
  counts_file: "output/rmd_runs/transcript-exon-primary-corrected45_20260809-153146/host_counts_no_rrna.tsv"
  min_total_count: 10
  top_variable_genes: 500
  output_tag: "count-qc-exploration"
---

## Purpose

::: {.walkthrough-grid}
::: {.walkthrough-step}
**Step 1 - Filter counts**

Remove rRNA and uninformative low-count rows while retaining every sample.
:::
::: {.walkthrough-step}
**Step 2 - Explore samples**

Use library depth, PCA, and sample distances to describe variation.
:::
::: {.walkthrough-step}
**Step 3 - Interpret QC**

Identify unusual libraries without excluding them from the main analysis.
:::
:::

This page checks count depth, low-count filtering, sample clustering, and PCA
before differential expression models are interpreted. These checks are not a
separate biological result; they are the audit trail that tells us whether the
diet and infection contrasts are being estimated from comparable samples or
whether a small number of unusual libraries could dominate the model.

## Setup and data loading

The count matrix and metadata are matched by sample label before any plotting.
This small bookkeeping step is important because RNA-seq models assume that the
columns of the count matrix and the rows of the metadata describe the same
samples in the same order.

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c("DESeq2", "ggplot2", "pheatmap", "scales")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

metadata <- read.delim(file.path(project_dir, params$metadata_file),
                       check.names = FALSE)
metadata$label <- as.character(metadata$label)
metadata$Diet <- factor(metadata$Diet, levels = c("33", "50", "83"))
metadata$Treatment <- factor(metadata$Treatment, levels = c("Control", "Infected"))

counts_raw <- read.delim(file.path(project_dir, params$counts_file),
                         check.names = FALSE)
counts <- as.matrix(counts_raw[, -1])
rownames(counts) <- counts_raw[[1]]
storage.mode(counts) <- "integer"

count_labels <- sub("_MERGE$", "", sub("^mehreen_", "", colnames(counts)))
metadata <- metadata[match(count_labels, metadata$label), ]
stopifnot(identical(metadata$label, count_labels))
rownames(metadata) <- metadata$label
colnames(counts) <- count_labels
```

## Library size

Library size is the first check because very low-depth samples can have noisy
gene estimates even when their biological label is correct. Here we compare the
total assigned counts across diet and infection groups, looking for samples that
fall far outside the rest of their condition.

```{r library-size}
library(ggplot2)

library_size <- data.frame(
  label = metadata$label,
  total_counts = colSums(counts),
  Diet = metadata$Diet,
  Treatment = metadata$Treatment
)

ggplot(library_size, aes(x = interaction(Diet, Treatment),
                         y = total_counts,
                         color = Treatment)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.12, height = 0, size = 2) +
  scale_y_continuous(labels = scales::comma) +
  theme_classic() +
  labs(
    title = "Total assigned counts per sample",
    x = "Diet and treatment",
    y = "Total counts",
    color = "Treatment"
  )

write.csv(library_size, file.path(out_dir, "library_size.csv"),
          row.names = FALSE)
```

## Low-count filtering

Genes with almost no reads across the whole experiment add multiple-testing
burden but usually cannot be estimated reliably. The filtering threshold keeps
genes with at least `r params$min_total_count` total counts, which is a light
screen intended to remove uninformative rows while preserving diet- and
infection-responsive genes for DESeq2.

```{r filtering}
keep <- rowSums(counts) >= params$min_total_count
filter_summary <- data.frame(
  min_total_count = params$min_total_count,
  genes_before = nrow(counts),
  genes_after = sum(keep),
  genes_removed = sum(!keep)
)

knitr::kable(filter_summary)
write.csv(filter_summary, file.path(out_dir, "filter_summary.csv"),
          row.names = FALSE)
```

## PCA

The PCA is calculated after a variance-stabilizing transformation (VST). VST
puts low-count and high-count genes onto a more comparable scale, so the plot is
useful for sample-level structure rather than being driven mostly by the most
abundant genes. We expect samples to show some organization by diet and/or
infection, but we also use this plot to identify samples that sit apart from
their expected group.

```{r pca}
dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = counts[keep, ],
  colData = metadata,
  design = ~ Diet + Treatment
)

vsd <- DESeq2::vst(dds, blind = TRUE)
pca_df <- DESeq2::plotPCA(vsd, intgroup = c("Diet", "Treatment"),
                          returnData = TRUE)
percent_var <- round(100 * attr(pca_df, "percentVar"))

p_pca <- ggplot(pca_df, aes(PC1, PC2, color = Diet, shape = Treatment)) +
  geom_point(size = 3) +
  theme_classic() +
  labs(
    title = "Variance-stabilized PCA",
    x = paste0("PC1: ", percent_var[1], "% variance"),
    y = paste0("PC2: ", percent_var[2], "% variance")
  )

p_pca
ggsave(file.path(out_dir, "pca_vst.png"), p_pca, width = 7, height = 5, dpi = 300)
```

## Sample distance heatmap

The sample distance heatmap asks the same QC question as the PCA in a different
format: which libraries are most similar after VST normalization? Samples from
the same diet and treatment should usually be closer to one another than to
unrelated groups. A sample that clusters away from its condition is a candidate
outlier and should be checked before it is allowed to drive differential
expression calls.

```{r sample-distance}
sample_dist <- dist(t(SummarizedExperiment::assay(vsd)))
sample_dist_matrix <- as.matrix(sample_dist)
rownames(sample_dist_matrix) <- paste(metadata$Diet, metadata$Treatment,
                                      metadata$label, sep = "_")
colnames(sample_dist_matrix) <- rownames(sample_dist_matrix)

png(file.path(out_dir, "sample_distance_heatmap.png"),
    width = 1600, height = 1400, res = 180)
pheatmap::pheatmap(sample_dist_matrix,
                   main = "Sample-to-sample distance")
dev.off()

pheatmap::pheatmap(sample_dist_matrix,
                   main = "Sample-to-sample distance")
```

## Most variable genes

This heatmap shows the `r params$top_variable_genes` genes with the strongest
sample-to-sample variation after VST transformation. It is not a DEG test. It is
a compact view of whether the largest expression patterns line up with the
experimental design and whether any sample carries a distinct expression profile
that might explain an outlier flag.

```{r variable-genes}
vst_mat <- SummarizedExperiment::assay(vsd)
gene_variance <- apply(vst_mat, 1, var)
top_n <- min(params$top_variable_genes, length(gene_variance))
top_genes <- names(sort(gene_variance, decreasing = TRUE))[seq_len(top_n)]

annotation_col <- data.frame(Diet = metadata$Diet,
                             Treatment = metadata$Treatment)
rownames(annotation_col) <- colnames(vst_mat)

png(file.path(out_dir, "top_variable_gene_heatmap.png"),
    width = 1600, height = 1600, res = 180)
pheatmap::pheatmap(vst_mat[top_genes, ],
                   show_rownames = FALSE,
                   annotation_col = annotation_col,
                   main = paste("Top", top_n, "variable genes"))
dev.off()

pheatmap::pheatmap(vst_mat[top_genes, ],
                   show_rownames = FALSE,
                   annotation_col = annotation_col,
                   main = paste("Top", top_n, "variable genes"))
```

```{r session-info}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
