---
title: "All samples + all scaffolds: diet pairwise DEGs"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  metadata_file: "data/metadata/mehreen_metadata_corrected_45.txt"
  counts_file: "output/rmd_runs/transcript-exon-primary-corrected45_20260809-153146/host_counts_no_rrna.tsv"
  treatment_subset: "All"
  min_total_count: 10
  alpha: 0.05
  lfc_threshold: 1
  output_tag: "all-scaffolds-diet-pairwise-deg"
---

## Purpose

::: {.walkthrough-grid}
::: {.walkthrough-step}
**Step 1 - Separate treatment**

Model diet contrasts within control and infected samples.
:::
::: {.walkthrough-step}
**Step 2 - Count DEGs**

Apply one adjusted-P-value and fold-change rule to each diet pair.
:::
::: {.walkthrough-step}
**Step 3 - Compare contexts**

Identify diet differences that depend on infection status.
:::
:::

This page tests pairwise diet contrasts. Set `treatment_subset` to `Control`,
`Infected`, or `All` in the YAML header before knitting.

All 45 samples are retained, and the fresh competitive-host count matrix keeps
all non-rRNA loci regardless of assembly placement. The matching
[placed-chromosome diet analysis](15-all-samples-diet-pairwise-deg.html) uses
the same samples but a narrower gene universe.

The aim is to separate the diet question from the infection question as much as
the design allows. When all samples are used, the model includes treatment as a
covariate so diet contrasts are not simply rediscovering the infected-control
difference. When one treatment is selected, the model asks which genes differ
among diets within that treatment subset only.

## Setup and data loading

The metadata and count matrix are matched by sample label before modelling.
This keeps the analysis reproducible and prevents silent sample-order mistakes,
which are especially easy to make when count columns carry sequencing-style
prefixes and suffixes.

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c("DESeq2", "ggplot2", "EnhancedVolcano")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}
library(ggplot2)

direction_colors <- c("Higher in first diet" = "#237C8E",
                      "Higher in second diet" = "#B8643C")

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, params$treatment_subset,
                           run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

metadata <- read.delim(file.path(project_dir, params$metadata_file),
                       check.names = FALSE)
metadata$label <- as.character(metadata$label)
metadata$Diet <- factor(metadata$Diet, levels = c("33", "50", "83"))
metadata$Treatment <- factor(metadata$Treatment, levels = c("Control", "Infected"))

counts_path <- file.path(project_dir, params$counts_file)
counts_raw <- if (grepl("[.]tsv$", counts_path, ignore.case = TRUE)) {
  read.delim(counts_path, check.names = FALSE)
} else {
  read.csv(counts_path, check.names = FALSE)
}
counts <- as.matrix(counts_raw[, -1])
rownames(counts) <- counts_raw[[1]]
storage.mode(counts) <- "integer"

count_labels <- sub("_MERGE$", "", sub("^mehreen_", "", colnames(counts)))
metadata <- metadata[match(count_labels, metadata$label), ]
stopifnot(identical(metadata$label, count_labels))
rownames(metadata) <- metadata$label
colnames(counts) <- count_labels

if (params$treatment_subset != "All") {
  keep_samples <- metadata$Treatment == params$treatment_subset
  counts <- counts[, keep_samples]
  metadata <- droplevels(metadata[keep_samples, ])
}

keep_genes <- rowSums(counts) >= params$min_total_count
```

## Fit model

DESeq2 models raw counts with library-size normalization and gene-wise
dispersion estimates. The `All` analysis uses `~ Treatment + Diet`, so each diet
coefficient is estimated after accounting for infection status. This gives a
cleaner diet comparison than pooling control and infected samples without a
treatment term.

```{r fit-model}
design_formula <- if (params$treatment_subset == "All") {
  ~ Treatment + Diet
} else {
  ~ Diet
}

dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = counts[keep_genes, ],
  colData = metadata,
  design = design_formula
)

dds <- DESeq2::DESeq(dds)
```

## Pairwise diet contrasts

Each contrast compares two diet levels at a time. A positive log2 fold-change
means the gene is higher in the first diet named in the contrast; a negative
log2 fold-change means it is higher in the second diet. The volcano plots show
both effect size and adjusted p-value, while the CSV files keep the full table
for downstream overlap and enrichment analyses.

```{r pairwise-contrasts, fig.height=7, fig.width=8}
diet_pairs <- combn(levels(droplevels(metadata$Diet)), 2, simplify = FALSE)

contrast_results <- list()
for (pair in diet_pairs) {
  contrast_label <- paste(pair[1], "vs", pair[2], sep = "_")
  message("Running diet contrast: ", contrast_label)

  res <- DESeq2::results(dds, contrast = c("Diet", pair[1], pair[2]),
                         alpha = params$alpha)
  res_df <- as.data.frame(res)
  res_df$gene_id <- rownames(res_df)
  res_df$contrast <- contrast_label
  res_df$significant <- !is.na(res_df$padj) &
    res_df$padj < params$alpha &
    abs(res_df$log2FoldChange) >= params$lfc_threshold

  contrast_results[[contrast_label]] <- res_df
  write.csv(res_df, file.path(out_dir, paste0("DESeq2_", contrast_label, ".csv")),
            row.names = FALSE)
  writeLines(res_df$gene_id[res_df$significant],
             file.path(out_dir, paste0("significant_genes_", contrast_label, ".txt")))

  print(EnhancedVolcano::EnhancedVolcano(
    res_df,
    lab = res_df$gene_id,
    x = "log2FoldChange",
    y = "padj",
    pCutoff = params$alpha,
    FCcutoff = params$lfc_threshold,
    title = paste("Diet", pair[1], "vs", pair[2]),
    subtitle = paste("Treatment subset:", params$treatment_subset)
  ))
}

all_pairwise <- do.call(rbind, contrast_results)
write.csv(all_pairwise, file.path(out_dir, "all_diet_pairwise_results.csv"),
          row.names = FALSE)
```

## DEG count barplot

Volcano plots are useful for seeing the distribution of effect sizes, but they
are not the fastest way to compare how many genes pass the current thresholds in
each contrast. This barplot reduces each contrast to the number of significant
DEGs and separates the direction of change.

```{r deg-count-barplot, fig.height=4.8, fig.width=7.5}
sig_pairwise <- all_pairwise[all_pairwise$significant &
                               !is.na(all_pairwise$log2FoldChange), ]
sig_pairwise$direction <- ifelse(sig_pairwise$log2FoldChange > 0,
                                 "Higher in first diet",
                                 "Higher in second diet")
sig_pairwise$contrast <- factor(sig_pairwise$contrast,
                                levels = names(contrast_results))
sig_pairwise$direction <- factor(sig_pairwise$direction,
                                 levels = c("Higher in first diet",
                                            "Higher in second diet"))

deg_bar_counts <- as.data.frame(table(
  contrast = sig_pairwise$contrast,
  direction = sig_pairwise$direction
))
names(deg_bar_counts)[3] <- "n_significant_genes"

p_deg_counts <- ggplot(deg_bar_counts,
                       aes(x = contrast, y = n_significant_genes,
                           fill = direction)) +
  geom_col(position = position_dodge(width = 0.78), width = 0.68) +
  geom_text(aes(label = n_significant_genes),
            position = position_dodge(width = 0.78),
            vjust = -0.35, size = 3.2) +
  scale_fill_manual(values = direction_colors) +
  theme_classic() +
  labs(
    title = "Significant diet DEGs by contrast and direction",
    subtitle = paste("Treatment subset:", params$treatment_subset),
    x = "Diet contrast",
    y = "Number of significant genes",
    fill = "Direction"
  )

p_deg_counts
ggsave(file.path(out_dir, "diet_pairwise_deg_count_barplot.png"),
       p_deg_counts, width = 7.5, height = 4.8, dpi = 300)
write.csv(deg_bar_counts, file.path(out_dir, "diet_pairwise_direction_counts.csv"),
          row.names = FALSE)
```

## Summary

The summary table reports the total number of genes that pass the adjusted
p-value and log2 fold-change thresholds for each diet contrast. It is saved as a
small machine-readable table because these counts are used again by the
publication-figure and enrichment pages.

```{r summary}
summary_table <- aggregate(significant ~ contrast, all_pairwise, sum)
names(summary_table)[2] <- "n_significant_genes"
knitr::kable(summary_table)

write.csv(summary_table, file.path(out_dir, "diet_pairwise_summary.csv"),
          row.names = FALSE)
```

```{r session-info}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
