---
title: "Infection within diet DEGs"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  metadata_file: "data/metadata/mehreen_metadata_fixed.txt"
  counts_file: "data/raw_read_counts/master_counts.csv"
  eggnog_file: "data/GO_Annotations/GCF_023897955.1_iqSchGreg1.2_Arthopoda_one2one.emapper.annotations"
  gff_file: "data/reference/GCF_023897955.1_iqSchGreg1.2_genomic.gff"
  min_total_count: 10
  top_variable_genes: 500
  top_curated_degs_per_direction: 12
  top_functional_degs_per_theme: 18
  alpha: 0.05
  lfc_threshold: 1
  output_tag: "infection-within-diet-deg"
---

## Purpose

This page tests infected versus control samples separately within each diet.
This is a simple and readable complement to the full interaction model.

The reason to split by diet is biological and statistical: infection may not
produce the same transcriptional response under every nutritional condition.
The full interaction model tests whether diet modifies the infection effect;
this page then turns that question into three direct infected-control
comparisons that are easier to inspect, plot, and interpret.

## Setup and data loading

The code first matches metadata rows to count-matrix columns using the sample
labels. This makes the analysis self-contained in the R Markdown file and keeps
the model inputs visible before any statistics are run.

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c("DESeq2", "ggplot2", "EnhancedVolcano", "pheatmap", "ggrepel",
              "dplyr", "SummarizedExperiment")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}
library(ggplot2)

diet_colors <- c("33" = "#E6862E", "50" = "#2F8F5B", "83" = "#7B4FA3")
treatment_colors <- c("Control" = "#7B8794", "Infected" = "#C85A54")
treatment_fills <- c("Control" = "#D8DEE2", "Infected" = "#F0C7BF")
condition_surface_colors <- c(
  "33 Control" = "#F7D6A6",
  "33 Infected" = "#C75C1B",
  "50 Control" = "#CDE8D5",
  "50 Infected" = "#1F6F43",
  "83 Control" = "#D8C7E8",
  "83 Infected" = "#5B2E83"
)

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

metadata <- read.delim(file.path(project_dir, params$metadata_file),
                       check.names = FALSE)
metadata$label <- as.character(metadata$label)
metadata$Diet <- factor(metadata$Diet, levels = c("33", "50", "83"))
metadata$Treatment <- relevel(factor(metadata$Treatment), ref = "Control")

counts_raw <- read.csv(file.path(project_dir, params$counts_file),
                       check.names = FALSE)
counts <- as.matrix(counts_raw[, -1])
rownames(counts) <- counts_raw[[1]]
storage.mode(counts) <- "integer"

count_labels <- sub("_MERGE$", "", sub("^mehreen_", "", colnames(counts)))
metadata <- metadata[match(count_labels, metadata$label), ]
stopifnot(identical(metadata$label, count_labels))
rownames(metadata) <- metadata$label
colnames(counts) <- count_labels

keep_vst <- rowSums(counts) >= params$min_total_count
dds_vst <- DESeq2::DESeqDataSetFromMatrix(
  countData = counts[keep_vst, ],
  colData = metadata,
  design = ~ Diet + Treatment
)

vsd <- DESeq2::vst(dds_vst, blind = TRUE)
vst_mat <- SummarizedExperiment::assay(vsd)
```

## VST PCA with condition centroids

PCA is the most transparent ordination for this step because the axes show how
much expression variance they explain. UMAP can be useful for pattern discovery,
but it does not give the same direct variance accounting and can change with
parameter choices. Here the small points are individual libraries, the larger
points are diet-by-treatment centroids, and the pale condition polygons connect
the outermost samples from each diet-by-treatment group in the first two PCs.

```{r vst-pca-centroids, fig.height=5.8, fig.width=8}
nonzero_variance <- apply(vst_mat, 1, var) > 0
pca <- prcomp(t(vst_mat[nonzero_variance, , drop = FALSE]), scale. = FALSE)
percent_var <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)

pca_scores <- data.frame(
  label = rownames(pca$x),
  PC1 = pca$x[, "PC1"],
  PC2 = pca$x[, "PC2"],
  Diet = metadata$Diet,
  Treatment = metadata$Treatment,
  check.names = FALSE
)
pca_scores$Condition <- interaction(pca_scores$Diet, pca_scores$Treatment,
                                    sep = " ")

pca_centroids <- aggregate(cbind(PC1, PC2) ~ Diet + Treatment + Condition,
                           pca_scores, mean)
names(pca_centroids)[names(pca_centroids) == "PC1"] <- "centroid_PC1"
names(pca_centroids)[names(pca_centroids) == "PC2"] <- "centroid_PC2"

make_condition_hull <- function(df) {
  if (nrow(df) < 3) return(df[0, ])
  df[chull(df$PC1, df$PC2), ]
}

pca_hulls <- do.call(rbind, lapply(split(pca_scores, pca_scores$Condition),
                                   make_condition_hull))

p_pca_centroids <- ggplot2::ggplot() +
  ggplot2::geom_polygon(
    data = pca_hulls,
    ggplot2::aes(x = PC1, y = PC2, fill = Condition, group = Condition),
    alpha = 0.20,
    color = NA
  ) +
  ggplot2::geom_point(
    data = pca_scores,
    ggplot2::aes(x = PC1, y = PC2, color = Diet, shape = Treatment),
    size = 3,
    alpha = 0.88
  ) +
  ggplot2::geom_point(
    data = pca_centroids,
    ggplot2::aes(x = centroid_PC1, y = centroid_PC2, color = Diet),
    shape = 21,
    fill = "white",
    stroke = 0.8,
    size = 5
  ) +
  ggrepel::geom_label_repel(
    data = pca_centroids,
    ggplot2::aes(x = centroid_PC1, y = centroid_PC2, label = Condition),
    color = "grey15",
    fill = "white",
    label.size = 0.18,
    size = 3.1,
    min.segment.length = 0
  ) +
  ggplot2::scale_color_manual(values = diet_colors) +
  ggplot2::scale_fill_manual(values = condition_surface_colors) +
  ggplot2::theme_classic() +
  ggplot2::theme(legend.position = "bottom") +
  ggplot2::labs(
    title = "VST PCA with diet-by-treatment centroids",
    x = paste0("PC1: ", percent_var[1], "% variance"),
    y = paste0("PC2: ", percent_var[2], "% variance"),
    color = "Diet",
    fill = "Condition polygon",
    shape = "Treatment"
  )

p_pca_centroids
ggplot2::ggsave(file.path(out_dir, "vst_pca_condition_centroids.png"),
                p_pca_centroids, width = 8, height = 5.8, dpi = 300)
write.csv(pca_scores, file.path(out_dir, "vst_pca_sample_scores.csv"),
          row.names = FALSE)
write.csv(pca_centroids, file.path(out_dir, "vst_pca_condition_centroids.csv"),
          row.names = FALSE)
```

## VST heatmap across samples

Before reading the infected-control tests, it helps to see the expression
structure across all samples. This heatmap uses the most variable
variance-stabilized genes, with diet and treatment annotations above the
columns. The VST transformation is used because raw RNA-seq counts have
mean-dependent variance; without transformation, very highly expressed genes can
overwhelm the visual structure.

```{r vst-sample-heatmap, fig.height=8, fig.width=9}
gene_variance <- apply(vst_mat, 1, var)
top_n <- min(params$top_variable_genes, length(gene_variance))
top_genes <- names(sort(gene_variance, decreasing = TRUE))[seq_len(top_n)]

annotation_col <- data.frame(
  Diet = metadata$Diet,
  Treatment = metadata$Treatment
)
rownames(annotation_col) <- colnames(vst_mat)

png(file.path(out_dir, "vst_top_variable_gene_heatmap.png"),
    width = 1800, height = 1600, res = 180)
pheatmap::pheatmap(
  vst_mat[top_genes, ],
  scale = "row",
  show_rownames = FALSE,
  annotation_col = annotation_col,
  main = paste("Top", top_n, "variable VST genes across samples")
)
dev.off()

pheatmap::pheatmap(
  vst_mat[top_genes, ],
  scale = "row",
  show_rownames = FALSE,
  annotation_col = annotation_col,
  main = paste("Top", top_n, "variable VST genes across samples")
)
```

## Run one infected-control test per diet

For each diet, the code keeps only samples from that diet and fits a simple
`~ Treatment` DESeq2 model. This avoids averaging the infection effect across
diets. A positive log2 fold-change means higher expression in infected samples;
a negative log2 fold-change means higher expression in controls. Genes are
called significant only when they pass both the adjusted p-value threshold and
the absolute log2 fold-change threshold.

```{r infection-tests, fig.height=7, fig.width=8}
diet_results <- list()

for (diet_level in levels(droplevels(metadata$Diet))) {
  message("Running infection contrast for diet ", diet_level)

  sample_keep <- metadata$Diet == diet_level
  metadata_sub <- droplevels(metadata[sample_keep, ])
  counts_sub <- counts[, sample_keep]
  gene_keep <- rowSums(counts_sub) >= params$min_total_count

  dds <- DESeq2::DESeqDataSetFromMatrix(
    countData = counts_sub[gene_keep, ],
    colData = metadata_sub,
    design = ~ Treatment
  )
  dds <- DESeq2::DESeq(dds)

  res <- DESeq2::results(dds,
                         contrast = c("Treatment", "Infected", "Control"),
                         alpha = params$alpha)
  res_df <- as.data.frame(res)
  res_df$gene_id <- rownames(res_df)
  res_df$diet <- diet_level
  res_df$significant <- !is.na(res_df$padj) &
    res_df$padj < params$alpha &
    abs(res_df$log2FoldChange) >= params$lfc_threshold

  diet_results[[diet_level]] <- res_df
  write.csv(res_df, file.path(out_dir,
                              paste0("DESeq2_infected_vs_control_diet_",
                                     diet_level, ".csv")),
            row.names = FALSE)
  writeLines(res_df$gene_id[res_df$significant],
             file.path(out_dir,
                       paste0("significant_genes_infection_diet_",
                              diet_level, ".txt")))

  print(EnhancedVolcano::EnhancedVolcano(
    res_df,
    lab = res_df$gene_id,
    x = "log2FoldChange",
    y = "padj",
    pCutoff = params$alpha,
    FCcutoff = params$lfc_threshold,
    title = paste("Infected vs control in diet", diet_level)
  ))
}

all_infection_results <- do.call(rbind, diet_results)
write.csv(all_infection_results,
          file.path(out_dir, "all_infection_within_diet_results.csv"),
          row.names = FALSE)
```

## Diet-specific DEG PCA

The PCA above uses all filtered genes and all samples, so it describes global
expression structure. This section asks a narrower question for each diet: if we
look only at the genes that are differentially expressed between infected and
control samples within that diet, do the samples separate by treatment? This is
closer to the DEG result itself, and it avoids using genes from another diet to
explain the infected-control pattern.

```{r diet-specific-deg-pca, fig.height=5.2, fig.width=7.2}
make_deg_pca <- function(diet_level) {
  res <- diet_results[[diet_level]]
  deg_genes <- res$gene_id[res$significant & !is.na(res$gene_id)]
  deg_genes <- intersect(deg_genes, rownames(vst_mat))
  sample_keep <- metadata$Diet == diet_level
  metadata_sub <- droplevels(metadata[sample_keep, ])

  if (length(deg_genes) < 3 || nrow(metadata_sub) < 4) {
    return(NULL)
  }

  mat <- vst_mat[deg_genes, metadata_sub$label, drop = FALSE]
  keep_gene <- apply(mat, 1, var) > 0
  mat <- mat[keep_gene, , drop = FALSE]
  if (nrow(mat) < 3) return(NULL)

  pca_local <- prcomp(t(mat), scale. = FALSE)
  percent_local <- round(100 * (pca_local$sdev^2 / sum(pca_local$sdev^2)), 1)
  scores <- data.frame(
    label = rownames(pca_local$x),
    PC1 = pca_local$x[, "PC1"],
    PC2 = pca_local$x[, "PC2"],
    Diet = metadata_sub$Diet,
    Treatment = metadata_sub$Treatment,
    check.names = FALSE
  )

  centroids <- aggregate(cbind(PC1, PC2) ~ Treatment, scores, mean)
  names(centroids)[names(centroids) == "PC1"] <- "centroid_PC1"
  names(centroids)[names(centroids) == "PC2"] <- "centroid_PC2"

  treatment_hulls <- do.call(rbind, lapply(split(scores, scores$Treatment),
                                           make_condition_hull))

  p <- ggplot2::ggplot(scores, ggplot2::aes(PC1, PC2)) +
    ggplot2::geom_polygon(
      data = treatment_hulls,
      ggplot2::aes(x = PC1, y = PC2, fill = Treatment, group = Treatment),
      alpha = 0.22,
      color = NA,
      inherit.aes = FALSE
    ) +
    ggplot2::geom_point(ggplot2::aes(color = Treatment, shape = Treatment),
                        size = 3.2, alpha = 0.9) +
    ggplot2::geom_point(
      data = centroids,
      ggplot2::aes(x = centroid_PC1, y = centroid_PC2, color = Treatment),
      shape = 21,
      fill = "white",
      stroke = 0.8,
      size = 5,
      inherit.aes = FALSE
    ) +
    ggrepel::geom_label_repel(
      data = centroids,
      ggplot2::aes(x = centroid_PC1, y = centroid_PC2, label = Treatment),
      color = "grey15",
      fill = "white",
      label.size = 0.18,
      size = 3.2,
      min.segment.length = 0,
      inherit.aes = FALSE
    ) +
    ggplot2::scale_color_manual(values = treatment_colors) +
    ggplot2::scale_fill_manual(values = treatment_fills) +
    ggplot2::theme_classic() +
    ggplot2::theme(legend.position = "bottom") +
    ggplot2::labs(
      title = paste("Diet", diet_level, "PCA using infection DEGs only"),
      subtitle = paste(length(deg_genes), "significant genes"),
      x = paste0("PC1: ", percent_local[1], "% variance"),
      y = paste0("PC2: ", percent_local[2], "% variance"),
      color = "Treatment",
      fill = "Treatment",
      shape = "Treatment"
    )

  ggsave(file.path(out_dir,
                   paste0("diet_", diet_level, "_infection_deg_pca.png")),
         p, width = 7.2, height = 5.2, dpi = 300)
  write.csv(scores, file.path(out_dir,
                              paste0("diet_", diet_level,
                                     "_infection_deg_pca_scores.csv")),
            row.names = FALSE)
  write.csv(centroids, file.path(out_dir,
                                 paste0("diet_", diet_level,
                                        "_infection_deg_pca_centroids.csv")),
            row.names = FALSE)
  p
}

for (diet_level in levels(droplevels(metadata$Diet))) {
  p_local <- make_deg_pca(diet_level)
  if (is.null(p_local)) {
    cat("Diet", diet_level,
        "does not have enough significant, variable genes for DEG-only PCA.\n")
  } else {
    print(p_local)
  }
}
```

## DEG count barplot

The volcano plots are good for seeing whether a contrast has many small effects
or a few strong effects. This barplot answers the simpler question: how many
genes pass the current DEG thresholds in each diet, and are they higher in
infected or control samples?

```{r deg-count-barplot, fig.height=4.8, fig.width=7.5}
sig_infection <- all_infection_results[all_infection_results$significant &
                                         !is.na(all_infection_results$log2FoldChange), ]
sig_infection$diet <- factor(sig_infection$diet,
                             levels = levels(droplevels(metadata$Diet)))
sig_infection$direction <- ifelse(sig_infection$log2FoldChange > 0,
                                  "Higher in infected",
                                  "Higher in control")
sig_infection$direction <- factor(sig_infection$direction,
                                  levels = c("Higher in infected",
                                             "Higher in control"))

deg_bar_counts <- as.data.frame(table(
  diet = sig_infection$diet,
  direction = sig_infection$direction
))
names(deg_bar_counts)[3] <- "n_significant_genes"

p_deg_counts <- ggplot(deg_bar_counts,
                       aes(x = diet, y = n_significant_genes,
                           fill = direction)) +
  geom_col(position = position_dodge(width = 0.78), width = 0.68) +
  geom_text(aes(label = n_significant_genes),
            position = position_dodge(width = 0.78),
            vjust = -0.35, size = 3.2) +
  scale_fill_manual(values = c("Higher in infected" = treatment_colors[["Infected"]],
                               "Higher in control" = treatment_colors[["Control"]])) +
  theme_classic() +
  labs(
    title = "Significant infection DEGs within each diet",
    x = "Diet",
    y = "Number of significant genes",
    fill = "Direction"
  )

p_deg_counts
ggsave(file.path(out_dir, "infection_within_diet_deg_count_barplot.png"),
       p_deg_counts, width = 7.5, height = 4.8, dpi = 300)
write.csv(deg_bar_counts,
          file.path(out_dir, "infection_within_diet_direction_counts.csv"),
          row.names = FALSE)
```

## Publication-style PCA and DEG burden panels

These two panels mirror the AGY figure style. The PCA panel keeps individual
samples visible, marks diet-by-treatment centroids, and draws one arrow per diet
from the control centroid to the infected centroid. The DEG burden panel uses a
diverging axis so genes higher in infected samples and genes higher in controls
are immediately comparable.

```{r pca-centroid-arrows, fig.height=6.2, fig.width=10}
centroid_control <- pca_centroids[pca_centroids$Treatment == "Control", ]
centroid_infected <- pca_centroids[pca_centroids$Treatment == "Infected", ]
centroid_arrows <- merge(
  centroid_control[, c("Diet", "centroid_PC1", "centroid_PC2")],
  centroid_infected[, c("Diet", "centroid_PC1", "centroid_PC2")],
  by = "Diet",
  suffixes = c("_control", "_infected")
)

p_pca_arrows <- ggplot2::ggplot() +
  ggplot2::geom_polygon(
    data = pca_hulls,
    ggplot2::aes(x = PC1, y = PC2, fill = Condition, group = Condition),
    alpha = 0.16,
    color = NA
  ) +
  ggplot2::geom_point(
    data = pca_scores,
    ggplot2::aes(x = PC1, y = PC2, color = Diet, shape = Treatment),
    size = 3,
    alpha = 0.86
  ) +
  ggplot2::geom_segment(
    data = centroid_arrows,
    ggplot2::aes(x = centroid_PC1_control, y = centroid_PC2_control,
                 xend = centroid_PC1_infected, yend = centroid_PC2_infected,
                 color = Diet),
    linewidth = 0.85,
    arrow = grid::arrow(length = grid::unit(0.18, "cm"))
  ) +
  ggplot2::geom_point(
    data = pca_centroids,
    ggplot2::aes(x = centroid_PC1, y = centroid_PC2, color = Diet),
    shape = 21,
    fill = "white",
    stroke = 1,
    size = 5
  ) +
  ggrepel::geom_label_repel(
    data = pca_centroids,
    ggplot2::aes(x = centroid_PC1, y = centroid_PC2, label = Condition),
    color = "grey15",
    fill = "white",
    label.size = 0.18,
    size = 3.0,
    min.segment.length = 0
  ) +
  ggplot2::scale_color_manual(values = diet_colors) +
  ggplot2::scale_fill_manual(values = condition_surface_colors) +
  ggplot2::theme_classic() +
  ggplot2::theme(legend.position = "bottom",
                 legend.box = "vertical") +
  ggplot2::guides(
    fill = ggplot2::guide_legend(nrow = 2),
    color = ggplot2::guide_legend(nrow = 1),
    shape = ggplot2::guide_legend(nrow = 1)
  ) +
  ggplot2::labs(
    title = "PCA centroid shift after infection",
    subtitle = "Arrows connect control and infected centroids within each diet",
    x = paste0("PC1: ", percent_var[1], "% variance"),
    y = paste0("PC2: ", percent_var[2], "% variance"),
    color = "Diet",
    fill = "Condition polygon",
    shape = "Treatment"
  )

p_pca_arrows
ggplot2::ggsave(file.path(out_dir, "pca_centroid_infection_arrows.png"),
                p_pca_arrows, width = 10, height = 6.2, dpi = 300)
write.csv(centroid_arrows,
          file.path(out_dir, "pca_centroid_infection_arrows.csv"),
          row.names = FALSE)
```

```{r diverging-deg-burden, fig.height=4.8, fig.width=8.6}
deg_burden <- deg_bar_counts
deg_burden$diet <- factor(deg_burden$diet,
                          levels = levels(droplevels(metadata$Diet)))
deg_burden$signed_n <- ifelse(deg_burden$direction == "Higher in infected",
                              deg_burden$n_significant_genes,
                              -deg_burden$n_significant_genes)
max_burden <- max(abs(deg_burden$signed_n), na.rm = TRUE)

p_deg_burden <- ggplot2::ggplot(
  deg_burden,
  ggplot2::aes(x = signed_n, y = diet, fill = direction)
) +
  ggplot2::geom_vline(xintercept = 0, linewidth = 0.35, color = "grey35") +
  ggplot2::geom_col(width = 0.68) +
  ggplot2::geom_text(
    ggplot2::aes(label = n_significant_genes,
                 hjust = ifelse(signed_n > 0, -0.15, 1.15)),
    size = 3.2
  ) +
  ggplot2::scale_x_continuous(labels = abs,
                              limits = c(-max_burden * 1.25,
                                         max_burden * 1.25)) +
  ggplot2::scale_fill_manual(values = c("Higher in infected" = treatment_colors[["Infected"]],
                                        "Higher in control" = treatment_colors[["Control"]])) +
  ggplot2::theme_classic() +
  ggplot2::labs(
    title = "Direction-aware infection DEG burden",
    subtitle = "Right: higher in infected; left: higher in control",
    x = "Number of significant genes",
    y = "Diet",
    fill = "Direction"
  )

p_deg_burden
ggplot2::ggsave(file.path(out_dir, "infection_deg_burden_diverging_barplot.png"),
                p_deg_burden, width = 8.6, height = 4.8, dpi = 300)
write.csv(deg_burden,
          file.path(out_dir, "infection_deg_burden_diverging_counts.csv"),
          row.names = FALSE)
```

## Curated DEG VST heatmap

The full VST heatmap is a broad QC view. This heatmap is narrower: it selects
the strongest infected-control DEGs from each diet and direction, then adds
significant DEGs with eggNOG descriptions matching immune or nutrition-related
keywords. This follows the same idea as the AGY candidate heatmaps: keep the
statistical DEG signal visible, but annotate genes that are biologically
relevant to the manuscript question. Samples are ordered by infection status
first and then diet instead of clustered, so the infected profiles appear
together and diet-specific substructure can be compared directly. Genes are
ordered first by how broadly they are detected as DEGs across diets, and then by
their functional theme.

```{r curated-deg-heatmap, fig.height=9, fig.width=10}
sig_for_heatmap <- all_infection_results[all_infection_results$significant &
                                           !is.na(all_infection_results$padj) &
                                           !is.na(all_infection_results$log2FoldChange), ]
sig_for_heatmap$direction <- ifelse(sig_for_heatmap$log2FoldChange > 0,
                                    "Higher in infected",
                                    "Higher in control")
sig_for_heatmap <- sig_for_heatmap[order(sig_for_heatmap$padj,
                                         -abs(sig_for_heatmap$log2FoldChange)), ]

select_top_degs <- function(tab, n) {
  tab <- tab[order(tab$padj, -abs(tab$log2FoldChange)), ]
  head(tab, n)
}

curated_pieces <- split(sig_for_heatmap,
                        paste(sig_for_heatmap$diet,
                              sig_for_heatmap$direction, sep = "__"))
top_deg_rows <- do.call(rbind, lapply(curated_pieces, select_top_degs,
                                      n = params$top_curated_degs_per_direction))
top_deg_rows$Selection <- "Top DEG"

read_emapper <- function(path) {
  cols <- c(
    "query", "seed_ortholog", "evalue", "score", "eggNOG_OGs",
    "max_annot_lvl", "COG_category", "Description", "Preferred_name",
    "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
    "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC", "CAZy",
    "BiGG_Reaction", "PFAMs"
  )
  tab <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", fill = TRUE)
  names(tab) <- cols[seq_len(ncol(tab))]
  tab$protein_id <- sub(".*(XP_[0-9]+[.][0-9]+).*", "\\1", tab$query)
  tab
}

parse_cds_gene_bridge <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  cds <- gff[gff$type == "CDS", ]
  data.frame(
    protein_id = sub(".*protein_id=([^;]+).*", "\\1", cds$attributes),
    gene_id = sub(".*gene=([^;]+).*", "\\1", cds$attributes),
    stringsAsFactors = FALSE
  ) |>
    dplyr::filter(grepl("^XP_", protein_id), grepl("^LOC", gene_id)) |>
    dplyr::distinct()
}

eggnog_path <- file.path(project_dir, params$eggnog_file)
gff_path <- file.path(project_dir, params$gff_file)
if (file.exists(eggnog_path) && file.exists(gff_path)) {
  eggnog <- read_emapper(eggnog_path)
  protein_to_gene <- parse_cds_gene_bridge(gff_path)
  gene_annotation <- eggnog |>
    dplyr::left_join(protein_to_gene, by = "protein_id") |>
    dplyr::filter(!is.na(gene_id)) |>
    dplyr::distinct(gene_id, .keep_all = TRUE) |>
    dplyr::select(gene_id, Preferred_name, Description, GOs, KEGG_ko,
                  KEGG_Pathway, PFAMs)
} else {
  gene_annotation <- data.frame(gene_id = character(),
                                Preferred_name = character(),
                                Description = character(),
                                GOs = character(),
                                KEGG_ko = character(),
                                KEGG_Pathway = character(),
                                PFAMs = character())
}

annotated_sig <- sig_for_heatmap |>
  dplyr::left_join(gene_annotation, by = "gene_id")
annotation_text <- apply(annotated_sig[, c("Preferred_name", "Description",
                                           "GOs", "KEGG_ko", "KEGG_Pathway",
                                           "PFAMs")],
                         1, paste, collapse = " ")

immune_keywords <- paste(c(
  "immune", "immun", "defensin", "antimicrobial", "lysozyme",
  "phenoloxidase", "prophenoloxidase", "toll", "imd", "jak", "stat",
  "lectin", "peptidoglycan", "hemocyte", "serpin", "clip", "thioester",
  "pathogen", "fungal", "fungus", "melanization"
), collapse = "|")

nutrition_keywords <- paste(c(
  "nutrition", "nutrient", "metabolism", "metabolic", "lipid", "fatty",
  "carbohydrate", "glucose", "glycogen", "trehalose", "insulin",
  "amino acid", "protease", "peptidase", "trypsin", "chymotrypsin",
  "lipase", "amylase", "digestion", "digestive", "transporter"
), collapse = "|")

annotated_sig$Functional_theme <- dplyr::case_when(
  grepl(immune_keywords, annotation_text, ignore.case = TRUE) ~ "Immune-related",
  grepl(nutrition_keywords, annotation_text, ignore.case = TRUE) ~ "Nutrition/metabolism",
  TRUE ~ "Other"
)

theme_rows <- annotated_sig |>
  dplyr::filter(Functional_theme != "Other") |>
  dplyr::group_by(Functional_theme) |>
  dplyr::arrange(padj, dplyr::desc(abs(log2FoldChange)), .by_group = TRUE) |>
  dplyr::slice_head(n = params$top_functional_degs_per_theme) |>
  dplyr::ungroup() |>
  dplyr::mutate(Selection = Functional_theme)

gene_deg_summary <- sig_for_heatmap |>
  dplyr::group_by(gene_id) |>
  dplyr::summarise(
    DEG_diet_count = dplyr::n_distinct(diet),
    DEG_diets = paste(sort(unique(as.character(diet))), collapse = ", "),
    DEG_direction_summary = dplyr::if_else(
      dplyr::n_distinct(direction) == 1,
      dplyr::first(direction),
      "Mixed directions"
    ),
    min_padj = min(padj, na.rm = TRUE),
    max_abs_lfc = max(abs(log2FoldChange), na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    DEG_status = dplyr::case_when(
      DEG_diet_count == length(levels(droplevels(metadata$Diet))) ~
        "DEG in all diets",
      DEG_diet_count > 1 ~ "DEG in multiple diets",
      DEG_diet_count == 1 ~ "DEG in one diet",
      TRUE ~ "Functional candidate (not DEG)"
    )
  )

deg_status_levels <- c("DEG in all diets", "DEG in multiple diets",
                       "DEG in one diet", "Functional candidate (not DEG)")
functional_theme_levels <- c("Immune-related", "Nutrition/metabolism",
                             "Top DEG", "Other")

curated_degs <- dplyr::bind_rows(top_deg_rows, theme_rows) |>
  dplyr::left_join(gene_annotation, by = "gene_id", suffix = c("", ".anno")) |>
  dplyr::left_join(gene_deg_summary, by = "gene_id") |>
  dplyr::mutate(
    Preferred_name = dplyr::coalesce(Preferred_name, Preferred_name.anno),
    Description = dplyr::coalesce(Description, Description.anno),
    GOs = dplyr::coalesce(GOs, GOs.anno),
    KEGG_ko = dplyr::coalesce(KEGG_ko, KEGG_ko.anno),
    KEGG_Pathway = dplyr::coalesce(KEGG_Pathway, KEGG_Pathway.anno),
    PFAMs = dplyr::coalesce(PFAMs, PFAMs.anno),
    curated_annotation_text = paste(Preferred_name, Description, GOs, KEGG_ko,
                                    KEGG_Pathway, PFAMs),
    Functional_theme = dplyr::case_when(
      grepl(immune_keywords, curated_annotation_text, ignore.case = TRUE) ~
        "Immune-related",
      grepl(nutrition_keywords, curated_annotation_text, ignore.case = TRUE) ~
        "Nutrition/metabolism",
      Selection %in% c("Immune-related", "Nutrition/metabolism") ~ Selection,
      TRUE ~ "Top DEG"
    ),
    DEG_status = factor(DEG_status, levels = deg_status_levels),
    Functional_theme = factor(Functional_theme, levels = functional_theme_levels)
  ) |>
  dplyr::arrange(DEG_status, Functional_theme, DEG_direction_summary,
                 DEG_diets, min_padj, dplyr::desc(max_abs_lfc)) |>
  dplyr::distinct(gene_id, .keep_all = TRUE)

curated_genes <- curated_degs$gene_id[curated_degs$gene_id %in% rownames(vst_mat)]
treatment_order <- factor(as.character(metadata$Treatment),
                          levels = c("Infected", "Control"))
sample_order <- metadata$label[order(treatment_order, metadata$Diet,
                                     metadata$label)]

sample_annotation <- data.frame(
  Treatment = metadata[sample_order, "Treatment"],
  Diet = metadata[sample_order, "Diet"]
)
rownames(sample_annotation) <- sample_order

gene_annotation_display <- curated_degs[match(curated_genes, curated_degs$gene_id),
                                        c("DEG_status",
                                          "DEG_direction_summary",
                                          "Selection",
                                          "Functional_theme"), drop = FALSE]
names(gene_annotation_display) <- c("DEG_status", "DEG_direction",
                                    "Selection", "Functional_theme")
rownames(gene_annotation_display) <- curated_genes

label_source <- curated_degs[match(curated_genes, curated_degs$gene_id), ]
label_text <- ifelse(!is.na(label_source$Preferred_name) &
                       label_source$Preferred_name != "-",
                     label_source$Preferred_name,
                     label_source$Description)
label_text[is.na(label_text) | label_text == "-" | label_text == ""] <- "unannotated"
label_text <- substr(label_text, 1, 45)
col_labels <- paste(curated_genes, label_text, sep = " | ")

find_gaps <- function(groups) {
  groups <- as.character(groups)
  if (length(groups) < 2) {
    return(integer())
  }
  which(groups[-length(groups)] != groups[-1])
}

sample_gaps <- find_gaps(paste(sample_annotation$Treatment,
                               sample_annotation$Diet))
gene_gaps <- find_gaps(paste(gene_annotation_display$DEG_status,
                             gene_annotation_display$Functional_theme))

annotation_colors <- list(
  Diet = diet_colors,
  Treatment = treatment_colors,
  DEG_direction = c("Higher in infected" = treatment_colors[["Infected"]],
                    "Higher in control" = treatment_colors[["Control"]],
                    "Mixed directions" = "#4B5563"),
  DEG_status = c("DEG in all diets" = "#111827",
                 "DEG in multiple diets" = "#4B5563",
                 "DEG in one diet" = "#9CA3AF",
                 "Functional candidate (not DEG)" = "#E5E7EB"),
  Selection = c("Top DEG" = "#374151",
                "Immune-related" = "#237C8E",
                "Nutrition/metabolism" = "#B8643C"),
  Functional_theme = c("Top DEG" = "#374151",
                       "Immune-related" = "#237C8E",
                       "Nutrition/metabolism" = "#B8643C",
                       "Other" = "#9CA3AF")
)

heatmap_expression_colors <- grDevices::colorRampPalette(
  c("#237C8E", "#F7F3E8", "#C85A54")
)(101)
heatmap_expression_breaks <- seq(-3, 3, length.out = 102)

heatmap_display <- pheatmap::pheatmap(
  t(vst_mat[curated_genes, sample_order, drop = FALSE]),
  scale = "column",
  color = heatmap_expression_colors,
  breaks = heatmap_expression_breaks,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  gaps_row = sample_gaps,
  gaps_col = gene_gaps,
  show_rownames = TRUE,
  show_colnames = FALSE,
  labels_col = col_labels,
  fontsize_row = 6,
  annotation_col = gene_annotation_display,
  annotation_row = sample_annotation,
  annotation_colors = annotation_colors,
  main = "Curated infection, immune, and nutrition DEGs",
  filename = file.path(out_dir, "curated_infection_deg_vst_heatmap.png"),
  width = 14,
  height = 11
)

display_dir <- file.path(project_dir, "docs", "figure",
                         "06-infection-within-diet-deg.Rmd")
dir.create(display_dir, recursive = TRUE, showWarnings = FALSE)
display_png <- file.path(display_dir, "curated-deg-heatmap-display.png")
file.copy(file.path(out_dir, "curated_infection_deg_vst_heatmap.png"),
          display_png, overwrite = TRUE)
curated_heatmap_src <- file.path("figure", "06-infection-within-diet-deg.Rmd",
                                 "curated-deg-heatmap-display.png")

write.csv(curated_degs,
          file.path(out_dir, "curated_infection_deg_heatmap_genes.csv"),
          row.names = FALSE)
```

```{r curated-deg-heatmap-display, echo=FALSE, results='asis'}
cat(paste0(
  '<p><img src="', curated_heatmap_src,
  '" alt="Curated infection, immune, and nutrition DEG VST heatmap" ',
  'style="display:block;margin:auto;max-width:100%;height:auto;" /></p>'
))
```

## Curated immune and nutrition DEG summary

This compact figure uses the same curated gene table as the heatmap, but
summarizes it by diet, direction, and biological theme. It is easier to read
quickly than the full heatmap and is the better candidate for a publication-style
overview panel.

```{r curated-theme-barplot, fig.height=5.2, fig.width=8.5}
curated_summary <- curated_degs |>
  dplyr::filter(Functional_theme %in% c("Immune-related",
                                        "Nutrition/metabolism")) |>
  dplyr::mutate(
    diet = factor(diet, levels = levels(droplevels(metadata$Diet))),
    direction = factor(direction, levels = c("Higher in infected",
                                             "Higher in control")),
    Functional_theme = factor(Functional_theme,
                              levels = c("Immune-related",
                                         "Nutrition/metabolism"))
  ) |>
  dplyr::count(diet, direction, Functional_theme,
               name = "n_curated_degs")

p_curated_summary <- ggplot(curated_summary,
                            aes(x = diet, y = n_curated_degs,
                                fill = Functional_theme)) +
  geom_col(position = position_dodge(width = 0.76), width = 0.68) +
  geom_text(aes(label = n_curated_degs),
            position = position_dodge(width = 0.76),
            vjust = -0.35, size = 3.1) +
  facet_wrap(~ direction) +
  scale_fill_manual(values = c("Immune-related" = "#237C8E",
                               "Nutrition/metabolism" = "#B8643C")) +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(
    title = "Curated immune and nutrition DEGs",
    subtitle = "Significant infected-control DEGs selected from eggNOG descriptions",
    x = "Diet",
    y = "Number of curated DEGs",
    fill = "Functional theme"
  )

p_curated_summary
ggsave(file.path(out_dir, "curated_immune_nutrition_deg_barplot.png"),
       p_curated_summary, width = 8.5, height = 5.2, dpi = 300)
write.csv(curated_summary,
          file.path(out_dir, "curated_immune_nutrition_deg_summary.csv"),
          row.names = FALSE)
```

## Summary

The summary table collapses each diet-specific result into a total DEG count.
It is deliberately simple, because the direction-aware barplot above carries
the visual comparison and the full CSV files keep the gene-level statistics.

```{r summary}
summary_table <- aggregate(significant ~ diet, all_infection_results, sum)
names(summary_table)[2] <- "n_significant_genes"
knitr::kable(summary_table)

write.csv(summary_table, file.path(out_dir, "infection_within_diet_summary.csv"),
          row.names = FALSE)
```

```{r session-info}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
