---
title: "DEG overlap and prioritization"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  results_root: "output/rmd_runs"
  alpha: 0.05
  lfc_threshold: 1
  output_tag: "deg-overlap-prioritization"
---

## Purpose

This page compares significant DEG lists from the model pages. It can be run
after the DESeq2 pages create result CSV files under `output/rmd_runs/`.

The goal is not to run a new statistical test. Instead, this page asks whether
the same genes repeatedly appear across diet contrasts, infection contrasts, and
model coefficients. Genes that recur across analyses can be useful candidates
for follow-up because they are less dependent on a single contrast definition.

## Setup

The page reads DEG result files from the timestamped run folders produced by the
analysis pages. This keeps the overlap step reproducible while avoiding hidden
helper scripts: the thresholds used to define a DEG remain visible in the YAML
header.

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c("ggplot2")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}
library(ggplot2)

overlap_colors <- c("Present" = "#237C8E", "Absent" = "grey92")
bar_color <- "#5B6B4A"

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

results_root <- file.path(project_dir, params$results_root)
```

## Find result files

Only CSV files that look like DESeq2 result tables are selected. The table below
is shown so the reader can see exactly which run outputs contributed to the
overlap analysis.

```{r find-files}
all_result_files <- list.files(results_root, pattern = "\\.csv$", recursive = TRUE,
                               full.names = TRUE)
all_result_files <- all_result_files[grepl("DESeq2_|all_.*results|coefficients",
                                           basename(all_result_files))]

file_info <- data.frame(
  file = all_result_files,
  short_name = basename(all_result_files),
  modified = file.info(all_result_files)$mtime,
  stringsAsFactors = FALSE
)
file_info <- file_info[order(file_info$short_name,
                             -as.numeric(file_info$modified)), ]

# Keep the newest run for each result table so old workflowR runs do not
# duplicate the same contrast many times in the overlap figures.
file_info <- file_info[!duplicated(file_info$short_name), ]
result_files <- file_info$file

file_table <- data.frame(
  file = result_files,
  short_name = basename(result_files),
  modified = file.info(result_files)$mtime
)

knitr::kable(file_table)
```

## Build DEG sets

Each result table is converted into a set of significant gene IDs. If a table
already contains a `significant` column, that column is used directly; otherwise
the page reconstructs the DEG call from adjusted p-value and log2 fold-change.
This lets the overlap page work with both the current self-contained Rmd outputs
and older result tables.

```{r build-sets}
read_deg_set <- function(path) {
  tab <- read.csv(path, check.names = FALSE)
  gene_col <- intersect(c("gene_id", "gene", "GeneID"), names(tab))[1]
  if (is.na(gene_col)) return(NULL)

  if ("significant" %in% names(tab)) {
    sig <- tab$significant
  } else if (all(c("padj", "log2FoldChange") %in% names(tab))) {
    sig <- !is.na(tab$padj) &
      tab$padj < params$alpha &
      abs(tab$log2FoldChange) >= params$lfc_threshold
  } else {
    return(NULL)
  }

  unique(as.character(tab[[gene_col]][sig]))
}

deg_sets <- lapply(result_files, read_deg_set)
names(deg_sets) <- tools::file_path_sans_ext(basename(result_files))
deg_sets <- deg_sets[lengths(deg_sets) > 0]

set_summary <- data.frame(
  set = names(deg_sets),
  n_genes = lengths(deg_sets),
  stringsAsFactors = FALSE
)
set_summary <- set_summary[order(set_summary$n_genes, decreasing = TRUE), ]
deg_sets <- deg_sets[set_summary$set]

knitr::kable(set_summary)
write.csv(set_summary, file.path(out_dir, "deg_set_summary.csv"),
          row.names = FALSE)
```

## DEG set size plot

This first figure is the fastest way to see which contrasts are driving most of
the DEG signal. It counts each DEG set independently, before asking whether
genes overlap between sets.

```{r deg-set-size-plot, fig.height=5.8, fig.width=9}
set_summary$set <- factor(set_summary$set, levels = rev(set_summary$set))

p_set_size <- ggplot(set_summary, aes(x = set, y = n_genes)) +
  geom_col(fill = bar_color, width = 0.72) +
  geom_text(aes(label = n_genes), hjust = -0.12, size = 3) +
  coord_flip() +
  theme_classic() +
  labs(
    title = "DEG set sizes",
    x = "DEG set",
    y = "Number of significant genes"
  )

p_set_size
ggsave(file.path(out_dir, "deg_set_size_barplot.png"),
       p_set_size, width = 9, height = 5.8, dpi = 300)
```

## Presence-absence matrix

The presence-absence matrix records whether each DEG appears in each result
set. The priority table then lists genes that occur in two or more sets, which
is a practical starting point for annotation, enrichment, and candidate-gene
figures.

```{r overlap-matrix}
if (length(deg_sets) == 0) {
  stop("No DEG sets found. Knit the DESeq2 pages first, then rerun this page.")
}

all_genes <- sort(unique(unlist(deg_sets)))
presence <- sapply(deg_sets, function(x) all_genes %in% x)
presence_df <- data.frame(gene_id = all_genes, presence, check.names = FALSE)

write.csv(presence_df, file.path(out_dir, "deg_presence_absence_matrix.csv"),
          row.names = FALSE)

overlap_count <- rowSums(presence)
priority_genes <- data.frame(
  gene_id = all_genes[overlap_count >= 2],
  n_sets = overlap_count[overlap_count >= 2],
  set_members = apply(presence[overlap_count >= 2, , drop = FALSE], 1,
                      function(x) paste(names(deg_sets)[x], collapse = "; ")),
  stringsAsFactors = FALSE
)
priority_genes <- priority_genes[order(priority_genes$n_sets, decreasing = TRUE), ]

knitr::kable(head(priority_genes, 25))
write.csv(priority_genes, file.path(out_dir, "genes_seen_in_two_or_more_sets.csv"),
          row.names = FALSE)
```

## Intersection size plot

This plot counts overlap patterns directly. A bar means that a group of genes is
shared by exactly the DEG sets named on the y-axis. This is easier to read than
a full gene-by-set heatmap when the result files contain many contrasts.

```{r intersection-size-plot, fig.height=6.5, fig.width=9}
intersection_labels <- apply(presence, 1, function(x) {
  paste(names(deg_sets)[x], collapse = " + ")
})
intersection_labels <- intersection_labels[nzchar(intersection_labels)]
intersection_summary <- as.data.frame(table(intersection = intersection_labels),
                                      stringsAsFactors = FALSE)
names(intersection_summary)[2] <- "n_genes"
intersection_summary$n_sets <- vapply(strsplit(intersection_summary$intersection,
                                               " [+] "),
                                      length, integer(1))
intersection_summary <- intersection_summary[order(intersection_summary$n_genes,
                                                   intersection_summary$n_sets,
                                                   decreasing = TRUE), ]
intersection_summary <- head(intersection_summary, 20)
intersection_summary$intersection_id <- sprintf("I%02d", seq_len(nrow(intersection_summary)))
intersection_summary$intersection_id <- factor(intersection_summary$intersection_id,
                                               levels = rev(intersection_summary$intersection_id))

p_intersections <- ggplot(intersection_summary,
                          aes(x = intersection_id, y = n_genes,
                              fill = factor(n_sets))) +
  geom_col(width = 0.72) +
  geom_text(aes(label = n_genes), hjust = -0.12, size = 3) +
  coord_flip() +
  scale_fill_manual(values = c("1" = "#7B8794", "2" = "#237C8E",
                               "3" = "#5B6B4A", "4" = "#B8643C",
                               "5" = "#8C6D9E", "6" = "#C85A54"),
                    na.value = "#374151") +
  theme_classic() +
  labs(
    title = "Top DEG intersections",
    subtitle = "Bars show exact overlap patterns across the latest DEG result tables",
    x = "Intersection ID",
    y = "Number of genes",
    fill = "Number of sets"
  )

p_intersections
ggsave(file.path(out_dir, "deg_intersection_size_barplot.png"),
       p_intersections, width = 9, height = 6.5, dpi = 300)
write.csv(intersection_summary,
          file.path(out_dir, "deg_intersection_summary.csv"),
          row.names = FALSE)

knitr::kable(intersection_summary[, c("intersection_id", "n_genes",
                                      "n_sets", "intersection")],
             caption = "Intersection IDs used in the plot")
```

## Overlap plot

The heatmap displays the densest part of the presence-absence matrix so the
website stays readable. Blue cells mean the gene is significant in that DEG set;
grey cells mean it is absent from that set under the current thresholds.

```{r overlap-plot, fig.height=6, fig.width=9}
plot_df <- data.frame(
  gene_id = rep(all_genes, times = length(deg_sets)),
  set = rep(names(deg_sets), each = length(all_genes)),
  present = as.vector(presence)
)

# Plot the densest part of the matrix so the website stays readable.
top_genes <- names(sort(overlap_count, decreasing = TRUE))[seq_len(min(100, length(overlap_count)))]
plot_df <- plot_df[plot_df$gene_id %in% top_genes, ]

p_overlap <- ggplot2::ggplot(plot_df,
                             ggplot2::aes(x = set, y = gene_id, fill = present)) +
  ggplot2::geom_tile() +
  ggplot2::scale_fill_manual(values = c("TRUE" = overlap_colors[["Present"]],
                                        "FALSE" = overlap_colors[["Absent"]])) +
  ggplot2::theme_classic() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
                 axis.text.y = ggplot2::element_blank(),
                 axis.ticks.y = ggplot2::element_blank()) +
  ggplot2::labs(
    title = "DEG overlap matrix",
    subtitle = "Top genes ranked by number of sets where they appear",
    x = "DEG set",
    y = "Genes",
    fill = "Present"
  )

p_overlap
ggplot2::ggsave(file.path(out_dir, "deg_overlap_matrix.png"),
                p_overlap, width = 9, height = 6, dpi = 300)
```

```{r session-info}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
