---
title: "Publication figure exports"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  metadata_file: "data/metadata/mehreen_metadata_fixed.txt"
  counts_file: "data/raw_read_counts/master_counts_no_rRNA.csv"
  gff_file: "data/reference/GCF_023897955.1_iqSchGreg1.2_genomic.gff"
  eggnog_file: "data/GO_Annotations/GCF_023897955.1_iqSchGreg1.2_Arthopoda_one2one.emapper.annotations"
  latest_results_root: "output/rmd_runs"
  min_total_count: 10
  alpha: 0.05
  lfc_threshold: 1
  output_tag: "publication-figures"
---

## Purpose

This page assembles manuscript-facing panels from the all-samples retained
analysis. It does not rerun the raw read-processing pipeline. It rebuilds plots
from the rRNA-filtered count matrix, metadata, GFF biotypes, and the current
all-samples DESeq2 result table.

Every knit writes to a new folder under `output/rmd_runs`, so previous figure
drafts remain available for comparison.

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c("DESeq2", "SummarizedExperiment", "dplyr", "tidyr", "stringr",
              "ggplot2", "ggrepel", "patchwork", "scales",
              "ggVennDiagram", "png", "grid")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

library(ggplot2)
library(patchwork)

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

results_root <- file.path(project_dir, params$latest_results_root)
metadata_file <- file.path(project_dir, params$metadata_file)
counts_file <- file.path(project_dir, params$counts_file)
gff_file <- file.path(project_dir, params$gff_file)
eggnog_file <- file.path(project_dir, params$eggnog_file)

diet_colors <- c("33" = "#E6862E", "50" = "#2F8F5B", "83" = "#7B4FA3")
treatment_colors <- c("Control" = "#FAEBD7", "Infected" = "#8B0000")
condition_colors <- c(
  "33 Control" = "#F7D6A6",
  "33 Infected" = "#C75C1B",
  "50 Control" = "#CDE8D5",
  "50 Infected" = "#1F6F43",
  "83 Control" = "#D8C7E8",
  "83 Infected" = "#5B2E83"
)

direction_biotype_colors <- c(
  "Upregulated - Coding genes" = "#B91C1C",
  "Upregulated - Long non-coding RNA" = "#F87171",
  "Upregulated - Other / unknown" = "#FCA5A5",
  "Downregulated - Coding genes" = "#1D4ED8",
  "Downregulated - Long non-coding RNA" = "#60A5FA",
  "Downregulated - Other / unknown" = "#BFDBFE"
)

physiology_theme_colors <- c(
  "Phase-related changes" = "#C51B7D",
  "Antimicrobial peptides" = "#D7191C",
  "Immune-related" = "#F58220",
  "Glycogen synthesis / storage" = "#FFD92F",
  "Lipid synthesis / metabolism" = "#1B9E77",
  "Protein synthesis / storage" = "#377EB8",
  "Detoxification / stress" = "#984EA3",
  "Other / unannotated genes" = "#BDBDBD"
)

theme_pub <- function(base_size = 12) {
  ggplot2::theme_classic(base_size = base_size) +
    ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold"),
      strip.background = ggplot2::element_rect(fill = "#F3F4F6",
                                               color = "#111827"),
      strip.text = ggplot2::element_text(face = "bold"),
      legend.position = "bottom",
      legend.title = ggplot2::element_text(face = "bold")
    )
}
```

## Helper functions

```{r helper-functions}
latest_dir <- function(pattern) {
  dirs <- list.dirs(results_root, recursive = FALSE, full.names = TRUE)
  dirs <- dirs[grepl(pattern, basename(dirs))]
  if (length(dirs) == 0) return(NA_character_)
  dirs[which.max(file.info(dirs)$mtime)]
}

save_png_svg <- function(plot, file_base, width, height, dpi = 300) {
  png_path <- file.path(out_dir, paste0(file_base, ".png"))
  svg_path <- file.path(out_dir, paste0(file_base, ".svg"))

  ggplot2::ggsave(png_path, plot, width = width, height = height,
                  dpi = dpi, limitsize = FALSE)

  grDevices::svg(svg_path, width = width, height = height, onefile = FALSE)
  on.exit(grDevices::dev.off(), add = TRUE)
  print(plot)

  invisible(c(png = png_path, svg = svg_path))
}

parse_gff_gene_annotation <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")

  gff |>
    dplyr::filter(type == "gene") |>
    dplyr::transmute(
      gene_id = dplyr::coalesce(
        stringr::str_match(attributes, "gene=([^;]+)")[, 2],
        stringr::str_match(attributes, "Name=([^;]+)")[, 2]
      ),
      Description = stringr::str_match(attributes, "description=([^;]+)")[, 2],
      gene_biotype = stringr::str_match(attributes, "gene_biotype=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(gene_id)) |>
    dplyr::mutate(
      Description = dplyr::if_else(
        is.na(Description) | Description == "",
        "no GFF description",
        utils::URLdecode(Description)
      ),
      gene_biotype = dplyr::coalesce(gene_biotype, "unknown")
    ) |>
    dplyr::distinct(gene_id, .keep_all = TRUE)
}

parse_gff_gene_coordinates <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")

  sequence_regions <- gff |>
    dplyr::filter(type == "region") |>
    dplyr::transmute(
      seqid,
      seq_start = start,
      seq_end = end,
      chromosome = stringr::str_match(attributes, "chromosome=([^;]+)")[, 2],
      region_name = stringr::str_match(attributes, "Name=([^;]+)")[, 2],
      genome = stringr::str_match(attributes, "genome=([^;]+)")[, 2]
    ) |>
    dplyr::mutate(
      chromosome = dplyr::coalesce(chromosome, region_name, seqid),
      chromosome = utils::URLdecode(chromosome),
      chromosome_class = dplyr::if_else(
        chromosome %in% c(as.character(1:11), "X"),
        "Named chromosome",
        "Unplaced scaffold"
      )
    ) |>
    dplyr::distinct(seqid, .keep_all = TRUE)

  gene_ranges <- gff |>
    dplyr::filter(type == "gene") |>
    dplyr::transmute(
      seqid,
      gene_id = dplyr::coalesce(
        stringr::str_match(attributes, "gene=([^;]+)")[, 2],
        stringr::str_match(attributes, "Name=([^;]+)")[, 2]
      ),
      start,
      end,
      midpoint = (start + end) / 2,
      strand
    ) |>
    dplyr::filter(!is.na(gene_id)) |>
    dplyr::distinct(gene_id, .keep_all = TRUE) |>
    dplyr::left_join(sequence_regions, by = "seqid")

  list(gene_ranges = gene_ranges, sequence_regions = sequence_regions)
}

first_nonempty <- function(x, fallback = "") {
  x <- x[!is.na(x) & x != "" & x != "-"]
  if (length(x) == 0) fallback else x[1]
}

read_emapper <- function(path) {
  cols <- c(
    "query", "seed_ortholog", "evalue", "score", "eggNOG_OGs",
    "max_annot_lvl", "COG_category", "eggNOG_Description",
    "eggNOG_name", "GOs", "EC", "KEGG_ko", "KEGG_Pathway",
    "KEGG_Module", "KEGG_Reaction", "KEGG_rclass", "BRITE",
    "KEGG_TC", "CAZy", "BiGG_Reaction", "PFAMs"
  )
  tab <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", fill = TRUE)
  names(tab) <- cols[seq_len(ncol(tab))]
  tab |>
    dplyr::mutate(protein_id = stringr::str_extract(query, "XP_[0-9]+[.][0-9]+"))
}

parse_cds_gene_bridge <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  gff |>
    dplyr::filter(type == "CDS") |>
    dplyr::transmute(
      protein_id = stringr::str_match(attributes, "protein_id=([^;]+)")[, 2],
      gene_id = stringr::str_match(attributes, "gene=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(protein_id), !is.na(gene_id)) |>
    dplyr::distinct()
}

build_gene_annotation <- function(gff_path, eggnog_path) {
  gff_annotation <- parse_gff_gene_annotation(gff_path)

  eggnog_annotation <- read_emapper(eggnog_path) |>
    dplyr::left_join(parse_cds_gene_bridge(gff_path), by = "protein_id") |>
    dplyr::filter(!is.na(gene_id)) |>
    dplyr::group_by(gene_id) |>
    dplyr::summarise(
      eggNOG_Description = first_nonempty(eggNOG_Description),
      eggNOG_name = first_nonempty(eggNOG_name),
      GOs = first_nonempty(GOs),
      KEGG_ko = first_nonempty(KEGG_ko),
      KEGG_Pathway = first_nonempty(KEGG_Pathway),
      PFAMs = first_nonempty(PFAMs),
      .groups = "drop"
    )

  gff_annotation |>
    dplyr::full_join(eggnog_annotation, by = "gene_id") |>
    dplyr::mutate(
      Description = dplyr::coalesce(Description, "no GFF description"),
      gene_biotype = dplyr::coalesce(gene_biotype, "unknown"),
      annotation_text = stringr::str_squish(paste(
        Description, eggNOG_Description, eggNOG_name, PFAMs,
        KEGG_Pathway, sep = " | "
      ))
    )
}

classify_physiology_theme <- function(annotation_text) {
  text <- stringr::str_to_lower(dplyr::coalesce(annotation_text, ""))
  dplyr::case_when(
    stringr::str_detect(
      text,
      "antimicrobial|defensin|attacin|cecropin|diptericin|gloverin|metchnikowin|thaumatin|termicin|lysozyme"
    ) ~ "Antimicrobial peptides",
    stringr::str_detect(
      text,
      "toll|imd|relish|myd88|cactus|dorsal|nf-kappa|nf kappa|jak|stat"
    ) ~ "Immune-related",
    stringr::str_detect(
      text,
      "immune|peptidoglycan|pgrp|gram-negative|gram negative|beta-1,3-glucan|glucan binding|phenoloxidase|prophenoloxidase|melanization|lectin|hemocyanin|complement|serpin"
    ) ~ "Immune-related",
    stringr::str_detect(
      text,
      "glycogen|glycogenin|branching enzyme|starch"
    ) ~ "Glycogen synthesis / storage",
    stringr::str_detect(
      text,
      "lipid|fatty acid|fatty-acid|acyl-coa|acyl coa|acetyl-coa carboxylase|elongase|desaturase|glycerolipid|phospholipid|triacylglycerol|diacylglycerol|sterol|lipase"
    ) ~ "Lipid synthesis / metabolism",
    stringr::str_detect(
      text,
      "protein synthesis|translation|ribosom|elongation factor|initiation factor|aminoacyl|trna synthetase|vitellogenin|hexamerin|arylphorin|storage protein"
    ) ~ "Protein synthesis / storage",
    stringr::str_detect(
      text,
      "detox|cytochrome p450|\\bcyp[0-9a-z]|glutathione|gst|carboxylesterase|esterase|udp-glucuronosyltransferase|\\bugt\\b|abc transporter|oxidoreductase|peroxidase|catalase|superoxide|heat shock|stress"
    ) ~ "Detoxification / stress",
    stringr::str_detect(text, "no gff description|^\\s*$") ~
      "Other / unannotated genes",
    TRUE ~ "Other / unannotated genes"
  )
}

classify_gene_class <- function(gene_biotype) {
  dplyr::case_when(
    gene_biotype == "protein_coding" ~ "Coding genes",
    gene_biotype == "lncRNA" ~ "Long non-coding RNA",
    TRUE ~ "Other / unknown"
  )
}

make_condition_label <- function(comparison_type, numerator, denominator, diet,
                                 side = c("numerator", "denominator")) {
  side <- match.arg(side)
  value <- if (side == "numerator") numerator else denominator

  dplyr::case_when(
    comparison_type == "Infection within diet" ~ paste(value, "diet", diet),
    grepl("control", comparison_type, ignore.case = TRUE) ~
      paste("Control diet", value),
    grepl("infected", comparison_type, ignore.case = TRUE) ~
      paste("Infected diet", value),
    TRUE ~ value
  )
}

make_condition_hull <- function(df) {
  if (nrow(df) < 3) return(df[0, ])
  df[chull(df$PC1, df$PC2), ]
}

copy_publication_asset <- function(source, file_base) {
  if (!file.exists(source)) return(NA_character_)
  target <- file.path(out_dir, paste0(file_base, ".", tools::file_ext(source)))
  file.copy(source, target, overwrite = TRUE)
  target
}
```

## Input data

```{r input-data}
scenario_dir <- latest_dir("^deg-sample-filter-scenarios_")
scenario_results_file <- file.path(scenario_dir, "all_deseq2_results.csv")
scenario_sample_counts_file <- file.path(scenario_dir,
                                         "sample_counts_by_scenario.csv")
model_dir <- latest_dir("^all-samples-deseq2-interaction-model_")
model_coefficients_file <- file.path(model_dir, "all_model_coefficients.csv")
logical_dir <- latest_dir("^deg-logical-set-construction_")
head_phase_file <- file.path(logical_dir, "step3_head_phase_degs.csv")
thorax_phase_file <- file.path(logical_dir, "step4_thorax_phase_degs.csv")
circled_cluster_dir <- latest_dir("^circled-cluster-enrichment_")
figure3_composite_source <- file.path(
  circled_cluster_dir,
  "expression_cluster_gene_level_GO_KEGG_lollipop.png"
)
figure3_reference_source <- file.path(
  circled_cluster_dir,
  "expression_cluster_gene_level_GO_KEGG_lollipop_with_references.png"
)

inputs <- data.frame(
  input = c("metadata", "rRNA-filtered counts", "GFF annotation",
            "eggNOG annotation", "all-samples DESeq2 scenario table",
            "sample-filter scenario counts",
            "all-samples model coefficients", "Head phase DEGs",
            "Thorax phase DEGs", "Figure 3 composite source"),
  path = c(metadata_file, counts_file, gff_file, eggnog_file,
           scenario_results_file, scenario_sample_counts_file,
           model_coefficients_file,
           head_phase_file, thorax_phase_file, figure3_composite_source),
  present = c(file.exists(metadata_file), file.exists(counts_file),
              file.exists(gff_file), file.exists(eggnog_file),
              file.exists(scenario_results_file),
              file.exists(scenario_sample_counts_file),
              file.exists(model_coefficients_file),
              file.exists(head_phase_file), file.exists(thorax_phase_file),
              file.exists(figure3_composite_source))
)

knitr::kable(inputs, caption = "Inputs used to assemble publication figures")
if (!all(inputs$present)) {
  stop("One or more publication figure inputs are missing.")
}
```

## Figure 1: design, fat body PCA, and infection DEGs

Figure 1 is a compact visual summary of the fat body experiment and the
infection-response contrasts.

- Panel A shows the balanced diet-by-infection design and the number of fat body
  RNA-seq libraries in each group.
- Panel B shows the VST PCA from all expressed, rRNA-filtered genes. Pale hulls
  show the spread of each diet-by-treatment group, black crosses mark
  centroids, and arrows show the shift from control to infected samples within
  each diet.
- Panel C summarizes the full contrast-level DEG burden across infection
  within diets, diet differences among controls, diet differences among
  infected samples, and the global infected-control model term. Red bars mark
  genes higher in the first condition named in the contrast. Blue bars mark
  genes higher in the second condition. Lighter red or blue segments mark long
  non-coding RNAs.

### Panel A

```{r figure-1-panel-a, fig.width=7.2, fig.height=4.6}
metadata <- read.delim(metadata_file, check.names = FALSE)
metadata <- metadata[, !is.na(names(metadata)) & nzchar(names(metadata)),
                     drop = FALSE]
metadata$label <- as.character(metadata$label)
metadata$Diet <- factor(metadata$Diet, levels = c("33", "50", "83"))
metadata$Treatment <- factor(metadata$Treatment,
                             levels = c("Control", "Infected"))

design_counts <- metadata |>
  dplyr::count(Diet, Treatment, name = "n_samples") |>
  tidyr::complete(Diet, Treatment, fill = list(n_samples = 0)) |>
  dplyr::mutate(
    Condition = paste(Diet, Treatment),
    label = paste0("Diet ", Diet, "\n", Treatment, "\nn = ", n_samples),
    text_color = dplyr::if_else(Treatment == "Infected", "white", "#111827")
  )

p_design <- ggplot2::ggplot(
  design_counts,
  ggplot2::aes(x = Treatment, y = Diet, fill = Condition)
) +
  ggplot2::geom_tile(width = 0.84, height = 0.72, color = "#111827",
                     linewidth = 0.45) +
  ggplot2::geom_text(
    ggplot2::aes(label = label, color = text_color),
    lineheight = 0.88,
    size = 3.6,
    fontface = "bold"
  ) +
  ggplot2::annotate(
    "label",
    x = 1.5, y = 4.18,
    label = paste(
      "Female S. gregaria fat body RNA-seq",
      "96 h after control or Metarhizium challenge",
      "all samples retained; rRNA loci removed bioinformatically",
      sep = "\n"
    ),
    label.size = 0.25,
    fill = "white",
    color = "#111827",
    size = 3.5,
    fontface = "bold",
    lineheight = 0.96
  ) +
  ggplot2::scale_fill_manual(values = condition_colors, drop = FALSE) +
  ggplot2::scale_color_identity() +
  ggplot2::scale_y_discrete(limits = rev(levels(metadata$Diet))) +
  ggplot2::coord_cartesian(clip = "off", ylim = c(0.55, 4.45)) +
  theme_pub(base_size = 12) +
  ggplot2::theme(
    axis.title = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_text(face = "bold", size = 12),
    axis.text.y = ggplot2::element_text(face = "bold", size = 12),
    axis.ticks = ggplot2::element_blank(),
    legend.position = "none",
    plot.margin = ggplot2::margin(8, 8, 8, 8)
  ) +
  ggplot2::labs(title = "A. Experimental design")

save_png_svg(p_design, "figure1_panel_A_experimental_design",
             width = 7.2, height = 4.6)
p_design
```

### Panel B

```{r figure-1-panel-b, fig.width=7.2, fig.height=5.2}
counts_raw <- read.csv(counts_file, check.names = FALSE)
counts <- as.matrix(counts_raw[, -1, drop = FALSE])
rownames(counts) <- counts_raw[[1]]
storage.mode(counts) <- "integer"

count_labels_all <- sub("_MERGE$", "", sub("^mehreen_", "", colnames(counts)))
keep_samples <- count_labels_all %in% metadata$label
counts <- counts[, keep_samples, drop = FALSE]
count_labels <- count_labels_all[keep_samples]
metadata_ordered <- metadata[match(count_labels, metadata$label), ]
stopifnot(identical(metadata_ordered$label, count_labels))
rownames(metadata_ordered) <- metadata_ordered$label
colnames(counts) <- count_labels

keep_vst <- rowSums(counts) >= params$min_total_count
dds_vst <- DESeq2::DESeqDataSetFromMatrix(
  countData = counts[keep_vst, , drop = FALSE],
  colData = metadata_ordered,
  design = ~ Diet + Treatment
)
vsd <- DESeq2::vst(dds_vst, blind = TRUE)
vst_mat <- SummarizedExperiment::assay(vsd)
nonzero_variance <- apply(vst_mat, 1, var) > 0
pca <- prcomp(t(vst_mat[nonzero_variance, , drop = FALSE]), scale. = FALSE)
percent_var <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)

pca_scores <- data.frame(
  label = rownames(pca$x),
  PC1 = pca$x[, "PC1"],
  PC2 = pca$x[, "PC2"],
  Diet = metadata_ordered$Diet,
  Treatment = metadata_ordered$Treatment,
  check.names = FALSE
)
pca_scores$Condition <- factor(
  paste(pca_scores$Diet, pca_scores$Treatment),
  levels = names(condition_colors)
)

pca_centroids <- aggregate(cbind(PC1, PC2) ~ Diet + Treatment + Condition,
                           pca_scores, mean)
pca_hulls <- do.call(rbind, lapply(split(pca_scores, pca_scores$Condition),
                                   make_condition_hull))
centroid_shift <- merge(
  pca_centroids[pca_centroids$Treatment == "Control",
                c("Diet", "PC1", "PC2")],
  pca_centroids[pca_centroids$Treatment == "Infected",
                c("Diet", "PC1", "PC2")],
  by = "Diet",
  suffixes = c("_control", "_infected")
)

p_pca <- ggplot2::ggplot() +
  ggplot2::geom_polygon(
    data = pca_hulls,
    ggplot2::aes(x = PC1, y = PC2, fill = Condition, group = Condition),
    alpha = 0.18,
    color = NA
  ) +
  ggplot2::geom_point(
    data = pca_scores,
    ggplot2::aes(x = PC1, y = PC2, fill = Condition, shape = Treatment),
    color = "#111827",
    size = 3.3,
    stroke = 0.35,
    alpha = 0.92
  ) +
  ggplot2::geom_segment(
    data = centroid_shift,
    ggplot2::aes(x = PC1_control, y = PC2_control,
                 xend = PC1_infected, yend = PC2_infected,
                 color = Diet),
    linewidth = 0.9,
    arrow = grid::arrow(length = grid::unit(0.18, "cm"))
  ) +
  ggplot2::geom_point(
    data = pca_centroids,
    ggplot2::aes(x = PC1, y = PC2),
    shape = 4,
    color = "#111827",
    stroke = 1.25,
    size = 5.2
  ) +
  ggplot2::scale_fill_manual(values = condition_colors, drop = FALSE) +
  ggplot2::scale_color_manual(values = diet_colors, drop = FALSE) +
  ggplot2::scale_shape_manual(values = c("Control" = 21, "Infected" = 24)) +
  theme_pub(base_size = 12) +
  ggplot2::theme(
    legend.position = "bottom",
    legend.box = "vertical",
    panel.grid.major = ggplot2::element_line(color = "#E5E7EB",
                                             linewidth = 0.25)
  ) +
  ggplot2::guides(
    fill = ggplot2::guide_legend(nrow = 2, title.position = "top"),
    color = ggplot2::guide_legend(nrow = 1, title.position = "top"),
    shape = ggplot2::guide_legend(nrow = 1, title.position = "top")
  ) +
  ggplot2::labs(
    title = "B. Fat body VST PCA with centroid shifts",
    x = paste0("PC1: ", percent_var[1], "% variance"),
    y = paste0("PC2: ", percent_var[2], "% variance"),
    fill = "Diet and treatment",
    color = "Centroid shift",
    shape = "Treatment"
  )

write.csv(pca_scores, file.path(out_dir, "figure1_panel_B_pca_scores.csv"),
          row.names = FALSE)
write.csv(pca_centroids, file.path(out_dir, "figure1_panel_B_pca_centroids.csv"),
          row.names = FALSE)
write.csv(centroid_shift, file.path(out_dir,
                                    "figure1_panel_B_pca_centroid_shifts.csv"),
          row.names = FALSE)
save_png_svg(p_pca, "figure1_panel_B_vst_pca_centroids",
             width = 7.2, height = 5.2)
p_pca
```

### Panel C

```{r figure-1-panel-c, fig.width=11.2, fig.height=8.4}
gene_biotypes <- parse_gff_gene_annotation(gff_file)

scenario_degs_for_barplot <- read.csv(scenario_results_file, check.names = FALSE) |>
  dplyr::filter(
    scenario == "All samples",
    significant,
    !is.na(gene_id),
    !is.na(padj),
    padj < params$alpha,
    abs(log2FoldChange) >= params$lfc_threshold
  )

model_degs_for_barplot <- read.csv(model_coefficients_file,
                                   check.names = FALSE) |>
  dplyr::filter(
    coefficient == "Treatment_Infected_vs_Control",
    significant,
    !is.na(gene_id),
    !is.na(padj),
    padj < params$alpha,
    abs(log2FoldChange) >= params$lfc_threshold
  )

deg_raw <- dplyr::bind_rows(
  scenario_degs_for_barplot |>
    dplyr::transmute(
      DEG_context = dplyr::case_when(
        comparison_type == "Infection within diet" ~
          "Infection within diets",
        comparison_type == "Diet contrasts among control" ~
          "Diet DEGs in controls",
        comparison_type == "Diet contrasts among infected" ~
          "Diet DEGs in infected",
        TRUE ~ comparison_type
      ),
      comparison_type,
      comparison,
      numerator,
      denominator,
      diet = as.character(diet),
      gene_id,
      log2FoldChange
    ),
  model_degs_for_barplot |>
    dplyr::transmute(
      DEG_context = "Infection vs control",
      comparison_type = "Infection vs control",
      comparison = "All diets: infected vs control",
      numerator = "Infected",
      denominator = "Control",
      diet = NA_character_,
      gene_id,
      log2FoldChange
    )
)

context_order <- c(
  "Infection within diets",
  "Diet DEGs in controls",
  "Diet DEGs in infected",
  "Infection vs control"
)

contrast_order <- c(
  "All diets\ninfected vs control",
  "Diet 83\ninfected vs control",
  "Diet 50\ninfected vs control",
  "Diet 33\ninfected vs control",
  "Control\n83 vs 50",
  "Control\n83 vs 33",
  "Control\n33 vs 50",
  "Infected\n83 vs 50",
  "Infected\n83 vs 33",
  "Infected\n33 vs 50"
)

deg_biotype_counts <- deg_raw |>
  dplyr::mutate(
    signed_direction = dplyr::if_else(log2FoldChange > 0,
                                      "Upregulated", "Downregulated"),
    numerator_condition = make_condition_label(
      comparison_type, numerator, denominator, diet, "numerator"
    ),
    denominator_condition = make_condition_label(
      comparison_type, numerator, denominator, diet, "denominator"
    ),
    higher_in = dplyr::if_else(log2FoldChange > 0,
                               numerator_condition, denominator_condition),
    contrast_label = dplyr::case_when(
      DEG_context == "Infection within diets" ~
        paste0("Diet ", diet, "\ninfected vs control"),
      DEG_context == "Diet DEGs in controls" ~
        paste0("Control\n", numerator, " vs ", denominator),
      DEG_context == "Diet DEGs in infected" ~
        paste0("Infected\n", numerator, " vs ", denominator),
      DEG_context == "Infection vs control" ~
        "All diets\ninfected vs control",
      TRUE ~ comparison
    )
  ) |>
  dplyr::left_join(gene_biotypes, by = "gene_id") |>
  dplyr::mutate(
    Gene_class = classify_gene_class(gene_biotype),
    DEG_context = factor(DEG_context, levels = context_order),
    contrast_label = factor(contrast_label, levels = contrast_order),
    signed_direction = factor(signed_direction,
                              levels = c("Downregulated", "Upregulated")),
    direction_class = paste(signed_direction, Gene_class, sep = " - ")
  ) |>
  dplyr::distinct(DEG_context, comparison, contrast_label, signed_direction,
                  gene_id, Gene_class, .keep_all = TRUE) |>
  dplyr::count(DEG_context, comparison, contrast_label,
               signed_direction, Gene_class,
               direction_class, name = "n_DEGs") |>
  dplyr::group_by(DEG_context, comparison, contrast_label, signed_direction) |>
  tidyr::complete(
    Gene_class = c("Coding genes", "Long non-coding RNA", "Other / unknown"),
    fill = list(n_DEGs = 0)
  ) |>
  dplyr::ungroup() |>
  dplyr::filter(!is.na(DEG_context), !is.na(contrast_label)) |>
  dplyr::mutate(
    direction_class = paste(signed_direction, Gene_class, sep = " - "),
    signed_n = dplyr::if_else(signed_direction == "Upregulated",
                              n_DEGs, -n_DEGs)
  )

deg_total_labels <- deg_biotype_counts |>
  dplyr::group_by(DEG_context, comparison, contrast_label, signed_direction) |>
  dplyr::summarise(total_DEGs = sum(n_DEGs), .groups = "drop") |>
  dplyr::mutate(
    signed_total = dplyr::if_else(signed_direction == "Upregulated",
                                  total_DEGs, -total_DEGs)
  )

context_max <- deg_total_labels |>
  dplyr::group_by(DEG_context) |>
  dplyr::summarise(max_total = max(abs(signed_total), 1),
                   .groups = "drop")

deg_total_labels <- deg_total_labels |>
  dplyr::left_join(context_max, by = "DEG_context") |>
  dplyr::filter(total_DEGs > 0) |>
  dplyr::mutate(
    label_x = signed_total + dplyr::if_else(
      signed_total > 0,
      max_total * 0.025,
      -max_total * 0.025
    ),
    hjust = dplyr::if_else(signed_total > 0, 0, 1)
  )

p_deg_biotype <- ggplot2::ggplot(
  deg_biotype_counts,
  ggplot2::aes(x = signed_n, y = contrast_label, fill = direction_class)
) +
  ggplot2::geom_vline(xintercept = 0, linewidth = 0.35, color = "#111827") +
  ggplot2::geom_col(width = 0.74, color = "white", linewidth = 0.2) +
  ggplot2::geom_text(
    data = deg_total_labels,
    ggplot2::aes(x = label_x, y = contrast_label, label = total_DEGs,
                 hjust = hjust),
    inherit.aes = FALSE,
    size = 3.1,
    fontface = "bold"
  ) +
  ggplot2::facet_wrap(~ DEG_context, ncol = 2, scales = "free") +
  ggplot2::scale_fill_manual(values = direction_biotype_colors, drop = FALSE) +
  ggplot2::scale_x_continuous(
    labels = abs,
    expand = ggplot2::expansion(mult = c(0.22, 0.22))
  ) +
  theme_pub(base_size = 12) +
  ggplot2::theme(
    axis.title.y = ggplot2::element_blank(),
    axis.text.y = ggplot2::element_text(size = 10.5, face = "bold"),
    strip.text = ggplot2::element_text(size = 10.5, face = "bold"),
    panel.spacing = grid::unit(0.55, "lines"),
    legend.position = "bottom",
    legend.text = ggplot2::element_text(size = 8),
    plot.margin = ggplot2::margin(6, 28, 6, 6)
  ) +
  ggplot2::guides(
    fill = ggplot2::guide_legend(nrow = 3, title.position = "top",
                                 byrow = TRUE)
  ) +
  ggplot2::labs(
    title = "C. Full DEG burden across experimental contrasts",
    subtitle = "Red: higher in first condition named; blue: higher in second condition",
    x = "Number of significant DEGs",
    fill = "Direction and gene class"
  )

write.csv(deg_biotype_counts,
          file.path(out_dir, "figure1_panel_C_deg_biotype_counts.csv"),
          row.names = FALSE)
write.csv(deg_raw,
          file.path(out_dir, "figure1_panel_C_exact_contrast_deg_gene_table.csv"),
          row.names = FALSE)
save_png_svg(p_deg_biotype, "figure1_panel_C_deg_biotype_burden",
             width = 11.2, height = 8.4)
p_deg_biotype
```

### Composite Figure 1

```{r figure-1-composite, fig.width=15.5, fig.height=10.5}
figure1 <- p_design / (p_pca | p_deg_biotype) +
  patchwork::plot_layout(heights = c(0.44, 0.56),
                         widths = c(1.0, 1.08)) +
  patchwork::plot_annotation(
    title = "Figure 1. Fat body design, transcriptome structure, and DEG burden",
    subtitle = "All samples retained; rRNA loci removed before expression analysis",
    theme = ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold", size = 18),
      plot.subtitle = ggplot2::element_text(size = 12)
    )
  )

save_png_svg(figure1, "figure1_design_fatbody_pca_infection_deg",
             width = 17.5, height = 12.6)
figure1
```

## Figure 2: DEG-family overlap and candidate composition

Figure 2 asks whether the major DEG families are distinct or reusing the same
genes. Panel A is a Venn diagram across four broad DEG contexts: infection
within diet, diet differences among controls, diet differences among infected
samples, and the global infected-control main effect. Panel B asks whether the
infection-within-diet DEGs overlap phase-change DEGs detected in independent
head and thorax solitarious-vs-gregarious references. Panel C asks what
fraction of each broad DEG family falls into hand-curated physiology categories
or the phase-related reference category.

```{r figure-2-overlap-composition, fig.width=14, fig.height=11, results='asis'}
scenario_degs <- read.csv(scenario_results_file, check.names = FALSE) |>
  dplyr::filter(
    scenario == "All samples",
    significant,
    !is.na(gene_id),
    !is.na(padj),
    padj < params$alpha,
    abs(log2FoldChange) >= params$lfc_threshold
  )

model_degs <- read.csv(model_coefficients_file, check.names = FALSE) |>
  dplyr::filter(
    coefficient == "Treatment_Infected_vs_Control",
    significant,
    !is.na(gene_id),
    !is.na(padj),
    padj < params$alpha,
    abs(log2FoldChange) >= params$lfc_threshold
  )

broad_deg_sets <- list(
  "Infection within diets" = scenario_degs |>
    dplyr::filter(comparison_type == "Infection within diet") |>
    dplyr::pull(gene_id) |>
    unique(),
  "Diet DEGs in controls" = scenario_degs |>
    dplyr::filter(comparison_type == "Diet contrasts among control") |>
    dplyr::pull(gene_id) |>
    unique(),
  "Diet DEGs in infected" = scenario_degs |>
    dplyr::filter(comparison_type == "Diet contrasts among infected") |>
    dplyr::pull(gene_id) |>
    unique(),
  "Infection vs control" = unique(model_degs$gene_id)
)
broad_deg_sets <- broad_deg_sets[lengths(broad_deg_sets) > 0]

all_overlap_genes <- sort(unique(unlist(broad_deg_sets, use.names = FALSE)))
overlap_membership <- data.frame(gene_id = all_overlap_genes)
for (set_name in names(broad_deg_sets)) {
  overlap_membership[[set_name]] <- overlap_membership$gene_id %in%
    broad_deg_sets[[set_name]]
}
overlap_membership$intersection <- apply(
  overlap_membership[names(broad_deg_sets)],
  1,
  function(x) paste(names(broad_deg_sets)[x], collapse = " + ")
)
write.csv(overlap_membership,
          file.path(out_dir, "figure2_deg_family_overlap_membership.csv"),
          row.names = FALSE)

overlap_summary <- overlap_membership |>
  dplyr::count(intersection, name = "n_genes") |>
  dplyr::arrange(dplyr::desc(n_genes), intersection)
write.csv(overlap_summary,
          file.path(out_dir, "figure2_deg_family_overlap_summary.csv"),
          row.names = FALSE)

head_phase_degs <- read.csv(head_phase_file, check.names = FALSE) |>
  dplyr::filter(
    significant,
    !is.na(gene_id),
    !is.na(padj),
    padj < params$alpha,
    abs(log2FoldChange) >= params$lfc_threshold
  )
thorax_phase_degs <- read.csv(thorax_phase_file, check.names = FALSE) |>
  dplyr::filter(
    significant,
    !is.na(gene_id),
    !is.na(padj),
    padj < params$alpha,
    abs(log2FoldChange) >= params$lfc_threshold
  )
head_phase_genes <- unique(head_phase_degs$gene_id)
thorax_phase_genes <- unique(thorax_phase_degs$gene_id)
phase_genes <- intersect(head_phase_genes, thorax_phase_genes)

phase_overlap_sets <- list(
  "Infection within diets" = broad_deg_sets[["Infection within diets"]],
  "Head phase DEGs" = head_phase_genes,
  "Thorax phase DEGs" = thorax_phase_genes
)

phase_overlap_genes <- sort(unique(unlist(phase_overlap_sets,
                                          use.names = FALSE)))
phase_overlap_membership <- data.frame(gene_id = phase_overlap_genes)
for (set_name in names(phase_overlap_sets)) {
  phase_overlap_membership[[set_name]] <-
    phase_overlap_membership$gene_id %in% phase_overlap_sets[[set_name]]
}
phase_overlap_membership$intersection <- apply(
  phase_overlap_membership[names(phase_overlap_sets)],
  1,
  function(x) paste(names(phase_overlap_sets)[x], collapse = " + ")
)
write.csv(phase_overlap_membership,
          file.path(out_dir, "figure2_infection_phase_overlap_membership.csv"),
          row.names = FALSE)

phase_overlap_summary <- phase_overlap_membership |>
  dplyr::count(intersection, name = "n_genes") |>
  dplyr::arrange(dplyr::desc(n_genes), intersection)
write.csv(phase_overlap_summary,
          file.path(out_dir, "figure2_infection_phase_overlap_summary.csv"),
          row.names = FALSE)

broad_deg_plot_sets <- broad_deg_sets
names(broad_deg_plot_sets) <- c(
  "Infection\nwithin diet",
  "Diet\ncontrols",
  "Diet\ninfected",
  "Infect.\nvs ctrl"
)

p_venn <- ggVennDiagram::ggVennDiagram(
  broad_deg_plot_sets,
  label = "count",
  label_alpha = 0,
  edge_size = 0.75,
  set_size = 4.0
) +
  ggplot2::scale_fill_gradient(low = "#FFF7D1", high = "#D9A21B") +
  ggplot2::theme_void(base_size = 12) +
  ggplot2::theme(
    legend.position = "none",
    plot.title = ggplot2::element_text(face = "bold", size = 16),
    plot.subtitle = ggplot2::element_text(size = 11),
    plot.background = ggplot2::element_rect(fill = "white", color = NA),
    panel.background = ggplot2::element_rect(fill = "white", color = NA),
    plot.margin = ggplot2::margin(8, 24, 8, 24)
  ) +
  ggplot2::labs(
    title = "A. Venn overlap among DEG contexts",
    subtitle = "Each circle is the union of significant genes in that context",
    fill = "Genes"
  )

save_png_svg(p_venn, "figure2_deg_family_overlap_venn",
             width = 9.4, height = 8.2)

phase_overlap_plot_sets <- phase_overlap_sets
names(phase_overlap_plot_sets) <- c(
  "Infection\nwithin diets",
  "Head phase\nDEGs",
  "Thorax phase\nDEGs"
)

p_phase_venn <- ggVennDiagram::ggVennDiagram(
  phase_overlap_plot_sets,
  label = "count",
  label_alpha = 0,
  edge_size = 0.75,
  set_size = 4.0
) +
  ggplot2::scale_fill_gradient(low = "#FFF7D1", high = "#D9A21B") +
  ggplot2::theme_void(base_size = 12) +
  ggplot2::theme(
    legend.position = "none",
    plot.title = ggplot2::element_text(face = "bold", size = 16),
    plot.subtitle = ggplot2::element_text(size = 11),
    plot.background = ggplot2::element_rect(fill = "white", color = NA),
    panel.background = ggplot2::element_rect(fill = "white", color = NA),
    plot.margin = ggplot2::margin(8, 24, 8, 24)
  ) +
  ggplot2::labs(
    title = "B. Infection DEGs overlapping phase-reference DEGs",
    subtitle = "Phase references are solitarious-vs-gregarious DEGs in head and thorax",
    fill = "Genes"
  )

save_png_svg(p_phase_venn, "figure2_infection_phase_overlap_venn",
             width = 9.4, height = 8.2)

gene_annotation <- build_gene_annotation(gff_file, eggnog_file)
deg_family_long <- dplyr::bind_rows(lapply(names(broad_deg_sets), function(set_name) {
  data.frame(
    DEG_family = set_name,
    gene_id = broad_deg_sets[[set_name]],
    stringsAsFactors = FALSE
  )
}))
deg_family_long$DEG_family <- factor(
  deg_family_long$DEG_family,
  levels = names(broad_deg_sets)
)

candidate_theme_table <- deg_family_long |>
  dplyr::left_join(gene_annotation, by = "gene_id") |>
  dplyr::mutate(
    annotation_text = dplyr::coalesce(annotation_text, ""),
    base_physiology_theme = classify_physiology_theme(annotation_text),
    phase_related = gene_id %in% phase_genes,
    phase_reference_source = dplyr::case_when(
      gene_id %in% head_phase_genes & gene_id %in% thorax_phase_genes ~
        "Head + thorax phase DEG",
      gene_id %in% head_phase_genes ~ "Head phase DEG",
      gene_id %in% thorax_phase_genes ~ "Thorax phase DEG",
      TRUE ~ "Not phase-related"
    ),
    physiology_theme = dplyr::case_when(
      base_physiology_theme != "Other / unannotated genes" ~
        base_physiology_theme,
      phase_related ~ "Phase-related changes",
      TRUE ~ base_physiology_theme
    ),
    physiology_theme = factor(physiology_theme,
                              levels = names(physiology_theme_colors))
  )

write.csv(candidate_theme_table,
          file.path(out_dir, "figure2_deg_family_candidate_theme_gene_table.csv"),
          row.names = FALSE)

composition_counts <- candidate_theme_table |>
  dplyr::distinct(DEG_family, gene_id, physiology_theme) |>
  dplyr::count(DEG_family, physiology_theme, name = "n_genes") |>
  dplyr::group_by(DEG_family) |>
  tidyr::complete(
    physiology_theme = factor(names(physiology_theme_colors),
                              levels = names(physiology_theme_colors)),
    fill = list(n_genes = 0)
  ) |>
  dplyr::mutate(
    total_genes = sum(n_genes),
    proportion = dplyr::if_else(total_genes > 0, n_genes / total_genes, 0)
  ) |>
  dplyr::ungroup()

write.csv(composition_counts,
          file.path(out_dir, "figure2_deg_family_candidate_theme_counts.csv"),
          row.names = FALSE)

candidate_theme_levels <- names(physiology_theme_colors)
candidate_composition_counts <- composition_counts |>
  dplyr::filter(physiology_theme %in% candidate_theme_levels) |>
  dplyr::group_by(DEG_family) |>
  dplyr::mutate(
    candidate_genes = sum(n_genes),
    candidate_proportion = dplyr::if_else(candidate_genes > 0,
                                          n_genes / candidate_genes, 0)
  ) |>
  dplyr::ungroup()

write.csv(candidate_composition_counts,
          file.path(out_dir,
                    "figure2_deg_family_candidate_theme_counts_all_categories.csv"),
          row.names = FALSE)

family_label_levels <- candidate_composition_counts |>
  dplyr::distinct(DEG_family, total_genes, candidate_genes) |>
  dplyr::arrange(DEG_family) |>
  dplyr::mutate(
    family_label = paste0(
      DEG_family,
      "\nall DEG n = ", total_genes
    )
  ) |>
  dplyr::pull(family_label)

pie_counts <- candidate_composition_counts |>
  dplyr::filter(n_genes > 0) |>
  dplyr::mutate(
    family_label = paste0(
      DEG_family,
      "\nall DEG n = ", total_genes
    ),
    family_label = factor(family_label, levels = family_label_levels)
  )

p_composition <- ggplot2::ggplot(
  pie_counts,
  ggplot2::aes(x = "", y = candidate_proportion,
               fill = physiology_theme)
) +
  ggplot2::geom_col(width = 1, color = "white", linewidth = 0.4) +
  ggplot2::coord_polar(theta = "y") +
  ggplot2::facet_wrap(~ family_label, nrow = 1) +
  ggplot2::scale_fill_manual(
    values = physiology_theme_colors[candidate_theme_levels],
    breaks = candidate_theme_levels,
    drop = FALSE
  ) +
  ggplot2::theme_void(base_size = 13) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold", size = 16),
    plot.subtitle = ggplot2::element_text(size = 11),
    strip.background = ggplot2::element_rect(fill = "#F3F4F6",
                                             color = "#111827"),
    strip.text = ggplot2::element_text(size = 10.5, face = "bold",
                                       lineheight = 0.95),
    legend.position = "bottom",
    legend.text = ggplot2::element_text(size = 9),
    legend.title = ggplot2::element_text(face = "bold"),
    plot.margin = ggplot2::margin(4, 8, 8, 8)
  ) +
  ggplot2::guides(
    fill = ggplot2::guide_legend(nrow = 3, byrow = TRUE,
                                 title.position = "top")
  ) +
  ggplot2::labs(
    title = "C. DEG physiology, phase-reference, and annotation composition",
    subtitle = "Immune and physiology annotations are prioritized; phase-related marks remaining DEGs common to head+thorax phase references",
    fill = "Gene category"
  )

save_png_svg(p_composition,
             "figure2_panel_C_candidate_physiology_phase_composition",
             width = 12, height = 5.6)

figure2 <- (p_venn | p_phase_venn) / p_composition +
  patchwork::plot_layout(heights = c(0.56, 0.44)) +
  patchwork::plot_annotation(
    title = "Figure 2. Overlap and physiology composition of major DEG families",
    subtitle = "All samples retained; broad DEG overlap and phase-reference overlap are followed by composition pies",
    theme = ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold", size = 18),
      plot.subtitle = ggplot2::element_text(size = 12)
    )
  )

save_png_svg(figure2, "figure2_deg_overlap_candidate_composition",
             width = 17.5, height = 11.4)
figure2

figure2_exports <- data.frame(
  figure = c("Figure 2 composite PNG", "Figure 2 composite SVG",
             "Figure 2 physiology composition PNG",
             "Figure 2 DEG context Venn PNG",
             "Figure 2 infection-phase Venn PNG",
             "Figure 2 gene membership table",
             "Figure 2 infection-phase membership table",
             "Figure 2 candidate category gene table"),
  path = c(
    file.path(out_dir, "figure2_deg_overlap_candidate_composition.png"),
    file.path(out_dir, "figure2_deg_overlap_candidate_composition.svg"),
    file.path(out_dir,
              "figure2_panel_C_candidate_physiology_phase_composition.png"),
    file.path(out_dir, "figure2_deg_family_overlap_venn.png"),
    file.path(out_dir, "figure2_infection_phase_overlap_venn.png"),
    file.path(out_dir, "figure2_deg_family_overlap_membership.csv"),
    file.path(out_dir, "figure2_infection_phase_overlap_membership.csv"),
    file.path(out_dir, "figure2_deg_family_candidate_theme_gene_table.csv")
  ),
  present = file.exists(c(
    file.path(out_dir, "figure2_deg_overlap_candidate_composition.png"),
    file.path(out_dir, "figure2_deg_overlap_candidate_composition.svg"),
    file.path(out_dir,
              "figure2_panel_C_candidate_physiology_phase_composition.png"),
    file.path(out_dir, "figure2_deg_family_overlap_venn.png"),
    file.path(out_dir, "figure2_infection_phase_overlap_venn.png"),
    file.path(out_dir, "figure2_deg_family_overlap_membership.csv"),
    file.path(out_dir, "figure2_infection_phase_overlap_membership.csv"),
    file.path(out_dir, "figure2_deg_family_candidate_theme_gene_table.csv")
  ))
)
knitr::kable(figure2_exports, caption = "Figure 2 overlap and composition exports")
```

## Figure 3: expression clusters and enrichment

Figure 3 keeps the expression-cluster composite from the circled-cluster page.
The fat body DEGs are shown as the gene-level heatmap, with GO and KEGG
enrichment summaries underneath.

```{r figure-3-composite, results='asis'}
figure3_png <- copy_publication_asset(
  figure3_composite_source,
  "figure3_expression_cluster_gene_level_GO_KEGG_lollipop"
)
figure3_reference_png <- copy_publication_asset(
  figure3_reference_source,
  "figure3_expression_cluster_gene_level_GO_KEGG_lollipop_with_references"
)

figure3_exports <- data.frame(
  figure = c("Figure 3 composite", "Figure 3 reference-context companion"),
  source = c(figure3_composite_source, figure3_reference_source),
  copied_to = c(figure3_png, figure3_reference_png),
  present = file.exists(c(figure3_png, figure3_reference_png))
)
knitr::kable(figure3_exports, caption = "Figure 3 copied exports")

if (file.exists(figure3_png)) {
  rel_figure3 <- sub(paste0("^", project_dir, "/"), "../", figure3_png)
  cat("![](", rel_figure3, ")\n\n", sep = "")
}
```

## Figure S1: outlier-removal sensitivity

Figure S1 compares DEG calls across sample-filtering scenarios. It is a
diagnostic companion to the main all-samples analysis: panel A shows how many
samples remain in each diet-by-infection group, panel B shows how the number of
DEGs changes across contrasts, and panel C shows whether genes are retained,
lost, or gained after removing the current outlier set (`1007`, `1036`, `1037`,
and `1039`).

```{r figure-s1-outlier-sensitivity, fig.width=14, fig.height=13, results='asis'}
s1_scenario_results <- read.csv(scenario_results_file, check.names = FALSE)
s1_sample_counts <- read.csv(scenario_sample_counts_file, check.names = FALSE)

s1_scenario_levels <- c("All samples", "Current outliers removed",
                        "Expanded outlier set")
s1_scenario_levels <- s1_scenario_levels[
  s1_scenario_levels %in% unique(s1_scenario_results$scenario)
]
s1_direction_levels <- c("Downregulated", "Upregulated")
s1_direction_colors <- c("Downregulated" = "#1D4ED8",
                         "Upregulated" = "#B91C1C")
s1_sensitivity_colors <- c(
  "Retained after outlier removal" = "#B7791F",
  "Lost after outlier removal" = "#B91C1C",
  "Gained after outlier removal" = "#2563EB"
)

format_comparison_label <- function(x) {
  dplyr::case_when(
    x == "Diet 33: infected vs control" ~ "Diet 33\ninfected vs control",
    x == "Diet 50: infected vs control" ~ "Diet 50\ninfected vs control",
    x == "Diet 83: infected vs control" ~ "Diet 83\ninfected vs control",
    x == "Control 33 vs 50" ~ "Control\n33 vs 50",
    x == "Control 83 vs 33" ~ "Control\n83 vs 33",
    x == "Control 83 vs 50" ~ "Control\n83 vs 50",
    x == "Infected 33 vs 50" ~ "Infected\n33 vs 50",
    x == "Infected 83 vs 33" ~ "Infected\n83 vs 33",
    x == "Infected 83 vs 50" ~ "Infected\n83 vs 50",
    TRUE ~ x
  )
}

s1_comparison_levels <- c(
  "Diet 33: infected vs control",
  "Diet 50: infected vs control",
  "Diet 83: infected vs control",
  "Control 33 vs 50",
  "Control 83 vs 33",
  "Control 83 vs 50",
  "Infected 33 vs 50",
  "Infected 83 vs 33",
  "Infected 83 vs 50"
)
s1_comparison_levels <- s1_comparison_levels[
  s1_comparison_levels %in% unique(s1_scenario_results$comparison)
]

s1_sample_counts <- s1_sample_counts |>
  dplyr::mutate(
    scenario = factor(scenario, levels = s1_scenario_levels),
    Diet = factor(Diet, levels = c("33", "50", "83")),
    Treatment = factor(Treatment, levels = c("Control", "Infected"))
  )

p_s1_samples <- ggplot2::ggplot(
  s1_sample_counts,
  ggplot2::aes(x = Diet, y = n_samples, fill = Treatment)
) +
  ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.75),
                    width = 0.66, color = "#111827", linewidth = 0.25) +
  ggplot2::geom_text(
    ggplot2::aes(label = n_samples),
    position = ggplot2::position_dodge(width = 0.75),
    vjust = -0.25,
    size = 3.6
  ) +
  ggplot2::facet_wrap(~ scenario, nrow = 1) +
  ggplot2::scale_fill_manual(values = treatment_colors, drop = FALSE) +
  ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0, 0.15))) +
  theme_pub(base_size = 12) +
  ggplot2::theme(
    strip.text = ggplot2::element_text(size = 11, face = "bold"),
    axis.text = ggplot2::element_text(color = "#111827"),
    legend.position = "bottom"
  ) +
  ggplot2::labs(
    title = "A. Samples retained per scenario",
    x = "Diet",
    y = "Number of samples",
    fill = "Treatment"
  )

s1_sig_results <- s1_scenario_results |>
  dplyr::filter(
    scenario %in% s1_scenario_levels,
    comparison %in% s1_comparison_levels,
    significant,
    !is.na(gene_id),
    !is.na(padj),
    padj < params$alpha,
    abs(log2FoldChange) >= params$lfc_threshold,
    direction %in% s1_direction_levels
  ) |>
  dplyr::mutate(
    scenario = factor(scenario, levels = s1_scenario_levels),
    comparison = factor(comparison, levels = s1_comparison_levels),
    comparison_label = factor(format_comparison_label(as.character(comparison)),
                              levels = rev(format_comparison_label(s1_comparison_levels))),
    direction = factor(direction, levels = s1_direction_levels)
  )

s1_deg_counts <- s1_sig_results |>
  dplyr::count(scenario, comparison_type, comparison_label, direction,
               name = "n_genes") |>
  tidyr::complete(
    scenario,
    tidyr::nesting(comparison_type, comparison_label),
    direction,
    fill = list(n_genes = 0)
  ) |>
  dplyr::mutate(
    signed_n_genes = dplyr::if_else(direction == "Downregulated",
                                    -n_genes, n_genes)
  )

write.csv(s1_sample_counts,
          file.path(out_dir, "figureS1_sample_counts_by_scenario.csv"),
          row.names = FALSE)
write.csv(s1_deg_counts,
          file.path(out_dir, "figureS1_deg_counts_by_scenario.csv"),
          row.names = FALSE)

p_s1_burden <- ggplot2::ggplot(
  s1_deg_counts,
  ggplot2::aes(x = signed_n_genes, y = comparison_label, fill = direction)
) +
  ggplot2::geom_vline(xintercept = 0, color = "#111827", linewidth = 0.3) +
  ggplot2::geom_col(width = 0.72, color = "white", linewidth = 0.2) +
  ggplot2::facet_wrap(~ scenario, nrow = 1) +
  ggplot2::scale_fill_manual(values = s1_direction_colors, drop = FALSE) +
  ggplot2::scale_x_continuous(labels = abs) +
  theme_pub(base_size = 12) +
  ggplot2::theme(
    axis.text.y = ggplot2::element_text(size = 10, color = "#111827"),
    strip.text = ggplot2::element_text(size = 10.5, face = "bold"),
    legend.position = "bottom"
  ) +
  ggplot2::labs(
    title = "B. Direction-aware DEG burden across filtering scenarios",
    x = "Number of significant DEGs",
    y = NULL,
    fill = "Direction"
  )

s1_compare_scenarios <- c("All samples", "Current outliers removed")
if (all(s1_compare_scenarios %in% unique(s1_sig_results$scenario))) {
  s1_retained_lost_gained <- do.call(rbind, lapply(s1_comparison_levels,
                                                   function(this_comparison) {
    all_genes <- unique(s1_sig_results$gene_id[
      s1_sig_results$scenario == "All samples" &
        s1_sig_results$comparison == this_comparison
    ])
    no_outlier_genes <- unique(s1_sig_results$gene_id[
      s1_sig_results$scenario == "Current outliers removed" &
        s1_sig_results$comparison == this_comparison
    ])
    data.frame(
      comparison = this_comparison,
      sensitivity_class = c("Retained after outlier removal",
                            "Lost after outlier removal",
                            "Gained after outlier removal"),
      n_genes = c(
        length(intersect(all_genes, no_outlier_genes)),
        length(setdiff(all_genes, no_outlier_genes)),
        length(setdiff(no_outlier_genes, all_genes))
      ),
      stringsAsFactors = FALSE
    )
  }))
} else {
  s1_retained_lost_gained <- data.frame(
    comparison = character(),
    sensitivity_class = character(),
    n_genes = integer()
  )
}

s1_retained_lost_gained <- s1_retained_lost_gained |>
  dplyr::mutate(
    comparison = factor(comparison, levels = s1_comparison_levels),
    comparison_label = factor(format_comparison_label(as.character(comparison)),
                              levels = rev(format_comparison_label(s1_comparison_levels))),
    sensitivity_class = factor(sensitivity_class,
                               levels = names(s1_sensitivity_colors))
  )

write.csv(s1_retained_lost_gained,
          file.path(out_dir, "figureS1_retained_lost_gained_degs.csv"),
          row.names = FALSE)

p_s1_retained <- ggplot2::ggplot(
  s1_retained_lost_gained,
  ggplot2::aes(x = n_genes, y = comparison_label, fill = sensitivity_class)
) +
  ggplot2::geom_col(width = 0.72, color = "white", linewidth = 0.2) +
  ggplot2::scale_fill_manual(values = s1_sensitivity_colors, drop = FALSE) +
  theme_pub(base_size = 12) +
  ggplot2::theme(
    axis.text.y = ggplot2::element_text(size = 10, color = "#111827"),
    legend.position = "bottom"
  ) +
  ggplot2::labs(
    title = "C. DEG calls retained, lost, or gained after current outlier removal",
    x = "Number of significant DEGs",
    y = NULL,
    fill = "Sensitivity class"
  )

figure_s1 <- p_s1_samples / p_s1_burden / p_s1_retained +
  patchwork::plot_layout(heights = c(0.22, 0.48, 0.30)) +
  patchwork::plot_annotation(
    title = "Figure S1. Outlier-removal sensitivity analysis",
    subtitle = "All comparisons use the rRNA-filtered count matrix and the same DEG thresholds as the main figures",
    theme = ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold", size = 18),
      plot.subtitle = ggplot2::element_text(size = 12)
    )
  )

save_png_svg(figure_s1, "figureS1_outlier_removal_sensitivity",
             width = 17.5, height = 15)
figure_s1

figure_s1_exports <- data.frame(
  figure = c("Figure S1 composite PNG", "Figure S1 composite SVG",
             "Figure S1 sample counts", "Figure S1 DEG counts",
             "Figure S1 retained/lost/gained counts"),
  path = c(
    file.path(out_dir, "figureS1_outlier_removal_sensitivity.png"),
    file.path(out_dir, "figureS1_outlier_removal_sensitivity.svg"),
    file.path(out_dir, "figureS1_sample_counts_by_scenario.csv"),
    file.path(out_dir, "figureS1_deg_counts_by_scenario.csv"),
    file.path(out_dir, "figureS1_retained_lost_gained_degs.csv")
  ),
  present = file.exists(c(
    file.path(out_dir, "figureS1_outlier_removal_sensitivity.png"),
    file.path(out_dir, "figureS1_outlier_removal_sensitivity.svg"),
    file.path(out_dir, "figureS1_sample_counts_by_scenario.csv"),
    file.path(out_dir, "figureS1_deg_counts_by_scenario.csv"),
    file.path(out_dir, "figureS1_retained_lost_gained_degs.csv")
  ))
)
knitr::kable(figure_s1_exports,
             caption = "Figure S1 outlier-sensitivity exports")
```

## Figure S2: DEG localization on chromosomes

Figure S2 maps significant DEGs from each broad context onto the chromosome
coordinates in the RefSeq GFF. Genes on unplaced scaffolds are retained in the
exported coordinate table, while the plotted chromosome tracks focus on
chromosomes 1-11 and X.

An ideal parallel validation would have quantified selected host loci and
*Metarhizium* load by qPCR from matched fat body tissue. That wet-lab validation
cannot be performed for this dataset because no remaining tissue is available,
so the origin of the unplaced-scaffold signal must be evaluated
bioinformatically with scaffold annotation, fungal similarity searches, and
competitive host-pathogen remapping.

```{r figure-s2-deg-genome-localization, fig.width=16, fig.height=10.5, results='asis'}
s2_scenario_degs <- read.csv(scenario_results_file, check.names = FALSE) |>
  dplyr::filter(
    scenario == "All samples",
    significant,
    !is.na(gene_id),
    !is.na(padj),
    padj < params$alpha,
    abs(log2FoldChange) >= params$lfc_threshold
  )

s2_model_degs <- read.csv(model_coefficients_file, check.names = FALSE) |>
  dplyr::filter(
    coefficient == "Treatment_Infected_vs_Control",
    significant,
    !is.na(gene_id),
    !is.na(padj),
    padj < params$alpha,
    abs(log2FoldChange) >= params$lfc_threshold
  )

s2_deg_sets <- list(
  "Infection within diets" = s2_scenario_degs |>
    dplyr::filter(comparison_type == "Infection within diet") |>
    dplyr::pull(gene_id) |>
    unique(),
  "Diet DEGs in controls" = s2_scenario_degs |>
    dplyr::filter(comparison_type == "Diet contrasts among control") |>
    dplyr::pull(gene_id) |>
    unique(),
  "Diet DEGs in infected" = s2_scenario_degs |>
    dplyr::filter(comparison_type == "Diet contrasts among infected") |>
    dplyr::pull(gene_id) |>
    unique(),
  "Infection vs control" = unique(s2_model_degs$gene_id)
)
s2_deg_sets <- s2_deg_sets[lengths(s2_deg_sets) > 0]
s2_context_levels <- names(s2_deg_sets)
s2_context_colors <- c(
  "Infection within diets" = "#C51B7D",
  "Diet DEGs in controls" = "#D9A21B",
  "Diet DEGs in infected" = "#1B9E77",
  "Infection vs control" = "#377EB8"
)

s2_deg_long <- dplyr::bind_rows(lapply(s2_context_levels, function(context) {
  data.frame(
    DEG_context = context,
    gene_id = s2_deg_sets[[context]],
    stringsAsFactors = FALSE
  )
})) |>
  dplyr::distinct(DEG_context, gene_id) |>
  dplyr::mutate(DEG_context = factor(DEG_context,
                                     levels = s2_context_levels))

gff_coordinates <- parse_gff_gene_coordinates(gff_file)
chromosome_order <- c(as.character(1:11), "X")

chromosome_regions <- gff_coordinates$sequence_regions |>
  dplyr::filter(chromosome %in% chromosome_order) |>
  dplyr::mutate(
    chromosome = factor(chromosome, levels = rev(chromosome_order)),
    length_mb = seq_end / 1e6
  )

annotated_gene_counts <- gff_coordinates$gene_ranges |>
  dplyr::filter(chromosome %in% chromosome_order) |>
  dplyr::count(chromosome, name = "annotated_genes")

s2_deg_coordinates <- s2_deg_long |>
  dplyr::left_join(gff_coordinates$gene_ranges, by = "gene_id") |>
  dplyr::mutate(
    chromosome = as.character(chromosome),
    chromosome_class = dplyr::coalesce(chromosome_class, "Missing from GFF"),
    position_mb = midpoint / 1e6
  )

write.csv(s2_deg_coordinates,
          file.path(out_dir, "figureS2_deg_coordinates_by_context.csv"),
          row.names = FALSE)

s2_unplaced_summary <- s2_deg_coordinates |>
  dplyr::filter(!(chromosome %in% chromosome_order)) |>
  dplyr::count(DEG_context, chromosome_class, name = "n_degs") |>
  dplyr::arrange(DEG_context, chromosome_class)

write.csv(s2_unplaced_summary,
          file.path(out_dir, "figureS2_unplaced_or_missing_deg_summary.csv"),
          row.names = FALSE)

s2_coordinate_class_counts <- s2_deg_coordinates |>
  dplyr::mutate(
    coordinate_class = dplyr::if_else(
      chromosome_class == "Named chromosome",
      "Named chromosomes 1-11/X",
      chromosome_class
    )
  ) |>
  dplyr::count(DEG_context, coordinate_class, name = "n_degs") |>
  dplyr::group_by(DEG_context) |>
  dplyr::mutate(proportion = n_degs / sum(n_degs)) |>
  dplyr::ungroup()

write.csv(s2_coordinate_class_counts,
          file.path(out_dir, "figureS2_coordinate_class_counts.csv"),
          row.names = FALSE)

s2_chromosome_degs <- s2_deg_coordinates |>
  dplyr::filter(chromosome %in% chromosome_order, !is.na(position_mb)) |>
  dplyr::mutate(
    chromosome = factor(chromosome, levels = rev(chromosome_order)),
    DEG_context = factor(DEG_context, levels = s2_context_levels)
  )

s2_chromosome_counts <- s2_chromosome_degs |>
  dplyr::count(DEG_context, chromosome, name = "deg_genes") |>
  tidyr::complete(
    DEG_context,
    chromosome = factor(chromosome_order, levels = rev(chromosome_order)),
    fill = list(deg_genes = 0)
  ) |>
  dplyr::mutate(chromosome = as.character(chromosome)) |>
  dplyr::left_join(annotated_gene_counts, by = "chromosome") |>
  dplyr::mutate(
    degs_per_1000_genes = 1000 * deg_genes / annotated_genes,
    chromosome = factor(chromosome, levels = rev(chromosome_order)),
    DEG_context = factor(DEG_context, levels = s2_context_levels)
  )

write.csv(s2_chromosome_counts,
          file.path(out_dir, "figureS2_chromosome_deg_density.csv"),
          row.names = FALSE)

p_s2_locations <- ggplot2::ggplot() +
  ggplot2::geom_segment(
    data = chromosome_regions,
    ggplot2::aes(x = 0, xend = length_mb,
                 y = chromosome, yend = chromosome),
    linewidth = 4.8,
    color = "#E5E7EB",
    lineend = "round"
  ) +
  ggplot2::geom_point(
    data = s2_chromosome_degs,
    ggplot2::aes(x = position_mb, y = chromosome, color = DEG_context),
    size = 0.55,
    alpha = 0.45,
    position = ggplot2::position_jitter(width = 0, height = 0.15, seed = 17)
  ) +
  ggplot2::facet_wrap(~ DEG_context, ncol = 2) +
  ggplot2::scale_color_manual(values = s2_context_colors, guide = "none") +
  ggplot2::scale_x_continuous(
    expand = ggplot2::expansion(mult = c(0.01, 0.04))
  ) +
  theme_pub(base_size = 12) +
  ggplot2::theme(
    panel.grid.major.y = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank(),
    strip.text = ggplot2::element_text(size = 12, face = "bold"),
    axis.text.y = ggplot2::element_text(size = 10, color = "#111827"),
    legend.position = "none"
  ) +
  ggplot2::labs(
    title = "A. Chromosomal positions of DEGs by context",
    subtitle = "Each point is one DEG-context membership; genes shared by contexts appear in each relevant panel",
    x = "Genomic position (Mb)",
    y = "Chromosome"
  )

p_s2_coordinate_class <- ggplot2::ggplot(
  s2_coordinate_class_counts,
  ggplot2::aes(x = n_degs, y = DEG_context, fill = coordinate_class)
) +
  ggplot2::geom_col(width = 0.68, color = "white", linewidth = 0.2) +
  ggplot2::scale_fill_manual(
    values = c("Named chromosomes 1-11/X" = "#4B5563",
               "Unplaced scaffold" = "#D9A21B",
               "Missing from GFF" = "#B91C1C"),
    drop = FALSE
  ) +
  theme_pub(base_size = 12) +
  ggplot2::theme(
    axis.text.y = ggplot2::element_text(size = 10, color = "#111827"),
    legend.position = "bottom"
  ) +
  ggplot2::labs(
    title = "C. DEG-context memberships with chromosome versus scaffold coordinates",
    x = "Number of DEG-context memberships",
    y = NULL,
    fill = "Coordinate class"
  )

figure_s2 <- p_s2_locations / p_s2_coordinate_class +
  patchwork::plot_layout(heights = c(0.78, 0.22)) +
  patchwork::plot_annotation(
    title = "Figure S2. Chromosomal localization of DEG contexts",
    subtitle = "Chromosome tracks use the RefSeq GFF; genes on unplaced scaffolds are summarized below and exported separately",
    theme = ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold", size = 18),
      plot.subtitle = ggplot2::element_text(size = 12)
    )
  )

save_png_svg(figure_s2, "figureS2_deg_chromosome_localization",
             width = 16, height = 10.5)
figure_s2

figure_s2_exports <- data.frame(
  figure = c("Figure S2 composite PNG", "Figure S2 composite SVG",
             "Figure S2 DEG coordinates", "Chromosome density table",
             "Figure S2 coordinate class counts",
             "Figure S2 unplaced/missing summary"),
  path = c(
    file.path(out_dir, "figureS2_deg_chromosome_localization.png"),
    file.path(out_dir, "figureS2_deg_chromosome_localization.svg"),
    file.path(out_dir, "figureS2_deg_coordinates_by_context.csv"),
    file.path(out_dir, "figureS2_chromosome_deg_density.csv"),
    file.path(out_dir, "figureS2_coordinate_class_counts.csv"),
    file.path(out_dir, "figureS2_unplaced_or_missing_deg_summary.csv")
  ),
  present = file.exists(c(
    file.path(out_dir, "figureS2_deg_chromosome_localization.png"),
    file.path(out_dir, "figureS2_deg_chromosome_localization.svg"),
    file.path(out_dir, "figureS2_deg_coordinates_by_context.csv"),
    file.path(out_dir, "figureS2_chromosome_deg_density.csv"),
    file.path(out_dir, "figureS2_coordinate_class_counts.csv"),
    file.path(out_dir, "figureS2_unplaced_or_missing_deg_summary.csv")
  ))
)
knitr::kable(figure_s2_exports,
             caption = "Figure S2 chromosome-localization exports")
knitr::kable(s2_unplaced_summary,
             caption = "DEG-context memberships on unplaced scaffolds or missing from the GFF")
```

## Exported files

```{r exported-files}
exports <- list.files(out_dir, full.names = FALSE)
knitr::kable(tibble::tibble(file = exports),
             caption = paste("Files written to", out_dir))
```

## Session information

```{r session-info}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
