---
title: "Quality sensitivity and functional enrichment"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  metadata_file: "data/metadata/mehreen_metadata_fixed.txt"
  metadata_no_outliers_file: "data/metadata/mehreen_metadata_no_outliers.txt"
  counts_file: "data/raw_read_counts/master_counts_no_rRNA.csv"
  eggnog_file: "data/GO_Annotations/GCF_023897955.1_iqSchGreg1.2_Arthopoda_one2one.emapper.annotations"
  gff_file: "data/reference/GCF_023897955.1_iqSchGreg1.2_genomic.gff"
  latest_results_root: "output/rmd_runs"
  alpha: 0.05
  lfc_threshold: 1
  top_terms: 15
  output_tag: "outliers-load-enrichment"
---

## Purpose

The biological goal is to identify genes that differ among diets, infection
status, and diet-specific infection responses. This page summarizes current
sample inclusion, lifts local eggNOG annotations onto the `LOC...` gene IDs used
by the count matrix, and runs GO and KEGG pathway enrichment on the current DEG
sets.

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c(
  "DESeq2", "dplyr", "tidyr", "stringr", "ggplot2", "forcats",
  "clusterProfiler", "GO.db", "AnnotationDbi", "enrichplot", "DT"
)
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

metadata_file <- file.path(project_dir, params$metadata_file)
metadata_no_outliers_file <- file.path(project_dir, params$metadata_no_outliers_file)
counts_file <- file.path(project_dir, params$counts_file)
eggnog_file <- file.path(project_dir, params$eggnog_file)
gff_file <- file.path(project_dir, params$gff_file)
```

## Current sample and outlier status

Outliers are handled explicitly because this experiment has a modest number of
replicates per diet-by-treatment group. A sample that separates strongly in PCA,
sample-distance heatmaps, or library-level QC can pull a group mean away from
the other biological replicates and inflate or hide DEG calls. The no-outlier
metadata file is therefore treated as a sensitivity dataset: it does not erase
the full dataset, but it lets the analysis ask whether the main conclusions
depend on the strongest QC outliers.

```{r outlier-status}
metadata <- read.delim(metadata_file, check.names = FALSE)
metadata_no_outliers <- read.delim(metadata_no_outliers_file, check.names = FALSE)
metadata <- metadata[, !is.na(names(metadata)) & nzchar(names(metadata)), drop = FALSE]
metadata_no_outliers <- metadata_no_outliers[, !is.na(names(metadata_no_outliers)) &
                                               nzchar(names(metadata_no_outliers)),
                                             drop = FALSE]
metadata$label <- as.character(metadata$label)
metadata_no_outliers$label <- as.character(metadata_no_outliers$label)

removed_samples <- metadata |>
  dplyr::filter(!label %in% metadata_no_outliers$label)

knitr::kable(removed_samples)

design_full <- as.data.frame.matrix(with(metadata, table(Diet, Treatment)))
design_no_outliers <- as.data.frame.matrix(with(metadata_no_outliers,
                                                table(Diet, Treatment)))

knitr::kable(design_full, caption = "Full metadata design")
knitr::kable(design_no_outliers, caption = "No-outlier metadata design")

write.csv(removed_samples, file.path(out_dir, "removed_outlier_samples.csv"),
          row.names = FALSE)
```

The no-outlier file removes samples `1036` and `1039`, both control samples from
diet 33, and samples `1007` and `1037`, both control samples from diet 50. These
are reported here because removing controls from two diet groups changes group
balance and should always be visible when interpreting diet-specific infection
results. The removal is used to reduce sensitivity to samples with atypical
global expression structure, not because infection age was uncontrolled; age was
controlled in the lab design.

## Current DEG set summary

Before enrichment, the page summarizes the most recent DEG outputs from the
model pages. This check keeps the enrichment step connected to the exact DEG
sets currently available on disk, rather than relying on a manually copied gene
list.

```{r current-deg-summary}
latest_file <- function(pattern) {
  files <- list.files(file.path(project_dir, params$latest_results_root),
                      pattern = pattern, recursive = TRUE, full.names = TRUE)
  if (length(files) == 0) return(NA_character_)
  files[order(file.info(files)$mtime, decreasing = TRUE)][1]
}

summary_files <- c(
  interaction = latest_file("significant_gene_counts[.]csv$"),
  diet_pairwise = latest_file("diet_pairwise_summary[.]csv$"),
  infection_within_diet = latest_file("infection_within_diet_summary[.]csv$")
)

for (nm in names(summary_files)) {
  if (!is.na(summary_files[[nm]])) {
    cat("\n\n###", gsub("_", " ", nm), "\n\n")
    print(knitr::kable(read.csv(summary_files[[nm]], check.names = FALSE)))
  }
}
```

## Build annotation maps from local eggNOG and GFF

The count matrix uses `LOC...` gene IDs, whereas the eggNOG file is annotated at
the protein-query level. This section uses the local GFF file as a bridge from
protein IDs to gene IDs, then keeps only annotations that can be lifted onto
genes present in the count matrix. That makes GO and KEGG enrichment compatible
with the DESeq2 result tables.

```{r annotation-maps}
read_emapper <- function(path) {
  cols <- c(
    "query", "seed_ortholog", "evalue", "score", "eggNOG_OGs",
    "max_annot_lvl", "COG_category", "Description", "Preferred_name",
    "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
    "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC", "CAZy",
    "BiGG_Reaction", "PFAMs"
  )
  tab <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", fill = TRUE)
  names(tab) <- cols[seq_len(ncol(tab))]
  tab |>
    dplyr::mutate(protein_id = stringr::str_extract(query, "XP_[0-9]+[.][0-9]+"))
}

parse_cds_gene_bridge <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  gff |>
    dplyr::filter(type == "CDS") |>
    dplyr::transmute(
      protein_id = stringr::str_match(attributes, "protein_id=([^;]+)")[, 2],
      gene_id = stringr::str_match(attributes, "gene=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(protein_id), !is.na(gene_id)) |>
    dplyr::distinct()
}

parse_gene_biotypes <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  gff |>
    dplyr::filter(type == "gene") |>
    dplyr::transmute(
      gene_id = stringr::str_match(attributes, "gene=([^;]+)")[, 2],
      gene_biotype = stringr::str_match(attributes, "gene_biotype=([^;]+)")[, 2],
      gene_description = stringr::str_match(attributes, "description=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(gene_id)) |>
    dplyr::distinct(gene_id, .keep_all = TRUE)
}

eggnog <- read_emapper(eggnog_file)
protein_to_gene <- parse_cds_gene_bridge(gff_file)
gene_biotypes <- parse_gene_biotypes(gff_file)

annotation <- eggnog |>
  dplyr::left_join(protein_to_gene, by = "protein_id") |>
  dplyr::filter(!is.na(gene_id)) |>
  dplyr::left_join(gene_biotypes, by = "gene_id") |>
  dplyr::distinct(gene_id, .keep_all = TRUE)

counts_raw <- read.csv(counts_file, check.names = FALSE)
gene_universe <- unique(counts_raw[[1]])
annotation <- annotation |>
  dplyr::filter(gene_id %in% gene_universe)

annotation_summary <- data.frame(
  genes_in_count_matrix = length(gene_universe),
  annotated_genes_lifted_to_LOC = dplyr::n_distinct(annotation$gene_id),
  genes_with_gff_biotype = sum(gene_universe %in% gene_biotypes$gene_id),
  genes_with_GO = sum(annotation$GOs != "-" & annotation$GOs != ""),
  genes_with_KEGG_KO = sum(annotation$KEGG_ko != "-" & annotation$KEGG_ko != ""),
  genes_with_KEGG_pathway = sum(annotation$KEGG_Pathway != "-" &
                                  annotation$KEGG_Pathway != "")
)

knitr::kable(annotation_summary)
write.csv(annotation_summary, file.path(out_dir, "annotation_summary.csv"),
          row.names = FALSE)
```

```{r make-term-maps}
term2gene_go <- annotation |>
  dplyr::select(gene_id, GOs) |>
  tidyr::separate_rows(GOs, sep = ",") |>
  dplyr::filter(grepl("^GO:", GOs)) |>
  dplyr::transmute(term = GOs, gene = gene_id) |>
  dplyr::distinct()

go_names <- AnnotationDbi::mapIds(
  GO.db::GO.db,
  keys = unique(term2gene_go$term),
  keytype = "GOID",
  column = "TERM",
  multiVals = "first"
)

go_ontology <- AnnotationDbi::mapIds(
  GO.db::GO.db,
  keys = unique(term2gene_go$term),
  keytype = "GOID",
  column = "ONTOLOGY",
  multiVals = "first"
)

term2name_go <- data.frame(
  term = names(go_names),
  name = unname(go_names),
  ontology = unname(go_ontology)
) |>
  dplyr::filter(!is.na(name))

term2gene_kegg_pathway <- annotation |>
  dplyr::select(gene_id, KEGG_Pathway) |>
  tidyr::separate_rows(KEGG_Pathway, sep = ",") |>
  dplyr::filter(grepl("^ko[0-9]+$", KEGG_Pathway)) |>
  dplyr::transmute(term = KEGG_Pathway, gene = gene_id) |>
  dplyr::distinct()

term2name_kegg_pathway <- term2gene_kegg_pathway |>
  dplyr::distinct(term) |>
  dplyr::mutate(name = term)

write.csv(term2gene_go, file.path(out_dir, "GO_TERM2GENE.csv"),
          row.names = FALSE)
write.csv(term2name_go, file.path(out_dir, "GO_TERM2NAME.csv"),
          row.names = FALSE)
write.csv(term2gene_kegg_pathway, file.path(out_dir, "KEGG_pathway_TERM2GENE.csv"),
          row.names = FALSE)

write.table(term2gene_go, file.path(out_dir, "eggNOG_GO_TERM2GENE.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
write.table(term2name_go, file.path(out_dir, "eggNOG_GO_TERM2NAME.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
write.table(term2gene_kegg_pathway,
            file.path(out_dir, "eggNOG_KEGG_pathway_TERM2GENE.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)

annotation_source_manifest <- data.frame(
  Annotation_source = c("eggNOG GO", "eggNOG KEGG pathway"),
  Source_file = c(eggnog_file, eggnog_file),
  Gene_id_bridge = c(gff_file, gff_file),
  Interpretation = c(
    "GO terms parsed from eggNOG protein annotations and lifted to LOC gene IDs.",
    "KEGG pathway IDs parsed from eggNOG protein annotations and lifted to LOC gene IDs."
  )
)
knitr::kable(annotation_source_manifest)
write.csv(annotation_source_manifest,
          file.path(out_dir, "annotation_source_manifest.csv"),
          row.names = FALSE)
```

## Enrichment input DEG sets

For enrichment, each DEG table is split by source and direction. Direction-aware
sets are helpful because genes higher in infected samples and genes higher in
controls can point to different biological processes, even when they come from
the same statistical contrast.

```{r deg-sets}
result_files <- list.files(file.path(project_dir, params$latest_results_root),
                           pattern = "\\.csv$", recursive = TRUE,
                           full.names = TRUE)
result_files <- result_files[grepl("DESeq2_|all_.*results|coefficients",
                                   basename(result_files))]

result_file_info <- data.frame(
  file = result_files,
  short_name = basename(result_files),
  modified = file.info(result_files)$mtime,
  stringsAsFactors = FALSE
)
result_file_info <- result_file_info[order(result_file_info$short_name,
                                           -as.numeric(result_file_info$modified)), ]
result_file_info <- result_file_info[!duplicated(result_file_info$short_name), ]
result_files <- result_file_info$file

read_deg_table <- function(path) {
  tab <- read.csv(path, check.names = FALSE)
  gene_col <- intersect(c("gene_id", "gene", "GeneID"), names(tab))[1]
  if (is.na(gene_col) || !"padj" %in% names(tab) ||
      !"log2FoldChange" %in% names(tab)) return(NULL)
  tab |>
    dplyr::transmute(
      gene_id = as.character(.data[[gene_col]]),
      log2FoldChange = log2FoldChange,
      padj = padj,
      source = tools::file_path_sans_ext(basename(path)),
      direction = dplyr::case_when(
        !is.na(padj) & padj < params$alpha &
          log2FoldChange >= params$lfc_threshold ~ "UP",
        !is.na(padj) & padj < params$alpha &
          log2FoldChange <= -params$lfc_threshold ~ "DOWN",
        !is.na(padj) & padj < params$alpha &
          abs(log2FoldChange) >= params$lfc_threshold ~ "ALL",
        TRUE ~ "NOT_SIGNIFICANT"
      )
    )
}

deg_tables <- lapply(result_files, read_deg_table)
deg_tables <- deg_tables[!vapply(deg_tables, is.null, logical(1))]
deg_long <- dplyr::bind_rows(deg_tables)
deg_long <- deg_long |>
  dplyr::filter(direction != "NOT_SIGNIFICANT") |>
  dplyr::mutate(source_direction = paste(source, direction, sep = "__"))

deg_set_summary <- deg_long |>
  dplyr::distinct(source_direction, gene_id) |>
  dplyr::count(source_direction, name = "n_genes") |>
  dplyr::arrange(dplyr::desc(n_genes))

DT::datatable(deg_set_summary, options = list(pageLength = 15))
write.csv(deg_set_summary, file.path(out_dir, "enrichment_deg_set_summary.csv"),
          row.names = FALSE)
```

## Coding and long non-coding DEG classes

This section mirrors the AGY DEG magnitude figure. It asks whether each DEG set
is mostly protein-coding, long non-coding RNA, or another/unknown gene class.
Gene class comes from `gene_biotype` in the local GFF, while DEG direction comes
from the current DESeq2 result tables.

```{r coding-lncrna-deg-summary, fig.height=7, fig.width=11}
classify_gene <- function(x) {
  dplyr::case_when(
    x == "protein_coding" ~ "Coding genes",
    x == "lncRNA" ~ "Long non-coding RNA",
    is.na(x) ~ "Other / unknown",
    TRUE ~ "Other non-coding"
  )
}

deg_biotype <- deg_long |>
  dplyr::filter(direction %in% c("UP", "DOWN")) |>
  dplyr::left_join(gene_biotypes, by = "gene_id") |>
  dplyr::mutate(
    Gene_class = classify_gene(gene_biotype),
    Direction = dplyr::recode(direction,
                              UP = "Higher in numerator/infected",
                              DOWN = "Higher in denominator/control"),
    signed_direction = ifelse(direction == "UP", 1, -1),
    source_label = stringr::str_replace_all(source, "_", " ")
  ) |>
  dplyr::distinct(source, source_label, direction, Direction,
                  gene_id, Gene_class, .keep_all = TRUE)

focus_sources <- c(
  "all_infection_within_diet_results",
  "DESeq2_infected_vs_control_diet_33",
  "DESeq2_infected_vs_control_diet_50",
  "DESeq2_infected_vs_control_diet_83",
  "all_model_coefficients",
  "interaction_coefficients",
  "all_diet_pairwise_results"
)

deg_biotype_focus <- deg_biotype |>
  dplyr::filter(source %in% focus_sources)

biotype_counts <- deg_biotype_focus |>
  dplyr::count(source, source_label, direction, Direction, Gene_class,
               name = "n_genes") |>
  dplyr::mutate(
    signed_n = ifelse(direction == "UP", n_genes, -n_genes),
    DEG_class = paste(Direction, Gene_class, sep = " - ")
  )

source_totals <- biotype_counts |>
  dplyr::group_by(source, source_label, Direction, direction) |>
  dplyr::summarise(
    n_genes = sum(n_genes),
    signed_n = ifelse(dplyr::first(direction) == "UP",
                      sum(n_genes), -sum(n_genes)),
    .groups = "drop"
  )

source_order <- biotype_counts |>
  dplyr::group_by(source_label) |>
  dplyr::summarise(total = sum(n_genes), .groups = "drop") |>
  dplyr::arrange(total) |>
  dplyr::pull(source_label)

biotype_counts$source_label <- factor(biotype_counts$source_label,
                                      levels = source_order)
source_totals$source_label <- factor(source_totals$source_label,
                                     levels = source_order)

class_colors <- c(
  "Higher in numerator/infected - Coding genes" = "#8B1E3F",
  "Higher in numerator/infected - Long non-coding RNA" = "#D98BA3",
  "Higher in numerator/infected - Other non-coding" = "#F2B8A0",
  "Higher in numerator/infected - Other / unknown" = "#E5B567",
  "Higher in denominator/control - Coding genes" = "#1F5F8B",
  "Higher in denominator/control - Long non-coding RNA" = "#93B7D4",
  "Higher in denominator/control - Other non-coding" = "#B8C7D9",
  "Higher in denominator/control - Other / unknown" = "#C9CED6"
)

max_biotype <- max(abs(source_totals$signed_n), 1)
source_totals$label_x <- ifelse(source_totals$signed_n >= 0,
                                source_totals$signed_n + 0.04 * max_biotype,
                                source_totals$signed_n - 0.04 * max_biotype)

p_biotype <- ggplot2::ggplot(
  biotype_counts,
  ggplot2::aes(x = signed_n, y = source_label, fill = DEG_class)
) +
  ggplot2::geom_vline(xintercept = 0, color = "#374151", linewidth = 0.35) +
  ggplot2::geom_col(width = 0.68) +
  ggplot2::geom_text(
    data = source_totals,
    ggplot2::aes(x = label_x, y = source_label, label = n_genes),
    inherit.aes = FALSE,
    size = 3,
    fontface = "bold"
  ) +
  ggplot2::scale_x_continuous(labels = abs,
                              limits = c(-1.2 * max_biotype,
                                         1.2 * max_biotype)) +
  ggplot2::scale_fill_manual(values = class_colors, drop = FALSE) +
  ggplot2::theme_classic() +
  ggplot2::theme(
    legend.position = "bottom",
    legend.title = ggplot2::element_blank(),
    legend.text = ggplot2::element_text(size = 8),
    axis.text.y = ggplot2::element_text(size = 8)
  ) +
  ggplot2::labs(
    title = "Coding and long non-coding RNA DEGs",
    subtitle = "Dark = coding genes; lighter shades = lncRNA, other non-coding, or unknown biotype",
    x = "Number of DEGs",
    y = NULL
  )

p_biotype
ggplot2::ggsave(file.path(out_dir, "coding_lncRNA_deg_barplot.png"),
                p_biotype, width = 11, height = 7, dpi = 300)
write.csv(biotype_counts,
          file.path(out_dir, "coding_lncRNA_deg_counts.csv"),
          row.names = FALSE)

lncrna_deg_table <- deg_biotype |>
  dplyr::filter(Gene_class == "Long non-coding RNA") |>
  dplyr::left_join(annotation |>
                     dplyr::select(gene_id, Preferred_name, Description,
                                   GOs, KEGG_ko, KEGG_Pathway),
                   by = "gene_id") |>
  dplyr::arrange(source, direction, gene_id)

DT::datatable(lncrna_deg_table,
              options = list(pageLength = 10, scrollX = TRUE))
write.csv(lncrna_deg_table,
          file.path(out_dir, "lncRNA_deg_table.csv"),
          row.names = FALSE)
```

## GO and KEGG enrichment

GO and KEGG enrichment asks whether a DEG set contains more genes from a
functional term than expected from the annotated gene universe. The universe is
restricted to genes that are both present in the count matrix and represented in
the corresponding annotation map, so the background matches what this dataset
could actually test.

```{r run-enrichment}
run_enrichment <- function(genes, term2gene, term2name, universe) {
  genes <- unique(genes[genes %in% universe])
  if (length(genes) < 5) return(NULL)
  clusterProfiler::enricher(
    gene = genes,
    universe = universe,
    TERM2GENE = term2gene,
    TERM2NAME = term2name[, c("term", "name")],
    pvalueCutoff = 0.1,
    pAdjustMethod = "BH",
    qvalueCutoff = 0.2
  )
}

enrichment_sets <- split(deg_long$gene_id, deg_long$source_direction)
universe_go <- unique(term2gene_go$gene)
universe_kegg <- unique(term2gene_kegg_pathway$gene)

go_results <- lapply(enrichment_sets, run_enrichment,
                     term2gene = term2gene_go,
                     term2name = term2name_go,
                     universe = universe_go)
kegg_results <- lapply(enrichment_sets, run_enrichment,
                       term2gene = term2gene_kegg_pathway,
                       term2name = term2name_kegg_pathway,
                       universe = universe_kegg)

enrich_to_df <- function(result_list, database) {
  pieces <- lapply(names(result_list), function(nm) {
    res <- result_list[[nm]]
    if (is.null(res)) return(NULL)
    df <- as.data.frame(res)
    if (nrow(df) == 0) return(NULL)
    df$set <- nm
    df$database <- database
    df
  })
  dplyr::bind_rows(pieces)
}

go_enrichment_df <- enrich_to_df(go_results, "GO")
kegg_enrichment_df <- enrich_to_df(kegg_results, "KEGG_pathway")
all_enrichment <- dplyr::bind_rows(go_enrichment_df, kegg_enrichment_df)

write.csv(go_enrichment_df, file.path(out_dir, "GO_enrichment_all_sets.csv"),
          row.names = FALSE)
write.csv(kegg_enrichment_df, file.path(out_dir, "KEGG_pathway_enrichment_all_sets.csv"),
          row.names = FALSE)
write.csv(all_enrichment, file.path(out_dir, "all_enrichment_results.csv"),
          row.names = FALSE)

if (nrow(all_enrichment) == 0) {
  cat("No enriched terms passed the current cutoffs. Try inspecting the DEG set sizes or relaxing qvalueCutoff.")
} else {
  DT::datatable(
    all_enrichment |>
      dplyr::select(database, set, ID, Description, GeneRatio, BgRatio,
                    p.adjust, qvalue, geneID) |>
      dplyr::arrange(p.adjust),
    options = list(pageLength = 15)
  )
}
```

## eggNOG GO enrichment by ontology

This panel follows the same logic used for the AGY gregaria eggNOG layer:
foreground genes come from the DESeq2 DEG sets, the background is the set of
tested genes with eggNOG GO annotations, and the GO ontology labels come from
`GO.db`. The enrichment is therefore a custom eggNOG GO enrichment, not an
NCBI/organism-specific `enrichGO()` call.

```{r eggnog-go-ontology-figure, fig.height=10, fig.width=13}
if (nrow(go_enrichment_df) > 0 && "ID" %in% names(go_enrichment_df)) {
  go_enrichment_ontology <- go_enrichment_df |>
    dplyr::left_join(term2name_go |> dplyr::select(term, ontology),
                     by = c("ID" = "term")) |>
    dplyr::mutate(
      ontology = dplyr::case_when(
        ontology == "BP" ~ "Biological process",
        ontology == "MF" ~ "Molecular function",
        ontology == "CC" ~ "Cellular component",
        TRUE ~ "GO term"
      ),
      neg_log10_padj = -log10(pmax(p.adjust, .Machine$double.xmin)),
      term_label = paste0(stringr::str_trunc(Description, 52), " [", ID, "]"),
      set_label = stringr::str_replace_all(set, "_", " ")
    )
} else {
  go_enrichment_ontology <- data.frame()
}

write.csv(go_enrichment_ontology,
          file.path(out_dir, "eggNOG_GO_enrichment_with_ontology.csv"),
          row.names = FALSE)

if (nrow(go_enrichment_ontology) > 0) {
  go_plot_df <- go_enrichment_ontology |>
    dplyr::filter(!is.na(p.adjust)) |>
    dplyr::group_by(set_label, ontology) |>
    dplyr::arrange(p.adjust, dplyr::desc(Count), .by_group = TRUE) |>
    dplyr::slice_head(n = 5) |>
    dplyr::ungroup() |>
    dplyr::arrange(ontology, set_label, p.adjust) |>
    dplyr::mutate(
      term_label = forcats::fct_reorder(term_label, neg_log10_padj)
    )

  p_go_ontology <- ggplot2::ggplot(
    go_plot_df,
    ggplot2::aes(x = neg_log10_padj, y = term_label)
  ) +
    ggplot2::geom_vline(xintercept = -log10(0.05),
                        linetype = "dashed", color = "#374151",
                        linewidth = 0.35) +
    ggplot2::geom_point(ggplot2::aes(size = Count, fill = ontology),
                        shape = 21, color = "#1f2937", alpha = 0.9,
                        stroke = 0.35) +
    ggplot2::facet_grid(ontology ~ set_label, scales = "free_y",
                        space = "free_y") +
    ggplot2::scale_fill_manual(values = c(
      "Biological process" = "#237C8E",
      "Molecular function" = "#B8643C",
      "Cellular component" = "#5B6B4A",
      "GO term" = "#7B8794"
    )) +
    ggplot2::scale_size(range = c(2, 7)) +
    ggplot2::theme_bw() +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      strip.background = ggplot2::element_rect(fill = "#F1F4F3",
                                               color = "#C8D0CC"),
      strip.text.x = ggplot2::element_text(size = 7, face = "bold"),
      strip.text.y = ggplot2::element_text(size = 8, face = "bold", angle = 0),
      axis.text.y = ggplot2::element_text(size = 6),
      legend.position = "bottom"
    ) +
    ggplot2::labs(
      title = "eggNOG GO enrichment by ontology",
      subtitle = "Top GO terms per DEG set and ontology; dashed line marks FDR = 0.05",
      x = "-log10 adjusted p-value",
      y = NULL,
      fill = "GO ontology",
      size = "DEGs in term"
    )

  print(p_go_ontology)
  ggplot2::ggsave(file.path(out_dir,
                            "eggNOG_GO_enrichment_by_ontology.png"),
                  p_go_ontology, width = 13, height = 10, dpi = 300)
}
```

## Enrichment overview figure

The overview dotplot shows the strongest enriched terms per DEG set. Dot size is
the number of genes from the DEG set in the term, and color is the adjusted
p-value strength, which makes it easier to compare both support and significance
across many DEG lists.

```{r enrichment-overview, fig.height=9, fig.width=11}
if (nrow(all_enrichment) > 0) {
  plot_df <- all_enrichment |>
    dplyr::group_by(database, set) |>
    dplyr::slice_min(p.adjust, n = params$top_terms, with_ties = FALSE) |>
    dplyr::ungroup() |>
    dplyr::mutate(
      term_label = stringr::str_trunc(Description, 55),
      term_label = forcats::fct_reorder(term_label, -log10(p.adjust)),
      set = stringr::str_replace_all(set, "_", " ")
    )

  p_enrich <- ggplot2::ggplot(
    plot_df,
    ggplot2::aes(x = set, y = term_label, size = Count,
                 color = -log10(p.adjust))
  ) +
    ggplot2::geom_point(alpha = 0.85) +
    ggplot2::facet_wrap(~ database, scales = "free_y") +
    ggplot2::scale_color_viridis_c(option = "magma") +
    ggplot2::theme_classic() +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)) +
    ggplot2::labs(
      title = "Top enriched GO terms and KEGG pathways",
      x = "DEG set and direction",
      y = "Enriched term",
      size = "Gene count",
      color = "-log10 adjusted P"
    )

  print(p_enrich)
  ggplot2::ggsave(file.path(out_dir, "enrichment_dotplot_by_set.png"),
                  p_enrich, width = 11, height = 9, dpi = 300)
}
```

## Term overlap across DEG sets

This final heatmap highlights functional terms that recur in multiple DEG sets.
Shared terms can point to processes that are robust across contrast definitions,
whereas terms seen in only one set may be more diet-, infection-, or
direction-specific.

```{r term-overlap, fig.height=7, fig.width=10}
if (nrow(all_enrichment) > 0) {
  term_overlap <- all_enrichment |>
    dplyr::filter(p.adjust < 0.1) |>
    dplyr::distinct(database, ID, Description, set) |>
    dplyr::add_count(database, ID, name = "n_sets") |>
    dplyr::filter(n_sets >= 2)

  write.csv(term_overlap, file.path(out_dir, "enriched_terms_seen_in_multiple_sets.csv"),
            row.names = FALSE)

  if (nrow(term_overlap) > 0) {
    top_overlap_terms <- term_overlap |>
      dplyr::distinct(database, ID, Description, n_sets) |>
      dplyr::arrange(dplyr::desc(n_sets)) |>
      head(40)

    overlap_plot_df <- term_overlap |>
      dplyr::semi_join(top_overlap_terms, by = c("database", "ID", "Description", "n_sets")) |>
      dplyr::mutate(
        term_label = stringr::str_trunc(Description, 60),
        set = stringr::str_replace_all(set, "_", " ")
      )

    p_overlap <- ggplot2::ggplot(overlap_plot_df,
                                 ggplot2::aes(x = set, y = term_label)) +
      ggplot2::geom_tile(fill = "#2b8cbe") +
      ggplot2::facet_wrap(~ database, scales = "free_y") +
      ggplot2::theme_classic() +
      ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)) +
      ggplot2::labs(
        title = "Enriched terms shared by two or more DEG sets",
        x = "DEG set and direction",
        y = "Shared enriched term"
      )

    print(p_overlap)
    ggplot2::ggsave(file.path(out_dir, "enriched_term_overlap_heatmap.png"),
                    p_overlap, width = 10, height = 7, dpi = 300)
  } else {
    cat("No enriched terms are currently shared by two or more DEG sets at p.adjust < 0.1.")
  }
}
```

```{r session-info}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
