---
title: "Main-chromosome GO term and KEGG enrichment"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  eggnog_file: "data/GO_Annotations/GCF_023897955.1_iqSchGreg1.2_Arthopoda_one2one.emapper.annotations"
  gff_file: "data/reference/GCF_023897955.1_iqSchGreg1.2_genomic.gff"
  latest_results_root: "output/rmd_runs"
  alpha: 0.05
  lfc_threshold: 1
  top_terms: 8
  output_tag: "go-term-enrichment"
---

## Purpose

<style>
.main-container table.table:not(.dataTable) {
  display: block;
  width: 100% !important;
  max-width: 100%;
  table-layout: fixed;
  overflow-x: auto;
}
.main-container table.table:not(.dataTable) th,
.main-container table.table:not(.dataTable) td {
  overflow-wrap: anywhere;
  white-space: normal;
}
.main-container .dataTables_wrapper {
  max-width: 100%;
  overflow-x: auto;
}
</style>

This page keeps the eggNOG-based functional enrichment with the conservative
main-nuclear-chromosome differential-expression results. It uses the newest
completed placed/unplaced analysis table on disk, retains only genes whose
official assembly-derived class is `Nuclear chromosome`, and lifts
eggNOG protein annotations onto the `LOC...` gene IDs in the count matrix, and
tests infection DEG sets for enriched GO terms and KEGG pathways.

The key point is that this is a custom enrichment workflow for this reference:
the foregrounds are current main-chromosome DEG sets, the background is the
expression-filtered main-chromosome gene universe tested by DESeq2, and GO
names/ontologies come from `GO.db`. Unplaced-scaffold and mitochondrial genes
are intentionally excluded from both foreground and background here and remain
available on the placement-sensitivity page.

The first enrichment section keeps the infection response non-directional,
which asks what functions are over-represented among infection DEGs regardless
of whether genes increase or decrease. A later section adds the requested
direction-aware contrasts: infected-versus-control within each diet, diet
contrasts among controls only, and diet contrasts among infected samples only.

Important KEGG scope note: the foreground genes and enrichment background are
from the *S. gregaria* fat body count matrix, but the KEGG pathway IDs are
`koXXXXX` KEGG Orthology reference pathways carried by eggNOG. These are not
Insecta-only organism pathways. Some enriched labels can therefore be broad
cross-domain maps or microbial/prokaryote-associated KEGG reference labels,
even though the genes tested here are locust gene IDs. The pathway-scope table
below flags those cases for interpretation.

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c(
  "dplyr", "tidyr", "stringr", "ggplot2", "forcats", "clusterProfiler",
  "GO.db", "AnnotationDbi", "DT"
)
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

eggnog_file <- file.path(project_dir, params$eggnog_file)
gff_file <- file.path(project_dir, params$gff_file)
results_root <- file.path(project_dir, params$latest_results_root)

placed_result_files <- list.files(
  results_root,
  pattern = "^placed_unplaced_all_deseq2_results[.]csv$",
  recursive = TRUE,
  full.names = TRUE
)
if (length(placed_result_files) == 0) {
  stop("Build analysis/28-placed-unplaced-scaffold-analysis.Rmd before this page.")
}
primary_results_file <- placed_result_files[
  which.max(file.info(placed_result_files)$mtime)
]
placed_results <- read.csv(primary_results_file, check.names = FALSE) |>
  dplyr::filter(
    mapping_branch == "Competitive host",
    analysis_scope == "Placed/host-assigned",
    sequence_class == "Nuclear chromosome"
  )
if (nrow(placed_results) == 0) {
  stop("The newest result file has no competitive main-chromosome rows.")
}

diet_colors <- c("33" = "#E6862E", "50" = "#2F8F5B", "83" = "#7B4FA3")
ontology_colors <- c(
  "Biological process" = "#2F8F5B",
  "Molecular function" = "#E6862E",
  "Cellular component" = "#7B4FA3"
)
kegg_theme_colors <- c(
  "Immune and stress signaling" = "#C23B22",
  "Storage, lipid transport, and endocrine nutrition" = "#D7832F",
  "Amino acid and nitrogen metabolism" = "#6F4AA8",
  "Carbohydrate and energy metabolism" = "#2C7FB8",
  "Lipid and membrane metabolism" = "#2F8F5B",
  "Detoxification and secondary metabolism" = "#8C6D31",
  "Transport and environmental response" = "#00A6A6",
  "Genetic information processing" = "#6B7280",
  "Cellular process" = "#9C5DA5",
  "Broad metabolic map" = "#4B5563",
  "Reference disease pathway" = "#A0AEC0",
  "Other KEGG pathway" = "#7B8794"
)
kegg_scope_colors <- c(
  "Eukaryote/metazoan-compatible" = "#2F8F5B",
  "Core metabolism or cross-domain" = "#2C7FB8",
  "Broad KEGG reference map" = "#4B5563",
  "Microbial/prokaryote-associated label" = "#D81B60",
  "Reference disease pathway" = "#7B4FA3",
  "Unclassified KEGG reference label" = "#F59E0B"
)

direction_colors <- c(
  "Upregulated" = "#E41A1C",
  "Downregulated" = "#1E88E5"
)
```

## Build eggNOG annotation maps

eggNOG reports annotations at the protein-query level, while DESeq2 reports
uses `LOC...` gene IDs. The GFF provides the bridge from CDS protein IDs to gene
IDs. Only expression-filtered genes tested in the competitive main-chromosome
analysis are kept in the enrichment background.

```{r annotation-maps}
read_emapper <- function(path) {
  cols <- c(
    "query", "seed_ortholog", "evalue", "score", "eggNOG_OGs",
    "max_annot_lvl", "COG_category", "Description", "Preferred_name",
    "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
    "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC", "CAZy",
    "BiGG_Reaction", "PFAMs"
  )
  tab <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", fill = TRUE)
  names(tab) <- cols[seq_len(ncol(tab))]
  tab |>
    dplyr::mutate(protein_id = stringr::str_extract(query, "XP_[0-9]+[.][0-9]+"))
}

parse_cds_gene_bridge <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  gff |>
    dplyr::filter(type == "CDS") |>
    dplyr::transmute(
      protein_id = stringr::str_match(attributes, "protein_id=([^;]+)")[, 2],
      gene_id = stringr::str_match(attributes, "gene=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(protein_id), !is.na(gene_id)) |>
    dplyr::distinct()
}

eggnog <- read_emapper(eggnog_file)
protein_to_gene <- parse_cds_gene_bridge(gff_file)
gene_universe <- unique(as.character(placed_results$gene_id))

annotation <- eggnog |>
  dplyr::left_join(protein_to_gene, by = "protein_id") |>
  dplyr::filter(!is.na(gene_id), gene_id %in% gene_universe) |>
  dplyr::distinct(gene_id, .keep_all = TRUE)

annotation_summary <- data.frame(
  placed_host_genes_tested_by_deseq2 = length(gene_universe),
  annotated_genes_lifted_to_LOC = dplyr::n_distinct(annotation$gene_id),
  genes_with_GO = sum(annotation$GOs != "-" & annotation$GOs != ""),
  genes_with_KEGG_KO = sum(annotation$KEGG_ko != "-" & annotation$KEGG_ko != ""),
  genes_with_KEGG_pathway = sum(annotation$KEGG_Pathway != "-" &
                                  annotation$KEGG_Pathway != "")
)

knitr::kable(annotation_summary)
write.csv(annotation_summary, file.path(out_dir, "annotation_summary.csv"),
          row.names = FALSE)
```

```{r term-maps}
term2gene_go <- annotation |>
  dplyr::select(gene_id, GOs) |>
  tidyr::separate_rows(GOs, sep = ",") |>
  dplyr::filter(grepl("^GO:", GOs)) |>
  dplyr::transmute(term = GOs, gene = gene_id) |>
  dplyr::distinct()

go_names <- AnnotationDbi::mapIds(
  GO.db::GO.db,
  keys = unique(term2gene_go$term),
  keytype = "GOID",
  column = "TERM",
  multiVals = "first"
)

go_ontology <- AnnotationDbi::mapIds(
  GO.db::GO.db,
  keys = unique(term2gene_go$term),
  keytype = "GOID",
  column = "ONTOLOGY",
  multiVals = "first"
)

term2name_go <- data.frame(
  term = names(go_names),
  name = unname(go_names),
  ontology = unname(go_ontology)
) |>
  dplyr::filter(!is.na(name))

term2gene_kegg_pathway <- annotation |>
  dplyr::select(gene_id, KEGG_Pathway) |>
  tidyr::separate_rows(KEGG_Pathway, sep = ",") |>
  dplyr::filter(grepl("^ko[0-9]+$", KEGG_Pathway)) |>
  dplyr::transmute(term = KEGG_Pathway, gene = gene_id) |>
  dplyr::distinct()

kegg_fallback_names <- c(
  ko00010 = "Glycolysis / Gluconeogenesis",
  ko00020 = "Citrate cycle (TCA cycle)",
  ko00030 = "Pentose phosphate pathway",
  ko00040 = "Pentose and glucuronate interconversions",
  ko00051 = "Fructose and mannose metabolism",
  ko00052 = "Galactose metabolism",
  ko00053 = "Ascorbate and aldarate metabolism",
  ko00061 = "Fatty acid biosynthesis",
  ko00062 = "Fatty acid elongation",
  ko00071 = "Fatty acid degradation",
  ko00190 = "Oxidative phosphorylation",
  ko00220 = "Arginine biosynthesis",
  ko00230 = "Purine metabolism",
  ko00240 = "Pyrimidine metabolism",
  ko00250 = "Alanine, aspartate and glutamate metabolism",
  ko00260 = "Glycine, serine and threonine metabolism",
  ko00270 = "Cysteine and methionine metabolism",
  ko00280 = "Valine, leucine and isoleucine degradation",
  ko00330 = "Arginine and proline metabolism",
  ko00340 = "Histidine metabolism",
  ko00350 = "Tyrosine metabolism",
  ko00360 = "Phenylalanine metabolism",
  ko00500 = "Starch and sucrose metabolism",
  ko00520 = "Amino sugar and nucleotide sugar metabolism",
  ko00561 = "Glycerolipid metabolism",
  ko00564 = "Glycerophospholipid metabolism",
  ko00565 = "Ether lipid metabolism",
  ko00590 = "Arachidonic acid metabolism",
  ko00591 = "Linoleic acid metabolism",
  ko00592 = "alpha-Linolenic acid metabolism",
  ko00620 = "Pyruvate metabolism",
  ko00630 = "Glyoxylate and dicarboxylate metabolism",
  ko00640 = "Propanoate metabolism",
  ko00650 = "Butanoate metabolism",
  ko00670 = "One carbon pool by folate",
  ko00730 = "Thiamine metabolism",
  ko00860 = "Porphyrin metabolism",
  ko00900 = "Terpenoid backbone biosynthesis",
  ko00910 = "Nitrogen metabolism",
  ko00920 = "Sulfur metabolism",
  ko00930 = "Caprolactam degradation",
  ko00970 = "Aminoacyl-tRNA biosynthesis",
  ko01040 = "Biosynthesis of unsaturated fatty acids",
  ko01100 = "Metabolic pathways",
  ko01110 = "Biosynthesis of secondary metabolites",
  ko01120 = "Microbial metabolism in diverse environments",
  ko01130 = "Biosynthesis of antibiotics",
  ko01200 = "Carbon metabolism",
  ko01212 = "Fatty acid metabolism",
  ko01220 = "Degradation of aromatic compounds",
  ko01230 = "Biosynthesis of amino acids",
  ko02010 = "ABC transporters",
  ko02020 = "Two-component system",
  ko03010 = "Ribosome",
  ko03013 = "Nucleocytoplasmic transport",
  ko03015 = "mRNA surveillance pathway",
  ko03018 = "RNA degradation",
  ko03022 = "Basal transcription factors",
  ko03030 = "DNA replication",
  ko03040 = "Spliceosome",
  ko03050 = "Proteasome",
  ko03060 = "Protein export",
  ko03320 = "PPAR signaling pathway",
  ko04010 = "MAPK signaling pathway",
  ko04014 = "Ras signaling pathway",
  ko04020 = "Calcium signaling pathway",
  ko04024 = "cAMP signaling pathway",
  ko04110 = "Cell cycle",
  ko04120 = "Ubiquitin mediated proteolysis",
  ko04130 = "SNARE interactions in vesicular transport",
  ko04141 = "Protein processing in endoplasmic reticulum",
  ko04142 = "Lysosome",
  ko04144 = "Endocytosis",
  ko04145 = "Phagosome",
  ko04146 = "Peroxisome",
  ko04147 = "Exosome",
  ko04150 = "mTOR signaling pathway",
  ko04151 = "PI3K-Akt signaling pathway",
  ko04152 = "AMPK signaling pathway",
  ko04210 = "Apoptosis",
  ko04213 = "Longevity regulating pathway - multiple species",
  ko04217 = "Necroptosis",
  ko04218 = "Cellular senescence",
  ko04270 = "Vascular smooth muscle contraction",
  ko04310 = "Wnt signaling pathway",
  ko04330 = "Notch signaling pathway",
  ko04340 = "Hedgehog signaling pathway",
  ko04350 = "TGF-beta signaling pathway",
  ko04370 = "VEGF signaling pathway",
  ko04371 = "Apelin signaling pathway",
  ko04510 = "Focal adhesion",
  ko04512 = "ECM-receptor interaction",
  ko04514 = "Cell adhesion molecules",
  ko04520 = "Adherens junction",
  ko04530 = "Tight junction",
  ko04630 = "JAK-STAT signaling pathway",
  ko04668 = "TNF signaling pathway",
  ko04710 = "Circadian rhythm",
  ko04713 = "Circadian entrainment",
  ko04714 = "Thermogenesis",
  ko04723 = "Retrograde endocannabinoid signaling",
  ko04730 = "Long-term depression",
  ko04910 = "Insulin signaling pathway",
  ko04920 = "Adipocytokine signaling pathway",
  ko04922 = "Glucagon signaling pathway",
  ko04926 = "Relaxin signaling pathway",
  ko04970 = "Salivary secretion",
  ko04972 = "Pancreatic secretion",
  ko04975 = "Fat digestion and absorption",
  ko05010 = "Alzheimer disease",
  ko05012 = "Parkinson disease",
  ko05014 = "Amyotrophic lateral sclerosis",
  ko05016 = "Huntington disease"
)

get_kegg_pathway_names <- function(ids) {
  ids <- unique(ids)
  resolved <- setNames(rep(NA_character_, length(ids)), ids)

  if (requireNamespace("KEGGREST", quietly = TRUE)) {
    kegg_list <- tryCatch(
      KEGGREST::keggList("pathway", "ko"),
      error = function(e) character()
    )
    if (length(kegg_list) > 0) {
      live_ids <- sub("^path:", "", names(kegg_list))
      live_names <- sub(" - Reference pathway$", "", unname(kegg_list))
      live_names <- sub(" - Homo sapiens \\(human\\)$", "", live_names)
      resolved[intersect(ids, live_ids)] <- live_names[match(intersect(ids, live_ids),
                                                             live_ids)]
    }
  }

  fallback_ids <- intersect(ids, names(kegg_fallback_names))
  resolved[fallback_ids] <- ifelse(
    is.na(resolved[fallback_ids]),
    unname(kegg_fallback_names[fallback_ids]),
    resolved[fallback_ids]
  )
  resolved[is.na(resolved)] <- ids[is.na(resolved)]
  resolved
}

classify_kegg_theme <- function(ids, names) {
  ids <- as.character(ids)
  names_lower <- stringr::str_to_lower(names)
  dplyr::case_when(
    stringr::str_detect(ids, "^ko046|^ko040|^ko042") |
      stringr::str_detect(names_lower, "toll|imd|tnf|jak-stat|immune|apoptosis|necrosis|stress") ~
      "Immune and stress signaling",
    stringr::str_detect(ids, "^ko049|^ko033|^ko00981") |
      stringr::str_detect(names_lower, "insulin|glucagon|adipocytokine|ppar|hormone|bile|mineral|fat digestion|secretion|absorption") ~
      "Storage, lipid transport, and endocrine nutrition",
    ids %in% c("ko00220", "ko00230", "ko00240", "ko00250", "ko00260",
               "ko00270", "ko00280", "ko00330", "ko00340", "ko00350",
               "ko00360", "ko00380", "ko00410", "ko01230") |
      stringr::str_detect(names_lower, "amino acid|arginine|proline|glycine|serine|threonine|tryptophan|tyrosine|nitrogen") ~
      "Amino acid and nitrogen metabolism",
    ids %in% c("ko00010", "ko00020", "ko00030", "ko00040", "ko00051",
               "ko00052", "ko00053", "ko00190", "ko00620", "ko00630",
               "ko00640", "ko00650", "ko00720", "ko01200") |
      stringr::str_detect(names_lower, "glycolysis|gluconeogenesis|citrate|carbon metabolism|pentose|pyruvate|oxidative phosphorylation|carbon fixation|ascorbate") ~
      "Carbohydrate and energy metabolism",
    ids %in% c("ko00061", "ko00062", "ko00071", "ko00561", "ko00564",
               "ko00565", "ko00590", "ko00591", "ko00592", "ko01040",
               "ko01212") |
      stringr::str_detect(names_lower, "fatty acid|lipid|glycerolipid|glycerophospholipid|linoleic|arachidonic") ~
      "Lipid and membrane metabolism",
    ids %in% c("ko00480", "ko00900", "ko00920", "ko00930", "ko00970",
               "ko00980", "ko01110", "ko01120", "ko01130", "ko01220") |
      stringr::str_detect(names_lower, "glutathione|xenobiotic|cytochrome|secondary metabolite|microbial metabolism|antibiotic|aromatic") ~
      "Detoxification and secondary metabolism",
    stringr::str_detect(ids, "^ko020") |
      stringr::str_detect(names_lower, "transporter|two-component|environmental") ~
      "Transport and environmental response",
    stringr::str_detect(ids, "^ko030") |
      stringr::str_detect(names_lower, "ribosome|spliceosome|rna|dna|replication|transcription|proteasome") ~
      "Genetic information processing",
    stringr::str_detect(ids, "^ko041|^ko043|^ko045") |
      stringr::str_detect(names_lower, "cell cycle|endocytosis|phagosome|peroxisome|junction|adhesion|notch|wnt|hedgehog") ~
      "Cellular process",
    ids == "ko01100" ~ "Broad metabolic map",
    stringr::str_detect(ids, "^ko050|^ko052") ~ "Reference disease pathway",
    TRUE ~ "Other KEGG pathway"
  )
}

classify_kegg_scope <- function(ids, names) {
  ids <- as.character(ids)
  names_lower <- stringr::str_to_lower(names)
  dplyr::case_when(
    ids %in% c("ko01100", "ko01110", "ko01120", "ko01130",
               "ko01200", "ko01210", "ko01212", "ko01220", "ko01230") ~
      "Broad KEGG reference map",
    stringr::str_detect(
      names_lower,
      "microbial|antibiotic|two-component|bacterial|quorum|peptidoglycan|lipopolysaccharide|aromatic compound"
    ) ~
      "Microbial/prokaryote-associated label",
    stringr::str_detect(ids, "^ko050|^ko052") ~
      "Reference disease pathway",
    stringr::str_detect(ids, "^ko040|^ko041|^ko042|^ko043|^ko045|^ko046|^ko049|^ko033") ~
      "Eukaryote/metazoan-compatible",
    TRUE ~ "Core metabolism or cross-domain"
  )
}

kegg_names <- get_kegg_pathway_names(unique(term2gene_kegg_pathway$term))

term2name_kegg_pathway <- term2gene_kegg_pathway |>
  dplyr::distinct(term) |>
  dplyr::mutate(
    name = unname(kegg_names[term]),
    pathway_theme = classify_kegg_theme(term, name),
    pathway_scope = classify_kegg_scope(term, name)
  )

write.csv(term2gene_go, file.path(out_dir, "eggNOG_GO_TERM2GENE.csv"),
          row.names = FALSE)
write.csv(term2name_go, file.path(out_dir, "eggNOG_GO_TERM2NAME.csv"),
          row.names = FALSE)
write.csv(term2gene_kegg_pathway,
          file.path(out_dir, "eggNOG_KEGG_pathway_TERM2GENE.csv"),
          row.names = FALSE)
write.csv(term2name_kegg_pathway,
          file.path(out_dir, "eggNOG_KEGG_pathway_TERM2NAME.csv"),
          row.names = FALSE)

kegg_scope_summary <- term2name_kegg_pathway |>
  dplyr::count(pathway_scope, pathway_theme, name = "n_pathways") |>
  dplyr::arrange(pathway_scope, pathway_theme)
write.csv(kegg_scope_summary,
          file.path(out_dir, "eggNOG_KEGG_pathway_scope_summary.csv"),
          row.names = FALSE)
```

## KEGG pathway scope

The KEGG pathway names used here are `ko` reference pathway labels from eggNOG,
not insect-only pathway labels. This table flags whether a pathway is best read
as a broad KEGG reference map, a microbial/prokaryote-associated label, a
reference disease pathway, a eukaryote/metazoan-compatible pathway, or a core
metabolism/cross-domain pathway. This flag is for interpretation; it does not
change the enrichment test.

```{r kegg-pathway-scope}
knitr::kable(
  kegg_scope_summary,
  caption = "Interpretive scope of KEGG reference pathways present in the eggNOG annotation map"
)
```

```{r kegg-pathway-scope-figures, fig.width=14, fig.height=10}
scope_levels <- names(kegg_scope_colors)

kegg_scope_summary_plot <- kegg_scope_summary |>
  dplyr::mutate(
    pathway_scope = factor(pathway_scope, levels = scope_levels),
    pathway_theme = stringr::str_wrap(pathway_theme, width = 38)
  )

write.csv(kegg_scope_summary_plot,
          file.path(out_dir, "eggNOG_KEGG_pathway_scope_summary_for_plot.csv"),
          row.names = FALSE)

scope_total_df <- kegg_scope_summary_plot |>
  dplyr::group_by(pathway_scope) |>
  dplyr::summarise(n_pathways = sum(n_pathways), .groups = "drop") |>
  tidyr::complete(
    pathway_scope = factor(scope_levels, levels = scope_levels),
    fill = list(n_pathways = 0)
  ) |>
  dplyr::mutate(
    pathway_scope = forcats::fct_reorder(pathway_scope, n_pathways)
  )

scope_total_plot <- scope_total_df |>
  ggplot2::ggplot(ggplot2::aes(x = pathway_scope, y = n_pathways,
                               fill = pathway_scope)) +
  ggplot2::geom_col(width = 0.72, color = "white", linewidth = 0.3) +
  ggplot2::geom_point(ggplot2::aes(color = pathway_scope), size = 3.2) +
  ggplot2::geom_text(ggplot2::aes(label = n_pathways), hjust = -0.25,
                     size = 4.2) +
  ggplot2::coord_flip() +
  ggplot2::scale_fill_manual(values = kegg_scope_colors, limits = scope_levels,
                             breaks = scope_levels, drop = FALSE) +
  ggplot2::scale_color_manual(values = kegg_scope_colors, limits = scope_levels,
                              breaks = scope_levels, drop = FALSE,
                              guide = "none") +
  ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0, 0.08))) +
  ggplot2::labs(
    title = "KEGG pathway scope in the eggNOG annotation map",
    subtitle = "Counts are unique KEGG pathway IDs linked to genes in the count-matrix background",
    x = NULL,
    y = "Number of KEGG pathway IDs"
  ) +
  ggplot2::theme_minimal(base_size = 15) +
  ggplot2::theme(
    legend.position = "none",
    panel.grid.major.y = ggplot2::element_blank(),
    axis.text.y = ggplot2::element_text(size = 13)
  )

scope_theme_plot <- kegg_scope_summary_plot |>
  dplyr::mutate(
    pathway_theme = forcats::fct_reorder(pathway_theme, n_pathways, sum)
  ) |>
  ggplot2::ggplot(ggplot2::aes(x = pathway_theme, y = n_pathways,
                               fill = pathway_scope)) +
  ggplot2::geom_col(width = 0.74, color = "white", linewidth = 0.25) +
  ggplot2::coord_flip() +
  ggplot2::scale_fill_manual(values = kegg_scope_colors, limits = scope_levels,
                             breaks = scope_levels, drop = FALSE) +
  ggplot2::labs(
    title = "KEGG pathway scope by biological theme",
    subtitle = "Theme groups remain useful, but the scope flag shows which labels are broad or cross-domain",
    x = NULL,
    y = "Number of KEGG pathway IDs",
    fill = "Pathway scope"
  ) +
  ggplot2::theme_minimal(base_size = 15) +
  ggplot2::theme(
    legend.position = "bottom",
    legend.title = ggplot2::element_text(size = 13),
    legend.text = ggplot2::element_text(size = 12),
    panel.grid.major.y = ggplot2::element_blank(),
    axis.text.y = ggplot2::element_text(size = 12)
  )

ggplot2::ggsave(file.path(out_dir, "eggNOG_KEGG_pathway_scope_counts.png"),
                scope_total_plot, width = 14, height = 6, dpi = 300)
ggplot2::ggsave(file.path(out_dir, "eggNOG_KEGG_pathway_scope_by_theme.png"),
                scope_theme_plot, width = 14, height = 8, dpi = 300)

print(scope_total_plot)
print(scope_theme_plot)
```

## Collect current infection DEG sets

The enrichment foregrounds are built from main-chromosome infected-versus-control
DEGs without
splitting genes by direction. This keeps the infection-response question simple:
which functional categories are over-represented among infection-responsive
genes in general and within each diet. All sets come from the same newest
competitive main-chromosome DESeq2 result file.

```{r deg-sets}
infection_contrasts <- c(
  "Diet 33" = "infection_diet33",
  "Diet 50" = "infection_diet50",
  "Diet 83" = "infection_diet83"
)

infection_sets_by_diet <- lapply(infection_contrasts, function(contrast_name) {
  placed_results |>
    dplyr::filter(.data$contrast_id == contrast_name, .data$significant) |>
    dplyr::pull(.data$gene_id) |>
    unique()
})

if (length(infection_sets_by_diet) == 0) {
  stop("No main-chromosome within-diet infection DEG sets were found.")
}

all_infection_genes <- sort(unique(unlist(infection_sets_by_diet, use.names = FALSE)))

enrichment_sets <- c(
  list("All infection DEGs" = all_infection_genes),
  infection_sets_by_diet
)

overview_set_label_levels <- stringr::str_replace_all(names(enrichment_sets),
                                                       "_", " ")

deg_set_summary <- data.frame(
  DEG_set = names(enrichment_sets),
  n_genes = lengths(enrichment_sets),
  stringsAsFactors = FALSE
)
deg_set_summary <- deg_set_summary[order(factor(
  deg_set_summary$DEG_set,
  levels = names(enrichment_sets)
)), ]

source_table <- data.frame(
  DEG_set = names(infection_sets_by_diet),
  contrast_id = unname(infection_contrasts),
  result_file = primary_results_file,
  modified = file.info(primary_results_file)$mtime,
  stringsAsFactors = FALSE
)

infection_deg_long <- placed_results |>
  dplyr::filter(.data$contrast_id %in% unname(infection_contrasts), .data$significant) |>
  dplyr::mutate(
    diet = stringr::str_remove(.data$contrast_id, "infection_diet"),
    DEG_set = paste("Diet", .data$diet)
  ) |>
  dplyr::select("gene_id", "diet", "DEG_set", "log2FoldChange", "padj", "significant")

DT::datatable(deg_set_summary, options = list(pageLength = 15))
write.csv(deg_set_summary, file.path(out_dir, "go_enrichment_deg_set_summary.csv"),
          row.names = FALSE)
write.csv(source_table, file.path(out_dir, "infection_deg_source_files.csv"),
          row.names = FALSE)
write.csv(infection_deg_long, file.path(out_dir, "infection_DEGs_long.csv"),
          row.names = FALSE)
```

## Collect requested direction-aware DEG sets

These foregrounds match the extra comparisons requested for discussion. The
requested figures use one column per contrast and use color/shape to show which
side of the contrast carries the enriched genes. For infection contrasts, red
means higher in infected samples and blue means higher in controls. For diet
contrasts, red means higher in the first diet named in the column and blue means
higher in the second diet. For example, in a `33 vs 83` column, blue points are
the genes higher in diet 83.

The diet-only contrasts use the same competitive main-chromosome DESeq2 catalogue
as the within-diet infection contrasts. They are kept separate from the
non-directional infection overview above because they answer a different
question: whether functional categories are enriched in one side of a contrast.

```{r directional-deg-sets}
make_directional_record <- function(set_id, family, comparison, direction,
                                    plot_label, genes, source_file) {
  data.frame(
    set_id = set_id,
    family = family,
    comparison = comparison,
    direction = direction,
    plot_label = plot_label,
    n_genes = length(genes),
    source_file = source_file,
    stringsAsFactors = FALSE
  )
}

directional_sets <- list()
directional_records <- list()

add_directional_sets <- function(contrast_key, set_prefix, family,
                                 comparison, plot_label) {
  tab <- placed_results |>
    dplyr::filter(.data$contrast_id == .env$contrast_key, .data$significant)
  up_genes <- unique(tab$gene_id[tab$log2FoldChange > 0])
  down_genes <- unique(tab$gene_id[tab$log2FoldChange < 0])

  list(
    sets = list(
      stats::setNames(list(up_genes), paste0(set_prefix, "_up")),
      stats::setNames(list(down_genes), paste0(set_prefix, "_down"))
    ) |>
      unlist(recursive = FALSE),
    records = list(
      make_directional_record(
        paste0(set_prefix, "_up"), family, comparison, "Upregulated",
        plot_label, up_genes, primary_results_file
      ),
      make_directional_record(
        paste0(set_prefix, "_down"), family, comparison, "Downregulated",
        plot_label, down_genes, primary_results_file
      )
    )
  )
}

directional_contrasts <- tibble::tribble(
  ~contrast_id, ~set_prefix, ~family, ~comparison, ~plot_label,
  "infection_diet33", "infection_diet33", "Infected within diet", "Diet 33 infected vs control", "Diet 33",
  "infection_diet50", "infection_diet50", "Infected within diet", "Diet 50 infected vs control", "Diet 50",
  "infection_diet83", "infection_diet83", "Infected within diet", "Diet 83 infected vs control", "Diet 83",
  "control_diet33_vs50", "control_diet33_vs_50", "Control animals only", "Control diet 33 vs 50", "33 vs 50",
  "control_diet83_vs50", "control_diet83_vs_50", "Control animals only", "Control diet 83 vs 50", "83 vs 50",
  "control_diet83_vs33", "control_diet83_vs_33", "Control animals only", "Control diet 83 vs 33", "83 vs 33",
  "infected_diet33_vs50", "infected_diet33_vs_50", "Infected animals only", "Infected diet 33 vs 50", "33 vs 50",
  "infected_diet83_vs50", "infected_diet83_vs_50", "Infected animals only", "Infected diet 83 vs 50", "83 vs 50",
  "infected_diet83_vs33", "infected_diet83_vs_33", "Infected animals only", "Infected diet 83 vs 33", "83 vs 33"
)

for (i in seq_len(nrow(directional_contrasts))) {
  added <- add_directional_sets(
    contrast_key = directional_contrasts$contrast_id[[i]],
    set_prefix = directional_contrasts$set_prefix[[i]],
    family = directional_contrasts$family[[i]],
    comparison = directional_contrasts$comparison[[i]],
    plot_label = directional_contrasts$plot_label[[i]]
  )
  directional_sets <- c(directional_sets, added$sets)
  directional_records <- c(directional_records, added$records)
}

directional_set_summary <- dplyr::bind_rows(directional_records) |>
  dplyr::mutate(
    family = factor(
      family,
      levels = c("Infected within diet", "Control animals only",
                 "Infected animals only")
    ),
    direction = factor(direction, levels = c("Upregulated", "Downregulated"))
  ) |>
  dplyr::arrange(family, comparison, direction)

directional_sets <- directional_sets[directional_set_summary$set_id]

DT::datatable(
  directional_set_summary |>
    dplyr::select(family, comparison, direction, plot_label, n_genes),
  rownames = FALSE,
  options = list(pageLength = 18, scrollX = TRUE)
)

write.csv(directional_set_summary,
          file.path(out_dir, "requested_directional_deg_set_summary.csv"),
          row.names = FALSE)
write.csv(
  data.frame(
    set_id = rep(names(directional_sets), lengths(directional_sets)),
    gene_id = unlist(directional_sets, use.names = FALSE)
  ),
  file.path(out_dir, "requested_directional_DEG_sets_long.csv"),
  row.names = FALSE
)
```

## Fat body immune and storage candidates

The fat body is central to immune defense, detoxification, lipid transport, and
nutrient storage. This table therefore flags significant infection-responsive
genes whose eggNOG annotation contains immune, storage-protein, lipid-transport,
endocrine-nutrition, or detoxification language. This is a transparent keyword
screen to help prioritize genes for biological interpretation; it does not
replace the GO or KEGG enrichment tests.

```{r fat-body-candidates}
classify_fat_body_candidate <- function(text) {
  text <- stringr::str_to_lower(text)
  immune <- stringr::str_detect(
    text,
    "immune|defen[cs]e|defensin|attacin|cecropin|gloverin|lysozyme|toll|imd|tnf|jak-stat|peptidoglycan|antimicrobial|phenoloxidase|prophenoloxidase|serpin|thioester|transferrin|lectin|phagosome"
  )
  storage <- stringr::str_detect(
    text,
    "storage protein|hexamerin|arylphorin|hemocyanin|vitellogenin|lipophorin|apolipophorin|nutrient reservoir"
  )
  detox <- stringr::str_detect(
    text,
    "detox|glutathione|cytochrome p450|p450|xenobiotic|oxidative stress|peroxidase|superoxide|catalase"
  )
  endocrine_nutrition <- stringr::str_detect(
    text,
    "juvenile hormone|insulin|glucagon|adipocytokine|ppar|bile secretion|mineral absorption|fat digestion|glycerolipid|amino acid|lipid transport"
  )

  dplyr::case_when(
    immune & storage ~ "Immune + storage/lipid transport",
    storage ~ "Storage / lipid transport",
    immune ~ "Immune / defense",
    detox ~ "Detoxification / oxidative stress",
    endocrine_nutrition ~ "Endocrine / nutrient metabolism",
    TRUE ~ NA_character_
  )
}

fat_body_candidates <- infection_deg_long |>
  dplyr::inner_join(
    annotation |>
      dplyr::select(gene_id, Preferred_name, Description, GOs, KEGG_ko,
                    KEGG_Pathway, BRITE, PFAMs),
    by = "gene_id"
  ) |>
  dplyr::mutate(
    annotation_text = paste(Preferred_name, Description, GOs, KEGG_ko,
                            KEGG_Pathway, BRITE, PFAMs, sep = " "),
    candidate_theme = classify_fat_body_candidate(annotation_text)
  ) |>
  dplyr::filter(!is.na(candidate_theme)) |>
  dplyr::arrange(candidate_theme, padj) |>
  dplyr::group_by(gene_id, candidate_theme, Preferred_name, Description, PFAMs) |>
  dplyr::summarise(
    DEG_sets = paste(unique(DEG_set), collapse = "; "),
    log2FC_by_diet = paste0(DEG_set, "=", round(log2FoldChange, 2),
                            collapse = "; "),
    min_padj = signif(min(padj, na.rm = TRUE), 3),
    max_abs_log2FC = round(max(abs(log2FoldChange), na.rm = TRUE), 2),
    KEGG_Pathway = paste(unique(KEGG_Pathway), collapse = "; "),
    .groups = "drop"
  ) |>
  dplyr::arrange(candidate_theme, min_padj)

write.csv(fat_body_candidates,
          file.path(out_dir, "fat_body_immune_storage_DEG_candidates.csv"),
          row.names = FALSE)

if (nrow(fat_body_candidates) > 0) {
  DT::datatable(
    fat_body_candidates,
    rownames = FALSE,
    options = list(pageLength = 15, scrollX = TRUE)
  )
} else {
  cat("No significant infection DEGs matched the current fat-body candidate keyword screen.")
}
```

## Run GO and KEGG enrichment

Enrichment tests whether a DEG set contains more genes from a term than expected
from the annotated background. Here the DEG sets are not separated by
upregulation or downregulation; each foreground contains significant
infected-versus-control DEGs regardless of fold-change direction.

```{r run-enrichment}
run_enrichment <- function(genes, term2gene, term2name, universe) {
  genes <- unique(genes[genes %in% universe])
  if (length(genes) < 5) return(NULL)
  clusterProfiler::enricher(
    gene = genes,
    universe = universe,
    TERM2GENE = term2gene,
    TERM2NAME = term2name[, c("term", "name")],
    pvalueCutoff = 0.1,
    pAdjustMethod = "BH",
    qvalueCutoff = 0.2
  )
}

universe_go <- unique(term2gene_go$gene)
universe_kegg <- unique(term2gene_kegg_pathway$gene)

go_results <- lapply(enrichment_sets, run_enrichment,
                     term2gene = term2gene_go,
                     term2name = term2name_go,
                     universe = universe_go)
kegg_results <- lapply(enrichment_sets, run_enrichment,
                       term2gene = term2gene_kegg_pathway,
                       term2name = term2name_kegg_pathway,
                       universe = universe_kegg)
directional_go_results <- lapply(directional_sets, run_enrichment,
                                 term2gene = term2gene_go,
                                 term2name = term2name_go,
                                 universe = universe_go)
directional_kegg_results <- lapply(directional_sets, run_enrichment,
                                   term2gene = term2gene_kegg_pathway,
                                   term2name = term2name_kegg_pathway,
                                   universe = universe_kegg)

enrich_to_df <- function(result_list, database) {
  pieces <- lapply(names(result_list), function(nm) {
    res <- result_list[[nm]]
    if (is.null(res)) return(NULL)
    df <- as.data.frame(res)
    if (nrow(df) == 0) return(NULL)
    df$set <- nm
    df$database <- database
    df
  })
  dplyr::bind_rows(pieces)
}

go_enrichment_df <- enrich_to_df(go_results, "GO")
kegg_enrichment_df <- enrich_to_df(kegg_results, "KEGG_pathway")
directional_go_enrichment_df <- enrich_to_df(directional_go_results, "GO")
directional_kegg_enrichment_df <- enrich_to_df(directional_kegg_results,
                                               "KEGG_pathway")

if (nrow(kegg_enrichment_df) > 0) {
  kegg_enrichment_df <- kegg_enrichment_df |>
    dplyr::left_join(
      term2name_kegg_pathway |>
        dplyr::select(term, pathway_theme, pathway_scope),
      by = c("ID" = "term")
    )
}
if (nrow(directional_kegg_enrichment_df) > 0) {
  directional_kegg_enrichment_df <- directional_kegg_enrichment_df |>
    dplyr::left_join(
      term2name_kegg_pathway |>
        dplyr::select(term, pathway_theme, pathway_scope),
      by = c("ID" = "term")
    )
}

all_enrichment <- dplyr::bind_rows(go_enrichment_df, kegg_enrichment_df)

if (nrow(directional_go_enrichment_df) > 0) {
  directional_go_enrichment_df <- directional_go_enrichment_df |>
    dplyr::left_join(directional_set_summary, by = c("set" = "set_id"))
}
if (nrow(directional_kegg_enrichment_df) > 0) {
  directional_kegg_enrichment_df <- directional_kegg_enrichment_df |>
    dplyr::left_join(directional_set_summary, by = c("set" = "set_id"))
}
directional_all_enrichment <- dplyr::bind_rows(
  directional_go_enrichment_df,
  directional_kegg_enrichment_df
)

write.csv(go_enrichment_df, file.path(out_dir, "GO_enrichment_all_sets.csv"),
          row.names = FALSE)
write.csv(kegg_enrichment_df,
          file.path(out_dir, "KEGG_pathway_enrichment_all_sets.csv"),
          row.names = FALSE)
write.csv(all_enrichment, file.path(out_dir, "all_enrichment_results.csv"),
          row.names = FALSE)
write.csv(directional_go_enrichment_df,
          file.path(out_dir, "requested_directional_GO_enrichment.csv"),
          row.names = FALSE)
write.csv(directional_kegg_enrichment_df,
          file.path(out_dir, "requested_directional_KEGG_pathway_enrichment.csv"),
          row.names = FALSE)
write.csv(directional_all_enrichment,
          file.path(out_dir, "requested_directional_all_enrichment.csv"),
          row.names = FALSE)

if (nrow(all_enrichment) == 0) {
  cat("No enriched terms passed the current cutoffs.")
} else {
  DT::datatable(
    all_enrichment |>
      dplyr::select(dplyr::any_of(c(
        "database", "set", "ID", "Description", "pathway_theme",
        "pathway_scope", "GeneRatio", "BgRatio", "p.adjust", "qvalue",
        "geneID"
      ))) |>
      dplyr::arrange(p.adjust),
    options = list(pageLength = 15, scrollX = TRUE)
  )
}

if (nrow(directional_all_enrichment) == 0) {
  cat("No requested direction-aware sets produced enriched terms at the current cutoffs.")
} else {
  DT::datatable(
    directional_all_enrichment |>
      dplyr::select(dplyr::any_of(c(
        "database", "family", "comparison", "direction", "plot_label",
        "ID", "Description", "pathway_theme", "pathway_scope",
        "GeneRatio", "BgRatio", "p.adjust", "qvalue", "geneID"
      ))) |>
      dplyr::arrange(family, comparison, direction, p.adjust),
    rownames = FALSE,
    options = list(pageLength = 18, scrollX = TRUE)
  )
}
```

## KEGG pathway scope among enriched terms

These figures show the same scope categories after enrichment. A pathway is
counted once per DEG set where it is significantly enriched, so the bars
summarize enriched KEGG pathway labels rather than the number of genes inside
each pathway.

```{r kegg-pathway-scope-enriched, fig.width=16, fig.height=10}
if (nrow(kegg_enrichment_df) > 0) {
  kegg_scope_enrichment_summary <- kegg_enrichment_df |>
    dplyr::mutate(
      set_label = stringr::str_replace_all(set, "_", " "),
      pathway_scope = dplyr::coalesce(as.character(pathway_scope),
                                      "Unclassified KEGG reference label")
    ) |>
    dplyr::distinct(set_label, ID, pathway_scope) |>
    dplyr::count(set_label, pathway_scope, name = "n_enriched_pathways") |>
    dplyr::mutate(
      set_label = factor(set_label, levels = overview_set_label_levels),
      pathway_scope = factor(pathway_scope, levels = scope_levels)
    )
} else {
  kegg_scope_enrichment_summary <- tibble::tibble(
    set_label = factor(character(), levels = overview_set_label_levels),
    pathway_scope = factor(character(), levels = scope_levels),
    n_enriched_pathways = integer()
  )
}

kegg_scope_enrichment_summary <- kegg_scope_enrichment_summary |>
  tidyr::complete(set_label, pathway_scope,
                  fill = list(n_enriched_pathways = 0))

write.csv(kegg_scope_enrichment_summary,
          file.path(out_dir, "KEGG_pathway_scope_enrichment_summary.csv"),
          row.names = FALSE)

if (nrow(kegg_scope_enrichment_summary) > 0) {
  kegg_scope_matrix_df <- dplyr::bind_rows(
    scope_total_df |>
      dplyr::mutate(
        pathway_scope = factor(as.character(pathway_scope),
                               levels = scope_levels),
        context = "Annotation map",
        n_pathways = as.integer(n_pathways)
      ) |>
      dplyr::select(context, pathway_scope, n_pathways),
    kegg_scope_enrichment_summary |>
      dplyr::group_by(pathway_scope) |>
      dplyr::summarise(n_pathways = sum(n_enriched_pathways),
                       .groups = "drop") |>
      dplyr::mutate(context = "Non-directional enrichments") |>
      dplyr::select(context, pathway_scope, n_pathways)
  ) |>
    tidyr::complete(
      context = c("Annotation map", "Non-directional enrichments"),
      pathway_scope = factor(scope_levels, levels = scope_levels),
      fill = list(n_pathways = 0)
    ) |>
    dplyr::mutate(
      context = factor(context,
                       levels = c("Annotation map",
                                  "Non-directional enrichments")),
      pathway_scope = factor(pathway_scope, levels = rev(scope_levels))
    )

  kegg_scope_matrix_plot <- kegg_scope_matrix_df |>
    ggplot2::ggplot(ggplot2::aes(x = context, y = pathway_scope,
                                 fill = pathway_scope)) +
    ggplot2::geom_tile(color = "white", linewidth = 0.7, width = 0.92,
                       height = 0.92) +
    ggplot2::geom_text(ggplot2::aes(label = n_pathways), size = 5,
                       fontface = "bold") +
    ggplot2::scale_fill_manual(values = kegg_scope_colors, limits = scope_levels,
                               breaks = scope_levels, drop = FALSE) +
    ggplot2::labs(
      title = "KEGG pathway scope color key and counts",
      subtitle = "Zero means that scope category is absent from the corresponding pathway set at the current cutoff",
      x = NULL,
      y = NULL,
      fill = "Pathway scope"
    ) +
    ggplot2::theme_minimal(base_size = 15) +
    ggplot2::theme(
      legend.position = "bottom",
      legend.title = ggplot2::element_text(size = 13),
      legend.text = ggplot2::element_text(size = 12),
      panel.grid = ggplot2::element_blank(),
      axis.text.x = ggplot2::element_text(size = 13, face = "bold"),
      axis.text.y = ggplot2::element_text(size = 12)
    )

  kegg_scope_enrichment_plot <- kegg_scope_enrichment_summary |>
    ggplot2::ggplot(ggplot2::aes(x = set_label, y = n_enriched_pathways,
                                 fill = pathway_scope)) +
    ggplot2::geom_col(width = 0.72, color = "white", linewidth = 0.25) +
    ggplot2::coord_flip() +
    ggplot2::scale_fill_manual(values = kegg_scope_colors, limits = scope_levels,
                               breaks = scope_levels, drop = FALSE) +
    ggplot2::labs(
      title = "KEGG pathway scope among enriched infection DEG sets",
      subtitle = "Counts are unique enriched KEGG pathway IDs per non-directional DEG set",
      x = NULL,
      y = "Enriched KEGG pathway IDs",
      fill = "Pathway scope"
    ) +
    ggplot2::theme_minimal(base_size = 15) +
    ggplot2::theme(
      legend.position = "bottom",
      legend.title = ggplot2::element_text(size = 13),
      legend.text = ggplot2::element_text(size = 12),
      panel.grid.major.y = ggplot2::element_blank(),
      axis.text.y = ggplot2::element_text(size = 13)
    )

  ggplot2::ggsave(
    file.path(out_dir, "KEGG_pathway_scope_color_key_counts.png"),
    kegg_scope_matrix_plot, width = 16, height = 6, dpi = 300
  )
  ggplot2::ggsave(
    file.path(out_dir, "KEGG_pathway_scope_enrichment_summary.png"),
    kegg_scope_enrichment_plot, width = 16, height = 7, dpi = 300
  )
  print(kegg_scope_matrix_plot)
  print(kegg_scope_enrichment_plot)
} else {
  cat("No non-directional KEGG enrichment rows were available for scope plotting.")
}
```

```{r requested-kegg-pathway-scope-enriched, fig.width=18, fig.height=12}
if (nrow(directional_kegg_enrichment_df) > 0) {
  requested_kegg_scope_counts <- directional_kegg_enrichment_df |>
    dplyr::mutate(
      pathway_scope = dplyr::coalesce(as.character(pathway_scope),
                                      "Unclassified KEGG reference label"),
      plot_label = stringr::str_wrap(plot_label, width = 28)
    ) |>
    dplyr::distinct(family, plot_label, direction, ID, pathway_scope) |>
    dplyr::count(family, plot_label, direction, pathway_scope,
                 name = "n_enriched_pathways")
  requested_kegg_scope_keys <- directional_set_summary |>
    dplyr::mutate(
      plot_label = stringr::str_wrap(plot_label, width = 28),
      direction = factor(direction, levels = c("Upregulated", "Downregulated"))
    ) |>
    dplyr::distinct(family, plot_label, direction)
  requested_kegg_scope_summary <- requested_kegg_scope_keys |>
    tidyr::crossing(pathway_scope = factor(scope_levels, levels = scope_levels)) |>
    dplyr::left_join(
      requested_kegg_scope_counts |>
        dplyr::mutate(
          direction = factor(direction,
                             levels = c("Upregulated", "Downregulated")),
          pathway_scope = factor(pathway_scope, levels = scope_levels)
        ),
      by = c("family", "plot_label", "direction", "pathway_scope")
    ) |>
    dplyr::mutate(
      n_enriched_pathways = dplyr::coalesce(n_enriched_pathways, 0L)
    )
} else {
  requested_kegg_scope_summary <- tibble::tibble(
    family = character(),
    plot_label = character(),
    direction = factor(character(),
                       levels = c("Upregulated", "Downregulated")),
    pathway_scope = factor(character(), levels = scope_levels),
    n_enriched_pathways = integer()
  )
}

write.csv(requested_kegg_scope_summary,
          file.path(out_dir, "requested_KEGG_pathway_scope_enrichment_summary.csv"),
          row.names = FALSE)

if (nrow(requested_kegg_scope_summary) > 0) {
  requested_kegg_scope_matrix_df <- requested_kegg_scope_summary |>
    dplyr::group_by(pathway_scope) |>
    dplyr::summarise(n_pathways = sum(n_enriched_pathways),
                     .groups = "drop") |>
    dplyr::mutate(
      context = "Direction-aware enrichments",
      pathway_scope = factor(pathway_scope, levels = rev(scope_levels))
    ) |>
    tidyr::complete(
      context = "Direction-aware enrichments",
      pathway_scope = factor(rev(scope_levels), levels = rev(scope_levels)),
      fill = list(n_pathways = 0)
    )

  requested_kegg_scope_matrix_plot <- requested_kegg_scope_matrix_df |>
    ggplot2::ggplot(ggplot2::aes(x = context, y = pathway_scope,
                                 fill = pathway_scope)) +
    ggplot2::geom_tile(color = "white", linewidth = 0.7, width = 0.92,
                       height = 0.92) +
    ggplot2::geom_text(ggplot2::aes(label = n_pathways), size = 5,
                       fontface = "bold") +
    ggplot2::scale_fill_manual(values = kegg_scope_colors, limits = scope_levels,
                               breaks = scope_levels, drop = FALSE) +
    ggplot2::labs(
      title = "KEGG pathway scope counts in requested direction-aware enrichments",
      subtitle = "Zero means that scope category is not enriched in any requested contrast at the current cutoff",
      x = NULL,
      y = NULL,
      fill = "Pathway scope"
    ) +
    ggplot2::theme_minimal(base_size = 15) +
    ggplot2::theme(
      legend.position = "bottom",
      legend.title = ggplot2::element_text(size = 13),
      legend.text = ggplot2::element_text(size = 12),
      panel.grid = ggplot2::element_blank(),
      axis.text.x = ggplot2::element_text(size = 13, face = "bold"),
      axis.text.y = ggplot2::element_text(size = 12)
    )

  requested_kegg_scope_plot <- requested_kegg_scope_summary |>
    ggplot2::ggplot(ggplot2::aes(x = plot_label, y = n_enriched_pathways,
                                 fill = pathway_scope)) +
    ggplot2::geom_col(width = 0.72, color = "white", linewidth = 0.25) +
    ggplot2::coord_flip() +
    ggplot2::facet_grid(direction ~ family, scales = "free_y",
                        space = "free_y") +
    ggplot2::scale_fill_manual(values = kegg_scope_colors, limits = scope_levels,
                               breaks = scope_levels, drop = FALSE) +
    ggplot2::labs(
      title = "KEGG pathway scope among requested direction-aware enrichments",
      subtitle = "Counts are unique enriched KEGG pathway IDs per contrast and direction",
      x = NULL,
      y = "Enriched KEGG pathway IDs",
      fill = "Pathway scope"
    ) +
    ggplot2::theme_minimal(base_size = 15) +
    ggplot2::theme(
      legend.position = "bottom",
      legend.title = ggplot2::element_text(size = 13),
      legend.text = ggplot2::element_text(size = 12),
      strip.text = ggplot2::element_text(size = 13, face = "bold"),
      panel.grid.major.y = ggplot2::element_blank(),
      axis.text.y = ggplot2::element_text(size = 12)
    )

  ggplot2::ggsave(
    file.path(out_dir, "requested_KEGG_pathway_scope_color_key_counts.png"),
    requested_kegg_scope_matrix_plot, width = 14, height = 6, dpi = 300
  )
  ggplot2::ggsave(
    file.path(out_dir, "requested_KEGG_pathway_scope_enrichment_summary.png"),
    requested_kegg_scope_plot, width = 18, height = 12, dpi = 300
  )
  print(requested_kegg_scope_matrix_plot)
  print(requested_kegg_scope_plot)
} else {
  cat("No requested direction-aware KEGG enrichment rows were available for scope plotting.")
}
```

## GO terms by ontology

This figure separates biological process, molecular function, and cellular
component terms. Each GO term is shown once per ontology row, so terms shared
by multiple DEG sets align across columns instead of being repeated as separate
y-axis rows. A few eggNOG GO IDs can be obsolete or unavailable in the local
`GO.db` ontology table; those terms are kept in the complete enrichment CSV but
omitted from this ontology-specific plot. The dashed vertical line marks FDR =
0.05. Ontology is shown by the right-side row label and colored panel frames so
the GO terms remain aligned across DEG-set columns.

```{r go-ontology-figure, fig.height=22, fig.width=30}
if (nrow(go_enrichment_df) > 0 && "ID" %in% names(go_enrichment_df)) {
  go_enrichment_ontology <- go_enrichment_df |>
    dplyr::left_join(term2name_go |> dplyr::select(term, ontology),
                     by = c("ID" = "term")) |>
    dplyr::mutate(
      ontology = dplyr::case_when(
        ontology == "BP" ~ "Biological process",
        ontology == "MF" ~ "Molecular function",
        ontology == "CC" ~ "Cellular component",
        TRUE ~ NA_character_
      ),
      neg_log10_padj = -log10(pmax(p.adjust, .Machine$double.xmin)),
      term_label = stringr::str_wrap(paste0(Description, " [", ID, "]"),
                                     width = 80),
      set_label = factor(stringr::str_replace_all(set, "_", " "),
                         levels = overview_set_label_levels)
    )
} else {
  go_enrichment_ontology <- data.frame()
}

write.csv(go_enrichment_ontology,
          file.path(out_dir, "eggNOG_GO_enrichment_with_ontology.csv"),
          row.names = FALSE)

unmapped_go_terms <- go_enrichment_ontology |>
  dplyr::filter(is.na(ontology)) |>
  dplyr::distinct(ID, Description, set, p.adjust, Count)
write.csv(unmapped_go_terms,
          file.path(out_dir, "eggNOG_GO_enrichment_unmapped_ontology.csv"),
          row.names = FALSE)

if (nrow(unmapped_go_terms) > 0) {
  cat(nrow(unmapped_go_terms),
      "enriched GO rows did not map to BP, MF, or CC in the local GO.db and are omitted from the ontology plot.\n")
}

if (nrow(go_enrichment_ontology) > 0) {
  go_plot_df <- go_enrichment_ontology |>
    dplyr::filter(!is.na(p.adjust), !is.na(ontology)) |>
    dplyr::group_by(set_label, ontology) |>
    dplyr::arrange(p.adjust, dplyr::desc(Count), .by_group = TRUE) |>
    dplyr::slice_head(n = params$top_terms) |>
    dplyr::ungroup() |>
    dplyr::arrange(set_label, ontology, p.adjust, dplyr::desc(Count)) |>
    dplyr::mutate(
      term_label_facet = paste(term_label, ontology, sep = "___")
    )

  go_term_levels <- go_plot_df |>
    dplyr::group_by(ontology, term_label_facet) |>
    dplyr::summarise(
      best_p = min(p.adjust, na.rm = TRUE),
      max_count = max(Count, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::arrange(ontology, best_p, dplyr::desc(max_count)) |>
    dplyr::pull(term_label_facet)

  go_plot_df <- go_plot_df |>
    dplyr::mutate(
      term_label_facet = factor(term_label_facet,
                                levels = rev(unique(go_term_levels)))
    )
  go_frame_df <- go_plot_df |>
    dplyr::distinct(ontology)
  go_missing_set_labels <- setdiff(
    overview_set_label_levels,
    as.character(stats::na.omit(unique(go_plot_df$set_label)))
  )
  go_column_dummy <- data.frame(
    set_label = factor(character(), levels = overview_set_label_levels),
    ontology = character(),
    term_label_facet = factor(character(),
                              levels = levels(go_plot_df$term_label_facet)),
    neg_log10_padj = numeric(),
    Count = numeric()
  )
  if (length(go_missing_set_labels) > 0) {
    go_column_dummy <- data.frame(
      set_label = factor(go_missing_set_labels,
                         levels = overview_set_label_levels),
      ontology = as.character(go_plot_df$ontology[1]),
      term_label_facet = factor(as.character(go_plot_df$term_label_facet[1]),
                                levels = levels(go_plot_df$term_label_facet)),
      neg_log10_padj = 0,
      Count = min(go_plot_df$Count, na.rm = TRUE)
    )
  }

  p_go_ontology <- ggplot2::ggplot(
    go_plot_df,
    ggplot2::aes(y = term_label_facet)
  ) +
    ggplot2::geom_blank(
      data = go_column_dummy,
      ggplot2::aes(x = neg_log10_padj, y = term_label_facet),
      inherit.aes = FALSE
    ) +
    ggplot2::geom_rect(
      data = go_frame_df,
      ggplot2::aes(color = ontology),
      xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf,
      fill = NA, linewidth = 0.95, inherit.aes = FALSE
    ) +
    ggplot2::geom_vline(xintercept = -log10(0.05),
                        linetype = "dashed", color = "#374151",
                        linewidth = 0.35) +
    ggplot2::geom_point(ggplot2::aes(x = neg_log10_padj, size = Count,
                                     fill = ontology),
                        shape = 21, color = "#1f2937",
                        alpha = 0.9,
                        stroke = 0.35) +
    ggplot2::facet_grid(ontology ~ set_label, scales = "free_y",
                        space = "free_y") +
    ggplot2::scale_color_manual(values = ontology_colors, drop = FALSE) +
    ggplot2::scale_fill_manual(values = ontology_colors, drop = FALSE) +
    ggplot2::scale_y_discrete(labels = function(x) sub("___.*$", "", x)) +
    ggplot2::scale_x_continuous(
      limits = c(0, NA),
      expand = ggplot2::expansion(mult = c(0.02, 0.05))
    ) +
    ggplot2::scale_size(range = c(4.5, 13)) +
    ggplot2::guides(
      color = "none",
      fill = ggplot2::guide_legend(nrow = 1, byrow = TRUE, order = 1),
      size = ggplot2::guide_legend(order = 2)
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      strip.background = ggplot2::element_rect(fill = "#F1F4F3",
                                               color = "#C8D0CC"),
      strip.text.x = ggplot2::element_text(size = 18, face = "bold"),
      strip.text.y = ggplot2::element_text(size = 18, face = "bold",
                                           angle = 90),
      axis.text.y = ggplot2::element_text(size = 18, lineheight = 0.95,
                                          hjust = 1),
      panel.border = ggplot2::element_blank(),
      legend.position = "bottom",
      legend.box = "vertical",
      legend.title = ggplot2::element_text(size = 12, face = "bold"),
      legend.text = ggplot2::element_text(size = 11),
      plot.margin = ggplot2::margin(10, 16, 10, 16)
    ) +
    ggplot2::labs(
      title = "eggNOG GO enrichment by ontology",
      subtitle = "Top GO terms per DEG set and ontology",
      x = "-log10 adjusted p-value",
      y = NULL,
      fill = "GO ontology",
      size = "DEGs in term"
    )

  print(p_go_ontology)
  ggplot2::ggsave(file.path(out_dir, "eggNOG_GO_enrichment_by_ontology.png"),
                  p_go_ontology, width = 30, height = 22, dpi = 300)
}
```

## Directional GO terms by ontology

These panels show the requested direction-aware foregrounds. The same annotated
background is used as above. Empty facets mean that no term passed the
enrichment cutoffs for that particular contrast. Colored panel frames mark GO
ontology, while the enrichment dots mark the side of the contrast carrying the
enriched genes.

```{r requested-go-ontology-setup}
if (nrow(directional_go_enrichment_df) > 0 && "ID" %in% names(directional_go_enrichment_df)) {
  directional_go_enrichment_ontology <- directional_go_enrichment_df |>
    dplyr::left_join(term2name_go |> dplyr::select(term, ontology),
                     by = c("ID" = "term")) |>
    dplyr::mutate(
      ontology = dplyr::case_when(
        ontology == "BP" ~ "Biological process",
        ontology == "MF" ~ "Molecular function",
        ontology == "CC" ~ "Cellular component",
        TRUE ~ NA_character_
      ),
      family = factor(
        as.character(family),
        levels = c("Infected within diet", "Control animals only",
                   "Infected animals only")
      ),
      direction = factor(as.character(direction),
                         levels = c("Upregulated", "Downregulated")),
      neg_log10_padj = -log10(pmax(p.adjust, .Machine$double.xmin)),
      term_label = stringr::str_wrap(paste0(Description, " [", ID, "]"),
                                     width = 80)
    )
} else {
  directional_go_enrichment_ontology <- data.frame()
}

write.csv(directional_go_enrichment_ontology,
          file.path(out_dir, "requested_directional_GO_enrichment_with_ontology.csv"),
          row.names = FALSE)

plot_directional_go_family <- function(family_name) {
  family_df <- directional_go_enrichment_ontology |>
    dplyr::filter(family == family_name, !is.na(ontology))
  if (nrow(family_df) == 0) {
    cat("No requested GO enrichment terms to plot for ", family_name, ".\n",
        sep = "")
    return(NULL)
  }

  plot_order <- directional_set_summary |>
    dplyr::filter(family == family_name) |>
    dplyr::pull(plot_label) |>
    unique()

  plot_df <- family_df |>
    dplyr::group_by(plot_label, direction, ontology) |>
    dplyr::arrange(p.adjust, dplyr::desc(Count), .by_group = TRUE) |>
    dplyr::slice_head(n = min(params$top_terms, 5)) |>
    dplyr::ungroup() |>
    dplyr::arrange(plot_label, direction, ontology, p.adjust,
                   dplyr::desc(Count)) |>
    dplyr::mutate(
      plot_label = factor(plot_label, levels = plot_order),
      term_label_facet = paste(term_label, ontology, sep = "___")
    )

  go_term_levels <- plot_df |>
    dplyr::group_by(ontology, term_label_facet) |>
    dplyr::summarise(
      best_p = min(p.adjust, na.rm = TRUE),
      max_count = max(Count, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::arrange(ontology, best_p, dplyr::desc(max_count)) |>
    dplyr::pull(term_label_facet)

  plot_df <- plot_df |>
    dplyr::mutate(
      term_label_facet = factor(term_label_facet,
                                levels = rev(unique(go_term_levels)))
    )
  go_frame_df <- plot_df |>
    dplyr::distinct(ontology)
  direction_legend_df <- data.frame(
    plot_label = factor(levels(plot_df$plot_label)[1],
                        levels = levels(plot_df$plot_label)),
    ontology = as.character(plot_df$ontology[1]),
    term_label_facet = plot_df$term_label_facet[1],
    neg_log10_padj = 0,
    Count = min(plot_df$Count, na.rm = TRUE),
    direction = factor(names(direction_colors), levels = names(direction_colors))
  )
  go_strip_x_size <- dplyr::case_when(
    family_name == "Infected within diet" ~ 18,
    family_name == "Control animals only" ~ 14,
    TRUE ~ 13
  )
  direction_legend_title <- if (family_name == "Infected within diet") {
    "Infection response"
  } else {
    "Diet side"
  }
  direction_legend_labels <- if (family_name == "Infected within diet") {
    c("Upregulated" = "Higher infected",
      "Downregulated" = "Higher control")
  } else {
    c("Upregulated" = "Higher first diet",
      "Downregulated" = "Higher second diet")
  }
  missing_plot_labels <- setdiff(
    as.character(plot_order),
    as.character(stats::na.omit(unique(plot_df$plot_label)))
  )
  no_terms_df <- data.frame(
    plot_label = factor(character(), levels = plot_order),
    ontology = character(),
    term_label_facet = factor(character(),
                              levels = levels(plot_df$term_label_facet)),
    label_x = numeric(),
    label = character()
  )
  if (length(missing_plot_labels) > 0) {
    label_x <- max(plot_df$neg_log10_padj, na.rm = TRUE) * 0.55
    if (!is.finite(label_x) || label_x <= 0) label_x <- 1
    go_y_by_row <- plot_df |>
      dplyr::group_by(ontology) |>
      dplyr::summarise(
        term_label_facet = dplyr::first(term_label_facet),
        .groups = "drop"
      )
    no_terms_df <- expand.grid(
      plot_label = missing_plot_labels,
      ontology = unique(as.character(go_y_by_row$ontology)),
      stringsAsFactors = FALSE
    ) |>
      dplyr::left_join(
        go_y_by_row |>
          dplyr::mutate(ontology = as.character(ontology)),
        by = "ontology"
      ) |>
      dplyr::mutate(
        plot_label = factor(plot_label, levels = plot_order),
        term_label_facet = factor(
          as.character(term_label_facet),
          levels = levels(plot_df$term_label_facet)
        ),
        label_x = label_x,
        label = "No enriched terms"
      )
  }

  p <- ggplot2::ggplot(
    plot_df,
    ggplot2::aes(y = term_label_facet)
  ) +
    ggplot2::geom_rect(
      data = go_frame_df,
      ggplot2::aes(color = ontology),
      xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf,
      fill = NA, linewidth = 0.95, inherit.aes = FALSE
    ) +
    ggplot2::geom_vline(xintercept = -log10(0.05),
                        linetype = "dashed", color = "#374151",
                        linewidth = 0.35) +
    ggplot2::geom_text(
      data = no_terms_df,
      ggplot2::aes(x = label_x, y = term_label_facet, label = label),
      inherit.aes = FALSE, color = "#6B7280", fontface = "italic",
      size = 5
    ) +
    ggplot2::geom_point(
      ggplot2::aes(x = neg_log10_padj, fill = direction,
                   shape = direction),
      color = "#1f2937", size = 4.5, alpha = 0.9, stroke = 0.35
    ) +
    ggplot2::geom_point(
      data = direction_legend_df,
      ggplot2::aes(x = neg_log10_padj, y = term_label_facet,
                   fill = direction, shape = direction),
      inherit.aes = FALSE, color = "#1f2937", size = 4.5,
      alpha = 0, show.legend = TRUE
    ) +
    ggplot2::facet_grid(ontology ~ plot_label, scales = "free_y",
                        space = "free_y") +
    ggplot2::scale_color_manual(values = ontology_colors, drop = FALSE) +
    ggplot2::scale_fill_manual(name = direction_legend_title,
                               values = direction_colors,
                               breaks = names(direction_colors),
                               labels = direction_legend_labels,
                               drop = FALSE) +
    ggplot2::scale_shape_manual(values = c("Upregulated" = 21,
                                           "Downregulated" = 24),
                                breaks = names(direction_colors),
                                name = direction_legend_title,
                                labels = direction_legend_labels,
                                drop = FALSE) +
    ggplot2::scale_y_discrete(labels = function(x) sub("___.*$", "", x)) +
    ggplot2::scale_x_continuous(
      limits = c(0, NA),
      expand = ggplot2::expansion(mult = c(0.02, 0.05))
    ) +
    ggplot2::guides(
      color = "none",
      fill = "none",
      shape = ggplot2::guide_legend(
        nrow = 1, byrow = TRUE, order = 2,
        override.aes = list(fill = unname(direction_colors), size = 4.5,
                            alpha = 1,
                            color = "#1f2937")
      )
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      strip.background = ggplot2::element_rect(fill = "#F1F4F3",
                                               color = "#C8D0CC"),
      strip.text.x = ggplot2::element_text(size = go_strip_x_size,
                                           face = "bold"),
      strip.text.y = ggplot2::element_text(size = 14, face = "bold",
                                           angle = 90),
      axis.text.y = ggplot2::element_text(size = 18, lineheight = 0.95,
                                          hjust = 1),
      panel.border = ggplot2::element_blank(),
      legend.position = "bottom",
      legend.box = "vertical",
      plot.margin = ggplot2::margin(10, 16, 10, 16)
    ) +
    ggplot2::labs(
      title = paste("eggNOG GO enrichment:", family_name),
      subtitle = "Directional DEG foregrounds; dashed line marks FDR = 0.05",
      x = "-log10 adjusted p-value",
      y = "GO term",
      color = "GO ontology",
      fill = direction_legend_title,
      shape = direction_legend_title
    )

  file_stub <- stringr::str_replace_all(stringr::str_to_lower(family_name),
                                        "[^a-z0-9]+", "_")
  figure_height <- dplyr::case_when(
    family_name == "Infected within diet" ~ 22,
    family_name == "Control animals only" ~ 10,
    TRUE ~ 18
  )
  figure_width <- dplyr::case_when(
    family_name == "Infected within diet" ~ 30,
    TRUE ~ 18
  )

  ggplot2::ggsave(
    file.path(out_dir, paste0("requested_GO_ontology_", file_stub, ".png")),
    p, width = figure_width, height = figure_height, dpi = 300
  )
  p
}
```

```{r requested-go-ontology-infected-within-diet, fig.height=22, fig.width=30}
plot_directional_go_family("Infected within diet")
```

```{r requested-go-ontology-control-animals-only, fig.height=10, fig.width=18}
plot_directional_go_family("Control animals only")
```

## KEGG pathway overview

KEGG pathways are labelled with descriptive pathway names and the pathway ID in
brackets. The names are resolved from KEGGREST when available, with an offline
fallback for common KEGG pathway IDs so the page remains reproducible. Pathways
are grouped into broad themes to keep immune, nutrition/storage, detoxification,
and core metabolic signals visually separate. Each pathway is shown once per
theme row, so shared pathways align across DEG-set columns. The biological
theme is shown by the right-side row label and colored panel frames.

```{r kegg-overview, fig.height=15, fig.width=25}
if (nrow(kegg_enrichment_df) > 0) {
  kegg_plot_df <- kegg_enrichment_df |>
    dplyr::mutate(
      pathway_theme = dplyr::coalesce(as.character(pathway_theme),
                                      "Other KEGG pathway"),
      pathway_scope = dplyr::coalesce(as.character(pathway_scope),
                                      "Unclassified KEGG reference label"),
      set_label = factor(stringr::str_replace_all(set, "_", " "),
                         levels = overview_set_label_levels),
      neg_log10_padj = -log10(pmax(p.adjust, .Machine$double.xmin))
    ) |>
    dplyr::group_by(set_label, pathway_theme) |>
    dplyr::arrange(p.adjust, dplyr::desc(Count), .by_group = TRUE) |>
    dplyr::slice_head(n = 5) |>
    dplyr::ungroup() |>
    dplyr::arrange(set_label, pathway_theme, p.adjust, dplyr::desc(Count)) |>
    dplyr::mutate(
      term_label = stringr::str_wrap(paste0(Description, " [", ID, "]"),
                                     width = 80),
      pathway_theme = factor(pathway_theme, levels = names(kegg_theme_colors)),
      term_label_facet = paste(term_label, pathway_theme, sep = "___")
    )

  kegg_term_levels <- kegg_plot_df |>
    dplyr::group_by(pathway_theme, term_label_facet) |>
    dplyr::summarise(
      best_p = min(p.adjust, na.rm = TRUE),
      max_count = max(Count, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::arrange(pathway_theme, best_p, dplyr::desc(max_count)) |>
    dplyr::pull(term_label_facet)

  kegg_plot_df <- kegg_plot_df |>
    dplyr::mutate(
      term_label_facet = factor(term_label_facet,
                                levels = rev(unique(kegg_term_levels)))
    )
  kegg_frame_df <- kegg_plot_df |>
    dplyr::distinct(pathway_theme)
  kegg_missing_set_labels <- setdiff(
    overview_set_label_levels,
    as.character(stats::na.omit(unique(kegg_plot_df$set_label)))
  )
  kegg_column_dummy <- data.frame(
    set_label = factor(character(), levels = overview_set_label_levels),
    pathway_theme = factor(character(), levels = names(kegg_theme_colors)),
    term_label_facet = factor(character(),
                              levels = levels(kegg_plot_df$term_label_facet)),
    neg_log10_padj = numeric(),
    Count = numeric()
  )
  if (length(kegg_missing_set_labels) > 0) {
    kegg_column_dummy <- data.frame(
      set_label = factor(kegg_missing_set_labels,
                         levels = overview_set_label_levels),
      pathway_theme = factor(as.character(kegg_plot_df$pathway_theme[1]),
                             levels = names(kegg_theme_colors)),
      term_label_facet = factor(as.character(kegg_plot_df$term_label_facet[1]),
                                levels = levels(kegg_plot_df$term_label_facet)),
      neg_log10_padj = 0,
      Count = min(kegg_plot_df$Count, na.rm = TRUE)
    )
  }

  p_kegg <- ggplot2::ggplot(
    kegg_plot_df,
    ggplot2::aes(y = term_label_facet)
  ) +
    ggplot2::geom_blank(
      data = kegg_column_dummy,
      ggplot2::aes(x = neg_log10_padj, y = term_label_facet),
      inherit.aes = FALSE
    ) +
    ggplot2::geom_rect(
      data = kegg_frame_df,
      ggplot2::aes(color = pathway_theme),
      xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf,
      fill = NA, linewidth = 1.2, inherit.aes = FALSE
    ) +
    ggplot2::geom_vline(xintercept = -log10(0.05),
                        linetype = "dashed", color = "#374151",
                        linewidth = 0.35) +
    ggplot2::geom_point(ggplot2::aes(x = neg_log10_padj, size = Count,
                                     fill = pathway_theme),
                        shape = 21, color = "#1f2937",
                        alpha = 0.9,
                        stroke = 0.35) +
    ggplot2::facet_grid(pathway_theme ~ set_label, scales = "free_y",
                        space = "free_y") +
    ggplot2::scale_color_manual(values = kegg_theme_colors) +
    ggplot2::scale_fill_manual(values = kegg_theme_colors) +
    ggplot2::scale_y_discrete(labels = function(x) sub("___.*$", "", x)) +
    ggplot2::scale_x_continuous(
      limits = c(0, NA),
      expand = ggplot2::expansion(mult = c(0.02, 0.05))
    ) +
    ggplot2::scale_size(range = c(4.5, 13)) +
    ggplot2::guides(
      color = "none",
      fill = "none",
      size = ggplot2::guide_legend(order = 1)
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      strip.background = ggplot2::element_rect(fill = "#F1F4F3",
                                               color = "#C8D0CC"),
      strip.text.x = ggplot2::element_text(size = 18, face = "bold"),
      strip.text.y = ggplot2::element_text(size = 14, face = "bold",
                                           angle = 0),
      axis.text.y = ggplot2::element_text(size = 18, lineheight = 0.95,
                                          hjust = 1),
      panel.border = ggplot2::element_blank(),
      legend.position = "bottom",
      legend.box = "vertical",
      legend.title = ggplot2::element_text(size = 12, face = "bold"),
      legend.text = ggplot2::element_text(size = 11),
      plot.margin = ggplot2::margin(10, 16, 10, 16)
    ) +
    ggplot2::labs(
      title = "eggNOG KEGG pathway enrichment by biological theme",
      subtitle = "Top KEGG pathways per DEG set and theme",
      x = "-log10 adjusted p-value",
      y = NULL,
      size = "DEGs in pathway",
      color = "KEGG theme"
    )

  print(p_kegg)
  ggplot2::ggsave(file.path(out_dir, "eggNOG_KEGG_pathway_enrichment.png"),
                  p_kegg, width = 25, height = 15, dpi = 300)
}
```

## Directional KEGG pathways by biological theme

These panels repeat the requested direction-aware contrasts for KEGG pathways.
The pathway descriptions are shown alongside IDs, and pathways are grouped by
the same broad biological themes used in the non-directional overview. The
colored panel frames mark pathway theme, and the enrichment dots mark the side
of the contrast carrying the enriched genes and the number of DEGs in the
pathway.

```{r requested-kegg-theme-setup}
if (nrow(directional_kegg_enrichment_df) > 0 && "ID" %in% names(directional_kegg_enrichment_df)) {
  directional_kegg_plot_source <- directional_kegg_enrichment_df |>
    dplyr::mutate(
      pathway_theme = dplyr::coalesce(as.character(pathway_theme),
                                      "Other KEGG pathway"),
      pathway_scope = dplyr::coalesce(as.character(pathway_scope),
                                      "Unclassified KEGG reference label"),
      pathway_theme = factor(pathway_theme, levels = names(kegg_theme_colors)),
      family = factor(
        as.character(family),
        levels = c("Infected within diet", "Control animals only",
                   "Infected animals only")
      ),
      direction = factor(as.character(direction),
                         levels = c("Upregulated", "Downregulated")),
      neg_log10_padj = -log10(pmax(p.adjust, .Machine$double.xmin))
    )
} else {
  directional_kegg_plot_source <- data.frame()
}

write.csv(directional_kegg_plot_source,
          file.path(out_dir, "requested_directional_KEGG_enrichment_with_themes.csv"),
          row.names = FALSE)

plot_directional_kegg_family <- function(family_name) {
  family_df <- directional_kegg_plot_source |>
    dplyr::filter(family == family_name, !is.na(pathway_theme))
  if (nrow(family_df) == 0) {
    cat("No requested KEGG enrichment terms to plot for ", family_name, ".\n",
        sep = "")
    return(NULL)
  }

  plot_order <- directional_set_summary |>
    dplyr::filter(family == family_name) |>
    dplyr::pull(plot_label) |>
    unique()

  plot_df <- family_df |>
    dplyr::group_by(plot_label, direction, pathway_theme) |>
    dplyr::arrange(p.adjust, dplyr::desc(Count), .by_group = TRUE) |>
    dplyr::slice_head(n = 5) |>
    dplyr::ungroup() |>
    dplyr::arrange(plot_label, direction, pathway_theme, p.adjust,
                   dplyr::desc(Count)) |>
    dplyr::mutate(
      plot_label = factor(plot_label, levels = plot_order),
      term_label = stringr::str_wrap(paste0(Description, " [", ID, "]"),
                                     width = 80),
      term_label_facet = paste(term_label, pathway_theme, sep = "___")
    )

  kegg_term_levels <- plot_df |>
    dplyr::group_by(pathway_theme, term_label_facet) |>
    dplyr::summarise(
      best_p = min(p.adjust, na.rm = TRUE),
      max_count = max(Count, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::arrange(pathway_theme, best_p, dplyr::desc(max_count)) |>
    dplyr::pull(term_label_facet)

  plot_df <- plot_df |>
    dplyr::mutate(
      term_label_facet = factor(term_label_facet,
                                levels = rev(unique(kegg_term_levels)))
    )
  kegg_frame_df <- plot_df |>
    dplyr::distinct(pathway_theme)
  direction_legend_df <- data.frame(
    plot_label = factor(levels(plot_df$plot_label)[1],
                        levels = levels(plot_df$plot_label)),
    pathway_theme = factor(as.character(plot_df$pathway_theme[1]),
                           levels = levels(plot_df$pathway_theme)),
    term_label_facet = plot_df$term_label_facet[1],
    neg_log10_padj = 0,
    Count = min(plot_df$Count, na.rm = TRUE),
    direction = factor(names(direction_colors), levels = names(direction_colors))
  )
  kegg_strip_x_size <- dplyr::case_when(
    family_name == "Infected within diet" ~ 16,
    TRUE ~ 18
  )
  direction_legend_title <- if (family_name == "Infected within diet") {
    "Infection response"
  } else {
    "Diet side"
  }
  direction_legend_labels <- if (family_name == "Infected within diet") {
    c("Upregulated" = "Higher infected",
      "Downregulated" = "Higher control")
  } else {
    c("Upregulated" = "Higher first diet",
      "Downregulated" = "Higher second diet")
  }
  missing_plot_labels <- setdiff(
    as.character(plot_order),
    as.character(stats::na.omit(unique(plot_df$plot_label)))
  )
  no_terms_df <- data.frame(
    plot_label = factor(character(), levels = plot_order),
    pathway_theme = factor(character(),
                           levels = levels(plot_df$pathway_theme)),
    term_label_facet = factor(character(),
                              levels = levels(plot_df$term_label_facet)),
    label_x = numeric(),
    label = character()
  )
  if (length(missing_plot_labels) > 0) {
    label_x <- max(plot_df$neg_log10_padj, na.rm = TRUE) * 0.55
    if (!is.finite(label_x) || label_x <= 0) label_x <- 1
    kegg_y_by_row <- plot_df |>
      dplyr::group_by(pathway_theme) |>
      dplyr::summarise(
        term_label_facet = dplyr::first(term_label_facet),
        .groups = "drop"
      )
    no_terms_df <- expand.grid(
      plot_label = missing_plot_labels,
      pathway_theme = unique(as.character(kegg_y_by_row$pathway_theme)),
      stringsAsFactors = FALSE
    ) |>
      dplyr::left_join(
        kegg_y_by_row |>
          dplyr::mutate(pathway_theme = as.character(pathway_theme)),
        by = "pathway_theme"
      ) |>
      dplyr::mutate(
        plot_label = factor(plot_label, levels = plot_order),
        pathway_theme = factor(
          pathway_theme,
          levels = levels(plot_df$pathway_theme)
        ),
        term_label_facet = factor(
          as.character(term_label_facet),
          levels = levels(plot_df$term_label_facet)
        ),
        label_x = label_x,
        label = "No enriched terms"
      )
  }

  p <- ggplot2::ggplot(
    plot_df,
    ggplot2::aes(y = term_label_facet)
  ) +
    ggplot2::geom_rect(
      data = kegg_frame_df,
      ggplot2::aes(color = pathway_theme),
      xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf,
      fill = NA, linewidth = 1.2, inherit.aes = FALSE
    ) +
    ggplot2::geom_vline(xintercept = -log10(0.05),
                        linetype = "dashed", color = "#374151",
                        linewidth = 0.35) +
    ggplot2::geom_text(
      data = no_terms_df,
      ggplot2::aes(x = label_x, y = term_label_facet, label = label),
      inherit.aes = FALSE, color = "#6B7280", fontface = "italic",
      size = 5
    ) +
    ggplot2::geom_point(
      ggplot2::aes(x = neg_log10_padj, size = Count, fill = direction,
                   shape = direction),
      color = "#1f2937", alpha = 0.9, stroke = 0.35
    ) +
    ggplot2::geom_point(
      data = direction_legend_df,
      ggplot2::aes(x = neg_log10_padj, y = term_label_facet,
                   fill = direction, shape = direction),
      inherit.aes = FALSE, color = "#1f2937", size = 5,
      alpha = 0, show.legend = TRUE
    ) +
    ggplot2::facet_grid(pathway_theme ~ plot_label, scales = "free_y",
                        space = "free_y") +
    ggplot2::scale_color_manual(values = kegg_theme_colors) +
    ggplot2::scale_fill_manual(name = direction_legend_title,
                               values = direction_colors,
                               breaks = names(direction_colors),
                               labels = direction_legend_labels,
                               drop = FALSE) +
    ggplot2::scale_shape_manual(values = c("Upregulated" = 21,
                                           "Downregulated" = 24),
                                breaks = names(direction_colors),
                                name = direction_legend_title,
                                labels = direction_legend_labels,
                                drop = FALSE) +
    ggplot2::scale_y_discrete(labels = function(x) sub("___.*$", "", x)) +
    ggplot2::scale_x_continuous(
      limits = c(0, NA),
      expand = ggplot2::expansion(mult = c(0.02, 0.05))
    ) +
    ggplot2::scale_size(range = c(4.5, 13)) +
    ggplot2::guides(
      color = "none",
      fill = "none",
      shape = ggplot2::guide_legend(
        nrow = 1, byrow = TRUE, order = 1,
        override.aes = list(fill = unname(direction_colors), size = 5,
                            alpha = 1,
                            color = "#1f2937")
      ),
      size = ggplot2::guide_legend(order = 2)
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      strip.background = ggplot2::element_rect(fill = "#F1F4F3",
                                               color = "#C8D0CC"),
      strip.text.x = ggplot2::element_text(size = kegg_strip_x_size,
                                           face = "bold"),
      strip.text.y = ggplot2::element_text(size = 14, face = "bold",
                                           angle = 0),
      axis.text.y = ggplot2::element_text(size = 18, lineheight = 0.95,
                                          hjust = 1),
      panel.border = ggplot2::element_blank(),
      legend.position = "bottom",
      legend.box = "vertical",
      plot.margin = ggplot2::margin(10, 16, 10, 16)
    ) +
    ggplot2::labs(
      title = paste("eggNOG KEGG enrichment:", family_name),
      subtitle = "Directional DEG foregrounds; pathway groups use broad biological themes",
      x = "-log10 adjusted p-value",
      y = "KEGG pathway",
      size = "DEGs in pathway",
      color = "KEGG theme",
      fill = direction_legend_title,
      shape = direction_legend_title
    )

  file_stub <- stringr::str_replace_all(stringr::str_to_lower(family_name),
                                        "[^a-z0-9]+", "_")
  figure_width <- dplyr::case_when(
    family_name == "Infected within diet" ~ 30,
    TRUE ~ 18
  )
  figure_height <- dplyr::case_when(
    family_name == "Infected within diet" ~ 18,
    family_name == "Infected animals only" ~ 15,
    TRUE ~ 12
  )
  ggplot2::ggsave(
    file.path(out_dir, paste0("requested_KEGG_theme_", file_stub, ".png")),
    p, width = figure_width, height = figure_height, dpi = 300
  )
  p
}
```

```{r requested-kegg-theme-infected-within-diet, fig.height=18, fig.width=30}
plot_directional_kegg_family("Infected within diet")
```

```{r requested-kegg-theme-control-animals-only, fig.height=12, fig.width=18}
plot_directional_kegg_family("Control animals only")
```

```{r requested-kegg-theme-infected-animals-only, fig.height=15, fig.width=18}
plot_directional_kegg_family("Infected animals only")
```

```{r session-info}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
