---
title: "GO term and KEGG enrichment"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  counts_file: "data/raw_read_counts/master_counts_no_rRNA.csv"
  eggnog_file: "data/GO_Annotations/GCF_023897955.1_iqSchGreg1.2_Arthopoda_one2one.emapper.annotations"
  gff_file: "data/reference/GCF_023897955.1_iqSchGreg1.2_genomic.gff"
  latest_results_root: "output/rmd_runs"
  alpha: 0.05
  lfc_threshold: 1
  top_terms: 8
  output_tag: "go-term-enrichment"
---

## Purpose

This page keeps the eggNOG-based functional enrichment with the differential
expression results. It uses the latest DESeq2 result tables on disk, lifts
eggNOG protein annotations onto the `LOC...` gene IDs in the count matrix, and
tests infection DEG sets for enriched GO terms and KEGG pathways.

The key point is that this is a custom enrichment workflow for this reference:
the foregrounds are current DEG sets, the background is the annotated gene
universe in this count matrix, and GO names/ontologies come from `GO.db`.

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c(
  "dplyr", "tidyr", "stringr", "ggplot2", "forcats", "clusterProfiler",
  "GO.db", "AnnotationDbi", "DT"
)
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

counts_file <- file.path(project_dir, params$counts_file)
eggnog_file <- file.path(project_dir, params$eggnog_file)
gff_file <- file.path(project_dir, params$gff_file)
results_root <- file.path(project_dir, params$latest_results_root)

diet_colors <- c("33" = "#E6862E", "50" = "#2F8F5B", "83" = "#7B4FA3")
ontology_colors <- c(
  "Biological process" = "#2F8F5B",
  "Molecular function" = "#E6862E",
  "Cellular component" = "#7B4FA3"
)
kegg_theme_colors <- c(
  "Immune and stress signaling" = "#C23B22",
  "Storage, lipid transport, and endocrine nutrition" = "#D7832F",
  "Amino acid and nitrogen metabolism" = "#6F4AA8",
  "Carbohydrate and energy metabolism" = "#2C7FB8",
  "Lipid and membrane metabolism" = "#2F8F5B",
  "Detoxification and secondary metabolism" = "#8C6D31",
  "Transport and environmental response" = "#00A6A6",
  "Genetic information processing" = "#6B7280",
  "Cellular process" = "#9C5DA5",
  "Broad metabolic map" = "#4B5563",
  "Reference disease pathway" = "#A0AEC0",
  "Other KEGG pathway" = "#7B8794"
)
```

## Build eggNOG annotation maps

eggNOG reports annotations at the protein-query level, while the count matrix
uses `LOC...` gene IDs. The GFF provides the bridge from CDS protein IDs to gene
IDs. Only genes present in the count matrix are kept in the enrichment
background.

```{r annotation-maps}
read_emapper <- function(path) {
  cols <- c(
    "query", "seed_ortholog", "evalue", "score", "eggNOG_OGs",
    "max_annot_lvl", "COG_category", "Description", "Preferred_name",
    "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
    "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC", "CAZy",
    "BiGG_Reaction", "PFAMs"
  )
  tab <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", fill = TRUE)
  names(tab) <- cols[seq_len(ncol(tab))]
  tab |>
    dplyr::mutate(protein_id = stringr::str_extract(query, "XP_[0-9]+[.][0-9]+"))
}

parse_cds_gene_bridge <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  gff |>
    dplyr::filter(type == "CDS") |>
    dplyr::transmute(
      protein_id = stringr::str_match(attributes, "protein_id=([^;]+)")[, 2],
      gene_id = stringr::str_match(attributes, "gene=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(protein_id), !is.na(gene_id)) |>
    dplyr::distinct()
}

eggnog <- read_emapper(eggnog_file)
protein_to_gene <- parse_cds_gene_bridge(gff_file)
counts_raw <- read.csv(counts_file, check.names = FALSE)
gene_universe <- unique(counts_raw[[1]])

annotation <- eggnog |>
  dplyr::left_join(protein_to_gene, by = "protein_id") |>
  dplyr::filter(!is.na(gene_id), gene_id %in% gene_universe) |>
  dplyr::distinct(gene_id, .keep_all = TRUE)

annotation_summary <- data.frame(
  genes_in_count_matrix = length(gene_universe),
  annotated_genes_lifted_to_LOC = dplyr::n_distinct(annotation$gene_id),
  genes_with_GO = sum(annotation$GOs != "-" & annotation$GOs != ""),
  genes_with_KEGG_KO = sum(annotation$KEGG_ko != "-" & annotation$KEGG_ko != ""),
  genes_with_KEGG_pathway = sum(annotation$KEGG_Pathway != "-" &
                                  annotation$KEGG_Pathway != "")
)

knitr::kable(annotation_summary)
write.csv(annotation_summary, file.path(out_dir, "annotation_summary.csv"),
          row.names = FALSE)
```

```{r term-maps}
term2gene_go <- annotation |>
  dplyr::select(gene_id, GOs) |>
  tidyr::separate_rows(GOs, sep = ",") |>
  dplyr::filter(grepl("^GO:", GOs)) |>
  dplyr::transmute(term = GOs, gene = gene_id) |>
  dplyr::distinct()

go_names <- AnnotationDbi::mapIds(
  GO.db::GO.db,
  keys = unique(term2gene_go$term),
  keytype = "GOID",
  column = "TERM",
  multiVals = "first"
)

go_ontology <- AnnotationDbi::mapIds(
  GO.db::GO.db,
  keys = unique(term2gene_go$term),
  keytype = "GOID",
  column = "ONTOLOGY",
  multiVals = "first"
)

term2name_go <- data.frame(
  term = names(go_names),
  name = unname(go_names),
  ontology = unname(go_ontology)
) |>
  dplyr::filter(!is.na(name))

term2gene_kegg_pathway <- annotation |>
  dplyr::select(gene_id, KEGG_Pathway) |>
  tidyr::separate_rows(KEGG_Pathway, sep = ",") |>
  dplyr::filter(grepl("^ko[0-9]+$", KEGG_Pathway)) |>
  dplyr::transmute(term = KEGG_Pathway, gene = gene_id) |>
  dplyr::distinct()

kegg_fallback_names <- c(
  ko00010 = "Glycolysis / Gluconeogenesis",
  ko00020 = "Citrate cycle (TCA cycle)",
  ko00030 = "Pentose phosphate pathway",
  ko00040 = "Pentose and glucuronate interconversions",
  ko00051 = "Fructose and mannose metabolism",
  ko00052 = "Galactose metabolism",
  ko00053 = "Ascorbate and aldarate metabolism",
  ko00061 = "Fatty acid biosynthesis",
  ko00062 = "Fatty acid elongation",
  ko00071 = "Fatty acid degradation",
  ko00190 = "Oxidative phosphorylation",
  ko00220 = "Arginine biosynthesis",
  ko00230 = "Purine metabolism",
  ko00240 = "Pyrimidine metabolism",
  ko00250 = "Alanine, aspartate and glutamate metabolism",
  ko00260 = "Glycine, serine and threonine metabolism",
  ko00270 = "Cysteine and methionine metabolism",
  ko00280 = "Valine, leucine and isoleucine degradation",
  ko00330 = "Arginine and proline metabolism",
  ko00340 = "Histidine metabolism",
  ko00350 = "Tyrosine metabolism",
  ko00360 = "Phenylalanine metabolism",
  ko00500 = "Starch and sucrose metabolism",
  ko00520 = "Amino sugar and nucleotide sugar metabolism",
  ko00561 = "Glycerolipid metabolism",
  ko00564 = "Glycerophospholipid metabolism",
  ko00565 = "Ether lipid metabolism",
  ko00590 = "Arachidonic acid metabolism",
  ko00591 = "Linoleic acid metabolism",
  ko00592 = "alpha-Linolenic acid metabolism",
  ko00620 = "Pyruvate metabolism",
  ko00630 = "Glyoxylate and dicarboxylate metabolism",
  ko00640 = "Propanoate metabolism",
  ko00650 = "Butanoate metabolism",
  ko00670 = "One carbon pool by folate",
  ko00730 = "Thiamine metabolism",
  ko00860 = "Porphyrin metabolism",
  ko00900 = "Terpenoid backbone biosynthesis",
  ko00910 = "Nitrogen metabolism",
  ko00920 = "Sulfur metabolism",
  ko00930 = "Caprolactam degradation",
  ko00970 = "Aminoacyl-tRNA biosynthesis",
  ko01040 = "Biosynthesis of unsaturated fatty acids",
  ko01100 = "Metabolic pathways",
  ko01110 = "Biosynthesis of secondary metabolites",
  ko01120 = "Microbial metabolism in diverse environments",
  ko01130 = "Biosynthesis of antibiotics",
  ko01200 = "Carbon metabolism",
  ko01212 = "Fatty acid metabolism",
  ko01220 = "Degradation of aromatic compounds",
  ko01230 = "Biosynthesis of amino acids",
  ko02010 = "ABC transporters",
  ko02020 = "Two-component system",
  ko03010 = "Ribosome",
  ko03013 = "Nucleocytoplasmic transport",
  ko03015 = "mRNA surveillance pathway",
  ko03018 = "RNA degradation",
  ko03022 = "Basal transcription factors",
  ko03030 = "DNA replication",
  ko03040 = "Spliceosome",
  ko03050 = "Proteasome",
  ko03060 = "Protein export",
  ko03320 = "PPAR signaling pathway",
  ko04010 = "MAPK signaling pathway",
  ko04014 = "Ras signaling pathway",
  ko04020 = "Calcium signaling pathway",
  ko04024 = "cAMP signaling pathway",
  ko04110 = "Cell cycle",
  ko04120 = "Ubiquitin mediated proteolysis",
  ko04130 = "SNARE interactions in vesicular transport",
  ko04141 = "Protein processing in endoplasmic reticulum",
  ko04142 = "Lysosome",
  ko04144 = "Endocytosis",
  ko04145 = "Phagosome",
  ko04146 = "Peroxisome",
  ko04147 = "Exosome",
  ko04150 = "mTOR signaling pathway",
  ko04151 = "PI3K-Akt signaling pathway",
  ko04152 = "AMPK signaling pathway",
  ko04210 = "Apoptosis",
  ko04213 = "Longevity regulating pathway - multiple species",
  ko04217 = "Necroptosis",
  ko04218 = "Cellular senescence",
  ko04270 = "Vascular smooth muscle contraction",
  ko04310 = "Wnt signaling pathway",
  ko04330 = "Notch signaling pathway",
  ko04340 = "Hedgehog signaling pathway",
  ko04350 = "TGF-beta signaling pathway",
  ko04370 = "VEGF signaling pathway",
  ko04371 = "Apelin signaling pathway",
  ko04510 = "Focal adhesion",
  ko04512 = "ECM-receptor interaction",
  ko04514 = "Cell adhesion molecules",
  ko04520 = "Adherens junction",
  ko04530 = "Tight junction",
  ko04630 = "JAK-STAT signaling pathway",
  ko04668 = "TNF signaling pathway",
  ko04710 = "Circadian rhythm",
  ko04713 = "Circadian entrainment",
  ko04714 = "Thermogenesis",
  ko04723 = "Retrograde endocannabinoid signaling",
  ko04730 = "Long-term depression",
  ko04910 = "Insulin signaling pathway",
  ko04920 = "Adipocytokine signaling pathway",
  ko04922 = "Glucagon signaling pathway",
  ko04926 = "Relaxin signaling pathway",
  ko04970 = "Salivary secretion",
  ko04972 = "Pancreatic secretion",
  ko04975 = "Fat digestion and absorption",
  ko05010 = "Alzheimer disease",
  ko05012 = "Parkinson disease",
  ko05014 = "Amyotrophic lateral sclerosis",
  ko05016 = "Huntington disease"
)

get_kegg_pathway_names <- function(ids) {
  ids <- unique(ids)
  resolved <- setNames(rep(NA_character_, length(ids)), ids)

  if (requireNamespace("KEGGREST", quietly = TRUE)) {
    kegg_list <- tryCatch(
      KEGGREST::keggList("pathway", "ko"),
      error = function(e) character()
    )
    if (length(kegg_list) > 0) {
      live_ids <- sub("^path:", "", names(kegg_list))
      live_names <- sub(" - Reference pathway$", "", unname(kegg_list))
      live_names <- sub(" - Homo sapiens \\(human\\)$", "", live_names)
      resolved[intersect(ids, live_ids)] <- live_names[match(intersect(ids, live_ids),
                                                             live_ids)]
    }
  }

  fallback_ids <- intersect(ids, names(kegg_fallback_names))
  resolved[fallback_ids] <- ifelse(
    is.na(resolved[fallback_ids]),
    unname(kegg_fallback_names[fallback_ids]),
    resolved[fallback_ids]
  )
  resolved[is.na(resolved)] <- ids[is.na(resolved)]
  resolved
}

classify_kegg_theme <- function(ids, names) {
  ids <- as.character(ids)
  names_lower <- stringr::str_to_lower(names)
  dplyr::case_when(
    stringr::str_detect(ids, "^ko046|^ko040|^ko042") |
      stringr::str_detect(names_lower, "toll|imd|tnf|jak-stat|immune|apoptosis|necrosis|stress") ~
      "Immune and stress signaling",
    stringr::str_detect(ids, "^ko049|^ko033|^ko00981") |
      stringr::str_detect(names_lower, "insulin|glucagon|adipocytokine|ppar|hormone|bile|mineral|fat digestion|secretion|absorption") ~
      "Storage, lipid transport, and endocrine nutrition",
    ids %in% c("ko00220", "ko00230", "ko00240", "ko00250", "ko00260",
               "ko00270", "ko00280", "ko00330", "ko00340", "ko00350",
               "ko00360", "ko00380", "ko00410", "ko01230") |
      stringr::str_detect(names_lower, "amino acid|arginine|proline|glycine|serine|threonine|tryptophan|tyrosine|nitrogen") ~
      "Amino acid and nitrogen metabolism",
    ids %in% c("ko00010", "ko00020", "ko00030", "ko00040", "ko00051",
               "ko00052", "ko00053", "ko00190", "ko00620", "ko00630",
               "ko00640", "ko00650", "ko00720", "ko01200") |
      stringr::str_detect(names_lower, "glycolysis|gluconeogenesis|citrate|carbon metabolism|pentose|pyruvate|oxidative phosphorylation|carbon fixation|ascorbate") ~
      "Carbohydrate and energy metabolism",
    ids %in% c("ko00061", "ko00062", "ko00071", "ko00561", "ko00564",
               "ko00565", "ko00590", "ko00591", "ko00592", "ko01040",
               "ko01212") |
      stringr::str_detect(names_lower, "fatty acid|lipid|glycerolipid|glycerophospholipid|linoleic|arachidonic") ~
      "Lipid and membrane metabolism",
    ids %in% c("ko00480", "ko00900", "ko00920", "ko00930", "ko00970",
               "ko00980", "ko01110", "ko01120", "ko01130", "ko01220") |
      stringr::str_detect(names_lower, "glutathione|xenobiotic|cytochrome|secondary metabolite|microbial metabolism|antibiotic|aromatic") ~
      "Detoxification and secondary metabolism",
    stringr::str_detect(ids, "^ko020") |
      stringr::str_detect(names_lower, "transporter|two-component|environmental") ~
      "Transport and environmental response",
    stringr::str_detect(ids, "^ko030") |
      stringr::str_detect(names_lower, "ribosome|spliceosome|rna|dna|replication|transcription|proteasome") ~
      "Genetic information processing",
    stringr::str_detect(ids, "^ko041|^ko043|^ko045") |
      stringr::str_detect(names_lower, "cell cycle|endocytosis|phagosome|peroxisome|junction|adhesion|notch|wnt|hedgehog") ~
      "Cellular process",
    ids == "ko01100" ~ "Broad metabolic map",
    stringr::str_detect(ids, "^ko050|^ko052") ~ "Reference disease pathway",
    TRUE ~ "Other KEGG pathway"
  )
}

kegg_names <- get_kegg_pathway_names(unique(term2gene_kegg_pathway$term))

term2name_kegg_pathway <- term2gene_kegg_pathway |>
  dplyr::distinct(term) |>
  dplyr::mutate(
    name = unname(kegg_names[term]),
    pathway_theme = classify_kegg_theme(term, name)
  )

write.csv(term2gene_go, file.path(out_dir, "eggNOG_GO_TERM2GENE.csv"),
          row.names = FALSE)
write.csv(term2name_go, file.path(out_dir, "eggNOG_GO_TERM2NAME.csv"),
          row.names = FALSE)
write.csv(term2gene_kegg_pathway,
          file.path(out_dir, "eggNOG_KEGG_pathway_TERM2GENE.csv"),
          row.names = FALSE)
write.csv(term2name_kegg_pathway,
          file.path(out_dir, "eggNOG_KEGG_pathway_TERM2NAME.csv"),
          row.names = FALSE)
```

## Collect current infection DEG sets

The enrichment foregrounds are built from infected-versus-control DEGs without
splitting genes by direction. This keeps the biological question simple: which
functional categories are over-represented among infection-responsive genes in
general, within each diet, and among genes shared across diets. The newest file
for each within-diet infection result table is used so older workflowR runs do
not duplicate a contrast.

```{r deg-sets}
infection_result_files <- list.files(
  results_root,
  pattern = "^DESeq2_infected_vs_control_diet_[0-9]+[.]csv$",
  recursive = TRUE,
  full.names = TRUE
)

infection_file_info <- data.frame(
  file = infection_result_files,
  short_name = basename(infection_result_files),
  modified = file.info(infection_result_files)$mtime,
  stringsAsFactors = FALSE
)
infection_file_info <- infection_file_info[order(infection_file_info$short_name,
                                                 -as.numeric(infection_file_info$modified)), ]
infection_file_info <- infection_file_info[!duplicated(infection_file_info$short_name), ]
infection_result_files <- infection_file_info$file

read_infection_deg_set <- function(path) {
  tab <- read.csv(path, check.names = FALSE)
  gene_col <- intersect(c("gene_id", "gene", "GeneID"), names(tab))[1]
  if (is.na(gene_col) || !"padj" %in% names(tab) ||
      !"log2FoldChange" %in% names(tab)) return(NULL)
  sig <- !is.na(tab$padj) &
    tab$padj < params$alpha &
    abs(tab$log2FoldChange) >= params$lfc_threshold
  unique(as.character(tab[[gene_col]][sig]))
}

infection_sets_by_diet <- lapply(infection_result_files, read_infection_deg_set)
names(infection_sets_by_diet) <- sub(
  "^DESeq2_infected_vs_control_diet_([0-9]+)[.]csv$",
  "Diet \\1",
  basename(infection_result_files)
)
infection_sets_by_diet <- infection_sets_by_diet[lengths(infection_sets_by_diet) > 0]
infection_sets_by_diet <- infection_sets_by_diet[order(names(infection_sets_by_diet))]

if (length(infection_sets_by_diet) == 0) {
  stop("No within-diet infection DEG result files were found. Build page 06 first.")
}

all_infection_genes <- sort(unique(unlist(infection_sets_by_diet, use.names = FALSE)))
diet_hit_count <- table(unlist(lapply(names(infection_sets_by_diet), function(nm) {
  unique(infection_sets_by_diet[[nm]])
}), use.names = FALSE))
shared_two_or_more <- names(diet_hit_count)[diet_hit_count >= 2]
shared_all_diets <- names(diet_hit_count)[diet_hit_count == length(infection_sets_by_diet)]

enrichment_sets <- c(
  list("All infection DEGs" = all_infection_genes),
  infection_sets_by_diet,
  list("Shared by at least two diets" = shared_two_or_more,
       "Shared by all diets" = shared_all_diets)
)

deg_set_summary <- data.frame(
  DEG_set = names(enrichment_sets),
  n_genes = lengths(enrichment_sets),
  stringsAsFactors = FALSE
)
deg_set_summary <- deg_set_summary[order(factor(
  deg_set_summary$DEG_set,
  levels = names(enrichment_sets)
)), ]

source_table <- data.frame(
  DEG_set = names(infection_sets_by_diet),
  result_file = infection_result_files,
  modified = file.info(infection_result_files)$mtime,
  stringsAsFactors = FALSE
)

read_infection_deg_table <- function(path) {
  tab <- read.csv(path, check.names = FALSE)
  gene_col <- intersect(c("gene_id", "gene", "GeneID"), names(tab))[1]
  if (is.na(gene_col) || !"padj" %in% names(tab) ||
      !"log2FoldChange" %in% names(tab)) return(NULL)
  diet <- sub("^DESeq2_infected_vs_control_diet_([0-9]+)[.]csv$",
              "\\1", basename(path))
  tab |>
    dplyr::transmute(
      gene_id = as.character(.data[[gene_col]]),
      diet = diet,
      DEG_set = paste("Diet", diet),
      log2FoldChange = log2FoldChange,
      padj = padj,
      significant = !is.na(padj) &
        padj < params$alpha &
        abs(log2FoldChange) >= params$lfc_threshold
    )
}

infection_deg_long <- dplyr::bind_rows(
  lapply(infection_result_files, read_infection_deg_table)
) |>
  dplyr::filter(significant)

DT::datatable(deg_set_summary, options = list(pageLength = 15))
write.csv(deg_set_summary, file.path(out_dir, "go_enrichment_deg_set_summary.csv"),
          row.names = FALSE)
write.csv(source_table, file.path(out_dir, "infection_deg_source_files.csv"),
          row.names = FALSE)
write.csv(infection_deg_long, file.path(out_dir, "infection_DEGs_long.csv"),
          row.names = FALSE)
```

## Fat body immune and storage candidates

The fat body is central to immune defense, detoxification, lipid transport, and
nutrient storage. This table therefore flags significant infection-responsive
genes whose eggNOG annotation contains immune, storage-protein, lipid-transport,
endocrine-nutrition, or detoxification language. This is a transparent keyword
screen to help prioritize genes for biological interpretation; it does not
replace the GO or KEGG enrichment tests.

```{r fat-body-candidates}
classify_fat_body_candidate <- function(text) {
  text <- stringr::str_to_lower(text)
  immune <- stringr::str_detect(
    text,
    "immune|defen[cs]e|defensin|attacin|cecropin|gloverin|lysozyme|toll|imd|tnf|jak-stat|peptidoglycan|antimicrobial|phenoloxidase|prophenoloxidase|serpin|thioester|transferrin|lectin|phagosome"
  )
  storage <- stringr::str_detect(
    text,
    "storage protein|hexamerin|arylphorin|hemocyanin|vitellogenin|lipophorin|apolipophorin|nutrient reservoir"
  )
  detox <- stringr::str_detect(
    text,
    "detox|glutathione|cytochrome p450|p450|xenobiotic|oxidative stress|peroxidase|superoxide|catalase"
  )
  endocrine_nutrition <- stringr::str_detect(
    text,
    "juvenile hormone|insulin|glucagon|adipocytokine|ppar|bile secretion|mineral absorption|fat digestion|glycerolipid|amino acid|lipid transport"
  )

  dplyr::case_when(
    immune & storage ~ "Immune + storage/lipid transport",
    storage ~ "Storage / lipid transport",
    immune ~ "Immune / defense",
    detox ~ "Detoxification / oxidative stress",
    endocrine_nutrition ~ "Endocrine / nutrient metabolism",
    TRUE ~ NA_character_
  )
}

fat_body_candidates <- infection_deg_long |>
  dplyr::inner_join(
    annotation |>
      dplyr::select(gene_id, Preferred_name, Description, GOs, KEGG_ko,
                    KEGG_Pathway, BRITE, PFAMs),
    by = "gene_id"
  ) |>
  dplyr::mutate(
    annotation_text = paste(Preferred_name, Description, GOs, KEGG_ko,
                            KEGG_Pathway, BRITE, PFAMs, sep = " "),
    candidate_theme = classify_fat_body_candidate(annotation_text)
  ) |>
  dplyr::filter(!is.na(candidate_theme)) |>
  dplyr::arrange(candidate_theme, padj) |>
  dplyr::group_by(gene_id, candidate_theme, Preferred_name, Description, PFAMs) |>
  dplyr::summarise(
    DEG_sets = paste(unique(DEG_set), collapse = "; "),
    log2FC_by_diet = paste0(DEG_set, "=", round(log2FoldChange, 2),
                            collapse = "; "),
    min_padj = signif(min(padj, na.rm = TRUE), 3),
    max_abs_log2FC = round(max(abs(log2FoldChange), na.rm = TRUE), 2),
    KEGG_Pathway = paste(unique(KEGG_Pathway), collapse = "; "),
    .groups = "drop"
  ) |>
  dplyr::arrange(candidate_theme, min_padj)

write.csv(fat_body_candidates,
          file.path(out_dir, "fat_body_immune_storage_DEG_candidates.csv"),
          row.names = FALSE)

if (nrow(fat_body_candidates) > 0) {
  DT::datatable(
    fat_body_candidates,
    rownames = FALSE,
    options = list(pageLength = 15, scrollX = TRUE)
  )
} else {
  cat("No significant infection DEGs matched the current fat-body candidate keyword screen.")
}
```

## Run GO and KEGG enrichment

Enrichment tests whether a DEG set contains more genes from a term than expected
from the annotated background. Here the DEG sets are not separated by
upregulation or downregulation; each foreground contains significant
infected-versus-control DEGs regardless of fold-change direction.

```{r run-enrichment}
run_enrichment <- function(genes, term2gene, term2name, universe) {
  genes <- unique(genes[genes %in% universe])
  if (length(genes) < 5) return(NULL)
  clusterProfiler::enricher(
    gene = genes,
    universe = universe,
    TERM2GENE = term2gene,
    TERM2NAME = term2name[, c("term", "name")],
    pvalueCutoff = 0.1,
    pAdjustMethod = "BH",
    qvalueCutoff = 0.2
  )
}

universe_go <- unique(term2gene_go$gene)
universe_kegg <- unique(term2gene_kegg_pathway$gene)

go_results <- lapply(enrichment_sets, run_enrichment,
                     term2gene = term2gene_go,
                     term2name = term2name_go,
                     universe = universe_go)
kegg_results <- lapply(enrichment_sets, run_enrichment,
                       term2gene = term2gene_kegg_pathway,
                       term2name = term2name_kegg_pathway,
                       universe = universe_kegg)

enrich_to_df <- function(result_list, database) {
  pieces <- lapply(names(result_list), function(nm) {
    res <- result_list[[nm]]
    if (is.null(res)) return(NULL)
    df <- as.data.frame(res)
    if (nrow(df) == 0) return(NULL)
    df$set <- nm
    df$database <- database
    df
  })
  dplyr::bind_rows(pieces)
}

go_enrichment_df <- enrich_to_df(go_results, "GO")
kegg_enrichment_df <- enrich_to_df(kegg_results, "KEGG_pathway")
all_enrichment <- dplyr::bind_rows(go_enrichment_df, kegg_enrichment_df)

write.csv(go_enrichment_df, file.path(out_dir, "GO_enrichment_all_sets.csv"),
          row.names = FALSE)
write.csv(kegg_enrichment_df,
          file.path(out_dir, "KEGG_pathway_enrichment_all_sets.csv"),
          row.names = FALSE)
write.csv(all_enrichment, file.path(out_dir, "all_enrichment_results.csv"),
          row.names = FALSE)

if (nrow(all_enrichment) == 0) {
  cat("No enriched terms passed the current cutoffs.")
} else {
  DT::datatable(
    all_enrichment |>
      dplyr::select(database, set, ID, Description, GeneRatio, BgRatio,
                    p.adjust, qvalue, geneID) |>
      dplyr::arrange(p.adjust),
    options = list(pageLength = 15, scrollX = TRUE)
  )
}
```

## GO terms by ontology

This figure separates biological process, molecular function, and cellular
component terms. A few eggNOG GO IDs can be obsolete or unavailable in the
local `GO.db` ontology table; those terms are kept in the complete enrichment
CSV but omitted from this ontology-specific plot. The dashed vertical line
marks FDR = 0.05.

```{r go-ontology-figure, fig.height=10, fig.width=13}
if (nrow(go_enrichment_df) > 0 && "ID" %in% names(go_enrichment_df)) {
  go_enrichment_ontology <- go_enrichment_df |>
    dplyr::left_join(term2name_go |> dplyr::select(term, ontology),
                     by = c("ID" = "term")) |>
    dplyr::mutate(
      ontology = dplyr::case_when(
        ontology == "BP" ~ "Biological process",
        ontology == "MF" ~ "Molecular function",
        ontology == "CC" ~ "Cellular component",
        TRUE ~ NA_character_
      ),
      neg_log10_padj = -log10(pmax(p.adjust, .Machine$double.xmin)),
      term_label = paste0(stringr::str_trunc(Description, 52), " [", ID, "]"),
      set_label = stringr::str_replace_all(set, "_", " ")
    )
} else {
  go_enrichment_ontology <- data.frame()
}

write.csv(go_enrichment_ontology,
          file.path(out_dir, "eggNOG_GO_enrichment_with_ontology.csv"),
          row.names = FALSE)

unmapped_go_terms <- go_enrichment_ontology |>
  dplyr::filter(is.na(ontology)) |>
  dplyr::distinct(ID, Description, set, p.adjust, Count)
write.csv(unmapped_go_terms,
          file.path(out_dir, "eggNOG_GO_enrichment_unmapped_ontology.csv"),
          row.names = FALSE)

if (nrow(unmapped_go_terms) > 0) {
  cat(nrow(unmapped_go_terms),
      "enriched GO rows did not map to BP, MF, or CC in the local GO.db and are omitted from the ontology plot.\n")
}

if (nrow(go_enrichment_ontology) > 0) {
  go_plot_df <- go_enrichment_ontology |>
    dplyr::filter(!is.na(p.adjust), !is.na(ontology)) |>
    dplyr::group_by(set_label, ontology) |>
    dplyr::arrange(p.adjust, dplyr::desc(Count), .by_group = TRUE) |>
    dplyr::slice_head(n = params$top_terms) |>
    dplyr::ungroup() |>
    dplyr::mutate(
      term_label = forcats::fct_reorder(term_label, neg_log10_padj)
    )

  p_go_ontology <- ggplot2::ggplot(
    go_plot_df,
    ggplot2::aes(x = neg_log10_padj, y = term_label)
  ) +
    ggplot2::geom_vline(xintercept = -log10(0.05),
                        linetype = "dashed", color = "#374151",
                        linewidth = 0.35) +
    ggplot2::geom_point(ggplot2::aes(size = Count, fill = ontology),
                        shape = 21, color = "#1f2937", alpha = 0.9,
                        stroke = 0.35) +
    ggplot2::facet_grid(ontology ~ set_label, scales = "free_y",
                        space = "free_y") +
    ggplot2::scale_fill_manual(values = ontology_colors) +
    ggplot2::scale_size(range = c(2, 7)) +
    ggplot2::theme_bw() +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      strip.background = ggplot2::element_rect(fill = "#F1F4F3",
                                               color = "#C8D0CC"),
      strip.text.x = ggplot2::element_text(size = 7, face = "bold"),
      strip.text.y = ggplot2::element_text(size = 8, face = "bold", angle = 0),
      axis.text.y = ggplot2::element_text(size = 6),
      legend.position = "bottom"
    ) +
    ggplot2::labs(
      title = "eggNOG GO enrichment by ontology",
      subtitle = "Top GO terms per DEG set and ontology",
      x = "-log10 adjusted p-value",
      y = NULL,
      fill = "GO ontology",
      size = "DEGs in term"
    )

  print(p_go_ontology)
  ggplot2::ggsave(file.path(out_dir, "eggNOG_GO_enrichment_by_ontology.png"),
                  p_go_ontology, width = 13, height = 10, dpi = 300)
}
```

## KEGG pathway overview

KEGG pathways are labelled with descriptive pathway names and the pathway ID in
brackets. The names are resolved from KEGGREST when available, with an offline
fallback for common KEGG pathway IDs so the page remains reproducible. Pathways
are grouped into broad themes to keep immune, nutrition/storage, detoxification,
and core metabolic signals visually separate.

```{r kegg-overview, fig.height=11, fig.width=14}
if (nrow(kegg_enrichment_df) > 0) {
  kegg_plot_df <- kegg_enrichment_df |>
    dplyr::left_join(
      term2name_kegg_pathway |> dplyr::select(term, pathway_theme),
      by = c("ID" = "term")
    ) |>
    dplyr::mutate(
      pathway_theme = ifelse(is.na(pathway_theme),
                             "Other KEGG pathway", pathway_theme),
      set_label = factor(stringr::str_replace_all(set, "_", " "),
                         levels = stringr::str_replace_all(
                           names(enrichment_sets), "_", " "
                         )),
      neg_log10_padj = -log10(pmax(p.adjust, .Machine$double.xmin))
    ) |>
    dplyr::group_by(set_label, pathway_theme) |>
    dplyr::arrange(p.adjust, dplyr::desc(Count), .by_group = TRUE) |>
    dplyr::slice_head(n = 5) |>
    dplyr::ungroup() |>
    dplyr::mutate(
      term_label = paste0(stringr::str_trunc(Description, 48), " [", ID, "]"),
      term_label = forcats::fct_reorder(term_label, neg_log10_padj),
      pathway_theme = factor(pathway_theme, levels = names(kegg_theme_colors))
    )

  p_kegg <- ggplot2::ggplot(
    kegg_plot_df,
    ggplot2::aes(x = neg_log10_padj, y = term_label)
  ) +
    ggplot2::geom_vline(xintercept = -log10(0.05),
                        linetype = "dashed", color = "#374151",
                        linewidth = 0.35) +
    ggplot2::geom_point(ggplot2::aes(size = Count, fill = pathway_theme),
                        shape = 21, color = "#1f2937", alpha = 0.9,
                        stroke = 0.35) +
    ggplot2::facet_grid(pathway_theme ~ set_label, scales = "free_y",
                        space = "free_y") +
    ggplot2::scale_fill_manual(values = kegg_theme_colors) +
    ggplot2::scale_size(range = c(2, 7)) +
    ggplot2::guides(fill = "none") +
    ggplot2::theme_bw() +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      strip.background = ggplot2::element_rect(fill = "#F1F4F3",
                                               color = "#C8D0CC"),
      strip.text.x = ggplot2::element_text(size = 7, face = "bold"),
      strip.text.y = ggplot2::element_text(size = 7, face = "bold", angle = 0),
      axis.text.y = ggplot2::element_text(size = 6),
      legend.position = "bottom",
      plot.margin = ggplot2::margin(10, 16, 10, 16)
    ) +
    ggplot2::labs(
      title = "eggNOG KEGG pathway enrichment by biological theme",
      subtitle = "Top KEGG pathways per DEG set and theme",
      x = "-log10 adjusted p-value",
      y = "KEGG pathway",
      size = "DEGs in pathway",
      fill = "KEGG theme"
    )

  print(p_kegg)
  ggplot2::ggsave(file.path(out_dir, "eggNOG_KEGG_pathway_enrichment.png"),
                  p_kegg, width = 14, height = 11, dpi = 300)
}
```

```{r session-info}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
