---
title: "DEG sample-filter scenarios"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  metadata_file: "data/metadata/mehreen_metadata_fixed.txt"
  counts_file: "data/raw_read_counts/master_counts_no_rRNA.csv"
  min_total_count: 10
  alpha: 0.05
  lfc_threshold: 1
  output_tag: "deg-sample-filter-scenarios"
---

## Purpose

This page compares three sample-filter definitions for the same rRNA-filtered
count matrix. The goal is to make the outlier choice visible before interpreting
infection, diet, and diet-by-infection expression patterns.

The three biological questions are kept constant across scenarios:

1. Does infection change fat body expression within each diet?
2. Do diets differ among control animals?
3. Do diets differ among infected animals?

Genes are called differentially expressed when `padj < alpha` and
`abs(log2FoldChange) >= lfc_threshold`.

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c("DESeq2", "ggplot2", "dplyr", "tidyr", "ggrepel",
              "SummarizedExperiment", "DT")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

metadata_file <- file.path(project_dir, params$metadata_file)
counts_file <- file.path(project_dir, params$counts_file)

diet_colors <- c("33" = "#E6862E", "50" = "#2F8F5B", "83" = "#7B4FA3")
treatment_colors <- c("Control" = "#7B8794", "Infected" = "#C85A54")
direction_colors <- c("Upregulated" = "#E41A1C",
                      "Downregulated" = "#1E88E5")
scenario_colors <- c(
  "All samples" = "#374151",
  "Current outliers removed" = "#2F8F5B",
  "Expanded outlier set" = "#7B4FA3"
)
```

## Scenario definitions

The current outlier-removal scenario uses the four samples that are absent from
`mehreen_metadata_no_outliers.txt`: 1007, 1036, 1037, and 1039. The expanded
scenario is intentionally defined in one vector at the top of the analysis so
additional sample IDs can be added without changing the statistical code.

```{r scenario-definitions}
current_outliers <- c("1007", "1036", "1037", "1039")
expanded_outliers <- current_outliers

scenario_tbl <- tibble::tibble(
  scenario_id = c("all_samples",
                  "current_outliers_removed",
                  "expanded_outlier_set"),
  scenario = c("All samples",
               "Current outliers removed",
               "Expanded outlier set"),
  excluded_samples = I(list(character(0), current_outliers, expanded_outliers)),
  rationale = c(
    "No samples removed.",
    "Removes the current four-sample outlier set.",
    "Uses the editable expanded_outliers vector."
  )
)

scenario_display <- scenario_tbl |>
  dplyr::mutate(
    excluded_samples = vapply(excluded_samples, function(x) {
      if (length(x) == 0) "None" else paste(x, collapse = ", ")
    }, character(1))
  )

knitr::kable(scenario_display, caption = "Sample-filter scenarios")
write.csv(scenario_display, file.path(out_dir, "sample_filter_scenarios.csv"),
          row.names = FALSE)
```

## Data loading

```{r data-loading}
read_metadata <- function(path) {
  tab <- read.delim(path, check.names = FALSE)
  tab <- tab[, !is.na(names(tab)) & nzchar(names(tab)), drop = FALSE]
  tab$label <- as.character(tab$label)
  tab$Diet <- factor(tab$Diet, levels = c("33", "50", "83"))
  tab$Treatment <- relevel(factor(tab$Treatment), ref = "Control")
  tab
}

metadata_full <- read_metadata(metadata_file)

counts_raw <- read.csv(counts_file, check.names = FALSE)
counts_all <- as.matrix(counts_raw[, -1])
rownames(counts_all) <- counts_raw[[1]]
storage.mode(counts_all) <- "integer"
colnames(counts_all) <- sub("_MERGE$", "", sub("^mehreen_", "",
                                                colnames(counts_all)))

missing_metadata <- setdiff(colnames(counts_all), metadata_full$label)
if (length(missing_metadata) > 0) {
  stop("Count columns missing from metadata: ",
       paste(missing_metadata, collapse = ", "))
}
```

## Analysis functions

```{r functions}
scenario_metadata <- function(excluded_samples) {
  meta <- metadata_full |>
    dplyr::filter(!label %in% excluded_samples)
  meta <- meta[match(colnames(counts_all)[colnames(counts_all) %in% meta$label],
                     meta$label), ]
  rownames(meta) <- meta$label
  meta
}

scenario_counts <- function(metadata) {
  counts_all[, metadata$label, drop = FALSE]
}

prepare_vst <- function(counts, metadata) {
  keep <- rowSums(counts) >= params$min_total_count
  dds <- DESeq2::DESeqDataSetFromMatrix(
    countData = counts[keep, ],
    colData = metadata,
    design = ~ Diet + Treatment
  )
  DESeq2::vst(dds, blind = TRUE)
}

run_infection_within_diet <- function(counts, metadata, scenario_label) {
  out <- list()
  for (diet_level in levels(droplevels(metadata$Diet))) {
    sample_keep <- metadata$Diet == diet_level
    meta_sub <- droplevels(metadata[sample_keep, ])
    counts_sub <- counts[, sample_keep, drop = FALSE]
    keep <- rowSums(counts_sub) >= params$min_total_count

    dds <- DESeq2::DESeqDataSetFromMatrix(
      countData = counts_sub[keep, ],
      colData = meta_sub,
      design = ~ Treatment
    )
    dds <- DESeq2::DESeq(dds)
    res <- DESeq2::results(dds, contrast = c("Treatment", "Infected", "Control"))
    tab <- as.data.frame(res)
    tab$gene_id <- rownames(tab)
    tab$scenario <- scenario_label
    tab$comparison_type <- "Infection within diet"
    tab$comparison <- paste0("Diet ", diet_level, ": infected vs control")
    tab$numerator <- "Infected"
    tab$denominator <- "Control"
    tab$diet <- diet_level
    out[[diet_level]] <- tab
  }
  dplyr::bind_rows(out)
}

run_diet_pairwise <- function(counts, metadata, treatment_level, scenario_label) {
  sample_keep <- metadata$Treatment == treatment_level
  meta_sub <- droplevels(metadata[sample_keep, ])
  counts_sub <- counts[, sample_keep, drop = FALSE]
  keep <- rowSums(counts_sub) >= params$min_total_count

  dds <- DESeq2::DESeqDataSetFromMatrix(
    countData = counts_sub[keep, ],
    colData = meta_sub,
    design = ~ Diet
  )
  dds <- DESeq2::DESeq(dds)

  contrasts <- list(
    "50 vs 33" = c("Diet", "50", "33"),
    "83 vs 33" = c("Diet", "83", "33"),
    "83 vs 50" = c("Diet", "83", "50")
  )

  dplyr::bind_rows(lapply(names(contrasts), function(label) {
    contrast <- contrasts[[label]]
    res <- DESeq2::results(dds, contrast = contrast)
    tab <- as.data.frame(res)
    tab$gene_id <- rownames(tab)
    tab$scenario <- scenario_label
    tab$comparison_type <- paste("Diet contrasts among", tolower(treatment_level))
    tab$comparison <- paste(treatment_level, label)
    tab$numerator <- contrast[2]
    tab$denominator <- contrast[3]
    tab$diet <- NA_character_
    tab
  }))
}

classify_degs <- function(tab) {
  tab |>
    dplyr::mutate(
      significant = !is.na(padj) &
        padj < params$alpha &
        abs(log2FoldChange) >= params$lfc_threshold,
      direction = dplyr::case_when(
        !significant ~ "Not DEG",
        log2FoldChange > 0 ~ "Upregulated",
        log2FoldChange < 0 ~ "Downregulated",
        TRUE ~ "Not DEG"
      )
    )
}

run_scenario <- function(row_index) {
  row <- scenario_tbl[row_index, ]
  metadata <- scenario_metadata(row$excluded_samples[[1]])
  counts <- scenario_counts(metadata)
  vsd <- prepare_vst(counts, metadata)

  infection <- run_infection_within_diet(counts, metadata, row$scenario)
  diet_control <- run_diet_pairwise(counts, metadata, "Control", row$scenario)
  diet_infected <- run_diet_pairwise(counts, metadata, "Infected", row$scenario)

  list(
    scenario = row$scenario,
    metadata = metadata,
    counts = counts,
    vsd = vsd,
    results = classify_degs(dplyr::bind_rows(infection, diet_control,
                                             diet_infected))
  )
}
```

## Run DESeq2 scenarios

```{r run-scenarios}
scenario_results <- lapply(seq_len(nrow(scenario_tbl)), run_scenario)
names(scenario_results) <- scenario_tbl$scenario

all_results <- dplyr::bind_rows(lapply(scenario_results, `[[`, "results"))

sample_counts <- dplyr::bind_rows(lapply(scenario_results, function(x) {
  x$metadata |>
    dplyr::count(Diet, Treatment, name = "n_samples") |>
    dplyr::mutate(scenario = x$scenario)
}))

all_contrasts <- all_results |>
  dplyr::distinct(scenario, comparison_type, comparison)

deg_summary_observed <- all_results |>
  dplyr::filter(significant) |>
  dplyr::count(scenario, comparison_type, comparison, direction,
               name = "n_degs")

deg_summary <- tidyr::expand_grid(
  all_contrasts,
  direction = c("Upregulated", "Downregulated")
) |>
  dplyr::left_join(
    deg_summary_observed,
    by = c("scenario", "comparison_type", "comparison", "direction")
  ) |>
  dplyr::mutate(n_degs = tidyr::replace_na(n_degs, 0L))

write.csv(sample_counts, file.path(out_dir, "sample_counts_by_scenario.csv"),
          row.names = FALSE)
write.csv(all_results, file.path(out_dir, "all_deseq2_results.csv"),
          row.names = FALSE)
write.csv(deg_summary, file.path(out_dir, "deg_summary_by_scenario.csv"),
          row.names = FALSE)

knitr::kable(sample_counts, caption = "Sample counts by scenario")
```

## PCA by scenario

```{r scenario-pca, fig.height=5.6, fig.width=11}
pca_scores <- dplyr::bind_rows(lapply(scenario_results, function(x) {
  vst_mat <- SummarizedExperiment::assay(x$vsd)
  keep <- apply(vst_mat, 1, var) > 0
  pca <- prcomp(t(vst_mat[keep, , drop = FALSE]), scale. = FALSE)
  percent_var <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)
  data.frame(
    scenario = x$scenario,
    label = rownames(pca$x),
    PC1 = pca$x[, "PC1"],
    PC2 = pca$x[, "PC2"],
    Diet = x$metadata$Diet,
    Treatment = x$metadata$Treatment,
    percent_PC1 = percent_var[1],
    percent_PC2 = percent_var[2],
    stringsAsFactors = FALSE
  )
}))

p_pca <- ggplot2::ggplot(
  pca_scores,
  ggplot2::aes(PC1, PC2, color = Diet, shape = Treatment)
) +
  ggplot2::geom_point(size = 2.7, alpha = 0.86) +
  ggplot2::facet_wrap(~ scenario, scales = "free") +
  ggplot2::scale_color_manual(values = diet_colors) +
  ggplot2::scale_shape_manual(values = c("Control" = 16, "Infected" = 17)) +
  ggplot2::labs(
    title = "VST PCA by sample-filter scenario",
    x = "PC1",
    y = "PC2"
  ) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    strip.background = ggplot2::element_rect(fill = "#EEF2F4", color = "#9CA3AF"),
    strip.text = ggplot2::element_text(face = "bold"),
    panel.grid.minor = ggplot2::element_blank()
  )

p_pca
ggplot2::ggsave(file.path(out_dir, "pca_by_sample_filter_scenario.png"),
                p_pca, width = 11, height = 5.6, dpi = 300)
write.csv(pca_scores, file.path(out_dir, "pca_scores_by_scenario.csv"),
          row.names = FALSE)
```

## DEG burden across scenarios

```{r deg-burden, fig.height=8.2, fig.width=12}
deg_summary$direction <- factor(deg_summary$direction,
                                levels = c("Upregulated", "Downregulated"))

p_deg_burden <- ggplot2::ggplot(
  deg_summary,
  ggplot2::aes(x = comparison, y = n_degs, fill = direction)
) +
  ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.72),
                    width = 0.65) +
  ggplot2::facet_grid(comparison_type ~ scenario,
                      scales = "free_y", space = "free_y") +
  ggplot2::scale_fill_manual(values = direction_colors) +
  ggplot2::labs(
    title = "DEG burden by contrast and sample-filter scenario",
    x = NULL,
    y = "Number of DEGs",
    fill = "Direction"
  ) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 35, hjust = 1),
    strip.background = ggplot2::element_rect(fill = "#EEF2F4", color = "#9CA3AF"),
    strip.text = ggplot2::element_text(face = "bold"),
    panel.grid.minor = ggplot2::element_blank(),
    legend.position = "bottom"
  )

p_deg_burden
ggplot2::ggsave(file.path(out_dir, "deg_burden_by_sample_filter_scenario.png"),
                p_deg_burden, width = 12, height = 8.2, dpi = 300)
```

## DEG summary table

```{r deg-summary-table}
DT::datatable(
  deg_summary |>
    dplyr::arrange(comparison_type, comparison, scenario, direction),
  rownames = FALSE,
  filter = "top",
  options = list(pageLength = 25, scrollX = TRUE),
  caption = "DEG counts by sample-filter scenario"
)
```

## Scenario overlap table

This table records which genes are significant in each sample-filter scenario
for each contrast. It is designed to show whether a gene-level signal is stable
or appears only under one sample-filter choice.

```{r scenario-overlap}
scenario_presence <- all_results |>
  dplyr::filter(significant) |>
  dplyr::transmute(comparison_type, comparison, gene_id, scenario,
                   direction, log2FoldChange, padj) |>
  dplyr::group_by(comparison_type, comparison, gene_id) |>
  dplyr::summarise(
    scenarios_detected = paste(sort(unique(scenario)), collapse = "; "),
    n_scenarios = dplyr::n_distinct(scenario),
    directions = paste(sort(unique(direction)), collapse = "; "),
    min_padj = min(padj, na.rm = TRUE),
    max_abs_lfc = max(abs(log2FoldChange), na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::arrange(comparison_type, comparison, dplyr::desc(n_scenarios),
                 min_padj)

write.csv(scenario_presence,
          file.path(out_dir, "deg_presence_across_sample_filter_scenarios.csv"),
          row.names = FALSE)

DT::datatable(
  scenario_presence,
  rownames = FALSE,
  filter = "top",
  options = list(pageLength = 25, scrollX = TRUE),
  caption = "Significant genes detected across sample-filter scenarios"
)
```

```{r session-info}
writeLines(capture.output(sessionInfo()), file.path(out_dir, "sessionInfo.txt"))
```
