---
title: "Physiology candidate genes: all scaffolds and placed chromosomes"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  counts_file: "output/rmd_runs/transcript-exon-primary-excluding1044_20260809-233420/host_counts_no_rrna.tsv"
  all_results_file: "output/rmd_runs/transcript-exon-primary-excluding1044_20260809-233420/host_deseq2_all_contrasts.csv"
  eggnog_file: "data/GO_Annotations/GCF_023897955.1_iqSchGreg1.2_Arthopoda_one2one.emapper.annotations"
  gff_file: "data/reference/GCF_023897955.1_iqSchGreg1.2_genomic.gff"
  curated_candidate_file: "data/external/curated_target_genes/2026-08-09_Immune_Protein_Anabolism_DEGs_v24.xlsx"
  latest_results_root: "output/rmd_runs"
  alpha: 0.05
  lfc_threshold: 1
  output_tag: "physiology-candidate-genes"
---

## Purpose

This page tests the biological hypothesis in a targeted way using the 44-sample
analysis cohort. The primary set contains 115 genes labelled **Manual curation
candidate** for immune and protein-anabolism functions. A second, explicitly
labelled tier adds **eggNOG-supported candidate** genes that are absent from the manual set but
have relevant eggNOG descriptions, GO terms, KEGG annotations, or PFAM domains.
The exact supporting annotation and matched rule are retained for every
eggNOG-supported assignment.

1. Protein and amino-acid availability that could affect fungal growth.
2. Energy stores and starvation resistance in the locust fat body.
3. Immune and stress responses during infection.

This is a candidate-gene screen, not a formal enrichment test. Darker candidate
marks denote manual curation candidates and lighter marks denote
eggNOG-supported candidates.

::: {.walkthrough-grid}
::: {.walkthrough-step}
**Step 1 - Define functions**

Search insect-relevant immune, protein, lipid, carbohydrate, and detoxification terms.
:::
::: {.walkthrough-step}
**Step 2 - Join DEGs**

Match those annotations to current all-scaffold and chromosome-only contrasts.
:::
::: {.walkthrough-step}
**Step 3 - Interpret diets**

Compare how diet changes the infection response within each physiological theme.
:::
:::

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c("dplyr", "tidyr", "stringr", "ggplot2", "forcats", "DT",
              "readxl")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

counts_file <- file.path(project_dir, params$counts_file)
eggnog_file <- file.path(project_dir, params$eggnog_file)
gff_file <- file.path(project_dir, params$gff_file)
curated_candidate_file <- file.path(project_dir, params$curated_candidate_file)
results_root <- file.path(project_dir, params$latest_results_root)

theme_colors <- c(
  "Protein and amino-acid availability" = "#6F4AA8",
  "Energy stores and starvation resistance" = "#2C7FB8",
  "Immune and stress response" = "#C23B22"
)
direction_colors <- c("Upregulated" = "#E41A1C",
                      "Downregulated" = "#1E88E5")
```

## Candidate terms

These rules are used only to identify the **eggNOG-supported candidate** tier.
They do not alter membership in the 115-gene manual curation set. The matched
rule and exact eggNOG fields are retained for every eggNOG-supported gene.

```{r candidate-keywords}
candidate_terms <- tibble::tribble(
  ~theme, ~term_group, ~pattern,
  "Immune pathways", "Recognition, signalling, and effectors", paste0(
    "antimicrobial|defensin|attacin|cecropin|diptericin|gloverin|lysozyme|",
    "toll|imd|relish|myd88|cactus|dorsal|jak|stat|immune|peptidoglycan|",
    "pgrp|gram-negative|beta-1,3-glucan|glucan binding|phenoloxidase|",
    "prophenoloxidase|melanization|lectin|serpin"
  ),
  "Protein storage and synthesis", "Protein and amino-acid functions", paste0(
    "protein synthesis|translation|ribosom|elongation factor|initiation factor|",
    "aminoacyl|trna synthetase|vitellogenin|hexamerin|arylphorin|storage protein"
  ),
  "Carbohydrate storage and metabolism", "Carbohydrate storage", "glycogen|glycogenin|branching enzyme|starch",
  "Lipid metabolism", "Lipid synthesis, transport, and storage", paste0(
    "lipid|fatty acid|fatty-acid|acyl-coa|acyl coa|acetyl-coa carboxylase|",
    "elongase|desaturase|glycerolipid|phospholipid|triacylglycerol|",
    "diacylglycerol|sterol|lipase"
  ),
  "Detoxification and stress response", "Detoxification and stress", paste0(
    "detox|cytochrome p450|glutathione|gst|carboxylesterase|esterase|",
    "udp-glucuronosyltransferase|abc transporter|oxidoreductase|peroxidase|",
    "catalase|superoxide|heat shock|stress"
  )
)

knitr::kable(candidate_terms, caption = "Candidate keyword rules")
```

## Annotation table

eggNOG is protein-centered, while the count matrix uses `LOC...` gene IDs. The
GFF file bridges protein IDs to gene IDs.

```{r annotation}
read_emapper <- function(path) {
  cols <- c(
    "query", "seed_ortholog", "evalue", "score", "eggNOG_OGs",
    "max_annot_lvl", "COG_category", "Description", "Preferred_name",
    "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
    "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC", "CAZy",
    "BiGG_Reaction", "PFAMs"
  )
  tab <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", fill = TRUE)
  names(tab) <- cols[seq_len(ncol(tab))]
  tab |>
    dplyr::mutate(protein_id = stringr::str_extract(query, "XP_[0-9]+[.][0-9]+"))
}

parse_cds_gene_bridge <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  gff |>
    dplyr::filter(type == "CDS") |>
    dplyr::transmute(
      protein_id = stringr::str_match(attributes, "protein_id=([^;]+)")[, 2],
      gene_id = stringr::str_match(attributes, "gene=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(protein_id), !is.na(gene_id)) |>
    dplyr::distinct()
}

counts_raw <- if (grepl("[.]tsv$", counts_file, ignore.case = TRUE)) {
  read.delim(counts_file, check.names = FALSE)
} else {
  read.csv(counts_file, check.names = FALSE)
}
gene_universe <- unique(counts_raw[[1]])
eggnog <- read_emapper(eggnog_file)
protein_to_gene <- parse_cds_gene_bridge(gff_file)

annotation <- eggnog |>
  dplyr::left_join(protein_to_gene, by = "protein_id") |>
  dplyr::filter(!is.na(gene_id), gene_id %in% gene_universe) |>
  dplyr::distinct(gene_id, .keep_all = TRUE) |>
  dplyr::mutate(
    annotation_text = paste(Preferred_name, Description, GOs, KEGG_ko,
                            KEGG_Pathway, PFAMs, sep = " | ")
  )
```

```{r candidate-screen}
candidate_hits <- dplyr::bind_rows(lapply(seq_len(nrow(candidate_terms)), function(i) {
  term <- candidate_terms[i, ]
  hit <- stringr::str_detect(annotation$annotation_text,
                             stringr::regex(term$pattern, ignore_case = TRUE))
  if (!any(hit)) return(NULL)
  annotation[hit, ] |>
    dplyr::transmute(
      gene_id, Preferred_name, Description, PFAMs,
      GOs, KEGG_ko, KEGG_Pathway,
      theme = term$theme,
      term_group = term$term_group,
      matched_pattern = term$pattern,
      matched_keyword = stringr::str_extract(
        stringr::str_to_lower(annotation_text),
        stringr::regex(term$pattern, ignore_case = TRUE)
      )
    )
})) |>
  dplyr::distinct()

curated_candidates <- dplyr::bind_rows(
  readxl::read_excel(curated_candidate_file, sheet = "Immune"),
  readxl::read_excel(curated_candidate_file, sheet = "Protein-Anabolism")
) |>
  dplyr::transmute(
    gene_id = trimws(as.character(`Gene ID`)),
    curated_description = trimws(as.character(`Gene Description`)),
    curated_umbrella = trimws(as.character(`Umbrella Category`)),
    curated_subcategory = trimws(as.character(`Sub-Category`)),
    physiology_theme = dplyr::case_when(
      stringr::str_detect(
        curated_umbrella,
        "Immune|Humoral|Cellular"
      ) ~ "Immune pathways",
      stringr::str_detect(
        curated_umbrella,
        "Nitrogen|Protein Synthesis|Storage Hexamerins"
      ) ~ "Protein storage and synthesis",
      TRUE ~ NA_character_
    )
  ) |>
  dplyr::filter(stringr::str_starts(gene_id, "LOC")) |>
  dplyr::distinct(gene_id, .keep_all = TRUE)

if (nrow(curated_candidates) != 115) {
  stop("Expected 115 unique genes in the manual candidate workbook; found ",
       nrow(curated_candidates), ".")
}

additional_candidate_summary <- candidate_hits |>
  dplyr::filter(!gene_id %in% curated_candidates$gene_id) |>
  dplyr::group_by(
    gene_id, Preferred_name, Description, GOs, KEGG_ko,
    KEGG_Pathway, PFAMs, theme
  ) |>
  dplyr::summarise(
    term_groups = paste(sort(unique(term_group)), collapse = "; "),
    matched_patterns = paste(sort(unique(matched_pattern)), collapse = "; "),
    matched_keywords = paste(sort(unique(matched_keyword)), collapse = "; "),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    evidence_tier = "eggNOG-supported candidate",
    candidate_reason = paste0(
      "eggNOG-supported candidate: ", theme,
      "; matched keyword(s): ", matched_keywords
    )
  )

candidate_gene_summary <- dplyr::bind_rows(
  curated_candidates |>
    dplyr::transmute(
      gene_id,
      Preferred_name = NA_character_,
      Description = curated_description,
      GOs = NA_character_,
      KEGG_ko = NA_character_,
      KEGG_Pathway = NA_character_,
      PFAMs = NA_character_,
      physiology_theme,
      evidence_tier = "Manual curation candidate",
      curated_umbrella,
      curated_subcategory,
      matched_keywords = NA_character_,
      candidate_reason = paste0(
        "Manual curation candidate: ", curated_umbrella,
        " / ", curated_subcategory
      )
    ),
  additional_candidate_summary |>
    dplyr::transmute(
      gene_id, Preferred_name, Description, GOs, KEGG_ko,
      KEGG_Pathway, PFAMs,
      physiology_theme = theme,
      evidence_tier,
      curated_umbrella = NA_character_,
      curated_subcategory = NA_character_,
      matched_keywords,
      candidate_reason
    )
  ) |>
  dplyr::mutate(
    evidence_tier = factor(
      evidence_tier,
      levels = c("Manual curation candidate", "eggNOG-supported candidate")
    )
  )

write.csv(candidate_hits, file.path(out_dir, "candidate_keyword_hits_long.csv"),
          row.names = FALSE)
write.csv(candidate_gene_summary,
          file.path(out_dir, "physiological_candidate_evidence_master.csv"),
          row.names = FALSE)

knitr::kable(
  dplyr::bind_rows(
    curated_candidates |>
      dplyr::count(physiology_theme, name = "n_candidate_genes") |>
      dplyr::mutate(evidence_tier = "Manual curation candidate"),
    additional_candidate_summary |>
      dplyr::count(theme, name = "n_candidate_genes") |>
      dplyr::rename(physiology_theme = theme) |>
      dplyr::mutate(evidence_tier = "eggNOG-supported candidate")
  ) |>
    dplyr::select(evidence_tier, physiology_theme, n_candidate_genes),
  caption = "Candidate genes by evidence tier and physiological category"
)
```

## Current all-sample DEG sources

The all-scaffold and placed-chromosome models use the same 44-sample analysis cohort and the
same significance thresholds. Their gene universes differ, so their adjusted
p-values are estimated independently and should be compared as analysis
scopes, not added together.

```{r load-current-deg-results}
latest_dir <- function(pattern) {
  dirs <- list.dirs(results_root, recursive = FALSE, full.names = TRUE)
  dirs <- dirs[grepl(pattern, basename(dirs))]
  if (length(dirs) == 0) stop("No output folder matched: ", pattern)
  dirs[which.max(file.info(dirs)$mtime)]
}

all_results_file <- file.path(project_dir, params$all_results_file)
placed_dir <- latest_dir("^placed-unplaced-scaffold-analysis_")
placed_results_file <- file.path(placed_dir, "placed_unplaced_all_deseq2_results.csv")

all_scope <- read.csv(all_results_file, check.names = FALSE) |>
  dplyr::filter(stringr::str_starts(mapping_branch, "competitive-host")) |>
  dplyr::mutate(analysis_scope = "All scaffolds")

placed_scope <- read.csv(placed_results_file, check.names = FALSE) |>
  dplyr::filter(
    mapping_branch == "Competitive host",
    analysis_scope == "Placed/host-assigned",
    sequence_class == "Nuclear chromosome"
  ) |>
  dplyr::mutate(analysis_scope = "Placed chromosomes")

deg_results <- dplyr::bind_rows(all_scope, placed_scope) |>
  dplyr::mutate(
    significant = !is.na(padj) & padj < params$alpha &
      abs(log2FoldChange) >= params$lfc_threshold
  )

knitr::kable(
  data.frame(
    scope = c("All scaffolds", "Placed chromosomes"),
    result_file = c(all_results_file, placed_results_file)
  ),
  caption = "Current all-sample DEG sources"
)
```

## Evidence-tiered physiological candidates

The 115 manual curation candidates define the primary immune and
protein-anabolism set. The eggNOG-supported candidate tier extends those
hypotheses to carbohydrate, lipid, detoxification, and additional immune or
protein genes. Only the eggNOG-supported tier is generated by annotation rules.

```{r four-theme-candidate-screen}
theme_levels <- c(
  "Immune pathways", "Protein storage and synthesis",
  "Carbohydrate storage and metabolism", "Lipid metabolism",
  "Detoxification and stress response"
)
scope_levels <- c("All scaffolds", "Placed chromosomes")
evidence_levels <- c("Manual curation candidate", "eggNOG-supported candidate")
evidence_alpha <- c(
  "Manual curation candidate" = 1,
  "eggNOG-supported candidate" = 0.42
)
response_colors <- c(
  "Higher in infected" = "#B91C1C",
  "Higher in control" = "#1D4ED8"
)

candidate_degs <- deg_results |>
  dplyr::filter(
    contrast_id %in% paste0("infection_diet", c("33", "50", "83")),
    significant
  ) |>
  dplyr::left_join(candidate_gene_summary, by = "gene_id") |>
  dplyr::filter(!is.na(evidence_tier), !is.na(physiology_theme)) |>
  dplyr::mutate(
    diet = stringr::str_extract(contrast_id, "[0-9]+$"),
    direction = ifelse(log2FoldChange > 0,
                       "Higher in infected", "Higher in control"),
    physiology_theme = factor(physiology_theme, levels = theme_levels),
    analysis_scope = factor(analysis_scope, levels = scope_levels),
    evidence_tier = factor(evidence_tier, levels = evidence_levels)
  )

candidate_counts <- candidate_degs |>
  dplyr::distinct(
    analysis_scope, physiology_theme, evidence_tier, diet, direction, gene_id
  ) |>
  dplyr::count(analysis_scope, physiology_theme, evidence_tier, diet, direction,
               name = "n_candidate_degs") |>
  tidyr::complete(
    analysis_scope = factor(scope_levels, levels = scope_levels),
    physiology_theme = factor(theme_levels, levels = theme_levels),
    evidence_tier = factor(evidence_levels, levels = evidence_levels),
    diet = c("33", "50", "83"),
    direction = c("Higher in infected", "Higher in control"),
    fill = list(n_candidate_degs = 0)
  )

write.csv(candidate_degs,
          file.path(out_dir, "physiology_candidate_degs_both_scopes.csv"),
          row.names = FALSE)
write.csv(candidate_counts,
          file.path(out_dir, "physiology_candidate_counts_both_scopes.csv"),
          row.names = FALSE)

additional_deg_ids <- candidate_degs |>
  dplyr::filter(evidence_tier == "eggNOG-supported candidate") |>
  dplyr::pull(gene_id) |>
  unique()

paper_candidate_master <- candidate_gene_summary |>
  dplyr::filter(
    evidence_tier == "Manual curation candidate" |
      gene_id %in% additional_deg_ids
  ) |>
  dplyr::arrange(evidence_tier, physiology_theme, gene_id)

write.csv(
  paper_candidate_master,
  file.path(out_dir, "paper_physiological_candidate_master.csv"),
  row.names = FALSE
)
```

## Diet-specific infection-response barplots

```{r physiology-response-barplots, fig.width=16, fig.height=10}
p_candidate <- ggplot2::ggplot(
  candidate_counts,
  ggplot2::aes(
    x = diet, y = n_candidate_degs, fill = direction,
    alpha = evidence_tier,
    group = interaction(direction, evidence_tier)
  )
) +
  ggplot2::geom_col(
    position = ggplot2::position_dodge2(width = 0.82, preserve = "single"),
    width = 0.64,
    color = "white"
  ) +
  ggplot2::geom_text(
    ggplot2::aes(label = ifelse(n_candidate_degs > 0, n_candidate_degs, "")),
    position = ggplot2::position_dodge2(width = 0.82, preserve = "single"),
    vjust = -0.25,
    size = 4.2
  ) +
  ggplot2::facet_grid(analysis_scope ~ physiology_theme, scales = "free_y") +
  ggplot2::scale_fill_manual(values = response_colors) +
  ggplot2::scale_alpha_manual(values = evidence_alpha, drop = FALSE) +
  ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0.02, 0.15))) +
  ggplot2::labs(
    title = "Diet-specific infection-response DEGs in physiological candidate sets",
    subtitle = "The 44-sample analysis cohort retained; panels differ only in the tested gene universe",
    x = "Diet", y = "Number of candidate DEGs", fill = NULL,
    alpha = "Candidate evidence"
  ) +
  ggplot2::theme_bw(base_size = 14) +
  ggplot2::theme(
    strip.text = ggplot2::element_text(face = "bold", size = 12),
    legend.position = "bottom",
    panel.grid.minor = ggplot2::element_blank()
  )

p_candidate
ggplot2::ggsave(
  file.path(out_dir, "physiology_response_counts_all_vs_placed.png"),
  p_candidate, width = 16, height = 10, dpi = 300, bg = "white"
)
ggplot2::ggsave(
  file.path(out_dir, "physiology_response_counts_all_vs_placed.svg"),
  p_candidate, width = 16, height = 10, bg = "white"
)
```

## Diet 50 versus diet 83

This contrast of counts asks whether each physiology class contains more
infection-responsive genes under diet 50 or diet 83. It is descriptive and does
not replace the gene-level diet-by-infection interaction test.

```{r diet50-vs-diet83-barplot, fig.width=12, fig.height=7}
diet_50_83 <- candidate_counts |>
  dplyr::filter(diet %in% c("50", "83")) |>
  tidyr::pivot_wider(
    names_from = diet,
    values_from = n_candidate_degs,
    values_fill = 0
  ) |>
  dplyr::mutate(diet50_minus_diet83 = `50` - `83`)

p_delta <- ggplot2::ggplot(
  diet_50_83,
  ggplot2::aes(
    x = physiology_theme, y = diet50_minus_diet83, fill = direction,
    alpha = evidence_tier,
    group = interaction(direction, evidence_tier)
  )
) +
  ggplot2::geom_hline(yintercept = 0, color = "#374151", linewidth = 0.7) +
  ggplot2::geom_col(
    position = ggplot2::position_dodge2(width = 0.82, preserve = "single"),
    width = 0.68
  ) +
  ggplot2::facet_wrap(~ analysis_scope, ncol = 1) +
  ggplot2::scale_fill_manual(values = response_colors) +
  ggplot2::scale_alpha_manual(values = evidence_alpha, drop = FALSE) +
  ggplot2::labs(
    title = "Difference in infection-response DEG counts: diet 50 minus diet 83",
    subtitle = "Positive bars indicate more candidate DEGs in diet 50; negative bars indicate more in diet 83",
    x = NULL, y = "Difference in DEG count", fill = NULL,
    alpha = "Candidate evidence"
  ) +
  ggplot2::theme_bw(base_size = 14) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 20, hjust = 1),
    strip.text = ggplot2::element_text(face = "bold"),
    legend.position = "bottom",
    panel.grid.minor = ggplot2::element_blank()
  )

p_delta
ggplot2::ggsave(
  file.path(out_dir, "diet50_vs_diet83_physiology_deg_count_difference.png"),
  p_delta, width = 12, height = 7, dpi = 300, bg = "white"
)
ggplot2::ggsave(
  file.path(out_dir, "diet50_vs_diet83_physiology_deg_count_difference.svg"),
  p_delta, width = 12, height = 7, bg = "white"
)
```

## Searchable candidate DEG table

```{r candidate-deg-table}
DT::datatable(
  candidate_degs |>
    dplyr::select(
      analysis_scope, physiology_theme, diet, direction, gene_id,
      evidence_tier, candidate_reason, Description, Preferred_name,
      matched_keywords, GOs, KEGG_ko, KEGG_Pathway, PFAMs,
      log2FoldChange, padj
    ) |>
    dplyr::arrange(analysis_scope, physiology_theme, diet, padj),
  rownames = FALSE,
  filter = "top",
  options = list(pageLength = 25, scrollX = TRUE),
  caption = "Candidate physiology genes significant within at least one diet"
)
```

```{r session-info}
writeLines(capture.output(sessionInfo()), file.path(out_dir, "sessionInfo.txt"))
```
