---
title: "Physiology candidate genes: all scaffolds and placed chromosomes"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  counts_file: "output/rmd_runs/transcript-exon-primary-deseq2_20260806-004620/host_counts_no_rrna.tsv"
  all_results_file: "output/rmd_runs/transcript-exon-primary-deseq2_20260806-004620/host_deseq2_all_contrasts.csv"
  eggnog_file: "data/GO_Annotations/GCF_023897955.1_iqSchGreg1.2_Arthopoda_one2one.emapper.annotations"
  gff_file: "data/reference/GCF_023897955.1_iqSchGreg1.2_genomic.gff"
  latest_results_root: "output/rmd_runs"
  alpha: 0.05
  lfc_threshold: 1
  output_tag: "physiology-candidate-genes"
---

## Purpose

This page tests the biological hypothesis in a targeted way using all 44 samples.
Instead of starting
from broad GO or KEGG enrichment labels, it searches the eggNOG annotation for
genes plausibly involved in insect-relevant immune and fat-body processes. The
same summaries are shown for the all-scaffold and placed-chromosome gene
universes so the effect of assembly placement remains visible.

1. Protein and amino-acid availability that could affect fungal growth.
2. Energy stores and starvation resistance in the locust fat body.
3. Immune and stress responses during infection.

This is a candidate-gene screen, not a formal enrichment test. It is designed to
produce an interpretable gene list that can be checked manually and refined as
we agree on more insect-specific marker genes.

::: {.walkthrough-grid}
::: {.walkthrough-step}
**Step 1 - Define functions**

Search insect-relevant immune, protein, lipid, carbohydrate, and detoxification terms.
:::
::: {.walkthrough-step}
**Step 2 - Join DEGs**

Match those annotations to current all-scaffold and chromosome-only contrasts.
:::
::: {.walkthrough-step}
**Step 3 - Interpret diets**

Compare how diet changes the infection response within each physiological theme.
:::
:::

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c("dplyr", "tidyr", "stringr", "ggplot2", "forcats", "DT")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

counts_file <- file.path(project_dir, params$counts_file)
eggnog_file <- file.path(project_dir, params$eggnog_file)
gff_file <- file.path(project_dir, params$gff_file)
results_root <- file.path(project_dir, params$latest_results_root)

theme_colors <- c(
  "Protein and amino-acid availability" = "#6F4AA8",
  "Energy stores and starvation resistance" = "#2C7FB8",
  "Immune and stress response" = "#C23B22"
)
direction_colors <- c("Upregulated" = "#E41A1C",
                      "Downregulated" = "#1E88E5")
```

## Candidate terms

These keyword groups deliberately favor insect physiology terms and fat-body
functions over broad pathway labels. The matched word is retained for each gene
so the screen remains auditable.

```{r candidate-keywords}
candidate_terms <- tibble::tribble(
  ~theme, ~term_group, ~pattern,
  "Protein and amino-acid availability", "Storage proteins", "storage protein|hexamerin|arylphorin|vitellogenin|yolk protein",
  "Protein and amino-acid availability", "Proteolysis", "protease|peptidase|trypsin|cathepsin|carboxypeptidase|metalloprotease|serine protease",
  "Protein and amino-acid availability", "Amino-acid transport/metabolism", "amino acid|amino-acid|peptide transport|solute carrier|transaminase|aminotransferase",
  "Energy stores and starvation resistance", "Lipid transport/storage", "lipophorin|apolipophorin|lipid|fatty acid|triacylglycerol|triglyceride|acyl-coa|beta-oxidation|lipase",
  "Energy stores and starvation resistance", "Carbohydrate stores", "glycogen|trehalose|trehalase|glucose|glycolysis|gluconeogenesis|pentose phosphate|citrate cycle|tca cycle",
  "Energy stores and starvation resistance", "Endocrine nutrition", "insulin|insulin-like|adipokinetic|juvenile hormone|tor signaling|target of rapamycin",
  "Immune and stress response", "Pattern recognition", "peptidoglycan recognition|pgrp|gram-negative binding|gnbp|beta-1,3-glucan|lectin|fibrinogen|scavenger receptor",
  "Immune and stress response", "Toll/IMD/JAK-STAT", "toll|imd|relish|cactus|jak|stat|dorsal|myd88|tube|pell[e]?",
  "Immune and stress response", "Effector and melanization", "defensin|cecropin|attacin|gloverin|lysozyme|antimicrobial|phenoloxidase|prophenoloxidase|melanization|serpin|clip-domain",
  "Immune and stress response", "Stress/detoxification", "heat shock|glutathione|peroxidase|cytochrome p450|detoxification|oxidative stress|reactive oxygen|autophagy|thioester"
)

knitr::kable(candidate_terms, caption = "Candidate keyword rules")
```

## Annotation table

eggNOG is protein-centered, while the count matrix uses `LOC...` gene IDs. The
GFF file bridges protein IDs to gene IDs.

```{r annotation}
read_emapper <- function(path) {
  cols <- c(
    "query", "seed_ortholog", "evalue", "score", "eggNOG_OGs",
    "max_annot_lvl", "COG_category", "Description", "Preferred_name",
    "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
    "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC", "CAZy",
    "BiGG_Reaction", "PFAMs"
  )
  tab <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", fill = TRUE)
  names(tab) <- cols[seq_len(ncol(tab))]
  tab |>
    dplyr::mutate(protein_id = stringr::str_extract(query, "XP_[0-9]+[.][0-9]+"))
}

parse_cds_gene_bridge <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  gff |>
    dplyr::filter(type == "CDS") |>
    dplyr::transmute(
      protein_id = stringr::str_match(attributes, "protein_id=([^;]+)")[, 2],
      gene_id = stringr::str_match(attributes, "gene=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(protein_id), !is.na(gene_id)) |>
    dplyr::distinct()
}

counts_raw <- if (grepl("[.]tsv$", counts_file, ignore.case = TRUE)) {
  read.delim(counts_file, check.names = FALSE)
} else {
  read.csv(counts_file, check.names = FALSE)
}
gene_universe <- unique(counts_raw[[1]])
eggnog <- read_emapper(eggnog_file)
protein_to_gene <- parse_cds_gene_bridge(gff_file)

annotation <- eggnog |>
  dplyr::left_join(protein_to_gene, by = "protein_id") |>
  dplyr::filter(!is.na(gene_id), gene_id %in% gene_universe) |>
  dplyr::distinct(gene_id, .keep_all = TRUE) |>
  dplyr::mutate(
    annotation_text = paste(Preferred_name, Description, GOs, KEGG_ko,
                            KEGG_Pathway, PFAMs, sep = " | ")
  )
```

```{r candidate-screen}
candidate_hits <- dplyr::bind_rows(lapply(seq_len(nrow(candidate_terms)), function(i) {
  term <- candidate_terms[i, ]
  hit <- stringr::str_detect(annotation$annotation_text,
                             stringr::regex(term$pattern, ignore_case = TRUE))
  if (!any(hit)) return(NULL)
  annotation[hit, ] |>
    dplyr::transmute(
      gene_id, Preferred_name, Description, PFAMs,
      theme = term$theme,
      term_group = term$term_group,
      matched_pattern = term$pattern
    )
})) |>
  dplyr::distinct()

candidate_gene_summary <- candidate_hits |>
  dplyr::group_by(gene_id, Preferred_name, Description) |>
  dplyr::summarise(
    themes = paste(sort(unique(theme)), collapse = "; "),
    term_groups = paste(sort(unique(term_group)), collapse = "; "),
    matched_patterns = paste(sort(unique(matched_pattern)), collapse = "; "),
    .groups = "drop"
  )

write.csv(candidate_hits, file.path(out_dir, "candidate_keyword_hits_long.csv"),
          row.names = FALSE)
write.csv(candidate_gene_summary,
          file.path(out_dir, "candidate_gene_annotation_summary.csv"),
          row.names = FALSE)

knitr::kable(
  candidate_hits |>
    dplyr::count(theme, term_group, name = "n_candidate_genes"),
  caption = "Candidate genes detected by keyword group"
)
```

## Current all-sample DEG sources

The all-scaffold and placed-chromosome models use the same 44 samples and the
same significance thresholds. Their gene universes differ, so their adjusted
p-values are estimated independently and should be compared as analysis
scopes, not added together.

```{r load-current-deg-results}
latest_dir <- function(pattern) {
  dirs <- list.dirs(results_root, recursive = FALSE, full.names = TRUE)
  dirs <- dirs[grepl(pattern, basename(dirs))]
  if (length(dirs) == 0) stop("No output folder matched: ", pattern)
  dirs[which.max(file.info(dirs)$mtime)]
}

all_results_file <- file.path(project_dir, params$all_results_file)
placed_dir <- latest_dir("^placed-unplaced-scaffold-analysis_")
placed_results_file <- file.path(placed_dir, "placed_unplaced_all_deseq2_results.csv")

all_scope <- read.csv(all_results_file, check.names = FALSE) |>
  dplyr::filter(mapping_branch == "competitive-host") |>
  dplyr::mutate(analysis_scope = "All scaffolds")

placed_scope <- read.csv(placed_results_file, check.names = FALSE) |>
  dplyr::filter(
    mapping_branch == "Competitive host",
    analysis_scope == "Placed/host-assigned",
    sequence_class == "Nuclear chromosome"
  ) |>
  dplyr::mutate(analysis_scope = "Placed chromosomes")

deg_results <- dplyr::bind_rows(all_scope, placed_scope) |>
  dplyr::mutate(
    significant = !is.na(padj) & padj < params$alpha &
      abs(log2FoldChange) >= params$lfc_threshold
  )

knitr::kable(
  data.frame(
    scope = c("All scaffolds", "Placed chromosomes"),
    result_file = c(all_results_file, placed_results_file)
  ),
  caption = "Current all-sample DEG sources"
)
```

## Four physiology themes

These four themes reflect the current biological hypotheses for fat body:
immune pathways, protein storage and synthesis, lipid metabolism, and
detoxification. Classification is based on local GFF/eggNOG description and
domain text and is therefore a transparent candidate screen rather than a
formal ontology.

```{r four-theme-candidate-screen}
classify_physiology_theme <- function(x) {
  x <- stringr::str_to_lower(dplyr::coalesce(x, ""))
  dplyr::case_when(
    stringr::str_detect(x, paste(
      "antimicrobial|defensin|attacin|cecropin|diptericin|gloverin|lysozyme|",
      "toll|imd|relish|myd88|cactus|dorsal|jak|stat|immune|peptidoglycan|",
      "pgrp|glucan|phenoloxidase|melanization|lectin|serpin", sep = ""
    )) ~ "Immune pathway",
    stringr::str_detect(x, paste(
      "protein synthesis|translation|ribosom|elongation factor|initiation factor|",
      "aminoacyl|trna synthetase|vitellogenin|hexamerin|arylphorin|storage protein",
      sep = ""
    )) ~ "Protein storage and synthesis",
    stringr::str_detect(x, paste(
      "lipid|fatty acid|acyl-coa|acetyl-coa carboxylase|elongase|desaturase|",
      "glycerolipid|phospholipid|triacylglycerol|sterol|lipase", sep = ""
    )) ~ "Lipid metabolism",
    stringr::str_detect(x, paste(
      "detox|cytochrome p450|glutathione|carboxylesterase|esterase|",
      "udp-glucuronosyltransferase|abc transporter|oxidoreductase|peroxidase|",
      "catalase|superoxide|heat shock|stress", sep = ""
    )) ~ "Detoxification",
    TRUE ~ NA_character_
  )
}

physiology_annotation <- annotation |>
  dplyr::transmute(
    gene_id,
    Description,
    Preferred_name,
    PFAMs,
    annotation_text,
    physiology_theme = classify_physiology_theme(annotation_text)
  ) |>
  dplyr::filter(!is.na(physiology_theme))

theme_levels <- c(
  "Immune pathway", "Protein storage and synthesis",
  "Lipid metabolism", "Detoxification"
)
scope_levels <- c("All scaffolds", "Placed chromosomes")
response_colors <- c(
  "Higher in infected" = "#B91C1C",
  "Higher in control" = "#1D4ED8"
)

candidate_degs <- deg_results |>
  dplyr::filter(
    contrast_id %in% paste0("infection_diet", c("33", "50", "83")),
    significant
  ) |>
  dplyr::left_join(physiology_annotation, by = "gene_id") |>
  dplyr::filter(!is.na(physiology_theme)) |>
  dplyr::mutate(
    diet = stringr::str_extract(contrast_id, "[0-9]+$"),
    direction = ifelse(log2FoldChange > 0,
                       "Higher in infected", "Higher in control"),
    physiology_theme = factor(physiology_theme, levels = theme_levels),
    analysis_scope = factor(analysis_scope, levels = scope_levels)
  )

candidate_counts <- candidate_degs |>
  dplyr::distinct(analysis_scope, physiology_theme, diet, direction, gene_id) |>
  dplyr::count(analysis_scope, physiology_theme, diet, direction,
               name = "n_candidate_degs") |>
  tidyr::complete(
    analysis_scope = factor(scope_levels, levels = scope_levels),
    physiology_theme = factor(theme_levels, levels = theme_levels),
    diet = c("33", "50", "83"),
    direction = c("Higher in infected", "Higher in control"),
    fill = list(n_candidate_degs = 0)
  )

write.csv(candidate_degs,
          file.path(out_dir, "physiology_candidate_degs_both_scopes.csv"),
          row.names = FALSE)
write.csv(candidate_counts,
          file.path(out_dir, "physiology_candidate_counts_both_scopes.csv"),
          row.names = FALSE)
```

## Diet-specific infection-response barplots

```{r physiology-response-barplots, fig.width=16, fig.height=10}
p_candidate <- ggplot2::ggplot(
  candidate_counts,
  ggplot2::aes(x = diet, y = n_candidate_degs, fill = direction)
) +
  ggplot2::geom_col(
    position = ggplot2::position_dodge(width = 0.74),
    width = 0.64,
    color = "white"
  ) +
  ggplot2::geom_text(
    ggplot2::aes(label = ifelse(n_candidate_degs > 0, n_candidate_degs, "")),
    position = ggplot2::position_dodge(width = 0.74),
    vjust = -0.25,
    size = 4.2
  ) +
  ggplot2::facet_grid(analysis_scope ~ physiology_theme, scales = "free_y") +
  ggplot2::scale_fill_manual(values = response_colors) +
  ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0.02, 0.15))) +
  ggplot2::labs(
    title = "Diet-specific infection-response DEGs in four physiology themes",
    subtitle = "All 44 samples retained; panels differ only in the tested gene universe",
    x = "Diet", y = "Number of candidate DEGs", fill = NULL
  ) +
  ggplot2::theme_bw(base_size = 14) +
  ggplot2::theme(
    strip.text = ggplot2::element_text(face = "bold", size = 12),
    legend.position = "bottom",
    panel.grid.minor = ggplot2::element_blank()
  )

p_candidate
ggplot2::ggsave(
  file.path(out_dir, "physiology_response_counts_all_vs_placed.png"),
  p_candidate, width = 16, height = 10, dpi = 300, bg = "white"
)
ggplot2::ggsave(
  file.path(out_dir, "physiology_response_counts_all_vs_placed.svg"),
  p_candidate, width = 16, height = 10, bg = "white"
)
```

## Diet 50 versus diet 83

This contrast of counts asks whether each physiology class contains more
infection-responsive genes under diet 50 or diet 83. It is descriptive and does
not replace the gene-level diet-by-infection interaction test.

```{r diet50-vs-diet83-barplot, fig.width=12, fig.height=7}
diet_50_83 <- candidate_counts |>
  dplyr::filter(diet %in% c("50", "83")) |>
  tidyr::pivot_wider(
    names_from = diet,
    values_from = n_candidate_degs,
    values_fill = 0
  ) |>
  dplyr::mutate(diet50_minus_diet83 = `50` - `83`)

p_delta <- ggplot2::ggplot(
  diet_50_83,
  ggplot2::aes(x = physiology_theme, y = diet50_minus_diet83, fill = direction)
) +
  ggplot2::geom_hline(yintercept = 0, color = "#374151", linewidth = 0.7) +
  ggplot2::geom_col(position = "dodge", width = 0.68) +
  ggplot2::facet_wrap(~ analysis_scope, ncol = 1) +
  ggplot2::scale_fill_manual(values = response_colors) +
  ggplot2::labs(
    title = "Difference in infection-response DEG counts: diet 50 minus diet 83",
    subtitle = "Positive bars indicate more candidate DEGs in diet 50; negative bars indicate more in diet 83",
    x = NULL, y = "Difference in DEG count", fill = NULL
  ) +
  ggplot2::theme_bw(base_size = 14) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 20, hjust = 1),
    strip.text = ggplot2::element_text(face = "bold"),
    legend.position = "bottom",
    panel.grid.minor = ggplot2::element_blank()
  )

p_delta
ggplot2::ggsave(
  file.path(out_dir, "diet50_vs_diet83_physiology_deg_count_difference.png"),
  p_delta, width = 12, height = 7, dpi = 300, bg = "white"
)
ggplot2::ggsave(
  file.path(out_dir, "diet50_vs_diet83_physiology_deg_count_difference.svg"),
  p_delta, width = 12, height = 7, bg = "white"
)
```

## Searchable candidate DEG table

```{r candidate-deg-table}
DT::datatable(
  candidate_degs |>
    dplyr::select(
      analysis_scope, physiology_theme, diet, direction, gene_id,
      Description, log2FoldChange, padj
    ) |>
    dplyr::arrange(analysis_scope, physiology_theme, diet, padj),
  rownames = FALSE,
  filter = "top",
  options = list(pageLength = 25, scrollX = TRUE),
  caption = "Candidate physiology genes significant within at least one diet"
)
```

```{r session-info}
writeLines(capture.output(sessionInfo()), file.path(out_dir, "sessionInfo.txt"))
```
