---
title: "Main-chromosome diet pairwise DEGs - all samples"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  metadata_file: "data/metadata/mehreen_metadata_fixed.txt"
  counts_file: "output/rmd_runs/transcript-exon-primary-deseq2_20260806-004620/host_counts_no_rrna.tsv"
  placement_file: "output/runs/host_pathogen_dual_20260727_000952/local-analysis/host-mapping-audit_20260729-163928/unified-model/audit-summary/host_gene_sequence_placement.tsv"
  rrna_file: "data/excluded_loci/gregaria_rrna_list.txt"
  treatment_subset: "All"
  min_count: 10
  min_samples: 3
  alpha: 0.05
  lfc_threshold: 1
  output_tag: "main-chromosome-all-samples-diet-pairwise-deg"
---

## Purpose

::: {.walkthrough-grid}
::: {.walkthrough-step}
**Step 1 - Restrict placement**

Use only main-chromosome non-rRNA genes while retaining all samples.
:::
::: {.walkthrough-step}
**Step 2 - Test diet pairs**

Compare diets separately among control and infected locusts.
:::
::: {.walkthrough-step}
**Step 3 - Interpret context**

Distinguish baseline diet effects from infection-dependent diet differences.
:::
:::

This is the all-samples, main-chromosome version of the diet pairwise DEG page. It keeps every
library in `mehreen_metadata_fixed.txt`, including the samples currently treated
as outliers in sensitivity analyses.

All tests use only non-rRNA genes assigned to nuclear chromosomes by the
official assembly-placement table. Unplaced scaffolds are retained only in the
separate origin-audit and sensitivity pages.

This page tests pairwise diet contrasts. Set `treatment_subset` to `Control`,
`Infected`, or `All` in the YAML header before knitting.

The aim is to separate the diet question from the infection question as much as
the design allows. When all samples are used, the model includes treatment as a
covariate so diet contrasts are not simply rediscovering the infected-control
difference. When one treatment is selected, the model asks which genes differ
among diets within that treatment subset only.

## Setup and data loading

The metadata and count matrix are matched by sample label before modelling.
This keeps the analysis reproducible and prevents silent sample-order mistakes,
which are especially easy to make when count columns carry sequencing-style
prefixes and suffixes.

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c("DESeq2", "ggplot2", "EnhancedVolcano", "readr", "dplyr")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}
library(ggplot2)

direction_colors <- c("Higher in first diet" = "#237C8E",
                      "Higher in second diet" = "#B8643C")

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, params$treatment_subset,
                           run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

metadata <- read.delim(file.path(project_dir, params$metadata_file),
                       check.names = FALSE)
metadata$label <- as.character(metadata$label)
metadata$Diet <- factor(metadata$Diet, levels = c("33", "50", "83"))
metadata$Treatment <- factor(metadata$Treatment, levels = c("Control", "Infected"))

counts_table <- readr::read_tsv(file.path(project_dir, params$counts_file),
                                show_col_types = FALSE)
stopifnot("gene_id" %in% names(counts_table))
counts <- as.matrix(counts_table[, setdiff(names(counts_table), "gene_id")])
rownames(counts) <- counts_table$gene_id
storage.mode(counts) <- "integer"

count_labels_all <- sub("_MERGE$", "", sub("^mehreen_", "", colnames(counts)))
keep_samples <- count_labels_all %in% metadata$label
counts <- counts[, keep_samples, drop = FALSE]
count_labels <- count_labels_all[keep_samples]
metadata <- metadata[match(count_labels, metadata$label), ]
stopifnot(identical(metadata$label, count_labels))
rownames(metadata) <- metadata$label
colnames(counts) <- count_labels

placement <- readr::read_tsv(file.path(project_dir, params$placement_file),
                             show_col_types = FALSE) |>
  dplyr::distinct(gene_id, .keep_all = TRUE)
rrna_ids <- trimws(readLines(file.path(project_dir, params$rrna_file), warn = FALSE))
rrna_ids <- rrna_ids[nzchar(rrna_ids) & !startsWith(rrna_ids, "#")]
main_chromosome_ids <- placement$gene_id[
  placement$sequence_class == "Nuclear chromosome" &
    !placement$gene_id %in% rrna_ids
]
counts <- counts[intersect(rownames(counts), main_chromosome_ids), , drop = FALSE]

if (params$treatment_subset != "All") {
  keep_samples <- metadata$Treatment == params$treatment_subset
  counts <- counts[, keep_samples]
  metadata <- droplevels(metadata[keep_samples, ])
}

keep_genes <- rowSums(counts >= params$min_count) >= params$min_samples
input_audit <- data.frame(
  treatment_subset = params$treatment_subset,
  samples_retained = ncol(counts),
  raw_count_rows = nrow(counts_table),
  main_chromosome_non_rRNA_rows = nrow(counts),
  genes_entering_DESeq2 = sum(keep_genes)
)
knitr::kable(input_audit, caption = "Main-chromosome analysis input audit")
write.csv(input_audit, file.path(out_dir, "main_chromosome_input_audit.csv"),
          row.names = FALSE)
```

## Fit model

DESeq2 models raw counts with library-size normalization and gene-wise
dispersion estimates. The `All` analysis uses `~ Treatment + Diet`, so each diet
coefficient is estimated after accounting for infection status. This gives a
cleaner diet comparison than pooling control and infected samples without a
treatment term.

```{r fit-model}
design_formula <- if (params$treatment_subset == "All") {
  ~ Treatment + Diet
} else {
  ~ Diet
}

dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = counts[keep_genes, ],
  colData = metadata,
  design = design_formula
)

dds <- DESeq2::DESeq(dds, minReplicatesForReplace = Inf, quiet = TRUE)
```

## Pairwise diet contrasts

Each contrast compares two diet levels at a time. A positive log2 fold-change
means the gene is higher in the first diet named in the contrast; a negative
log2 fold-change means it is higher in the second diet. The volcano plots show
both effect size and adjusted p-value, while the CSV files keep the full table
for downstream overlap and enrichment analyses.

```{r pairwise-contrasts, fig.height=7, fig.width=8}
diet_pairs <- combn(levels(droplevels(metadata$Diet)), 2, simplify = FALSE)

contrast_results <- list()
for (pair in diet_pairs) {
  contrast_label <- paste(pair[1], "vs", pair[2], sep = "_")
  message("Running diet contrast: ", contrast_label)

  res <- DESeq2::results(dds, contrast = c("Diet", pair[1], pair[2]),
                         alpha = params$alpha, cooksCutoff = FALSE)
  res_df <- as.data.frame(res)
  res_df$gene_id <- rownames(res_df)
  res_df$contrast <- contrast_label
  res_df$significant <- !is.na(res_df$padj) &
    res_df$padj < params$alpha &
    abs(res_df$log2FoldChange) >= params$lfc_threshold

  contrast_results[[contrast_label]] <- res_df
  write.csv(res_df, file.path(out_dir, paste0("DESeq2_", contrast_label, ".csv")),
            row.names = FALSE)
  writeLines(res_df$gene_id[res_df$significant],
             file.path(out_dir, paste0("significant_genes_", contrast_label, ".txt")))

  print(EnhancedVolcano::EnhancedVolcano(
    res_df,
    lab = res_df$gene_id,
    x = "log2FoldChange",
    y = "padj",
    pCutoff = params$alpha,
    FCcutoff = params$lfc_threshold,
    title = paste("Diet", pair[1], "vs", pair[2]),
    subtitle = paste("Treatment subset:", params$treatment_subset)
  ))
}

all_pairwise <- do.call(rbind, contrast_results)
write.csv(all_pairwise, file.path(out_dir, "all_diet_pairwise_results.csv"),
          row.names = FALSE)
```

## DEG count barplot

Volcano plots are useful for seeing the distribution of effect sizes, but they
are not the fastest way to compare how many genes pass the current thresholds in
each contrast. This barplot reduces each contrast to the number of significant
DEGs and separates the direction of change.

```{r deg-count-barplot, fig.height=4.8, fig.width=7.5}
sig_pairwise <- all_pairwise[all_pairwise$significant &
                               !is.na(all_pairwise$log2FoldChange), ]
sig_pairwise$direction <- ifelse(sig_pairwise$log2FoldChange > 0,
                                 "Higher in first diet",
                                 "Higher in second diet")
sig_pairwise$contrast <- factor(sig_pairwise$contrast,
                                levels = names(contrast_results))
sig_pairwise$direction <- factor(sig_pairwise$direction,
                                 levels = c("Higher in first diet",
                                            "Higher in second diet"))

deg_bar_counts <- as.data.frame(table(
  contrast = sig_pairwise$contrast,
  direction = sig_pairwise$direction
))
names(deg_bar_counts)[3] <- "n_significant_genes"

p_deg_counts <- ggplot(deg_bar_counts,
                       aes(x = contrast, y = n_significant_genes,
                           fill = direction)) +
  geom_col(position = position_dodge(width = 0.78), width = 0.68) +
  geom_text(aes(label = n_significant_genes),
            position = position_dodge(width = 0.78),
            vjust = -0.35, size = 3.2) +
  scale_fill_manual(values = direction_colors) +
  theme_classic() +
  labs(
    title = "Significant diet DEGs by contrast and direction",
    subtitle = paste("Treatment subset:", params$treatment_subset),
    x = "Diet contrast",
    y = "Number of significant genes",
    fill = "Direction"
  )

p_deg_counts
ggsave(file.path(out_dir, "diet_pairwise_deg_count_barplot.png"),
       p_deg_counts, width = 7.5, height = 4.8, dpi = 300)
write.csv(deg_bar_counts, file.path(out_dir, "diet_pairwise_direction_counts.csv"),
          row.names = FALSE)
```

## Summary

The summary table reports the total number of genes that pass the adjusted
p-value and log2 fold-change thresholds for each diet contrast. It is saved as a
small machine-readable table because these counts are used again by the
publication-figure and enrichment pages.

```{r summary}
summary_table <- aggregate(significant ~ contrast, all_pairwise, sum)
names(summary_table)[2] <- "n_significant_genes"
knitr::kable(summary_table)

write.csv(summary_table, file.path(out_dir, "diet_pairwise_summary.csv"),
          row.names = FALSE)
```

```{r session-info}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
