---
title: "Main-chromosome infection within diet DEGs - all samples"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  metadata_file: "data/metadata/mehreen_metadata_corrected_45.txt"
  counts_file: "output/rmd_runs/transcript-exon-primary-corrected45_20260809-153146/host_counts_no_rrna.tsv"
  placement_file: "output/runs/host_pathogen_dual_20260727_000952/local-analysis/host-mapping-audit_20260729-163928/unified-model/audit-summary/host_gene_sequence_placement.tsv"
  rrna_file: "data/excluded_loci/gregaria_rrna_list.txt"
  eggnog_file: "data/GO_Annotations/GCF_023897955.1_iqSchGreg1.2_Arthopoda_one2one.emapper.annotations"
  gff_file: "data/reference/GCF_023897955.1_iqSchGreg1.2_genomic.gff"
  min_count: 10
  min_samples: 3
  top_variable_genes: 500
  top_curated_degs_per_direction: 12
  top_functional_degs_per_theme: 18
  alpha: 0.05
  lfc_threshold: 1
  output_tag: "main-chromosome-all-samples-infection-within-diet-deg"
---

## Purpose

::: {.walkthrough-grid}
::: {.walkthrough-step}
**Step 1 - Restrict placement**

Use main-chromosome non-rRNA genes and retain all 45 fat body samples.
:::
::: {.walkthrough-step}
**Step 2 - Test infection**

Compare infected and control samples within each diet.
:::
::: {.walkthrough-step}
**Step 3 - Interpret response**

Read PCA, DEG burden, heatmaps, and functional candidate summaries together.
:::
:::

This is the all-samples, main-chromosome version of the infection-within-diet
DEG page. It keeps
every library in `mehreen_metadata_corrected_45.txt`, including the samples currently
treated as outliers in sensitivity analyses.

Every plot and test uses only non-rRNA genes assigned to nuclear chromosomes in
the official assembly-placement table. Unplaced scaffolds are not allowed to
enter normalization, VST, PCA, heatmaps, or the DEG hypothesis family here.

This page tests infected versus control samples separately within each diet.
This is a simple and readable complement to the full interaction model.

The reason to split by diet is biological and statistical: infection may not
produce the same transcriptional response under every nutritional condition.
The full interaction model tests whether diet modifies the infection effect;
this page then turns that question into three direct infected-control
comparisons that are easier to inspect, plot, and interpret.

## Setup and data loading

The code first matches metadata rows to count-matrix columns using the sample
labels. This makes the analysis self-contained in the R Markdown file and keeps
the model inputs visible before any statistics are run.

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c("DESeq2", "ggplot2", "EnhancedVolcano", "pheatmap", "ggrepel",
              "dplyr", "SummarizedExperiment", "DT", "readr")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}
library(ggplot2)

diet_colors <- c("33" = "#E6862E", "50" = "#2F8F5B", "83" = "#7B4FA3")
treatment_colors <- c("Control" = "#7B8794", "Infected" = "#C85A54")
treatment_fills <- c("Control" = "#D8DEE2", "Infected" = "#F0C7BF")
condition_surface_colors <- c(
  "33 Control" = "#F7D6A6",
  "33 Infected" = "#C75C1B",
  "50 Control" = "#CDE8D5",
  "50 Infected" = "#1F6F43",
  "83 Control" = "#D8C7E8",
  "83 Infected" = "#5B2E83"
)

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

metadata <- read.delim(file.path(project_dir, params$metadata_file),
                       check.names = FALSE)
metadata$label <- as.character(metadata$label)
metadata$Diet <- factor(metadata$Diet, levels = c("33", "50", "83"))
metadata$Treatment <- relevel(factor(metadata$Treatment), ref = "Control")

counts_table <- readr::read_tsv(file.path(project_dir, params$counts_file),
                                show_col_types = FALSE)
stopifnot("gene_id" %in% names(counts_table))
counts <- as.matrix(counts_table[, setdiff(names(counts_table), "gene_id")])
rownames(counts) <- counts_table$gene_id
storage.mode(counts) <- "integer"

count_labels_all <- sub("_MERGE$", "", sub("^mehreen_", "", colnames(counts)))
keep_samples <- count_labels_all %in% metadata$label
counts <- counts[, keep_samples, drop = FALSE]
count_labels <- count_labels_all[keep_samples]
metadata <- metadata[match(count_labels, metadata$label), ]
stopifnot(identical(metadata$label, count_labels))
rownames(metadata) <- metadata$label
colnames(counts) <- count_labels

placement <- readr::read_tsv(file.path(project_dir, params$placement_file),
                             show_col_types = FALSE) |>
  dplyr::distinct(gene_id, .keep_all = TRUE)
rrna_ids <- trimws(readLines(file.path(project_dir, params$rrna_file), warn = FALSE))
rrna_ids <- rrna_ids[nzchar(rrna_ids) & !startsWith(rrna_ids, "#")]
main_chromosome_ids <- placement$gene_id[
  placement$sequence_class == "Nuclear chromosome" &
    !placement$gene_id %in% rrna_ids
]
counts <- counts[intersect(rownames(counts), main_chromosome_ids), , drop = FALSE]

keep_vst <- rowSums(counts >= params$min_count) >= params$min_samples
input_audit <- data.frame(
  samples_retained = ncol(counts),
  raw_count_rows = nrow(counts_table),
  main_chromosome_non_rRNA_rows = nrow(counts),
  genes_entering_VST = sum(keep_vst)
)
knitr::kable(input_audit, caption = "Main-chromosome analysis input audit")
write.csv(input_audit, file.path(out_dir, "main_chromosome_input_audit.csv"),
          row.names = FALSE)

metadata$group <- factor(
  paste0("diet", metadata$Diet, "_", tolower(as.character(metadata$Treatment))),
  levels = c(
    "diet50_control", "diet33_control", "diet83_control",
    "diet50_infected", "diet33_infected", "diet83_infected"
  )
)

dds_global <- DESeq2::DESeqDataSetFromMatrix(
  countData = counts[keep_vst, ],
  colData = metadata,
  design = ~ 0 + group
)
dds_global <- DESeq2::DESeq(
  dds_global,
  minReplicatesForReplace = Inf,
  quiet = TRUE
)

vsd <- DESeq2::vst(dds_global, blind = TRUE)
vst_mat <- SummarizedExperiment::assay(vsd)
```

## VST PCA with condition centroids

PCA is the most transparent ordination for this step because the axes show how
much expression variance they explain. UMAP can be useful for pattern discovery,
but it does not give the same direct variance accounting and can change with
parameter choices. Here the small points are individual libraries, the larger
points are diet-by-treatment centroids, and the pale condition polygons connect
the outermost samples from each diet-by-treatment group in the first two PCs.

```{r vst-pca-centroids, fig.height=5.8, fig.width=8}
nonzero_variance <- apply(vst_mat, 1, var) > 0
pca <- prcomp(t(vst_mat[nonzero_variance, , drop = FALSE]), scale. = FALSE)
percent_var <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)

pca_scores <- data.frame(
  label = rownames(pca$x),
  PC1 = pca$x[, "PC1"],
  PC2 = pca$x[, "PC2"],
  Diet = metadata$Diet,
  Treatment = metadata$Treatment,
  check.names = FALSE
)
pca_scores$Condition <- interaction(pca_scores$Diet, pca_scores$Treatment,
                                    sep = " ")

pca_centroids <- aggregate(cbind(PC1, PC2) ~ Diet + Treatment + Condition,
                           pca_scores, mean)
names(pca_centroids)[names(pca_centroids) == "PC1"] <- "centroid_PC1"
names(pca_centroids)[names(pca_centroids) == "PC2"] <- "centroid_PC2"

make_condition_hull <- function(df) {
  if (nrow(df) < 3) return(df[0, ])
  df[chull(df$PC1, df$PC2), ]
}

pca_hulls <- do.call(rbind, lapply(split(pca_scores, pca_scores$Condition),
                                   make_condition_hull))

p_pca_centroids <- ggplot2::ggplot() +
  ggplot2::geom_polygon(
    data = pca_hulls,
    ggplot2::aes(x = PC1, y = PC2, fill = Condition, group = Condition),
    alpha = 0.20,
    color = NA
  ) +
  ggplot2::geom_point(
    data = pca_scores,
    ggplot2::aes(x = PC1, y = PC2, color = Diet, shape = Treatment),
    size = 3,
    alpha = 0.88
  ) +
  ggplot2::geom_point(
    data = pca_centroids,
    ggplot2::aes(x = centroid_PC1, y = centroid_PC2, color = Diet),
    shape = 21,
    fill = "white",
    stroke = 0.8,
    size = 5
  ) +
  ggrepel::geom_label_repel(
    data = pca_centroids,
    ggplot2::aes(x = centroid_PC1, y = centroid_PC2, label = Condition),
    color = "grey15",
    fill = "white",
    label.size = 0.18,
    size = 3.1,
    min.segment.length = 0
  ) +
  ggplot2::scale_color_manual(values = diet_colors) +
  ggplot2::scale_fill_manual(values = condition_surface_colors) +
  ggplot2::theme_classic() +
  ggplot2::theme(legend.position = "bottom") +
  ggplot2::labs(
    title = "VST PCA with diet-by-treatment centroids",
    x = paste0("PC1: ", percent_var[1], "% variance"),
    y = paste0("PC2: ", percent_var[2], "% variance"),
    color = "Diet",
    fill = "Condition polygon",
    shape = "Treatment"
  )

p_pca_centroids
ggplot2::ggsave(file.path(out_dir, "vst_pca_condition_centroids.png"),
                p_pca_centroids, width = 8, height = 5.8, dpi = 300)
write.csv(pca_scores, file.path(out_dir, "vst_pca_sample_scores.csv"),
          row.names = FALSE)
write.csv(pca_centroids, file.path(out_dir, "vst_pca_condition_centroids.csv"),
          row.names = FALSE)
```

## VST heatmap across samples

Before reading the infected-control tests, it helps to see the expression
structure across all samples. This heatmap uses the most variable
variance-stabilized genes, with diet and treatment annotations above the
columns. The VST transformation is used because raw RNA-seq counts have
mean-dependent variance; without transformation, very highly expressed genes can
overwhelm the visual structure.

```{r vst-sample-heatmap, fig.height=8, fig.width=9}
gene_variance <- apply(vst_mat, 1, var)
top_n <- min(params$top_variable_genes, length(gene_variance))
top_genes <- names(sort(gene_variance, decreasing = TRUE))[seq_len(top_n)]

annotation_col <- data.frame(
  Diet = metadata$Diet,
  Treatment = metadata$Treatment
)
rownames(annotation_col) <- colnames(vst_mat)

png(file.path(out_dir, "vst_top_variable_gene_heatmap.png"),
    width = 1800, height = 1600, res = 180)
pheatmap::pheatmap(
  vst_mat[top_genes, ],
  scale = "row",
  show_rownames = FALSE,
  annotation_col = annotation_col,
  main = paste("Top", top_n, "variable VST genes across samples")
)
dev.off()

pheatmap::pheatmap(
  vst_mat[top_genes, ],
  scale = "row",
  show_rownames = FALSE,
  annotation_col = annotation_col,
  main = paste("Top", top_n, "variable VST genes across samples")
)
```

## Run one infected-control test per diet

One joint `~ 0 + group` DESeq2 model is fitted across all 45 samples, using a
separate coefficient for every diet-by-treatment group. Each displayed result
then contrasts infected and control coefficients within one diet. This keeps
normalization, dispersion estimation, filtering, and multiple-testing behavior
consistent across the interaction, catalogue, enrichment, and publication
pages. A positive log2 fold-change means higher expression in infected samples;
a negative value means higher expression in controls.

```{r infection-tests, fig.height=7, fig.width=8}
diet_results <- list()

for (diet_level in levels(droplevels(metadata$Diet))) {
  message("Running infection contrast for diet ", diet_level)

  coefficient_names <- DESeq2::resultsNames(dds_global)
  contrast_vector <- stats::setNames(rep(0, length(coefficient_names)),
                                     coefficient_names)
  infected_name <- paste0("groupdiet", diet_level, "_infected")
  control_name <- paste0("groupdiet", diet_level, "_control")
  if (!all(c(infected_name, control_name) %in% coefficient_names)) {
    stop("Missing group coefficient for diet ", diet_level)
  }
  contrast_vector[infected_name] <- 1
  contrast_vector[control_name] <- -1

  res <- DESeq2::results(
    dds_global,
    contrast = contrast_vector,
    alpha = params$alpha,
    cooksCutoff = FALSE
  )
  res_df <- as.data.frame(res)
  res_df$gene_id <- rownames(res_df)
  res_df$diet <- diet_level
  res_df$significant <- !is.na(res_df$padj) &
    res_df$padj < params$alpha &
    abs(res_df$log2FoldChange) >= params$lfc_threshold

  diet_results[[diet_level]] <- res_df
  write.csv(res_df, file.path(out_dir,
                              paste0("DESeq2_infected_vs_control_diet_",
                                     diet_level, ".csv")),
            row.names = FALSE)
  writeLines(res_df$gene_id[res_df$significant],
             file.path(out_dir,
                       paste0("significant_genes_infection_diet_",
                              diet_level, ".txt")))

  print(EnhancedVolcano::EnhancedVolcano(
    res_df,
    lab = res_df$gene_id,
    x = "log2FoldChange",
    y = "padj",
    pCutoff = params$alpha,
    FCcutoff = params$lfc_threshold,
    title = paste("Infected vs control in diet", diet_level)
  ))
}

all_infection_results <- do.call(rbind, diet_results)
write.csv(all_infection_results,
          file.path(out_dir, "all_infection_within_diet_results.csv"),
          row.names = FALSE)
```

## Diet-specific DEG PCA

The PCA above uses all filtered genes and all samples, so it describes global
expression structure. This section asks a narrower question for each diet: if we
look only at the genes that are differentially expressed between infected and
control samples within that diet, do the samples separate by treatment? This is
closer to the DEG result itself, and it avoids using genes from another diet to
explain the infected-control pattern.

```{r diet-specific-deg-pca, fig.height=5.2, fig.width=7.2}
make_deg_pca <- function(diet_level) {
  res <- diet_results[[diet_level]]
  deg_genes <- res$gene_id[res$significant & !is.na(res$gene_id)]
  deg_genes <- intersect(deg_genes, rownames(vst_mat))
  sample_keep <- metadata$Diet == diet_level
  metadata_sub <- droplevels(metadata[sample_keep, ])

  if (length(deg_genes) < 3 || nrow(metadata_sub) < 4) {
    return(NULL)
  }

  mat <- vst_mat[deg_genes, metadata_sub$label, drop = FALSE]
  keep_gene <- apply(mat, 1, var) > 0
  mat <- mat[keep_gene, , drop = FALSE]
  if (nrow(mat) < 3) return(NULL)

  pca_local <- prcomp(t(mat), scale. = FALSE)
  percent_local <- round(100 * (pca_local$sdev^2 / sum(pca_local$sdev^2)), 1)
  scores <- data.frame(
    label = rownames(pca_local$x),
    PC1 = pca_local$x[, "PC1"],
    PC2 = pca_local$x[, "PC2"],
    Diet = metadata_sub$Diet,
    Treatment = metadata_sub$Treatment,
    check.names = FALSE
  )

  centroids <- aggregate(cbind(PC1, PC2) ~ Treatment, scores, mean)
  names(centroids)[names(centroids) == "PC1"] <- "centroid_PC1"
  names(centroids)[names(centroids) == "PC2"] <- "centroid_PC2"

  treatment_hulls <- do.call(rbind, lapply(split(scores, scores$Treatment),
                                           make_condition_hull))

  p <- ggplot2::ggplot(scores, ggplot2::aes(PC1, PC2)) +
    ggplot2::geom_polygon(
      data = treatment_hulls,
      ggplot2::aes(x = PC1, y = PC2, fill = Treatment, group = Treatment),
      alpha = 0.22,
      color = NA,
      inherit.aes = FALSE
    ) +
    ggplot2::geom_point(ggplot2::aes(color = Treatment, shape = Treatment),
                        size = 3.2, alpha = 0.9) +
    ggplot2::geom_point(
      data = centroids,
      ggplot2::aes(x = centroid_PC1, y = centroid_PC2, color = Treatment),
      shape = 21,
      fill = "white",
      stroke = 0.8,
      size = 5,
      inherit.aes = FALSE
    ) +
    ggrepel::geom_label_repel(
      data = centroids,
      ggplot2::aes(x = centroid_PC1, y = centroid_PC2, label = Treatment),
      color = "grey15",
      fill = "white",
      label.size = 0.18,
      size = 3.2,
      min.segment.length = 0,
      inherit.aes = FALSE
    ) +
    ggplot2::scale_color_manual(values = treatment_colors) +
    ggplot2::scale_fill_manual(values = treatment_fills) +
    ggplot2::theme_classic() +
    ggplot2::theme(legend.position = "bottom") +
    ggplot2::labs(
      title = paste("Diet", diet_level, "PCA using infection DEGs only"),
      subtitle = paste(length(deg_genes), "significant genes"),
      x = paste0("PC1: ", percent_local[1], "% variance"),
      y = paste0("PC2: ", percent_local[2], "% variance"),
      color = "Treatment",
      fill = "Treatment",
      shape = "Treatment"
    )

  ggsave(file.path(out_dir,
                   paste0("diet_", diet_level, "_infection_deg_pca.png")),
         p, width = 7.2, height = 5.2, dpi = 300)
  write.csv(scores, file.path(out_dir,
                              paste0("diet_", diet_level,
                                     "_infection_deg_pca_scores.csv")),
            row.names = FALSE)
  write.csv(centroids, file.path(out_dir,
                                 paste0("diet_", diet_level,
                                        "_infection_deg_pca_centroids.csv")),
            row.names = FALSE)
  p
}

for (diet_level in levels(droplevels(metadata$Diet))) {
  p_local <- make_deg_pca(diet_level)
  if (is.null(p_local)) {
    cat("Diet", diet_level,
        "does not have enough significant, variable genes for DEG-only PCA.\n")
  } else {
    print(p_local)
  }
}
```

## Within-diet MA plots and DEG heatmaps

The PCA shows whether samples separate using infection-responsive genes. The MA
plots add the gene-level view for each diet: effect size on the y-axis and mean
expression on the x-axis. The heatmaps then show the strongest infected-control
DEGs within each diet using only samples from that diet, with the same
blue-white-red VST palette used for the curated publication-style heatmap.

```{r within-diet-ma-plots, fig.height=4.8, fig.width=7.2}
make_ma_plot <- function(diet_level) {
  res <- diet_results[[diet_level]]
  if (is.null(res) || nrow(res) == 0) return(NULL)

  res$ma_class <- "Not significant"
  res$ma_class[res$significant & res$log2FoldChange > 0] <- "Higher in infected"
  res$ma_class[res$significant & res$log2FoldChange < 0] <- "Higher in control"
  res$ma_class <- factor(res$ma_class,
                         levels = c("Higher in infected",
                                    "Higher in control",
                                    "Not significant"))

  ggplot2::ggplot(res, ggplot2::aes(x = log10(baseMean + 1),
                                    y = log2FoldChange,
                                    color = ma_class)) +
    ggplot2::geom_hline(yintercept = c(-params$lfc_threshold,
                                       params$lfc_threshold),
                        linetype = "dashed", linewidth = 0.3,
                        color = "#6B7280") +
    ggplot2::geom_hline(yintercept = 0, linewidth = 0.25,
                        color = "#111827") +
    ggplot2::geom_point(alpha = 0.45, size = 0.7) +
    ggplot2::scale_color_manual(values = c(
      "Higher in infected" = treatment_colors[["Infected"]],
      "Higher in control" = "#237C8E",
      "Not significant" = "#D1D5DB"
    )) +
    ggplot2::theme_classic() +
    ggplot2::theme(legend.position = "bottom") +
    ggplot2::labs(
      title = paste("Diet", diet_level, "infected-control MA plot"),
      subtitle = paste(sum(res$significant, na.rm = TRUE),
                       "significant DEGs"),
      x = "log10 mean normalized count + 1",
      y = "log2 fold-change: infected / control",
      color = "DEG class"
    )
}

for (diet_level in levels(droplevels(metadata$Diet))) {
  p_ma <- make_ma_plot(diet_level)
  if (!is.null(p_ma)) {
    print(p_ma)
    ggplot2::ggsave(file.path(out_dir,
                              paste0("diet_", diet_level,
                                     "_infection_ma_plot.png")),
                    p_ma, width = 7.2, height = 4.8, dpi = 300)
  }
}
```

```{r within-diet-deg-heatmaps, results='asis'}
within_diet_heatmap_colors <- grDevices::colorRampPalette(
  c("#237C8E", "#F7F3E8", "#C85A54")
)(101)
within_diet_heatmap_breaks <- seq(-3, 3, length.out = 102)
within_diet_display_dir <- file.path(project_dir, "docs", "figure",
                                     "16-all-samples-infection-within-diet-deg.Rmd")
dir.create(within_diet_display_dir, recursive = TRUE, showWarnings = FALSE)

make_within_diet_heatmap <- function(diet_level, n_genes = 60) {
  res <- diet_results[[diet_level]]
  if (is.null(res)) return(NULL)

  degs <- res[res$significant & !is.na(res$padj) &
                !is.na(res$log2FoldChange), ]
  degs <- degs[order(degs$padj, -abs(degs$log2FoldChange)), ]
  deg_genes <- intersect(head(degs$gene_id, n_genes), rownames(vst_mat))
  if (length(deg_genes) < 3) return(NULL)

  sample_keep <- metadata$Diet == diet_level
  sample_order_diet <- metadata$label[sample_keep][order(
    factor(as.character(metadata$Treatment[sample_keep]),
           levels = c("Infected", "Control")),
    metadata$label[sample_keep]
  )]

  sample_annotation_diet <- data.frame(
    Treatment = metadata[sample_order_diet, "Treatment"]
  )
  rownames(sample_annotation_diet) <- sample_order_diet

  gene_direction <- degs[match(deg_genes, degs$gene_id), ]
  gene_annotation_diet <- data.frame(
    Direction = ifelse(gene_direction$log2FoldChange > 0,
                       "Higher in infected", "Higher in control")
  )
  rownames(gene_annotation_diet) <- deg_genes

  pheatmap::pheatmap(
    vst_mat[deg_genes, sample_order_diet, drop = FALSE],
    scale = "row",
    color = within_diet_heatmap_colors,
    breaks = within_diet_heatmap_breaks,
    cluster_rows = TRUE,
    cluster_cols = FALSE,
    show_rownames = FALSE,
    show_colnames = TRUE,
    fontsize_col = 7,
    annotation_col = sample_annotation_diet,
    annotation_row = gene_annotation_diet,
    annotation_colors = list(
      Treatment = treatment_colors,
      Direction = c("Higher in infected" = treatment_colors[["Infected"]],
                    "Higher in control" = "#237C8E")
    ),
    main = paste("Diet", diet_level, "top infection DEGs"),
    filename = file.path(out_dir,
                         paste0("diet_", diet_level,
                                "_infection_deg_vst_heatmap.png")),
    width = 8,
    height = 7.5
  )
}

for (diet_level in levels(droplevels(metadata$Diet))) {
  hm <- make_within_diet_heatmap(diet_level)
  if (is.null(hm)) {
    cat("Diet", diet_level,
        "does not have enough significant DEGs for a within-diet heatmap.\n")
  } else {
    print(hm)
    heatmap_out <- file.path(out_dir,
                             paste0("diet_", diet_level,
                                    "_infection_deg_vst_heatmap.png"))
    heatmap_display <- file.path(within_diet_display_dir,
                                 paste0("diet-", diet_level,
                                        "-infection-deg-vst-heatmap.png"))
    file.copy(heatmap_out, heatmap_display, overwrite = TRUE)
    heatmap_src <- file.path("figure", "16-all-samples-infection-within-diet-deg.Rmd",
                             basename(heatmap_display))
    cat(paste0(
      '<h4>Diet ', diet_level, '</h4>',
      '<p><img src="', heatmap_src,
      '" alt="Diet ', diet_level, ' infection DEG VST heatmap" ',
      'style="display:block;margin:auto;max-width:100%;height:auto;" /></p>'
    ))
  }
}
```

## DEG count barplot

The volcano plots are good for seeing whether a contrast has many small effects
or a few strong effects. This barplot answers the simpler question: how many
genes pass the current DEG thresholds in each diet, and are they higher in
infected or control samples?

```{r deg-count-barplot, fig.height=4.8, fig.width=7.5}
sig_infection <- all_infection_results[all_infection_results$significant &
                                         !is.na(all_infection_results$log2FoldChange), ]
sig_infection$diet <- factor(sig_infection$diet,
                             levels = levels(droplevels(metadata$Diet)))
sig_infection$direction <- ifelse(sig_infection$log2FoldChange > 0,
                                  "Higher in infected",
                                  "Higher in control")
sig_infection$direction <- factor(sig_infection$direction,
                                  levels = c("Higher in infected",
                                             "Higher in control"))

deg_bar_counts <- as.data.frame(table(
  diet = sig_infection$diet,
  direction = sig_infection$direction
))
names(deg_bar_counts)[3] <- "n_significant_genes"

p_deg_counts <- ggplot(deg_bar_counts,
                       aes(x = diet, y = n_significant_genes,
                           fill = direction)) +
  geom_col(position = position_dodge(width = 0.78), width = 0.68) +
  geom_text(aes(label = n_significant_genes),
            position = position_dodge(width = 0.78),
            vjust = -0.35, size = 3.2) +
  scale_fill_manual(values = c("Higher in infected" = treatment_colors[["Infected"]],
                               "Higher in control" = treatment_colors[["Control"]])) +
  theme_classic() +
  labs(
    title = "Significant infection DEGs within each diet",
    x = "Diet",
    y = "Number of significant genes",
    fill = "Direction"
  )

p_deg_counts
ggsave(file.path(out_dir, "infection_within_diet_deg_count_barplot.png"),
       p_deg_counts, width = 7.5, height = 4.8, dpi = 300)
write.csv(deg_bar_counts,
          file.path(out_dir, "infection_within_diet_direction_counts.csv"),
          row.names = FALSE)
```

## Publication-style PCA and DEG burden panels

These two panels mirror the AGY figure style. The PCA panel keeps individual
samples visible, marks diet-by-treatment centroids, and draws one arrow per diet
from the control centroid to the infected centroid. The DEG burden panel uses a
diverging axis so genes higher in infected samples and genes higher in controls
are immediately comparable.

```{r pca-centroid-arrows, fig.height=6.2, fig.width=10}
centroid_control <- pca_centroids[pca_centroids$Treatment == "Control", ]
centroid_infected <- pca_centroids[pca_centroids$Treatment == "Infected", ]
centroid_arrows <- merge(
  centroid_control[, c("Diet", "centroid_PC1", "centroid_PC2")],
  centroid_infected[, c("Diet", "centroid_PC1", "centroid_PC2")],
  by = "Diet",
  suffixes = c("_control", "_infected")
)

p_pca_arrows <- ggplot2::ggplot() +
  ggplot2::geom_polygon(
    data = pca_hulls,
    ggplot2::aes(x = PC1, y = PC2, fill = Condition, group = Condition),
    alpha = 0.16,
    color = NA
  ) +
  ggplot2::geom_point(
    data = pca_scores,
    ggplot2::aes(x = PC1, y = PC2, color = Diet, shape = Treatment),
    size = 3,
    alpha = 0.86
  ) +
  ggplot2::geom_segment(
    data = centroid_arrows,
    ggplot2::aes(x = centroid_PC1_control, y = centroid_PC2_control,
                 xend = centroid_PC1_infected, yend = centroid_PC2_infected,
                 color = Diet),
    linewidth = 0.85,
    arrow = grid::arrow(length = grid::unit(0.18, "cm"))
  ) +
  ggplot2::geom_point(
    data = pca_centroids,
    ggplot2::aes(x = centroid_PC1, y = centroid_PC2, color = Diet),
    shape = 21,
    fill = "white",
    stroke = 1,
    size = 5
  ) +
  ggrepel::geom_label_repel(
    data = pca_centroids,
    ggplot2::aes(x = centroid_PC1, y = centroid_PC2, label = Condition),
    color = "grey15",
    fill = "white",
    label.size = 0.18,
    size = 3.0,
    min.segment.length = 0
  ) +
  ggplot2::scale_color_manual(values = diet_colors) +
  ggplot2::scale_fill_manual(values = condition_surface_colors) +
  ggplot2::theme_classic() +
  ggplot2::theme(legend.position = "bottom",
                 legend.box = "vertical") +
  ggplot2::guides(
    fill = ggplot2::guide_legend(nrow = 2),
    color = ggplot2::guide_legend(nrow = 1),
    shape = ggplot2::guide_legend(nrow = 1)
  ) +
  ggplot2::labs(
    title = "PCA centroid shift after infection",
    subtitle = "Arrows connect control and infected centroids within each diet",
    x = paste0("PC1: ", percent_var[1], "% variance"),
    y = paste0("PC2: ", percent_var[2], "% variance"),
    color = "Diet",
    fill = "Condition polygon",
    shape = "Treatment"
  )

p_pca_arrows
ggplot2::ggsave(file.path(out_dir, "pca_centroid_infection_arrows.png"),
                p_pca_arrows, width = 10, height = 6.2, dpi = 300)
write.csv(centroid_arrows,
          file.path(out_dir, "pca_centroid_infection_arrows.csv"),
          row.names = FALSE)
```

```{r diverging-deg-burden, fig.height=4.8, fig.width=8.6}
deg_burden <- deg_bar_counts
deg_burden$diet <- factor(deg_burden$diet,
                          levels = levels(droplevels(metadata$Diet)))
deg_burden$signed_n <- ifelse(deg_burden$direction == "Higher in infected",
                              deg_burden$n_significant_genes,
                              -deg_burden$n_significant_genes)
max_burden <- max(abs(deg_burden$signed_n), na.rm = TRUE)

p_deg_burden <- ggplot2::ggplot(
  deg_burden,
  ggplot2::aes(x = signed_n, y = diet, fill = direction)
) +
  ggplot2::geom_vline(xintercept = 0, linewidth = 0.35, color = "grey35") +
  ggplot2::geom_col(width = 0.68) +
  ggplot2::geom_text(
    ggplot2::aes(label = n_significant_genes,
                 hjust = ifelse(signed_n > 0, -0.15, 1.15)),
    size = 3.2
  ) +
  ggplot2::scale_x_continuous(labels = abs,
                              limits = c(-max_burden * 1.25,
                                         max_burden * 1.25)) +
  ggplot2::scale_fill_manual(values = c("Higher in infected" = treatment_colors[["Infected"]],
                                        "Higher in control" = treatment_colors[["Control"]])) +
  ggplot2::theme_classic() +
  ggplot2::labs(
    title = "Direction-aware infection DEG burden",
    subtitle = "Right: higher in infected; left: higher in control",
    x = "Number of significant genes",
    y = "Diet",
    fill = "Direction"
  )

p_deg_burden
ggplot2::ggsave(file.path(out_dir, "infection_deg_burden_diverging_barplot.png"),
                p_deg_burden, width = 8.6, height = 4.8, dpi = 300)
write.csv(deg_burden,
          file.path(out_dir, "infection_deg_burden_diverging_counts.csv"),
          row.names = FALSE)
```

```{r diverging-deg-burden-by-biotype, fig.height=5.2, fig.width=9.4}
parse_gene_biotypes <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  genes <- gff[gff$type == "gene", ]
  data.frame(
    gene_id = sub(".*gene=([^;]+).*", "\\1", genes$attributes),
    gene_biotype = sub(".*gene_biotype=([^;]+).*", "\\1", genes$attributes),
    raw_attributes = genes$attributes,
    stringsAsFactors = FALSE
  ) |>
    dplyr::filter(grepl("^LOC", gene_id)) |>
    dplyr::mutate(gene_biotype = ifelse(gene_biotype == raw_attributes,
                                        NA_character_, gene_biotype)) |>
    dplyr::select(-raw_attributes) |>
    dplyr::distinct()
}

classify_gene_class <- function(x) {
  dplyr::case_when(
    x == "protein_coding" ~ "Coding genes",
    x == "lncRNA" ~ "Long non-coding RNA",
    x == "rRNA" ~ "rRNA",
    is.na(x) ~ "Other / unknown",
    TRUE ~ "Other non-coding"
  )
}

gene_biotypes <- parse_gene_biotypes(file.path(project_dir, params$gff_file))

biotype_burden <- sig_infection |>
  dplyr::left_join(gene_biotypes, by = "gene_id") |>
  dplyr::mutate(
    Gene_class = classify_gene_class(gene_biotype),
    signed_direction = ifelse(direction == "Higher in infected", 1, -1)
  ) |>
  dplyr::distinct(diet, direction, gene_id, Gene_class, .keep_all = TRUE) |>
  dplyr::count(diet, direction, signed_direction, Gene_class,
               name = "n_significant_genes") |>
  dplyr::mutate(
    signed_n = signed_direction * n_significant_genes,
    burden_class = paste(direction, Gene_class, sep = " - ")
  )

biotype_totals <- biotype_burden |>
  dplyr::group_by(diet, direction, signed_direction) |>
  dplyr::summarise(
    n_significant_genes = sum(n_significant_genes),
    signed_n = dplyr::first(signed_direction) * n_significant_genes,
    .groups = "drop"
  )

max_biotype_burden <- max(abs(biotype_totals$signed_n), 1)
biotype_totals$label_x <- ifelse(
  biotype_totals$signed_n > 0,
  biotype_totals$signed_n + 0.04 * max_biotype_burden,
  biotype_totals$signed_n - 0.04 * max_biotype_burden
)

biotype_burden$diet <- factor(biotype_burden$diet,
                              levels = levels(droplevels(metadata$Diet)))
biotype_totals$diet <- factor(biotype_totals$diet,
                              levels = levels(droplevels(metadata$Diet)))

biotype_colors <- c(
  "Higher in infected - Coding genes" = "#8B1E3F",
  "Higher in infected - Long non-coding RNA" = "#D98BA3",
  "Higher in infected - rRNA" = "#F2B8A0",
  "Higher in infected - Other non-coding" = "#F2B8A0",
  "Higher in infected - Other / unknown" = "#E5B567",
  "Higher in control - Coding genes" = "#1F5F8B",
  "Higher in control - Long non-coding RNA" = "#93B7D4",
  "Higher in control - rRNA" = "#B8C7D9",
  "Higher in control - Other non-coding" = "#B8C7D9",
  "Higher in control - Other / unknown" = "#C9CED6"
)

p_biotype_burden <- ggplot2::ggplot(
  biotype_burden,
  ggplot2::aes(x = signed_n, y = diet, fill = burden_class)
) +
  ggplot2::geom_vline(xintercept = 0, linewidth = 0.35, color = "grey35") +
  ggplot2::geom_col(width = 0.68) +
  ggplot2::geom_text(
    data = biotype_totals,
    ggplot2::aes(x = label_x, y = diet, label = n_significant_genes),
    inherit.aes = FALSE,
    size = 3.2
  ) +
  ggplot2::scale_x_continuous(labels = abs,
                              limits = c(-max_biotype_burden * 1.25,
                                         max_biotype_burden * 1.25)) +
  ggplot2::scale_fill_manual(values = biotype_colors, drop = FALSE) +
  ggplot2::theme_classic() +
  ggplot2::theme(
    legend.position = "bottom",
    legend.title = ggplot2::element_blank(),
    legend.text = ggplot2::element_text(size = 8)
  ) +
  ggplot2::labs(
    title = "Direction-aware infection DEG burden by gene class",
    subtitle = "This uses the rRNA-filtered master count matrix; rRNA should be absent after filtering",
    x = "Number of significant genes",
    y = "Diet"
  )

p_biotype_burden
ggplot2::ggsave(file.path(out_dir,
                          "infection_deg_burden_by_biotype_diverging_barplot.png"),
                p_biotype_burden, width = 9.4, height = 5.2, dpi = 300)
write.csv(biotype_burden,
          file.path(out_dir, "infection_deg_burden_by_biotype_counts.csv"),
          row.names = FALSE)
```

## Curated DEG VST heatmap

The full VST heatmap is a broad QC view. This heatmap is narrower: it selects
strong infection-responsive DEGs and fat-body-relevant genes with eggNOG
descriptions matching immune, storage, nutrition, lipid-transport, or
detoxification keywords. Samples are ordered by infection status first and then
diet instead of clustered, so the infected profiles appear together and
diet-specific substructure can be compared directly. The gene annotations are
kept deliberately simple: whether a gene is a significant infection DEG in at
least one diet, and its functional theme. Genes are ordered first by DEG class
and then by functional theme.

```{r curated-deg-heatmap, fig.height=9, fig.width=10}
sig_for_heatmap <- all_infection_results[all_infection_results$significant &
                                           !is.na(all_infection_results$padj) &
                                           !is.na(all_infection_results$log2FoldChange), ]
sig_for_heatmap$direction <- ifelse(sig_for_heatmap$log2FoldChange > 0,
                                    "Higher in infected",
                                    "Higher in control")
sig_for_heatmap <- sig_for_heatmap[order(sig_for_heatmap$padj,
                                         -abs(sig_for_heatmap$log2FoldChange)), ]

select_top_degs <- function(tab, n) {
  tab <- tab[order(tab$padj, -abs(tab$log2FoldChange)), ]
  head(tab, n)
}

curated_pieces <- split(sig_for_heatmap,
                        paste(sig_for_heatmap$diet,
                              sig_for_heatmap$direction, sep = "__"))
top_deg_rows <- do.call(rbind, lapply(curated_pieces, select_top_degs,
                                      n = params$top_curated_degs_per_direction))
top_deg_rows$Selection <- "Top DEG"

read_emapper <- function(path) {
  cols <- c(
    "query", "seed_ortholog", "evalue", "score", "eggNOG_OGs",
    "max_annot_lvl", "COG_category", "Description", "Preferred_name",
    "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
    "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC", "CAZy",
    "BiGG_Reaction", "PFAMs"
  )
  tab <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", fill = TRUE)
  names(tab) <- cols[seq_len(ncol(tab))]
  tab$protein_id <- sub(".*(XP_[0-9]+[.][0-9]+).*", "\\1", tab$query)
  tab
}

parse_cds_gene_bridge <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  cds <- gff[gff$type == "CDS", ]
  data.frame(
    protein_id = sub(".*protein_id=([^;]+).*", "\\1", cds$attributes),
    gene_id = sub(".*gene=([^;]+).*", "\\1", cds$attributes),
    stringsAsFactors = FALSE
  ) |>
    dplyr::filter(grepl("^XP_", protein_id), grepl("^LOC", gene_id)) |>
    dplyr::distinct()
}

eggnog_path <- file.path(project_dir, params$eggnog_file)
gff_path <- file.path(project_dir, params$gff_file)
if (file.exists(eggnog_path) && file.exists(gff_path)) {
  eggnog <- read_emapper(eggnog_path)
  protein_to_gene <- parse_cds_gene_bridge(gff_path)
  gene_annotation <- eggnog |>
    dplyr::left_join(protein_to_gene, by = "protein_id") |>
    dplyr::filter(!is.na(gene_id)) |>
    dplyr::distinct(gene_id, .keep_all = TRUE) |>
    dplyr::select(gene_id, Preferred_name, Description, GOs, KEGG_ko,
                  KEGG_Pathway, PFAMs)
} else {
  gene_annotation <- data.frame(gene_id = character(),
                                Preferred_name = character(),
                                Description = character(),
                                GOs = character(),
                                KEGG_ko = character(),
                                KEGG_Pathway = character(),
                                PFAMs = character())
}

annotated_sig <- sig_for_heatmap |>
  dplyr::left_join(gene_annotation, by = "gene_id")
annotation_text <- apply(annotated_sig[, c("Preferred_name", "Description",
                                           "GOs", "KEGG_ko", "KEGG_Pathway",
                                           "PFAMs")],
                         1, paste, collapse = " ")

immune_keywords <- paste(c(
  "immune", "immun", "defensin", "antimicrobial", "lysozyme",
  "phenoloxidase", "prophenoloxidase", "toll", "imd", "jak", "stat",
  "lectin", "peptidoglycan", "hemocyte", "serpin", "clip", "thioester",
  "pathogen", "fungal", "fungus", "melanization"
), collapse = "|")

nutrition_keywords <- paste(c(
  "nutrition", "nutrient", "metabolism", "metabolic", "lipid", "fatty",
  "carbohydrate", "glucose", "glycogen", "trehalose", "insulin",
  "amino acid", "protease", "peptidase", "trypsin", "chymotrypsin",
  "lipase", "amylase", "digestion", "digestive", "transporter"
), collapse = "|")

storage_keywords <- paste(c(
  "storage protein", "hexamerin", "arylphorin", "hemocyanin",
  "vitellogenin", "lipophorin", "apolipophorin", "nutrient reservoir"
), collapse = "|")

detox_keywords <- paste(c(
  "detox", "xenobiotic", "cytochrome p450", "p450", "glutathione",
  "peroxidase", "oxidative stress", "superoxide", "catalase"
), collapse = "|")

classify_functional_theme <- function(text) {
  dplyr::case_when(
    grepl(immune_keywords, text, ignore.case = TRUE) ~ "Immune-related",
    grepl(storage_keywords, text, ignore.case = TRUE) ~ "Storage/lipid transport",
    grepl(detox_keywords, text, ignore.case = TRUE) ~ "Detoxification/stress",
    grepl(nutrition_keywords, text, ignore.case = TRUE) ~ "Nutrition/metabolism",
    TRUE ~ "Other strong DEG"
  )
}

gene_deg_summary <- sig_for_heatmap |>
  dplyr::group_by(gene_id) |>
  dplyr::summarise(
    DEG_diet_count = dplyr::n_distinct(diet),
    DEG_diets = paste(sort(unique(as.character(diet))), collapse = ", "),
    DEG_direction_summary = dplyr::if_else(
      dplyr::n_distinct(direction) == 1,
      dplyr::first(direction),
      "Mixed directions"
    ),
    min_padj = min(padj, na.rm = TRUE),
    max_abs_lfc = max(abs(log2FoldChange), na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    DEG_class = "DEG"
  )

annotation_for_heatmap <- gene_annotation |>
  dplyr::filter(gene_id %in% rownames(vst_mat)) |>
  dplyr::mutate(
    curated_annotation_text = paste(Preferred_name, Description, GOs, KEGG_ko,
                                    KEGG_Pathway, PFAMs),
    Functional_theme = classify_functional_theme(curated_annotation_text)
  )

theme_rows <- annotation_for_heatmap |>
  dplyr::left_join(gene_deg_summary, by = "gene_id") |>
  dplyr::mutate(
    DEG_class = dplyr::if_else(is.na(DEG_class), "Not DEG", DEG_class),
    min_padj = dplyr::if_else(is.na(min_padj), Inf, min_padj),
    max_abs_lfc = dplyr::if_else(is.na(max_abs_lfc), 0, max_abs_lfc),
    vst_variance = apply(vst_mat[gene_id, , drop = FALSE], 1, var)
  ) |>
  dplyr::filter(Functional_theme != "Other strong DEG") |>
  dplyr::group_by(Functional_theme, DEG_class) |>
  dplyr::arrange(min_padj, dplyr::desc(max_abs_lfc),
                 dplyr::desc(vst_variance), .by_group = TRUE) |>
  dplyr::slice_head(n = params$top_functional_degs_per_theme) |>
  dplyr::ungroup() |>
  dplyr::mutate(Selection = "Functional candidate")

curated_degs <- top_deg_rows |>
  dplyr::select(gene_id, diet, direction, padj, log2FoldChange, Selection) |>
  dplyr::left_join(annotation_for_heatmap, by = "gene_id") |>
  dplyr::left_join(gene_deg_summary, by = "gene_id", suffix = c("", ".summary")) |>
  dplyr::mutate(
    Functional_theme = dplyr::if_else(is.na(Functional_theme),
                                      "Other strong DEG",
                                      Functional_theme),
    DEG_class = "DEG"
  ) |>
  dplyr::bind_rows(theme_rows) |>
  dplyr::mutate(
    DEG_class = dplyr::if_else(is.na(DEG_class), "Not DEG", DEG_class),
    min_padj = dplyr::if_else(is.na(min_padj), Inf, min_padj),
    max_abs_lfc = dplyr::if_else(is.na(max_abs_lfc), 0, max_abs_lfc),
    Functional_theme = factor(
      Functional_theme,
      levels = c("Immune-related", "Storage/lipid transport",
                 "Nutrition/metabolism", "Detoxification/stress",
                 "Other strong DEG")
    ),
    DEG_class = factor(DEG_class, levels = c("DEG", "Not DEG"))
  ) |>
  dplyr::arrange(DEG_class, Functional_theme, min_padj,
                 dplyr::desc(max_abs_lfc)) |>
  dplyr::distinct(gene_id, .keep_all = TRUE)

curated_genes <- curated_degs$gene_id[curated_degs$gene_id %in% rownames(vst_mat)]
treatment_order <- factor(as.character(metadata$Treatment),
                          levels = c("Infected", "Control"))
sample_order <- metadata$label[order(treatment_order, metadata$Diet,
                                     metadata$label)]

sample_annotation <- data.frame(
  Treatment = metadata[sample_order, "Treatment"],
  Diet = metadata[sample_order, "Diet"]
)
rownames(sample_annotation) <- sample_order

gene_annotation_display <- curated_degs[match(curated_genes, curated_degs$gene_id),
                                        c("DEG_class",
                                          "Functional_theme"), drop = FALSE]
rownames(gene_annotation_display) <- curated_genes

label_source <- curated_degs[match(curated_genes, curated_degs$gene_id), ]
label_text <- ifelse(!is.na(label_source$Preferred_name) &
                       label_source$Preferred_name != "-",
                     label_source$Preferred_name,
                     label_source$Description)
label_text[is.na(label_text) | label_text == "-" | label_text == ""] <- "unannotated"
label_text <- substr(label_text, 1, 45)
col_labels <- paste(curated_genes, label_text, sep = " | ")

find_gaps <- function(groups) {
  groups <- as.character(groups)
  if (length(groups) < 2) {
    return(integer())
  }
  which(groups[-length(groups)] != groups[-1])
}

sample_gaps <- find_gaps(paste(sample_annotation$Treatment,
                               sample_annotation$Diet))
gene_gaps <- find_gaps(paste(gene_annotation_display$DEG_class,
                             gene_annotation_display$Functional_theme))

annotation_colors <- list(
  Diet = diet_colors,
  Treatment = treatment_colors,
  DEG_class = c("DEG" = "#111827",
                "Not DEG" = "#D1D5DB"),
  Functional_theme = c("Immune-related" = "#237C8E",
                       "Storage/lipid transport" = "#D7832F",
                       "Nutrition/metabolism" = "#B8643C",
                       "Detoxification/stress" = "#8C6D31",
                       "Other strong DEG" = "#9CA3AF")
)

heatmap_expression_colors <- grDevices::colorRampPalette(
  c("#237C8E", "#F7F3E8", "#C85A54")
)(101)
heatmap_expression_breaks <- seq(-3, 3, length.out = 102)

heatmap_display <- pheatmap::pheatmap(
  t(vst_mat[curated_genes, sample_order, drop = FALSE]),
  scale = "column",
  color = heatmap_expression_colors,
  breaks = heatmap_expression_breaks,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  gaps_row = sample_gaps,
  gaps_col = gene_gaps,
  show_rownames = TRUE,
  show_colnames = FALSE,
  labels_col = col_labels,
  fontsize_row = 6,
  annotation_col = gene_annotation_display,
  annotation_row = sample_annotation,
  annotation_colors = annotation_colors,
  main = "Curated fat body infection-response genes",
  filename = file.path(out_dir, "curated_infection_deg_vst_heatmap.png"),
  width = 14,
  height = 11
)

labeled_heatmap_display <- pheatmap::pheatmap(
  t(vst_mat[curated_genes, sample_order, drop = FALSE]),
  scale = "column",
  color = heatmap_expression_colors,
  breaks = heatmap_expression_breaks,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  gaps_row = sample_gaps,
  gaps_col = gene_gaps,
  show_rownames = TRUE,
  show_colnames = TRUE,
  labels_col = col_labels,
  fontsize_row = 6,
  fontsize_col = 4.5,
  angle_col = 90,
  annotation_col = gene_annotation_display,
  annotation_row = sample_annotation,
  annotation_colors = annotation_colors,
  main = "Curated fat body infection-response genes with annotations",
  filename = file.path(out_dir,
                       "curated_infection_deg_vst_heatmap_labeled.png"),
  width = 22,
  height = 12
)

display_dir <- file.path(project_dir, "docs", "figure",
                         "16-all-samples-infection-within-diet-deg.Rmd")
dir.create(display_dir, recursive = TRUE, showWarnings = FALSE)
display_png <- file.path(display_dir, "curated-deg-heatmap-display.png")
file.copy(file.path(out_dir, "curated_infection_deg_vst_heatmap.png"),
          display_png, overwrite = TRUE)
curated_heatmap_src <- file.path("figure", "16-all-samples-infection-within-diet-deg.Rmd",
                                 "curated-deg-heatmap-display.png")

labeled_display_png <- file.path(display_dir,
                                 "curated-deg-heatmap-labeled-display.png")
file.copy(file.path(out_dir, "curated_infection_deg_vst_heatmap_labeled.png"),
          labeled_display_png, overwrite = TRUE)
labeled_curated_heatmap_src <- file.path(
  "figure", "16-all-samples-infection-within-diet-deg.Rmd",
  "curated-deg-heatmap-labeled-display.png"
)

write.csv(curated_degs,
          file.path(out_dir, "curated_infection_deg_heatmap_genes.csv"),
          row.names = FALSE)
```

```{r curated-deg-heatmap-display, echo=FALSE, results='asis'}
cat(paste0(
  '<p><img src="', curated_heatmap_src,
  '" alt="Curated infection, immune, and nutrition DEG VST heatmap" ',
  'style="display:block;margin:auto;max-width:100%;height:auto;" /></p>'
))
```

```{r curated-deg-heatmap-labeled-display, echo=FALSE, results='asis'}
cat(paste0(
  '<h3>Labeled curated heatmap</h3>',
  '<p><img src="', labeled_curated_heatmap_src,
  '" alt="Curated infection DEG VST heatmap with gene descriptions" ',
  'style="display:block;margin:auto;max-width:100%;height:auto;" /></p>'
))
```

## Full non-curated infection DEG heatmap

This diagnostic heatmap uses every significant infected-control DEG detected
within any diet on this page. It is meant for pattern checking rather than for a
publication panel: rows are clustered so unusual gene blocks can emerge, while
sample columns remain ordered by infection status and diet.

Because the full DEG set is large, gene labels are hidden on the image. The table
below keeps the clustered heatmap row order, gene description, the diets where
each gene is a DEG, and the sample with the strongest row-z-score deviation.

```{r full-infection-deg-diagnostic-heatmap}
full_infection_deg_table <- sig_for_heatmap |>
  dplyr::filter(gene_id %in% rownames(vst_mat)) |>
  dplyr::group_by(gene_id) |>
  dplyr::summarise(
    n_infection_diets = dplyr::n_distinct(diet),
    infection_diets = paste(sort(unique(as.character(diet))), collapse = "; "),
    directions = paste(sort(unique(direction)), collapse = "; "),
    comparisons = paste0(
      "Diet ", paste(sort(unique(as.character(diet))), collapse = "; Diet "),
      ": infected vs control"
    ),
    max_abs_log2FC = max(abs(log2FoldChange), na.rm = TRUE),
    min_padj = min(padj, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::left_join(gene_annotation |> dplyr::select(gene_id, Description),
                   by = "gene_id") |>
  dplyr::mutate(
    Description = dplyr::coalesce(Description, "no description"),
    DEG_scope = dplyr::case_when(
      n_infection_diets >= 3 ~ "DEG in 3 diets",
      n_infection_diets == 2 ~ "DEG in 2 diets",
      TRUE ~ "DEG in 1 diet"
    ),
    DEG_scope = factor(DEG_scope,
                       levels = c("DEG in 3 diets", "DEG in 2 diets",
                                  "DEG in 1 diet"))
  ) |>
  dplyr::arrange(DEG_scope, min_padj, dplyr::desc(max_abs_log2FC))

full_infection_deg_ids <- full_infection_deg_table$gene_id

if (length(full_infection_deg_ids) > 1) {
  full_infection_gene_annotation <- full_infection_deg_table |>
    dplyr::transmute(DEG_scope = as.character(DEG_scope))
  full_infection_gene_annotation <- as.data.frame(
    full_infection_gene_annotation,
    stringsAsFactors = FALSE
  )
  rownames(full_infection_gene_annotation) <- full_infection_deg_ids

  full_infection_mat <- vst_mat[
    full_infection_deg_ids,
    sample_order,
    drop = FALSE
  ]
  full_infection_zmat <- t(scale(t(full_infection_mat)))
  full_infection_zmat[is.na(full_infection_zmat)] <- 0
  storage.mode(full_infection_zmat) <- "double"
  max_deviation_index <- max.col(abs(full_infection_zmat),
                                 ties.method = "first")
  max_deviation_value <- full_infection_zmat[
    cbind(seq_len(nrow(full_infection_zmat)), max_deviation_index)
  ]
  full_infection_zmat_plot <- pmax(pmin(full_infection_zmat, 3), -3)
  full_infection_zmat_plot <- as.matrix(full_infection_zmat_plot)
  storage.mode(full_infection_zmat_plot) <- "double"

  full_infection_annotation_colors <- annotation_colors
  full_infection_annotation_colors$DEG_scope <- c(
    "DEG in 3 diets" = "#111827",
    "DEG in 2 diets" = "#6B7280",
    "DEG in 1 diet" = "#D1D5DB"
  )

  full_infection_heatmap <- pheatmap::pheatmap(
    full_infection_zmat_plot,
    color = heatmap_expression_colors,
    breaks = heatmap_expression_breaks,
    cluster_rows = TRUE,
    cluster_cols = FALSE,
    gaps_col = sample_gaps,
    show_rownames = FALSE,
    show_colnames = TRUE,
    fontsize_col = 6,
    annotation_row = full_infection_gene_annotation,
    annotation_col = sample_annotation,
    annotation_colors = full_infection_annotation_colors,
    main = "Full infection-within-diet DEG heatmap",
    filename = file.path(out_dir, "full_infection_deg_diagnostic_heatmap.png"),
    width = 16,
    height = 22,
    silent = TRUE
  )

  full_infection_row_order <- if (!is.null(full_infection_heatmap$tree_row)) {
    rownames(full_infection_zmat_plot)[full_infection_heatmap$tree_row$order]
  } else {
    rownames(full_infection_zmat_plot)
  }

  full_infection_gene_table_ordered <- full_infection_deg_table |>
    dplyr::mutate(
      max_abs_sample_z = abs(max_deviation_value[
        match(gene_id, rownames(full_infection_zmat))
      ]),
      sample_with_max_abs_z = sample_order[
        max_deviation_index[match(gene_id, rownames(full_infection_zmat))]
      ],
      sample_deviation_direction = dplyr::if_else(
        max_deviation_value[match(gene_id, rownames(full_infection_zmat))] > 0,
        "Higher than gene average",
        "Lower than gene average"
      )
    ) |>
    dplyr::slice(match(full_infection_row_order, gene_id)) |>
    dplyr::mutate(heatmap_row_order = dplyr::row_number()) |>
    dplyr::select(heatmap_row_order, gene_id, Description, DEG_scope,
                  n_infection_diets, infection_diets, directions, comparisons,
                  max_abs_log2FC, min_padj, max_abs_sample_z,
                  sample_with_max_abs_z, sample_deviation_direction)

  write.csv(
    full_infection_gene_table_ordered,
    file.path(out_dir, "full_infection_deg_diagnostic_heatmap_gene_order.csv"),
    row.names = FALSE
  )
  utils::write.csv(
    tibble::rownames_to_column(as.data.frame(full_infection_zmat), "gene_id"),
    gzfile(file.path(out_dir, "full_infection_deg_diagnostic_z_vst_matrix.csv.gz")),
    row.names = FALSE
  )

  full_infection_display_png <- file.path(
    display_dir,
    "full-infection-deg-diagnostic-heatmap-display.png"
  )
  file.copy(
    file.path(out_dir, "full_infection_deg_diagnostic_heatmap.png"),
    full_infection_display_png,
    overwrite = TRUE
  )
  full_infection_heatmap_src <- file.path(
    "figure", basename(display_dir),
    "full-infection-deg-diagnostic-heatmap-display.png"
  )
} else {
  message("Not enough infection-within-diet DEGs for the full diagnostic heatmap.")
}
```

```{r full-infection-deg-diagnostic-heatmap-display, echo=FALSE, results='asis'}
if (exists("full_infection_heatmap_src")) {
  cat(paste0(
    '<p><img src="', full_infection_heatmap_src,
    '" alt="Full non-curated infection DEG diagnostic heatmap" ',
    'style="display:block;margin:auto;max-width:100%;height:auto;" /></p>'
  ))
}
```

```{r full-infection-deg-diagnostic-table}
if (exists("full_infection_gene_table_ordered")) {
  DT::datatable(
    full_infection_gene_table_ordered,
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 20, scrollX = TRUE),
    caption = "Full infection DEG heatmap gene order and strongest single-sample deviations"
  )
}
```

## Curated immune and nutrition DEG summary

This compact figure uses the same curated gene table as the heatmap, but
summarizes it by diet, direction, and biological theme. It is easier to read
quickly than the full heatmap and is the better candidate for a publication-style
overview panel.

```{r curated-theme-barplot, fig.height=5.2, fig.width=8.5}
curated_summary <- curated_degs |>
  dplyr::filter(Functional_theme %in% c("Immune-related",
                                        "Nutrition/metabolism")) |>
  dplyr::mutate(
    diet = factor(diet, levels = levels(droplevels(metadata$Diet))),
    direction = factor(direction, levels = c("Higher in infected",
                                             "Higher in control")),
    Functional_theme = factor(Functional_theme,
                              levels = c("Immune-related",
                                         "Nutrition/metabolism"))
  ) |>
  dplyr::count(diet, direction, Functional_theme,
               name = "n_curated_degs")

p_curated_summary <- ggplot(curated_summary,
                            aes(x = diet, y = n_curated_degs,
                                fill = Functional_theme)) +
  geom_col(position = position_dodge(width = 0.76), width = 0.68) +
  geom_text(aes(label = n_curated_degs),
            position = position_dodge(width = 0.76),
            vjust = -0.35, size = 3.1) +
  facet_wrap(~ direction) +
  scale_fill_manual(values = c("Immune-related" = "#237C8E",
                               "Nutrition/metabolism" = "#B8643C")) +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(
    title = "Curated immune and nutrition DEGs",
    subtitle = "Significant infected-control DEGs selected from eggNOG descriptions",
    x = "Diet",
    y = "Number of curated DEGs",
    fill = "Functional theme"
  )

p_curated_summary
ggsave(file.path(out_dir, "curated_immune_nutrition_deg_barplot.png"),
       p_curated_summary, width = 8.5, height = 5.2, dpi = 300)
write.csv(curated_summary,
          file.path(out_dir, "curated_immune_nutrition_deg_summary.csv"),
          row.names = FALSE)
```

```{r curated-immune-nutrition-gene-list}
curated_gene_list <- curated_degs |>
  dplyr::filter(Functional_theme %in% c("Immune-related",
                                        "Nutrition/metabolism",
                                        "Storage/lipid transport")) |>
  dplyr::mutate(
    Preferred_name = ifelse(is.na(Preferred_name) | Preferred_name == "-",
                            "", Preferred_name),
    Description = ifelse(is.na(Description) | Description == "-",
                         "", Description),
    DEG_diets = ifelse(is.na(DEG_diets), "", DEG_diets),
    DEG_direction_summary = ifelse(is.na(DEG_direction_summary), "",
                                   DEG_direction_summary),
    min_padj = ifelse(is.infinite(min_padj), NA_real_, min_padj)
  ) |>
  dplyr::select(
    gene_id, Preferred_name, Description, Functional_theme, DEG_class,
    DEG_diets, DEG_direction_summary, min_padj, max_abs_lfc
  ) |>
  dplyr::arrange(DEG_class, Functional_theme, min_padj,
                 dplyr::desc(max_abs_lfc))

DT::datatable(
  curated_gene_list,
  rownames = FALSE,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = "Curated immune, nutrition, and storage/lipid-transport genes shown in the heatmap"
)
write.csv(curated_gene_list,
          file.path(out_dir, "curated_immune_nutrition_gene_list.csv"),
          row.names = FALSE)
```

## Summary

The summary table collapses each diet-specific result into a total DEG count.
It is deliberately simple, because the direction-aware barplot above carries
the visual comparison and the full CSV files keep the gene-level statistics.

```{r summary}
summary_table <- aggregate(significant ~ diet, all_infection_results, sum)
names(summary_table)[2] <- "n_significant_genes"
knitr::kable(summary_table)

write.csv(summary_table, file.path(out_dir, "infection_within_diet_summary.csv"),
          row.names = FALSE)
```

```{r session-info}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
