---
title: "Phase-signature overlap with outliers removed"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  latest_results_root: "output/rmd_runs"
  phase_reference_dir: "data/external/phase_reference"
  counts_file: "data/raw_read_counts/master_counts_no_rRNA.csv"
  metadata_file: "data/metadata/mehreen_metadata_fixed.txt"
  timecourse_fatbody_reference_dir: "data/external/timecourse_fatbody_reference"
  timecourse_head_thorax_counts_dir: "data/raw_read_counts/03-gregaria-DESeq2"
  curated_infection_heatmap_gene_table: ""
  eggnog_file: "data/GO_Annotations/GCF_023897955.1_iqSchGreg1.2_Arthopoda_one2one.emapper.annotations"
  gff_file: "data/reference/GCF_023897955.1_iqSchGreg1.2_genomic.gff"
  alpha: 0.05
  lfc_threshold: 1
  phase_alpha: 0.05
  phase_lfc_threshold: 1
  min_total_count: 10
  heatmap_gene_n: 55
  timecourse_vst_delta_threshold: 1
  timecourse_top_gene_n: 500
  cross_tissue_top_variable_gene_n: 120
  cross_tissue_fatbody_deg_heatmap_n: 150
  analysis_scenario: "Current outliers removed"
  exclude_experiment_outliers: true
  tissue_vst_range_threshold: 2
  source_partition_top_gene_n: 80
  display_rmd_dir: "21-phase-signature-overlap-no-outliers.Rmd"
  output_tag: "phase-signature-overlap-no-outliers"
---

## Purpose

This page asks whether the fat body DEGs from the diet-by-infection experiment
also overlap with independent phase-responsive genes from *S. gregaria* head
and thorax transcriptomes.

The concern is biological rather than technical: fat bodies were sampled after
animals experienced the diet and infection treatments, but the experimental
context may also include a shift toward semi-isolated conditions. If some fat
body DEGs are also known phase-transition genes, those genes should be treated
as potentially density-state sensitive. They are not removed from the analysis;
they are flagged so the infection and diet interpretation can be more careful.

In the comparative phase dataset, isolated or solitarious animals were the
reference state. Therefore, a positive phase log2 fold-change means a gene is
higher in crowded/gregarious animals, and a negative phase log2 fold-change
means a gene is higher in isolated/solitarious animals.

This rendered page uses the fat body DEG scenario
`r params$analysis_scenario`. Experiment fat body sample displays use
`r if (isTRUE(params$exclude_experiment_outliers)) "the current outlier-removed sample set" else "all experiment samples, including the current outlier flags"`.

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c("DESeq2", "dplyr", "tidyr", "stringr", "ggplot2", "forcats",
              "DT", "SummarizedExperiment", "tibble", "scales", "pheatmap")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

library(ggplot2)

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

results_root <- file.path(project_dir, params$latest_results_root)
phase_dir <- file.path(project_dir, params$phase_reference_dir)
counts_file <- file.path(project_dir, params$counts_file)
metadata_file <- file.path(project_dir, params$metadata_file)
timecourse_fatbody_dir <- file.path(project_dir, params$timecourse_fatbody_reference_dir)
timecourse_head_thorax_dir <- file.path(project_dir, params$timecourse_head_thorax_counts_dir)
curated_infection_heatmap_gene_table <- params$curated_infection_heatmap_gene_table
eggnog_file <- file.path(project_dir, params$eggnog_file)
gff_file <- file.path(project_dir, params$gff_file)

diet_colors <- c("33" = "#E6862E", "50" = "#2F8F5B", "83" = "#7B4FA3")
treatment_colors <- c("Control" = "#7B8794", "Infected" = "#C85A54")

phase_colors <- c(
  "Gregarious-up" = "#B23A48",
  "Solitarious-up" = "#2166AC",
  "Discordant phase direction" = "#6B7280",
  "No phase DEG" = "#D1D5DB",
  "No phase reference" = "#F3F4F6"
)

shift_colors <- c(
  "Numerator more gregarious-like" = "#B23A48",
  "Numerator more solitarious-like" = "#2166AC",
  "Mixed head/thorax phase signal" = "#6B7280",
  "Not phase-associated" = "#D1D5DB"
)

fat_direction_colors <- c(
  "Upregulated" = "#E41A1C",
  "Downregulated" = "#1E88E5"
)

condition_colors <- c(
  "Infected_33" = "#C45A1A",
  "Control_33" = "#F3B36B",
  "Infected_50" = "#1E6F46",
  "Control_50" = "#8DD3A7",
  "Infected_83" = "#5F3A8A",
  "Control_83" = "#C4A4E0",
  "Gregarious reference" = "#111827",
  "Solitarious reference" = "#6B7280"
)

cross_tissue_colors <- c(
  condition_colors[c(
    "Infected_33", "Control_33", "Infected_50",
    "Control_50", "Infected_83", "Control_83"
  )],
  "Fat body: gregarious/crowded control" = "#111827",
  "Fat body: solitarious/isolated control" = "#6B7280",
  "Head: crowded/gregarious" = "#009E9A",
  "Head: isolated/solitarious" = "#8DDDD8",
  "Thorax: crowded/gregarious" = "#C51B7D",
  "Thorax: isolated/solitarious" = "#F1A6C8"
)

current_outliers <- c("1007", "1036", "1037", "1039")
sample_filter_label <- if (isTRUE(params$exclude_experiment_outliers)) {
  "current outliers removed"
} else {
  "all experiment samples included"
}
analysis_scenario_label <- as.character(params$analysis_scenario)
```

## Input data

The phase-reference tables were copied from `locust-comparative-genomics` into
this repository. This page only reads those copied files, so the comparative
repository is not modified.

```{r input-files}
phase_files <- tibble::tibble(
  tissue = c("Head", "Thorax"),
  result_file = file.path(
    phase_dir,
    c("DESeq2_results_sva_Head_gregaria_togregaria.csv",
      "DESeq2_results_sva_Thorax_gregaria_togregaria.csv")
  ),
  significant_file = file.path(
    phase_dir,
    c("DESeq2_sigresults_sva_Head_gregaria_togregaria.csv",
      "DESeq2_sigresults_sva_Thorax_gregaria_togregaria.csv")
  )
)

if (any(!file.exists(phase_files$result_file))) {
  stop("Missing phase-reference files: ",
       paste(phase_files$result_file[!file.exists(phase_files$result_file)],
             collapse = ", "))
}

knitr::kable(
  phase_files |>
    dplyr::mutate(
      result_file = basename(result_file),
      significant_file = basename(significant_file)
    ),
  caption = "Copied phase-reference DEG files"
)
```

## Helper functions

```{r helper-functions}
latest_dir <- function(pattern) {
  dirs <- list.dirs(results_root, recursive = FALSE, full.names = TRUE)
  dirs <- dirs[grepl(pattern, basename(dirs))]
  if (length(dirs) == 0) return(NA_character_)
  dirs[which.max(file.info(dirs)$mtime)]
}

resolve_curated_infection_gene_table <- function(path_param) {
  explicit_path <- if (is.null(path_param) || !nzchar(path_param)) {
    NA_character_
  } else if (grepl("^/", path_param)) {
    path_param
  } else {
    file.path(project_dir, path_param)
  }
  if (!is.na(explicit_path) && file.exists(explicit_path)) {
    return(explicit_path)
  }
  infection_dir <- latest_dir("^infection-within-diet-deg_")
  if (is.na(infection_dir)) {
    return(explicit_path)
  }
  candidate <- file.path(infection_dir, "curated_infection_deg_heatmap_genes.csv")
  if (file.exists(candidate)) {
    return(candidate)
  }
  explicit_path
}

curated_infection_heatmap_gene_table <- resolve_curated_infection_gene_table(
  curated_infection_heatmap_gene_table
)

read_phase_table <- function(path, tissue) {
  tab <- read.csv(path, check.names = FALSE)
  bad_names <- is.na(names(tab)) | names(tab) == ""
  names(tab)[bad_names] <- paste0("row_id", seq_len(sum(bad_names)))
  if (!"GeneID" %in% names(tab)) {
    tab$GeneID <- tab[[1]]
  }
  tab |>
    dplyr::transmute(
      gene_id = as.character(GeneID),
      tissue = tissue,
      phase_baseMean = baseMean,
      phase_log2FoldChange = log2FoldChange,
      phase_padj = padj,
      phase_significant = !is.na(padj) &
        padj < params$phase_alpha &
        abs(log2FoldChange) >= params$phase_lfc_threshold,
      phase_direction = dplyr::case_when(
        !phase_significant ~ "No phase DEG",
        log2FoldChange > 0 ~ "Gregarious-up",
        log2FoldChange < 0 ~ "Solitarious-up",
        TRUE ~ "No phase DEG"
      )
    )
}

read_emapper <- function(path) {
  cols <- c(
    "query", "seed_ortholog", "evalue", "score", "eggNOG_OGs",
    "max_annot_lvl", "COG_category", "Description", "Preferred_name",
    "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
    "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC", "CAZy",
    "BiGG_Reaction", "PFAMs"
  )
  tab <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", fill = TRUE)
  names(tab) <- cols[seq_len(ncol(tab))]
  tab |>
    dplyr::mutate(protein_id = stringr::str_extract(query, "XP_[0-9]+[.][0-9]+"))
}

parse_cds_gene_bridge <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  gff |>
    dplyr::filter(type == "CDS") |>
    dplyr::transmute(
      protein_id = stringr::str_match(attributes, "protein_id=([^;]+)")[, 2],
      gene_id = stringr::str_match(attributes, "gene=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(protein_id), !is.na(gene_id)) |>
    dplyr::distinct()
}

safe_neglog10 <- function(x) {
  x <- ifelse(is.na(x), NA_real_, x)
  -log10(pmax(x, .Machine$double.xmin))
}

read_featurecounts_gene_counts <- function(path, sample_id) {
  tab <- read.delim(path, comment.char = "#", check.names = FALSE)
  if ("Geneid" %in% names(tab)) {
    count_col <- names(tab)[ncol(tab)]
    out <- tibble::tibble(
      gene_id = as.character(tab[["Geneid"]]),
      count = as.integer(tab[[count_col]])
    )
  } else {
    tab <- read.delim(path, comment.char = "#", header = FALSE,
                      check.names = FALSE)
    if (ncol(tab) < 2) {
      stop("Count file must contain at least gene ID and count columns: ", path)
    }
    out <- tibble::tibble(
      gene_id = as.character(tab[[1]]),
      count = as.integer(tab[[ncol(tab)]])
    )
  }
  out <- out |>
    dplyr::filter(!is.na(gene_id), nzchar(gene_id)) |>
    dplyr::group_by(gene_id) |>
    dplyr::summarise(count = sum(count, na.rm = TRUE), .groups = "drop")
  names(out)[2] <- sample_id
  out
}

parse_head_thorax_reference_metadata <- function(reference_dir) {
  files <- list.files(
    reference_dir,
    pattern = "^SGRE-(HEAD|THOX)-(CRD|ISO)-[0-9]+_MERGE_featurecounts[.]txt$",
    full.names = TRUE
  )
  if (length(files) == 0) {
    return(tibble::tibble(
      sample_id = character(),
      count_file = character(),
      tissue = character(),
      phase = character(),
      rearing_condition = character(),
      replicate = integer(),
      source = character(),
      display_group = character(),
      reference_role = character(),
      sample_label = character(),
      sample_block = character()
    ))
  }

  parsed <- stringr::str_match(
    basename(files),
    "^SGRE-(HEAD|THOX)-(CRD|ISO)-([0-9]+)_MERGE_featurecounts[.]txt$"
  )
  tibble::tibble(
    sample_id = stringr::str_remove(basename(files), "_featurecounts[.]txt$"),
    count_file = files,
    tissue_code = parsed[, 2],
    rearing_code = parsed[, 3],
    replicate = as.integer(parsed[, 4])
  ) |>
    dplyr::filter(!is.na(tissue_code), !is.na(rearing_code)) |>
    dplyr::mutate(
      tissue = dplyr::case_when(
        tissue_code == "HEAD" ~ "Head",
        tissue_code == "THOX" ~ "Thorax",
        TRUE ~ tissue_code
      ),
      phase = dplyr::case_when(
        rearing_code == "CRD" ~ "Gregarious",
        rearing_code == "ISO" ~ "Solitarious",
        TRUE ~ rearing_code
      ),
      rearing_condition = dplyr::case_when(
        rearing_code == "CRD" ~ "Crowded",
        rearing_code == "ISO" ~ "Isolated",
        TRUE ~ rearing_code
      ),
      source = "Time-course head/thorax phase reference",
      display_group = paste0(
        tissue, ": ",
        dplyr::if_else(rearing_condition == "Crowded",
                       "crowded/gregarious", "isolated/solitarious")
      ),
      reference_role = paste(tissue, tolower(rearing_condition),
                             "rep", replicate),
      sample_label = paste0(substr(tissue, 1, 1), "-",
                            rearing_code, "-", replicate),
      sample_block = paste(tissue, "reference")
    ) |>
    dplyr::arrange(tissue, rearing_condition, replicate)
}

find_gaps <- function(groups) {
  groups <- as.character(groups)
  if (length(groups) < 2) {
    return(integer())
  }
  which(groups[-length(groups)] != groups[-1])
}
```

## Phase signatures

```{r phase-signatures}
phase_all <- dplyr::bind_rows(lapply(seq_len(nrow(phase_files)), function(i) {
  read_phase_table(phase_files$result_file[i], phase_files$tissue[i])
}))

phase_sig <- phase_all |>
  dplyr::filter(phase_significant)

phase_by_gene <- phase_sig |>
  dplyr::group_by(gene_id) |>
  dplyr::summarise(
    n_phase_tissues = dplyr::n_distinct(tissue),
    phase_tissues = paste(sort(unique(tissue)), collapse = " + "),
    phase_directions = paste(sort(unique(phase_direction)), collapse = " + "),
    phase_mean_log2FC = mean(phase_log2FoldChange, na.rm = TRUE),
    phase_max_abs_log2FC = max(abs(phase_log2FoldChange), na.rm = TRUE),
    head_phase_log2FC = phase_log2FoldChange[tissue == "Head"][1],
    thorax_phase_log2FC = phase_log2FoldChange[tissue == "Thorax"][1],
    head_phase_padj = phase_padj[tissue == "Head"][1],
    thorax_phase_padj = phase_padj[tissue == "Thorax"][1],
    .groups = "drop"
  ) |>
  dplyr::mutate(
    phase_consensus = dplyr::case_when(
      stringr::str_detect(phase_directions, "[+]") ~ "Discordant phase direction",
      phase_directions == "Gregarious-up" ~ "Gregarious-up",
      phase_directions == "Solitarious-up" ~ "Solitarious-up",
      TRUE ~ "No phase DEG"
    ),
    phase_class = dplyr::case_when(
      n_phase_tissues >= 2 ~ "Shared head/thorax phase DEG",
      phase_tissues == "Head" ~ "Head phase DEG only",
      phase_tissues == "Thorax" ~ "Thorax phase DEG only",
      TRUE ~ "No phase DEG"
    )
  )

phase_gene_all <- tibble::tibble(gene_id = unique(phase_all$gene_id)) |>
  dplyr::left_join(phase_by_gene, by = "gene_id") |>
  dplyr::mutate(
    n_phase_tissues = dplyr::coalesce(n_phase_tissues, 0L),
    phase_tissues = dplyr::coalesce(phase_tissues, "None"),
    phase_directions = dplyr::coalesce(phase_directions, "No phase DEG"),
    phase_consensus = dplyr::coalesce(phase_consensus, "No phase DEG"),
    phase_class = dplyr::coalesce(phase_class, "No phase DEG")
  )

phase_summary <- phase_sig |>
  dplyr::count(tissue, phase_direction, name = "n_genes")

write.csv(phase_summary, file.path(out_dir, "phase_signature_summary.csv"),
          row.names = FALSE)
write.csv(phase_gene_all, file.path(out_dir, "phase_signature_by_gene.csv"),
          row.names = FALSE)

knitr::kable(phase_summary, caption = "Phase DEG counts from the copied reference")
```

```{r phase-summary-plot, fig.width=8, fig.height=4.8}
phase_summary_plot <- ggplot(phase_summary,
                             aes(x = tissue, y = n_genes,
                                 fill = phase_direction)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.68,
           color = "grey25", linewidth = 0.25) +
  scale_fill_manual(values = phase_colors) +
  labs(
    title = "Reference phase signatures",
    subtitle = "Positive phase log2FC means higher in crowded/gregarious animals",
    x = NULL, y = "Number of phase DEGs", fill = "Phase direction"
  ) +
  theme_classic(base_size = 13) +
  theme(legend.position = "bottom")

phase_summary_plot
ggsave(file.path(out_dir, "phase_signature_summary.png"), phase_summary_plot,
       width = 8, height = 4.8, dpi = 300)
```

## Fat body DEG results

The page uses the newest `DEG sample-filter scenarios` output if it exists.
That lets this phase audit cover all samples, the current outlier-filtered
analysis, and the editable expanded-outlier scenario with the same code.

```{r fat-body-results}
scenario_dir <- latest_dir("^deg-sample-filter-scenarios_")
if (is.na(scenario_dir)) {
  stop("Render 12-deg-sample-filter-scenarios.Rmd before this page.")
}

fat_source <- file.path(scenario_dir, "all_deseq2_results.csv")
fat_results <- read.csv(fat_source, check.names = FALSE) |>
  dplyr::mutate(
    significant = !is.na(padj) &
      padj < params$alpha &
      abs(log2FoldChange) >= params$lfc_threshold,
    direction = dplyr::case_when(
      !significant ~ "Not DEG",
      log2FoldChange > 0 ~ "Upregulated",
      log2FoldChange < 0 ~ "Downregulated",
      TRUE ~ "Not DEG"
    )
  )

fat_sig <- fat_results |>
  dplyr::filter(significant)

knitr::kable(
  fat_sig |>
    dplyr::count(scenario, comparison_type, name = "n_significant_rows"),
  caption = "Significant fat body DEG rows used for phase layering"
)
```

## Layer phase signatures onto fat body DEGs

For every fat body comparison, the numerator is the condition named first in the
contrast. For example, `Diet 33: infected vs control` has infected as the
numerator. If a phase-linked fat body DEG moves in the same sign as the
gregarious-vs-solitarious phase signature, the numerator is described as more
gregarious-like for that gene. If it moves in the opposite sign, the numerator
is described as more solitarious-like.

```{r layered-results}
phase_layered <- fat_results |>
  dplyr::left_join(phase_gene_all, by = "gene_id") |>
  dplyr::mutate(
    phase_class = dplyr::coalesce(phase_class, "No phase reference"),
    phase_consensus = dplyr::coalesce(phase_consensus, "No phase reference"),
    n_phase_tissues = dplyr::coalesce(n_phase_tissues, 0L),
    phase_shift = dplyr::case_when(
      !significant | n_phase_tissues == 0 ~ "Not phase-associated",
      phase_consensus == "Discordant phase direction" ~ "Mixed head/thorax phase signal",
      sign(log2FoldChange) == sign(phase_mean_log2FC) ~ "Numerator more gregarious-like",
      sign(log2FoldChange) == -sign(phase_mean_log2FC) ~ "Numerator more solitarious-like",
      TRUE ~ "Not phase-associated"
    )
  )

phase_layered_sig <- phase_layered |>
  dplyr::filter(significant)

write.csv(phase_layered,
          gzfile(file.path(out_dir, "fat_body_degs_with_phase_layer_all_rows.csv.gz")),
          row.names = FALSE)
write.csv(phase_layered_sig, file.path(out_dir, "fat_body_degs_with_phase_layer_significant.csv"),
          row.names = FALSE)

phase_overlap_summary <- phase_layered_sig |>
  dplyr::count(scenario, comparison_type, comparison, phase_class, phase_shift,
               direction, name = "n_genes")

write.csv(phase_overlap_summary,
          file.path(out_dir, "phase_overlap_summary_by_comparison.csv"),
          row.names = FALSE)
```

```{r phase-overlap-table}
main_overlap_table <- phase_layered_sig |>
  dplyr::count(scenario, comparison_type, comparison, phase_class, name = "n_genes") |>
  tidyr::pivot_wider(names_from = phase_class, values_from = n_genes,
                     values_fill = 0) |>
  dplyr::arrange(scenario, comparison_type, comparison)

DT::datatable(
  main_overlap_table,
  rownames = FALSE,
  filter = "top",
  options = list(pageLength = 12, scrollX = TRUE),
  caption = "Fat body DEGs classified by independent phase signature"
)
```

```{r infection-phase-burden, fig.width=14, fig.height=7}
infection_phase_burden <- phase_layered_sig |>
  dplyr::filter(comparison_type == "Infection within diet") |>
  dplyr::mutate(
    comparison_plot = stringr::str_remove(comparison, ": infected vs control"),
    phase_class = factor(
      phase_class,
      levels = c("No phase DEG", "Head phase DEG only",
                 "Thorax phase DEG only", "Shared head/thorax phase DEG",
                 "No phase reference")
    )
  ) |>
  dplyr::count(scenario, comparison_plot, phase_class, name = "n_genes")

infection_phase_burden_plot <- ggplot(
  infection_phase_burden,
  aes(x = comparison_plot, y = n_genes, fill = phase_class)
) +
  geom_col(color = "grey20", linewidth = 0.2) +
  facet_wrap(~ scenario, nrow = 1) +
  scale_fill_manual(
    values = c(
      "No phase DEG" = "#D1D5DB",
      "Head phase DEG only" = "#F59E0B",
      "Thorax phase DEG only" = "#10B981",
      "Shared head/thorax phase DEG" = "#7C3AED",
      "No phase reference" = "#F3F4F6"
    )
  ) +
  labs(
    title = "Infection-within-diet DEG burden with phase-signature layer",
    subtitle = "Phase-linked DEGs are retained but flagged as possible density-state sensitive genes",
    x = NULL, y = "Number of significant fat body DEGs", fill = "Phase class"
  ) +
  theme_classic(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )

infection_phase_burden_plot
ggsave(file.path(out_dir, "infection_within_diet_phase_burden.png"),
       infection_phase_burden_plot, width = 14, height = 7, dpi = 300)
```

```{r infection-phase-shift, fig.width=13, fig.height=7}
phase_shift_counts <- phase_layered_sig |>
  dplyr::filter(comparison_type == "Infection within diet",
                phase_shift != "Not phase-associated") |>
  dplyr::mutate(
    comparison_plot = stringr::str_remove(comparison, ": infected vs control")
  ) |>
  dplyr::count(scenario, comparison_plot, phase_shift, direction, name = "n_genes")

phase_shift_plot <- ggplot(
  phase_shift_counts,
  aes(x = comparison_plot, y = n_genes, fill = phase_shift)
) +
  geom_col(position = position_dodge(width = 0.72), width = 0.64,
           color = "grey20", linewidth = 0.2) +
  facet_grid(direction ~ scenario, scales = "free_y") +
  scale_fill_manual(values = shift_colors) +
  labs(
    title = "Direction of phase-linked infection DEGs",
    subtitle = "For infection contrasts, the numerator is infected and the denominator is control",
    x = NULL, y = "Number of phase-linked DEGs", fill = "Phase interpretation"
  ) +
  theme_classic(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )

phase_shift_plot
ggsave(file.path(out_dir, "infection_phase_direction_shift.png"),
       phase_shift_plot, width = 13, height = 7, dpi = 300)
```

## Enrichment and concordance tests

The overlap tests ask whether fat body DEGs are enriched for phase-DEG genes
more than expected from the analyzed gene universe. The sign tests ask whether
overlapping genes tend to move in the same direction as the phase signature or
the opposite direction.

```{r overlap-statistics}
universe_genes <- intersect(unique(fat_results$gene_id), unique(phase_gene_all$gene_id))

phase_sets <- list(
  "Any phase DEG" = phase_gene_all$gene_id[phase_gene_all$n_phase_tissues > 0],
  "Head phase DEG" = phase_sig$gene_id[phase_sig$tissue == "Head"],
  "Thorax phase DEG" = phase_sig$gene_id[phase_sig$tissue == "Thorax"],
  "Shared head/thorax phase DEG" =
    phase_gene_all$gene_id[phase_gene_all$phase_class == "Shared head/thorax phase DEG"]
)

fisher_one <- function(sig_genes, phase_genes, universe) {
  sig_genes <- intersect(unique(sig_genes), universe)
  phase_genes <- intersect(unique(phase_genes), universe)
  overlap <- length(intersect(sig_genes, phase_genes))
  tab <- matrix(c(
    overlap,
    length(setdiff(sig_genes, phase_genes)),
    length(setdiff(phase_genes, sig_genes)),
    length(setdiff(universe, union(sig_genes, phase_genes)))
  ), nrow = 2)
  test <- fisher.test(tab, alternative = "greater")
  tibble::tibble(
    n_fat_degs = length(sig_genes),
    n_phase_genes = length(phase_genes),
    n_overlap = overlap,
    odds_ratio = unname(test$estimate),
    p_value = test$p.value
  )
}

overlap_tests <- phase_layered_sig |>
  dplyr::group_by(scenario, comparison_type, comparison) |>
  dplyr::summarise(sig_genes = list(unique(gene_id)), .groups = "drop") |>
  tidyr::crossing(phase_set = names(phase_sets)) |>
  dplyr::rowwise() |>
  dplyr::mutate(stats = list(fisher_one(sig_genes, phase_sets[[phase_set]],
                                        universe_genes))) |>
  tidyr::unnest(stats) |>
  dplyr::ungroup() |>
  dplyr::mutate(p_adj = p.adjust(p_value, method = "BH")) |>
  dplyr::select(-sig_genes) |>
  dplyr::arrange(p_adj, p_value)

write.csv(overlap_tests, file.path(out_dir, "phase_overlap_fisher_tests.csv"),
          row.names = FALSE)

DT::datatable(
  overlap_tests,
  rownames = FALSE,
  filter = "top",
  options = list(pageLength = 12, scrollX = TRUE),
  caption = "Fisher exact tests for enrichment of phase genes among fat body DEGs"
)
```

```{r concordance-statistics}
phase_long_for_tests <- phase_all |>
  dplyr::filter(phase_significant) |>
  dplyr::select(gene_id, phase_tissue = tissue,
                phase_log2FoldChange, phase_direction)

phase_fat_long <- phase_layered_sig |>
  dplyr::inner_join(phase_long_for_tests, by = "gene_id") |>
  dplyr::mutate(
    same_direction = sign(log2FoldChange) == sign(phase_log2FoldChange)
  )

sign_test_one <- function(same_direction) {
  same_direction <- same_direction[!is.na(same_direction)]
  if (length(same_direction) < 2) {
    return(tibble::tibble(n_overlap = length(same_direction),
                          n_same_direction = sum(same_direction),
                          n_opposite_direction = sum(!same_direction),
                          p_value = NA_real_))
  }
  test <- binom.test(sum(same_direction), length(same_direction), p = 0.5)
  tibble::tibble(n_overlap = length(same_direction),
                 n_same_direction = sum(same_direction),
                 n_opposite_direction = sum(!same_direction),
                 p_value = test$p.value)
}

concordance_tests <- phase_fat_long |>
  dplyr::group_by(scenario, comparison_type, comparison, phase_tissue) |>
  dplyr::summarise(stats = list(sign_test_one(same_direction)),
                   .groups = "drop") |>
  tidyr::unnest(stats) |>
  dplyr::mutate(p_adj = p.adjust(p_value, method = "BH")) |>
  dplyr::arrange(p_adj, p_value)

write.csv(concordance_tests,
          file.path(out_dir, "phase_fatbody_sign_concordance_tests.csv"),
          row.names = FALSE)

DT::datatable(
  concordance_tests,
  rownames = FALSE,
  filter = "top",
  options = list(pageLength = 12, scrollX = TRUE),
  caption = "Sign concordance between fat body DEGs and head/thorax phase signatures"
)
```

## Fat body log2FC versus phase log2FC

This scatterplot focuses on the selected infection-within-diet DEG scenario:
`r params$analysis_scenario`. Points to the same side of zero on both axes indicate that infected
animals move in the same direction as the gregarious phase signature for that
gene. Points in opposite quadrants indicate an infected state moving opposite
to the gregarious signature, which is the more solitarization-like pattern.

```{r lfc-scatter, fig.width=13, fig.height=8}
scatter_data <- phase_fat_long |>
  dplyr::filter(
    scenario == params$analysis_scenario,
    comparison_type == "Infection within diet"
  ) |>
  dplyr::mutate(
    diet = stringr::str_extract(comparison, "Diet [0-9]+"),
    direction_relation = dplyr::case_when(
      same_direction ~ "Numerator more gregarious-like",
      !same_direction ~ "Numerator more solitarious-like",
      TRUE ~ "Not phase-associated"
    )
  )

lfc_scatter_plot <- ggplot(
  scatter_data,
  aes(x = phase_log2FoldChange, y = log2FoldChange,
      color = direction_relation)
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey45") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey45") +
  geom_point(alpha = 0.72, size = 2.2) +
  facet_grid(phase_tissue ~ diet) +
  scale_color_manual(values = shift_colors) +
  labs(
    title = "Infection response compared with independent phase signatures",
    subtitle = paste0(
      "DEG scenario: ", params$analysis_scenario,
      "; positive phase log2FC means gregarious-up"
    ),
    x = "Phase log2FC: crowded/gregarious vs isolated/solitarious",
    y = "Fat body log2FC: infected vs control",
    color = "Interpretation"
  ) +
  theme_classic(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )

lfc_scatter_plot
ggsave(file.path(out_dir, "infection_fatbody_vs_phase_lfc_scatter.png"),
       lfc_scatter_plot, width = 13, height = 8, dpi = 300)
```

## Annotated candidate gene table

The table below highlights significant fat body DEGs that are also phase DEGs.
The priority score favors genes with phase support in both tissues, large fat
body effect size, strong fat body adjusted p-value, and large phase effect.

```{r gene-annotation}
if (file.exists(eggnog_file) && file.exists(gff_file)) {
  eggnog <- read_emapper(eggnog_file)
  protein_to_gene <- parse_cds_gene_bridge(gff_file)
  annotation <- eggnog |>
    dplyr::left_join(protein_to_gene, by = "protein_id") |>
    dplyr::filter(!is.na(gene_id)) |>
    dplyr::distinct(gene_id, .keep_all = TRUE) |>
    dplyr::transmute(
      gene_id,
      Preferred_name = dplyr::na_if(Preferred_name, "-"),
      Description = dplyr::na_if(Description, "-"),
      GOs = dplyr::na_if(GOs, "-"),
      KEGG_ko = dplyr::na_if(KEGG_ko, "-"),
      KEGG_Pathway = dplyr::na_if(KEGG_Pathway, "-"),
      PFAMs = dplyr::na_if(PFAMs, "-")
    )
} else {
  annotation <- tibble::tibble(
    gene_id = character(),
    Preferred_name = character(),
    Description = character(),
    GOs = character(),
    KEGG_ko = character(),
    KEGG_Pathway = character(),
    PFAMs = character()
  )
}

candidate_phase_degs <- phase_layered_sig |>
  dplyr::filter(n_phase_tissues > 0) |>
  dplyr::left_join(annotation, by = "gene_id") |>
  dplyr::mutate(
    neglog10_padj = safe_neglog10(padj),
    priority_score = n_phase_tissues * 10 +
      abs(log2FoldChange) +
      phase_max_abs_log2FC +
      dplyr::coalesce(neglog10_padj, 0),
    phase_interpretation = dplyr::case_when(
      phase_shift == "Numerator more solitarious-like" &
        comparison_type == "Infection within diet" ~
        "Infected expression is more solitarious-like than control",
      phase_shift == "Numerator more gregarious-like" &
        comparison_type == "Infection within diet" ~
        "Infected expression is more gregarious-like than control",
      TRUE ~ phase_shift
    )
  ) |>
  dplyr::arrange(dplyr::desc(priority_score), scenario, comparison)

write.csv(candidate_phase_degs,
          file.path(out_dir, "phase_linked_fat_body_deg_candidates.csv"),
          row.names = FALSE)

DT::datatable(
  candidate_phase_degs |>
    dplyr::select(
      scenario, comparison_type, comparison, gene_id, Preferred_name,
      Description, log2FoldChange, padj, direction, phase_class,
      phase_consensus, phase_shift, phase_interpretation, phase_tissues,
      phase_mean_log2FC, priority_score
    ),
  rownames = FALSE,
  filter = "top",
  options = list(pageLength = 20, scrollX = TRUE),
  caption = "Phase-linked fat body DEG candidates"
)
```

## Curated VST heatmap

This heatmap uses the `r params$analysis_scenario` DEG scenario and
`r sample_filter_label`. Genes are selected from significant
infection-within-diet DEGs that also overlap a phase signature, with shared
head/thorax phase DEGs prioritized first. Samples are grouped by infection
status and then diet.

```{r phase-linked-vst-heatmap, fig.width=15, fig.height=12}
read_metadata <- function(path) {
  tab <- read.delim(path, check.names = FALSE)
  tab <- tab[, !is.na(names(tab)) & nzchar(names(tab)), drop = FALSE]
  tab$label <- as.character(tab$label)
  tab$Diet <- factor(tab$Diet, levels = c("33", "50", "83"))
  tab$Treatment <- factor(tab$Treatment, levels = c("Infected", "Control"))
  tab
}

metadata_full <- read_metadata(metadata_file)
if (isTRUE(params$exclude_experiment_outliers)) {
  metadata_full <- metadata_full |>
    dplyr::filter(!label %in% current_outliers)
}
metadata_full <- metadata_full |>
  dplyr::arrange(Treatment, Diet, label)

counts_raw <- read.csv(counts_file, check.names = FALSE)
counts_all <- as.matrix(counts_raw[, -1])
rownames(counts_all) <- counts_raw[[1]]
storage.mode(counts_all) <- "integer"
colnames(counts_all) <- sub("_MERGE$", "", sub("^mehreen_", "",
                                                colnames(counts_all)))
counts_heat <- counts_all[, metadata_full$label, drop = FALSE]
rownames(metadata_full) <- metadata_full$label

heatmap_genes <- candidate_phase_degs |>
  dplyr::filter(
    scenario == params$analysis_scenario,
    comparison_type == "Infection within diet"
  ) |>
  dplyr::group_by(gene_id) |>
  dplyr::slice_max(order_by = priority_score, n = 1, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::arrange(
    dplyr::desc(n_phase_tissues),
    phase_consensus,
    dplyr::desc(priority_score)
  ) |>
  dplyr::slice_head(n = params$heatmap_gene_n)

if (nrow(heatmap_genes) > 1) {
  keep <- rowSums(counts_heat) >= 10
  dds_heat <- DESeq2::DESeqDataSetFromMatrix(
    countData = counts_heat[keep, ],
    colData = metadata_full,
    design = ~ Diet + Treatment
  )
  vst_heat <- DESeq2::vst(dds_heat, blind = TRUE)
  vst_mat <- SummarizedExperiment::assay(vst_heat)
  heatmap_gene_ids <- intersect(heatmap_genes$gene_id, rownames(vst_mat))
  zmat <- t(scale(t(vst_mat[heatmap_gene_ids, , drop = FALSE])))
  zmat[is.na(zmat)] <- 0

  gene_labels <- heatmap_genes |>
    dplyr::filter(gene_id %in% heatmap_gene_ids) |>
    dplyr::mutate(
      label = paste0(
        gene_id, " | ",
        dplyr::coalesce(Preferred_name, "unannotated"), " | ",
        stringr::str_trunc(dplyr::coalesce(Description, "no description"), 48)
      )
    ) |>
    dplyr::select(gene_id, label, phase_consensus, phase_class, phase_shift)

  heatmap_long <- as.data.frame(zmat) |>
    tibble::rownames_to_column("gene_id") |>
    tidyr::pivot_longer(-gene_id, names_to = "sample", values_to = "z_vst") |>
    dplyr::left_join(gene_labels, by = "gene_id") |>
    dplyr::left_join(
      metadata_full |>
        tibble::rownames_to_column("sample") |>
        dplyr::select(sample, Diet, Treatment),
      by = "sample"
    ) |>
    dplyr::mutate(
      sample = factor(sample, levels = metadata_full$label),
      label = factor(label, levels = rev(gene_labels$label)),
      condition = paste(Treatment, Diet, sep = "_")
    )

  phase_heatmap_plot <- ggplot(heatmap_long,
                               aes(x = sample, y = label, fill = z_vst)) +
    geom_tile(color = "white", linewidth = 0.08) +
    geom_tile(aes(y = label, x = sample, color = condition),
              fill = NA, linewidth = 0.18, show.legend = FALSE) +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                         midpoint = 0, limits = c(-2.5, 2.5),
                         oob = scales::squish, name = "Row-z VST") +
    scale_color_manual(values = condition_colors) +
    labs(
      title = "Curated phase-linked infection DEG heatmap",
      subtitle = paste0(
        "DEG scenario: ", params$analysis_scenario,
        "; ", sample_filter_label,
        "; samples grouped by infection status and diet"
      ),
      x = NULL, y = NULL
    ) +
    theme_classic(base_size = 11) +
    theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
      axis.text.y = element_text(size = 7),
      legend.position = "bottom",
      plot.title = element_text(face = "bold")
    )

  phase_heatmap_plot
  ggsave(file.path(out_dir, "phase_linked_infection_deg_vst_heatmap.png"),
         phase_heatmap_plot, width = 15, height = 12, dpi = 300)
  write.csv(heatmap_genes,
            file.path(out_dir, "phase_linked_heatmap_genes.csv"),
            row.names = FALSE)
} else {
  message("Not enough phase-linked infection DEGs for the heatmap.")
}
```

## Full experiment DEG diagnostic heatmap

This heatmap is deliberately not curated. It uses every significant fat body DEG
from the selected scenario, across infection-within-diet and diet-contrast
comparisons. The DEG calls are still from the outlier-removed DESeq2 scenario,
but the heatmap displays all experiment libraries, including the outlier-flagged
samples, so their expression profiles can be inspected against the retained
samples. Sample columns are kept in experimental order: infected samples first,
then controls, with diet substructure inside each treatment. Gene rows are
clustered to help reveal unusual expression blocks or single-sample effects.

Because the full DEG set contains thousands of genes, row names are hidden in the
image. The searchable table below the figure keeps the heatmap row order,
annotation, DEG context, and the sample with the strongest row-z-score deviation
for each gene.

```{r full-experiment-deg-diagnostic-heatmap, fig.width=18, fig.height=22}
if (exists("metadata_full")) {
  full_experiment_display_metadata <- read_metadata(metadata_file) |>
    dplyr::mutate(
      Outlier_status = dplyr::if_else(
        label %in% current_outliers,
        "Outlier displayed only",
        "Included in DESeq2 model"
      ),
      Outlier_status = factor(
        Outlier_status,
        levels = c("Included in DESeq2 model", "Outlier displayed only")
      )
    ) |>
    dplyr::arrange(Treatment, Diet, Outlier_status, label)

  full_experiment_counts_raw <- read.csv(counts_file, check.names = FALSE)
  full_experiment_counts_all <- as.matrix(full_experiment_counts_raw[, -1])
  rownames(full_experiment_counts_all) <- full_experiment_counts_raw[[1]]
  storage.mode(full_experiment_counts_all) <- "integer"
  colnames(full_experiment_counts_all) <- sub(
    "_MERGE$",
    "",
    sub("^mehreen_", "", colnames(full_experiment_counts_all))
  )

  full_experiment_display_counts <- full_experiment_counts_all[
    ,
    full_experiment_display_metadata$label,
    drop = FALSE
  ]
  rownames(full_experiment_display_metadata) <-
    full_experiment_display_metadata$label

  keep_full_experiment_display <- rowSums(full_experiment_display_counts) >=
    params$min_total_count
  dds_full_experiment_display <- DESeq2::DESeqDataSetFromMatrix(
    countData = full_experiment_display_counts[keep_full_experiment_display, ],
    colData = full_experiment_display_metadata,
    design = ~ Diet + Treatment
  )
  vst_full_experiment_display <- DESeq2::vst(dds_full_experiment_display,
                                             blind = TRUE)
  vst_full_experiment_display_mat <- SummarizedExperiment::assay(
    vst_full_experiment_display
  )

  full_experiment_deg_gene_table <- phase_layered_sig |>
    dplyr::filter(scenario == params$analysis_scenario) |>
    dplyr::group_by(gene_id) |>
    dplyr::summarise(
      n_deg_contrasts = dplyr::n_distinct(comparison),
      has_infection_deg = any(comparison_type == "Infection within diet"),
      has_control_diet_deg = any(comparison_type == "Diet contrasts among control"),
      has_infected_diet_deg = any(comparison_type == "Diet contrasts among infected"),
      comparison_types = paste(sort(unique(comparison_type)), collapse = "; "),
      comparisons = paste(sort(unique(comparison)), collapse = "; "),
      directions = paste(sort(unique(direction)), collapse = "; "),
      max_abs_log2FC = max(abs(log2FoldChange), na.rm = TRUE),
      min_padj = min(padj, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::left_join(annotation |> dplyr::select(gene_id, Description),
                     by = "gene_id") |>
    dplyr::left_join(phase_gene_all |> dplyr::select(gene_id, phase_class),
                     by = "gene_id") |>
    dplyr::mutate(
      Description = dplyr::coalesce(Description, "no description"),
      DEG_context = dplyr::case_when(
        has_infection_deg & (has_control_diet_deg | has_infected_diet_deg) ~
          "Infection + diet",
        has_infection_deg ~ "Infection within diet",
        has_control_diet_deg & has_infected_diet_deg ~
          "Diet in controls + infected",
        has_control_diet_deg ~ "Diet in controls",
        has_infected_diet_deg ~ "Diet in infected",
        TRUE ~ "Other fat body DEG"
      ),
      DEG_context = factor(
        DEG_context,
        levels = c("Infection + diet", "Infection within diet",
                   "Diet in controls", "Diet in infected",
                   "Diet in controls + infected", "Other fat body DEG")
      ),
      phase_class = dplyr::coalesce(phase_class, "No phase DEG"),
      phase_class = factor(
        phase_class,
        levels = c("Shared head/thorax phase DEG", "Head phase DEG only",
                   "Thorax phase DEG only", "No phase DEG")
      )
    ) |>
    dplyr::arrange(DEG_context, phase_class, min_padj,
                   dplyr::desc(max_abs_log2FC))

  full_experiment_deg_ids <- intersect(
    full_experiment_deg_gene_table$gene_id,
    rownames(vst_full_experiment_display_mat)
  )

  if (length(full_experiment_deg_ids) > 1) {
    full_experiment_deg_gene_table <- full_experiment_deg_gene_table |>
      dplyr::filter(gene_id %in% full_experiment_deg_ids)

    full_experiment_sample_order <- full_experiment_display_metadata$label
    full_experiment_sample_annotation <- full_experiment_display_metadata[
      full_experiment_sample_order,
    ] |>
      dplyr::transmute(
        Treatment = as.character(Treatment),
        Diet = as.character(Diet),
        Outlier_status = as.character(Outlier_status)
      )
    full_experiment_sample_annotation <- as.data.frame(
      full_experiment_sample_annotation,
      stringsAsFactors = FALSE
    )
    rownames(full_experiment_sample_annotation) <- full_experiment_sample_order

    full_experiment_gene_annotation <- full_experiment_deg_gene_table |>
      dplyr::transmute(
        DEG_context = as.character(DEG_context),
        Phase_class = as.character(phase_class)
      )
    full_experiment_gene_annotation <- as.data.frame(
      full_experiment_gene_annotation,
      stringsAsFactors = FALSE
    )
    rownames(full_experiment_gene_annotation) <- full_experiment_deg_ids

    full_experiment_sample_gaps <- find_gaps(paste(
      full_experiment_sample_annotation$Treatment,
      full_experiment_sample_annotation$Diet,
      full_experiment_sample_annotation$Outlier_status
    ))

    full_experiment_mat <- vst_full_experiment_display_mat[
      full_experiment_deg_ids,
      full_experiment_sample_order,
      drop = FALSE
    ]
    full_experiment_zmat <- t(scale(t(full_experiment_mat)))
    full_experiment_zmat[is.na(full_experiment_zmat)] <- 0
    storage.mode(full_experiment_zmat) <- "double"
    max_deviation_index <- max.col(abs(full_experiment_zmat),
                                   ties.method = "first")
    max_deviation_value <- full_experiment_zmat[
      cbind(seq_len(nrow(full_experiment_zmat)), max_deviation_index)
    ]
    full_experiment_zmat_plot <- pmax(pmin(full_experiment_zmat, 3), -3)
    full_experiment_zmat_plot <- as.matrix(full_experiment_zmat_plot)
    storage.mode(full_experiment_zmat_plot) <- "double"

    full_experiment_heatmap_colors <- grDevices::colorRampPalette(
      c("#237C8E", "#F7F3E8", "#C85A54")
    )(101)
    full_experiment_heatmap_breaks <- seq(-3, 3, length.out = 102)

    full_experiment_annotation_colors <- list(
      Treatment = treatment_colors,
      Diet = diet_colors,
      DEG_context = c(
        "Infection + diet" = "#111827",
        "Infection within diet" = "#C85A54",
        "Diet in controls" = "#2F8F5B",
        "Diet in infected" = "#7B4FA3",
        "Diet in controls + infected" = "#E6862E",
        "Other fat body DEG" = "#9CA3AF"
      ),
      Phase_class = c(
        "Shared head/thorax phase DEG" = "#3B1F6D",
        "Head phase DEG only" = "#009E9A",
        "Thorax phase DEG only" = "#C51B7D",
        "No phase DEG" = "#D1D5DB"
      ),
      Outlier_status = c(
        "Included in DESeq2 model" = "#D1D5DB",
        "Outlier displayed only" = "#E41A1C"
      )
    )

    full_experiment_heatmap <- pheatmap::pheatmap(
      full_experiment_zmat_plot,
      color = full_experiment_heatmap_colors,
      breaks = full_experiment_heatmap_breaks,
      cluster_rows = TRUE,
      cluster_cols = FALSE,
      gaps_col = full_experiment_sample_gaps,
      show_rownames = FALSE,
      show_colnames = TRUE,
      fontsize_col = 6,
      annotation_row = full_experiment_gene_annotation,
      annotation_col = full_experiment_sample_annotation,
      annotation_colors = full_experiment_annotation_colors,
      main = paste0("Full fat body DEG heatmap - ",
                    params$analysis_scenario,
                    " with outliers displayed"),
      filename = file.path(out_dir,
                           "full_experiment_deg_diagnostic_heatmap.png"),
      width = 18,
      height = 22,
      silent = TRUE
    )

    heatmap_row_order <- if (!is.null(full_experiment_heatmap$tree_row)) {
      rownames(full_experiment_zmat_plot)[full_experiment_heatmap$tree_row$order]
    } else {
      rownames(full_experiment_zmat_plot)
    }

    full_experiment_gene_table_ordered <- full_experiment_deg_gene_table |>
      dplyr::mutate(
        max_abs_sample_z = abs(max_deviation_value[
          match(gene_id, rownames(full_experiment_zmat))
        ]),
        sample_with_max_abs_z = full_experiment_sample_order[
          max_deviation_index[match(gene_id, rownames(full_experiment_zmat))]
        ],
        sample_deviation_direction = dplyr::if_else(
          max_deviation_value[match(gene_id, rownames(full_experiment_zmat))] > 0,
          "Higher than gene average",
          "Lower than gene average"
        )
      ) |>
      dplyr::slice(match(heatmap_row_order, gene_id)) |>
      dplyr::mutate(heatmap_row_order = dplyr::row_number()) |>
      dplyr::select(heatmap_row_order, gene_id, Description, DEG_context,
                    phase_class, n_deg_contrasts, comparisons, directions,
                    max_abs_log2FC, min_padj, max_abs_sample_z,
                    sample_with_max_abs_z, sample_deviation_direction)

    write.csv(
      full_experiment_gene_table_ordered,
      file.path(out_dir,
                "full_experiment_deg_diagnostic_heatmap_gene_order.csv"),
      row.names = FALSE
    )
    utils::write.csv(
      tibble::rownames_to_column(as.data.frame(full_experiment_zmat),
                                 "gene_id"),
      gzfile(file.path(out_dir,
                       "full_experiment_deg_diagnostic_z_vst_matrix.csv.gz")),
      row.names = FALSE
    )
    write.csv(
      full_experiment_display_metadata,
      file.path(out_dir,
                "full_experiment_deg_diagnostic_heatmap_display_samples.csv"),
      row.names = FALSE
    )

    full_experiment_display_dir <- file.path(
      project_dir, "docs", "figure", params$display_rmd_dir
    )
    dir.create(full_experiment_display_dir, recursive = TRUE,
               showWarnings = FALSE)
    full_experiment_display_png <- file.path(
      full_experiment_display_dir,
      "full-experiment-deg-diagnostic-heatmap-display.png"
    )
    file.copy(
      file.path(out_dir, "full_experiment_deg_diagnostic_heatmap.png"),
      full_experiment_display_png,
      overwrite = TRUE
    )
    full_experiment_heatmap_src <- file.path(
      "figure", params$display_rmd_dir,
      "full-experiment-deg-diagnostic-heatmap-display.png"
    )
  } else {
    message("Not enough fat body DEGs were available for a full diagnostic heatmap.")
  }
}
```

```{r full-experiment-deg-diagnostic-heatmap-display, echo=FALSE, results='asis'}
if (exists("full_experiment_heatmap_src")) {
  cat(paste0(
    '<p><img src="', full_experiment_heatmap_src,
    '" alt="Full non-curated fat body DEG diagnostic heatmap" ',
    'style="display:block;margin:auto;max-width:100%;height:auto;" /></p>'
  ))
}
```

```{r full-experiment-deg-diagnostic-table}
if (exists("full_experiment_gene_table_ordered")) {
  DT::datatable(
    full_experiment_gene_table_ordered,
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 20, scrollX = TRUE),
    caption = "Full diagnostic heatmap gene order and strongest single-sample deviations"
  )
}
```

## Time-course fat body single-pair reference

The time-course project contains two non-`FULL` *S. gregaria* fat body samples:
one gregarious/crowded control and one solitarious/isolated control. This is
not enough replication for a formal DEG analysis, but it is useful as a
tissue-matched qualitative phase anchor.

The main time-course metadata labels `GREG_S_ICT_9_FAT_MERGE` as gregarious,
but the sample name and the derived time-course metadata identify it as the
isolated/solitarious control. This page uses the corrected local metadata below.

```{r timecourse-fatbody-reference-files}
timecourse_reference_metadata <- file.path(
  timecourse_fatbody_dir,
  "gregaria_FAT_phase_reference_metadata.csv"
)

timecourse_ref_meta <- read.csv(timecourse_reference_metadata,
                                check.names = FALSE) |>
  dplyr::filter(tissue == "FAT")

timecourse_ref_meta <- timecourse_ref_meta |>
  dplyr::mutate(
    count_file = file.path(timecourse_fatbody_dir, "readcounts", file_name),
    file_present = file.exists(count_file)
  )

knitr::kable(
  timecourse_ref_meta |>
    dplyr::select(sample_id, phase, rearing_condition, reference_role,
                  file_name, file_present),
  caption = "Time-course fat body reference samples"
)
```

```{r timecourse-fatbody-pca-heatmap, fig.width=10, fig.height=8}
if (all(timecourse_ref_meta$file_present)) {
  reference_count_tables <- Map(
    read_featurecounts_gene_counts,
    timecourse_ref_meta$count_file,
    timecourse_ref_meta$sample_id
  )
  timecourse_counts_df <- Reduce(function(x, y) dplyr::full_join(x, y, by = "gene_id"),
                                 reference_count_tables)
  timecourse_counts <- as.matrix(timecourse_counts_df[, -1, drop = FALSE])
  rownames(timecourse_counts) <- timecourse_counts_df$gene_id
  timecourse_counts[is.na(timecourse_counts)] <- 0L
  storage.mode(timecourse_counts) <- "integer"

  diet_counts_current <- counts_all[, metadata_full$label, drop = FALSE]
  shared_genes <- intersect(rownames(diet_counts_current), rownames(timecourse_counts))
  combined_counts <- cbind(
    diet_counts_current[shared_genes, , drop = FALSE],
    timecourse_counts[shared_genes, timecourse_ref_meta$sample_id, drop = FALSE]
  )

  diet_metadata_plot <- metadata_full |>
    dplyr::mutate(
      sample_id = label,
      source = "Diet-infection fat body",
      display_group = paste(Treatment, Diet, sep = "_"),
      reference_role = paste(Treatment, "diet", Diet),
      phase = NA_character_,
      rearing_condition = NA_character_
    ) |>
    dplyr::select(sample_id, source, display_group, reference_role,
                  phase, rearing_condition)

  timecourse_metadata_plot <- timecourse_ref_meta |>
    dplyr::mutate(
      source = "Time-course single-pair fat body",
      display_group = dplyr::case_when(
        phase == "Gregarious" ~ "Gregarious reference",
        phase == "Solitarious" ~ "Solitarious reference",
        TRUE ~ "Reference"
      )
    ) |>
    dplyr::select(sample_id, source, display_group, reference_role,
                  phase, rearing_condition)

  combined_metadata <- dplyr::bind_rows(diet_metadata_plot, timecourse_metadata_plot)
  combined_metadata <- combined_metadata[match(colnames(combined_counts),
                                               combined_metadata$sample_id), ]
  rownames(combined_metadata) <- combined_metadata$sample_id

  keep_combined <- rowSums(combined_counts) >= 10
  dds_combined <- DESeq2::DESeqDataSetFromMatrix(
    countData = combined_counts[keep_combined, , drop = FALSE],
    colData = combined_metadata,
    design = ~ 1
  )
  vst_combined <- DESeq2::vst(dds_combined, blind = TRUE)
  vst_combined_mat <- SummarizedExperiment::assay(vst_combined)

  pca <- prcomp(t(vst_combined_mat), center = TRUE, scale. = FALSE)
  pct <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)
  pca_scores <- as.data.frame(pca$x[, 1:4, drop = FALSE]) |>
    tibble::rownames_to_column("sample_id") |>
    dplyr::left_join(combined_metadata |>
                       tibble::rownames_to_column("sample_id_meta"),
                     by = c("sample_id" = "sample_id_meta"))

  write.csv(pca_scores,
            file.path(out_dir, "timecourse_fatbody_reference_combined_pca_scores.csv"),
            row.names = FALSE)

  pca_reference_labels <- pca_scores |>
    dplyr::filter(source == "Time-course single-pair fat body") |>
    dplyr::mutate(
      label_x = PC1,
      label_y = PC2 + dplyr::if_else(phase == "Gregarious", 7, -7)
    )

  pca_plot <- ggplot(
    pca_scores,
    aes(x = PC1, y = PC2, color = display_group, shape = source)
  ) +
    geom_point(size = 3.2, alpha = 0.84) +
    geom_segment(
      data = pca_reference_labels,
      aes(x = PC1, y = PC2, xend = label_x, yend = label_y),
      inherit.aes = FALSE,
      color = "#4B5563",
      linewidth = 0.25
    ) +
    geom_label(
      data = pca_reference_labels,
      aes(x = label_x, y = label_y, label = reference_role,
          color = display_group),
      inherit.aes = FALSE,
      size = 3.5,
      fill = "white",
      label.size = 0.15,
      show.legend = FALSE
    ) +
    scale_color_manual(values = condition_colors) +
    scale_x_continuous(expand = ggplot2::expansion(mult = 0.16)) +
    scale_y_continuous(expand = ggplot2::expansion(mult = 0.12)) +
    scale_shape_manual(values = c("Diet-infection fat body" = 16,
                                  "Time-course single-pair fat body" = 17)) +
    labs(
      title = "PCA with time-course fat body phase reference",
      subtitle = "Exploratory only: one gregarious/crowded and one solitarious/isolated fat body sample",
      x = paste0("PC1 (", pct[1], "%)"),
      y = paste0("PC2 (", pct[2], "%)"),
      color = "Sample group",
      shape = "Source"
    ) +
    guides(
      color = ggplot2::guide_legend(nrow = 2, byrow = TRUE, order = 2),
      shape = ggplot2::guide_legend(order = 1)
    ) +
    coord_cartesian(clip = "off") +
    theme_classic(base_size = 13) +
    theme(
      legend.position = "bottom",
      legend.box = "vertical",
      plot.margin = ggplot2::margin(10, 80, 10, 10)
    )

  print(pca_plot)
  ggsave(file.path(out_dir, "timecourse_fatbody_reference_combined_pca.png"),
         pca_plot, width = 13, height = 8, dpi = 300)

  timecourse_direction_all <- tibble::tibble(
    gene_id = rownames(vst_combined_mat),
    timecourse_fatbody_gregarious_minus_solitarious =
      vst_combined_mat[, "GREG_G_CCT_11_FAT"] -
      vst_combined_mat[, "GREG_S_ICT_9_FAT"]
  ) |>
    dplyr::mutate(
      abs_timecourse_delta = abs(timecourse_fatbody_gregarious_minus_solitarious),
      tissue_matched_direction = dplyr::case_when(
        timecourse_fatbody_gregarious_minus_solitarious > 0 ~
          "Higher in gregarious/crowded reference",
        timecourse_fatbody_gregarious_minus_solitarious < 0 ~
          "Higher in solitarious/isolated reference",
        TRUE ~ "No difference in single-pair reference"
      ),
      exploratory_shift = abs_timecourse_delta >= params$timecourse_vst_delta_threshold
    )

  write.csv(
    timecourse_direction_all,
    file.path(out_dir, "timecourse_fatbody_single_pair_vst_direction_all_genes.csv"),
    row.names = FALSE
  )

  if (exists("heatmap_genes") && nrow(heatmap_genes) > 1) {
    ref_heatmap_genes <- heatmap_genes |>
      dplyr::filter(gene_id %in% rownames(vst_combined_mat)) |>
      dplyr::arrange(dplyr::desc(n_phase_tissues),
                     phase_consensus,
                     dplyr::desc(priority_score)) |>
      dplyr::slice_head(n = params$heatmap_gene_n)

    ref_gene_ids <- ref_heatmap_genes$gene_id
    ref_zmat <- t(scale(t(vst_combined_mat[ref_gene_ids, , drop = FALSE])))
    ref_zmat[is.na(ref_zmat)] <- 0

    ref_gene_labels <- ref_heatmap_genes |>
      dplyr::mutate(
        label = paste0(
          gene_id, " | ",
          dplyr::coalesce(Preferred_name, "unannotated"), " | ",
          stringr::str_trunc(dplyr::coalesce(Description, "no description"), 54)
        )
      ) |>
      dplyr::select(gene_id, label)

    ref_heatmap_long <- as.data.frame(
      ref_zmat[, timecourse_ref_meta$sample_id, drop = FALSE]
    ) |>
      tibble::rownames_to_column("gene_id") |>
      tidyr::pivot_longer(-gene_id, names_to = "sample_id",
                          values_to = "z_vst") |>
      dplyr::left_join(ref_gene_labels, by = "gene_id") |>
      dplyr::left_join(timecourse_metadata_plot, by = "sample_id") |>
      dplyr::mutate(
        label = factor(label, levels = rev(ref_gene_labels$label)),
        sample_label = factor(reference_role,
                              levels = timecourse_ref_meta$reference_role)
      )

    ref_heatmap_plot <- ggplot(ref_heatmap_long,
                               aes(x = sample_label, y = label, fill = z_vst)) +
      geom_tile(color = "white", linewidth = 0.12) +
      scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                           midpoint = 0, limits = c(-2.5, 2.5),
                           oob = scales::squish, name = "Row-z VST") +
      labs(
        title = "Phase-linked candidate genes in time-course fat body controls",
        subtitle = "Rows are z-scored using the combined diet-infection plus time-course VST matrix",
        x = NULL, y = NULL
      ) +
      theme_classic(base_size = 11) +
      theme(
        axis.text.x = element_text(angle = 20, hjust = 1, size = 10),
        axis.text.y = element_text(size = 7),
        legend.position = "bottom",
        plot.title = element_text(face = "bold")
      )

    print(ref_heatmap_plot)
    ggsave(file.path(out_dir, "timecourse_fatbody_reference_phase_linked_heatmap.png"),
           ref_heatmap_plot, width = 10, height = 12, dpi = 300)

    reference_direction <- ref_heatmap_genes |>
      dplyr::select(gene_id, Preferred_name, Description, phase_class,
                    phase_consensus, phase_mean_log2FC, priority_score) |>
      dplyr::left_join(timecourse_direction_all, by = "gene_id")

    write.csv(reference_direction,
              file.path(out_dir, "timecourse_fatbody_single_pair_phase_direction.csv"),
              row.names = FALSE)
  }

  timecourse_gene_sets <- list(
    "Any large single-pair shift" =
      timecourse_direction_all$gene_id[timecourse_direction_all$exploratory_shift],
    "Higher in gregarious/crowded reference" =
      timecourse_direction_all$gene_id[
        timecourse_direction_all$exploratory_shift &
          timecourse_direction_all$tissue_matched_direction ==
          "Higher in gregarious/crowded reference"
      ],
    "Higher in solitarious/isolated reference" =
      timecourse_direction_all$gene_id[
        timecourse_direction_all$exploratory_shift &
          timecourse_direction_all$tissue_matched_direction ==
          "Higher in solitarious/isolated reference"
      ],
    "Top absolute single-pair shifts" =
      timecourse_direction_all |>
        dplyr::slice_max(order_by = abs_timecourse_delta,
                         n = params$timecourse_top_gene_n,
                         with_ties = FALSE) |>
        dplyr::pull(gene_id)
  )

  overlap_one_set <- function(deg_genes, reference_genes) {
    deg_genes <- unique(deg_genes)
    reference_genes <- unique(reference_genes)
    overlap_genes <- intersect(deg_genes, reference_genes)
    tibble::tibble(
      n_fat_body_degs = length(deg_genes),
      n_reference_genes = length(reference_genes),
      n_overlap = length(overlap_genes),
      percent_fat_body_degs = 100 * length(overlap_genes) / max(length(deg_genes), 1),
      percent_reference_genes = 100 * length(overlap_genes) / max(length(reference_genes), 1),
      jaccard = length(overlap_genes) / max(length(union(deg_genes, reference_genes)), 1),
      overlap_gene_ids = paste(sort(overlap_genes), collapse = ";")
    )
  }

  timecourse_overlap_summary <- phase_layered_sig |>
    dplyr::group_by(scenario, comparison_type, comparison, direction) |>
    dplyr::summarise(fat_body_deg_genes = list(unique(gene_id)), .groups = "drop") |>
    tidyr::crossing(timecourse_reference_set = names(timecourse_gene_sets)) |>
    dplyr::rowwise() |>
    dplyr::mutate(
      stats = list(overlap_one_set(
        fat_body_deg_genes,
        timecourse_gene_sets[[timecourse_reference_set]]
      ))
    ) |>
    tidyr::unnest(stats) |>
    dplyr::ungroup() |>
    dplyr::select(-fat_body_deg_genes) |>
    dplyr::arrange(scenario, comparison_type, comparison,
                   dplyr::desc(n_overlap))

  write.csv(
    timecourse_overlap_summary,
    file.path(out_dir, "timecourse_fatbody_single_pair_overlap_summary.csv"),
    row.names = FALSE
  )

  timecourse_overlap_plot_data <- timecourse_overlap_summary |>
    dplyr::filter(comparison_type == "Infection within diet",
                  timecourse_reference_set != "Top absolute single-pair shifts") |>
    dplyr::mutate(
      comparison_plot = stringr::str_remove(comparison, ": infected vs control"),
      timecourse_reference_set = factor(
        timecourse_reference_set,
        levels = c("Any large single-pair shift",
                   "Higher in gregarious/crowded reference",
                   "Higher in solitarious/isolated reference")
      )
    )

  timecourse_overlap_plot <- ggplot(
    timecourse_overlap_plot_data,
    aes(x = comparison_plot, y = n_overlap, fill = timecourse_reference_set)
  ) +
    geom_col(position = position_dodge(width = 0.75), width = 0.68,
             color = "grey25", linewidth = 0.2) +
    facet_grid(direction ~ scenario, scales = "free_y") +
    scale_fill_manual(
      values = c(
        "Any large single-pair shift" = "#6B7280",
        "Higher in gregarious/crowded reference" = "#B23A48",
        "Higher in solitarious/isolated reference" = "#2166AC"
      )
    ) +
    labs(
      title = "Overlap with exploratory time-course fat body phase shifts",
      subtitle = paste0(
        "Large single-pair shifts use |gregarious - solitarious VST| >= ",
        params$timecourse_vst_delta_threshold
      ),
      x = NULL,
      y = "Number of overlapping genes",
      fill = "Time-course fat body reference"
    ) +
    theme_classic(base_size = 13) +
    theme(
      axis.text.x = element_text(angle = 35, hjust = 1),
      legend.position = "bottom",
      strip.text = element_text(face = "bold")
    )

  print(timecourse_overlap_plot)
  ggsave(file.path(out_dir, "timecourse_fatbody_single_pair_overlap.png"),
         timecourse_overlap_plot, width = 13, height = 8, dpi = 300)

  DT::datatable(
    timecourse_overlap_summary |>
      dplyr::select(-overlap_gene_ids),
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 12, scrollX = TRUE),
    caption = "Exploratory overlap between fat body DEGs and time-course single-pair phase shifts"
  )
} else {
  missing_files <- timecourse_ref_meta |>
    dplyr::filter(!file_present) |>
    dplyr::transmute(
      missing_featurecounts_file = file.path("data/external/timecourse_fatbody_reference",
                                             "readcounts", file_name)
    )

  knitr::kable(
    missing_files,
    caption = "Copy these featureCounts files to enable the exploratory PCA and heatmap"
  )
}
```

## Cross-tissue VST overview

This exploratory view combines the diet-infection fat body experiment, the two
time-course fat body controls, and the time-course gregaria head/thorax
phase-reference samples. The purpose is to ask where the present fat body
samples sit relative to independent tissue and phase references. Because tissue
identity and library context can dominate this kind of visualization, the PCA
and heatmap should be read as a structure/QC view, not as a formal cross-tissue
differential-expression test.

The heatmap includes all sample columns, but only the most variable shared genes
are drawn as rows so that the plot remains readable. The complete combined VST
matrix is written to the output folder for follow-up checks.

```{r cross-tissue-reference-files}
head_thorax_ref_meta <- parse_head_thorax_reference_metadata(
  timecourse_head_thorax_dir
) |>
  dplyr::mutate(file_present = file.exists(count_file))

fatbody_ref_meta_cross <- timecourse_ref_meta |>
  dplyr::mutate(
    tissue = "Fat body",
    source = "Time-course fat body phase control",
    display_group = dplyr::case_when(
      phase == "Gregarious" ~ "Fat body: gregarious/crowded control",
      phase == "Solitarious" ~ "Fat body: solitarious/isolated control",
      TRUE ~ "Fat body: reference control"
    ),
    sample_label = dplyr::case_when(
      phase == "Gregarious" ~ "FAT-G-ctrl",
      phase == "Solitarious" ~ "FAT-S-ctrl",
      TRUE ~ sample_id
    ),
    sample_block = "FAT controls",
    infection_status = "Control",
    diet = "Reference",
    outlier_status = "Reference sample"
  ) |>
  dplyr::select(
    sample_id, count_file, tissue, phase, rearing_condition, replicate,
    source, display_group, reference_role, sample_label, sample_block,
    infection_status, diet, outlier_status, file_present
  )

head_thorax_ref_meta_cross <- head_thorax_ref_meta |>
  dplyr::mutate(
    infection_status = "Control",
    diet = "Reference",
    outlier_status = "Reference sample"
  ) |>
  dplyr::select(
    sample_id, count_file, tissue, phase, rearing_condition, replicate,
    source, display_group, reference_role, sample_label, sample_block,
    infection_status, diet, outlier_status, file_present
  )

cross_reference_manifest <- dplyr::bind_rows(
  fatbody_ref_meta_cross,
  head_thorax_ref_meta_cross
)

knitr::kable(
  cross_reference_manifest |>
    dplyr::count(sample_block, tissue, phase, rearing_condition,
                 file_present, name = "n_samples"),
  caption = "Reference samples available for the cross-tissue VST overview"
)
```

```{r cross-tissue-vst-build}
cross_tissue_ready <- nrow(cross_reference_manifest) > 0 &&
  all(cross_reference_manifest$file_present)

if (cross_tissue_ready) {
  metadata_all_samples <- read_metadata(metadata_file)
  if (isTRUE(params$exclude_experiment_outliers)) {
    metadata_all_samples <- metadata_all_samples |>
      dplyr::filter(!label %in% current_outliers)
  }
  metadata_all_samples <- metadata_all_samples |>
    dplyr::arrange(Treatment, Diet, label)

  experiment_counts <- counts_all[, metadata_all_samples$label, drop = FALSE]

  experiment_meta_cross <- metadata_all_samples |>
    dplyr::mutate(
      sample_id = label,
      count_file = NA_character_,
      tissue = "Fat body",
      phase = "Diet-infection experiment",
      rearing_condition = "Semi-isolated experiment",
      replicate = NA_integer_,
      source = "Diet-infection fat body experiment",
      display_group = paste(Treatment, Diet, sep = "_"),
      reference_role = paste("Fat body", Treatment, "diet", Diet),
      sample_label = paste0(label, "_",
                            dplyr::if_else(Treatment == "Infected", "I", "C"),
                            Diet),
      sample_block = "Experiment fat body",
      infection_status = as.character(Treatment),
      diet = as.character(Diet),
      outlier_status = dplyr::if_else(
        label %in% current_outliers,
        "Current outlier flag",
        "Retained in current analysis"
      ),
      file_present = TRUE
    ) |>
    dplyr::select(
      sample_id, count_file, tissue, phase, rearing_condition, replicate,
      source, display_group, reference_role, sample_label, sample_block,
      infection_status, diet, outlier_status, file_present
    )

  reference_count_tables <- Map(
    read_featurecounts_gene_counts,
    cross_reference_manifest$count_file,
    cross_reference_manifest$sample_id
  )
  reference_counts_df <- Reduce(
    function(x, y) dplyr::full_join(x, y, by = "gene_id"),
    reference_count_tables
  )
  reference_counts <- as.matrix(reference_counts_df[, -1, drop = FALSE])
  rownames(reference_counts) <- reference_counts_df$gene_id
  reference_counts[is.na(reference_counts)] <- 0L
  storage.mode(reference_counts) <- "integer"

  shared_cross_genes <- Reduce(
    intersect,
    list(rownames(experiment_counts), rownames(reference_counts))
  )

  cross_counts <- cbind(
    experiment_counts[shared_cross_genes, , drop = FALSE],
    reference_counts[
      shared_cross_genes,
      cross_reference_manifest$sample_id,
      drop = FALSE
    ]
  )

  cross_metadata <- dplyr::bind_rows(
    experiment_meta_cross,
    cross_reference_manifest
  ) |>
    dplyr::mutate(
      sample_block = factor(
        sample_block,
        levels = c(
          "Experiment fat body",
          "FAT controls",
          "Head reference",
          "Thorax reference"
        )
      ),
      tissue = factor(tissue, levels = c("Fat body", "Head", "Thorax")),
      display_group = factor(display_group, levels = names(cross_tissue_colors))
    )

  cross_metadata <- cross_metadata[match(colnames(cross_counts),
                                         cross_metadata$sample_id), ]
  rownames(cross_metadata) <- cross_metadata$sample_id

  keep_cross <- rowSums(cross_counts) >= 10
  dds_cross <- DESeq2::DESeqDataSetFromMatrix(
    countData = cross_counts[keep_cross, , drop = FALSE],
    colData = cross_metadata,
    design = ~ 1
  )
  vst_cross <- DESeq2::vst(dds_cross, blind = TRUE)
  vst_cross_mat <- SummarizedExperiment::assay(vst_cross)

  write.csv(
    cross_metadata,
    file.path(out_dir, "cross_tissue_reference_sample_metadata.csv"),
    row.names = FALSE
  )
  utils::write.csv(
    tibble::rownames_to_column(as.data.frame(vst_cross_mat), "gene_id"),
    gzfile(file.path(out_dir, "cross_tissue_reference_vst_matrix_all_genes.csv.gz")),
    row.names = FALSE
  )

  cross_pca <- prcomp(t(vst_cross_mat), center = TRUE, scale. = FALSE)
  cross_pct <- round(100 * (cross_pca$sdev^2 / sum(cross_pca$sdev^2)), 1)
  cross_pca_scores <- as.data.frame(cross_pca$x[, 1:4, drop = FALSE]) |>
    tibble::rownames_to_column("sample_id") |>
    dplyr::left_join(
      cross_metadata |>
        tibble::rownames_to_column("sample_id_meta"),
      by = c("sample_id" = "sample_id_meta")
    )

  write.csv(
    cross_pca_scores,
    file.path(out_dir, "cross_tissue_reference_combined_pca_scores.csv"),
    row.names = FALSE
  )

  pca_fatbody_reference_labels <- cross_pca_scores |>
    dplyr::filter(sample_block == "FAT controls") |>
    dplyr::mutate(
      label_x = PC1,
      label_y = PC2 + dplyr::if_else(phase == "Gregarious", 6, -6)
    )

  cross_pca_plot <- ggplot(
    cross_pca_scores,
    aes(x = PC1, y = PC2, color = display_group, shape = sample_block)
  ) +
    geom_point(size = 3.1, alpha = 0.86) +
    geom_segment(
      data = pca_fatbody_reference_labels,
      aes(x = PC1, y = PC2, xend = label_x, yend = label_y),
      inherit.aes = FALSE,
      color = "#4B5563",
      linewidth = 0.25
    ) +
    geom_label(
      data = pca_fatbody_reference_labels,
      aes(x = label_x, y = label_y, label = reference_role,
          color = display_group),
      inherit.aes = FALSE,
      fill = "white",
      label.size = 0.15,
      size = 3.2,
      show.legend = FALSE
    ) +
    scale_color_manual(values = cross_tissue_colors, drop = FALSE) +
    scale_shape_manual(
      values = c(
        "Experiment fat body" = 16,
        "FAT controls" = 17,
        "Head reference" = 15,
        "Thorax reference" = 18
      ),
      drop = FALSE
    ) +
    scale_x_continuous(expand = ggplot2::expansion(mult = 0.12)) +
    scale_y_continuous(expand = ggplot2::expansion(mult = 0.12)) +
    labs(
      title = "PCA of experiment fat body with fat body, head, and thorax references",
      subtitle = paste0(
        "Exploratory VST PCA across shared expressed genes; ",
        sample_filter_label
      ),
      x = paste0("PC1 (", cross_pct[1], "%)"),
      y = paste0("PC2 (", cross_pct[2], "%)"),
      color = "Sample group",
      shape = "Sample block"
    ) +
    guides(
      color = ggplot2::guide_legend(ncol = 2, byrow = TRUE, order = 2),
      shape = ggplot2::guide_legend(nrow = 2, byrow = TRUE, order = 1)
    ) +
    coord_cartesian(clip = "off") +
    theme_classic(base_size = 13) +
    theme(
      legend.position = "bottom",
      legend.box = "vertical",
      plot.title = element_text(face = "bold"),
      plot.margin = ggplot2::margin(10, 70, 10, 10)
    )

  cross_gene_variance <- apply(vst_cross_mat, 1, stats::var)
  cross_top_gene_ids <- names(sort(cross_gene_variance, decreasing = TRUE))[
    seq_len(min(params$cross_tissue_top_variable_gene_n,
                length(cross_gene_variance)))
  ]

  cross_heatmap_genes <- tibble::tibble(
    gene_id = cross_top_gene_ids,
    vst_variance = cross_gene_variance[cross_top_gene_ids]
  ) |>
    dplyr::left_join(annotation, by = "gene_id") |>
    dplyr::left_join(phase_gene_all, by = "gene_id") |>
    dplyr::mutate(
      Preferred_name = dplyr::coalesce(Preferred_name, "unannotated"),
      Description = dplyr::coalesce(Description, "no description"),
      phase_class = dplyr::coalesce(phase_class, "No phase DEG"),
      label = paste0(
        gene_id, " | ", Preferred_name, " | ",
        stringr::str_trunc(Description, 42)
      )
    )

  write.csv(
    cross_heatmap_genes,
    file.path(out_dir, "cross_tissue_top_variable_vst_heatmap_genes.csv"),
    row.names = FALSE
  )

  cross_zmat <- t(scale(t(vst_cross_mat[cross_top_gene_ids, , drop = FALSE])))
  cross_zmat[is.na(cross_zmat)] <- 0

  cross_heatmap_long <- as.data.frame(cross_zmat) |>
    tibble::rownames_to_column("gene_id") |>
    tidyr::pivot_longer(-gene_id, names_to = "sample_id",
                        values_to = "z_vst") |>
    dplyr::left_join(
      cross_heatmap_genes |>
        dplyr::select(gene_id, label, vst_variance, phase_class),
      by = "gene_id"
    ) |>
    dplyr::left_join(
      cross_metadata |>
        dplyr::select(sample_id, sample_label, sample_block, display_group,
                      tissue, outlier_status),
      by = "sample_id"
    ) |>
    dplyr::mutate(
      sample_label = factor(sample_label, levels = cross_metadata$sample_label),
      label = factor(label, levels = rev(cross_heatmap_genes$label))
    )

  cross_heatmap_plot <- ggplot(
    cross_heatmap_long,
    aes(x = sample_label, y = label, fill = z_vst)
  ) +
    geom_tile(color = "white", linewidth = 0.03) +
    facet_grid(. ~ sample_block, scales = "free_x", space = "free_x") +
    scale_fill_gradient2(
      low = "#2166AC", mid = "white", high = "#B2182B",
      midpoint = 0, limits = c(-2.5, 2.5),
      oob = scales::squish, name = "Row-z VST"
    ) +
    labs(
      title = "Cross-tissue VST heatmap with experiment and phase-reference samples",
      subtitle = paste0(
        "Top ", length(cross_top_gene_ids),
        " most variable shared genes; ", sample_filter_label
      ),
      x = NULL,
      y = NULL
    ) +
    theme_classic(base_size = 10) +
    theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 5.5),
      axis.text.y = element_text(size = 4.8),
      legend.position = "bottom",
      strip.text.x = element_text(face = "bold", size = 11),
      panel.spacing.x = grid::unit(0.6, "lines"),
      plot.title = element_text(face = "bold")
    )
} else {
  missing_cross_files <- cross_reference_manifest |>
    dplyr::filter(!file_present) |>
    dplyr::select(sample_id, sample_block, count_file)

  knitr::kable(
    missing_cross_files,
    caption = "Missing reference files for the cross-tissue VST overview"
  )
}
```

```{r cross-tissue-pca, fig.width=13, fig.height=8}
if (exists("cross_pca_plot")) {
  print(cross_pca_plot)
  ggsave(
    file.path(out_dir, "cross_tissue_reference_combined_pca.png"),
    cross_pca_plot,
    width = 13,
    height = 8,
    dpi = 300
  )
}
```

```{r cross-tissue-vst-heatmap, fig.width=24, fig.height=20}
if (exists("cross_heatmap_plot")) {
  print(cross_heatmap_plot)
  ggsave(
    file.path(out_dir, "cross_tissue_top_variable_vst_heatmap.png"),
    cross_heatmap_plot,
    width = 24,
    height = 20,
    dpi = 300
  )
}
```

## Fat body DEGs in reference samples

The heatmap above uses the most variable shared genes, which is useful for
seeing sample structure. The plot below asks a narrower biological question:
among genes called differentially expressed in the fat body experiment, how do
those same genes look in the independent reference samples?

Rows are selected from the `r params$analysis_scenario` fat body DEG results.
Columns are only the reference samples: the two time-course fat body controls
plus the head and thorax phase-reference samples. Row z-scores are calculated
across reference samples only, so red and blue indicate relative expression
within the reference panel for each experimental DEG.

```{r fatbody-degs-in-reference-samples, fig.width=18, fig.height=22}
if (exists("vst_cross_mat") && exists("cross_metadata")) {
  fatbody_deg_reference_genes_all <- phase_layered_sig |>
    dplyr::filter(scenario == params$analysis_scenario) |>
    dplyr::group_by(gene_id) |>
    dplyr::summarise(
      n_deg_contrasts = dplyr::n_distinct(comparison),
      has_infection_deg = any(comparison_type == "Infection within diet"),
      comparison_types = paste(sort(unique(comparison_type)), collapse = "; "),
      comparisons = paste(sort(unique(comparison)), collapse = "; "),
      directions = paste(sort(unique(direction)), collapse = "; "),
      max_abs_log2FC = max(abs(log2FoldChange), na.rm = TRUE),
      min_padj = min(padj, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::mutate(
      deg_priority_score = n_deg_contrasts * 10 +
        dplyr::if_else(has_infection_deg, 5, 0) +
        max_abs_log2FC + safe_neglog10(min_padj)
    ) |>
    dplyr::left_join(annotation, by = "gene_id") |>
    dplyr::left_join(phase_gene_all, by = "gene_id") |>
    dplyr::mutate(
      Preferred_name = dplyr::coalesce(Preferred_name, "unannotated"),
      Description = dplyr::coalesce(Description, "no description"),
      phase_class = dplyr::coalesce(phase_class, "No phase DEG"),
      phase_consensus = dplyr::coalesce(phase_consensus, "No phase DEG")
    ) |>
    dplyr::arrange(
      dplyr::desc(has_infection_deg),
      dplyr::desc(n_deg_contrasts),
      min_padj,
      dplyr::desc(max_abs_log2FC)
    )

  write.csv(
    fatbody_deg_reference_genes_all,
    file.path(out_dir, "cross_tissue_fatbody_deg_reference_heatmap_all_deg_genes.csv"),
    row.names = FALSE
  )

  fatbody_deg_reference_genes <- fatbody_deg_reference_genes_all |>
    dplyr::filter(gene_id %in% rownames(vst_cross_mat)) |>
    dplyr::slice_head(n = params$cross_tissue_fatbody_deg_heatmap_n) |>
    dplyr::mutate(
      label = paste0(
        gene_id, " | ", Preferred_name, " | ",
        stringr::str_trunc(Description, 44)
      )
    )

  reference_only_metadata <- cross_metadata |>
    tibble::rownames_to_column("sample_id_row") |>
    dplyr::filter(sample_block != "Experiment fat body") |>
    dplyr::select(-sample_id_row)

  reference_sample_ids <- reference_only_metadata$sample_id
  reference_deg_ids <- intersect(
    fatbody_deg_reference_genes$gene_id,
    rownames(vst_cross_mat)
  )

  if (length(reference_deg_ids) > 1) {
    deg_reference_zmat <- t(scale(t(
      vst_cross_mat[reference_deg_ids, reference_sample_ids, drop = FALSE]
    )))
    deg_reference_zmat[is.na(deg_reference_zmat)] <- 0

    write.csv(
      fatbody_deg_reference_genes,
      file.path(out_dir, "cross_tissue_fatbody_deg_reference_heatmap_genes.csv"),
      row.names = FALSE
    )

    deg_reference_long <- as.data.frame(deg_reference_zmat) |>
      tibble::rownames_to_column("gene_id") |>
      tidyr::pivot_longer(-gene_id, names_to = "sample_id",
                          values_to = "z_vst") |>
      dplyr::left_join(
        fatbody_deg_reference_genes |>
          dplyr::select(
            gene_id, label, has_infection_deg, n_deg_contrasts,
            comparison_types, phase_class, phase_consensus
          ),
        by = "gene_id"
      ) |>
      dplyr::left_join(
        reference_only_metadata |>
          dplyr::select(sample_id, sample_label, sample_block, display_group,
                        tissue, phase, rearing_condition),
        by = "sample_id"
      ) |>
      dplyr::mutate(
        sample_label = factor(sample_label,
                              levels = reference_only_metadata$sample_label),
        label = factor(label, levels = rev(fatbody_deg_reference_genes$label))
      )

    fatbody_deg_reference_heatmap <- ggplot(
      deg_reference_long,
      aes(x = sample_label, y = label, fill = z_vst)
    ) +
      geom_tile(color = "white", linewidth = 0.04) +
      facet_grid(. ~ sample_block, scales = "free_x", space = "free_x") +
      scale_fill_gradient2(
        low = "#2166AC", mid = "white", high = "#B2182B",
        midpoint = 0, limits = c(-2.5, 2.5),
        oob = scales::squish, name = "Reference row-z VST"
      ) +
      labs(
        title = "Experimental fat body DEGs in reference samples",
        subtitle = paste0(
          "Top ", length(reference_deg_ids),
          " ", params$analysis_scenario,
          " fat body DEGs; columns show only FAT/head/thorax references"
        ),
        x = NULL,
        y = NULL
      ) +
      theme_classic(base_size = 10) +
      theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 7),
        axis.text.y = element_text(size = 5),
        legend.position = "bottom",
        strip.text.x = element_text(face = "bold", size = 12),
        panel.spacing.x = grid::unit(0.8, "lines"),
        plot.title = element_text(face = "bold")
      )

    print(fatbody_deg_reference_heatmap)
    ggsave(
      file.path(out_dir, "cross_tissue_fatbody_deg_reference_heatmap.png"),
      fatbody_deg_reference_heatmap,
      width = 18,
      height = 22,
      dpi = 300
    )
  } else {
    message("Not enough fat body DEGs were available for the reference heatmap.")
  }
}
```

## Curated infection-response heatmap with references

This version uses the same curated infection-response gene table as the
publication-style heatmap in the infection-within-diet page, then appends the
FAT, head, and thorax reference samples as extra sample rows. This keeps the
gene order and functional annotations comparable to the original curated
figure, while showing how those same genes behave in the reference samples.

The values are VST-normalized counts from the combined count matrix, scaled
within each gene across the displayed samples. This is therefore a visualization
of relative expression patterns, not a new differential-expression test.

```{r curated-infection-reference-heatmap, fig.width=26, fig.height=16}
if (exists("vst_cross_mat") && exists("cross_metadata")) {
  if (!file.exists(curated_infection_heatmap_gene_table)) {
    stop("Missing curated heatmap gene table: ",
         curated_infection_heatmap_gene_table)
  }

  curated_reference_gene_table <- read.csv(
    curated_infection_heatmap_gene_table,
    check.names = FALSE
  ) |>
    dplyr::distinct(gene_id, .keep_all = TRUE)

  curated_reference_genes <- curated_reference_gene_table$gene_id[
    curated_reference_gene_table$gene_id %in% rownames(vst_cross_mat)
  ]

  curated_reference_gene_table <- curated_reference_gene_table |>
    dplyr::filter(gene_id %in% curated_reference_genes) |>
    dplyr::mutate(
      Functional_theme = factor(
        Functional_theme,
        levels = c(
          "Immune-related",
          "Storage/lipid transport",
          "Nutrition/metabolism",
          "Detoxification/stress",
          "Other strong DEG"
        )
      ),
      DEG_class = factor(DEG_class, levels = c("DEG", "Not DEG"))
    )

  curated_reference_genes <- curated_reference_gene_table$gene_id
  curated_reference_sample_ids <- cross_metadata$sample_id
  curated_reference_sample_labels <- dplyr::if_else(
    cross_metadata$sample_block == "Experiment fat body",
    cross_metadata$sample_id,
    cross_metadata$sample_label
  )

  curated_reference_mat <- t(vst_cross_mat[
    curated_reference_genes,
    curated_reference_sample_ids,
    drop = FALSE
  ])
  rownames(curated_reference_mat) <- curated_reference_sample_labels

  curated_reference_sample_annotation <- cross_metadata |>
    dplyr::transmute(
      Sample_block = sample_block,
      Diet = dplyr::if_else(
        sample_block == "Experiment fat body",
        as.character(diet),
        "Reference"
      ),
      Treatment = dplyr::if_else(
        sample_block == "Experiment fat body",
        infection_status,
        "Reference"
      ),
      Phase = dplyr::if_else(
        sample_block == "Experiment fat body",
        "Experiment",
        phase
      )
    )
  rownames(curated_reference_sample_annotation) <- rownames(curated_reference_mat)

  curated_reference_gene_annotation <- curated_reference_gene_table |>
    dplyr::select(DEG_class, Functional_theme)
  rownames(curated_reference_gene_annotation) <- curated_reference_genes

  label_text <- dplyr::if_else(
    !is.na(curated_reference_gene_table$Preferred_name) &
      curated_reference_gene_table$Preferred_name != "-",
    curated_reference_gene_table$Preferred_name,
    curated_reference_gene_table$Description
  )
  label_text[is.na(label_text) | label_text == "-" | label_text == ""] <-
    "unannotated"
  label_text <- substr(label_text, 1, 45)
  curated_reference_col_labels <- paste(curated_reference_genes,
                                        label_text, sep = " | ")

  curated_reference_sample_gaps <- find_gaps(paste(
    curated_reference_sample_annotation$Sample_block,
    curated_reference_sample_annotation$Treatment,
    curated_reference_sample_annotation$Diet,
    curated_reference_sample_annotation$Phase
  ))
  curated_reference_gene_gaps <- find_gaps(paste(
    curated_reference_gene_annotation$DEG_class,
    curated_reference_gene_annotation$Functional_theme
  ))

  curated_reference_annotation_colors <- list(
    Sample_block = c(
      "Experiment fat body" = "#4B5563",
      "FAT controls" = "#111827",
      "Head reference" = "#009E9A",
      "Thorax reference" = "#C51B7D"
    ),
    Diet = c(diet_colors, "Reference" = "#D1D5DB"),
    Treatment = c(treatment_colors, "Reference" = "#D1D5DB"),
    Phase = c(
      "Experiment" = "#4B5563",
      "Gregarious" = "#B23A48",
      "Solitarious" = "#2166AC"
    ),
    DEG_class = c("DEG" = "#111827",
                  "Not DEG" = "#D1D5DB"),
    Functional_theme = c(
      "Immune-related" = "#237C8E",
      "Storage/lipid transport" = "#D7832F",
      "Nutrition/metabolism" = "#B8643C",
      "Detoxification/stress" = "#8C6D31",
      "Other strong DEG" = "#9CA3AF"
    )
  )

  curated_reference_expression_colors <- grDevices::colorRampPalette(
    c("#237C8E", "#F7F3E8", "#C85A54")
  )(101)
  curated_reference_expression_breaks <- seq(-3, 3, length.out = 102)

  curated_reference_heatmap <- pheatmap::pheatmap(
    curated_reference_mat,
    scale = "column",
    color = curated_reference_expression_colors,
    breaks = curated_reference_expression_breaks,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    gaps_row = curated_reference_sample_gaps,
    gaps_col = curated_reference_gene_gaps,
    show_rownames = TRUE,
    show_colnames = FALSE,
    labels_col = curated_reference_col_labels,
    fontsize_row = 5,
    annotation_col = curated_reference_gene_annotation,
    annotation_row = curated_reference_sample_annotation,
    annotation_colors = curated_reference_annotation_colors,
    main = "Curated fat body infection-response genes with reference samples",
    filename = file.path(out_dir,
                         "curated_infection_deg_vst_heatmap_with_references.png"),
    width = 26,
    height = 16,
    silent = TRUE
  )

  curated_reference_labeled_heatmap <- pheatmap::pheatmap(
    curated_reference_mat,
    scale = "column",
    color = curated_reference_expression_colors,
    breaks = curated_reference_expression_breaks,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    gaps_row = curated_reference_sample_gaps,
    gaps_col = curated_reference_gene_gaps,
    show_rownames = TRUE,
    show_colnames = TRUE,
    labels_col = curated_reference_col_labels,
    fontsize_row = 5,
    fontsize_col = 4.2,
    angle_col = 90,
    annotation_col = curated_reference_gene_annotation,
    annotation_row = curated_reference_sample_annotation,
    annotation_colors = curated_reference_annotation_colors,
    main = "Curated fat body infection-response genes with reference samples",
    filename = file.path(
      out_dir,
      "curated_infection_deg_vst_heatmap_with_references_labeled.png"
    ),
    width = 28,
    height = 16,
    silent = TRUE
  )

  write.csv(
    curated_reference_gene_table,
    file.path(out_dir,
              "curated_infection_deg_heatmap_with_references_genes.csv"),
    row.names = FALSE
  )
  write.csv(
    cbind(sample_id = curated_reference_sample_ids,
          sample_label = curated_reference_sample_labels,
          curated_reference_sample_annotation),
    file.path(out_dir,
              "curated_infection_deg_heatmap_with_references_samples.csv"),
    row.names = FALSE
  )

  curated_reference_display_dir <- file.path(
    project_dir,
    "docs", "figure", params$display_rmd_dir
  )
  dir.create(curated_reference_display_dir, recursive = TRUE,
             showWarnings = FALSE)
  curated_reference_labeled_display_png <- file.path(
    curated_reference_display_dir,
    "curated-infection-reference-heatmap-labeled-display.png"
  )
  file.copy(
    file.path(out_dir,
              "curated_infection_deg_vst_heatmap_with_references_labeled.png"),
    curated_reference_labeled_display_png,
    overwrite = TRUE
  )
  curated_reference_labeled_heatmap_src <- file.path(
    "figure", params$display_rmd_dir,
    "curated-infection-reference-heatmap-labeled-display.png"
  )

  grid::grid.newpage()
  grid::grid.draw(curated_reference_labeled_heatmap$gtable)
}
```

```{r curated-infection-reference-heatmap-display, echo=FALSE, results='asis'}
if (exists("curated_reference_labeled_heatmap_src")) {
  cat(paste0(
    '<p><img src="', curated_reference_labeled_heatmap_src,
    '" style="width:100%;"></p>'
  ))
}
```

## Full fat body DEG heatmap with references

The curated heatmap above is easier to read because it uses a selected set of
infection-response and fat-body-relevant genes. This companion figure uses the
full `r params$analysis_scenario` fat body DEG set instead. The sample rows are
the same as in the curated reference heatmap: the selected diet-infection fat
body samples, then the two FAT controls, then head and thorax reference samples.

Individual gene labels are hidden because thousands of DEG columns cannot be
read as text in a single figure. The exported gene table preserves the ordering,
DEG context, phase class, adjusted p-values, effect sizes, and annotations.

```{r full-fatbody-deg-reference-heatmap, fig.width=34, fig.height=16}
if (exists("vst_cross_mat") && exists("cross_metadata") &&
    exists("fatbody_deg_reference_genes_all")) {
  full_deg_reference_gene_table <- fatbody_deg_reference_genes_all |>
    dplyr::filter(gene_id %in% rownames(vst_cross_mat)) |>
    dplyr::mutate(
      DEG_context = dplyr::case_when(
        has_infection_deg &
          stringr::str_detect(comparison_types, "Diet contrasts") ~
          "Infection and diet DEGs",
        has_infection_deg ~ "Infection DEG",
        TRUE ~ "Diet contrast DEG"
      ),
      DEG_context = factor(
        DEG_context,
        levels = c("Infection and diet DEGs",
                   "Infection DEG",
                   "Diet contrast DEG")
      ),
      phase_class = factor(
        dplyr::case_when(
          is.na(phase_class) ~ "No phase DEG",
          phase_class %in% c("Shared head/thorax phase DEG",
                             "Head phase DEG only",
                             "Thorax phase DEG only",
                             "No phase DEG") ~ phase_class,
          TRUE ~ "No phase DEG"
        ),
        levels = c("Shared head/thorax phase DEG",
                   "Head phase DEG only",
                   "Thorax phase DEG only",
                   "No phase DEG")
      )
    ) |>
    dplyr::arrange(
      DEG_context,
      phase_class,
      min_padj,
      dplyr::desc(max_abs_log2FC)
    )

  full_deg_reference_genes <- full_deg_reference_gene_table$gene_id
  full_deg_reference_sample_ids <- cross_metadata$sample_id
  full_deg_reference_sample_labels <- dplyr::if_else(
    cross_metadata$sample_block == "Experiment fat body",
    cross_metadata$sample_id,
    cross_metadata$sample_label
  )

  full_deg_reference_mat <- t(vst_cross_mat[
    full_deg_reference_genes,
    full_deg_reference_sample_ids,
    drop = FALSE
  ])
  storage.mode(full_deg_reference_mat) <- "double"
  rownames(full_deg_reference_mat) <- full_deg_reference_sample_labels

  full_deg_reference_sample_annotation <- cross_metadata |>
    dplyr::transmute(
      Sample_block = sample_block,
      Diet = dplyr::if_else(
        sample_block == "Experiment fat body",
        as.character(diet),
        "Reference"
      ),
      Treatment = dplyr::if_else(
        sample_block == "Experiment fat body",
        infection_status,
        "Reference"
      ),
      Phase = dplyr::if_else(
        sample_block == "Experiment fat body",
        "Experiment",
        phase
      )
    ) |>
    dplyr::mutate(dplyr::across(dplyr::everything(), as.character))
  rownames(full_deg_reference_sample_annotation) <-
    rownames(full_deg_reference_mat)

  full_deg_reference_gene_annotation <- full_deg_reference_gene_table |>
    dplyr::transmute(
      DEG_context = as.character(DEG_context),
      phase_class = as.character(phase_class)
    )
  rownames(full_deg_reference_gene_annotation) <- full_deg_reference_genes

  full_deg_reference_sample_gaps <- find_gaps(paste(
    full_deg_reference_sample_annotation$Sample_block,
    full_deg_reference_sample_annotation$Treatment,
    full_deg_reference_sample_annotation$Diet,
    full_deg_reference_sample_annotation$Phase
  ))
  full_deg_reference_gene_gaps <- find_gaps(paste(
    full_deg_reference_gene_annotation$DEG_context,
    full_deg_reference_gene_annotation$phase_class
  ))

  full_deg_reference_zmat <- scale(full_deg_reference_mat)
  full_deg_reference_zmat[is.na(full_deg_reference_zmat)] <- 0
  full_deg_reference_zmat <- pmax(pmin(full_deg_reference_zmat, 3), -3)

  full_deg_gene_order <- tibble::tibble(
    gene_id = full_deg_reference_genes,
    gene_index = seq_along(full_deg_reference_genes)
  ) |>
    dplyr::left_join(
      full_deg_reference_gene_table |>
        dplyr::select(gene_id, DEG_context, phase_class),
      by = "gene_id"
    )

  full_deg_sample_order <- tibble::tibble(
    sample_label = rownames(full_deg_reference_zmat),
    sample_index = seq_len(nrow(full_deg_reference_zmat)),
    y_plot = nrow(full_deg_reference_zmat) - sample_index + 1
  ) |>
    dplyr::bind_cols(full_deg_reference_sample_annotation |>
                       tibble::rownames_to_column("sample_label_check") |>
                       dplyr::select(-sample_label_check))

  full_deg_reference_long <- as.data.frame(full_deg_reference_zmat) |>
    tibble::rownames_to_column("sample_label") |>
    tidyr::pivot_longer(-sample_label, names_to = "gene_id",
                        values_to = "z_vst") |>
    dplyr::left_join(full_deg_gene_order, by = "gene_id") |>
    dplyr::left_join(full_deg_sample_order, by = "sample_label")

  sample_gap_positions <- if (length(full_deg_reference_sample_gaps) > 0) {
    nrow(full_deg_reference_zmat) - full_deg_reference_sample_gaps + 0.5
  } else {
    numeric()
  }
  gene_gap_positions <- if (length(full_deg_reference_gene_gaps) > 0) {
    full_deg_reference_gene_gaps + 0.5
  } else {
    numeric()
  }

  full_deg_reference_heatmap <- ggplot(
    full_deg_reference_long,
    aes(x = gene_index, y = y_plot, fill = z_vst)
  ) +
    geom_raster() +
    geom_hline(yintercept = sample_gap_positions,
               color = "white", linewidth = 0.35) +
    geom_vline(xintercept = gene_gap_positions,
               color = "white", linewidth = 0.35) +
    scale_fill_gradient2(
      low = "#237C8E", mid = "#F7F3E8", high = "#C85A54",
      midpoint = 0, limits = c(-3, 3),
      oob = scales::squish, name = "Row-z VST"
    ) +
    scale_x_continuous(
      breaks = NULL,
      expand = c(0, 0)
    ) +
    scale_y_continuous(
      breaks = full_deg_sample_order$y_plot,
      labels = full_deg_sample_order$sample_label,
      expand = c(0, 0)
    ) +
    labs(
      title = "Full fat body DEG set with reference samples",
      subtitle = paste0(
        length(full_deg_reference_genes),
        " ", params$analysis_scenario,
        " fat body DEGs; genes ordered by DEG context and phase class"
      ),
      x = "Fat body DEG genes",
      y = NULL
    ) +
    theme_classic(base_size = 10) +
    theme(
      axis.text.y = element_text(size = 5),
      axis.ticks.y = element_blank(),
      axis.line = element_blank(),
      legend.position = "bottom",
      plot.title = element_text(face = "bold"),
      plot.margin = ggplot2::margin(10, 10, 10, 10)
    )

  ggsave(
    file.path(out_dir, "full_fatbody_deg_vst_heatmap_with_references.png"),
    full_deg_reference_heatmap,
    width = 34,
    height = 16,
    dpi = 300
  )

  write.csv(
    full_deg_reference_gene_table,
    file.path(out_dir,
              "full_fatbody_deg_heatmap_with_references_genes.csv"),
    row.names = FALSE
  )
  write.csv(
    cbind(sample_id = full_deg_reference_sample_ids,
          sample_label = full_deg_reference_sample_labels,
          full_deg_reference_sample_annotation),
    file.path(out_dir,
              "full_fatbody_deg_heatmap_with_references_samples.csv"),
    row.names = FALSE
  )

  full_deg_reference_display_dir <- file.path(
    project_dir,
    "docs", "figure", params$display_rmd_dir
  )
  dir.create(full_deg_reference_display_dir, recursive = TRUE,
             showWarnings = FALSE)
  full_deg_reference_display_png <- file.path(
    full_deg_reference_display_dir,
    "full-fatbody-deg-reference-heatmap-display.png"
  )
  file.copy(
    file.path(out_dir,
              "full_fatbody_deg_vst_heatmap_with_references.png"),
    full_deg_reference_display_png,
    overwrite = TRUE
  )
  full_deg_reference_heatmap_src <- file.path(
    "figure", params$display_rmd_dir,
    "full-fatbody-deg-reference-heatmap-display.png"
  )

  print(full_deg_reference_heatmap)
}
```

```{r full-fatbody-deg-reference-heatmap-display, echo=FALSE, results='asis'}
if (exists("full_deg_reference_heatmap_src")) {
  cat(paste0(
    '<p><img src="', full_deg_reference_heatmap_src,
    '" style="width:100%;"></p>'
  ))
}
```

## Gene-source partition

This section separates three evidence layers that could otherwise be mixed
together in interpretation:

- tissue-associated reference expression patterns, from the FAT/head/thorax
  VST reference panel;
- phase-associated genes, from the independent head/thorax phase DEG lists;
- diet/infection-associated genes, from the selected fat body DEG scenario
  (`r params$analysis_scenario`).

The tissue layer is treated as an exploratory expression-pattern layer rather
than a formal tissue DEG test, because the reference samples were not generated
as a balanced replicated tissue-DEG experiment. A gene is called
tissue-associated here when its mean VST differs by at least
`r params$tissue_vst_range_threshold` across reference tissues.

```{r source-partition, fig.width=12, fig.height=7}
if (exists("vst_cross_mat") && exists("cross_metadata") &&
    exists("phase_gene_all") && exists("phase_layered_sig")) {
  reference_tissue_metadata <- as.data.frame(cross_metadata) |>
    dplyr::filter(sample_block != "Experiment fat body") |>
    dplyr::mutate(tissue = as.character(tissue))

  reference_tissue_sample_ids <- intersect(
    reference_tissue_metadata$sample_id,
    colnames(vst_cross_mat)
  )

  reference_tissue_long <- as.data.frame(
    vst_cross_mat[, reference_tissue_sample_ids, drop = FALSE]
  ) |>
    tibble::rownames_to_column("gene_id") |>
    tidyr::pivot_longer(-gene_id, names_to = "sample_id",
                        values_to = "vst") |>
    dplyr::left_join(
      reference_tissue_metadata |>
        dplyr::select(sample_id, tissue),
      by = "sample_id"
    )

  reference_tissue_signal <- reference_tissue_long |>
    dplyr::group_by(gene_id, tissue) |>
    dplyr::summarise(mean_vst = mean(vst, na.rm = TRUE), .groups = "drop") |>
    dplyr::group_by(gene_id) |>
    dplyr::summarise(
      n_reference_tissues = dplyr::n_distinct(tissue),
      tissue_vst_range = diff(range(mean_vst, na.rm = TRUE)),
      highest_reference_tissue = tissue[which.max(mean_vst)][1],
      lowest_reference_tissue = tissue[which.min(mean_vst)][1],
      reference_tissue_means = paste0(
        tissue, "=", round(mean_vst, 2), collapse = "; "
      ),
      .groups = "drop"
    )

  tissue_pattern_gene_ids <- reference_tissue_signal |>
    dplyr::filter(
      n_reference_tissues >= 2,
      tissue_vst_range >= params$tissue_vst_range_threshold
    ) |>
    dplyr::pull(gene_id)

  phase_gene_evidence <- phase_gene_all |>
    dplyr::transmute(
      gene_id,
      phase_associated = n_phase_tissues > 0,
      phase_class,
      phase_consensus,
      phase_tissues,
      phase_mean_log2FC,
      phase_max_abs_log2FC
    )

  diet_infection_evidence <- phase_layered_sig |>
    dplyr::filter(scenario == params$analysis_scenario) |>
    dplyr::group_by(gene_id) |>
    dplyr::summarise(
      diet_infection_associated = TRUE,
      infection_associated = any(comparison_type == "Infection within diet"),
      diet_associated = any(comparison_type != "Infection within diet"),
      n_deg_contrasts = dplyr::n_distinct(comparison),
      deg_comparison_types = paste(sort(unique(comparison_type)),
                                   collapse = "; "),
      deg_comparisons = paste(sort(unique(comparison)), collapse = "; "),
      deg_directions = paste(sort(unique(direction)), collapse = "; "),
      max_abs_log2FC = max(abs(log2FoldChange), na.rm = TRUE),
      min_padj = min(padj, na.rm = TRUE),
      .groups = "drop"
    )

  partition_gene_ids <- Reduce(
    union,
    list(
      tissue_pattern_gene_ids,
      phase_gene_evidence$gene_id[phase_gene_evidence$phase_associated],
      diet_infection_evidence$gene_id
    )
  )

  source_partition_levels <- c(
    "Tissue reference pattern only",
    "Phase signature only",
    "Diet/infection DEG only",
    "Tissue + phase",
    "Tissue + diet/infection",
    "Phase + diet/infection",
    "Tissue + phase + diet/infection"
  )

  gene_source_partition <- tibble::tibble(gene_id = partition_gene_ids) |>
    dplyr::left_join(reference_tissue_signal, by = "gene_id") |>
    dplyr::left_join(phase_gene_evidence, by = "gene_id") |>
    dplyr::left_join(diet_infection_evidence, by = "gene_id") |>
    dplyr::left_join(annotation, by = "gene_id") |>
    dplyr::mutate(
      tissue_associated = gene_id %in% tissue_pattern_gene_ids,
      phase_associated = dplyr::coalesce(phase_associated, FALSE),
      diet_infection_associated =
        dplyr::coalesce(diet_infection_associated, FALSE),
      infection_associated = dplyr::coalesce(infection_associated, FALSE),
      diet_associated = dplyr::coalesce(diet_associated, FALSE),
      source_class = dplyr::case_when(
        tissue_associated & !phase_associated & !diet_infection_associated ~
          "Tissue reference pattern only",
        !tissue_associated & phase_associated & !diet_infection_associated ~
          "Phase signature only",
        !tissue_associated & !phase_associated & diet_infection_associated ~
          "Diet/infection DEG only",
        tissue_associated & phase_associated & !diet_infection_associated ~
          "Tissue + phase",
        tissue_associated & !phase_associated & diet_infection_associated ~
          "Tissue + diet/infection",
        !tissue_associated & phase_associated & diet_infection_associated ~
          "Phase + diet/infection",
        tissue_associated & phase_associated & diet_infection_associated ~
          "Tissue + phase + diet/infection",
        TRUE ~ "No selected evidence"
      ),
      source_class = factor(source_class, levels = source_partition_levels),
      Preferred_name = dplyr::coalesce(Preferred_name, "unannotated"),
      Description = dplyr::coalesce(Description, "no description"),
      priority_score = dplyr::coalesce(n_deg_contrasts, 0) * 10 +
        dplyr::coalesce(tissue_vst_range, 0) +
        dplyr::coalesce(phase_max_abs_log2FC, 0) +
        dplyr::coalesce(safe_neglog10(min_padj), 0)
    ) |>
    dplyr::arrange(source_class, dplyr::desc(priority_score), gene_id)

  source_partition_counts <- gene_source_partition |>
    dplyr::count(source_class, name = "n_genes") |>
    dplyr::filter(!is.na(source_class))

  source_partition_plot <- ggplot(
    source_partition_counts,
    aes(x = n_genes, y = source_class, fill = source_class)
  ) +
    geom_col(width = 0.72, color = "grey25", linewidth = 0.25) +
    geom_text(aes(label = scales::comma(n_genes)),
              hjust = -0.15, size = 3.8) +
    scale_fill_manual(
      values = c(
        "Tissue reference pattern only" = "#009E9A",
        "Phase signature only" = "#7C3AED",
        "Diet/infection DEG only" = "#C85A54",
        "Tissue + phase" = "#4C78A8",
        "Tissue + diet/infection" = "#D7832F",
        "Phase + diet/infection" = "#B23A48",
        "Tissue + phase + diet/infection" = "#111827"
      ),
      guide = "none"
    ) +
    scale_x_continuous(
      labels = scales::comma,
      expand = ggplot2::expansion(mult = c(0, 0.14))
    ) +
    labs(
      title = "Partition of genes by tissue, phase, and diet/infection evidence",
      subtitle = paste0(
        "DEG scenario: ", params$analysis_scenario,
        "; tissue pattern threshold: VST range >= ",
        params$tissue_vst_range_threshold
      ),
      x = "Number of genes",
      y = NULL
    ) +
    theme_classic(base_size = 13) +
    theme(
      axis.text.y = element_text(size = 11),
      plot.title = element_text(face = "bold")
    )

  print(source_partition_plot)
  ggsave(
    file.path(out_dir, "gene_source_partition_counts.png"),
    source_partition_plot,
    width = 12,
    height = 7,
    dpi = 300
  )

  write.csv(
    reference_tissue_signal,
    file.path(out_dir, "reference_tissue_vst_signal_all_genes.csv"),
    row.names = FALSE
  )
  write.csv(
    gene_source_partition,
    file.path(out_dir, "gene_source_partition_gene_table.csv"),
    row.names = FALSE
  )
  write.csv(
    source_partition_counts,
    file.path(out_dir, "gene_source_partition_counts.csv"),
    row.names = FALSE
  )

  strict_source_classes <- c(
    "Tissue reference pattern only",
    "Phase signature only",
    "Diet/infection DEG only"
  )

  strict_source_partition <- gene_source_partition |>
    dplyr::filter(source_class %in% strict_source_classes) |>
    dplyr::group_by(source_class) |>
    dplyr::slice_head(n = params$source_partition_top_gene_n) |>
    dplyr::ungroup()

  write.csv(
    strict_source_partition,
    file.path(out_dir, "gene_source_partition_strict_gene_table.csv"),
    row.names = FALSE
  )

  DT::datatable(
    strict_source_partition |>
      dplyr::select(
        source_class, gene_id, Preferred_name, Description,
        tissue_vst_range, highest_reference_tissue, phase_class,
        phase_consensus, n_deg_contrasts, deg_comparison_types,
        deg_comparisons, max_abs_log2FC, min_padj
      ),
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 20, scrollX = TRUE),
    caption = paste0(
      "Top strict source-partition genes per class for ",
      params$analysis_scenario
    )
  )
}
```

## Interpretation guide

The most useful output is not the raw overlap count alone, but the direction of
the overlap:

- `Numerator more gregarious-like` means the numerator condition moves in the
  same direction as crowded/gregarious animals in the independent phase dataset.
- `Numerator more solitarious-like` means the numerator condition moves opposite
  to the crowded/gregarious signature, which is more consistent with an
  isolated/solitarious-like expression direction.
- `Mixed head/thorax phase signal` means the gene is phase responsive, but the
  head and thorax reference signatures disagree in sign.

These genes should not be discarded automatically. Instead, they should be
highlighted in candidate-gene interpretation as possible density-state
sensitive genes that could confound or mediate diet and infection effects.

## Exported files

```{r exported-files}
exports <- list.files(out_dir, full.names = FALSE)
knitr::kable(
  tibble::tibble(file = exports),
  caption = paste("Files written to", out_dir)
)
```
