---
title: "Logical DEG set construction"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  latest_results_root: "output/rmd_runs"
  counts_file: "data/raw_read_counts/master_counts_no_rRNA.csv"
  metadata_file: "data/metadata/mehreen_metadata_fixed.txt"
  phase_reference_dir: "data/external/phase_reference"
  fatbody_reference_dir: "data/raw_read_counts/03-FAT-DESeq2"
  fatbody_reference_metadata: "data/external/timecourse_fatbody_reference/gregaria_FAT_phase_reference_metadata.csv"
  head_thorax_counts_dir: "data/raw_read_counts/03-gregaria-DESeq2"
  eggnog_file: "data/GO_Annotations/GCF_023897955.1_iqSchGreg1.2_Arthopoda_one2one.emapper.annotations"
  gff_file: "data/reference/GCF_023897955.1_iqSchGreg1.2_genomic.gff"
  analysis_scenario: "Current outliers removed"
  alpha: 0.05
  lfc_threshold: 1
  tissue_alpha: 0.05
  tissue_lfc_threshold: 1
  phase_alpha: 0.05
  phase_lfc_threshold: 1
  tissue_heatmap_gene_n: 100
  experiment_heatmap_gene_n: 120
  phase_heatmap_gene_n: 80
  category_heatmap_genes_per_class: 18
  output_tag: "deg-logical-set-construction"
---

## Purpose

This page separates the DEG questions before comparing them. The fat body
experiment remains the primary dataset for diet and infection, but the reference
data are now modeled independently instead of being forced to pass through the
fat body DEG list first.

The logic is:

1. Fit DESeq2 on the reference tissues only, using all copied fat body, head, and
   thorax reference samples regardless of phase. This gives tissue-associated
   DEGs.
2. Keep head and thorax phase-associated DEGs as the phase reference set. This
   asks whether a fat body diet/infection DEG is also phase-related in another
   sampled tissue.
3. Keep fat body experiment DEGs as separate sets for diet among controls and
   infection within each diet.
4. Only after those independent sets exist, compare them with Venn/UpSet plots
   and heatmaps.

Here, DEG categories are set-membership categories, not causal labels. A gene
can therefore belong to more than one biological hypothesis class. For example,
a fat body infection-response DEG can also be a tissue DEG or a phase DEG in
the reference data.

The experiment fat body and the reference fat body are kept as separate sample
blocks in PCA and heatmap annotations because they come from different
experimental sources, even though both are fat body tissue.

## DEG categories used here

```{r deg-vocabulary-table}
deg_vocabulary <- tibble::tribble(
  ~category, ~status_in_this_page, ~definition,
  "DEG Diet (controls)",
  "Computed",
  "Fat body genes significant in pairwise diet contrasts using control animals only.",
	  "DEG infection x diet",
	  "Final overlap set",
	  "Fat body genes significant between infected and control animals within one diet.",
	  "Pooled infection-only DEGs",
	  "Computed but not used in final four-set overlap",
	  "Fat body genes significant for infected versus control across all diets after accounting for diet with a DESeq2 model of ~ Diet + Treatment.",
	  "Fat body experiment DEGs",
	  "Visual QC set",
	  "Union of fat body experiment DEG contrasts used for broad fat body heatmaps; the final overlap uses only DEG Diet (controls) and DEG infection x diet.",
  "DEG tissues",
  "Independent reference set",
  "Genes significant in pairwise DESeq2 tissue contrasts among copied fat body, head, and thorax reference count files using the model ~ tissue. Phase is ignored for this first reference question.",
	  "Head phase DEGs",
	  "Included in phase reference set",
	  "Head genes significant between gregarious and solitarious reference samples.",
	  "Thorax phase DEGs",
	  "Included in phase reference set",
	  "Thorax genes significant between gregarious and solitarious reference samples.",
	  "DEG phase changes",
	  "Final overlap set",
	  "Union of head phase DEGs and thorax phase DEGs. The exported table keeps head and thorax membership separately.",
	  "Overlap categories",
	  "Compared after independent DEG calling",
	  "The final Venn/UpSet plots compare four independent sets: DEG tissues, DEG phase changes, DEG Diet (controls), and DEG infection x diet."
)

knitr::kable(
  deg_vocabulary,
  caption = "DEG vocabulary used to avoid mixing diet, infection, phase, and tissue questions"
)
```

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c(
  "DESeq2", "dplyr", "tidyr", "stringr", "ggplot2", "forcats",
  "DT", "SummarizedExperiment", "tibble", "scales", "UpSetR",
  "ggVennDiagram"
)
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

library(ggplot2)

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(
  project_dir, "output", "rmd_runs",
  paste(params$output_tag, run_stamp, sep = "_")
)
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

results_root <- file.path(project_dir, params$latest_results_root)
counts_file <- file.path(project_dir, params$counts_file)
metadata_file <- file.path(project_dir, params$metadata_file)
phase_dir <- file.path(project_dir, params$phase_reference_dir)
fatbody_reference_dir <- file.path(project_dir, params$fatbody_reference_dir)
fatbody_reference_metadata <- file.path(
  project_dir,
  params$fatbody_reference_metadata
)
head_thorax_counts_dir <- file.path(project_dir, params$head_thorax_counts_dir)
eggnog_file <- file.path(project_dir, params$eggnog_file)
gff_file <- file.path(project_dir, params$gff_file)

current_outliers <- c("1007", "1036", "1037", "1039")

diet_colors <- c("33" = "#E6862E", "50" = "#2F8F5B", "83" = "#7B4FA3")
treatment_colors <- c("Control" = "#7B8794", "Infected" = "#C85A54")
phase_colors <- c("Gregarious-up" = "#B23A48",
                  "Solitarious-up" = "#2166AC")
set_colors <- c(
  "DEG tissues" = "#4C78A8",
  "DEG phase changes" = "#C51B7D",
  "DEG Diet (controls)" = "#D7832F",
  "DEG infection x diet" = "#C85A54"
)
category_colors <- c(
  "DEG tissues" = "#4C78A8",
  "DEG phase changes" = "#C51B7D",
  "DEG Diet (controls)" = "#D7832F",
  "DEG infection x diet" = "#C85A54",
  "Other DEG" = "#6B7280"
)
```

## Helper functions

```{r helper-functions}
latest_dir <- function(pattern) {
  dirs <- list.dirs(results_root, recursive = FALSE, full.names = TRUE)
  dirs <- dirs[grepl(pattern, basename(dirs))]
  if (length(dirs) == 0) return(NA_character_)
  dirs[which.max(file.info(dirs)$mtime)]
}

safe_neglog10 <- function(x) {
  x <- ifelse(is.na(x), NA_real_, x)
  -log10(pmax(x, .Machine$double.xmin))
}

read_metadata <- function(path) {
  tab <- read.delim(path, check.names = FALSE)
  tab <- tab[, !is.na(names(tab)) & nzchar(names(tab)), drop = FALSE]
  tab$label <- as.character(tab$label)
  tab$Diet <- factor(tab$Diet, levels = c("33", "50", "83"))
  tab$Treatment <- factor(tab$Treatment, levels = c("Infected", "Control"))
  tab
}

read_phase_table <- function(path, tissue) {
  tab <- read.csv(path, check.names = FALSE)
  bad_names <- is.na(names(tab)) | names(tab) == ""
  names(tab)[bad_names] <- paste0("row_id", seq_len(sum(bad_names)))
  if (!"GeneID" %in% names(tab)) {
    tab$GeneID <- tab[[1]]
  }
  tab |>
    dplyr::transmute(
      gene_id = as.character(GeneID),
      tissue = tissue,
      baseMean = baseMean,
      log2FoldChange = log2FoldChange,
      pvalue = pvalue,
      padj = padj,
      significant = !is.na(padj) &
        padj < params$phase_alpha &
        abs(log2FoldChange) >= params$phase_lfc_threshold,
      direction = dplyr::case_when(
        !significant ~ "Not DEG",
        log2FoldChange > 0 ~ "Gregarious-up",
        log2FoldChange < 0 ~ "Solitarious-up",
        TRUE ~ "Not DEG"
      )
    )
}

read_featurecounts_gene_counts <- function(path, sample_id) {
  tab <- read.delim(path, comment.char = "#", check.names = FALSE)
  if ("Geneid" %in% names(tab)) {
    count_col <- names(tab)[ncol(tab)]
    out <- tibble::tibble(
      gene_id = as.character(tab[["Geneid"]]),
      count = as.integer(tab[[count_col]])
    )
  } else {
    tab <- read.delim(path, comment.char = "#", header = FALSE,
                      check.names = FALSE)
    out <- tibble::tibble(
      gene_id = as.character(tab[[1]]),
      count = as.integer(tab[[ncol(tab)]])
    )
  }
  out <- out |>
    dplyr::filter(!is.na(gene_id), nzchar(gene_id)) |>
    dplyr::group_by(gene_id) |>
    dplyr::summarise(count = sum(count, na.rm = TRUE), .groups = "drop")
  names(out)[2] <- sample_id
  out
}

parse_head_thorax_reference_metadata <- function(reference_dir) {
  files <- list.files(
    reference_dir,
    pattern = "^SGRE-(HEAD|THOX)-(CRD|ISO)-[0-9]+_MERGE_featurecounts[.]txt$",
    full.names = TRUE
  )
  parsed <- stringr::str_match(
    basename(files),
    "^SGRE-(HEAD|THOX)-(CRD|ISO)-([0-9]+)_MERGE_featurecounts[.]txt$"
  )
  tibble::tibble(
    sample_id = stringr::str_remove(basename(files), "_featurecounts[.]txt$"),
    count_file = files,
    tissue_code = parsed[, 2],
    rearing_code = parsed[, 3],
    replicate = as.integer(parsed[, 4])
  ) |>
    dplyr::filter(!is.na(tissue_code), !is.na(rearing_code)) |>
    dplyr::mutate(
      tissue = dplyr::case_when(
        tissue_code == "HEAD" ~ "Head",
        tissue_code == "THOX" ~ "Thorax",
        TRUE ~ tissue_code
      ),
      phase = dplyr::case_when(
        rearing_code == "CRD" ~ "Gregarious",
        rearing_code == "ISO" ~ "Solitarious",
        TRUE ~ rearing_code
      ),
      rearing_condition = dplyr::case_when(
        rearing_code == "CRD" ~ "Crowded",
        rearing_code == "ISO" ~ "Isolated",
        TRUE ~ rearing_code
      ),
      sample_label = paste0(tissue, "_", rearing_condition, "_", replicate),
      sample_block = paste(tissue, "phase reference")
    ) |>
    dplyr::arrange(tissue, phase, replicate)
}

parse_fatbody_reference_metadata <- function(metadata_path, count_dir) {
  tab <- read.csv(metadata_path, check.names = FALSE) |>
    dplyr::filter(tissue == "FAT")

  tab |>
    dplyr::mutate(
      count_file = file.path(count_dir, file_name),
      tissue = "Fat body",
      source = "Time-course fat body phase reference",
      display_group = dplyr::case_when(
        phase == "Gregarious" ~ "Reference fat body: crowded/gregarious",
        phase == "Solitarious" ~ "Reference fat body: isolated/solitarious",
        TRUE ~ "Reference fat body"
      ),
      sample_label = dplyr::case_when(
        phase == "Gregarious" ~ "FAT-G-control",
        phase == "Solitarious" ~ "FAT-S-control",
        TRUE ~ sample_id
      ),
      sample_block = "Reference fat body"
    ) |>
    dplyr::select(
      sample_id, count_file, tissue, phase, rearing_condition, replicate,
      source, display_group, reference_role, sample_label, sample_block
    ) |>
    dplyr::arrange(tissue, phase, replicate)
}

read_emapper <- function(path) {
  cols <- c(
    "query", "seed_ortholog", "evalue", "score", "eggNOG_OGs",
    "max_annot_lvl", "COG_category", "Description", "Preferred_name",
    "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
    "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC", "CAZy",
    "BiGG_Reaction", "PFAMs"
  )
  tab <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", fill = TRUE)
  names(tab) <- cols[seq_len(ncol(tab))]
  tab |>
    dplyr::mutate(protein_id = stringr::str_extract(query, "XP_[0-9]+[.][0-9]+"))
}

parse_cds_gene_bridge <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  gff |>
    dplyr::filter(type == "CDS") |>
    dplyr::transmute(
      protein_id = stringr::str_match(attributes, "protein_id=([^;]+)")[, 2],
      gene_id = stringr::str_match(attributes, "gene=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(protein_id), !is.na(gene_id)) |>
    dplyr::distinct()
}

make_set_membership <- function(gene_sets) {
  set_names <- names(gene_sets)
  membership <- tibble::tibble(gene_id = Reduce(union, gene_sets))
  for (nm in set_names) {
    membership[[nm]] <- membership$gene_id %in% gene_sets[[nm]]
  }
  membership |>
    dplyr::rowwise() |>
    dplyr::mutate(
      membership_label = paste(set_names[dplyr::c_across(dplyr::all_of(set_names))],
                               collapse = " + ")
    ) |>
    dplyr::ungroup()
}

write_upset_png <- function(gene_sets, file, width = 3000, height = 2200) {
  png(file, width = width, height = height, res = 300)
  print(
    UpSetR::upset(
      UpSetR::fromList(gene_sets),
      nsets = length(gene_sets),
      nintersects = NA,
      order.by = "freq",
      keep.order = TRUE,
      mainbar.y.label = "Intersection size",
      sets.x.label = "Set size",
      text.scale = c(1.4, 1.4, 1.2, 1.2, 1.4, 1.2)
    )
  )
  dev.off()
}

plot_deg_sets_venn <- function(gene_sets, title, subtitle) {
  ggVennDiagram::ggVennDiagram(
    gene_sets,
    category.names = names(gene_sets),
    label = "count",
    label_alpha = 0,
    set_size = 4.2,
    edge_size = 0.7
	  ) +
	    scale_fill_gradient(low = "#F7F3E8", high = "#4C78A8") +
	    coord_equal(clip = "off") +
    labs(
      title = title,
      subtitle = subtitle
    ) +
	    theme_void(base_size = 13) +
	    theme(
	      plot.title = element_text(face = "bold"),
	      plot.subtitle = element_text(hjust = 0.5),
	      plot.background = element_rect(fill = "white", color = NA),
	      panel.background = element_rect(fill = "white", color = NA),
	      legend.background = element_rect(fill = "white", color = NA),
	      plot.margin = ggplot2::margin(30, 130, 30, 130)
	    )
}

run_pairwise_tissue_deseq <- function(count_mat, sample_metadata) {
  sample_metadata <- sample_metadata |>
    dplyr::mutate(
      phase = factor(phase, levels = c("Gregarious", "Solitarious")),
      tissue = factor(tissue, levels = c("Fat body", "Head", "Thorax"))
    )
  rownames(sample_metadata) <- sample_metadata$sample_id

	  keep <- rowSums(count_mat) >= 10
	  dds <- DESeq2::DESeqDataSetFromMatrix(
	    countData = count_mat[keep, rownames(sample_metadata), drop = FALSE],
	    colData = sample_metadata,
	    design = ~ tissue
	  )
  dds <- DESeq2::DESeq(dds, quiet = TRUE)

  contrasts <- tibble::tribble(
    ~comparison, ~numerator, ~denominator,
    "Fat body vs Head", "Fat body", "Head",
    "Fat body vs Thorax", "Fat body", "Thorax",
    "Head vs Thorax", "Head", "Thorax"
  )

  dplyr::bind_rows(lapply(seq_len(nrow(contrasts)), function(i) {
    res <- DESeq2::results(
      dds,
      contrast = c("tissue", contrasts$numerator[i], contrasts$denominator[i])
    )
    as.data.frame(res) |>
      tibble::rownames_to_column("gene_id") |>
      dplyr::mutate(
        comparison = contrasts$comparison[i],
	        numerator = contrasts$numerator[i],
	        denominator = contrasts$denominator[i],
	        comparison_type = "Tissue contrast",
	        model_formula = "~ tissue",
        significant = !is.na(padj) &
          padj < params$tissue_alpha &
          abs(log2FoldChange) >= params$tissue_lfc_threshold,
        direction = dplyr::case_when(
          !significant ~ "Not DEG",
          log2FoldChange > 0 ~ paste("Higher in", contrasts$numerator[i]),
          log2FoldChange < 0 ~ paste("Higher in", contrasts$denominator[i]),
          TRUE ~ "Not DEG"
        )
      )
  }))
}

run_pooled_infection_deseq <- function(count_mat, sample_metadata,
                                       scenario_label) {
  sample_metadata <- sample_metadata |>
    dplyr::mutate(
      Diet = factor(Diet, levels = c("33", "50", "83")),
      Treatment = factor(Treatment, levels = c("Control", "Infected"))
    )
  rownames(sample_metadata) <- sample_metadata$label

  keep <- rowSums(count_mat) >= 10
  dds <- DESeq2::DESeqDataSetFromMatrix(
    countData = count_mat[keep, rownames(sample_metadata), drop = FALSE],
    colData = sample_metadata,
    design = ~ Diet + Treatment
  )
  dds <- DESeq2::DESeq(dds, quiet = TRUE)
  res <- DESeq2::results(dds, contrast = c("Treatment", "Infected", "Control"))

  as.data.frame(res) |>
    tibble::rownames_to_column("gene_id") |>
    dplyr::mutate(
      scenario = scenario_label,
      comparison_type = "Pooled infection across diets",
      comparison = "All diets: infected vs control",
      numerator = "Infected",
      denominator = "Control",
      diet = "All",
      significant = !is.na(padj) &
        padj < params$alpha &
        abs(log2FoldChange) >= params$lfc_threshold,
      direction = dplyr::case_when(
        !significant ~ "Not DEG",
        log2FoldChange > 0 ~ "Upregulated",
        log2FoldChange < 0 ~ "Downregulated",
        TRUE ~ "Not DEG"
      ),
      comparison_family = "Pooled infection across diets"
    )
}

plot_sample_matrix <- function(sample_metadata, title, output_file = NULL,
                               width = 10, height = 6) {
  sample_counts <- sample_metadata |>
    dplyr::count(sample_block, tissue, phase, rearing_condition,
                 name = "n_samples") |>
    dplyr::mutate(
      sample_group = paste(tissue, phase, sep = "\n"),
      sample_group = factor(sample_group, levels = unique(sample_group))
    )

  p <- ggplot(
    sample_counts,
    aes(x = sample_group, y = n_samples, fill = tissue)
  ) +
    geom_col(width = 0.68, color = "grey25", linewidth = 0.25) +
    geom_text(aes(label = n_samples), vjust = -0.35, size = 4) +
    scale_fill_manual(
      values = c("Fat body" = "#4C78A8", "Head" = "#009E9A",
                 "Thorax" = "#C51B7D"),
      guide = "none"
    ) +
    scale_y_continuous(expand = ggplot2::expansion(mult = c(0, 0.15))) +
    labs(title = title, x = NULL, y = "Number of samples") +
    theme_classic(base_size = 13) +
    theme(
      axis.text.x = element_text(size = 11),
      plot.title = element_text(face = "bold")
    )

  if (!is.null(output_file)) {
    ggsave(output_file, p, width = width, height = height, dpi = 300)
  }
  p
}

plot_fatbody_design_matrix <- function(metadata_exp, title, output_file = NULL,
                                       width = 8, height = 5.5) {
  design_counts <- metadata_exp |>
    dplyr::mutate(
      outlier_flag = label %in% current_outliers,
      Diet = factor(as.character(Diet), levels = c("33", "50", "83")),
      Treatment = factor(as.character(Treatment), levels = c("Infected", "Control"))
    ) |>
    dplyr::group_by(Treatment, Diet) |>
    dplyr::summarise(
      n_samples = dplyr::n(),
      n_current_outliers = sum(outlier_flag),
      .groups = "drop"
    ) |>
    dplyr::mutate(
      label = dplyr::if_else(
        n_current_outliers > 0,
        paste0(n_samples, "\n", n_current_outliers, " flagged"),
        as.character(n_samples)
      )
    )

  p <- ggplot(design_counts, aes(x = Diet, y = Treatment, fill = Diet)) +
    geom_tile(color = "white", linewidth = 1.1) +
    geom_text(aes(label = label), size = 4.3, lineheight = 0.9) +
    scale_fill_manual(values = diet_colors, guide = "none") +
    labs(title = title, x = "Diet", y = NULL) +
    coord_equal() +
    theme_classic(base_size = 13) +
    theme(
      axis.text = element_text(size = 12),
      plot.title = element_text(face = "bold")
    )

  if (!is.null(output_file)) {
    ggsave(output_file, p, width = width, height = height, dpi = 300)
  }
  p
}

plot_experiment_pca <- function(vst_mat, sample_metadata, title, subtitle,
                                output_file = NULL, width = 10, height = 7) {
  pca <- prcomp(t(vst_mat), center = TRUE, scale. = FALSE)
  pct <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)
  pca_scores <- as.data.frame(pca$x[, 1:4, drop = FALSE]) |>
    tibble::rownames_to_column("sample_id") |>
    dplyr::left_join(sample_metadata, by = "sample_id") |>
    dplyr::mutate(
      condition = paste(Treatment, "diet", Diet),
      outlier_flag = stringr::str_remove(sample_label, "[*]$") %in%
        current_outliers
    )

  ellipse_scores <- pca_scores |>
    dplyr::group_by(condition) |>
    dplyr::filter(dplyr::n() >= 3) |>
    dplyr::ungroup()

  p <- ggplot(
    pca_scores,
    aes(x = PC1, y = PC2, color = Diet, shape = Treatment)
  ) +
    stat_ellipse(
      data = ellipse_scores,
      aes(group = condition, color = Diet),
      linewidth = 0.45,
      linetype = "dashed",
      show.legend = FALSE
    ) +
    geom_point(aes(size = outlier_flag), alpha = 0.9) +
    geom_text(aes(label = sample_label), vjust = -0.9, size = 3,
              show.legend = FALSE) +
    scale_color_manual(values = diet_colors, drop = FALSE) +
    scale_shape_manual(values = c("Infected" = 16, "Control" = 17),
                       drop = FALSE) +
    scale_size_manual(values = c(`FALSE` = 2.8, `TRUE` = 4.2),
                      guide = "none") +
    labs(
      title = title,
      subtitle = subtitle,
      x = paste0("PC1 (", pct[1], "%)"),
      y = paste0("PC2 (", pct[2], "%)"),
      color = "Diet",
      shape = "Treatment"
    ) +
    coord_cartesian(clip = "off") +
    theme_classic(base_size = 13) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold"),
      plot.margin = ggplot2::margin(10, 40, 10, 10)
    )

  if (!is.null(output_file)) {
    ggsave(output_file, p, width = width, height = height, dpi = 300)
  }
  p
}

plot_combined_sample_pca <- function(vst_mat, sample_metadata, title, subtitle,
                                     output_file = NULL, width = 12, height = 8) {
  pca <- prcomp(t(vst_mat), center = TRUE, scale. = FALSE)
  pct <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)
  pca_scores <- as.data.frame(pca$x[, 1:4, drop = FALSE]) |>
    tibble::rownames_to_column("sample_id") |>
    dplyr::left_join(sample_metadata, by = "sample_id") |>
    dplyr::mutate(
      sample_block = factor(
        sample_block,
        levels = c(
          "Experiment fat body",
          "Reference fat body",
          "Head reference",
          "Thorax reference"
        )
      ),
      outlier_flag = stringr::str_remove(sample_label, "[*]$") %in%
        current_outliers
    )

  ellipse_scores <- pca_scores |>
    dplyr::group_by(sample_block) |>
    dplyr::filter(dplyr::n() >= 3) |>
    dplyr::ungroup()
  centroid_scores <- pca_scores |>
    dplyr::group_by(sample_block) |>
    dplyr::summarise(
      PC1 = mean(PC1, na.rm = TRUE),
      PC2 = mean(PC2, na.rm = TRUE),
      .groups = "drop"
    )
  point_labels <- pca_scores |>
    dplyr::filter(outlier_flag)

  p <- ggplot(
    pca_scores,
    aes(x = PC1, y = PC2, color = sample_block, shape = sample_block)
  ) +
    stat_ellipse(
      data = ellipse_scores,
      aes(group = sample_block, color = sample_block),
      linewidth = 0.5,
      linetype = "dashed",
      show.legend = FALSE
    ) +
    geom_point(aes(size = outlier_flag), alpha = 0.88) +
    geom_text(
      data = point_labels,
      aes(label = sample_label),
      vjust = -0.9,
      size = 3.3,
      show.legend = FALSE
    ) +
    geom_label(
      data = centroid_scores,
      aes(label = sample_block, fill = sample_block),
      color = "#111827",
      label.size = 0.15,
      alpha = 0.88,
      size = 3.6,
      show.legend = FALSE
    ) +
    scale_color_manual(
      values = c(
        "Experiment fat body" = "#C85A54",
        "Reference fat body" = "#4C78A8",
        "Head reference" = "#009E9A",
        "Thorax reference" = "#C51B7D"
      ),
      drop = FALSE
    ) +
    scale_shape_manual(
      values = c(
        "Experiment fat body" = 16,
        "Reference fat body" = 17,
        "Head reference" = 15,
        "Thorax reference" = 18
      ),
      drop = FALSE
    ) +
    scale_fill_manual(
      values = c(
        "Experiment fat body" = "#F4C7B8",
        "Reference fat body" = "#C7D7F2",
        "Head reference" = "#BEE8E6",
        "Thorax reference" = "#F1C3DA"
      ),
      drop = FALSE
    ) +
    scale_size_manual(values = c(`FALSE` = 3.0, `TRUE` = 4.4),
                      guide = "none") +
    labs(
      title = title,
      subtitle = subtitle,
      x = paste0("PC1 (", pct[1], "%)"),
      y = paste0("PC2 (", pct[2], "%)"),
      color = "Sample block",
      shape = "Sample block"
    ) +
    coord_cartesian(clip = "off") +
    theme_classic(base_size = 13) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold"),
      plot.margin = ggplot2::margin(10, 50, 10, 10)
    )

  if (!is.null(output_file)) {
    ggsave(output_file, p, width = width, height = height, dpi = 300)
  }
  p
}

plot_pca_from_vst <- function(vst_mat, sample_metadata, title, subtitle,
                              output_file = NULL, width = 10, height = 7) {
  pca <- prcomp(t(vst_mat), center = TRUE, scale. = FALSE)
  pct <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)
  pca_scores <- as.data.frame(pca$x[, 1:4, drop = FALSE]) |>
    tibble::rownames_to_column("sample_id") |>
    dplyr::left_join(sample_metadata, by = "sample_id")
  ellipse_scores <- pca_scores |>
    dplyr::group_by(tissue) |>
    dplyr::filter(dplyr::n() >= 3) |>
    dplyr::ungroup()

  p <- ggplot(
    pca_scores,
    aes(x = PC1, y = PC2, color = tissue, shape = phase)
  ) +
    geom_point(size = 3.6, alpha = 0.9) +
    geom_text(
      aes(label = sample_label),
      vjust = -0.9,
      size = 3,
      show.legend = FALSE
    ) +
    stat_ellipse(data = ellipse_scores,
                 aes(group = tissue, color = tissue),
                 linewidth = 0.45, linetype = "dashed", show.legend = FALSE) +
    scale_color_manual(
      values = c("Fat body" = "#4C78A8", "Head" = "#009E9A",
                 "Thorax" = "#C51B7D"),
      drop = FALSE
    ) +
    labs(
      title = title,
      subtitle = subtitle,
      x = paste0("PC1 (", pct[1], "%)"),
      y = paste0("PC2 (", pct[2], "%)"),
      color = "Tissue",
      shape = "Phase"
    ) +
    coord_cartesian(clip = "off") +
    theme_classic(base_size = 13) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold"),
      plot.margin = ggplot2::margin(10, 40, 10, 10)
    )

  if (!is.null(output_file)) {
    ggsave(output_file, p, width = width, height = height, dpi = 300)
  }
  p
}

make_vst_heatmap_long <- function(vst_mat, gene_table, sample_metadata,
                                  gene_order, sample_order) {
  zmat <- t(scale(t(vst_mat[gene_order, sample_order, drop = FALSE])))
  zmat[is.na(zmat)] <- 0
  as.data.frame(zmat) |>
    tibble::rownames_to_column("gene_id") |>
    tidyr::pivot_longer(-gene_id, names_to = "sample_id",
                        values_to = "z_vst") |>
    dplyr::left_join(gene_table, by = "gene_id") |>
    dplyr::left_join(sample_metadata, by = "sample_id") |>
    dplyr::mutate(
      sample_label = factor(sample_label,
                            levels = sample_metadata$sample_label[
                              match(sample_order, sample_metadata$sample_id)
                            ]),
      gene_label = factor(gene_label, levels = rev(gene_table$gene_label))
    )
}

plot_annotated_heatmap <- function(vst_mat, gene_table, sample_metadata,
                                   title, subtitle, output_file = NULL,
                                   width = 18, height = 14) {
  gene_order <- gene_table$gene_id[gene_table$gene_id %in% rownames(vst_mat)]
  sample_order <- sample_metadata$sample_id[
    sample_metadata$sample_id %in% colnames(vst_mat)
  ]
  gene_table <- gene_table |>
    dplyr::filter(gene_id %in% gene_order)
  if (length(gene_order) < 2 || length(sample_order) < 2) {
    return(NULL)
  }

  heatmap_long <- make_vst_heatmap_long(
    vst_mat,
    gene_table,
    sample_metadata,
    gene_order,
    sample_order
  ) |>
    dplyr::mutate(
      z_vst = pmax(pmin(z_vst, 3), -3),
      column_group = factor(column_group, levels = unique(sample_metadata$column_group))
    )

  p <- ggplot(
    heatmap_long,
    aes(x = sample_label, y = gene_label, fill = z_vst)
  ) +
    geom_tile(color = "white", linewidth = 0.035) +
    facet_grid(. ~ column_group, scales = "free_x", space = "free_x") +
    scale_fill_gradient2(
      low = "#237C8E", mid = "#F7F3E8", high = "#C85A54",
      midpoint = 0, limits = c(-3, 3),
      oob = scales::squish, name = "Row-z VST"
    ) +
    labs(
      title = title,
      subtitle = subtitle,
      x = NULL,
      y = NULL
    ) +
    theme_classic(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5,
                                 hjust = 1, size = 7),
      axis.text.y = element_text(size = 6.5),
      legend.position = "bottom",
      strip.text.x = element_text(face = "bold", size = 11),
      panel.spacing.x = grid::unit(0.3, "lines"),
      plot.title = element_text(face = "bold", size = 15),
      plot.subtitle = element_text(size = 12)
    )

  if (!is.null(output_file)) {
    ggsave(output_file, p, width = width, height = height, dpi = 300)
  }
  p
}

add_overlap_evidence_annotations <- function(gene_table) {
  gene_table |>
    dplyr::mutate(
      tissue_reference_deg = dplyr::coalesce(tissue_reference_deg, FALSE),
      head_phase_deg = dplyr::coalesce(head_phase_deg, FALSE),
      thorax_phase_deg = dplyr::coalesce(thorax_phase_deg, FALSE),
      phase_reference_deg = dplyr::coalesce(
        phase_reference_deg,
        head_phase_deg | thorax_phase_deg
      ),
	      diet_control_deg = dplyr::coalesce(diet_control_deg, FALSE),
	      diet_infected_deg = dplyr::coalesce(diet_infected_deg, FALSE),
	      fatbody_experiment_deg = dplyr::coalesce(fatbody_experiment_deg, FALSE),
	      any_phase_change = phase_reference_deg,
	      fatbody_diet_deg = diet_control_deg |
	        stringr::str_detect(dplyr::coalesce(experiment_families, ""),
	                            "Diet contrasts: controls"),
	      fatbody_infection_deg = dplyr::coalesce(fatbody_infection_deg, FALSE) |
	        stringr::str_detect(dplyr::coalesce(experiment_families, ""),
	                            "Infection within diet"),
	      has_tissue_overlay = tissue_reference_deg,
	      has_phase_overlay = any_phase_change,
	      primary_annotation_group = dplyr::case_when(
	        tissue_reference_deg ~ "DEG tissues",
	        phase_reference_deg ~ "DEG phase changes",
	        diet_control_deg ~ "DEG Diet (controls)",
	        fatbody_infection_deg ~ "DEG infection x diet",
	        TRUE ~ "Other DEG"
	      ),
      primary_annotation_group = factor(
	        primary_annotation_group,
	        levels = c(
	          "DEG tissues",
	          "DEG phase changes",
	          "DEG Diet (controls)",
	          "DEG infection x diet",
	          "Other DEG"
	        )
	      )
    )
}

plot_final_overlap_annotated_heatmap <- function(vst_mat, gene_table,
                                                 sample_metadata, title,
                                                 subtitle,
                                                 output_file = NULL,
                                                 width = 34,
                                                 height = 30) {
  display_group <- NULL
  display_group_levels <- NULL
  if ("primary_annotation_group" %in% names(gene_table)) {
    display_group <- as.character(gene_table$primary_annotation_group)
    display_group_levels <- levels(gene_table$primary_annotation_group)
    if (is.null(display_group_levels)) {
      display_group_levels <- unique(display_group)
    }
  }
  gene_table <- add_overlap_evidence_annotations(gene_table)
  if (!is.null(display_group)) {
    gene_table <- gene_table |>
      dplyr::mutate(
        primary_annotation_group = factor(display_group,
                                          levels = display_group_levels),
        logical_category = display_group
      )
  }
  gene_order <- gene_table$gene_id[gene_table$gene_id %in% rownames(vst_mat)]
  sample_order <- sample_metadata$sample_id[
    sample_metadata$sample_id %in% colnames(vst_mat)
  ]
  gene_table <- gene_table |>
    dplyr::filter(gene_id %in% gene_order)
  if (length(gene_order) < 2 || length(sample_order) < 2) {
    return(NULL)
  }

  column_levels <- c(
    "Row evidence",
    unique(as.character(sample_metadata$column_group))
  )

  heatmap_long <- make_vst_heatmap_long(
    vst_mat,
    gene_table,
    sample_metadata,
    gene_order,
    sample_order
  ) |>
    dplyr::mutate(
      z_vst = pmax(pmin(z_vst, 3), -3),
      column_group = factor(column_group, levels = column_levels),
      primary_annotation_group = factor(
        primary_annotation_group,
        levels = levels(gene_table$primary_annotation_group)
      )
    )

  evidence_levels <- c(
    "DEG tissues", "DEG phase changes",
    "DEG Diet (controls)", "DEG infection x diet"
  )
  evidence_colors <- c(
    "DEG tissues" = "#4C78A8",
    "DEG phase changes" = "#C51B7D",
    "DEG Diet (controls)" = "#D7832F",
    "DEG infection x diet" = "#C85A54"
  )

  annotation_background <- tidyr::expand_grid(
    gene_id = gene_order,
    evidence = factor(evidence_levels, levels = evidence_levels)
  ) |>
    dplyr::left_join(
	      gene_table |>
	        dplyr::select(
	          gene_id, gene_label, primary_annotation_group,
	          tissue_reference_deg, phase_reference_deg,
	          diet_control_deg, fatbody_infection_deg
	        ),
      by = "gene_id"
    ) |>
    dplyr::mutate(
	      gene_label = factor(gene_label, levels = rev(gene_table$gene_label)),
	      column_group = factor("Row evidence", levels = column_levels),
	      present = dplyr::case_when(
	        evidence == "DEG tissues" ~ tissue_reference_deg,
	        evidence == "DEG phase changes" ~ phase_reference_deg,
	        evidence == "DEG Diet (controls)" ~ diet_control_deg,
	        evidence == "DEG infection x diet" ~ fatbody_infection_deg,
	        TRUE ~ FALSE
	      )
	    )

  annotation_present <- annotation_background |>
    dplyr::filter(present)

  p <- ggplot() +
    geom_point(
      data = annotation_background,
      aes(x = evidence, y = gene_label),
      shape = 15,
      size = 3.2,
      color = "#E5E7EB"
    ) +
    geom_point(
      data = annotation_present,
      aes(x = evidence, y = gene_label, color = evidence),
      shape = 15,
      size = 3.4
    ) +
    geom_tile(
      data = heatmap_long,
      aes(x = sample_label, y = gene_label, fill = z_vst),
      color = "white",
      linewidth = 0.035
    ) +
    facet_grid(
      primary_annotation_group ~ column_group,
      scales = "free",
      space = "free"
    ) +
    scale_color_manual(values = evidence_colors, name = "Row evidence") +
    scale_fill_gradient2(
      low = "#237C8E", mid = "#F7F3E8", high = "#C85A54",
      midpoint = 0, limits = c(-3, 3),
      oob = scales::squish, name = "Row-z VST"
    ) +
    labs(title = title, subtitle = subtitle, x = NULL, y = NULL) +
    theme_classic(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5,
                                 hjust = 1, size = 7),
      axis.text.y = element_text(size = 6.5),
      legend.position = "bottom",
      legend.box = "vertical",
      strip.text.x = element_text(face = "bold", size = 11),
      strip.text.y = element_text(face = "bold", size = 9.5, angle = 0),
      panel.spacing.x = grid::unit(0.35, "lines"),
      panel.spacing.y = grid::unit(0.35, "lines"),
      plot.title = element_text(face = "bold", size = 15),
      plot.subtitle = element_text(size = 12)
    )

  if (!is.null(output_file)) {
    ggsave(output_file, p, width = width, height = height, dpi = 300)
  }
  p
}
```

## Input files

```{r input-files}
scenario_dir <- latest_dir("^deg-sample-filter-scenarios_")
if (is.na(scenario_dir)) {
  stop("Render 12-deg-sample-filter-scenarios.Rmd before this page.")
}

fat_results_file <- file.path(scenario_dir, "all_deseq2_results.csv")
head_phase_file <- file.path(
  phase_dir,
  "DESeq2_results_sva_Head_gregaria_togregaria.csv"
)
thorax_phase_file <- file.path(
  phase_dir,
  "DESeq2_results_sva_Thorax_gregaria_togregaria.csv"
)

input_files <- tibble::tibble(
  input = c("Fat body scenario DESeq2 results",
            "Head phase DESeq2 results",
            "Thorax phase DESeq2 results",
            "Fat body reference metadata",
            "Fat body reference count folder",
            "Fat body no-rRNA count matrix",
            "Fat body metadata",
            "Head/thorax featureCounts folder"),
  path = c(fat_results_file, head_phase_file, thorax_phase_file,
           fatbody_reference_metadata, fatbody_reference_dir,
           counts_file, metadata_file, head_thorax_counts_dir),
  present = c(file.exists(fat_results_file), file.exists(head_phase_file),
              file.exists(thorax_phase_file),
              file.exists(fatbody_reference_metadata),
              dir.exists(fatbody_reference_dir), file.exists(counts_file),
              file.exists(metadata_file), dir.exists(head_thorax_counts_dir))
)

knitr::kable(input_files, caption = "Files used to build the logical DEG sets")
```

```{r load-annotations}
if (file.exists(eggnog_file) && file.exists(gff_file)) {
  eggnog <- read_emapper(eggnog_file)
  protein_gene_bridge <- parse_cds_gene_bridge(gff_file)
  annotation <- protein_gene_bridge |>
    dplyr::left_join(eggnog, by = "protein_id") |>
    dplyr::group_by(gene_id) |>
    dplyr::summarise(
      Preferred_name = dplyr::first(Preferred_name[!is.na(Preferred_name) &
                                                    Preferred_name != "-"]),
      Description = dplyr::first(Description[!is.na(Description) &
                                             Description != "-"]),
      .groups = "drop"
    )
} else {
  annotation <- tibble::tibble(
    gene_id = character(),
    Preferred_name = character(),
    Description = character()
  )
}
```

## Deduction workflow

The figure below is the logical overview for the main all-samples analysis. Each
DEG question is computed as its own set first. The overlap is then interpreted
afterward, rather than forcing reference genes to pass through a fat body DEG
filter.

For this page, `DEG Diet (controls)` means diet contrasts among control fat
body samples, and `DEG infection x diet` means infected-versus-control contrasts
computed within each diet. `DEG tissues` means a cautious exploratory tissue
contrast computed from the copied fat body, head, and thorax reference count
files using the model `~ tissue`, with phase ignored for this tissue-only
question. `DEG phase changes` is the union of head and thorax phase DEGs.

```{r top-deg-set-definitions, include=FALSE}
fat_results <- read.csv(fat_results_file, check.names = FALSE) |>
  dplyr::mutate(
    diet = as.character(diet),
    numerator = as.character(numerator),
    denominator = as.character(denominator),
    significant = !is.na(padj) &
      padj < params$alpha &
      abs(log2FoldChange) >= params$lfc_threshold,
    direction = dplyr::case_when(
      !significant ~ "Not DEG",
      log2FoldChange > 0 ~ "Upregulated",
      log2FoldChange < 0 ~ "Downregulated",
      TRUE ~ "Not DEG"
    ),
    comparison_family = dplyr::case_when(
      comparison_type == "Infection within diet" ~ "Infection within diet",
      comparison_type == "Diet contrasts among control" ~
        "Diet contrasts: controls",
      comparison_type == "Diet contrasts among infected" ~
        "Diet contrasts: infected",
      TRUE ~ comparison_type
    )
  )

metadata_all_base <- read_metadata(metadata_file)

counts_raw <- read.csv(counts_file, check.names = FALSE)
counts_all <- as.matrix(counts_raw[, -1, drop = FALSE])
rownames(counts_all) <- counts_raw[[1]]
storage.mode(counts_all) <- "integer"
colnames(counts_all) <- sub("_MERGE$", "", sub("^mehreen_", "",
                                                colnames(counts_all)))

make_pooled_infection_results_for_scenario <- function(scenario_label,
                                                       exclude_outliers) {
  metadata_exp <- metadata_all_base
  if (exclude_outliers) {
    metadata_exp <- metadata_exp |>
      dplyr::filter(!label %in% current_outliers)
  }
  metadata_exp <- metadata_exp |>
    dplyr::arrange(Treatment, Diet, label)
  experiment_counts <- counts_all[, metadata_exp$label, drop = FALSE]
  run_pooled_infection_deseq(
    experiment_counts,
    metadata_exp,
    scenario_label
  )
}

pooled_infection_results <- dplyr::bind_rows(
  make_pooled_infection_results_for_scenario("All samples", FALSE),
  make_pooled_infection_results_for_scenario("Current outliers removed", TRUE)
)

fat_results_all <- dplyr::bind_rows(
  fat_results,
  pooled_infection_results
)

fatbody_reference_meta_top <- parse_fatbody_reference_metadata(
  fatbody_reference_metadata,
  fatbody_reference_dir
)
head_thorax_reference_meta_top <- parse_head_thorax_reference_metadata(
  head_thorax_counts_dir
)
reference_meta_top <- dplyr::bind_rows(
  fatbody_reference_meta_top,
  head_thorax_reference_meta_top
) |>
  dplyr::mutate(
    tissue = factor(tissue, levels = c("Fat body", "Head", "Thorax")),
    phase = factor(phase, levels = c("Gregarious", "Solitarious"))
  ) |>
  dplyr::arrange(tissue, phase, replicate)

reference_count_tables_top <- Map(
  read_featurecounts_gene_counts,
  reference_meta_top$count_file,
  reference_meta_top$sample_id
)
reference_counts_df_top <- Reduce(
  function(x, y) dplyr::full_join(x, y, by = "gene_id"),
  reference_count_tables_top
)
reference_counts_all_top <- as.matrix(reference_counts_df_top[, -1, drop = FALSE])
rownames(reference_counts_all_top) <- reference_counts_df_top$gene_id
reference_counts_all_top[is.na(reference_counts_all_top)] <- 0L
storage.mode(reference_counts_all_top) <- "integer"

tissue_results <- run_pairwise_tissue_deseq(
  reference_counts_all_top,
  as.data.frame(reference_meta_top)
)
tissue_sig <- tissue_results |>
  dplyr::filter(significant)
tissue_genes <- unique(tissue_sig$gene_id)

head_phase <- read_phase_table(head_phase_file, "Head")
head_sig <- head_phase |>
  dplyr::filter(significant)
head_genes <- unique(head_sig$gene_id)

thorax_phase <- read_phase_table(thorax_phase_file, "Thorax")
thorax_sig <- thorax_phase |>
  dplyr::filter(significant)
thorax_genes <- unique(thorax_sig$gene_id)
phase_genes <- union(head_genes, thorax_genes)

make_experiment_gene_sets_for_scenario <- function(scenario_label) {
  scenario_sig <- fat_results_all |>
    dplyr::filter(scenario == scenario_label, significant)

	  list(
	    "DEG Diet (controls)" = unique(scenario_sig$gene_id[
	      scenario_sig$comparison_family == "Diet contrasts: controls"
	    ]),
	    "DEG infection x diet" = unique(scenario_sig$gene_id[
	      scenario_sig$comparison_family == "Infection within diet"
	    ])
	  )
}

make_independent_gene_sets_for_scenario <- function(scenario_label) {
  c(
    list(
      "DEG tissues" = tissue_genes,
      "DEG phase changes" = phase_genes
    ),
    make_experiment_gene_sets_for_scenario(scenario_label)
  )
}

make_independent_set_counts_for_scenario <- function(scenario_label) {
  gene_sets <- make_independent_gene_sets_for_scenario(scenario_label)
  tibble::tibble(
    scenario = scenario_label,
    logical_set = names(gene_sets),
    n_genes = vapply(gene_sets, length, integer(1))
  )
}

workflow_scenarios <- c("All samples", "Current outliers removed")
workflow_category_counts <- dplyr::bind_rows(lapply(
  workflow_scenarios,
  make_independent_set_counts_for_scenario
))

scenario_set_counts <- tibble::tibble(
  scenario = workflow_scenarios,
  tissue_reference_deg_n = length(tissue_genes),
  head_phase_deg_n = length(head_genes),
  thorax_phase_deg_n = length(thorax_genes),
  phase_reference_deg_n = length(phase_genes),
  infection_within_diet_deg_n = vapply(
    workflow_scenarios,
    function(x) length(make_experiment_gene_sets_for_scenario(x)[[
      "DEG infection x diet"
    ]]),
    integer(1)
  ),
	  diet_control_deg_n = vapply(
	    workflow_scenarios,
	    function(x) length(make_experiment_gene_sets_for_scenario(x)[[
	      "DEG Diet (controls)"
	    ]]),
	    integer(1)
	  )
)

main_workflow_counts <- scenario_set_counts |>
  dplyr::filter(scenario == "All samples")

workflow_nodes <- tibble::tibble(
  node = c("Fat body QC", "Fat body experiment sets", "Reference tissue set",
           "Phase reference set", "Four-set overlap", "Sensitivity"),
  node_type = c("QC", "Experiment", "Reference", "Reference", "Output",
                "Sensitivity"),
  x = c(1, 3.35, 5.7, 8.05, 10.4, 12.75),
  y = c(1, 1, 1, 1, 1, 1),
  label = c(
    paste0("Step 1\nFat body experiment QC\n",
           "sample matrix + PCA\n",
           "outliers retained and marked"),
	    paste0("Step 2\nFat body DEGs\n",
	           "control diet n = ", scales::comma(main_workflow_counts$diet_control_deg_n),
	           "\ninfection within diet n = ",
	           scales::comma(main_workflow_counts$infection_within_diet_deg_n)),
    paste0("Step 3\nDEG tissues\n",
           "model ~ tissue\n",
           "n = ", scales::comma(main_workflow_counts$tissue_reference_deg_n)),
	    paste0("Step 4\nDEG phase changes\n",
	           "head or thorax\n",
	           "n = ", scales::comma(main_workflow_counts$phase_reference_deg_n)),
	    paste0("Step 5\nFour-set overlap\n",
	           "Venn / UpSet\n",
	           "category heatmaps"),
    paste0("Step 6\nSensitivity\n",
           "same logic\n",
           "without current outliers")
  )
)

workflow_edges <- tibble::tibble(
  x = workflow_nodes$x[-nrow(workflow_nodes)] + 0.75,
  y = workflow_nodes$y[-nrow(workflow_nodes)],
  xend = workflow_nodes$x[-1] - 0.75,
  yend = workflow_nodes$y[-1]
)
```

```{r deduction-workflow-summary, fig.width=26, fig.height=6.5}
workflow_plot <- ggplot() +
  geom_segment(
    data = workflow_edges,
    aes(x = x, y = y, xend = xend, yend = yend),
    arrow = grid::arrow(length = grid::unit(0.14, "inches")),
    linewidth = 0.45,
    color = "#4B5563"
  ) +
  geom_label(
    data = workflow_nodes,
    aes(x = x, y = y, label = label, fill = node_type),
    label.size = 0.25,
    color = "#111827",
    size = 4.7,
    lineheight = 0.94,
    label.padding = grid::unit(0.2, "lines")
  ) +
  scale_fill_manual(
    values = c(
      "QC" = "#E5E7EB",
      "Experiment" = "#F4C7B8",
      "Reference" = "#C7D7F2",
      "Output" = "#F7F3E8",
      "Sensitivity" = "#DDD6FE"
    ),
    guide = "none"
  ) +
  coord_cartesian(xlim = c(0.1, 13.7), ylim = c(0.25, 1.75),
                  clip = "off") +
  labs(
    title = "Independent DEG set construction workflow",
    subtitle = "Compute tissue, phase, infection, and diet DEG sets independently; compare them only after each set is defined",
    x = NULL,
    y = NULL
  ) +
  theme_void(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, size = 13),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.margin = ggplot2::margin(10, 20, 10, 20)
  )

workflow_plot
  ggsave(
  file.path(out_dir, "step0_deduction_workflow_all_samples_with_outliers.png"),
  workflow_plot,
  width = 26,
  height = 6.5,
  dpi = 300
)
write.csv(
  workflow_category_counts,
  file.path(out_dir, "step0_deduction_workflow_category_counts_by_scenario.csv"),
  row.names = FALSE
)
```

## Stepwise computations and visual checks

The figures below show each independently computed set before the overlap
summary. Sample labels with `*` are the current outlier-flagged fat body samples.
The column strips above each heatmap are the sample annotations: diet and
infection status for the fat body experiment, and tissue plus
gregarious/solitarious phase for the reference samples.

```{r scenario-annotated-heatmap-prep, include=FALSE}
make_vst_matrix <- function(count_mat, coldata, design_formula = ~ 1) {
  keep <- rowSums(count_mat) >= 10
  dds <- DESeq2::DESeqDataSetFromMatrix(
    countData = count_mat[keep, , drop = FALSE],
    colData = coldata,
    design = design_formula
  )
  vst <- DESeq2::vst(dds, blind = TRUE)
  SummarizedExperiment::assay(vst)
}

tissue_reference_coldata_top <- as.data.frame(reference_meta_top) |>
  dplyr::select(sample_id, tissue, phase, rearing_condition, replicate)
rownames(tissue_reference_coldata_top) <- tissue_reference_coldata_top$sample_id

vst_tissue_reference_top <- make_vst_matrix(
  reference_counts_all_top,
  tissue_reference_coldata_top,
  ~ tissue
)

tissue_sample_matrix_plot <- plot_sample_matrix(
  reference_meta_top,
  "Reference model: tissue sample matrix",
  file.path(out_dir, "step2_tissue_reference_sample_matrix.png"),
  width = 10,
  height = 6
)

tissue_pca_plot <- plot_pca_from_vst(
  vst_tissue_reference_top,
  as.data.frame(reference_meta_top),
  "Reference overlay: PCA of fat body, head, and thorax reference samples",
  "VST PCA of copied reference count files used for the tissue DEG model",
  file.path(out_dir, "step2_tissue_reference_pca.png"),
  width = 11,
  height = 7
)

make_reference_vst <- function(tissue_name) {
  tissue_meta <- reference_meta_top |>
    dplyr::filter(tissue == tissue_name) |>
    dplyr::arrange(phase, replicate)
  tissue_counts <- reference_counts_all_top[
    ,
    tissue_meta$sample_id,
    drop = FALSE
  ]
  tissue_coldata <- tissue_meta |>
    dplyr::select(sample_id, phase, rearing_condition, replicate) |>
    dplyr::mutate(
      phase = factor(phase, levels = c("Gregarious", "Solitarious"))
    )
  rownames(tissue_coldata) <- tissue_coldata$sample_id
  make_vst_matrix(tissue_counts, tissue_coldata, ~ phase)
}

vst_head_top <- make_reference_vst("Head")
vst_thorax_top <- make_reference_vst("Thorax")

reference_sample_metadata_top <- reference_meta_top |>
  dplyr::transmute(
    sample_id,
    sample_label,
    sample_block = dplyr::case_when(
      as.character(tissue) == "Fat body" ~ "Reference fat body",
      TRUE ~ paste(as.character(tissue), "reference")
    ),
    tissue,
    phase,
    Treatment = "Reference",
    Diet = "Reference",
    column_group = paste(tissue, phase)
  )

tissue_gene_summary_top <- tissue_sig |>
  dplyr::group_by(gene_id) |>
  dplyr::summarise(
    n_tissue_contrasts = dplyr::n_distinct(comparison),
    tissue_comparisons = paste(sort(unique(comparison)), collapse = "; "),
    tissue_directions = paste(sort(unique(direction)), collapse = "; "),
    tissue_max_abs_log2FC = max(abs(log2FoldChange), na.rm = TRUE),
    tissue_min_padj = min(padj, na.rm = TRUE),
    .groups = "drop"
  )

head_gene_summary_top <- head_sig |>
  dplyr::group_by(gene_id) |>
  dplyr::summarise(
    head_log2FC = log2FoldChange[which.min(padj)][1],
    head_padj = min(padj, na.rm = TRUE),
    head_direction = direction[which.min(padj)][1],
    .groups = "drop"
  )

thorax_gene_summary_top <- thorax_sig |>
  dplyr::group_by(gene_id) |>
  dplyr::summarise(
    thorax_log2FC = log2FoldChange[which.min(padj)][1],
    thorax_padj = min(padj, na.rm = TRUE),
    thorax_direction = direction[which.min(padj)][1],
    .groups = "drop"
  )

make_tissue_heatmap_genes <- function(tissue_summary, vst_mat) {
  tissue_summary |>
    dplyr::filter(gene_id %in% rownames(vst_mat)) |>
    dplyr::left_join(annotation, by = "gene_id") |>
    dplyr::mutate(
      Preferred_name = dplyr::coalesce(Preferred_name, "unannotated"),
      Description = dplyr::coalesce(Description, "no description"),
      priority_score = n_tissue_contrasts * 10 +
        tissue_max_abs_log2FC +
        dplyr::coalesce(safe_neglog10(tissue_min_padj), 0),
      gene_label = paste0(
        gene_id, " | ", Preferred_name, " | ",
        stringr::str_trunc(Description, 42)
      )
    ) |>
    dplyr::arrange(dplyr::desc(n_tissue_contrasts),
                   tissue_min_padj,
                   dplyr::desc(tissue_max_abs_log2FC)) |>
    dplyr::slice_head(n = params$tissue_heatmap_gene_n)
}

make_phase_heatmap_genes <- function(sig_table, vst_mat, label_prefix) {
  sig_table |>
    dplyr::filter(gene_id %in% rownames(vst_mat)) |>
    dplyr::left_join(annotation, by = "gene_id") |>
    dplyr::mutate(
      Preferred_name = dplyr::coalesce(Preferred_name, "unannotated"),
      Description = dplyr::coalesce(Description, "no description"),
      gene_label = paste0(
        gene_id, " | ", Preferred_name, " | ",
        stringr::str_trunc(Description, 42)
      )
    ) |>
    dplyr::arrange(padj, dplyr::desc(abs(log2FoldChange))) |>
    dplyr::slice_head(n = params$phase_heatmap_gene_n) |>
    dplyr::mutate(source_set = label_prefix)
}

tissue_heatmap_genes_top <- make_tissue_heatmap_genes(
  tissue_gene_summary_top,
  vst_tissue_reference_top
)
head_heatmap_genes_top <- make_phase_heatmap_genes(
  head_sig,
  vst_head_top,
  "Head phase DEG"
)
thorax_heatmap_genes_top <- make_phase_heatmap_genes(
  thorax_sig,
  vst_thorax_top,
  "Thorax phase DEG"
)

make_experiment_sample_metadata <- function(metadata_exp) {
  metadata_exp |>
    dplyr::transmute(
      sample_id = label,
      sample_label = dplyr::if_else(
        label %in% current_outliers,
        paste0(label, "*"),
        label
      ),
      sample_block = "Experiment fat body",
      tissue = "Fat body",
      phase = "Experiment",
      Treatment = as.character(Treatment),
      Diet = as.character(Diet),
      column_group = paste(Treatment, "diet", Diet)
    )
}

make_experiment_gene_summary_for_scenario <- function(scenario_label) {
  fat_results_all |>
    dplyr::filter(scenario == scenario_label, significant) |>
    dplyr::group_by(gene_id) |>
    dplyr::summarise(
      experiment_deg = TRUE,
      n_experiment_contrasts = dplyr::n_distinct(comparison),
      experiment_families = paste(sort(unique(comparison_family)), collapse = "; "),
      experiment_comparisons = paste(sort(unique(comparison)), collapse = "; "),
      experiment_directions = paste(sort(unique(direction)), collapse = "; "),
      experiment_max_abs_log2FC = max(abs(log2FoldChange), na.rm = TRUE),
      experiment_min_padj = min(padj, na.rm = TRUE),
      .groups = "drop"
    )
}

make_experiment_heatmap_genes <- function(experiment_summary, vst_mat) {
  experiment_summary |>
    dplyr::filter(gene_id %in% rownames(vst_mat)) |>
    dplyr::left_join(annotation, by = "gene_id") |>
    dplyr::mutate(
      Preferred_name = dplyr::coalesce(Preferred_name, "unannotated"),
      Description = dplyr::coalesce(Description, "no description"),
      priority_score = n_experiment_contrasts * 10 +
        experiment_max_abs_log2FC +
        dplyr::coalesce(safe_neglog10(experiment_min_padj), 0),
      gene_label = paste0(
        gene_id, " | ", Preferred_name, " | ",
        stringr::str_trunc(Description, 42)
      )
    ) |>
    dplyr::arrange(dplyr::desc(n_experiment_contrasts),
                   experiment_min_padj,
                   dplyr::desc(experiment_max_abs_log2FC)) |>
    dplyr::slice_head(n = params$experiment_heatmap_gene_n)
}

make_independent_membership_for_scenario <- function(scenario_label,
                                                     experiment_summary) {
  independent_sets <- make_independent_gene_sets_for_scenario(scenario_label)

  make_set_membership(independent_sets) |>
    dplyr::rename(
      tissue_reference_deg = `DEG tissues`,
      phase_reference_deg = `DEG phase changes`,
      diet_control_deg = `DEG Diet (controls)`,
      fatbody_infection_deg = `DEG infection x diet`
    ) |>
    dplyr::mutate(
      diet_infected_deg = FALSE,
      head_phase_deg = gene_id %in% head_genes,
      thorax_phase_deg = gene_id %in% thorax_genes,
      phase_reference_source = dplyr::case_when(
        head_phase_deg & thorax_phase_deg ~ "Head + thorax",
        head_phase_deg ~ "Head",
        thorax_phase_deg ~ "Thorax",
        TRUE ~ "Not phase"
      ),
      fatbody_experiment_deg = fatbody_infection_deg | diet_control_deg
    ) |>
    dplyr::left_join(tissue_gene_summary_top, by = "gene_id") |>
    dplyr::left_join(experiment_summary, by = "gene_id") |>
    dplyr::left_join(head_gene_summary_top, by = "gene_id") |>
    dplyr::left_join(thorax_gene_summary_top, by = "gene_id") |>
    dplyr::left_join(annotation, by = "gene_id") |>
    dplyr::mutate(
      Preferred_name = dplyr::coalesce(Preferred_name, "unannotated"),
      Description = dplyr::coalesce(Description, "no description"),
      combined_priority_score =
        dplyr::coalesce(n_tissue_contrasts, 0) * 8 +
        dplyr::coalesce(n_experiment_contrasts, 0) * 10 +
        dplyr::coalesce(tissue_max_abs_log2FC, 0) +
        dplyr::coalesce(experiment_max_abs_log2FC, 0) +
        dplyr::coalesce(abs(head_log2FC), 0) +
        dplyr::coalesce(abs(thorax_log2FC), 0) +
        dplyr::coalesce(safe_neglog10(tissue_min_padj), 0) +
        dplyr::coalesce(safe_neglog10(experiment_min_padj), 0) +
        dplyr::coalesce(safe_neglog10(head_padj), 0) +
        dplyr::coalesce(safe_neglog10(thorax_padj), 0)
    ) |>
    add_overlap_evidence_annotations() |>
    dplyr::mutate(logical_category = as.character(primary_annotation_group))
}

make_category_heatmap_genes <- function(independent_membership_scenario,
                                        vst_mat) {
  display_levels <- c(
    "DEG tissues only",
    "DEG phase changes, no fat body DEG",
    "DEG Diet (controls), no phase",
    "DEG infection x diet, no phase",
    "DEG Diet (controls) + infection x diet, no phase",
    "DEG phase changes + Diet (controls)",
    "DEG phase changes + infection x diet",
    "DEG phase changes + Diet (controls) + infection x diet"
  )

  annotated <- independent_membership_scenario |>
    add_overlap_evidence_annotations() |>
    dplyr::filter(gene_id %in% rownames(vst_mat)) |>
    dplyr::mutate(
      heatmap_category = dplyr::case_when(
        tissue_reference_deg & !phase_reference_deg &
          !diet_control_deg & !fatbody_infection_deg ~ "DEG tissues only",
        phase_reference_deg &
          !diet_control_deg & !fatbody_infection_deg ~
          "DEG phase changes, no fat body DEG",

        !fatbody_infection_deg & diet_control_deg &
          !phase_reference_deg ~
          "DEG Diet (controls), no phase",
        !fatbody_infection_deg & diet_control_deg &
          phase_reference_deg ~
          "DEG phase changes + Diet (controls)",

        fatbody_infection_deg & !diet_control_deg &
          !phase_reference_deg ~
          "DEG infection x diet, no phase",
        fatbody_infection_deg & !diet_control_deg &
          phase_reference_deg ~
          "DEG phase changes + infection x diet",

        fatbody_infection_deg & diet_control_deg &
          !phase_reference_deg ~
          "DEG Diet (controls) + infection x diet, no phase",
        fatbody_infection_deg & diet_control_deg &
          phase_reference_deg ~
          "DEG phase changes + Diet (controls) + infection x diet",
        TRUE ~ NA_character_
      ),
      heatmap_category = factor(heatmap_category, levels = display_levels)
    ) |>
    dplyr::filter(!is.na(heatmap_category))

  annotated |>
    dplyr::group_by(heatmap_category) |>
    dplyr::arrange(dplyr::desc(combined_priority_score), .by_group = TRUE) |>
    dplyr::slice_head(n = params$category_heatmap_genes_per_class) |>
    dplyr::ungroup() |>
    dplyr::mutate(
      primary_annotation_group = heatmap_category,
      logical_category = as.character(heatmap_category),
      gene_label = paste0(
        gene_id, " | ", Preferred_name, " | ",
        stringr::str_trunc(Description, 38)
      )
    ) |>
    dplyr::arrange(
      primary_annotation_group,
      dplyr::desc(combined_priority_score)
    )
}

make_scenario_bundle <- function(scenario_label, exclude_outliers) {
  metadata_exp <- metadata_all_base
  if (exclude_outliers) {
    metadata_exp <- metadata_exp |>
      dplyr::filter(!label %in% current_outliers)
  }
  metadata_exp <- metadata_exp |>
    dplyr::arrange(Treatment, Diet, label)

  experiment_counts_scenario <- counts_all[, metadata_exp$label, drop = FALSE]
  rownames(metadata_exp) <- metadata_exp$label

  experiment_sample_metadata_scenario <- make_experiment_sample_metadata(
    metadata_exp
  )
  rownames(experiment_sample_metadata_scenario) <-
    experiment_sample_metadata_scenario$sample_id

  vst_experiment_scenario <- make_vst_matrix(
    experiment_counts_scenario,
    metadata_exp,
    ~ Diet + Treatment
  )

  shared_genes <- intersect(
    rownames(experiment_counts_scenario),
    rownames(reference_counts_all_top)
  )
  combined_counts_scenario <- cbind(
    experiment_counts_scenario[shared_genes, , drop = FALSE],
    reference_counts_all_top[
      shared_genes,
      reference_meta_top$sample_id,
      drop = FALSE
    ]
  )
  combined_sample_metadata_scenario <- dplyr::bind_rows(
    experiment_sample_metadata_scenario,
    reference_sample_metadata_top
  )
  combined_sample_metadata_scenario <-
    combined_sample_metadata_scenario[
      match(colnames(combined_counts_scenario),
            combined_sample_metadata_scenario$sample_id),
    ]
  rownames(combined_sample_metadata_scenario) <-
    combined_sample_metadata_scenario$sample_id

  vst_combined_scenario <- make_vst_matrix(
    combined_counts_scenario,
    combined_sample_metadata_scenario,
    ~ 1
  )

  experiment_summary_scenario <-
    make_experiment_gene_summary_for_scenario(scenario_label)
  experiment_heatmap_genes_scenario <- make_experiment_heatmap_genes(
    experiment_summary_scenario,
    vst_experiment_scenario
  )
  independent_sets_scenario <- make_independent_gene_sets_for_scenario(
    scenario_label
  )
  independent_membership_scenario <- make_independent_membership_for_scenario(
    scenario_label,
	    experiment_summary_scenario
	  )
	  category_heatmap_genes_scenario <- make_category_heatmap_genes(
	    independent_membership_scenario,
	    vst_combined_scenario
	  )

  scenario_slug <- dplyr::case_when(
    scenario_label == "All samples" ~ "all_samples",
    scenario_label == "Current outliers removed" ~ "outliers_removed",
    TRUE ~ stringr::str_replace_all(tolower(scenario_label), "[^a-z0-9]+", "_")
  )
  scenario_subtitle <- paste0("Scenario: ", scenario_label)

  list(
    scenario = scenario_label,
	    slug = scenario_slug,
	    experiment_summary = experiment_summary_scenario,
	    independent_sets = independent_sets_scenario,
	    independent_membership = independent_membership_scenario,
	    category_heatmap_genes = category_heatmap_genes_scenario,
    plots = list(
      tissue_reference_samples = plot_annotated_heatmap(
	        vst_tissue_reference_top,
	        tissue_heatmap_genes_top,
	        reference_sample_metadata_top,
	        "Reference model: DEG tissues in fat body, head, and thorax references",
	        "Rows are genes significant in at least one pairwise tissue contrast from the DESeq2 model ~ tissue",
        file.path(out_dir, paste0(scenario_slug,
                                  "_step1_tissue_reference_degs_heatmap.png")),
        width = 20,
        height = 18
      ),
      fatbody_sample_matrix = plot_fatbody_design_matrix(
        metadata_exp,
        paste0("Fat body experiment: sample matrix - ", scenario_label),
        file.path(out_dir, paste0(scenario_slug,
                                  "_step2_fatbody_sample_matrix.png")),
        width = 8,
        height = 5.5
      ),
      fatbody_pca = plot_experiment_pca(
        vst_experiment_scenario,
        experiment_sample_metadata_scenario,
        paste0("Fat body experiment: PCA - ", scenario_label),
        "VST PCA of the fat body experiment count matrix",
        file.path(out_dir, paste0(scenario_slug,
                                  "_step2_fatbody_experiment_pca.png")),
        width = 10,
        height = 7
      ),
      combined_pca = plot_combined_sample_pca(
        vst_combined_scenario,
        combined_sample_metadata_scenario,
        paste0("Reference overlay: experiment fat body plus reference samples - ",
               scenario_label),
        "VST PCA of fat body experiment samples together with fat body, head, and thorax references",
        file.path(out_dir, paste0(scenario_slug,
                                  "_step2_experiment_fatbody_plus_reference_tissues_pca.png")),
        width = 12,
        height = 8
      ),
      experiment_samples_only = plot_annotated_heatmap(
        vst_experiment_scenario,
        experiment_heatmap_genes_scenario,
        experiment_sample_metadata_scenario,
        "Fat body experiment DEGs in fat body samples",
        paste0(scenario_subtitle, "; column strips show infection status and diet"),
        file.path(out_dir, paste0(scenario_slug,
                                  "_step5_experiment_degs_fatbody_samples_heatmap.png")),
        width = 22,
        height = 20
      ),
      experiment_all_samples = plot_annotated_heatmap(
        vst_combined_scenario,
        experiment_heatmap_genes_scenario,
        combined_sample_metadata_scenario,
        "Fat body experiment DEGs projected across all sample types",
        paste0(scenario_subtitle, "; fat body experiment, head, and thorax columns shown together"),
        file.path(out_dir, paste0(scenario_slug,
                                  "_step5_experiment_degs_all_samples_heatmap.png")),
        width = 30,
        height = 22
      ),
      head_samples_only = plot_annotated_heatmap(
        vst_head_top,
        head_heatmap_genes_top,
        reference_sample_metadata_top |>
          dplyr::filter(tissue == "Head"),
        "Reference overlay: DEG phase changes in head samples",
        "Column strips show gregarious versus solitarious head references",
        file.path(out_dir, paste0(scenario_slug,
                                  "_step4_head_degs_head_samples_heatmap.png")),
        width = 14,
        height = 16
      ),
      head_all_samples = plot_annotated_heatmap(
        vst_combined_scenario,
        head_heatmap_genes_top,
        combined_sample_metadata_scenario,
        "Reference overlay: DEG phase changes from head across all sample types",
        paste0(scenario_subtitle, "; fat body experiment, head, and thorax columns shown together"),
        file.path(out_dir, paste0(scenario_slug,
                                  "_step4_head_degs_all_samples_heatmap.png")),
        width = 28,
        height = 16
      ),
      thorax_samples_only = plot_annotated_heatmap(
        vst_thorax_top,
        thorax_heatmap_genes_top,
        reference_sample_metadata_top |>
          dplyr::filter(tissue == "Thorax"),
        "Reference overlay: DEG phase changes in thorax samples",
        "Column strips show gregarious versus solitarious thorax references",
        file.path(out_dir, paste0(scenario_slug,
                                  "_step4_thorax_degs_thorax_samples_heatmap.png")),
        width = 14,
        height = 16
      ),
      thorax_all_samples = plot_annotated_heatmap(
        vst_combined_scenario,
        thorax_heatmap_genes_top,
        combined_sample_metadata_scenario,
        "Reference overlay: DEG phase changes from thorax across all sample types",
        paste0(scenario_subtitle, "; fat body experiment, head, and thorax columns shown together"),
        file.path(out_dir, paste0(scenario_slug,
                                  "_step4_thorax_degs_all_samples_heatmap.png")),
        width = 28,
        height = 16
      ),
      category_all_samples = plot_annotated_heatmap(
	        vst_combined_scenario,
	        category_heatmap_genes_scenario,
	        combined_sample_metadata_scenario,
	        "Final: phase-centered DEG categories across all sample types",
	        paste0(
	          scenario_subtitle,
	          "; row blocks emphasize phase overlap with diet and infection while tissue remains an evidence track"
	        ),
        file.path(out_dir, paste0(scenario_slug,
                                  "_final_logical_category_all_samples_heatmap.png")),
        width = 30,
        height = 28
      ),
      category_annotated_all_samples = plot_final_overlap_annotated_heatmap(
	        vst_combined_scenario,
	        category_heatmap_genes_scenario,
	        combined_sample_metadata_scenario,
	        "Final: annotated independent DEG overlap heatmap",
	        paste0(
	          scenario_subtitle,
	          "; phase-centered row blocks and evidence squares mark all set memberships"
	        ),
        file.path(out_dir, paste0(scenario_slug,
                                  "_final_annotated_overlap_heatmap.png")),
        width = 34,
        height = 30
      )
    )
  )
}

scenario_bundles <- list(
  "All samples" = make_scenario_bundle("All samples", exclude_outliers = FALSE),
  "Current outliers removed" = make_scenario_bundle(
    "Current outliers removed",
    exclude_outliers = TRUE
  )
)

write.csv(
  tissue_results,
  file.path(out_dir, "step3_tissue_deseq2_pairwise_results.csv"),
  row.names = FALSE
)
write.csv(
  tissue_gene_summary_top,
  file.path(out_dir, "step3_tissue_deg_gene_summary.csv"),
  row.names = FALSE
)
write.csv(
  head_sig,
  file.path(out_dir, "step3_head_phase_degs.csv"),
  row.names = FALSE
)
write.csv(
  thorax_sig,
  file.path(out_dir, "step4_thorax_phase_degs.csv"),
  row.names = FALSE
)
write.csv(
  pooled_infection_results,
  file.path(out_dir, "step5_pooled_infection_across_diets_deseq2_results.csv"),
  row.names = FALSE
)
write.csv(
  scenario_bundles[["All samples"]]$experiment_summary,
  file.path(out_dir, "all_samples_step5_experiment_deg_gene_summary.csv"),
  row.names = FALSE
)
write.csv(
  scenario_bundles[["Current outliers removed"]]$experiment_summary,
  file.path(out_dir,
            "outliers_removed_step5_experiment_deg_gene_summary.csv"),
  row.names = FALSE
)
write.csv(
  scenario_bundles[["All samples"]]$independent_membership,
  file.path(out_dir, "all_samples_independent_set_membership.csv"),
  row.names = FALSE
)
write.csv(
  scenario_bundles[["All samples"]]$category_heatmap_genes,
  file.path(out_dir, "all_samples_final_annotated_overlap_heatmap_genes.csv"),
  row.names = FALSE
)
write.csv(
  scenario_bundles[["Current outliers removed"]]$independent_membership,
  file.path(out_dir, "outliers_removed_independent_set_membership.csv"),
  row.names = FALSE
)
write.csv(
  scenario_bundles[["Current outliers removed"]]$category_heatmap_genes,
  file.path(out_dir,
            "outliers_removed_final_annotated_overlap_heatmap_genes.csv"),
  row.names = FALSE
)
```

## Main analysis: all samples with outliers retained

This section keeps all fat body experiment samples. The current outlier-flagged
samples are retained and marked with `*` in the fat body figures. The order is
intentionally stepwise: first inspect the fat body experiment, then compute the
reference tissue and phase sets independently, then compare all sets.

### Step 1: fat body experiment sample structure

These figures describe only the diet-infection fat body experiment. They are
placed first because the fat body experiment is the primary analysis; the
reference samples are interpretive overlays used later.

```{r all-samples-step2-fatbody-sample-matrix, fig.width=8, fig.height=5.5}
scenario_bundles[["All samples"]]$plots$fatbody_sample_matrix
```

```{r all-samples-step2-fatbody-pca, fig.width=10, fig.height=7}
scenario_bundles[["All samples"]]$plots$fatbody_pca
```

### Step 2: fat body diet/infection DEGs {.tabset}

#### DEG counts by contrast

```{r all-samples-step5-counts}
all_samples_fatbody_contrast_counts <- fat_results_all |>
  dplyr::filter(scenario == "All samples", significant) |>
  dplyr::count(comparison_family, comparison, direction,
               name = "n_deg_rows") |>
  dplyr::arrange(comparison_family, comparison, direction)

knitr::kable(
  all_samples_fatbody_contrast_counts,
  caption = "Fat body DEG counts, including the pooled infection contrast"
)
```

#### Fat body samples only

```{r all-samples-step5-fatbody-only, fig.width=22, fig.height=20}
scenario_bundles[["All samples"]]$plots$experiment_samples_only
```

### Step 3: independent tissue and phase reference sets {.tabset}

These signatures are computed independently from copied reference files. They
are not used to call fat body experiment DEGs. Instead, they define reference
tissue and phase DEG sets. The simplified final overlap uses `DEG tissues` and
`DEG phase changes`; the phase set is the union of head phase DEGs and thorax
phase DEGs.

The tissue DEG model is fitted only on the copied reference count files:
reference fat body, head, and thorax. Its design is `~ tissue`; phase is ignored
for this first tissue-only question. The reported contrasts are Fat body vs
Head, Fat body vs Thorax, and Head vs Thorax. The head and thorax phase DEG
tables are also exported separately, so genes in `DEG phase changes` can still
be traced back to head, thorax, or both.

#### Reference sample matrix

```{r all-samples-step2-reference-sample-matrix, fig.width=10, fig.height=6}
tissue_sample_matrix_plot
```

#### Reference PCA

```{r all-samples-step2-reference-pca, fig.width=11, fig.height=7}
tissue_pca_plot
```

#### Experiment plus references PCA

```{r all-samples-step2-combined-pca, fig.width=12, fig.height=8}
scenario_bundles[["All samples"]]$plots$combined_pca
```

#### DEG tissues counts

```{r all-samples-step1-tissue-counts}
tissue_contrast_counts <- tissue_sig |>
  dplyr::count(comparison, direction, name = "n_deg_rows") |>
  dplyr::arrange(comparison, direction)

knitr::kable(
  tissue_contrast_counts,
  caption = "DEG tissues counts by pairwise tissue contrast from the DESeq2 model ~ tissue"
)
```

#### DEG tissues heatmap

```{r all-samples-step1-tissue-heatmap, fig.width=20, fig.height=18}
scenario_bundles[["All samples"]]$plots$tissue_reference_samples
```

#### DEG phase changes: head

```{r all-samples-step4-head-only, fig.width=14, fig.height=16}
scenario_bundles[["All samples"]]$plots$head_samples_only
```

#### DEG phase changes: thorax

```{r all-samples-step4-thorax-only, fig.width=14, fig.height=16}
scenario_bundles[["All samples"]]$plots$thorax_samples_only
```

### Step 4: independent DEG set overlap {.tabset}

The Venn and UpSet plots in this section use four independent sets directly:
`DEG tissues`, `DEG phase changes`, `DEG Diet (controls)`, and
`DEG infection x diet`. These are not filtered through a fat body DEG list
before overlap is computed. In this page, `DEG infection x diet` means
infected-vs-control DEGs computed separately within each diet.

For the final heatmaps, row blocks are phase-centered because the main
biological concern is that solitarization or phase change may share many genes
with the fat body experiment. `DEG tissues` is kept as a separate evidence
track instead of being merged into the row names. This lets the rows highlight
`DEG phase changes + Diet (controls)`, `DEG phase changes + infection x diet`,
and `DEG phase changes + Diet (controls) + infection x diet`, while the blue
evidence square still shows whether the same gene is also a tissue DEG.

#### Independent set sizes

```{r all-samples-final-independent-set-counts}
all_samples_independent_set_counts <- workflow_category_counts |>
  dplyr::filter(scenario == "All samples") |>
  dplyr::arrange(logical_set)

knitr::kable(
  all_samples_independent_set_counts,
  caption = "All-samples independent DEG set sizes before overlap"
)
```

#### Heatmap display categories

```{r all-samples-final-display-category-counts}
all_samples_display_category_counts <-
  scenario_bundles[["All samples"]]$category_heatmap_genes |>
  dplyr::count(primary_annotation_group, name = "n_genes") |>
  dplyr::arrange(primary_annotation_group)

knitr::kable(
  all_samples_display_category_counts,
  caption = "Phase-centered display groups used to order the final heatmaps"
)
```

#### Fat body experiment DEGs across all sample types

```{r all-samples-step5-all-samples, fig.width=30, fig.height=22}
scenario_bundles[["All samples"]]$plots$experiment_all_samples
```

#### Venn diagram

```{r all-samples-final-venn, fig.width=11, fig.height=8}
all_samples_independent_sets <-
  scenario_bundles[["All samples"]]$independent_sets
all_samples_final_venn <- plot_deg_sets_venn(
  all_samples_independent_sets,
  "All samples: independent DEG set overlap",
  "Sets are computed independently before this overlap is drawn"
)
all_samples_final_venn
ggsave(
  file.path(out_dir, "all_samples_final_logical_sets_venn.png"),
  all_samples_final_venn,
  width = 11,
  height = 8,
  dpi = 300
)
```

#### UpSetR plot

```{r all-samples-final-upset, fig.width=11, fig.height=7}
write_upset_png(
  all_samples_independent_sets,
  file.path(out_dir, "all_samples_final_logical_sets_upset.png")
)
UpSetR::upset(
  UpSetR::fromList(all_samples_independent_sets),
  nsets = length(all_samples_independent_sets),
  nintersects = NA,
  order.by = "freq",
  keep.order = TRUE,
  mainbar.y.label = "Intersection size",
  sets.x.label = "Set size",
  text.scale = c(1.4, 1.4, 1.2, 1.2, 1.4, 1.2)
)
```

#### Category heatmap

```{r all-samples-final-category-heatmap, fig.width=30, fig.height=28}
scenario_bundles[["All samples"]]$plots$category_all_samples
```

#### Annotated category heatmap

Colored squares on the left mark which independent DEG sets contain each gene:
`DEG tissues`, `DEG phase changes`, `DEG Diet (controls)`, and
`DEG infection x diet`.

Rows are grouped by mutually exclusive interpretation classes. For example,
`DEG phase changes + infection x diet` means a gene is phase-related in the
head/thorax reference data and is also a fat body infected-vs-control DEG
within at least one diet. The `DEG tissues` evidence square is independent, so
the same row can still show whether that gene is also tissue-associated.

```{r all-samples-final-annotated-category-heatmap, fig.width=34, fig.height=30}
scenario_bundles[["All samples"]]$plots$category_annotated_all_samples
```

## Outlier-removed sensitivity

This section repeats the fat body and overlap parts after removing the current
outlier set (`1007`, `1036`, `1037`, and `1039`). The tissue and phase-change
reference sets are unchanged; only the diet-control and infection-within-diet
fat body sets and the combined cross-sample heatmaps change.

### Step 1: filtered fat body sample structure

```{r outliers-removed-step2-fatbody-sample-matrix, fig.width=8, fig.height=5.5}
scenario_bundles[["Current outliers removed"]]$plots$fatbody_sample_matrix
```

```{r outliers-removed-step2-fatbody-pca, fig.width=10, fig.height=7}
scenario_bundles[["Current outliers removed"]]$plots$fatbody_pca
```

### Step 2: filtered fat body diet/infection DEGs {.tabset}

This repeats the fat body DEG definition after removing the current outlier
set. The reference overlays are held constant so the sensitivity question is
focused on how the fat body experiment DEG set changes.

#### DEG counts by contrast

```{r outliers-removed-step5-counts}
outliers_removed_fatbody_contrast_counts <- fat_results_all |>
  dplyr::filter(scenario == "Current outliers removed", significant) |>
  dplyr::count(comparison_family, comparison, direction,
               name = "n_deg_rows") |>
  dplyr::arrange(comparison_family, comparison, direction)

knitr::kable(
  outliers_removed_fatbody_contrast_counts,
  caption = "Filtered fat body DEG counts, including the pooled infection contrast"
)
```

#### Fat body samples only

```{r outliers-removed-step5-fatbody-only, fig.width=22, fig.height=20}
scenario_bundles[["Current outliers removed"]]$plots$experiment_samples_only
```

### Step 3: filtered independent DEG set overlap {.tabset}

As above, the Venn and UpSet plots compare four independent DEG sets:
`DEG tissues`, `DEG phase changes`, `DEG Diet (controls)`, and
`DEG infection x diet`. The two reference sets are held constant;
diet-control and infection-within-diet fat body sets are recomputed after
removing the current outliers.

#### Experiment plus references PCA

```{r outliers-removed-step2-combined-pca, fig.width=12, fig.height=8}
scenario_bundles[["Current outliers removed"]]$plots$combined_pca
```

#### Fat body experiment DEGs across all sample types

```{r outliers-removed-step5-all-samples, fig.width=30, fig.height=22}
scenario_bundles[["Current outliers removed"]]$plots$experiment_all_samples
```

#### Independent set sizes

```{r outliers-removed-final-independent-set-counts}
outliers_removed_independent_set_counts <- workflow_category_counts |>
  dplyr::filter(scenario == "Current outliers removed") |>
  dplyr::arrange(logical_set)

knitr::kable(
  outliers_removed_independent_set_counts,
  caption = "Outlier-removed independent DEG set sizes before overlap"
)
```

#### Heatmap display categories

```{r outliers-removed-final-display-category-counts}
outliers_removed_display_category_counts <-
  scenario_bundles[["Current outliers removed"]]$category_heatmap_genes |>
  dplyr::count(primary_annotation_group, name = "n_genes") |>
  dplyr::arrange(primary_annotation_group)

knitr::kable(
  outliers_removed_display_category_counts,
  caption = "Phase-centered display groups used to order the outlier-removed heatmaps"
)
```

#### Venn diagram

```{r outliers-removed-final-venn, fig.width=11, fig.height=8}
outliers_removed_independent_sets <-
  scenario_bundles[["Current outliers removed"]]$independent_sets
outliers_removed_final_venn <- plot_deg_sets_venn(
  outliers_removed_independent_sets,
  "Current outliers removed: independent DEG set overlap",
  "Reference sets are fixed; fat body sets are recomputed after outlier removal"
)
outliers_removed_final_venn
ggsave(
  file.path(out_dir, "outliers_removed_final_logical_sets_venn.png"),
  outliers_removed_final_venn,
  width = 11,
  height = 8,
  dpi = 300
)
```

#### UpSetR plot

```{r outliers-removed-final-upset, fig.width=11, fig.height=7}
write_upset_png(
  outliers_removed_independent_sets,
  file.path(out_dir, "outliers_removed_final_logical_sets_upset.png")
)
UpSetR::upset(
  UpSetR::fromList(outliers_removed_independent_sets),
  nsets = length(outliers_removed_independent_sets),
  nintersects = NA,
  order.by = "freq",
  keep.order = TRUE,
  mainbar.y.label = "Intersection size",
  sets.x.label = "Set size",
  text.scale = c(1.4, 1.4, 1.2, 1.2, 1.4, 1.2)
)
```

#### Category heatmap

```{r outliers-removed-final-category-heatmap, fig.width=30, fig.height=28}
scenario_bundles[["Current outliers removed"]]$plots$category_all_samples
```

#### Annotated category heatmap

Colored squares on the left mark which independent DEG sets contain each gene:
`DEG tissues`, `DEG phase changes`, `DEG Diet (controls)`, and
`DEG infection x diet`.

Rows are grouped by mutually exclusive interpretation classes. For example,
`DEG phase changes + infection x diet` means a gene is phase-related in the
head/thorax reference data and is also a fat body infected-vs-control DEG
within at least one diet. The `DEG tissues` evidence square is independent, so
the same row can still show whether that gene is also tissue-associated.

```{r outliers-removed-final-annotated-category-heatmap, fig.width=34, fig.height=30}
scenario_bundles[["Current outliers removed"]]$plots$category_annotated_all_samples
```

## Exported files

```{r exported-files}
exports <- list.files(out_dir, full.names = FALSE)
knitr::kable(
  tibble::tibble(file = exports),
  caption = paste("Files written to", out_dir)
)
```
