---
title: "Main-chromosome all-samples DEG catalogue"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  placement_file: "output/runs/host_pathogen_dual_20260727_000952/local-analysis/host-mapping-audit_20260729-163928/unified-model/audit-summary/host_gene_sequence_placement.tsv"
  eggnog_file: "data/GO_Annotations/GCF_023897955.1_iqSchGreg1.2_Arthopoda_one2one.emapper.annotations"
  gff_file: "data/reference/GCF_023897955.1_iqSchGreg1.2_genomic.gff"
  latest_results_root: "output/rmd_runs"
  results_pattern: "^placed-unplaced-scaffold-analysis_"
  cluster_results_pattern: "^main-chromosome-cluster-enrichment_"
  exon_results_pattern: "^exon-only-deseq2-analysis_"
  primary_counts: "output/rmd_runs/transcript-exon-primary-excluding1044_20260809-233420/host_counts_no_rrna.tsv"
  metadata: "data/metadata/mehreen_metadata_analysis_44_excluding_1044.txt"
  alpha: 0.05
  lfc_threshold: 1
  large_lfc_threshold: 2
  high_expression_threshold: 100
  output_tag: "main-chromosome-all-deg-catalogue"
---

## Purpose

::: {.walkthrough-grid}
::: {.walkthrough-step}
**Step 1 - Collect contrasts**

Gather every significant main-chromosome DEG call from the joint model.
:::
::: {.walkthrough-step}
**Step 2 - Add annotation**

Join exact GFF descriptions, biotype, effect size, direction, and functional IDs.
:::
::: {.walkthrough-step}
**Step 3 - Search genes**

Filter the table by gene, contrast, description, direction, or expression cluster.
:::
:::

This page is a searchable catalogue of all significant DEGs from the primary
all-samples retained, main-chromosome analysis. It uses the joint competitive
host DESeq2 model from page 28 and restricts the catalogue to genes assigned to
nuclear chromosomes by the official placement table.

The table shows the gene ID, RefSeq/GFF description, contrast, log2 fold
change, and which side of the contrast has higher expression. Positive log2
fold change is colored red; negative log2 fold change is colored blue.

The `Description` column is taken directly from the gene-level `description=`
attribute in the RefSeq/GFF annotation from the same *S. gregaria* assembly.
eggNOG is still used only for the GO, KEGG, and PFAM annotation columns.

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c("dplyr", "tidyr", "stringr", "DT", "knitr", "ggplot2", "ggrepel",
              "scales", "readr", "tibble")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

placement_file <- file.path(project_dir, params$placement_file)
eggnog_file <- file.path(project_dir, params$eggnog_file)
gff_file <- file.path(project_dir, params$gff_file)
results_root <- file.path(project_dir, params$latest_results_root)

direction_colors <- c("Upregulated" = "#E41A1C",
                      "Downregulated" = "#1E88E5")
cluster_colors <- c("DEG cluster 1" = "#315C8A",
                    "DEG cluster 2" = "#C85A54",
                    "DEG cluster 3+4" = "#2F8F5B")
```

## Helper functions

```{r helpers}
latest_dir <- function(pattern) {
  dirs <- list.dirs(results_root, recursive = FALSE, full.names = TRUE)
  dirs <- dirs[grepl(pattern, basename(dirs))]
  if (length(dirs) == 0) {
    stop("No output folder matched pattern: ", pattern)
  }
  dirs[which.max(file.info(dirs)$mtime)]
}

read_emapper <- function(path) {
  cols <- c(
    "query", "seed_ortholog", "evalue", "score", "eggNOG_OGs",
    "max_annot_lvl", "COG_category", "eggNOG_Description", "eggNOG_name",
    "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
    "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC", "CAZy",
    "BiGG_Reaction", "PFAMs"
  )
  tab <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", fill = TRUE)
  names(tab) <- cols[seq_len(ncol(tab))]
  tab |>
    dplyr::mutate(protein_id = stringr::str_extract(query, "XP_[0-9]+[.][0-9]+"))
}

parse_cds_gene_bridge <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  gff |>
    dplyr::filter(type == "CDS") |>
    dplyr::transmute(
      protein_id = stringr::str_match(attributes, "protein_id=([^;]+)")[, 2],
      gene_id = stringr::str_match(attributes, "gene=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(protein_id), !is.na(gene_id)) |>
    dplyr::distinct()
}

parse_gff_gene_annotation <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  decoded <- gff |>
    dplyr::filter(type == "gene") |>
    dplyr::transmute(
      gene_id = dplyr::coalesce(
        stringr::str_match(attributes, "gene=([^;]+)")[, 2],
        stringr::str_match(attributes, "Name=([^;]+)")[, 2]
      ),
      gff_description = stringr::str_match(attributes, "description=([^;]+)")[, 2],
      gene_biotype = stringr::str_match(attributes, "gene_biotype=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(gene_id)) |>
    dplyr::distinct(gene_id, .keep_all = TRUE)

  decoded$gene_id <- vapply(
    decoded$gene_id,
    utils::URLdecode,
    character(1),
    USE.NAMES = FALSE
  )
  has_description <- !is.na(decoded$gff_description)
  decoded$gff_description[has_description] <- vapply(
    decoded$gff_description[has_description],
    utils::URLdecode,
    character(1),
    USE.NAMES = FALSE
  )
  decoded
}

first_nonempty <- function(x, fallback = "unannotated") {
  x <- x[!is.na(x) & x != "" & x != "-"]
  if (length(x) == 0) fallback else x[1]
}

make_condition_label <- function(comparison_type, numerator, denominator, diet,
                                 side = c("numerator", "denominator")) {
  side <- match.arg(side)
  value <- if (side == "numerator") numerator else denominator
  dplyr::case_when(
    comparison_type == "Infection within diet" ~
      paste(value, "diet", diet),
    grepl("control", comparison_type, ignore.case = TRUE) ~
      paste("Control diet", value),
    grepl("infected", comparison_type, ignore.case = TRUE) ~
      paste("Infected diet", value),
    TRUE ~ value
  )
}

save_png_svg <- function(plot, file_base, width, height, dpi = 300) {
  png_path <- file.path(out_dir, paste0(file_base, ".png"))
  svg_path <- file.path(out_dir, paste0(file_base, ".svg"))

  ggplot2::ggsave(png_path, plot, width = width, height = height,
                  dpi = dpi, limitsize = FALSE)

  grDevices::svg(svg_path, width = width, height = height, onefile = FALSE)
  on.exit(grDevices::dev.off(), add = TRUE)
  print(plot)

  invisible(c(png = png_path, svg = svg_path))
}

slugify <- function(x) {
  x |>
    stringr::str_to_lower() |>
    stringr::str_replace_all("[^a-z0-9]+", "_") |>
    stringr::str_replace_all("^_|_$", "")
}

read_count_matrix <- function(path) {
  table <- readr::read_tsv(path, show_col_types = FALSE)
  if (!"gene_id" %in% names(table)) stop("Count matrix lacks gene_id: ", path)
  ids <- as.character(table$gene_id)
  matrix <- as.matrix(table[, setdiff(names(table), "gene_id")])
  storage.mode(matrix) <- "numeric"
  rownames(matrix) <- ids
  matrix
}
```

## Input files

```{r input-files}
placed_dir <- latest_dir(params$results_pattern)
scenario_results_file <- file.path(placed_dir, "placed_unplaced_all_deseq2_results.csv")
size_factor_file <- file.path(
  placed_dir,
  "competitive_host_placed_gene_size_factors.tsv"
)
exon_dir <- latest_dir(params$exon_results_pattern)
exon_results_file <- file.path(exon_dir, "placed_unplaced_all_deseq2_results.csv")
primary_counts_file <- file.path(project_dir, params$primary_counts)
metadata_file <- file.path(project_dir, params$metadata)

input_files <- data.frame(
  input = c(
    "Joint main-chromosome DESeq2 table", "Placed-gene DESeq2 size factors",
    "Transcript-plus-exon count matrix", "Analysis metadata",
    "Exon-only DESeq2 sensitivity table", "Official gene placement",
    "eggNOG functional annotation", "GFF gene descriptions"
  ),
  path = c(
    scenario_results_file, size_factor_file, primary_counts_file, metadata_file,
    exon_results_file, placement_file, eggnog_file, gff_file
  ),
  present = c(file.exists(scenario_results_file),
              file.exists(size_factor_file), file.exists(primary_counts_file),
              file.exists(metadata_file), file.exists(exon_results_file),
              file.exists(placement_file), file.exists(eggnog_file),
              file.exists(gff_file))
)

knitr::kable(input_files, caption = "Input files used for the DEG catalogue")
if (!all(input_files$present)) {
  stop("One or more required inputs are missing.")
}
```

## Annotation and DEG catalogue

```{r build-catalogue}
placement <- readr::read_tsv(placement_file, show_col_types = FALSE) |>
  dplyr::distinct(gene_id, .keep_all = TRUE)
gene_universe <- placement$gene_id[
  placement$sequence_class == "Nuclear chromosome"
]

gff_annotation <- parse_gff_gene_annotation(gff_file) |>
  dplyr::filter(gene_id %in% gene_universe)

eggnog_annotation <- read_emapper(eggnog_file) |>
  dplyr::left_join(parse_cds_gene_bridge(gff_file), by = "protein_id") |>
  dplyr::filter(!is.na(gene_id), gene_id %in% gene_universe) |>
  dplyr::group_by(gene_id) |>
  dplyr::summarise(
    GOs = first_nonempty(GOs, ""),
    KEGG_ko = first_nonempty(KEGG_ko, ""),
    KEGG_Pathway = first_nonempty(KEGG_Pathway, ""),
    PFAMs = first_nonempty(PFAMs, ""),
    .groups = "drop"
  )

annotation <- data.frame(gene_id = gene_universe) |>
  dplyr::left_join(gff_annotation, by = "gene_id") |>
  dplyr::left_join(eggnog_annotation, by = "gene_id") |>
  dplyr::mutate(
    Description = dplyr::coalesce(gff_description, "no GFF description"),
    gene_biotype = dplyr::coalesce(gene_biotype, "unknown")
  ) |>
  dplyr::select(
    gene_id, Description, gene_biotype,
    GOs, KEGG_ko, KEGG_Pathway, PFAMs
  )

contrast_map <- tibble::tribble(
  ~contrast_id, ~comparison_type, ~comparison,
  "infection_diet33", "Infection within diet", "Diet 33: infected vs control",
  "infection_diet50", "Infection within diet", "Diet 50: infected vs control",
  "infection_diet83", "Infection within diet", "Diet 83: infected vs control",
  "control_diet33_vs50", "Diet contrasts among control", "Control 33 vs 50",
  "control_diet83_vs33", "Diet contrasts among control", "Control 83 vs 33",
  "control_diet83_vs50", "Diet contrasts among control", "Control 83 vs 50",
  "infected_diet33_vs50", "Diet contrasts among infected", "Infected 33 vs 50",
  "infected_diet83_vs33", "Diet contrasts among infected", "Infected 83 vs 33",
  "infected_diet83_vs50", "Diet contrasts among infected", "Infected 83 vs 50",
  "infection_average_all_diets", "Overall infection", "Average infected vs control",
  "interaction_diet33_vs50", "Diet by infection interaction", "Infection response: 33 vs 50",
  "interaction_diet83_vs50", "Diet by infection interaction", "Infection response: 83 vs 50",
  "interaction_diet83_vs33", "Diet by infection interaction", "Infection response: 83 vs 33"
)

deg_raw <- read.csv(scenario_results_file, check.names = FALSE) |>
  dplyr::filter(
    mapping_branch == "Competitive host",
    analysis_scope == "Placed/host-assigned",
    gene_id %in% gene_universe,
    significant,
    !is.na(gene_id),
    !is.na(padj),
    padj < params$alpha,
    abs(log2FoldChange) >= params$lfc_threshold
  ) |>
  dplyr::inner_join(contrast_map, by = "contrast_id") |>
  # The searchable table uses the exact current GFF annotation joined below.
  dplyr::select(-dplyr::any_of(c("description", "gene_biotype")))

deg_catalogue <- deg_raw |>
  dplyr::mutate(
    numerator_condition = numerator,
    denominator_condition = denominator,
    higher_in = dplyr::if_else(log2FoldChange > 0,
                               numerator_condition, denominator_condition),
    lower_in = dplyr::if_else(log2FoldChange > 0,
                              denominator_condition, numerator_condition),
    signed_direction = dplyr::if_else(log2FoldChange > 0,
                                      "Upregulated", "Downregulated")
  ) |>
  dplyr::left_join(annotation, by = "gene_id") |>
  dplyr::mutate(
    Description = dplyr::coalesce(Description, "no GFF description")
  ) |>
  dplyr::select(
    gene_id, contrast_id, Description,
    gene_biotype,
    comparison_type, comparison, numerator_condition, denominator_condition,
    higher_in, lower_in, signed_direction, log2FoldChange, lfcSE, baseMean,
    stat, pvalue, padj, GOs, KEGG_ko, KEGG_Pathway, PFAMs
  ) |>
  dplyr::arrange(comparison_type, comparison, padj,
                 dplyr::desc(abs(log2FoldChange)))

write.csv(deg_catalogue, file.path(out_dir, "all_samples_deg_catalogue.csv"),
          row.names = FALSE)

catalogue_summary <- deg_catalogue |>
  dplyr::count(comparison_type, comparison, signed_direction,
               name = "n_DEGs") |>
  dplyr::arrange(comparison_type, comparison, signed_direction)

comparison_type_order <- c("Diet contrasts among control",
                           "Diet contrasts among infected",
                           "Infection within diet",
                           "Overall infection",
                           "Diet by infection interaction")
comparison_order <- catalogue_summary |>
  dplyr::distinct(comparison_type, comparison) |>
  dplyr::arrange(factor(comparison_type, levels = comparison_type_order),
                 comparison) |>
  dplyr::pull(comparison)

catalogue_summary <- catalogue_summary |>
  dplyr::mutate(
    comparison_type = factor(comparison_type,
                             levels = comparison_type_order),
    comparison = factor(comparison, levels = comparison_order),
    signed_direction = factor(signed_direction,
                              levels = c("Downregulated", "Upregulated"))
  )

write.csv(catalogue_summary,
          file.path(out_dir, "all_samples_deg_catalogue_summary.csv"),
          row.names = FALSE)

knitr::kable(
  catalogue_summary,
  caption = "Significant DEG counts in the all-samples retained catalogue"
)
```

```{r deg-summary-plot, fig.width=12, fig.height=7}
p_summary <- ggplot2::ggplot(
  catalogue_summary,
  ggplot2::aes(x = comparison, y = n_DEGs, fill = signed_direction)
) +
  ggplot2::geom_col(position = "dodge", width = 0.75) +
  ggplot2::facet_wrap(~ comparison_type, scales = "free_x") +
  ggplot2::scale_fill_manual(values = direction_colors) +
  ggplot2::labs(
    title = "All-samples retained DEG catalogue",
    subtitle = "Counts use padj < alpha and absolute log2FC >= lfc_threshold",
    x = NULL,
    y = "Number of significant DEGs",
    fill = NULL
  ) +
  ggplot2::theme_classic(base_size = 12) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 35, hjust = 1),
    legend.position = "bottom",
    plot.title = ggplot2::element_text(face = "bold")
  )

save_png_svg(p_summary, "all_samples_deg_catalogue_summary",
             width = 12, height = 7, dpi = 300)
p_summary
```

## DEG context zooms and expression-cluster proportions

The full DEG catalogue plot is useful for comparing all contrasts at once, but
the large infected-diet contrast compresses the smaller contexts. The exported
zoom panels below repeat the same counts one context at a time, with an
independent y-axis. The proportion plots then ask which expression cluster
contains the genes from each DEG contrast.

```{r deg-summary-context-exports, fig.width=13, fig.height=8}
make_context_count_plot <- function(context_label) {
  context_df <- catalogue_summary |>
    dplyr::filter(comparison_type == context_label)

  ggplot2::ggplot(
    context_df,
    ggplot2::aes(x = comparison, y = n_DEGs, fill = signed_direction)
  ) +
    ggplot2::geom_col(position = "dodge", width = 0.75) +
    ggplot2::geom_text(
      ggplot2::aes(label = n_DEGs),
      position = ggplot2::position_dodge(width = 0.75),
      vjust = -0.3,
      size = 3.8,
      fontface = "bold"
    ) +
    ggplot2::scale_fill_manual(values = direction_colors, drop = FALSE) +
    ggplot2::scale_y_continuous(
      expand = ggplot2::expansion(mult = c(0, 0.12))
    ) +
    ggplot2::labs(
      title = "All-samples retained DEG catalogue",
      subtitle = paste0(as.character(context_label),
                        " - independent y-axis zoom"),
      x = NULL,
      y = "Number of significant DEGs",
      fill = NULL
    ) +
    ggplot2::theme_classic(base_size = 13) +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 35, hjust = 1),
      legend.position = "bottom",
      plot.title = ggplot2::element_text(face = "bold")
    )
}

context_count_plots <- lapply(levels(catalogue_summary$comparison_type),
                              make_context_count_plot)
names(context_count_plots) <- levels(catalogue_summary$comparison_type)

for (context_label in names(context_count_plots)) {
  n_comparisons <- dplyr::n_distinct(
    catalogue_summary$comparison[catalogue_summary$comparison_type == context_label]
  )
  save_png_svg(
    context_count_plots[[context_label]],
    paste0("all_samples_deg_catalogue_summary_zoom_",
           slugify(context_label)),
    width = max(6.5, 2.3 * n_comparisons),
    height = 5.2,
    dpi = 300
  )
}

cluster_assignment_dir <- latest_dir(params$cluster_results_pattern)
cluster_assignment_file <- if (is.na(cluster_assignment_dir)) {
  NA_character_
} else {
  file.path(cluster_assignment_dir, "full_heatmap_expression_cluster_assignments.csv")
}

cluster_export_note <- data.frame(
  input = "Expression-cluster assignments",
  path = cluster_assignment_file,
  present = file.exists(cluster_assignment_file)
)
knitr::kable(cluster_export_note,
             caption = "Cluster assignment file used for proportion plots")

if (file.exists(cluster_assignment_file)) {
  cluster_levels <- names(cluster_colors)
  cluster_assignments <- read.csv(cluster_assignment_file,
                                  check.names = FALSE) |>
    dplyr::transmute(
      gene_id,
      expression_cluster = factor(expression_cluster,
                                  levels = cluster_levels)
    ) |>
    dplyr::filter(!is.na(expression_cluster)) |>
    dplyr::distinct()

  deg_cluster_counts <- deg_catalogue |>
    dplyr::inner_join(cluster_assignments, by = "gene_id") |>
    dplyr::mutate(
      comparison_type = factor(comparison_type,
                               levels = comparison_type_order),
      comparison = factor(comparison, levels = comparison_order),
      signed_direction = factor(signed_direction,
                                levels = c("Downregulated", "Upregulated"))
    ) |>
    dplyr::distinct(comparison_type, comparison, signed_direction,
                    gene_id, expression_cluster) |>
    dplyr::count(comparison_type, comparison, signed_direction,
                 expression_cluster, name = "n_genes") |>
    dplyr::group_by(comparison_type, comparison, signed_direction) |>
    tidyr::complete(
      expression_cluster = factor(cluster_levels, levels = cluster_levels),
      fill = list(n_genes = 0)
    ) |>
    dplyr::mutate(
      total_DEGs = sum(n_genes),
      proportion = dplyr::if_else(total_DEGs > 0, n_genes / total_DEGs, 0)
    ) |>
    dplyr::ungroup() |>
    dplyr::arrange(comparison_type, comparison, signed_direction,
                   expression_cluster)

  write.csv(
    deg_cluster_counts,
    file.path(out_dir, "all_samples_deg_catalogue_cluster_proportions.csv"),
    row.names = FALSE
  )

  p_cluster_proportion <- ggplot2::ggplot(
    deg_cluster_counts,
    ggplot2::aes(x = comparison, y = proportion, fill = expression_cluster)
  ) +
    ggplot2::geom_col(width = 0.78, color = "white", linewidth = 0.4) +
    ggplot2::facet_grid(
      signed_direction ~ comparison_type,
      scales = "free_x",
      space = "free_x"
    ) +
    ggplot2::scale_fill_manual(values = cluster_colors, drop = FALSE) +
    ggplot2::scale_y_continuous(
      labels = scales::percent_format(accuracy = 1),
      limits = c(0, 1),
      expand = c(0, 0)
    ) +
    ggplot2::labs(
      title = "All-samples DEG catalogue by expression cluster",
      subtitle = "Each bar sums to 100% of significant DEGs for one contrast and direction",
      x = NULL,
      y = "Proportion of significant DEGs",
      fill = "Expression cluster"
    ) +
    ggplot2::theme_classic(base_size = 13) +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 35, hjust = 1),
      strip.background = ggplot2::element_rect(fill = "#F3F4F6",
                                               color = "#111827"),
      strip.text = ggplot2::element_text(face = "bold"),
      legend.position = "bottom",
      plot.title = ggplot2::element_text(face = "bold")
    )

  save_png_svg(
    p_cluster_proportion,
    "all_samples_deg_catalogue_cluster_proportions",
    width = 13,
    height = 8,
    dpi = 300
  )

  make_context_cluster_plot <- function(context_label) {
    context_df <- deg_cluster_counts |>
      dplyr::filter(comparison_type == context_label)

    ggplot2::ggplot(
      context_df,
      ggplot2::aes(x = comparison, y = proportion, fill = expression_cluster)
    ) +
      ggplot2::geom_col(width = 0.78, color = "white", linewidth = 0.4) +
      ggplot2::facet_wrap(~ signed_direction, nrow = 1) +
      ggplot2::scale_fill_manual(values = cluster_colors, drop = FALSE) +
      ggplot2::scale_y_continuous(
        labels = scales::percent_format(accuracy = 1),
        limits = c(0, 1),
        expand = c(0, 0)
      ) +
      ggplot2::labs(
        title = "Expression-cluster proportions by DEG context",
        subtitle = paste0(as.character(context_label),
                          " - independent context zoom"),
        x = NULL,
        y = "Proportion of significant DEGs",
        fill = "Expression cluster"
      ) +
      ggplot2::theme_classic(base_size = 13) +
      ggplot2::theme(
        axis.text.x = ggplot2::element_text(angle = 35, hjust = 1),
        strip.background = ggplot2::element_rect(fill = "#F3F4F6",
                                                 color = "#111827"),
        strip.text = ggplot2::element_text(face = "bold"),
        legend.position = "bottom",
        plot.title = ggplot2::element_text(face = "bold")
      )
  }

  context_cluster_plots <- lapply(levels(catalogue_summary$comparison_type),
                                  make_context_cluster_plot)
  names(context_cluster_plots) <- levels(catalogue_summary$comparison_type)

  for (context_label in names(context_cluster_plots)) {
    n_comparisons <- dplyr::n_distinct(
      catalogue_summary$comparison[catalogue_summary$comparison_type == context_label]
    )
    save_png_svg(
      context_cluster_plots[[context_label]],
      paste0("all_samples_deg_catalogue_cluster_proportions_zoom_",
             slugify(context_label)),
      width = max(7, 2.5 * n_comparisons),
      height = 5.4,
      dpi = 300
    )
  }

  knitr::kable(
    deg_cluster_counts,
    caption = "Significant DEG proportions assigned to expression clusters"
  )
  p_cluster_proportion
} else {
  message("Cluster proportion plots skipped because the cluster assignment file was not found.")
}
```

## High-effect, high-expression candidates {.tabset}

DEG totals describe the breadth of a response, but they do not identify genes
that combine a large change with abundant expression. This exploratory ranking
therefore adds condition-specific expression to the DESeq2 results. It is not a
second significance test and does not replace the complete catalogue below.

A high-priority candidate must meet the study-wide DEG thresholds
(`padj < 0.05` and `|log2FC| >= 1`), show at least a fourfold difference
(`|log2FC| >= 2`), and have a mean of at least 100 size-factor-normalized
fragments in the higher-expression condition. Within each contrast, candidates
are then ranked jointly by effect size, expression, and adjusted significance.
The thresholds are deliberately shown here so they can be changed without
altering the DESeq2 models.

```{r candidate-ranking}
metadata_candidates <- readr::read_tsv(metadata_file, show_col_types = FALSE) |>
  dplyr::transmute(
    sample_id = as.character(.data$label),
    diet = as.character(.data$Diet),
    treatment = as.character(.data$Treatment)
  )

raw_candidate_counts <- read_count_matrix(primary_counts_file)
size_factors <- readr::read_tsv(size_factor_file, show_col_types = FALSE) |>
  dplyr::transmute(
    sample_id = as.character(.data$sample_id),
    size_factor = as.numeric(.data$placed_gene_size_factor)
  )

if (!setequal(metadata_candidates$sample_id, colnames(raw_candidate_counts)) ||
    !setequal(metadata_candidates$sample_id, size_factors$sample_id)) {
  stop("Candidate-ranking counts, metadata, and size factors have different samples.")
}

sample_order <- metadata_candidates$sample_id
raw_candidate_counts <- raw_candidate_counts[, sample_order, drop = FALSE]
sample_size_factors <- size_factors$size_factor[
  match(sample_order, size_factors$sample_id)
]
normalized_candidate_counts <- sweep(
  raw_candidate_counts,
  2,
  sample_size_factors,
  "/"
)

samples_for <- function(treatment = NULL, diet = NULL) {
  keep <- rep(TRUE, nrow(metadata_candidates))
  if (!is.null(treatment)) keep <- keep & metadata_candidates$treatment == treatment
  if (!is.null(diet)) keep <- keep & metadata_candidates$diet == as.character(diet)
  metadata_candidates$sample_id[keep]
}

contrast_sample_sets <- list(
  infection_diet33 = list(
    numerator = samples_for("Infected", "33"),
    denominator = samples_for("Control", "33")
  ),
  infection_diet50 = list(
    numerator = samples_for("Infected", "50"),
    denominator = samples_for("Control", "50")
  ),
  infection_diet83 = list(
    numerator = samples_for("Infected", "83"),
    denominator = samples_for("Control", "83")
  ),
  control_diet33_vs50 = list(
    numerator = samples_for("Control", "33"),
    denominator = samples_for("Control", "50")
  ),
  control_diet83_vs33 = list(
    numerator = samples_for("Control", "83"),
    denominator = samples_for("Control", "33")
  ),
  control_diet83_vs50 = list(
    numerator = samples_for("Control", "83"),
    denominator = samples_for("Control", "50")
  ),
  infected_diet33_vs50 = list(
    numerator = samples_for("Infected", "33"),
    denominator = samples_for("Infected", "50")
  ),
  infected_diet83_vs33 = list(
    numerator = samples_for("Infected", "83"),
    denominator = samples_for("Infected", "33")
  ),
  infected_diet83_vs50 = list(
    numerator = samples_for("Infected", "83"),
    denominator = samples_for("Infected", "50")
  ),
  infection_average_all_diets = list(
    numerator = samples_for("Infected"),
    denominator = samples_for("Control")
  )
)

condition_mean <- function(gene_id, contrast_id, side) {
  samples <- contrast_sample_sets[[contrast_id]][[side]]
  if (!gene_id %in% rownames(normalized_candidate_counts)) return(NA_real_)
  mean(normalized_candidate_counts[gene_id, samples], na.rm = TRUE)
}

candidate_pool <- deg_catalogue |>
  dplyr::filter(.data$contrast_id %in% names(contrast_sample_sets))

candidate_pool$mean_normalized_numerator <- mapply(
  condition_mean,
  candidate_pool$gene_id,
  candidate_pool$contrast_id,
  MoreArgs = list(side = "numerator")
)
candidate_pool$mean_normalized_denominator <- mapply(
  condition_mean,
  candidate_pool$gene_id,
  candidate_pool$contrast_id,
  MoreArgs = list(side = "denominator")
)

candidate_pool <- candidate_pool |>
  dplyr::mutate(
    mean_normalized_higher_condition = dplyr::if_else(
      .data$log2FoldChange > 0,
      .data$mean_normalized_numerator,
      .data$mean_normalized_denominator
    ),
    large_high_expression =
      abs(.data$log2FoldChange) >= as.numeric(params$large_lfc_threshold) &
      .data$mean_normalized_higher_condition >=
        as.numeric(params$high_expression_threshold)
  ) |>
  dplyr::group_by(.data$contrast_id) |>
  dplyr::mutate(
    effect_percentile = dplyr::percent_rank(abs(.data$log2FoldChange)),
    expression_percentile = dplyr::percent_rank(
      log10(.data$mean_normalized_higher_condition + 1)
    ),
    significance_percentile = dplyr::percent_rank(-log10(.data$padj)),
    priority_score = rowMeans(
      cbind(
        .data$effect_percentile,
        .data$expression_percentile,
        .data$significance_percentile
      ),
      na.rm = TRUE
    )
  ) |>
  dplyr::ungroup()

exon_sensitivity <- read.csv(exon_results_file, check.names = FALSE) |>
  dplyr::filter(
    .data$mapping_branch == "Competitive host",
    .data$analysis_scope == "Placed/host-assigned"
  ) |>
  dplyr::transmute(
    gene_id,
    contrast_id,
    exon_log2FoldChange = log2FoldChange,
    exon_padj = padj,
    exon_significant = significant
  )

candidate_pool <- candidate_pool |>
  dplyr::left_join(exon_sensitivity, by = c("gene_id", "contrast_id")) |>
  dplyr::mutate(
    exon_same_direction = !is.na(.data$exon_log2FoldChange) &
      sign(.data$exon_log2FoldChange) == sign(.data$log2FoldChange),
    exon_concordant_DEG = .data$exon_significant %in% TRUE &
      .data$exon_same_direction
  )

high_priority_candidates <- candidate_pool |>
  dplyr::filter(.data$large_high_expression) |>
  dplyr::arrange(.data$comparison_type, .data$comparison,
                 dplyr::desc(.data$priority_score))

readr::write_csv(
  high_priority_candidates,
  file.path(out_dir, "high_effect_high_expression_candidate_catalogue.csv"),
  na = ""
)

candidate_summary <- candidate_pool |>
  dplyr::group_by(.data$comparison_type, .data$comparison) |>
  dplyr::summarise(
    significant_DEGs = dplyr::n(),
    high_effect_high_expression = sum(.data$large_high_expression),
    exon_concordant_high_priority = sum(
      .data$large_high_expression & .data$exon_concordant_DEG,
      na.rm = TRUE
    ),
    .groups = "drop"
  )
readr::write_csv(
  candidate_summary,
  file.path(out_dir, "high_expression_candidate_summary.csv"),
  na = ""
)

knitr::kable(
  candidate_summary,
  caption = "Large-effect, high-expression candidates by direct contrast"
)
```

### Diet-associated candidates

The plot below ranks genes that differ among diets in control animals or among
infected animals. Point position gives the DESeq2 effect size, while point size
shows mean normalized expression in whichever diet has higher expression. The
same gene can appear in more than one panel when it separates multiple diets.

```{r diet-high-expression-candidates, fig.width=15, fig.height=15}
diet_candidates <- high_priority_candidates |>
  dplyr::filter(
    .data$comparison_type %in%
      c("Diet contrasts among control", "Diet contrasts among infected")
  ) |>
  dplyr::group_by(.data$comparison) |>
  dplyr::slice_max(.data$priority_score, n = 5, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    candidate_label = paste0(.data$Description, " [", .data$gene_id, "]")
  )

readr::write_csv(
  diet_candidates,
  file.path(out_dir, "top_diet_high_expression_candidates.csv"),
  na = ""
)

p_diet_candidates <- ggplot2::ggplot(
  diet_candidates,
  ggplot2::aes(
    x = .data$log2FoldChange,
    y = stats::reorder(.data$candidate_label, .data$log2FoldChange),
    size = .data$mean_normalized_higher_condition,
    color = .data$log2FoldChange
  )
) +
  ggplot2::geom_vline(xintercept = 0, color = "#6B7280", linewidth = 0.45) +
  ggplot2::geom_segment(
    ggplot2::aes(x = 0, xend = .data$log2FoldChange, yend = .data$candidate_label),
    color = "#B8B8B8",
    linewidth = 0.55
  ) +
  ggplot2::geom_point(alpha = 0.9) +
  ggplot2::facet_wrap(~ comparison, scales = "free_y", ncol = 2) +
  ggplot2::scale_color_gradient2(
    low = "#2166AC", mid = "#F7F7F7", high = "#B2182B", midpoint = 0
  ) +
  ggplot2::scale_size_continuous(
    trans = "log10",
    range = c(3.5, 10),
    labels = scales::label_comma()
  ) +
  ggplot2::labs(
    title = "Large and abundant diet-associated expression differences",
    subtitle = "Top five ranked candidates per diet contrast; main chromosomes only",
    x = "log2 fold change: numerator relative to denominator",
    y = NULL,
    color = "log2 fold change",
    size = "Mean normalized fragments\nin higher condition"
  ) +
  ggplot2::theme_bw(base_size = 13) +
  ggplot2::theme(
    strip.text = ggplot2::element_text(face = "bold", size = 12),
    axis.text.y = ggplot2::element_text(size = 10),
    legend.position = "bottom",
    plot.title = ggplot2::element_text(face = "bold")
  )

save_png_svg(
  p_diet_candidates,
  "top_diet_high_expression_candidates",
  width = 15,
  height = 15,
  dpi = 300
)
p_diet_candidates
```

The strongest annotated diet candidates include
`r paste(unique(diet_candidates$Description[diet_candidates$Description != "no GFF description"])[1:5], collapse = "; ")`.
These are candidates for follow-up, not evidence that diet directly regulates
the annotated biological process.

### Infection-induced candidates

For infection, the clearest candidates are genes that are both induced and
abundant in infected fat body. The matrix below shows the strongest candidates
that met those criteria across all three diets. Color gives the infection log2
fold change and point size is the mean normalized expression among infected
samples of that diet.

```{r infection-high-expression-candidates, fig.width=16, fig.height=12}
infection_gene_summary <- candidate_pool |>
  dplyr::filter(
    .data$contrast_id %in% c("infection_diet33", "infection_diet50", "infection_diet83"),
    .data$log2FoldChange > 0,
    .data$large_high_expression
  ) |>
  dplyr::group_by(.data$gene_id, .data$Description, .data$gene_biotype) |>
  dplyr::summarise(
    diets_with_high_priority_induction = dplyr::n_distinct(.data$contrast_id),
    maximum_priority_score = max(.data$priority_score, na.rm = TRUE),
    maximum_infected_mean = max(.data$mean_normalized_numerator, na.rm = TRUE),
    minimum_infection_log2FC = min(.data$log2FoldChange, na.rm = TRUE),
    exon_concordant_diets = sum(.data$exon_concordant_DEG, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::arrange(
    dplyr::desc(.data$diets_with_high_priority_induction),
    dplyr::desc(.data$maximum_priority_score),
    dplyr::desc(.data$maximum_infected_mean)
  ) |>
  dplyr::slice_head(n = 20) |>
  dplyr::mutate(
    candidate_label = paste0(.data$Description, " [", .data$gene_id, "]")
  )

all_primary_results <- read.csv(scenario_results_file, check.names = FALSE) |>
  dplyr::filter(
    .data$mapping_branch == "Competitive host",
    .data$analysis_scope == "Placed/host-assigned",
    .data$contrast_id %in% c("infection_diet33", "infection_diet50", "infection_diet83"),
    .data$gene_id %in% infection_gene_summary$gene_id
  ) |>
  dplyr::left_join(
    infection_gene_summary |>
      dplyr::select(.data$gene_id, .data$candidate_label),
    by = "gene_id"
  ) |>
  dplyr::mutate(
    diet = stringr::str_remove(.data$contrast_id, "infection_diet"),
    significant_DEG = !is.na(.data$padj) &
      .data$padj < as.numeric(params$alpha) &
      abs(.data$log2FoldChange) >= as.numeric(params$lfc_threshold)
  )

all_primary_results$infected_mean <- mapply(
  condition_mean,
  all_primary_results$gene_id,
  all_primary_results$contrast_id,
  MoreArgs = list(side = "numerator")
)

infection_label_order <- rev(infection_gene_summary$candidate_label)
all_primary_results <- all_primary_results |>
  dplyr::mutate(
    candidate_label = factor(.data$candidate_label, levels = infection_label_order),
    diet = factor(.data$diet, levels = c("33", "50", "83"))
  )

readr::write_csv(
  infection_gene_summary,
  file.path(out_dir, "top_infection_induced_high_expression_candidates.csv"),
  na = ""
)

p_infection_candidates <- ggplot2::ggplot(
  all_primary_results,
  ggplot2::aes(
    x = .data$diet,
    y = .data$candidate_label,
    color = .data$log2FoldChange,
    size = .data$infected_mean
  )
) +
  ggplot2::geom_point(alpha = 0.92) +
  ggplot2::scale_color_gradient2(
    low = "#2166AC", mid = "#F7F7F7", high = "#B2182B", midpoint = 0
  ) +
  ggplot2::scale_size_continuous(
    trans = "log10",
    range = c(3.5, 11),
    labels = scales::label_comma()
  ) +
  ggplot2::labs(
    title = "Highly induced and abundant infection-response candidates",
    subtitle = "Candidates selected from placed-chromosome DEGs; all diet-specific estimates are shown",
    x = "Diet",
    y = NULL,
    color = "Infected vs control\nlog2 fold change",
    size = "Mean normalized fragments\nin infected samples"
  ) +
  ggplot2::guides(
    size = ggplot2::guide_legend(order = 1, nrow = 1),
    color = ggplot2::guide_colorbar(
      order = 2,
      title.position = "top",
      barwidth = grid::unit(7, "cm")
    )
  ) +
  ggplot2::theme_bw(base_size = 13) +
  ggplot2::theme(
    axis.text.y = ggplot2::element_text(size = 10.5),
    legend.position = "bottom",
    legend.box = "vertical",
    plot.title = ggplot2::element_text(face = "bold")
  )

save_png_svg(
  p_infection_candidates,
  "top_infection_induced_high_expression_candidates",
  width = 16,
  height = 12,
  dpi = 300
)
p_infection_candidates
```

Among the strongest annotated infection-induced candidates are
`r paste(unique(infection_gene_summary$Description[infection_gene_summary$Description != "no GFF description"])[1:7], collapse = "; ")`.
Genes appearing across all three diets provide the most reproducible candidates
for a general infection response, whereas genes restricted to one diet are
better treated as possible diet-dependent responses.

### Candidate volcano plots

These volcano plots place the prioritized genes back into the complete tested
gene universe. Gray points were not called as DEGs; blue and red points are the
ordinary significant DEGs, and gold points are the large-effect,
high-expression candidates defined above. Solid vertical lines show the
study-wide `|log2FC| = 1` DEG threshold, while dotted lines show the more
stringent `|log2FC| = 2` candidate threshold. Labels are limited to the
highest-ranked candidates in each contrast, but every prioritized candidate is
highlighted and available in the searchable table.

```{r candidate-volcano-data}
volcano_contrasts <- names(contrast_sample_sets)

volcano_data <- read.csv(scenario_results_file, check.names = FALSE) |>
  dplyr::filter(
    .data$mapping_branch == "Competitive host",
    .data$analysis_scope == "Placed/host-assigned",
    .data$contrast_id %in% volcano_contrasts,
    !is.na(.data$padj),
    !is.na(.data$log2FoldChange)
  ) |>
  dplyr::inner_join(contrast_map, by = "contrast_id") |>
  dplyr::left_join(
    annotation |>
      dplyr::select(.data$gene_id, .data$Description, .data$gene_biotype),
    by = "gene_id"
  ) |>
  dplyr::left_join(
    candidate_pool |>
      dplyr::select(
        .data$gene_id, .data$contrast_id,
        .data$large_high_expression, .data$priority_score,
        .data$mean_normalized_higher_condition,
        .data$exon_concordant_DEG
      ),
    by = c("gene_id", "contrast_id")
  ) |>
  dplyr::mutate(
    Description = dplyr::coalesce(.data$Description, "no GFF description"),
    large_high_expression = dplyr::coalesce(
      .data$large_high_expression,
      FALSE
    ),
    minus_log10_padj = -log10(pmax(.data$padj, .Machine$double.xmin)),
    DEG_status = dplyr::case_when(
      .data$large_high_expression ~ "High-effect, high-expression",
      .data$significant & .data$log2FoldChange > 0 ~ "Significant: higher in numerator",
      .data$significant & .data$log2FoldChange < 0 ~ "Significant: higher in denominator",
      TRUE ~ "Not significant"
    ),
    volcano_label = dplyr::if_else(
      stringr::str_detect(.data$Description, "^(uncharacterized|no GFF)"),
      .data$gene_id,
      paste0(.data$Description, " [", .data$gene_id, "]")
    ) |>
      stringr::str_wrap(width = 38)
  )

volcano_colors <- c(
  "Not significant" = "#C9CDD2",
  "Significant: higher in denominator" = "#3B82C4",
  "Significant: higher in numerator" = "#D8584C",
  "High-effect, high-expression" = "#F2B134"
)

make_volcano_plot <- function(plot_data, label_data, title, ncol) {
  ggplot2::ggplot(
    plot_data,
    ggplot2::aes(x = .data$log2FoldChange, y = .data$minus_log10_padj)
  ) +
    ggplot2::geom_hline(
      yintercept = -log10(as.numeric(params$alpha)),
      linetype = "dashed",
      color = "#5F6368",
      linewidth = 0.45
    ) +
    ggplot2::geom_vline(
      xintercept = c(-as.numeric(params$lfc_threshold),
                     as.numeric(params$lfc_threshold)),
      color = "#5F6368",
      linewidth = 0.45
    ) +
    ggplot2::geom_vline(
      xintercept = c(-as.numeric(params$large_lfc_threshold),
                     as.numeric(params$large_lfc_threshold)),
      linetype = "dotted",
      color = "#B7791F",
      linewidth = 0.65
    ) +
    ggplot2::geom_point(
      ggplot2::aes(color = .data$DEG_status),
      size = 1.05,
      alpha = 0.58
    ) +
    ggplot2::geom_point(
      data = plot_data |>
        dplyr::filter(.data$large_high_expression),
      shape = 21,
      fill = "#F2B134",
      color = "#202124",
      stroke = 0.45,
      size = 2.4,
      alpha = 0.92
    ) +
    ggrepel::geom_label_repel(
      data = label_data,
      ggplot2::aes(label = .data$volcano_label),
      size = 3.3,
      label.size = 0.18,
      fill = scales::alpha("white", 0.9),
      color = "#202124",
      min.segment.length = 0,
      box.padding = 0.35,
      point.padding = 0.2,
      max.overlaps = Inf,
      seed = 19,
      show.legend = FALSE
    ) +
    ggplot2::facet_wrap(~ comparison, ncol = ncol, scales = "free_y") +
    ggplot2::scale_color_manual(values = volcano_colors, drop = FALSE) +
    ggplot2::labs(
      title = title,
      subtitle = "Gold points combine at least fourfold change with abundant normalized expression",
      x = "log2 fold change: numerator relative to denominator",
      y = expression(-log[10](adjusted~italic(P))),
      color = NULL
    ) +
    ggplot2::theme_bw(base_size = 13) +
    ggplot2::theme(
      strip.text = ggplot2::element_text(face = "bold", size = 12),
      legend.position = "bottom",
      plot.title = ggplot2::element_text(face = "bold"),
      panel.grid.minor = ggplot2::element_blank()
    )
}
```

```{r infection-candidate-volcano, fig.width=18, fig.height=8}
infection_volcano_data <- volcano_data |>
  dplyr::filter(
    .data$contrast_id %in% c("infection_diet33", "infection_diet50", "infection_diet83")
  )

infection_volcano_labels <- infection_volcano_data |>
  dplyr::filter(.data$large_high_expression) |>
  dplyr::group_by(.data$contrast_id) |>
  dplyr::slice_max(.data$priority_score, n = 7, with_ties = FALSE) |>
  dplyr::ungroup()

p_infection_volcano <- make_volcano_plot(
  infection_volcano_data,
  infection_volcano_labels,
  "Infection-response volcano plots with prioritized candidates",
  ncol = 3
)

save_png_svg(
  p_infection_volcano,
  "infection_high_expression_candidate_volcanoes",
  width = 18,
  height = 8,
  dpi = 300
)
p_infection_volcano
```

```{r diet-candidate-volcano, fig.width=18, fig.height=14}
diet_volcano_data <- volcano_data |>
  dplyr::filter(
    .data$comparison_type %in%
      c("Diet contrasts among control", "Diet contrasts among infected")
  )

diet_volcano_labels <- diet_volcano_data |>
  dplyr::filter(.data$large_high_expression) |>
  dplyr::group_by(.data$contrast_id) |>
  dplyr::slice_max(.data$priority_score, n = 5, with_ties = FALSE) |>
  dplyr::ungroup()

p_diet_volcano <- make_volcano_plot(
  diet_volcano_data,
  diet_volcano_labels,
  "Diet-comparison volcano plots with prioritized candidates",
  ncol = 2
)

save_png_svg(
  p_diet_volcano,
  "diet_high_expression_candidate_volcanoes",
  width = 18,
  height = 14,
  dpi = 300
)
p_diet_volcano
```

### Searchable candidate table

```{r searchable-high-expression-candidates}
candidate_display <- high_priority_candidates |>
  dplyr::mutate(
    log2FoldChange = round(.data$log2FoldChange, 3),
    padj = signif(.data$padj, 3),
    mean_normalized_numerator = round(.data$mean_normalized_numerator, 1),
    mean_normalized_denominator = round(.data$mean_normalized_denominator, 1),
    mean_normalized_higher_condition = round(
      .data$mean_normalized_higher_condition,
      1
    ),
    exon_log2FoldChange = round(.data$exon_log2FoldChange, 3),
    exon_padj = signif(.data$exon_padj, 3)
  ) |>
  dplyr::select(
    .data$gene_id, .data$Description, .data$gene_biotype,
    .data$comparison_type, .data$comparison, .data$higher_in,
    .data$log2FoldChange, .data$padj,
    .data$mean_normalized_numerator, .data$mean_normalized_denominator,
    .data$mean_normalized_higher_condition,
    .data$exon_log2FoldChange, .data$exon_padj, .data$exon_concordant_DEG
  )

DT::datatable(
  candidate_display,
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    pageLength = 20,
    scrollX = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv")
  )
) |>
  DT::formatStyle(
    "log2FoldChange",
    color = DT::styleInterval(0, c("#1E88E5", "#E41A1C")),
    fontWeight = "bold"
  )
```

## Searchable DEG catalogue

Use the search boxes to find a gene ID, annotation term, contrast, or condition.
The `higher_in` column is the plain-language interpretation of the log2 fold
change sign.

```{r searchable-deg-table}
display_table <- deg_catalogue |>
  dplyr::mutate(
    log2FoldChange = round(log2FoldChange, 3),
    lfcSE = round(lfcSE, 3),
    baseMean = round(baseMean, 2),
    stat = round(stat, 3),
    pvalue = signif(pvalue, 3),
    padj = signif(padj, 3)
  ) |>
  dplyr::select(
    gene_id, Description, gene_biotype, comparison_type, comparison,
    higher_in, lower_in, signed_direction, log2FoldChange, padj, baseMean,
    numerator_condition, denominator_condition, GOs, KEGG_ko, KEGG_Pathway,
    PFAMs
  )

DT::datatable(
  display_table,
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv")
  )
) |>
  DT::formatStyle(
    "log2FoldChange",
    color = DT::styleInterval(0, c("#1E88E5", "#E41A1C")),
    fontWeight = "bold"
  ) |>
  DT::formatStyle(
    "signed_direction",
    color = DT::styleEqual(names(direction_colors), direction_colors),
    fontWeight = "bold"
  )
```

## Exported files

```{r exported-files}
exports <- list.files(out_dir, full.names = FALSE)
knitr::kable(tibble::tibble(file = exports),
             caption = paste("Files written to", out_dir))
```

## Session information

```{r session-info}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
