---
title: "Main-chromosome GO and KEGG term-gene index"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  counts_file: "output/rmd_runs/transcript-exon-primary-corrected45_20260809-153146/host_counts_no_rrna.tsv"
  eggnog_file: "data/GO_Annotations/GCF_023897955.1_iqSchGreg1.2_Arthopoda_one2one.emapper.annotations"
  gff_file: "data/reference/GCF_023897955.1_iqSchGreg1.2_genomic.gff"
  latest_results_root: "output/rmd_runs"
  enrichment_results_pattern: "^go-term-enrichment_"
  output_tag: "enrichment-term-gene-index"
---

## Purpose

::: {.walkthrough-grid}
::: {.walkthrough-step}
**Step 1 - Read enrichment**

Load the current chromosome-only GO and KEGG result tables.
:::
::: {.walkthrough-step}
**Step 2 - Reverse the mapping**

Expand every enriched term into the DEG IDs that contributed to it.
:::
::: {.walkthrough-step}
**Step 3 - Inspect genes**

Search terms and descriptions to move from a pathway back to individual genes.
:::
:::

This page reverses the GO/KEGG enrichment output. Instead of starting from a
gene and asking which annotations it has, it starts from each enriched GO term
or KEGG pathway and lists the DEG IDs that contributed to that enrichment.

The gene lists come directly from the `geneID` column written by
`clusterProfiler::enricher()` on the GO/KEGG enrichment page. Those are the
foreground genes that overlap each enriched term.

Because the source enrichment page now restricts both foregrounds and
backgrounds to official nuclear-chromosome genes, every gene listed here follows
the same primary paper inclusion rule.

Gene descriptions are taken from eggNOG when available and otherwise fall back
to the RefSeq/GFF gene description from the same *S. gregaria* assembly.

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c("dplyr", "tidyr", "stringr", "DT", "knitr")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

counts_file <- file.path(project_dir, params$counts_file)
eggnog_file <- file.path(project_dir, params$eggnog_file)
gff_file <- file.path(project_dir, params$gff_file)
results_root <- file.path(project_dir, params$latest_results_root)
```

## Helper functions

```{r helpers}
latest_dir <- function(pattern) {
  dirs <- list.dirs(results_root, recursive = FALSE, full.names = TRUE)
  dirs <- dirs[grepl(pattern, basename(dirs))]
  if (length(dirs) == 0) {
    stop("No output folder matched pattern: ", pattern)
  }
  dirs[which.max(file.info(dirs)$mtime)]
}

read_optional_csv <- function(path) {
  if (!file.exists(path)) {
    return(data.frame())
  }
  read.csv(path, check.names = FALSE)
}

read_emapper <- function(path) {
  cols <- c(
    "query", "seed_ortholog", "evalue", "score", "eggNOG_OGs",
    "max_annot_lvl", "COG_category", "Description", "Preferred_name",
    "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
    "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC", "CAZy",
    "BiGG_Reaction", "PFAMs"
  )
  tab <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", fill = TRUE)
  names(tab) <- cols[seq_len(ncol(tab))]
  tab |>
    dplyr::mutate(protein_id = stringr::str_extract(query, "XP_[0-9]+[.][0-9]+"))
}

parse_cds_gene_bridge <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  gff |>
    dplyr::filter(type == "CDS") |>
    dplyr::transmute(
      protein_id = stringr::str_match(attributes, "protein_id=([^;]+)")[, 2],
      gene_id = stringr::str_match(attributes, "gene=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(protein_id), !is.na(gene_id)) |>
    dplyr::distinct()
}

parse_gff_gene_annotation <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  gff |>
    dplyr::filter(type == "gene") |>
    dplyr::transmute(
      gene_id = dplyr::coalesce(
        stringr::str_match(attributes, "gene=([^;]+)")[, 2],
        stringr::str_match(attributes, "Name=([^;]+)")[, 2]
      ),
      gff_description = stringr::str_match(attributes, "description=([^;]+)")[, 2],
      gene_biotype = stringr::str_match(attributes, "gene_biotype=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(gene_id)) |>
    dplyr::mutate(
      gff_description = dplyr::if_else(
        is.na(gff_description), NA_character_, utils::URLdecode(gff_description)
      )
    ) |>
    dplyr::distinct(gene_id, .keep_all = TRUE)
}

first_nonempty <- function(x, fallback = "unannotated") {
  x <- x[!is.na(x) & x != "" & x != "-"]
  if (length(x) == 0) fallback else x[1]
}

expand_enrichment_genes <- function(tab, enrichment_context) {
  if (nrow(tab) == 0 || !"geneID" %in% names(tab)) {
    return(data.frame())
  }

  # Non-directional and direction-aware enrichment tables carry different
  # metadata columns; add missing columns so the same code can read both.
  optional_cols <- c("family", "comparison", "direction", "plot_label",
                     "pathway_theme", "pathway_scope")
  for (nm in optional_cols) {
    if (!nm %in% names(tab)) {
      tab[[nm]] <- NA_character_
    }
  }

  tab |>
    dplyr::mutate(
      enrichment_context = enrichment_context,
      geneID = as.character(geneID)
    ) |>
    tidyr::separate_rows(geneID, sep = "/") |>
    dplyr::rename(gene_id = geneID) |>
    dplyr::filter(!is.na(gene_id), gene_id != "") |>
    dplyr::transmute(
      enrichment_context,
      database,
      DEG_set = set,
      family = as.character(family),
      comparison = as.character(comparison),
      direction = as.character(direction),
      plot_label = as.character(plot_label),
      term_id = ID,
      term_description = Description,
      gene_id,
      GeneRatio,
      BgRatio,
      FoldEnrichment,
      pvalue,
      p.adjust,
      qvalue,
      Count,
      pathway_theme = as.character(pathway_theme),
      pathway_scope = as.character(pathway_scope)
    )
}
```

## Input files

```{r input-files}
go_dir <- latest_dir(params$enrichment_results_pattern)

input_files <- data.frame(
  input = c(
    "GO/KEGG enrichment output folder",
    "Non-directional enrichment results",
    "Direction-aware enrichment results",
    "GO term names",
    "KEGG pathway names",
    "eggNOG annotation",
    "GFF bridge and gene descriptions"
  ),
  path = c(
    go_dir,
    file.path(go_dir, "all_enrichment_results.csv"),
    file.path(go_dir, "requested_directional_all_enrichment.csv"),
    file.path(go_dir, "eggNOG_GO_TERM2NAME.csv"),
    file.path(go_dir, "eggNOG_KEGG_pathway_TERM2NAME.csv"),
    eggnog_file,
    gff_file
  ),
  present = c(
    dir.exists(go_dir),
    file.exists(file.path(go_dir, "all_enrichment_results.csv")),
    file.exists(file.path(go_dir, "requested_directional_all_enrichment.csv")),
    file.exists(file.path(go_dir, "eggNOG_GO_TERM2NAME.csv")),
    file.exists(file.path(go_dir, "eggNOG_KEGG_pathway_TERM2NAME.csv")),
    file.exists(eggnog_file),
    file.exists(gff_file)
  )
)

knitr::kable(input_files, caption = "Files used for the enriched term gene index")
if (!all(input_files$present)) {
  stop("One or more required inputs are missing. Build page 10 first.")
}
```

## Build term-to-gene index

```{r build-index}
counts_raw <- read.delim(counts_file, check.names = FALSE)
gene_universe <- unique(counts_raw[[1]])

gff_annotation <- parse_gff_gene_annotation(gff_file) |>
  dplyr::filter(gene_id %in% gene_universe)

eggnog_annotation <- read_emapper(eggnog_file) |>
  dplyr::left_join(parse_cds_gene_bridge(gff_file), by = "protein_id") |>
  dplyr::filter(!is.na(gene_id), gene_id %in% gene_universe) |>
  dplyr::group_by(gene_id) |>
  dplyr::summarise(
    Preferred_name = first_nonempty(Preferred_name),
    gene_description = first_nonempty(Description, "no description"),
    GOs = first_nonempty(GOs, ""),
    KEGG_ko = first_nonempty(KEGG_ko, ""),
    KEGG_Pathway = first_nonempty(KEGG_Pathway, ""),
    PFAMs = first_nonempty(PFAMs, ""),
    .groups = "drop"
  )

annotation <- data.frame(gene_id = gene_universe) |>
  dplyr::left_join(gff_annotation, by = "gene_id") |>
  dplyr::left_join(eggnog_annotation, by = "gene_id") |>
  dplyr::mutate(
    Preferred_name = dplyr::coalesce(Preferred_name, "unannotated"),
    gene_description = dplyr::case_when(
      !is.na(gene_description) & gene_description != "" &
        gene_description != "no description" ~ gene_description,
      !is.na(gff_description) & gff_description != "" ~ gff_description,
      TRUE ~ "no description"
    ),
    gene_biotype = dplyr::coalesce(gene_biotype, "unknown")
  ) |>
  dplyr::select(
    gene_id, Preferred_name, gene_description, gene_biotype,
    GOs, KEGG_ko, KEGG_Pathway, PFAMs
  )

go_term_names <- read.csv(file.path(go_dir, "eggNOG_GO_TERM2NAME.csv"),
                          check.names = FALSE) |>
  dplyr::rename(term_id = term, GO_name = name, GO_ontology = ontology)

kegg_term_names <- read.csv(file.path(go_dir, "eggNOG_KEGG_pathway_TERM2NAME.csv"),
                            check.names = FALSE) |>
  dplyr::rename(
    term_id = term,
    KEGG_name = name,
    KEGG_theme = pathway_theme,
    KEGG_scope = pathway_scope
  )

non_directional <- read_optional_csv(file.path(go_dir, "all_enrichment_results.csv"))
directional <- read_optional_csv(file.path(go_dir, "requested_directional_all_enrichment.csv"))

term_gene_long <- dplyr::bind_rows(
  expand_enrichment_genes(non_directional, "Non-directional enrichment"),
  expand_enrichment_genes(directional, "Direction-aware enrichment")
) |>
  dplyr::left_join(go_term_names, by = "term_id") |>
  dplyr::left_join(kegg_term_names, by = "term_id") |>
  dplyr::mutate(
    term_name = dplyr::coalesce(GO_name, KEGG_name, term_description),
    ontology_or_theme = dplyr::coalesce(GO_ontology, KEGG_theme, pathway_theme),
    pathway_scope = dplyr::coalesce(KEGG_scope, pathway_scope)
  ) |>
  dplyr::left_join(annotation, by = "gene_id") |>
  dplyr::mutate(
    Preferred_name = dplyr::coalesce(Preferred_name, "unannotated"),
    gene_description = dplyr::coalesce(gene_description, "no description"),
    family = dplyr::coalesce(family, DEG_set),
    comparison = dplyr::coalesce(comparison, DEG_set),
    direction = dplyr::coalesce(direction, "Not split by direction"),
    plot_label = dplyr::coalesce(plot_label, DEG_set)
  ) |>
  dplyr::select(
    enrichment_context, database, family, comparison, direction, plot_label,
    DEG_set, term_id, term_name, term_description, ontology_or_theme,
    pathway_scope, GeneRatio, BgRatio, FoldEnrichment, p.adjust, qvalue,
    Count, gene_id, Preferred_name, gene_description, GOs, KEGG_ko,
    gene_biotype,
    KEGG_Pathway, PFAMs
  ) |>
  dplyr::arrange(database, family, comparison, direction, p.adjust, term_id,
                 gene_id)

term_gene_compact <- term_gene_long |>
  dplyr::group_by(
    enrichment_context, database, family, comparison, direction, plot_label,
    DEG_set, term_id, term_name, term_description, ontology_or_theme,
    pathway_scope, GeneRatio, BgRatio, FoldEnrichment, p.adjust, qvalue, Count
  ) |>
  dplyr::summarise(
    n_listed_genes = dplyr::n_distinct(gene_id),
    DEG_gene_IDs = paste(sort(unique(gene_id)), collapse = "; "),
    DEG_gene_names = paste(sort(unique(Preferred_name)), collapse = "; "),
    .groups = "drop"
  ) |>
  dplyr::arrange(database, family, comparison, direction, p.adjust, term_id)

write.csv(term_gene_long,
          file.path(out_dir, "enriched_term_gene_reverse_index_long.csv"),
          row.names = FALSE)
write.csv(term_gene_compact,
          file.path(out_dir, "enriched_term_gene_reverse_index_compact.csv"),
          row.names = FALSE)

summary_by_context <- term_gene_compact |>
  dplyr::count(enrichment_context, database, family,
               name = "n_enriched_terms") |>
  dplyr::arrange(enrichment_context, database, family)

knitr::kable(summary_by_context,
             caption = "Number of enriched terms represented in the reverse index")
write.csv(summary_by_context,
          file.path(out_dir, "enriched_term_gene_index_summary.csv"),
          row.names = FALSE)
```

## Compact term-to-gene list

This table has one row per enriched term per DEG set. The `DEG_gene_IDs` column
is the reverse list used to trace each enriched term back to its contributing
DEGs.

```{r compact-table}
compact_display <- term_gene_compact |>
  dplyr::mutate(
    p.adjust = signif(p.adjust, 3),
    qvalue = signif(qvalue, 3),
    FoldEnrichment = round(FoldEnrichment, 2)
  ) |>
  dplyr::select(
    enrichment_context, database, family, comparison, direction, term_id,
    term_name, ontology_or_theme, pathway_scope, p.adjust, FoldEnrichment,
    Count, n_listed_genes, DEG_gene_IDs
  )

DT::datatable(
  compact_display,
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    pageLength = 15,
    scrollX = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv")
  )
)
```

## Expanded term-gene table

This table has one row per enriched term per contributing DEG. It is easier for
gene-level filtering because gene IDs and descriptions are not collapsed.

```{r expanded-table}
long_display <- term_gene_long |>
  dplyr::mutate(
    p.adjust = signif(p.adjust, 3),
    qvalue = signif(qvalue, 3),
    FoldEnrichment = round(FoldEnrichment, 2)
  ) |>
  dplyr::select(
    enrichment_context, database, family, comparison, direction, term_id,
    term_name, ontology_or_theme, pathway_scope, p.adjust, FoldEnrichment,
    gene_id, Preferred_name, gene_description
  )

DT::datatable(
  long_display,
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv")
  )
)
```

## Exported files

```{r exported-files}
exports <- list.files(out_dir, full.names = FALSE)
knitr::kable(tibble::tibble(file = exports),
             caption = paste("Files written to", out_dir))
```

## Session information

```{r session-info}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
