---
title: "Circled heatmap cluster enrichment"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  latest_results_root: "output/rmd_runs"
  cluster_gene_file: ""
  counts_file: "data/raw_read_counts/master_counts_no_rRNA.csv"
  metadata_file: "data/metadata/mehreen_metadata_fixed.txt"
  timecourse_fatbody_reference_dir: "data/external/timecourse_fatbody_reference"
  timecourse_head_thorax_counts_dir: "data/raw_read_counts/03-gregaria-DESeq2"
  eggnog_file: "data/GO_Annotations/GCF_023897955.1_iqSchGreg1.2_Arthopoda_one2one.emapper.annotations"
  gff_file: "data/reference/GCF_023897955.1_iqSchGreg1.2_genomic.gff"
  min_total_count: 10
  alpha: 0.05
  top_terms_per_panel: 18
  output_tag: "circled-cluster-enrichment"
---

## Purpose

This page asks what functions are enriched in the upper heatmap branch that was
circled in the full all-samples fat body DEG diagnostic heatmap.

The foreground is the exported gene list for that circled branch. The enrichment
is run in two ways:

- against all expressed genes in the rRNA-filtered count matrix;
- against all genes in the full all-samples fat body DEG heatmap.

The first test asks what functions are over-represented compared with the whole
expressed transcriptome. The second test asks what makes this heatmap branch
different from the other fat body DEGs.

For GO enrichment, only terms that resolve to Biological process, Cellular
component, or Molecular function in the local GO database are kept in the final
tables and plots. Unresolved GO identifiers are omitted here because they do not
provide an interpretable functional label.

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c("clusterProfiler", "DESeq2", "dplyr", "tidyr", "stringr",
              "ggplot2", "DT", "GO.db", "AnnotationDbi", "tibble",
              "scales", "SummarizedExperiment")
required <- c(required, "patchwork", "ggdendro")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

library(ggplot2)
library(patchwork)

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

results_root <- file.path(project_dir, params$latest_results_root)
counts_file <- file.path(project_dir, params$counts_file)
metadata_file <- file.path(project_dir, params$metadata_file)
timecourse_fatbody_dir <- file.path(
  project_dir,
  params$timecourse_fatbody_reference_dir
)
timecourse_head_thorax_dir <- file.path(
  project_dir,
  params$timecourse_head_thorax_counts_dir
)
eggnog_file <- file.path(project_dir, params$eggnog_file)
gff_file <- file.path(project_dir, params$gff_file)

current_outliers <- c("1007", "1036", "1037", "1039")

go_ontology_colors <- c(
  "Biological process" = "#252525",
  "Cellular component" = "#737373",
  "Molecular function" = "#BDBDBD"
)

kegg_scope_colors <- c(
  "Eukaryote/metazoan-compatible" = "#1E88E5",
  "Core metabolism or cross-domain" = "#2F8F5B",
  "Broad KEGG reference map" = "#4B5563",
  "Microbial/prokaryote-associated label" = "#E41A1C",
  "Reference disease pathway" = "#7B4FA3",
  "Unclassified KEGG reference label" = "#F59E0B"
)
kegg_lollipop_color <- "#8B5A2B"
```

## Helper functions

```{r helper-functions}
latest_dir <- function(pattern) {
  dirs <- list.dirs(results_root, recursive = FALSE, full.names = TRUE)
  dirs <- dirs[grepl(pattern, basename(dirs))]
  if (length(dirs) == 0) return(NA_character_)
  dirs[which.max(file.info(dirs)$mtime)]
}

resolve_cluster_file <- function(path_param) {
  explicit_path <- if (is.null(path_param) || !nzchar(path_param)) {
    NA_character_
  } else if (grepl("^/", path_param)) {
    path_param
  } else {
    file.path(project_dir, path_param)
  }
  if (!is.na(explicit_path) && file.exists(explicit_path)) {
    return(explicit_path)
  }

  phase_dir <- latest_dir("^phase-signature-overlap_")
  if (is.na(phase_dir)) return(explicit_path)

  candidate <- file.path(
    phase_dir,
    "circled_upper_cluster_audit",
    "circled_upper_cluster_genes.csv"
  )
  candidate
}

read_emapper <- function(path) {
  cols <- c(
    "query", "seed_ortholog", "evalue", "score", "eggNOG_OGs",
    "max_annot_lvl", "COG_category", "Description", "Preferred_name",
    "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
    "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC", "CAZy",
    "BiGG_Reaction", "PFAMs"
  )
  tab <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", fill = TRUE)
  names(tab) <- cols[seq_len(ncol(tab))]
  tab |>
    dplyr::mutate(protein_id = stringr::str_extract(query, "XP_[0-9]+[.][0-9]+"))
}

parse_cds_gene_bridge <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  gff |>
    dplyr::filter(type == "CDS") |>
    dplyr::transmute(
      protein_id = stringr::str_match(attributes, "protein_id=([^;]+)")[, 2],
      gene_id = stringr::str_match(attributes, "gene=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(protein_id), !is.na(gene_id)) |>
    dplyr::distinct()
}

parse_gff_gene_annotation <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  gff |>
    dplyr::filter(type == "gene") |>
    dplyr::transmute(
      gene_id = dplyr::coalesce(
        stringr::str_match(attributes, "gene=([^;]+)")[, 2],
        stringr::str_match(attributes, "Name=([^;]+)")[, 2]
      ),
      Description = stringr::str_match(attributes, "description=([^;]+)")[, 2],
      gene_biotype = stringr::str_match(attributes, "gene_biotype=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(gene_id)) |>
    dplyr::mutate(
      Description = dplyr::if_else(
        is.na(Description) | Description == "",
        "no GFF description",
        utils::URLdecode(Description)
      ),
      gene_biotype = dplyr::coalesce(gene_biotype, "unknown")
    ) |>
    dplyr::distinct(gene_id, .keep_all = TRUE)
}

read_metadata <- function(path) {
  tab <- read.delim(path, check.names = FALSE)
  tab <- tab[, !is.na(names(tab)) & nzchar(names(tab)), drop = FALSE]
  tab$label <- as.character(tab$label)
  tab$Diet <- factor(tab$Diet, levels = c("33", "50", "83"))
  tab$Treatment <- factor(tab$Treatment, levels = c("Infected", "Control"))
  tab
}

read_featurecounts_gene_counts <- function(path, sample_id) {
  tab <- read.delim(path, comment.char = "#", check.names = FALSE)
  if ("Geneid" %in% names(tab)) {
    count_col <- names(tab)[ncol(tab)]
    out <- tibble::tibble(
      gene_id = as.character(tab[["Geneid"]]),
      count = as.integer(tab[[count_col]])
    )
  } else {
    tab <- read.delim(path, comment.char = "#", header = FALSE,
                      check.names = FALSE)
    if (ncol(tab) < 2) {
      stop("Count file must contain at least gene ID and count columns: ", path)
    }
    out <- tibble::tibble(
      gene_id = as.character(tab[[1]]),
      count = as.integer(tab[[ncol(tab)]])
    )
  }
  out <- out |>
    dplyr::filter(!is.na(gene_id), nzchar(gene_id)) |>
    dplyr::group_by(gene_id) |>
    dplyr::summarise(count = sum(count, na.rm = TRUE), .groups = "drop")
  names(out)[2] <- sample_id
  out
}

parse_head_thorax_reference_metadata <- function(reference_dir) {
  files <- list.files(
    reference_dir,
    pattern = "^SGRE-(HEAD|THOX)-(CRD|ISO)-[0-9]+_MERGE_featurecounts[.]txt$",
    full.names = TRUE
  )
  if (length(files) == 0) {
    return(tibble::tibble(
      sample_id = character(),
      count_file = character(),
      tissue = character(),
      phase = character(),
      rearing_condition = character(),
      replicate = integer(),
      source = character(),
      display_group = character(),
      reference_role = character(),
      sample_label = character(),
      sample_block = character()
    ))
  }

  parsed <- stringr::str_match(
    basename(files),
    "^SGRE-(HEAD|THOX)-(CRD|ISO)-([0-9]+)_MERGE_featurecounts[.]txt$"
  )
  tibble::tibble(
    sample_id = stringr::str_remove(basename(files), "_featurecounts[.]txt$"),
    count_file = files,
    tissue_code = parsed[, 2],
    rearing_code = parsed[, 3],
    replicate = as.integer(parsed[, 4])
  ) |>
    dplyr::filter(!is.na(tissue_code), !is.na(rearing_code)) |>
    dplyr::mutate(
      tissue = dplyr::case_when(
        tissue_code == "HEAD" ~ "Head",
        tissue_code == "THOX" ~ "Thorax",
        TRUE ~ tissue_code
      ),
      phase = dplyr::case_when(
        rearing_code == "CRD" ~ "Gregarious",
        rearing_code == "ISO" ~ "Solitarious",
        TRUE ~ rearing_code
      ),
      rearing_condition = dplyr::case_when(
        rearing_code == "CRD" ~ "Crowded",
        rearing_code == "ISO" ~ "Isolated",
        TRUE ~ rearing_code
      ),
      source = "Time-course head/thorax phase reference",
      display_group = paste0(
        tissue, ": ",
        dplyr::if_else(rearing_condition == "Crowded",
                       "crowded/gregarious", "isolated/solitarious")
      ),
      reference_role = paste(tissue, tolower(rearing_condition),
                             "rep", replicate),
      sample_label = paste0(substr(tissue, 1, 1), "-",
                            rearing_code, "-", replicate),
      sample_block = paste(tissue, "reference")
    ) |>
    dplyr::arrange(tissue, rearing_condition, replicate)
}

first_nonempty <- function(x, fallback = "") {
  x <- x[!is.na(x) & x != "" & x != "-"]
  if (length(x) == 0) fallback else x[1]
}

build_term_maps_from_eggnog <- function() {
  counts_raw <- read.csv(counts_file, check.names = FALSE)
  gene_universe <- unique(counts_raw[[1]])
  annotation <- read_emapper(eggnog_file) |>
    dplyr::left_join(parse_cds_gene_bridge(gff_file), by = "protein_id") |>
    dplyr::filter(!is.na(gene_id), gene_id %in% gene_universe) |>
    dplyr::group_by(gene_id) |>
    dplyr::summarise(
      GOs = first_nonempty(GOs, ""),
      KEGG_Pathway = first_nonempty(KEGG_Pathway, ""),
      .groups = "drop"
    )

  term2gene_go <- annotation |>
    dplyr::select(gene_id, GOs) |>
    tidyr::separate_rows(GOs, sep = ",") |>
    dplyr::filter(grepl("^GO:", GOs)) |>
    dplyr::transmute(term = GOs, gene = gene_id) |>
    dplyr::distinct()

  go_names <- AnnotationDbi::mapIds(
    GO.db::GO.db,
    keys = unique(term2gene_go$term),
    keytype = "GOID",
    column = "TERM",
    multiVals = "first"
  )
  go_ontology <- AnnotationDbi::mapIds(
    GO.db::GO.db,
    keys = unique(term2gene_go$term),
    keytype = "GOID",
    column = "ONTOLOGY",
    multiVals = "first"
  )
  term2name_go <- data.frame(
    term = names(go_names),
    name = unname(go_names),
    ontology = unname(go_ontology)
  ) |>
    dplyr::filter(!is.na(name))

  term2gene_kegg <- annotation |>
    dplyr::select(gene_id, KEGG_Pathway) |>
    tidyr::separate_rows(KEGG_Pathway, sep = ",") |>
    dplyr::filter(grepl("^ko[0-9]+$", KEGG_Pathway)) |>
    dplyr::transmute(term = KEGG_Pathway, gene = gene_id) |>
    dplyr::distinct()

  term2name_kegg <- term2gene_kegg |>
    dplyr::distinct(term) |>
    dplyr::mutate(
      name = term,
      pathway_theme = "Other KEGG pathway",
      pathway_scope = "Unclassified KEGG reference label"
    )

  list(
    term2gene_go = term2gene_go,
    term2name_go = term2name_go,
    term2gene_kegg = term2gene_kegg,
    term2name_kegg = term2name_kegg
  )
}

load_term_maps <- function() {
  go_dir <- latest_dir("^go-term-enrichment_")
  expected <- file.path(
    go_dir,
    c("eggNOG_GO_TERM2GENE.csv", "eggNOG_GO_TERM2NAME.csv",
      "eggNOG_KEGG_pathway_TERM2GENE.csv",
      "eggNOG_KEGG_pathway_TERM2NAME.csv")
  )
  if (!is.na(go_dir) && all(file.exists(expected))) {
    return(list(
      term2gene_go = read.csv(expected[1], check.names = FALSE),
      term2name_go = read.csv(expected[2], check.names = FALSE),
      term2gene_kegg = read.csv(expected[3], check.names = FALSE),
      term2name_kegg = read.csv(expected[4], check.names = FALSE)
    ))
  }
  build_term_maps_from_eggnog()
}

run_enrichment <- function(genes, universe, term2gene, term2name, database,
                           background_label) {
  term2gene <- term2gene |>
    dplyr::filter(gene %in% universe) |>
    dplyr::distinct(term, gene)
  term2name_min <- term2name |>
    dplyr::distinct(term, .keep_all = TRUE) |>
    dplyr::select(term, name)

  tested_universe <- intersect(universe, term2gene$gene)
  tested_genes <- intersect(genes, tested_universe)
  if (length(tested_genes) < 5 || length(tested_universe) < 10) {
    return(data.frame())
  }

  res <- clusterProfiler::enricher(
    gene = tested_genes,
    universe = tested_universe,
    TERM2GENE = term2gene,
    TERM2NAME = term2name_min,
    pvalueCutoff = 1,
    pAdjustMethod = "BH",
    qvalueCutoff = 1,
    minGSSize = 5,
    maxGSSize = 5000
  )
  if (is.null(res)) return(data.frame())
  out <- as.data.frame(res)
  if (nrow(out) == 0) return(data.frame())
  out |>
    dplyr::mutate(
      database = database,
      background = background_label,
      foreground_genes_tested = length(tested_genes),
      background_genes_tested = length(tested_universe)
    )
}

expand_term_genes <- function(tab, gene_annotation) {
  if (nrow(tab) == 0) {
    return(tibble::tibble())
  }
  tab |>
    dplyr::filter(p.adjust <= params$alpha) |>
    dplyr::select(database, background, ID, Description, GeneRatio, BgRatio,
                  pvalue, p.adjust, qvalue, Count, geneID) |>
    tidyr::separate_rows(geneID, sep = "/") |>
    dplyr::rename(term_id = ID, term_name = Description, gene_id = geneID) |>
    dplyr::left_join(gene_annotation, by = "gene_id")
}
```

## Inputs

```{r inputs}
cluster_file <- resolve_cluster_file(params$cluster_gene_file)
if (is.na(cluster_file) || !file.exists(cluster_file)) {
  stop("Could not find circled cluster gene file.")
}

cluster_run_dir <- dirname(dirname(cluster_file))
full_heatmap_gene_file <- file.path(
  cluster_run_dir,
  "full_experiment_deg_diagnostic_heatmap_gene_order.csv"
)
if (!file.exists(full_heatmap_gene_file)) {
  stop("Could not find full heatmap gene-order file: ", full_heatmap_gene_file)
}
full_heatmap_z_file <- file.path(
  cluster_run_dir,
  "full_experiment_deg_diagnostic_z_vst_matrix.csv.gz"
)
row_cluster_span_file <- file.path(
  cluster_run_dir,
  "circled_upper_cluster_audit",
  "row_cluster_spans_k4.csv"
)

input_files <- data.frame(
  input = c("Circled upper heatmap branch", "Full heatmap DEG order",
            "Full heatmap z-scored VST matrix", "Heatmap row-cluster spans",
            "Sample metadata", "rRNA-filtered count matrix",
            "eggNOG annotation", "GFF gene descriptions"),
  path = c(cluster_file, full_heatmap_gene_file, full_heatmap_z_file,
           row_cluster_span_file, metadata_file, counts_file, eggnog_file,
           gff_file),
  present = c(file.exists(cluster_file), file.exists(full_heatmap_gene_file),
              file.exists(full_heatmap_z_file),
              file.exists(row_cluster_span_file), file.exists(metadata_file),
              file.exists(counts_file), file.exists(eggnog_file),
              file.exists(gff_file))
)
knitr::kable(input_files, caption = "Inputs for circled-cluster enrichment")
if (!all(input_files$present)) {
  stop("One or more required inputs are missing.")
}
```

```{r build-input-sets}
circled_cluster <- read.csv(cluster_file, check.names = FALSE)
full_heatmap_genes <- read.csv(full_heatmap_gene_file, check.names = FALSE)
gff_annotation <- parse_gff_gene_annotation(gff_file)

counts_raw <- read.csv(counts_file, check.names = FALSE)
counts <- as.matrix(counts_raw[, -1, drop = FALSE])
rownames(counts) <- counts_raw[[1]]
storage.mode(counts) <- "numeric"
expressed_universe <- rownames(counts)[rowSums(counts, na.rm = TRUE) >=
                                        params$min_total_count]

cluster_genes <- unique(circled_cluster$gene_id)
full_deg_universe <- unique(full_heatmap_genes$gene_id)

set_summary <- data.frame(
  set = c("Circled upper heatmap branch",
          "All expressed rRNA-filtered genes",
          "All all-samples fat body DEGs in heatmap"),
  n_genes = c(length(cluster_genes), length(expressed_universe),
              length(full_deg_universe))
)

write.csv(set_summary, file.path(out_dir, "circled_cluster_set_summary.csv"),
          row.names = FALSE)
knitr::kable(set_summary, caption = "Gene sets used for enrichment")
```

## Enrichment tests

```{r run-enrichment}
maps <- load_term_maps()

maps$term2gene_go <- maps$term2gene_go |>
  dplyr::filter(gene %in% expressed_universe)
maps$term2gene_kegg <- maps$term2gene_kegg |>
  dplyr::filter(gene %in% expressed_universe)

backgrounds <- list(
  "Expressed rRNA-filtered genes" = expressed_universe,
  "All fat body DEGs in heatmap" = intersect(full_deg_universe, expressed_universe)
)

go_enrichment <- dplyr::bind_rows(lapply(names(backgrounds), function(bg) {
  run_enrichment(
    genes = cluster_genes,
    universe = backgrounds[[bg]],
    term2gene = maps$term2gene_go,
    term2name = maps$term2name_go,
    database = "GO",
    background_label = bg
  )
}))

kegg_enrichment <- dplyr::bind_rows(lapply(names(backgrounds), function(bg) {
  run_enrichment(
    genes = cluster_genes,
    universe = backgrounds[[bg]],
    term2gene = maps$term2gene_kegg,
    term2name = maps$term2name_kegg,
    database = "KEGG_pathway",
    background_label = bg
  )
}))

if (nrow(go_enrichment) > 0) {
  go_enrichment <- go_enrichment |>
    dplyr::left_join(
      maps$term2name_go |>
        dplyr::distinct(term, .keep_all = TRUE) |>
        dplyr::transmute(
          ID = term,
          ontology = dplyr::recode(
            ontology,
            BP = "Biological process",
            CC = "Cellular component",
            MF = "Molecular function",
            .default = NA_character_
          )
        ),
      by = "ID"
    ) |>
    dplyr::filter(!is.na(ontology))
}

if (nrow(kegg_enrichment) > 0) {
  kegg_enrichment <- kegg_enrichment |>
    dplyr::left_join(
      maps$term2name_kegg |>
        dplyr::distinct(term, .keep_all = TRUE) |>
        dplyr::transmute(ID = term, pathway_theme, pathway_scope),
      by = "ID"
    ) |>
    dplyr::mutate(
      pathway_theme = dplyr::coalesce(pathway_theme, "Other KEGG pathway"),
      pathway_scope = dplyr::coalesce(pathway_scope,
                                      "Unclassified KEGG reference label")
    )
}

all_enrichment <- dplyr::bind_rows(go_enrichment, kegg_enrichment)
significant_enrichment <- all_enrichment |>
  dplyr::filter(p.adjust <= params$alpha) |>
  dplyr::arrange(database, background, p.adjust)

write.csv(go_enrichment,
          file.path(out_dir, "circled_cluster_GO_enrichment_all_terms.csv"),
          row.names = FALSE)
write.csv(kegg_enrichment,
          file.path(out_dir, "circled_cluster_KEGG_enrichment_all_terms.csv"),
          row.names = FALSE)
write.csv(all_enrichment,
          file.path(out_dir, "circled_cluster_enrichment_all_terms.csv"),
          row.names = FALSE)
write.csv(significant_enrichment,
          file.path(out_dir, "circled_cluster_enrichment_significant_terms.csv"),
          row.names = FALSE)

enrichment_count_summary <- significant_enrichment |>
  dplyr::count(database, background, name = "n_significant_terms") |>
  tidyr::complete(database = c("GO", "KEGG_pathway"),
                  background = names(backgrounds),
                  fill = list(n_significant_terms = 0)) |>
  dplyr::arrange(database, background)

write.csv(enrichment_count_summary,
          file.path(out_dir, "circled_cluster_enrichment_count_summary.csv"),
          row.names = FALSE)
knitr::kable(enrichment_count_summary,
             caption = "Significant enriched terms in the circled cluster")
```

## Expression-cluster landscape

The full diagnostic heatmap was first divided into the four major row branches
from the same dendrogram used to identify the circled cluster. Because the
cluster-level cladogram showed the original branches 3 and 4 were the closest
pair, they are merged here into a single interpreted `Cluster 3+4`. The top panel
below shows the z-scored VST expression landscape for all fat body DEGs, split
by the three interpreted expression clusters. The two lollipop panels below
summarize significant GO terms for each expression cluster, using the full fat
body DEG heatmap as the background. This asks what makes each expression block
functionally different from the other DEG blocks in the same heatmap.

```{r expression-cluster-landscape, fig.width=24, fig.height=26}
full_heatmap_z <- read.csv(gzfile(full_heatmap_z_file), check.names = FALSE,
                           row.names = 1)
full_heatmap_z <- as.matrix(full_heatmap_z)
storage.mode(full_heatmap_z) <- "numeric"

raw_sample_order <- colnames(full_heatmap_z)
sample_metadata <- read_metadata(metadata_file) |>
  dplyr::filter(label %in% raw_sample_order) |>
  dplyr::mutate(
    original_sample_index = match(label, raw_sample_order),
    Treatment_plot = factor(as.character(Treatment),
                            levels = c("Control", "Infected")),
    Diet_plot = factor(as.character(Diet), levels = c("33", "50", "83"))
  ) |>
  dplyr::arrange(Treatment_plot, Diet_plot, original_sample_index)
sample_metadata <- sample_metadata |>
  dplyr::mutate(
    sample_index = dplyr::row_number(),
    sample_group = factor(
      paste(Treatment_plot, Diet_plot),
      levels = c("Control 33", "Control 50", "Control 83",
                 "Infected 33", "Infected 50", "Infected 83")
    ),
    outlier_status = dplyr::if_else(
      label %in% current_outliers,
      "Discussed outlier",
      "Other sample"
    )
  )
sample_order <- sample_metadata$label

row_cluster_spans <- read.csv(row_cluster_span_file, check.names = FALSE) |>
  dplyr::arrange(start) |>
  dplyr::mutate(
    original_expression_cluster_order = dplyr::row_number(),
    original_expression_cluster = paste0(
      "Original branch ", original_expression_cluster_order,
      "\n(n = ", n, " genes)"
    ),
    expression_cluster_order = dplyr::case_when(
      original_expression_cluster_order %in% c(1, 2) ~ original_expression_cluster_order,
      original_expression_cluster_order %in% c(3, 4) ~ 3L
    )
  )

merged_cluster_spans <- row_cluster_spans |>
  dplyr::group_by(expression_cluster_order) |>
  dplyr::summarise(
    start = min(start),
    end = max(end),
    n = sum(n),
    original_branches = paste(original_expression_cluster_order, collapse = "+"),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    expression_cluster = dplyr::case_when(
      expression_cluster_order == 1 ~ "DEG cluster 1",
      expression_cluster_order == 2 ~ "DEG cluster 2",
      expression_cluster_order == 3 ~ "DEG cluster 3+4"
    )
  )
cluster_levels <- merged_cluster_spans$expression_cluster
expression_cluster_colors <- stats::setNames(
  c("#315C8A", "#C85A54", "#2F8F5B"),
  cluster_levels
)
cluster_index <- findInterval(
  full_heatmap_genes$heatmap_row_order,
  row_cluster_spans$start
)
cluster_index <- pmin(pmax(cluster_index, 1), nrow(row_cluster_spans))
merged_cluster_index <- row_cluster_spans$expression_cluster_order[cluster_index]
merged_cluster_match <- match(
  merged_cluster_index,
  merged_cluster_spans$expression_cluster_order
)

deg_context_levels <- c(
  "Diet in controls",
  "Diet in infected",
  "Infection within diet",
  "Infection + diet",
  "Diet in controls + infected",
  "Other fat body DEG"
)
deg_context_labels <- c(
  "Diet in controls" = "Diet DEG in controls",
  "Diet in infected" = "Diet DEG in infected",
  "Infection within diet" = "Infected vs control within diet",
  "Infection + diet" = "Infection + diet DEG",
  "Diet in controls + infected" = "Diet DEG in controls + infected",
  "Other fat body DEG" = "Other fat body DEG"
)
deg_context_colors <- c(
  "Diet in controls" = "#00A7B5",
  "Diet in infected" = "#E377C2",
  "Infection within diet" = "#B6A500",
  "Infection + diet" = "#111827",
  "Diet in controls + infected" = "#2B6CB0",
  "Other fat body DEG" = "#C8CDD2"
)
deg_context_plot_levels <- c(
  "~ diet (controls)",
  "~ diet (infected)",
  "~ diet + infection",
  "shared DEGs",
  "Other fat body DEG"
)
deg_context_plot_colors <- c(
  "~ diet (controls)" = "#00A7B5",
  "~ diet (infected)" = "#E377C2",
  "~ diet + infection" = "#B6A500",
  "shared DEGs" = "#111827",
  "Other fat body DEG" = "#C8CDD2"
)

expression_cluster_assignments <- full_heatmap_genes |>
  dplyr::mutate(
    expression_cluster_order = merged_cluster_index,
    expression_cluster = factor(merged_cluster_spans$expression_cluster[merged_cluster_match],
                                levels = cluster_levels),
    original_expression_cluster_order =
      row_cluster_spans$original_expression_cluster_order[cluster_index],
    original_expression_cluster =
      row_cluster_spans$original_expression_cluster[cluster_index],
    original_pheatmap_cluster_id = row_cluster_spans$cluster_id[cluster_index],
    within_cluster_order = heatmap_row_order -
      merged_cluster_spans$start[merged_cluster_match] + 1,
    original_within_cluster_order = heatmap_row_order -
      row_cluster_spans$start[cluster_index] + 1,
    DEG_context = dplyr::coalesce(DEG_context, "Other fat body DEG"),
    DEG_context = dplyr::if_else(
      DEG_context %in% deg_context_levels,
      DEG_context,
      "Other fat body DEG"
    ),
    DEG_context = factor(DEG_context, levels = deg_context_levels)
  ) |>
  dplyr::arrange(
    expression_cluster_order,
    DEG_context,
    min_padj,
    dplyr::desc(max_abs_log2FC),
    original_within_cluster_order
  ) |>
  dplyr::group_by(expression_cluster) |>
  dplyr::mutate(within_cluster_order = dplyr::row_number()) |>
  dplyr::ungroup() |>
  dplyr::select(heatmap_row_order, gene_id, expression_cluster_order,
                expression_cluster, original_expression_cluster_order,
                original_expression_cluster, original_pheatmap_cluster_id,
                within_cluster_order, original_within_cluster_order,
                Description, DEG_context, phase_class,
                n_deg_contrasts, comparisons, directions, max_abs_log2FC,
                min_padj)

write.csv(
  expression_cluster_assignments,
  file.path(out_dir, "full_heatmap_expression_cluster_assignments.csv"),
  row.names = FALSE
)

cluster_go_enrichment <- dplyr::bind_rows(lapply(cluster_levels, function(cl) {
  genes <- expression_cluster_assignments$gene_id[
    expression_cluster_assignments$expression_cluster == cl
  ]
  run_enrichment(
    genes = genes,
    universe = backgrounds[["All fat body DEGs in heatmap"]],
    term2gene = maps$term2gene_go,
    term2name = maps$term2name_go,
    database = "GO",
    background_label = "All fat body DEGs in heatmap"
  ) |>
    dplyr::mutate(
      expression_cluster = cl,
      n_cluster_genes = length(genes)
    )
}))

if (nrow(cluster_go_enrichment) > 0) {
  cluster_go_enrichment <- cluster_go_enrichment |>
    dplyr::left_join(
      maps$term2name_go |>
        dplyr::distinct(term, .keep_all = TRUE) |>
        dplyr::transmute(
          ID = term,
          ontology = dplyr::recode(
            ontology,
            BP = "Biological process",
            CC = "Cellular component",
            MF = "Molecular function",
            .default = NA_character_
          )
        ),
      by = "ID"
    ) |>
    dplyr::filter(!is.na(ontology)) |>
    dplyr::mutate(
      expression_cluster = factor(expression_cluster, levels = cluster_levels),
      ontology = factor(ontology, levels = names(go_ontology_colors))
    )
}

cluster_go_significant <- cluster_go_enrichment |>
  dplyr::filter(p.adjust <= params$alpha) |>
  dplyr::arrange(expression_cluster, ontology, p.adjust)

cluster_go_summary <- cluster_go_significant |>
  dplyr::count(expression_cluster, ontology, name = "n_significant_terms") |>
  tidyr::complete(expression_cluster = factor(cluster_levels,
                                             levels = cluster_levels),
                  ontology = factor(names(go_ontology_colors),
                                    levels = names(go_ontology_colors)),
                  fill = list(n_significant_terms = 0)) |>
  dplyr::arrange(expression_cluster, ontology)

write.csv(cluster_go_enrichment,
          file.path(out_dir,
                    "expression_cluster_GO_enrichment_all_terms.csv"),
          row.names = FALSE)
write.csv(cluster_go_significant,
          file.path(out_dir,
                    "expression_cluster_GO_enrichment_significant_terms.csv"),
          row.names = FALSE)
write.csv(cluster_go_summary,
          file.path(out_dir,
                    "expression_cluster_GO_enrichment_count_summary.csv"),
          row.names = FALSE)

cluster_gene_annotation <- expression_cluster_assignments |>
  dplyr::select(gene_id, DEG_context, phase_class, n_deg_contrasts,
                comparisons, directions, max_abs_log2FC, min_padj) |>
  dplyr::left_join(
    gff_annotation |> dplyr::select(gene_id, Description, gene_biotype),
    by = "gene_id"
  ) |>
  dplyr::mutate(
    Description = dplyr::coalesce(Description, "no GFF description"),
    gene_biotype = dplyr::coalesce(gene_biotype, "unknown")
  )

cluster_term_gene_long <- cluster_go_enrichment |>
  dplyr::filter(p.adjust <= params$alpha) |>
  dplyr::select(expression_cluster, database, background, ID, Description,
                GeneRatio, BgRatio, pvalue, p.adjust, qvalue, Count, geneID) |>
  tidyr::separate_rows(geneID, sep = "/") |>
  dplyr::rename(term_id = ID, term_name = Description, gene_id = geneID) |>
  dplyr::left_join(cluster_gene_annotation, by = "gene_id") |>
  dplyr::arrange(expression_cluster, p.adjust, term_id, gene_id)

cluster_term_gene_compact <- cluster_term_gene_long |>
  dplyr::group_by(expression_cluster, database, background, term_id, term_name,
                  GeneRatio, BgRatio, p.adjust, Count) |>
  dplyr::summarise(
    DEG_gene_IDs = paste(sort(unique(gene_id)), collapse = "; "),
    DEG_gene_descriptions = paste(sort(unique(Description)), collapse = "; "),
    DEG_contexts = paste(sort(unique(DEG_context)), collapse = "; "),
    .groups = "drop"
  ) |>
  dplyr::arrange(expression_cluster, p.adjust)

write.csv(cluster_term_gene_long,
          file.path(out_dir,
                    "expression_cluster_GO_term_gene_index_long.csv"),
          row.names = FALSE)
write.csv(cluster_term_gene_compact,
          file.path(out_dir,
                    "expression_cluster_GO_term_gene_index_compact.csv"),
          row.names = FALSE)

sample_group_ranges <- sample_metadata |>
  dplyr::group_by(Treatment_plot, Diet_plot, sample_group) |>
  dplyr::summarise(start = min(sample_index), end = max(sample_index),
                   .groups = "drop") |>
  dplyr::arrange(Treatment_plot, Diet_plot)
sample_gaps <- sample_group_ranges$end[-nrow(sample_group_ranges)] + 0.5
treatment_gaps <- sample_group_ranges$end[
  which(as.character(sample_group_ranges$Treatment_plot[-nrow(sample_group_ranges)]) !=
          as.character(sample_group_ranges$Treatment_plot[-1]))
] + 0.5
diet_gaps <- setdiff(sample_gaps, treatment_gaps)

annotation_fill_colors <- c(
  "Treatment: Infected" = "#8B0000",
  "Treatment: Control" = "#FAEBD7",
  "Diet: 33" = "#E6862E",
  "Diet: 50" = "#2F8F5B",
  "Diet: 83" = "#7B4FA3"
)

sample_x_limits <- c(-1.25, length(sample_order) + 0.5)

sample_annotation_long <- sample_metadata |>
  dplyr::transmute(
    sample_index,
    Treatment = as.character(Treatment),
    Diet = as.character(Diet)
  ) |>
  tidyr::pivot_longer(
    cols = c(Treatment, Diet),
    names_to = "track",
    values_to = "value"
  ) |>
  dplyr::mutate(
    track = factor(track, levels = c("Diet", "Treatment")),
    track_y = as.numeric(track),
    fill_key = paste(track, value, sep = ": ")
  )

sample_annotation_plot <- ggplot2::ggplot(
  sample_annotation_long,
  ggplot2::aes(x = sample_index, y = track_y, fill = fill_key)
) +
  ggplot2::geom_tile(width = 0.95, height = 0.9) +
  ggplot2::geom_vline(xintercept = diet_gaps, color = "black",
                      linewidth = 0.35) +
  ggplot2::geom_vline(xintercept = treatment_gaps, color = "black",
                      linewidth = 0.8) +
  ggplot2::scale_fill_manual(values = annotation_fill_colors, drop = FALSE) +
  ggplot2::scale_x_continuous(
    limits = sample_x_limits,
    breaks = sample_metadata$sample_index,
    labels = sample_metadata$label,
    expand = c(0, 0),
    position = "top"
  ) +
  ggplot2::scale_y_continuous(
    breaks = seq_len(2),
    labels = c("Diet", "Treatment"),
    limits = c(0.45, 2.55),
    expand = c(0, 0),
    sec.axis = ggplot2::dup_axis(
      breaks = seq_len(2),
      labels = c("Diet", "Treatment")
    )
  ) +
  ggplot2::labs(
    x = NULL,
    y = NULL,
    title = "A. Full fat body DEG expression landscape split by row clusters",
    fill = "Sample annotation"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank(),
    axis.text.y = ggplot2::element_text(face = "bold", size = 11),
    axis.text.y.right = ggplot2::element_text(face = "bold", size = 11),
    legend.position = "bottom",
    legend.title = ggplot2::element_text(face = "bold"),
    plot.title = ggplot2::element_text(face = "bold", size = 15),
    plot.margin = ggplot2::margin(3, 4, 0, 4)
  )

heatmap_z_plot <- full_heatmap_z[
  expression_cluster_assignments$gene_id,
  sample_order,
  drop = FALSE
]
heatmap_z_plot <- pmax(pmin(heatmap_z_plot, 3), -3)
heatmap_df <- as.data.frame(heatmap_z_plot, check.names = FALSE) |>
  tibble::rownames_to_column("gene_id") |>
  dplyr::left_join(
    expression_cluster_assignments |>
      dplyr::select(gene_id, expression_cluster, within_cluster_order,
                    DEG_context),
    by = "gene_id"
  ) |>
  tidyr::pivot_longer(
    cols = dplyr::all_of(sample_order),
    names_to = "sample_id",
    values_to = "z"
  ) |>
  dplyr::mutate(sample_index = match(sample_id, sample_order))

heatmap_colors <- grDevices::colorRampPalette(
  c("#237C8E", "#F7F3E8", "#C85A54")
)(101)

gene_context_strip_df <- expression_cluster_assignments |>
  dplyr::select(gene_id, expression_cluster, within_cluster_order,
                DEG_context) |>
  dplyr::mutate(
    DEG_context_plot = dplyr::case_when(
      DEG_context == "Diet in controls" ~
        "~ diet (controls)",
      DEG_context == "Diet in infected" ~
        "~ diet (infected)",
      DEG_context == "Infection within diet" ~
        "~ diet + infection",
      DEG_context %in% c("Infection + diet",
                         "Diet in controls + infected") ~
        "shared DEGs",
      TRUE ~ "Other fat body DEG"
    ),
    DEG_context_plot = factor(DEG_context_plot,
                              levels = deg_context_plot_levels)
  )
present_deg_context_levels <- deg_context_plot_levels[
  deg_context_plot_levels %in%
    as.character(gene_context_strip_df$DEG_context_plot)
]
present_deg_context_labels <- stringr::str_wrap(
  present_deg_context_levels,
  width = 26
)

cluster_heatmap_plot <- ggplot2::ggplot(
  heatmap_df,
  ggplot2::aes(x = sample_index, y = within_cluster_order, fill = z)
) +
  ggplot2::geom_segment(
    data = gene_context_strip_df,
    ggplot2::aes(
      x = -1.1,
      xend = 0.35,
      y = within_cluster_order,
      yend = within_cluster_order,
      color = DEG_context_plot
    ),
    inherit.aes = FALSE,
    linewidth = 1.05,
    lineend = "butt"
  ) +
  ggplot2::geom_raster() +
  ggplot2::geom_vline(xintercept = 0.5, color = "#9CA3AF",
                      linewidth = 0.4) +
  ggplot2::geom_vline(xintercept = diet_gaps, color = "black",
                      linewidth = 0.25) +
  ggplot2::geom_vline(xintercept = treatment_gaps, color = "black",
                      linewidth = 0.65) +
  ggplot2::facet_grid(expression_cluster ~ ., scales = "free_y",
                      switch = "y") +
  ggplot2::scale_fill_gradientn(
    colors = heatmap_colors,
    limits = c(-3, 3),
    oob = scales::squish,
    name = "Row z-score"
  ) +
  ggplot2::scale_color_manual(
    values = deg_context_plot_colors,
    breaks = present_deg_context_levels,
    labels = present_deg_context_labels,
    drop = TRUE,
    name = "DEG context"
  ) +
  ggplot2::scale_x_continuous(
    limits = sample_x_limits,
    breaks = sample_metadata$sample_index,
    labels = sample_metadata$label,
    expand = c(0, 0)
  ) +
  ggplot2::scale_y_reverse(expand = c(0, 0)) +
  ggplot2::labs(
    x = NULL,
    y = NULL,
    title = NULL,
    subtitle = NULL
  ) +
  ggplot2::theme_classic(base_size = 13) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, hjust = 1,
                                        vjust = 0.5, size = 13),
    axis.text.y = ggplot2::element_blank(),
    axis.ticks.y = ggplot2::element_blank(),
    strip.placement = "outside",
    strip.background.y = ggplot2::element_rect(fill = "#F3F4F6",
                                               color = "#9CA3AF"),
    strip.text.y.left = ggplot2::element_text(angle = 90, face = "bold",
                                              size = 13),
    panel.spacing.y = grid::unit(0.25, "lines"),
    legend.position = "right",
    legend.box = "vertical",
    legend.title = ggplot2::element_text(face = "bold"),
    legend.text = ggplot2::element_text(size = 9),
    plot.title = ggplot2::element_text(face = "bold"),
    plot.margin = ggplot2::margin(0, 4, 3, 4)
  ) +
  ggplot2::guides(
    fill = ggplot2::guide_colorbar(
      order = 1,
      title.position = "top",
      barheight = grid::unit(3.0, "cm"),
      barwidth = grid::unit(0.35, "cm")
    ),
    color = ggplot2::guide_legend(
      order = 2,
      title.position = "top",
      override.aes = list(linewidth = 3)
    )
  )

ggplot2::ggsave(
  file.path(out_dir, "expression_cluster_gene_level_heatmap.png"),
  cluster_heatmap_plot,
  width = 20,
  height = 12,
  dpi = 260
)

prepare_lollipop_df <- function(tab, ontologies, n_terms,
                                terms_per_ontology = FALSE) {
  group_vars <- if (terms_per_ontology) {
    c("expression_cluster", "ontology")
  } else {
    "expression_cluster"
  }

  tab |>
    dplyr::filter(p.adjust <= params$alpha, ontology %in% ontologies) |>
    dplyr::group_by(dplyr::across(dplyr::all_of(group_vars))) |>
    dplyr::slice_min(p.adjust, n = n_terms, with_ties = FALSE) |>
    dplyr::ungroup() |>
    dplyr::arrange(expression_cluster, ontology, p.adjust) |>
    dplyr::group_by(expression_cluster) |>
    dplyr::mutate(
      neg_log10_padj = -log10(pmax(p.adjust, .Machine$double.xmin)),
      term_label = stringr::str_wrap(Description, 46),
      term_plot_id = paste(term_label, dplyr::row_number(),
                           expression_cluster, sep = "___")
    ) |>
    dplyr::ungroup() |>
    dplyr::mutate(
      term_plot_id = factor(term_plot_id, levels = rev(unique(term_plot_id))),
      expression_cluster = factor(expression_cluster, levels = cluster_levels)
    )
}

clean_term_labels <- function(x) {
  labels <- sub("___.*$", "", x)
  dplyr::if_else(labels == "No significant GO terms", "", labels)
}

make_lollipop_plot <- function(plot_df, title, subtitle = NULL) {
  if (nrow(plot_df) == 0) {
    return(
      ggplot2::ggplot() +
        ggplot2::annotate("text", x = 0, y = 0,
                          label = "No significant GO terms",
                          size = 6, fontface = "bold") +
        ggplot2::theme_void() +
        ggplot2::labs(title = title, subtitle = subtitle)
      )
  }

  x_axis_max <- max(
    c(plot_df$neg_log10_padj, -log10(params$alpha)),
    na.rm = TRUE
  )
  x_axis_limit <- x_axis_max + max(0.8, x_axis_max * 0.18)
  label_space <- x_axis_max * 2.4
  x_breaks <- pretty(c(0, x_axis_limit), n = 5)
  x_breaks <- x_breaks[x_breaks >= 0 & x_breaks <= x_axis_limit]
  missing_label_x <- max(
    c(plot_df$neg_log10_padj, -log10(params$alpha)),
    na.rm = TRUE
  ) / 2
  missing_cluster_df <- tibble::tibble(
    expression_cluster = factor(
      setdiff(cluster_levels, as.character(unique(plot_df$expression_cluster))),
      levels = cluster_levels
    ),
    term_plot_id = "No significant GO terms",
    x = missing_label_x,
    label = "No significant GO terms"
  ) |>
    dplyr::filter(!is.na(expression_cluster))

  ggplot2::ggplot(
    plot_df,
    ggplot2::aes(x = neg_log10_padj, y = term_plot_id)
  ) +
    ggplot2::geom_vline(
      xintercept = -log10(params$alpha),
      linetype = "dashed",
      color = "#6B7280",
      linewidth = 0.35
    ) +
    ggplot2::geom_vline(
      xintercept = 0,
      color = "#9CA3AF",
      linewidth = 0.35
    ) +
    ggplot2::geom_text(
      data = plot_df,
      ggplot2::aes(
        x = -label_space * 0.98,
        y = term_plot_id,
        label = clean_term_labels(term_plot_id)
      ),
      inherit.aes = FALSE,
      hjust = 0,
      size = 5.2,
      lineheight = 0.84,
      color = "#111827"
    ) +
    ggplot2::geom_segment(
      ggplot2::aes(x = 0, xend = neg_log10_padj,
                   y = term_plot_id, yend = term_plot_id,
                   color = ontology),
      linewidth = 0.7,
      alpha = 0.75
    ) +
    ggplot2::geom_point(
      ggplot2::aes(size = Count, color = ontology),
      alpha = 0.95
    ) +
    ggplot2::geom_text(
      data = missing_cluster_df,
      ggplot2::aes(x = x, y = term_plot_id, label = label),
      inherit.aes = FALSE,
      size = 4,
      color = "#6B7280",
      fontface = "bold"
    ) +
    ggplot2::facet_wrap(~ expression_cluster, nrow = 1,
                        scales = "free_y", drop = FALSE) +
    ggplot2::scale_color_manual(values = go_ontology_colors, drop = FALSE) +
    ggplot2::scale_y_discrete(labels = function(x) rep("", length(x))) +
    ggplot2::scale_x_continuous(
      limits = c(-label_space, x_axis_limit),
      breaks = x_breaks,
      expand = c(0, 0)
    ) +
    ggplot2::scale_size_continuous(range = c(3.5, 9)) +
    ggplot2::labs(
      title = title,
      subtitle = subtitle,
      x = "-log10 adjusted p-value",
      y = NULL,
      color = "GO ontology",
      size = "Cluster genes in term"
    ) +
    ggplot2::theme_classic(base_size = 13) +
    ggplot2::theme(
      axis.text.y = ggplot2::element_blank(),
      axis.ticks.y = ggplot2::element_blank(),
      axis.line = ggplot2::element_blank(),
      axis.line.x.bottom = ggplot2::element_line(color = "black",
                                                 linewidth = 0.35),
      panel.border = ggplot2::element_blank(),
      strip.background = ggplot2::element_rect(fill = "#F3F4F6",
                                               color = "#9CA3AF"),
      strip.text = ggplot2::element_text(face = "bold", size = 12),
      legend.position = "bottom",
      plot.title = ggplot2::element_text(face = "bold")
    )
}

go_lollipop_df <- prepare_lollipop_df(
  cluster_go_significant,
  ontologies = names(go_ontology_colors),
  n_terms = 4,
  terms_per_ontology = TRUE
)

go_lollipop_plot <- make_lollipop_plot(
  go_lollipop_df,
  title = "B. GO terms by expression cluster",
  subtitle = "Foreground = genes in each expression cluster; background = all fat body DEGs in the heatmap"
)

gene_level_GO_figure <- sample_annotation_plot /
  cluster_heatmap_plot /
  go_lollipop_plot +
  patchwork::plot_layout(heights = c(0.95, 6.8, 5.6)) +
  patchwork::plot_annotation(
    title = "Full gene-level expression and GO enrichment by DEG cluster",
    subtitle = "All samples are retained; GO lollipops combine BP, CC, and MF terms in each cluster"
  )

ggplot2::ggsave(
  file.path(out_dir, "expression_cluster_gene_level_GO_lollipop.png"),
  gene_level_GO_figure,
  width = 21,
  height = 19,
  dpi = 220
)

knitr::kable(
  cluster_go_summary,
  caption = "Significant GO terms by expression cluster and ontology"
)
```

## Combined GO and KEGG figure

The main figure below keeps the full gene-by-sample heatmap so every DEG in the
circled branch remains visible, then summarizes GO terms and KEGG pathways for
the same expression clusters. KEGG terms are eggNOG-linked `ko` reference
pathways, not insect-only organism pathways. The pathway-scope category is kept
in the exported KEGG tables, but panel C uses one brown color so the pathway
labels remain visually simple. The code also exports optional
per-sample and treatment-average cluster-score summaries as secondary diagnostic
files.

### DEG context annotation in panel A

The `DEG context` strip on the left of panel A is a contrast-family annotation.
It tells which type of DESeq2 contrast detected each gene as significant before
the heatmap was clustered. These labels are useful for reading the heatmap, but
they should not be interpreted as mutually exclusive biological mechanisms.

`~ diet (controls)` means the gene was significant in diet comparisons among
control animals only, for example control 33 vs 50, 83 vs 33, or 83 vs 50. This
is the closest label to a baseline diet effect without infection.

`~ diet (infected)` means the gene was significant in diet comparisons among
infected animals only. This asks whether infected animals on different diets
have different expression.

`~ diet + infection` means the gene was significant when infected and
control animals were compared within at least one diet. This is the most direct
infection-response label in this heatmap, but the comparison is diet-stratified,
so the label keeps both diet and infection in view.

`shared DEGs` means the gene was detected in more than one contrast family. This
combines genes that were diet-associated in both control and infected
backgrounds, and genes that were detected in both infection and diet contrast
families. It is an overlap label and does not by itself prove a formal
diet-by-infection interaction.

`Other fat body DEG` is a fallback category for genes in the full fat-body DEG
heatmap that do not fall cleanly into the simplified contrast families above. It
is retained in the code for completeness, but it may not appear in every figure.

```{r expression-cluster-treatment-summary, fig.width=18, fig.height=18}
cluster_sample_scores <- heatmap_df |>
  dplyr::group_by(expression_cluster, sample_id) |>
  dplyr::summarise(cluster_mean_z = mean(z, na.rm = TRUE),
                   .groups = "drop") |>
  dplyr::left_join(
    sample_metadata |>
      dplyr::select(label, Treatment, Diet, sample_group, outlier_status),
    by = c("sample_id" = "label")
  ) |>
  dplyr::mutate(
    expression_cluster = factor(expression_cluster, levels = cluster_levels),
    sample_group = factor(sample_group,
                          levels = unique(sample_metadata$sample_group)),
    sample_label = dplyr::if_else(
      outlier_status == "Discussed outlier",
      paste0(sample_id, " *"),
      sample_id
    ),
    sample_label = factor(
      sample_label,
      levels = rev(dplyr::if_else(
        sample_metadata$outlier_status == "Discussed outlier",
        paste0(sample_metadata$label, " *"),
        sample_metadata$label
      ))
    )
  )

treatment_cluster_summary <- cluster_sample_scores |>
  dplyr::group_by(expression_cluster, Treatment, Diet, sample_group) |>
  dplyr::summarise(
    n_samples = dplyr::n(),
    mean_z = mean(cluster_mean_z, na.rm = TRUE),
    sd_z = stats::sd(cluster_mean_z, na.rm = TRUE),
    se_z = sd_z / sqrt(n_samples),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    sample_group_plot = factor(sample_group,
                               levels = rev(unique(sample_metadata$sample_group))),
    tile_label = sprintf("%.2f", mean_z)
  )

write.csv(cluster_sample_scores,
          file.path(out_dir,
                    "expression_cluster_sample_scores.csv"),
          row.names = FALSE)
write.csv(treatment_cluster_summary,
          file.path(out_dir,
                    "expression_cluster_treatment_summary.csv"),
          row.names = FALSE)

sample_cluster_heatmap_plot <- ggplot2::ggplot(
  cluster_sample_scores,
  ggplot2::aes(x = expression_cluster, y = sample_label, fill = cluster_mean_z)
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.8) +
  ggplot2::facet_grid(sample_group ~ ., scales = "free_y",
                      space = "free_y", switch = "y") +
  ggplot2::scale_fill_gradientn(
    colors = heatmap_colors,
    limits = c(-2, 2),
    oob = scales::squish,
    name = "Sample cluster\nscore"
  ) +
  ggplot2::scale_x_discrete(position = "top") +
  ggplot2::labs(
    title = "A. Sample-level cluster expression heatmap",
    subtitle = "Each tile is one sample's mean score for a gene-expression cluster; * marks discussed outliers",
    x = "Expression cluster",
    y = "Sample grouped by diet/treatment"
  ) +
  ggplot2::theme_classic(base_size = 14) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(face = "bold", size = 12),
    axis.text.y = ggplot2::element_text(face = "bold", size = 8),
    axis.title = ggplot2::element_text(face = "bold"),
    strip.placement = "outside",
    strip.background.y = ggplot2::element_rect(fill = "#F3F4F6",
                                               color = "#9CA3AF"),
    strip.text.y.left = ggplot2::element_text(angle = 0, face = "bold",
                                              size = 11),
    panel.spacing.y = grid::unit(0.15, "lines"),
    legend.position = "right",
    plot.title = ggplot2::element_text(face = "bold")
  )

ggplot2::ggsave(
  file.path(out_dir, "expression_cluster_sample_level_heatmap.png"),
  sample_cluster_heatmap_plot,
  width = 10,
  height = 10,
  dpi = 300
)

treatment_cluster_heatmap_plot <- ggplot2::ggplot(
  treatment_cluster_summary,
  ggplot2::aes(x = expression_cluster, y = sample_group_plot, fill = mean_z)
) +
  ggplot2::geom_tile(color = "white", linewidth = 1.2) +
  ggplot2::geom_text(
    ggplot2::aes(label = tile_label),
    size = 5,
    fontface = "bold",
    color = "#111827"
  ) +
  ggplot2::scale_fill_gradientn(
    colors = heatmap_colors,
    limits = c(-2, 2),
    oob = scales::squish,
    name = "Mean row\nz-score"
  ) +
  ggplot2::scale_x_discrete(position = "top") +
  ggplot2::labs(
    title = "A. Treatment-by-cluster expression summary",
    subtitle = "Each tile is the mean cluster score across samples in that diet/treatment group",
    x = "Expression cluster",
    y = "Diet/treatment group"
  ) +
  ggplot2::theme_classic(base_size = 14) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(face = "bold", size = 12),
    axis.text.y = ggplot2::element_text(face = "bold", size = 12),
    axis.title = ggplot2::element_text(face = "bold"),
    legend.position = "right",
    plot.title = ggplot2::element_text(face = "bold")
  )

ggplot2::ggsave(
  file.path(out_dir, "expression_cluster_treatment_average_heatmap.png"),
  treatment_cluster_heatmap_plot,
  width = 10,
  height = 5.5,
  dpi = 300
)

cluster_kegg_enrichment <- dplyr::bind_rows(lapply(cluster_levels, function(cl) {
  genes <- expression_cluster_assignments$gene_id[
    expression_cluster_assignments$expression_cluster == cl
  ]
  run_enrichment(
    genes = genes,
    universe = backgrounds[["All fat body DEGs in heatmap"]],
    term2gene = maps$term2gene_kegg,
    term2name = maps$term2name_kegg,
    database = "KEGG_pathway",
    background_label = "All fat body DEGs in heatmap"
  ) |>
    dplyr::mutate(
      expression_cluster = cl,
      n_cluster_genes = length(genes)
    )
}))

if (nrow(cluster_kegg_enrichment) > 0) {
  cluster_kegg_enrichment <- cluster_kegg_enrichment |>
    dplyr::left_join(
      maps$term2name_kegg |>
        dplyr::distinct(term, .keep_all = TRUE) |>
        dplyr::transmute(ID = term, pathway_theme, pathway_scope),
      by = "ID"
    ) |>
    dplyr::mutate(
      pathway_theme = dplyr::coalesce(pathway_theme, "Other KEGG pathway"),
      pathway_scope = dplyr::coalesce(pathway_scope,
                                      "Unclassified KEGG reference label"),
      expression_cluster = factor(expression_cluster, levels = cluster_levels),
      pathway_scope = factor(pathway_scope, levels = names(kegg_scope_colors))
    )
}

cluster_kegg_significant <- cluster_kegg_enrichment |>
  dplyr::filter(p.adjust <= params$alpha) |>
  dplyr::arrange(expression_cluster, pathway_theme, p.adjust)

cluster_kegg_summary <- cluster_kegg_significant |>
  dplyr::count(expression_cluster, pathway_theme, pathway_scope,
               name = "n_significant_pathways") |>
  dplyr::arrange(expression_cluster, pathway_theme, pathway_scope)

write.csv(cluster_kegg_enrichment,
          file.path(out_dir,
                    "expression_cluster_KEGG_enrichment_all_terms.csv"),
          row.names = FALSE)
write.csv(cluster_kegg_significant,
          file.path(out_dir,
                    "expression_cluster_KEGG_enrichment_significant_terms.csv"),
          row.names = FALSE)
write.csv(cluster_kegg_summary,
          file.path(out_dir,
                    "expression_cluster_KEGG_enrichment_count_summary.csv"),
          row.names = FALSE)

cluster_kegg_term_gene_long <- cluster_kegg_enrichment |>
  dplyr::filter(p.adjust <= params$alpha) |>
  dplyr::select(expression_cluster, database, background, pathway_theme,
                pathway_scope, ID, Description, GeneRatio, BgRatio, pvalue,
                p.adjust, qvalue, Count, geneID) |>
  tidyr::separate_rows(geneID, sep = "/") |>
  dplyr::rename(term_id = ID, term_name = Description, gene_id = geneID) |>
  dplyr::left_join(cluster_gene_annotation, by = "gene_id") |>
  dplyr::arrange(expression_cluster, p.adjust, term_id, gene_id)

cluster_kegg_term_gene_compact <- cluster_kegg_term_gene_long |>
  dplyr::group_by(expression_cluster, database, background, pathway_theme,
                  pathway_scope, term_id, term_name, GeneRatio, BgRatio,
                  p.adjust, Count) |>
  dplyr::summarise(
    DEG_gene_IDs = paste(sort(unique(gene_id)), collapse = "; "),
    DEG_gene_descriptions = paste(sort(unique(Description)), collapse = "; "),
    DEG_contexts = paste(sort(unique(DEG_context)), collapse = "; "),
    .groups = "drop"
  ) |>
  dplyr::arrange(expression_cluster, p.adjust)

write.csv(cluster_kegg_term_gene_long,
          file.path(out_dir,
                    "expression_cluster_KEGG_term_gene_index_long.csv"),
          row.names = FALSE)
write.csv(cluster_kegg_term_gene_compact,
          file.path(out_dir,
                    "expression_cluster_KEGG_term_gene_index_compact.csv"),
          row.names = FALSE)

prepare_kegg_lollipop_df <- function(tab, n_terms) {
  tab |>
    dplyr::filter(p.adjust <= params$alpha) |>
    dplyr::group_by(expression_cluster) |>
    dplyr::slice_min(p.adjust, n = n_terms, with_ties = FALSE) |>
    dplyr::ungroup() |>
    dplyr::arrange(expression_cluster, pathway_theme, p.adjust) |>
    dplyr::group_by(expression_cluster) |>
    dplyr::mutate(
      neg_log10_padj = -log10(pmax(p.adjust, .Machine$double.xmin)),
      pathway_label = stringr::str_wrap(Description, 34),
      pathway_plot_id = paste(pathway_label, dplyr::row_number(),
                              expression_cluster, sep = "___")
    ) |>
    dplyr::ungroup() |>
    dplyr::mutate(
      pathway_plot_id = factor(pathway_plot_id,
                               levels = rev(unique(pathway_plot_id))),
      expression_cluster = factor(expression_cluster, levels = cluster_levels)
    )
}

clean_kegg_labels <- function(x) {
  labels <- sub("___.*$", "", x)
  dplyr::if_else(labels == "No significant KEGG pathways", "", labels)
}

make_kegg_lollipop_plot <- function(plot_df, title, subtitle = NULL) {
  if (nrow(plot_df) == 0) {
    return(
      ggplot2::ggplot() +
        ggplot2::annotate("text", x = 0, y = 0,
                          label = "No significant KEGG pathways",
                          size = 6, fontface = "bold") +
        ggplot2::theme_void() +
        ggplot2::labs(title = title, subtitle = subtitle)
      )
  }

  x_axis_max <- max(
    c(plot_df$neg_log10_padj, -log10(params$alpha)),
    na.rm = TRUE
  )
  x_axis_limit <- x_axis_max + max(0.8, x_axis_max * 0.18)
  label_space <- x_axis_max * 2.4
  x_breaks <- pretty(c(0, x_axis_limit), n = 5)
  x_breaks <- x_breaks[x_breaks >= 0 & x_breaks <= x_axis_limit]
  missing_label_x <- max(
    c(plot_df$neg_log10_padj, -log10(params$alpha)),
    na.rm = TRUE
  ) / 2
  missing_cluster_df <- tibble::tibble(
    expression_cluster = factor(
      setdiff(cluster_levels, as.character(unique(plot_df$expression_cluster))),
      levels = cluster_levels
    ),
    pathway_plot_id = "No significant KEGG pathways",
    x = missing_label_x,
    label = "No significant KEGG pathways"
  ) |>
    dplyr::filter(!is.na(expression_cluster))

  ggplot2::ggplot(
    plot_df,
    ggplot2::aes(x = neg_log10_padj, y = pathway_plot_id)
  ) +
    ggplot2::geom_vline(
      xintercept = -log10(params$alpha),
      linetype = "dashed",
      color = "#6B7280",
      linewidth = 0.35
    ) +
    ggplot2::geom_vline(
      xintercept = 0,
      color = "#9CA3AF",
      linewidth = 0.35
    ) +
    ggplot2::geom_text(
      data = plot_df,
      ggplot2::aes(
        x = -label_space * 0.98,
        y = pathway_plot_id,
        label = clean_kegg_labels(pathway_plot_id)
      ),
      inherit.aes = FALSE,
      hjust = 0,
      size = 5.2,
      lineheight = 0.84,
      color = "#111827"
    ) +
    ggplot2::geom_segment(
      ggplot2::aes(x = 0, xend = neg_log10_padj,
                   y = pathway_plot_id, yend = pathway_plot_id),
      color = kegg_lollipop_color,
      linewidth = 0.8,
      alpha = 0.78
    ) +
    ggplot2::geom_point(
      ggplot2::aes(size = Count),
      color = kegg_lollipop_color,
      alpha = 0.95
    ) +
    ggplot2::geom_text(
      data = missing_cluster_df,
      ggplot2::aes(x = x, y = pathway_plot_id, label = label),
      inherit.aes = FALSE,
      size = 4,
      color = "#6B7280",
      fontface = "bold"
    ) +
    ggplot2::facet_wrap(~ expression_cluster, nrow = 1,
                        scales = "free_y", drop = FALSE) +
    ggplot2::scale_y_discrete(labels = function(x) rep("", length(x))) +
    ggplot2::scale_x_continuous(
      limits = c(-label_space, x_axis_limit),
      breaks = x_breaks,
      expand = c(0, 0)
    ) +
    ggplot2::scale_size_continuous(range = c(4, 10)) +
    ggplot2::labs(
      title = title,
      subtitle = subtitle,
      x = "-log10 adjusted p-value",
      y = NULL,
      size = "Cluster genes in pathway"
    ) +
    ggplot2::theme_classic(base_size = 13) +
    ggplot2::theme(
      axis.text.y = ggplot2::element_blank(),
      axis.ticks.y = ggplot2::element_blank(),
      axis.line = ggplot2::element_blank(),
      axis.line.x.bottom = ggplot2::element_line(color = "black",
                                                 linewidth = 0.35),
      panel.border = ggplot2::element_blank(),
      strip.background = ggplot2::element_rect(fill = "#F3F4F6",
                                               color = "#9CA3AF"),
      strip.text = ggplot2::element_text(face = "bold", size = 12),
      legend.position = "bottom",
      legend.box = "vertical",
      plot.title = ggplot2::element_text(face = "bold")
    )
}

kegg_lollipop_df <- prepare_kegg_lollipop_df(
  cluster_kegg_significant,
  n_terms = 8
)
kegg_lollipop_plot <- make_kegg_lollipop_plot(
  kegg_lollipop_df,
  title = "C. KEGG pathways by expression cluster",
  subtitle = "Foreground = genes in each expression cluster; background = all fat body DEGs in the heatmap"
)

gene_level_KEGG_figure <- sample_annotation_plot /
  cluster_heatmap_plot /
  kegg_lollipop_plot +
  patchwork::plot_layout(heights = c(0.95, 6.8, 3.7)) +
  patchwork::plot_annotation(
    title = "Full gene-level expression and KEGG enrichment by expression cluster",
    subtitle = "Top panel keeps all genes in the circled heatmap branch; lower panel shows KEGG pathways enriched in each cluster"
  )

ggplot2::ggsave(
  file.path(out_dir,
            "expression_cluster_gene_level_KEGG_lollipop.png"),
  gene_level_KEGG_figure,
  width = 21,
  height = 15,
  dpi = 220
)

gene_level_GO_KEGG_figure <- sample_annotation_plot /
  cluster_heatmap_plot /
  go_lollipop_plot /
  kegg_lollipop_plot +
  patchwork::plot_layout(heights = c(0.95, 6.8, 5.6, 3.7)) +
  patchwork::plot_annotation(
    title = "Full gene-level expression and functional enrichment by DEG cluster"
  )

print(gene_level_GO_KEGG_figure)
ggplot2::ggsave(
  file.path(out_dir,
            "expression_cluster_gene_level_GO_KEGG_lollipop.png"),
  gene_level_GO_KEGG_figure,
  width = 21,
  height = 22,
  dpi = 220
)

knitr::kable(
  cluster_kegg_summary,
  caption = "Significant KEGG pathways by expression cluster"
)
```

## Composite with reference tissues

This companion composite keeps the same fat-body DEG rows as panel A above, but
adds VST expression columns for the available reference samples: the two
time-course FAT phase controls and the copied head/thorax phase-reference
samples. These reference samples are not used to call the DEGs in this figure;
they are shown only to ask whether the fat-body DEG clusters also have visible
expression structure in other tissues or in the independent FAT controls.

```{r expression-cluster-reference-composite, fig.width=24, fig.height=22}
timecourse_reference_metadata <- file.path(
  timecourse_fatbody_dir,
  "gregaria_FAT_phase_reference_metadata.csv"
)

timecourse_ref_meta <- if (file.exists(timecourse_reference_metadata)) {
  read.csv(timecourse_reference_metadata, check.names = FALSE) |>
    dplyr::mutate(
      count_file = file.path(timecourse_fatbody_dir, "readcounts", file_name),
      file_present = file.exists(count_file)
    )
} else {
  tibble::tibble()
}

head_thorax_ref_meta <- parse_head_thorax_reference_metadata(
  timecourse_head_thorax_dir
) |>
  dplyr::mutate(file_present = file.exists(count_file))

fatbody_ref_meta <- if (nrow(timecourse_ref_meta) > 0) {
  timecourse_ref_meta |>
    dplyr::mutate(
      tissue = "Fat body",
      source = "Time-course fat body phase control",
      display_group = dplyr::case_when(
        phase == "Gregarious" ~ "Fat body: gregarious/crowded control",
        phase == "Solitarious" ~ "Fat body: solitarious/isolated control",
        TRUE ~ "Fat body: reference control"
      ),
      sample_label = dplyr::case_when(
        phase == "Gregarious" ~ "FAT-G",
        phase == "Solitarious" ~ "FAT-S",
        TRUE ~ sample_id
      ),
      sample_block = "FAT controls",
      treatment_or_phase = phase,
      diet = "Reference"
    ) |>
    dplyr::select(
      sample_id, count_file, tissue, phase, rearing_condition, replicate,
      source, display_group, reference_role, sample_label, sample_block,
      treatment_or_phase, diet, file_present
    )
} else {
  tibble::tibble()
}

head_thorax_ref_meta <- head_thorax_ref_meta |>
  dplyr::mutate(
    treatment_or_phase = phase,
    diet = "Reference"
  ) |>
  dplyr::select(
    sample_id, count_file, tissue, phase, rearing_condition, replicate,
    source, display_group, reference_role, sample_label, sample_block,
    treatment_or_phase, diet, file_present
  )

reference_manifest <- dplyr::bind_rows(
  fatbody_ref_meta,
  head_thorax_ref_meta
)

reference_inputs <- reference_manifest |>
  dplyr::count(sample_block, tissue, treatment_or_phase, file_present,
               name = "n_samples")
knitr::kable(
  reference_inputs,
  caption = "Reference samples available for the expanded panel A"
)

reference_ready <- nrow(reference_manifest) > 0 &&
  all(reference_manifest$file_present)

if (reference_ready) {
  experiment_counts_raw <- read.csv(counts_file, check.names = FALSE)
  experiment_counts <- as.matrix(experiment_counts_raw[, -1, drop = FALSE])
  rownames(experiment_counts) <- experiment_counts_raw[[1]]
  colnames(experiment_counts) <- sub(
    "_MERGE$",
    "",
    sub("^mehreen_", "", colnames(experiment_counts))
  )
  storage.mode(experiment_counts) <- "integer"
  experiment_counts <- experiment_counts[, sample_order, drop = FALSE]

  experiment_reference_meta <- sample_metadata |>
    dplyr::transmute(
      sample_id = label,
      count_file = NA_character_,
      tissue = "Fat body",
      phase = "Diet-infection experiment",
      rearing_condition = "Semi-isolated experiment",
      replicate = NA_integer_,
      source = "Diet-infection fat body experiment",
      display_group = paste(Treatment, Diet, sep = "_"),
      reference_role = paste("Fat body", Treatment, "diet", Diet),
      sample_label = label,
      sample_block = "Experiment fat body",
      treatment_or_phase = as.character(Treatment),
      diet = as.character(Diet),
      file_present = TRUE
    )

  reference_manifest <- reference_manifest |>
    dplyr::mutate(
      sample_block = factor(
        sample_block,
        levels = c("FAT controls", "Head reference", "Thorax reference")
      ),
      tissue = factor(tissue, levels = c("Fat body", "Head", "Thorax")),
      treatment_or_phase = factor(
        treatment_or_phase,
        levels = c("Control", "Infected", "Gregarious", "Solitarious")
      )
    ) |>
    dplyr::arrange(sample_block, tissue, treatment_or_phase, replicate)

  combined_reference_meta <- dplyr::bind_rows(
    experiment_reference_meta,
    reference_manifest
  )

  reference_count_tables <- Map(
    read_featurecounts_gene_counts,
    reference_manifest$count_file,
    reference_manifest$sample_id
  )
  reference_counts_df <- Reduce(
    function(x, y) dplyr::full_join(x, y, by = "gene_id"),
    reference_count_tables
  )
  reference_counts <- as.matrix(reference_counts_df[, -1, drop = FALSE])
  rownames(reference_counts) <- reference_counts_df$gene_id
  reference_counts[is.na(reference_counts)] <- 0L
  storage.mode(reference_counts) <- "integer"

  shared_reference_genes <- Reduce(
    intersect,
    list(rownames(experiment_counts), rownames(reference_counts))
  )

  combined_counts <- cbind(
    experiment_counts[shared_reference_genes, , drop = FALSE],
    reference_counts[
      shared_reference_genes,
      reference_manifest$sample_id,
      drop = FALSE
    ]
  )

  combined_reference_meta <- combined_reference_meta[
    match(colnames(combined_counts), combined_reference_meta$sample_id),
  ]
  combined_reference_meta$sample_block <- factor(
    combined_reference_meta$sample_block,
    levels = c(
      "Experiment fat body",
      "FAT controls",
      "Head reference",
      "Thorax reference"
    )
  )
  combined_reference_meta$tissue <- factor(
    combined_reference_meta$tissue,
    levels = c("Fat body", "Head", "Thorax")
  )
  combined_reference_meta$treatment_or_phase <- factor(
    combined_reference_meta$treatment_or_phase,
    levels = c("Control", "Infected", "Gregarious", "Solitarious")
  )
  combined_reference_meta$diet <- factor(
    combined_reference_meta$diet,
    levels = c("33", "50", "83", "Reference")
  )
  rownames(combined_reference_meta) <- combined_reference_meta$sample_id

  keep_reference <- rowSums(combined_counts) >= params$min_total_count
  dds_reference <- DESeq2::DESeqDataSetFromMatrix(
    countData = combined_counts[keep_reference, , drop = FALSE],
    colData = as.data.frame(combined_reference_meta),
    design = ~ 1
  )
  vst_reference <- DESeq2::vst(dds_reference, blind = TRUE)
  vst_reference_mat <- SummarizedExperiment::assay(vst_reference)

  reference_gene_table <- expression_cluster_assignments |>
    dplyr::filter(gene_id %in% rownames(vst_reference_mat)) |>
    dplyr::arrange(expression_cluster_order, within_cluster_order) |>
    dplyr::group_by(expression_cluster) |>
    dplyr::mutate(reference_within_cluster_order = dplyr::row_number()) |>
    dplyr::ungroup()

  reference_sample_order <- combined_reference_meta$sample_id
  reference_heatmap_z <- vst_reference_mat[
    reference_gene_table$gene_id,
    reference_sample_order,
    drop = FALSE
  ]
  reference_heatmap_z <- t(scale(t(reference_heatmap_z)))
  reference_heatmap_z[is.na(reference_heatmap_z)] <- 0
  reference_heatmap_z <- pmax(pmin(reference_heatmap_z, 3), -3)

  reference_sample_metadata <- combined_reference_meta |>
    tibble::rownames_to_column("sample_id_check") |>
    dplyr::select(-sample_id_check) |>
    dplyr::mutate(
      sample_index = dplyr::row_number(),
      sample_group_key = paste(sample_block, treatment_or_phase, diet, sep = "|")
    )

  reference_sample_ranges <- reference_sample_metadata |>
    dplyr::group_by(sample_group_key, sample_block) |>
    dplyr::summarise(start = min(sample_index), end = max(sample_index),
                     .groups = "drop") |>
    dplyr::arrange(start)
  reference_group_gaps <- reference_sample_ranges$end[
    -nrow(reference_sample_ranges)
  ] + 0.5
  reference_block_gaps <- reference_sample_ranges$end[
    which(
      as.character(reference_sample_ranges$sample_block[-nrow(reference_sample_ranges)]) !=
        as.character(reference_sample_ranges$sample_block[-1])
    )
  ] + 0.5
  reference_group_gaps <- setdiff(reference_group_gaps, reference_block_gaps)

  reference_annotation_colors <- c(
    "Tissue: Fat body" = "#D9A441",
    "Tissue: Head" = "#009E9A",
    "Tissue: Thorax" = "#C51B7D",
    "Diet: 33" = "#E6862E",
    "Diet: 50" = "#2F8F5B",
    "Diet: 83" = "#7B4FA3",
    "Diet: Reference" = "#D1D5DB",
    "State: Control" = "#FAEBD7",
    "State: Infected" = "#8B0000",
    "State: Gregarious" = "#111827",
    "State: Solitarious" = "#6B7280"
  )

  reference_annotation_long <- reference_sample_metadata |>
    dplyr::transmute(
      sample_index,
      Tissue = as.character(tissue),
      Diet = as.character(diet),
      State = as.character(treatment_or_phase)
    ) |>
    tidyr::pivot_longer(
      cols = c(Tissue, Diet, State),
      names_to = "track",
      values_to = "value"
    ) |>
    dplyr::mutate(
      track = factor(track, levels = c("Tissue", "Diet", "State")),
      track_y = as.numeric(track),
      fill_key = paste(track, value, sep = ": ")
    )

  reference_x_limits <- c(-1.25, length(reference_sample_order) + 0.5)

  reference_sample_annotation_plot <- ggplot2::ggplot(
    reference_annotation_long,
    ggplot2::aes(x = sample_index, y = track_y, fill = fill_key)
  ) +
    ggplot2::geom_tile(width = 0.95, height = 0.9) +
    ggplot2::geom_vline(xintercept = reference_group_gaps,
                        color = "black", linewidth = 0.25) +
    ggplot2::geom_vline(xintercept = reference_block_gaps,
                        color = "black", linewidth = 0.8) +
    ggplot2::scale_fill_manual(values = reference_annotation_colors,
                               drop = FALSE) +
    ggplot2::scale_x_continuous(
      limits = reference_x_limits,
      breaks = reference_sample_metadata$sample_index,
      labels = reference_sample_metadata$sample_label,
      expand = c(0, 0),
      position = "top"
    ) +
    ggplot2::scale_y_continuous(
      breaks = seq_len(3),
      labels = c("Tissue", "Diet", "State"),
      limits = c(0.45, 3.55),
      expand = c(0, 0)
    ) +
    ggplot2::labs(x = NULL, y = NULL, fill = "Sample annotation") +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      panel.grid = ggplot2::element_blank(),
      axis.text.x = ggplot2::element_blank(),
      axis.ticks.x = ggplot2::element_blank(),
      axis.text.y = ggplot2::element_text(face = "bold", size = 9),
      legend.position = "bottom",
      legend.title = ggplot2::element_text(face = "bold"),
      legend.text = ggplot2::element_text(size = 8)
    ) +
    ggplot2::guides(
      fill = ggplot2::guide_legend(nrow = 2, byrow = TRUE)
    )

  reference_gene_context_strip_df <- reference_gene_table |>
    dplyr::select(gene_id, expression_cluster, reference_within_cluster_order,
                  DEG_context) |>
    dplyr::mutate(
      DEG_context_plot = dplyr::case_when(
        DEG_context == "Diet in controls" ~ "~ diet (controls)",
        DEG_context == "Diet in infected" ~ "~ diet (infected)",
        DEG_context == "Infection within diet" ~ "~ diet + infection",
        DEG_context %in% c("Infection + diet",
                           "Diet in controls + infected") ~ "shared DEGs",
        TRUE ~ "Other fat body DEG"
      ),
      DEG_context_plot = factor(DEG_context_plot,
                                levels = deg_context_plot_levels)
    )
  present_reference_deg_context_levels <- deg_context_plot_levels[
    deg_context_plot_levels %in%
      as.character(reference_gene_context_strip_df$DEG_context_plot)
  ]
  present_reference_deg_context_labels <- stringr::str_wrap(
    present_reference_deg_context_levels,
    width = 26
  )

  reference_heatmap_df <- as.data.frame(reference_heatmap_z,
                                        check.names = FALSE) |>
    tibble::rownames_to_column("gene_id") |>
    dplyr::left_join(
      reference_gene_table |>
        dplyr::select(gene_id, expression_cluster,
                      reference_within_cluster_order),
      by = "gene_id"
    ) |>
    tidyr::pivot_longer(
      cols = dplyr::all_of(reference_sample_order),
      names_to = "sample_id",
      values_to = "z"
    ) |>
    dplyr::mutate(sample_index = match(sample_id, reference_sample_order))

  reference_cluster_heatmap_plot <- ggplot2::ggplot(
    reference_heatmap_df,
    ggplot2::aes(x = sample_index, y = reference_within_cluster_order,
                 fill = z)
  ) +
    ggplot2::geom_segment(
      data = reference_gene_context_strip_df,
      ggplot2::aes(
        x = -1.1,
        xend = 0.35,
        y = reference_within_cluster_order,
        yend = reference_within_cluster_order,
        color = DEG_context_plot
      ),
      inherit.aes = FALSE,
      linewidth = 1.05,
      lineend = "butt"
    ) +
    ggplot2::geom_raster() +
    ggplot2::geom_vline(xintercept = 0.5, color = "#9CA3AF",
                        linewidth = 0.4) +
    ggplot2::geom_vline(xintercept = reference_group_gaps,
                        color = "black", linewidth = 0.25) +
    ggplot2::geom_vline(xintercept = reference_block_gaps,
                        color = "black", linewidth = 0.8) +
    ggplot2::facet_grid(expression_cluster ~ ., scales = "free_y",
                        switch = "y") +
    ggplot2::scale_fill_gradientn(
      colors = heatmap_colors,
      limits = c(-3, 3),
      oob = scales::squish,
      name = "Row z-score"
    ) +
    ggplot2::scale_color_manual(
      values = deg_context_plot_colors,
      breaks = present_reference_deg_context_levels,
      labels = present_reference_deg_context_labels,
      drop = TRUE,
      name = "DEG context"
    ) +
    ggplot2::scale_x_continuous(
      limits = reference_x_limits,
      breaks = reference_sample_metadata$sample_index,
      labels = reference_sample_metadata$sample_label,
      expand = c(0, 0)
    ) +
    ggplot2::scale_y_reverse(expand = c(0, 0)) +
    ggplot2::labs(
      x = NULL,
      y = NULL,
      title = "A. Fat body DEG expression landscape with reference tissues",
      subtitle = paste0(
        "Rows are ", nrow(reference_gene_table),
        " fat-body DEGs; head, thorax, and FAT references are VST context columns"
      )
    ) +
    ggplot2::theme_classic(base_size = 13) +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 90, hjust = 1,
                                          vjust = 0.5, size = 7),
      axis.text.y = ggplot2::element_blank(),
      axis.ticks.y = ggplot2::element_blank(),
      strip.placement = "outside",
      strip.background.y = ggplot2::element_rect(fill = "#F3F4F6",
                                                 color = "#9CA3AF"),
      strip.text.y.left = ggplot2::element_text(angle = 90, face = "bold",
                                                size = 13),
      panel.spacing.y = grid::unit(0.25, "lines"),
      legend.position = "right",
      legend.box = "vertical",
      legend.title = ggplot2::element_text(face = "bold"),
      legend.text = ggplot2::element_text(size = 9),
      plot.title = ggplot2::element_text(face = "bold")
    ) +
    ggplot2::guides(
      fill = ggplot2::guide_colorbar(
        order = 1,
        title.position = "top",
        barheight = grid::unit(3.0, "cm"),
        barwidth = grid::unit(0.35, "cm")
      ),
      color = ggplot2::guide_legend(
        order = 2,
        title.position = "top",
        override.aes = list(linewidth = 3)
      )
    )

  reference_gene_level_GO_KEGG_figure <- reference_sample_annotation_plot /
    reference_cluster_heatmap_plot /
    go_lollipop_plot /
    kegg_lollipop_plot +
    patchwork::plot_layout(heights = c(1.05, 6.8, 5.6, 3.7)) +
    patchwork::plot_annotation(
      title = "Fat body DEG expression clusters with reference tissues",
      subtitle = paste0(
        "DEGs were called only in the fat body diet-infection experiment; ",
        "reference columns show VST row-z-scores for context"
      )
    )

  print(reference_gene_level_GO_KEGG_figure)
  ggplot2::ggsave(
    file.path(out_dir,
              "expression_cluster_gene_level_GO_KEGG_lollipop_with_references.png"),
    reference_gene_level_GO_KEGG_figure,
    width = 24,
    height = 22,
    dpi = 220
  )
  grDevices::svg(
    file.path(out_dir,
              "expression_cluster_gene_level_GO_KEGG_lollipop_with_references.svg"),
    width = 24,
    height = 22,
    onefile = FALSE
  )
  print(reference_gene_level_GO_KEGG_figure)
  grDevices::dev.off()

  ggplot2::ggsave(
    file.path(out_dir,
              "expression_cluster_reference_panel_A_heatmap.png"),
    reference_sample_annotation_plot / reference_cluster_heatmap_plot +
      patchwork::plot_layout(heights = c(1.05, 6.8)),
    width = 24,
    height = 11,
    dpi = 260
  )

  write.csv(
    reference_gene_table,
    file.path(out_dir,
              "expression_cluster_reference_panel_A_genes.csv"),
    row.names = FALSE
  )
  write.csv(
    combined_reference_meta,
    file.path(out_dir,
              "expression_cluster_reference_panel_A_samples.csv"),
    row.names = FALSE
  )
} else {
  message("Reference composite skipped because one or more reference count files are missing.")
}
```

## Cluster cladogram

This dendrogram is a cluster-level summary of the heatmap row branches. It is
not a phylogeny; it clusters the average row-z-scored expression profile of each
expression cluster across the 44 fat body samples. It is useful for checking
which major heatmap blocks have the most similar expression patterns.

```{r expression-cluster-cladogram, fig.width=9, fig.height=6}
cluster_short_labels <- gsub("\n.*$", "", cluster_levels)
cluster_label_map <- stats::setNames(cluster_short_labels, cluster_levels)
cluster_size_map <- expression_cluster_assignments |>
  dplyr::count(expression_cluster, name = "n_genes") |>
  dplyr::mutate(cluster_short = cluster_label_map[as.character(expression_cluster)])

cluster_centroids <- do.call(rbind, lapply(cluster_levels, function(cl) {
  genes <- expression_cluster_assignments$gene_id[
    expression_cluster_assignments$expression_cluster == cl
  ]
  colMeans(full_heatmap_z[genes, sample_order, drop = FALSE], na.rm = TRUE)
}))
rownames(cluster_centroids) <- cluster_short_labels

cluster_centroid_dist <- stats::dist(cluster_centroids, method = "euclidean")
cluster_centroid_hclust <- stats::hclust(cluster_centroid_dist,
                                         method = "complete")
cluster_dendro_data <- ggdendro::dendro_data(
  stats::as.dendrogram(cluster_centroid_hclust),
  type = "rectangle"
)

cluster_pairwise_distances <- as.data.frame(as.table(
  as.matrix(cluster_centroid_dist)
)) |>
  dplyr::rename(cluster_a = Var1, cluster_b = Var2,
                centroid_distance = Freq) |>
  dplyr::filter(as.character(cluster_a) < as.character(cluster_b)) |>
  dplyr::arrange(centroid_distance)
closest_cluster_pair <- cluster_pairwise_distances |>
  dplyr::slice(1)

write.csv(cluster_centroids,
          file.path(out_dir, "expression_cluster_centroid_profiles.csv"))
write.csv(cluster_pairwise_distances,
          file.path(out_dir, "expression_cluster_centroid_distances.csv"),
          row.names = FALSE)

label_df <- ggdendro::label(cluster_dendro_data) |>
  dplyr::left_join(
    cluster_size_map |>
      dplyr::select(cluster_short, n_genes),
    by = c("label" = "cluster_short")
  ) |>
  dplyr::mutate(
    label = paste0(label, "\n", n_genes, " genes"),
    label_x = dplyr::case_when(
      x == min(x) ~ x + 0.18,
      x == max(x) ~ x - 0.18,
      TRUE ~ x
    )
  )

cluster_cladogram_plot <- ggplot2::ggplot() +
  ggplot2::geom_segment(
    data = ggdendro::segment(cluster_dendro_data),
    ggplot2::aes(x = x, y = y, xend = xend, yend = yend),
    linewidth = 0.9,
    color = "#374151"
  ) +
  ggplot2::geom_label(
    data = label_df,
    ggplot2::aes(x = label_x, y = y - 0.18, label = label),
    size = 4.5,
    fontface = "bold",
    fill = "#F8FAFC",
    color = "#111827",
    label.size = 0.35
  ) +
  ggplot2::coord_cartesian(clip = "off") +
  ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0.2, 0.08))) +
  ggplot2::labs(
    title = "Expression-cluster cladogram",
    subtitle = paste0(
      "Closest pair: ", closest_cluster_pair$cluster_a, " and ",
      closest_cluster_pair$cluster_b, " (centroid distance = ",
      round(closest_cluster_pair$centroid_distance, 2), ")"
    ),
    x = NULL,
    y = "Euclidean distance between cluster centroids"
  ) +
  ggplot2::theme_classic(base_size = 14) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank(),
    plot.title = ggplot2::element_text(face = "bold"),
    plot.margin = grid::unit(c(0.5, 0.5, 1.1, 0.5), "lines")
  )

print(cluster_cladogram_plot)
ggplot2::ggsave(
  file.path(out_dir, "expression_cluster_centroid_cladogram.png"),
  cluster_cladogram_plot,
  width = 9,
  height = 6,
  dpi = 300
)

knitr::kable(
  cluster_pairwise_distances,
  digits = 3,
  caption = "Pairwise Euclidean distances between expression-cluster centroids"
)
```

## Full gene-level cladogram

This is the full gene-level row dendrogram for all genes in the diagnostic
heatmap. Gene labels are hidden because there are 3,620 leaves; the colored strip
under the tree marks the three interpreted expression clusters used above, with
the original branches 3 and 4 merged. The companion CSV records the leaf order,
the merged cluster, and the original branch assignment so individual genes can
still be looked up.

```{r full-gene-level-cladogram, fig.width=24, fig.height=8}
full_gene_hclust <- stats::hclust(stats::dist(heatmap_z_plot),
                                  method = "complete")
full_gene_dendro_data <- ggdendro::dendro_data(
  stats::as.dendrogram(full_gene_hclust),
  type = "rectangle"
)

full_gene_leaf_df <- ggdendro::label(full_gene_dendro_data) |>
  dplyr::rename(gene_id = label, leaf_x = x, leaf_y = y) |>
  dplyr::left_join(expression_cluster_assignments, by = "gene_id") |>
  dplyr::arrange(leaf_x) |>
  dplyr::mutate(full_cladogram_leaf_order = dplyr::row_number()) |>
  dplyr::select(full_cladogram_leaf_order, leaf_x, gene_id,
                expression_cluster_order, expression_cluster,
                original_pheatmap_cluster_id, heatmap_row_order,
                Description, DEG_context, phase_class, n_deg_contrasts,
                comparisons, directions, max_abs_log2FC, min_padj)

full_gene_order_check <- data.frame(
  check = "Full gene cladogram order matches exported pheatmap row order",
  value = identical(full_gene_leaf_df$gene_id,
                    expression_cluster_assignments$gene_id)
)

write.csv(
  full_gene_leaf_df,
  file.path(out_dir, "full_gene_level_expression_cladogram_leaf_order.csv"),
  row.names = FALSE
)
write.csv(
  full_gene_order_check,
  file.path(out_dir, "full_gene_level_expression_cladogram_order_check.csv"),
  row.names = FALSE
)

full_gene_cluster_spans <- full_gene_leaf_df |>
  dplyr::group_by(expression_cluster) |>
  dplyr::summarise(
    xmin = min(leaf_x) - 0.5,
    xmax = max(leaf_x) + 0.5,
    mid = (min(leaf_x) + max(leaf_x)) / 2,
    n_genes = dplyr::n(),
    .groups = "drop"
  ) |>
  dplyr::mutate(label = paste0(
    gsub("\n.*$", "", as.character(expression_cluster)),
    "\n", n_genes, " genes"
  ))

full_gene_max_height <- max(ggdendro::segment(full_gene_dendro_data)$y,
                            na.rm = TRUE)
full_gene_strip_ymin <- -full_gene_max_height * 0.14
full_gene_strip_ymax <- -full_gene_max_height * 0.08
full_gene_label_y <- -full_gene_max_height * 0.18

full_gene_level_cladogram_plot <- ggplot2::ggplot() +
  ggplot2::geom_segment(
    data = ggdendro::segment(full_gene_dendro_data),
    ggplot2::aes(x = x, y = y, xend = xend, yend = yend),
    linewidth = 0.13,
    color = "#263241",
    lineend = "square"
  ) +
  ggplot2::geom_rect(
    data = full_gene_cluster_spans,
    ggplot2::aes(xmin = xmin, xmax = xmax,
                 ymin = full_gene_strip_ymin, ymax = full_gene_strip_ymax,
                 fill = expression_cluster),
    inherit.aes = FALSE,
    color = "white",
    linewidth = 0.25
  ) +
  ggplot2::geom_text(
    data = full_gene_cluster_spans,
    ggplot2::aes(x = mid, y = full_gene_label_y, label = label),
    inherit.aes = FALSE,
    size = 4,
    fontface = "bold",
    lineheight = 0.9
  ) +
  ggplot2::scale_fill_manual(values = expression_cluster_colors,
                             guide = "none") +
  ggplot2::coord_cartesian(
    ylim = c(full_gene_max_height * -0.22, full_gene_max_height * 1.03),
    clip = "off"
  ) +
  ggplot2::labs(
    title = "Full gene-level cladogram for the diagnostic heatmap",
    subtitle = paste0(
      nrow(full_gene_leaf_df),
      " DEG leaves; colored strip marks the three interpreted expression clusters"
    ),
    x = NULL,
    y = "Euclidean distance between gene expression profiles"
  ) +
  ggplot2::theme_classic(base_size = 14) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank(),
    plot.title = ggplot2::element_text(face = "bold"),
    plot.margin = grid::unit(c(0.5, 0.5, 1.6, 0.5), "lines")
  )

print(full_gene_level_cladogram_plot)
ggplot2::ggsave(
  file.path(out_dir, "full_gene_level_expression_cladogram.png"),
  full_gene_level_cladogram_plot,
  width = 24,
  height = 8,
  dpi = 300
)
ggplot2::ggsave(
  file.path(out_dir, "full_gene_level_expression_cladogram.pdf"),
  full_gene_level_cladogram_plot,
  width = 24,
  height = 8
)

knitr::kable(
  full_gene_order_check,
  caption = "Sanity check for the full gene-level cladogram"
)
```

```{r expression-cluster-go-table}
if (nrow(cluster_go_significant) > 0) {
  DT::datatable(
    cluster_go_significant |>
      dplyr::arrange(expression_cluster, ontology, p.adjust) |>
      dplyr::select(expression_cluster, ontology, ID, Description, GeneRatio,
                    BgRatio, pvalue, p.adjust, qvalue, Count, geneID),
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 20, scrollX = TRUE),
    caption = "Cluster-specific significant GO terms"
  )
}
```

```{r expression-cluster-go-reverse-index}
if (nrow(cluster_term_gene_compact) > 0) {
  DT::datatable(
    cluster_term_gene_compact,
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 20, scrollX = TRUE),
    caption = "Reverse index for expression-cluster GO terms"
  )
}
```

## GO enrichment

```{r go-enrichment-plot, fig.width=18, fig.height=14}
go_plot_df <- go_enrichment |>
  dplyr::filter(p.adjust <= params$alpha) |>
  dplyr::group_by(background, ontology) |>
  dplyr::slice_min(p.adjust, n = params$top_terms_per_panel,
                   with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    neg_log10_padj = -log10(pmax(p.adjust, .Machine$double.xmin)),
    term_label = stringr::str_wrap(Description, 58),
    term_label = factor(term_label, levels = rev(unique(term_label))),
    ontology = factor(ontology, levels = names(go_ontology_colors))
  )

if (nrow(go_plot_df) > 0) {
  go_plot <- ggplot2::ggplot(
    go_plot_df,
    ggplot2::aes(x = neg_log10_padj, y = term_label)
  ) +
    ggplot2::geom_vline(
      xintercept = -log10(params$alpha),
      linetype = "dashed",
      color = "#6B7280",
      linewidth = 0.35
    ) +
    ggplot2::geom_point(
      ggplot2::aes(size = Count, color = ontology),
      alpha = 0.9
    ) +
    ggplot2::facet_grid(ontology ~ background,
                        scales = "free_y", space = "free_y") +
    ggplot2::scale_color_manual(values = go_ontology_colors, drop = FALSE) +
    ggplot2::scale_size_continuous(range = c(3.5, 11)) +
    ggplot2::labs(
      title = "GO enrichment in the circled heatmap branch",
      subtitle = "Foreground is the upper branch from the all-samples full DEG heatmap",
      x = "-log10 adjusted p-value",
      y = NULL,
      color = "GO ontology",
      size = "Cluster genes in term"
    ) +
    ggplot2::theme_classic(base_size = 14) +
    ggplot2::theme(
      axis.text.y = ggplot2::element_text(size = 11),
      strip.text = ggplot2::element_text(face = "bold", size = 12),
      legend.position = "bottom",
      plot.title = ggplot2::element_text(face = "bold")
    )
  print(go_plot)
  ggplot2::ggsave(file.path(out_dir, "circled_cluster_GO_enrichment.png"),
                  go_plot, width = 18, height = 14, dpi = 300)
} else {
  cat("No GO terms passed the adjusted p-value cutoff for this cluster.")
}
```

```{r go-enrichment-table}
if (nrow(go_enrichment) > 0) {
  DT::datatable(
    go_enrichment |>
      dplyr::arrange(p.adjust) |>
      dplyr::select(background, ontology, ID, Description, GeneRatio, BgRatio,
                    pvalue, p.adjust, qvalue, Count, geneID),
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 20, scrollX = TRUE),
    caption = "GO enrichment for the circled heatmap branch"
  )
}
```

## KEGG pathway enrichment

```{r kegg-enrichment-plot, fig.width=18, fig.height=12}
kegg_plot_df <- kegg_enrichment |>
  dplyr::filter(p.adjust <= params$alpha) |>
  dplyr::group_by(background, pathway_theme) |>
  dplyr::slice_min(p.adjust, n = params$top_terms_per_panel,
                   with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    neg_log10_padj = -log10(pmax(p.adjust, .Machine$double.xmin)),
    term_label = stringr::str_wrap(Description, 55),
    term_label = factor(term_label, levels = rev(unique(term_label))),
    pathway_scope = factor(pathway_scope, levels = names(kegg_scope_colors))
  )

if (nrow(kegg_plot_df) > 0) {
  kegg_plot <- ggplot2::ggplot(
    kegg_plot_df,
    ggplot2::aes(x = neg_log10_padj, y = term_label)
  ) +
    ggplot2::geom_vline(
      xintercept = -log10(params$alpha),
      linetype = "dashed",
      color = "#6B7280",
      linewidth = 0.35
    ) +
    ggplot2::geom_point(
      ggplot2::aes(size = Count),
      color = kegg_lollipop_color,
      alpha = 0.9
    ) +
    ggplot2::facet_grid(pathway_theme ~ background,
                        scales = "free_y", space = "free_y") +
    ggplot2::scale_size_continuous(range = c(4, 12)) +
    ggplot2::labs(
      title = "KEGG pathway enrichment in the circled heatmap branch",
      subtitle = "All KEGG pathway points are shown in brown; pathway scope remains in the exported tables",
      x = "-log10 adjusted p-value",
      y = NULL,
      size = "Cluster genes in pathway"
    ) +
    ggplot2::theme_classic(base_size = 14) +
    ggplot2::theme(
      axis.text.y = ggplot2::element_text(size = 12),
      strip.text = ggplot2::element_text(face = "bold", size = 11),
      legend.position = "bottom",
      plot.title = ggplot2::element_text(face = "bold")
    )
  print(kegg_plot)
  ggplot2::ggsave(file.path(out_dir, "circled_cluster_KEGG_enrichment.png"),
                  kegg_plot, width = 18, height = 12, dpi = 300)
} else {
  cat("No KEGG pathways passed the adjusted p-value cutoff for this cluster.")
}
```

```{r kegg-enrichment-table}
if (nrow(kegg_enrichment) > 0) {
  DT::datatable(
    kegg_enrichment |>
      dplyr::arrange(p.adjust) |>
      dplyr::select(background, pathway_theme, pathway_scope, ID, Description,
                    GeneRatio, BgRatio, pvalue, p.adjust, qvalue, Count,
                    geneID),
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 20, scrollX = TRUE),
    caption = "KEGG pathway enrichment for the circled heatmap branch"
  )
}
```

## Reverse term-to-gene index

```{r reverse-term-gene-index}
gene_annotation <- circled_cluster |>
  dplyr::select(gene_id, DEG_context, phase_class, n_deg_contrasts,
                comparisons, directions, max_abs_log2FC, min_padj,
                max_abs_sample_z, sample_with_max_abs_z,
                sample_deviation_direction) |>
  dplyr::distinct(gene_id, .keep_all = TRUE) |>
  dplyr::left_join(
    gff_annotation |> dplyr::select(gene_id, Description, gene_biotype),
    by = "gene_id"
  ) |>
  dplyr::mutate(
    Description = dplyr::coalesce(Description, "no GFF description"),
    gene_biotype = dplyr::coalesce(gene_biotype, "unknown")
  )

term_gene_long <- dplyr::bind_rows(
  expand_term_genes(go_enrichment, gene_annotation) |>
    dplyr::mutate(database = "GO"),
  expand_term_genes(kegg_enrichment, gene_annotation) |>
    dplyr::mutate(database = "KEGG_pathway")
) |>
  dplyr::arrange(database, background, p.adjust, term_id, gene_id)

term_gene_compact <- term_gene_long |>
  dplyr::group_by(database, background, term_id, term_name, GeneRatio, BgRatio,
                  p.adjust, Count) |>
  dplyr::summarise(
    DEG_gene_IDs = paste(sort(unique(gene_id)), collapse = "; "),
    DEG_gene_descriptions = paste(sort(unique(Description)), collapse = "; "),
    DEG_contexts = paste(sort(unique(DEG_context)), collapse = "; "),
    strongest_deviation_samples = paste(
      sort(unique(as.character(sample_with_max_abs_z))),
      collapse = "; "
    ),
    .groups = "drop"
  ) |>
  dplyr::arrange(database, background, p.adjust)

write.csv(term_gene_long,
          file.path(out_dir,
                    "circled_cluster_enriched_term_gene_index_long.csv"),
          row.names = FALSE)
write.csv(term_gene_compact,
          file.path(out_dir,
                    "circled_cluster_enriched_term_gene_index_compact.csv"),
          row.names = FALSE)

if (nrow(term_gene_compact) > 0) {
  DT::datatable(
    term_gene_compact,
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 20, scrollX = TRUE),
    caption = "Reverse index: enriched terms and the circled-cluster genes contributing to them"
  )
} else {
  cat("No significant enriched terms were available for the reverse gene index.")
}
```

## Cluster gene table

```{r cluster-gene-table}
cluster_gene_display <- circled_cluster |>
  dplyr::left_join(
    gff_annotation |> dplyr::select(gene_id, GFF_Description = Description,
                                    gene_biotype),
    by = "gene_id"
  ) |>
  dplyr::mutate(
    GFF_Description = dplyr::coalesce(GFF_Description, "no GFF description"),
    gene_biotype = dplyr::coalesce(gene_biotype, "unknown")
  ) |>
  dplyr::select(
    heatmap_row_order, gene_id, GFF_Description, gene_biotype, DEG_context,
    phase_class, n_deg_contrasts, comparisons, directions, max_abs_log2FC,
    min_padj, max_abs_sample_z, sample_with_max_abs_z,
    sample_deviation_direction
  )

write.csv(cluster_gene_display,
          file.path(out_dir, "circled_cluster_gene_table_with_gff_descriptions.csv"),
          row.names = FALSE)

DT::datatable(
  cluster_gene_display,
  rownames = FALSE,
  filter = "top",
  options = list(pageLength = 25, scrollX = TRUE),
  caption = "Genes in the circled heatmap branch"
)
```
