---
title: "Main-chromosome DEG expression clusters and enrichment"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  latest_results_root: "output/rmd_runs"
  counts_file: "output/rmd_runs/transcript-exon-primary-deseq2_20260806-004620/host_counts_no_rrna.tsv"
  mapping_summary_file: "output/runs/host_pathogen_dual_20260727_000952/05-mapping-comparison/competitive_mapping_summary.tsv"
  placement_file: "output/runs/host_pathogen_dual_20260727_000952/local-analysis/host-mapping-audit_20260729-163928/unified-model/audit-summary/host_gene_sequence_placement.tsv"
  rrna_file: "data/excluded_loci/gregaria_rrna_list.txt"
  metadata_file: "data/metadata/mehreen_metadata_fixed.txt"
  timecourse_fatbody_reference_dir: "data/external/timecourse_fatbody_reference"
  timecourse_head_thorax_counts_dir: "data/raw_read_counts/03-gregaria-DESeq2"
  eggnog_file: "data/GO_Annotations/GCF_023897955.1_iqSchGreg1.2_Arthopoda_one2one.emapper.annotations"
  gff_file: "data/reference/GCF_023897955.1_iqSchGreg1.2_genomic.gff"
  min_count: 10
  min_samples: 3
  alpha: 0.05
  top_terms_per_panel: 18
  output_tag: "main-chromosome-cluster-enrichment"
---

## Purpose

::: {.walkthrough-grid}
::: {.walkthrough-step}
**Step 1 - Cluster expression**

Group main-chromosome DEGs by their row-scaled VST profiles across all samples.
:::
::: {.walkthrough-step}
**Step 2 - Test functions**

Run GO and eukaryote-inclusive KEGG enrichment within each expression cluster.
:::
::: {.walkthrough-step}
**Step 3 - Interpret patterns**

Read expression, DEG context, fungal signal, and enrichment as one figure.
:::
:::

This page rebuilds the gene-level expression clusters using only significant
main-chromosome fat body DEGs from the joint all-samples model. The current
publication composite retains the earlier three-block display for continuity,
but that grouping is provisional. The full unpartitioned gene dendrogram and
its exactly aligned heatmap are shown below so that new boundaries can be chosen
from the current chromosome-only data rather than inherited from the older run.

The foreground is the exported gene list for that circled branch. The enrichment
is run in two ways:

- against all expressed main-chromosome genes in the rRNA-filtered count matrix;
- against all genes in the full all-samples fat body DEG heatmap.

The first test asks what functions are over-represented compared with the whole
expressed transcriptome. The second test asks what makes this heatmap branch
different from the other fat body DEGs.

For GO enrichment, only terms that resolve to Biological process, Cellular
component, or Molecular function in the local GO database are kept in the final
tables and plots. Unresolved GO identifiers are omitted here because they do not
provide an interpretable functional label.

**Current cluster decision:** [Open the searchable unmerged four-branch
review](#unmerged-four-branch-review) before deciding which branches should be
merged in the publication figure.

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c("clusterProfiler", "DESeq2", "dplyr", "readr", "tidyr", "stringr",
              "ggplot2", "DT", "GO.db", "AnnotationDbi", "tibble",
              "scales", "SummarizedExperiment", "cluster", "png")
required <- c(required, "patchwork", "ggdendro")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

library(ggplot2)
library(patchwork)

figure3_path_ancestors <- function(path) {
  path <- normalizePath(path, mustWork = FALSE)
  ancestors <- path
  repeat {
    parent <- dirname(path)
    if (identical(parent, path)) break
    ancestors <- c(ancestors, parent)
    path <- parent
  }
  ancestors
}

is_figure3_project <- function(path) {
  file.exists(file.path(path, "analysis", "26-circled-cluster-enrichment.Rmd")) &&
    dir.exists(file.path(path, "data")) &&
    dir.exists(file.path(path, "output"))
}

find_figure3_project <- function() {
  starts <- character()

  explicit_root <- Sys.getenv("GREGARIA_DIET_PROJECT_DIR", unset = "")
  if (nzchar(explicit_root)) starts <- c(starts, explicit_root)

  input_file <- tryCatch(knitr::current_input(dir = TRUE),
                         error = function(e) "")
  if (length(input_file) == 1 && nzchar(input_file)) {
    starts <- c(starts, dirname(input_file))
  }

  if (requireNamespace("rstudioapi", quietly = TRUE) &&
      rstudioapi::isAvailable()) {
    active_file <- tryCatch(
      rstudioapi::getSourceEditorContext()$path,
      error = function(e) ""
    )
    if (length(active_file) == 1 && nzchar(active_file)) {
      starts <- c(starts, dirname(active_file))
    }
  }

  starts <- unique(c(starts, getwd()))
  candidates <- unique(unlist(
    lapply(starts, figure3_path_ancestors),
    use.names = FALSE
  ))
  matches <- candidates[vapply(candidates, is_figure3_project, logical(1))]

  if (length(matches) == 0) {
    stop(
      "Could not locate gregaria-diet-infection-interaction. ",
      "Open this Rmd from that project, set the working directory to the ",
      "repository root, or set GREGARIA_DIET_PROJECT_DIR explicitly."
    )
  }

  normalizePath(matches[1], mustWork = TRUE)
}

project_dir <- find_figure3_project()
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

results_root <- file.path(project_dir, params$latest_results_root)
counts_file <- file.path(project_dir, params$counts_file)
mapping_summary_file <- file.path(project_dir, params$mapping_summary_file)
placement_file <- file.path(project_dir, params$placement_file)
rrna_file <- file.path(project_dir, params$rrna_file)
metadata_file <- file.path(project_dir, params$metadata_file)
timecourse_fatbody_dir <- file.path(
  project_dir,
  params$timecourse_fatbody_reference_dir
)
timecourse_head_thorax_dir <- file.path(
  project_dir,
  params$timecourse_head_thorax_counts_dir
)
eggnog_file <- file.path(project_dir, params$eggnog_file)
gff_file <- file.path(project_dir, params$gff_file)
figure3_locust_path <- file.path(
  project_dir,
  "analysis", "assets", "figure3", "schistocerca_pathway_base.png"
)
if (!file.exists(figure3_locust_path)) {
  stop("Missing Figure 3 locust illustration: ", figure3_locust_path)
}

current_outliers <- c("1007", "1036", "1037", "1039")

go_ontology_colors <- c(
  "Biological process" = "#111111",
  "Cellular component" = "#8B5A2B",
  "Molecular function" = "#8C8C8C"
)

kegg_scope_colors <- c(
  "Eukaryote/metazoan-compatible" = "#1E88E5",
  "Core metabolism or cross-domain" = "#2F8F5B",
  "Broad KEGG reference map" = "#4B5563",
  "Microbial/prokaryote-associated label" = "#E41A1C",
  "Reference disease pathway" = "#7B4FA3",
  "Unclassified KEGG reference label" = "#F59E0B"
)
kegg_lollipop_color <- "#8B5A2B"
```

## Helper functions

```{r helper-functions}
latest_dir <- function(pattern) {
  dirs <- list.dirs(results_root, recursive = FALSE, full.names = TRUE)
  dirs <- dirs[grepl(pattern, basename(dirs))]
  if (length(dirs) == 0) return(NA_character_)
  dirs[which.max(file.info(dirs)$mtime)]
}

resolve_cluster_file <- function(path_param) {
  explicit_path <- if (is.null(path_param) || !nzchar(path_param)) {
    NA_character_
  } else if (grepl("^/", path_param)) {
    path_param
  } else {
    file.path(project_dir, path_param)
  }
  if (!is.na(explicit_path) && file.exists(explicit_path)) {
    return(explicit_path)
  }

  phase_dir <- latest_dir("^phase-signature-overlap_")
  if (is.na(phase_dir)) return(explicit_path)

  candidate <- file.path(
    phase_dir,
    "circled_upper_cluster_audit",
    "circled_upper_cluster_genes.csv"
  )
  candidate
}

read_emapper <- function(path) {
  cols <- c(
    "query", "seed_ortholog", "evalue", "score", "eggNOG_OGs",
    "max_annot_lvl", "COG_category", "Description", "Preferred_name",
    "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
    "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC", "CAZy",
    "BiGG_Reaction", "PFAMs"
  )
  tab <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", fill = TRUE)
  names(tab) <- cols[seq_len(ncol(tab))]
  tab |>
    dplyr::mutate(protein_id = stringr::str_extract(query, "XP_[0-9]+[.][0-9]+"))
}

parse_cds_gene_bridge <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  gff |>
    dplyr::filter(type == "CDS") |>
    dplyr::transmute(
      protein_id = stringr::str_match(attributes, "protein_id=([^;]+)")[, 2],
      gene_id = stringr::str_match(attributes, "gene=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(protein_id), !is.na(gene_id)) |>
    dplyr::distinct()
}

parse_gff_gene_annotation <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  gff |>
    dplyr::filter(type == "gene") |>
    dplyr::transmute(
      gene_id = dplyr::coalesce(
        stringr::str_match(attributes, "gene=([^;]+)")[, 2],
        stringr::str_match(attributes, "Name=([^;]+)")[, 2]
      ),
      Description = stringr::str_match(attributes, "description=([^;]+)")[, 2],
      gene_biotype = stringr::str_match(attributes, "gene_biotype=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(gene_id)) |>
    dplyr::mutate(
      Description = dplyr::if_else(
        is.na(Description) | Description == "",
        "no GFF description",
        utils::URLdecode(Description)
      ),
      gene_biotype = dplyr::coalesce(gene_biotype, "unknown")
    ) |>
    dplyr::distinct(gene_id, .keep_all = TRUE)
}

read_metadata <- function(path) {
  tab <- read.delim(path, check.names = FALSE)
  tab <- tab[, !is.na(names(tab)) & nzchar(names(tab)), drop = FALSE]
  tab$label <- as.character(tab$label)
  tab$Diet <- factor(tab$Diet, levels = c("33", "50", "83"))
  tab$Treatment <- factor(tab$Treatment, levels = c("Infected", "Control"))
  tab
}

read_featurecounts_gene_counts <- function(path, sample_id) {
  tab <- read.delim(path, comment.char = "#", check.names = FALSE)
  if ("Geneid" %in% names(tab)) {
    count_col <- names(tab)[ncol(tab)]
    out <- tibble::tibble(
      gene_id = as.character(tab[["Geneid"]]),
      count = as.integer(tab[[count_col]])
    )
  } else {
    tab <- read.delim(path, comment.char = "#", header = FALSE,
                      check.names = FALSE)
    if (ncol(tab) < 2) {
      stop("Count file must contain at least gene ID and count columns: ", path)
    }
    out <- tibble::tibble(
      gene_id = as.character(tab[[1]]),
      count = as.integer(tab[[ncol(tab)]])
    )
  }
  out <- out |>
    dplyr::filter(!is.na(gene_id), nzchar(gene_id)) |>
    dplyr::group_by(gene_id) |>
    dplyr::summarise(count = sum(count, na.rm = TRUE), .groups = "drop")
  names(out)[2] <- sample_id
  out
}

parse_head_thorax_reference_metadata <- function(reference_dir) {
  files <- list.files(
    reference_dir,
    pattern = "^SGRE-(HEAD|THOX)-(CRD|ISO)-[0-9]+_MERGE_featurecounts[.]txt$",
    full.names = TRUE
  )
  if (length(files) == 0) {
    return(tibble::tibble(
      sample_id = character(),
      count_file = character(),
      tissue = character(),
      phase = character(),
      rearing_condition = character(),
      replicate = integer(),
      source = character(),
      display_group = character(),
      reference_role = character(),
      sample_label = character(),
      sample_block = character()
    ))
  }

  parsed <- stringr::str_match(
    basename(files),
    "^SGRE-(HEAD|THOX)-(CRD|ISO)-([0-9]+)_MERGE_featurecounts[.]txt$"
  )
  tibble::tibble(
    sample_id = stringr::str_remove(basename(files), "_featurecounts[.]txt$"),
    count_file = files,
    tissue_code = parsed[, 2],
    rearing_code = parsed[, 3],
    replicate = as.integer(parsed[, 4])
  ) |>
    dplyr::filter(!is.na(tissue_code), !is.na(rearing_code)) |>
    dplyr::mutate(
      tissue = dplyr::case_when(
        tissue_code == "HEAD" ~ "Head",
        tissue_code == "THOX" ~ "Thorax",
        TRUE ~ tissue_code
      ),
      phase = dplyr::case_when(
        rearing_code == "CRD" ~ "Gregarious",
        rearing_code == "ISO" ~ "Solitarious",
        TRUE ~ rearing_code
      ),
      rearing_condition = dplyr::case_when(
        rearing_code == "CRD" ~ "Crowded",
        rearing_code == "ISO" ~ "Isolated",
        TRUE ~ rearing_code
      ),
      source = "Time-course head/thorax phase reference",
      display_group = paste0(
        tissue, ": ",
        dplyr::if_else(rearing_condition == "Crowded",
                       "crowded/gregarious", "isolated/solitarious")
      ),
      reference_role = paste(tissue, tolower(rearing_condition),
                             "rep", replicate),
      sample_label = paste0(substr(tissue, 1, 1), "-",
                            rearing_code, "-", replicate),
      sample_block = paste(tissue, "reference")
    ) |>
    dplyr::arrange(tissue, rearing_condition, replicate)
}

first_nonempty <- function(x, fallback = "") {
  x <- x[!is.na(x) & x != "" & x != "-"]
  if (length(x) == 0) fallback else x[1]
}

build_term_maps_from_eggnog <- function() {
  counts_raw <- read.csv(counts_file, check.names = FALSE)
  gene_universe <- unique(counts_raw[[1]])
  annotation <- read_emapper(eggnog_file) |>
    dplyr::left_join(parse_cds_gene_bridge(gff_file), by = "protein_id") |>
    dplyr::filter(!is.na(gene_id), gene_id %in% gene_universe) |>
    dplyr::group_by(gene_id) |>
    dplyr::summarise(
      GOs = first_nonempty(GOs, ""),
      KEGG_Pathway = first_nonempty(KEGG_Pathway, ""),
      .groups = "drop"
    )

  term2gene_go <- annotation |>
    dplyr::select(gene_id, GOs) |>
    tidyr::separate_rows(GOs, sep = ",") |>
    dplyr::filter(grepl("^GO:", GOs)) |>
    dplyr::transmute(term = GOs, gene = gene_id) |>
    dplyr::distinct()

  go_names <- AnnotationDbi::mapIds(
    GO.db::GO.db,
    keys = unique(term2gene_go$term),
    keytype = "GOID",
    column = "TERM",
    multiVals = "first"
  )
  go_ontology <- AnnotationDbi::mapIds(
    GO.db::GO.db,
    keys = unique(term2gene_go$term),
    keytype = "GOID",
    column = "ONTOLOGY",
    multiVals = "first"
  )
  term2name_go <- data.frame(
    term = names(go_names),
    name = unname(go_names),
    ontology = unname(go_ontology)
  ) |>
    dplyr::filter(!is.na(name))

  term2gene_kegg <- annotation |>
    dplyr::select(gene_id, KEGG_Pathway) |>
    tidyr::separate_rows(KEGG_Pathway, sep = ",") |>
    dplyr::filter(grepl("^ko[0-9]+$", KEGG_Pathway)) |>
    dplyr::transmute(term = KEGG_Pathway, gene = gene_id) |>
    dplyr::distinct()

  term2name_kegg <- term2gene_kegg |>
    dplyr::distinct(term) |>
    dplyr::mutate(
      name = term,
      pathway_theme = "Other KEGG pathway",
      pathway_scope = "Unclassified KEGG reference label"
    )

  list(
    term2gene_go = term2gene_go,
    term2name_go = term2name_go,
    term2gene_kegg = term2gene_kegg,
    term2name_kegg = term2name_kegg
  )
}

load_term_maps <- function() {
  go_dir <- latest_dir("^go-term-enrichment_")
  expected <- file.path(
    go_dir,
    c("eggNOG_GO_TERM2GENE.csv", "eggNOG_GO_TERM2NAME.csv",
      "eggNOG_KEGG_pathway_TERM2GENE.csv",
      "eggNOG_KEGG_pathway_TERM2NAME.csv")
  )
  if (!is.na(go_dir) && all(file.exists(expected))) {
    return(list(
      term2gene_go = read.csv(expected[1], check.names = FALSE),
      term2name_go = read.csv(expected[2], check.names = FALSE),
      term2gene_kegg = read.csv(expected[3], check.names = FALSE),
      term2name_kegg = read.csv(expected[4], check.names = FALSE)
    ))
  }
  build_term_maps_from_eggnog()
}

run_enrichment <- function(genes, universe, term2gene, term2name, database,
                           background_label) {
  term2gene <- term2gene |>
    dplyr::filter(gene %in% universe) |>
    dplyr::distinct(term, gene)
  term2name_min <- term2name |>
    dplyr::distinct(term, .keep_all = TRUE) |>
    dplyr::select(term, name)

  tested_universe <- intersect(universe, term2gene$gene)
  tested_genes <- intersect(genes, tested_universe)
  if (length(tested_genes) < 5 || length(tested_universe) < 10) {
    return(data.frame())
  }

  res <- clusterProfiler::enricher(
    gene = tested_genes,
    universe = tested_universe,
    TERM2GENE = term2gene,
    TERM2NAME = term2name_min,
    pvalueCutoff = 1,
    pAdjustMethod = "BH",
    qvalueCutoff = 1,
    minGSSize = 5,
    maxGSSize = 5000
  )
  if (is.null(res)) return(data.frame())
  out <- as.data.frame(res)
  if (nrow(out) == 0) return(data.frame())
  out |>
    dplyr::mutate(
      database = database,
      background = background_label,
      foreground_genes_tested = length(tested_genes),
      background_genes_tested = length(tested_universe)
    )
}

expand_term_genes <- function(tab, gene_annotation) {
  if (nrow(tab) == 0) {
    return(tibble::tibble())
  }
  tab |>
    dplyr::filter(p.adjust <= params$alpha) |>
    dplyr::select(database, background, ID, Description, GeneRatio, BgRatio,
                  pvalue, p.adjust, qvalue, Count, geneID) |>
    tidyr::separate_rows(geneID, sep = "/") |>
    dplyr::rename(term_id = ID, term_name = Description, gene_id = geneID) |>
    dplyr::left_join(gene_annotation, by = "gene_id")
}
```

## Inputs

```{r inputs}
primary_run_dir <- latest_dir("^placed-unplaced-scaffold-analysis_")
if (is.na(primary_run_dir)) {
  stop("Build the placed/unplaced audit page before this cluster page.")
}
primary_result_file <- file.path(
  primary_run_dir,
  "placed_unplaced_all_deseq2_results.csv"
)

input_files <- data.frame(
  input = c("Joint all-contrast DESeq2 results",
            "Sample metadata", "Fresh competitive host count matrix",
            "Competitive mapping summary",
            "Official gene placement", "Annotated rRNA exclusion list",
            "eggNOG annotation", "GFF gene descriptions"),
  path = c(primary_result_file, metadata_file, counts_file,
           mapping_summary_file,
           placement_file, rrna_file,
           eggnog_file, gff_file),
  present = c(file.exists(primary_result_file), file.exists(metadata_file),
              file.exists(counts_file),
              file.exists(mapping_summary_file),
              file.exists(placement_file), file.exists(rrna_file),
              file.exists(eggnog_file), file.exists(gff_file))
)
knitr::kable(input_files, caption = "Inputs for main-chromosome cluster enrichment")
if (!all(input_files$present)) {
  stop("One or more required inputs are missing.")
}
```

```{r build-input-sets}
gff_annotation <- parse_gff_gene_annotation(gff_file)
counts_table <- readr::read_tsv(counts_file, show_col_types = FALSE)
counts <- as.matrix(counts_table[, setdiff(names(counts_table), "gene_id"), drop = FALSE])
rownames(counts) <- counts_table$gene_id
storage.mode(counts) <- "integer"

placement <- readr::read_tsv(placement_file, show_col_types = FALSE) |>
  dplyr::distinct(gene_id, .keep_all = TRUE)
rrna_ids <- trimws(readLines(rrna_file, warn = FALSE))
rrna_ids <- rrna_ids[nzchar(rrna_ids) & !startsWith(rrna_ids, "#")]
main_chromosome_ids <- placement$gene_id[
  placement$sequence_class == "Nuclear chromosome" &
    !placement$gene_id %in% rrna_ids
]
counts <- counts[intersect(rownames(counts), main_chromosome_ids), , drop = FALSE]
expressed_universe <- rownames(counts)[
  rowSums(counts >= params$min_count, na.rm = TRUE) >= params$min_samples
]
counts <- counts[expressed_universe, , drop = FALSE]

metadata <- read_metadata(metadata_file) |>
  dplyr::filter(label %in% colnames(counts)) |>
  dplyr::arrange(match(label, colnames(counts))) |>
  dplyr::mutate(
    group = factor(
      paste0("diet", Diet, "_", tolower(as.character(Treatment))),
      levels = c(
        "diet33_control", "diet50_control", "diet83_control",
        "diet33_infected", "diet50_infected", "diet83_infected"
      )
    )
  )
if (!setequal(metadata$label, colnames(counts))) {
  stop("Metadata and fresh competitive counts have different sample IDs.")
}
counts <- counts[, metadata$label, drop = FALSE]

dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = counts,
  colData = data.frame(
    group = metadata$group,
    row.names = metadata$label
  ),
  design = ~ 0 + group
)
dds <- DESeq2::DESeq(
  dds,
  minReplicatesForReplace = Inf,
  quiet = TRUE
)
vst_matrix <- SummarizedExperiment::assay(DESeq2::vst(dds, blind = TRUE))

primary_results <- readr::read_csv(primary_result_file, show_col_types = FALSE) |>
  dplyr::filter(
    mapping_branch == "Competitive host",
    analysis_scope == "Placed/host-assigned",
    significant,
    gene_id %in% main_chromosome_ids,
    gene_id %in% rownames(vst_matrix)
  )
if (nrow(primary_results) == 0) {
  stop("No significant main-chromosome DEGs were found in the primary results.")
}

context_membership <- primary_results |>
  dplyr::mutate(
    context_family = dplyr::case_when(
      contrast_family == "Diet among controls" ~ "Diet in controls",
      contrast_family == "Diet among infected" ~ "Diet in infected",
      contrast_family %in% c(
        "Infection within diet",
        "Average infection response",
        "Diet by infection interaction"
      ) ~ "Infection within diet",
      TRUE ~ "Other fat body DEG"
    )
  ) |>
  dplyr::distinct(gene_id, context_family)

gene_context <- context_membership |>
  dplyr::group_by(gene_id) |>
  dplyr::summarise(
    n_context_families = dplyr::n_distinct(context_family),
    single_context = dplyr::first(context_family),
    DEG_context = dplyr::case_when(
      n_context_families > 1 ~ "Infection + diet",
      TRUE ~ single_context
    ),
    .groups = "drop"
  )

full_heatmap_genes <- primary_results |>
  dplyr::group_by(gene_id) |>
  dplyr::summarise(
    n_deg_contrasts = dplyr::n_distinct(contrast_id),
    comparisons = paste(sort(unique(contrast_id)), collapse = "; "),
    directions = paste(sort(unique(higher_in)), collapse = "; "),
    max_abs_log2FC = max(abs(log2FoldChange), na.rm = TRUE),
    min_padj = min(padj, na.rm = TRUE),
    phase_class = "Not evaluated in this chromosome-only figure",
    .groups = "drop"
  ) |>
  dplyr::left_join(gene_context, by = "gene_id") |>
  dplyr::left_join(
    gff_annotation |> dplyr::select(gene_id, Description),
    by = "gene_id"
  ) |>
  dplyr::mutate(
    Description = dplyr::coalesce(Description, "no GFF description")
  )

full_heatmap_z_input <- vst_matrix[full_heatmap_genes$gene_id, , drop = FALSE]
full_heatmap_z_input <- t(scale(t(full_heatmap_z_input)))
keep_finite <- apply(full_heatmap_z_input, 1, function(x) all(is.finite(x)))
full_heatmap_z_input <- full_heatmap_z_input[keep_finite, , drop = FALSE]
full_heatmap_genes <- full_heatmap_genes |>
  dplyr::filter(gene_id %in% rownames(full_heatmap_z_input))

sample_max_index <- max.col(abs(full_heatmap_z_input), ties.method = "first")
sample_diagnostics <- tibble::tibble(
  gene_id = rownames(full_heatmap_z_input),
  max_abs_sample_z = apply(abs(full_heatmap_z_input), 1, max),
  sample_with_max_abs_z = colnames(full_heatmap_z_input)[sample_max_index],
  sample_deviation_direction = dplyr::if_else(
    full_heatmap_z_input[cbind(seq_len(nrow(full_heatmap_z_input)), sample_max_index)] > 0,
    "Higher than gene mean",
    "Lower than gene mean"
  )
)
full_heatmap_genes <- full_heatmap_genes |>
  dplyr::left_join(sample_diagnostics, by = "gene_id")

row_distance <- stats::dist(full_heatmap_z_input)
row_tree <- stats::hclust(row_distance, method = "complete")
ordered_genes <- rownames(full_heatmap_z_input)[row_tree$order]
raw_cluster <- stats::cutree(row_tree, k = 4)[ordered_genes]
cluster_runs <- rle(raw_cluster)
row_cluster_spans <- data.frame(
  cluster_id = cluster_runs$values,
  start = cumsum(c(1L, head(cluster_runs$lengths, -1L))),
  end = cumsum(cluster_runs$lengths),
  n = cluster_runs$lengths
)
if (nrow(row_cluster_spans) != 4) {
  stop("The four-cluster tree did not produce four contiguous heatmap branches.")
}

full_heatmap_genes <- full_heatmap_genes |>
  dplyr::mutate(heatmap_row_order = match(gene_id, ordered_genes)) |>
  dplyr::arrange(heatmap_row_order)
full_heatmap_z_input <- full_heatmap_z_input[ordered_genes, , drop = FALSE]

full_heatmap_gene_file <- file.path(
  out_dir,
  "full_main_chromosome_deg_diagnostic_heatmap_gene_order.csv"
)
full_heatmap_z_file <- file.path(
  out_dir,
  "full_main_chromosome_deg_diagnostic_z_vst_matrix.csv.gz"
)
row_cluster_span_file <- file.path(out_dir, "row_cluster_spans_k4.csv")
write.csv(full_heatmap_genes, full_heatmap_gene_file, row.names = FALSE)
write.csv(full_heatmap_z_input, gzfile(full_heatmap_z_file))
write.csv(row_cluster_spans, row_cluster_span_file, row.names = FALSE)

cluster_genes <- ordered_genes[raw_cluster == cluster_runs$values[[1]]]
circled_cluster <- full_heatmap_genes |>
  dplyr::filter(gene_id %in% cluster_genes)
full_deg_universe <- unique(full_heatmap_genes$gene_id)

set_summary <- data.frame(
  set = c("First main-chromosome expression branch",
          "All expressed main-chromosome genes",
          "Union of all significant main-chromosome fat body DEGs"),
  n_genes = c(length(cluster_genes), length(expressed_universe),
              length(full_deg_universe))
)

write.csv(set_summary, file.path(out_dir, "circled_cluster_set_summary.csv"),
          row.names = FALSE)
knitr::kable(set_summary, caption = "Gene sets used for enrichment")
```

## Enrichment tests

```{r run-enrichment}
maps <- load_term_maps()

maps$term2gene_go <- maps$term2gene_go |>
  dplyr::filter(gene %in% expressed_universe)
maps$term2gene_kegg <- maps$term2gene_kegg |>
  dplyr::filter(gene %in% expressed_universe)

backgrounds <- list(
  "Expressed rRNA-filtered genes" = expressed_universe,
  "All fat body DEGs in heatmap" = intersect(full_deg_universe, expressed_universe)
)

go_enrichment <- dplyr::bind_rows(lapply(names(backgrounds), function(bg) {
  run_enrichment(
    genes = cluster_genes,
    universe = backgrounds[[bg]],
    term2gene = maps$term2gene_go,
    term2name = maps$term2name_go,
    database = "GO",
    background_label = bg
  )
}))

kegg_enrichment <- dplyr::bind_rows(lapply(names(backgrounds), function(bg) {
  run_enrichment(
    genes = cluster_genes,
    universe = backgrounds[[bg]],
    term2gene = maps$term2gene_kegg,
    term2name = maps$term2name_kegg,
    database = "KEGG_pathway",
    background_label = bg
  )
}))

if (nrow(go_enrichment) > 0) {
  go_enrichment <- go_enrichment |>
    dplyr::left_join(
      maps$term2name_go |>
        dplyr::distinct(term, .keep_all = TRUE) |>
        dplyr::transmute(
          ID = term,
          ontology = dplyr::recode(
            ontology,
            BP = "Biological process",
            CC = "Cellular component",
            MF = "Molecular function",
            .default = NA_character_
          )
        ),
      by = "ID"
    ) |>
    dplyr::filter(!is.na(ontology))
}

if (nrow(kegg_enrichment) > 0) {
  kegg_enrichment <- kegg_enrichment |>
    dplyr::left_join(
      maps$term2name_kegg |>
        dplyr::distinct(term, .keep_all = TRUE) |>
        dplyr::transmute(ID = term, pathway_theme, pathway_scope),
      by = "ID"
    ) |>
    dplyr::mutate(
      pathway_theme = dplyr::coalesce(pathway_theme, "Other KEGG pathway"),
      pathway_scope = dplyr::coalesce(pathway_scope,
                                      "Unclassified KEGG reference label")
    )
}

all_enrichment <- dplyr::bind_rows(go_enrichment, kegg_enrichment)
significant_enrichment <- all_enrichment |>
  dplyr::filter(p.adjust <= params$alpha) |>
  dplyr::arrange(database, background, p.adjust)

write.csv(go_enrichment,
          file.path(out_dir, "circled_cluster_GO_enrichment_all_terms.csv"),
          row.names = FALSE)
write.csv(kegg_enrichment,
          file.path(out_dir, "circled_cluster_KEGG_enrichment_all_terms.csv"),
          row.names = FALSE)
write.csv(all_enrichment,
          file.path(out_dir, "circled_cluster_enrichment_all_terms.csv"),
          row.names = FALSE)
write.csv(significant_enrichment,
          file.path(out_dir, "circled_cluster_enrichment_significant_terms.csv"),
          row.names = FALSE)

enrichment_count_summary <- significant_enrichment |>
  dplyr::count(database, background, name = "n_significant_terms") |>
  tidyr::complete(database = c("GO", "KEGG_pathway"),
                  background = names(backgrounds),
                  fill = list(n_significant_terms = 0)) |>
  dplyr::arrange(database, background)

write.csv(enrichment_count_summary,
          file.path(out_dir, "circled_cluster_enrichment_count_summary.csv"),
          row.names = FALSE)
knitr::kable(enrichment_count_summary,
             caption = "Significant enriched terms in the circled cluster")
```

## Expression-cluster landscape

The full diagnostic heatmap is first divided into four row branches from the
current gene-level dendrogram. The three-block display below keeps the earlier
publication layout by provisionally combining branches 3 and 4; it should be
treated as a visual review, not as the final clustering decision. The later
[full heatmap and gene-level cladogram](#Cluster_cladogram) preserve the original
tree continuously and provide the appropriate view for selecting new
boundaries. The lollipop panels summarize significant GO terms using all fat
body DEGs in the heatmap as the enrichment background.

```{r expression-cluster-landscape, fig.width=24, fig.height=26}
full_heatmap_z <- read.csv(gzfile(full_heatmap_z_file), check.names = FALSE,
                           row.names = 1)
full_heatmap_z <- as.matrix(full_heatmap_z)
storage.mode(full_heatmap_z) <- "numeric"

raw_sample_order <- colnames(full_heatmap_z)
sample_metadata <- read_metadata(metadata_file) |>
  dplyr::filter(label %in% raw_sample_order) |>
  dplyr::mutate(
    original_sample_index = match(label, raw_sample_order),
    Treatment_plot = factor(as.character(Treatment),
                            levels = c("Control", "Infected")),
    Diet_plot = factor(as.character(Diet), levels = c("33", "50", "83"))
  ) |>
  dplyr::arrange(Treatment_plot, Diet_plot, original_sample_index)
sample_metadata <- sample_metadata |>
  dplyr::mutate(
    sample_index = dplyr::row_number(),
    sample_group = factor(
      paste(Treatment_plot, Diet_plot),
      levels = c("Control 33", "Control 50", "Control 83",
                 "Infected 33", "Infected 50", "Infected 83")
    ),
    outlier_status = dplyr::if_else(
      label %in% current_outliers,
      "Discussed outlier",
      "Other sample"
    )
  )
sample_order <- sample_metadata$label

row_cluster_spans <- read.csv(row_cluster_span_file, check.names = FALSE) |>
  dplyr::arrange(start) |>
  dplyr::mutate(
    original_expression_cluster_order = dplyr::row_number(),
    original_expression_cluster = paste0(
      "Original branch ", original_expression_cluster_order,
      "\n(n = ", n, " genes)"
    ),
    expression_cluster_order = dplyr::case_when(
      original_expression_cluster_order %in% c(1, 2) ~ original_expression_cluster_order,
      original_expression_cluster_order %in% c(3, 4) ~ 3L
    )
  )

merged_cluster_spans <- row_cluster_spans |>
  dplyr::group_by(expression_cluster_order) |>
  dplyr::summarise(
    start = min(start),
    end = max(end),
    n = sum(n),
    original_branches = paste(original_expression_cluster_order, collapse = "+"),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    expression_cluster = dplyr::case_when(
      expression_cluster_order == 1 ~ "DEG cluster 1",
      expression_cluster_order == 2 ~ "DEG cluster 2",
      expression_cluster_order == 3 ~ "DEG cluster 3+4"
    )
  )
cluster_levels <- merged_cluster_spans$expression_cluster
expression_cluster_colors <- stats::setNames(
  c("#315C8A", "#C85A54", "#2F8F5B"),
  cluster_levels
)
cluster_index <- findInterval(
  full_heatmap_genes$heatmap_row_order,
  row_cluster_spans$start
)
cluster_index <- pmin(pmax(cluster_index, 1), nrow(row_cluster_spans))
merged_cluster_index <- row_cluster_spans$expression_cluster_order[cluster_index]
merged_cluster_match <- match(
  merged_cluster_index,
  merged_cluster_spans$expression_cluster_order
)

deg_context_levels <- c(
  "Diet in controls",
  "Diet in infected",
  "Infection within diet",
  "Infection + diet",
  "Diet in controls + infected",
  "Other fat body DEG"
)
deg_context_labels <- c(
  "Diet in controls" = "Diet DEG in controls",
  "Diet in infected" = "Diet DEG in infected",
  "Infection within diet" = "Infected vs control within diet",
  "Infection + diet" = "Infection + diet DEG",
  "Diet in controls + infected" = "Diet DEG in controls + infected",
  "Other fat body DEG" = "Other fat body DEG"
)
deg_context_colors <- c(
  "Diet in controls" = "#DDE3EA",
  "Diet in infected" = "#AEBBC8",
  "Infection within diet" = "#7D8FA3",
  "Infection + diet" = "#4D6278",
  "Diet in controls + infected" = "#2F4052",
  "Other fat body DEG" = "#F2F4F6"
)
deg_context_plot_levels <- c(
  "~ diet (controls)",
  "~ diet (infected)",
  "~ diet + infection",
  "shared DEGs",
  "Other fat body DEG"
)
deg_context_plot_colors <- c(
  "~ diet (controls)" = "#DDE3EA",
  "~ diet (infected)" = "#AEBBC8",
  "~ diet + infection" = "#65788C",
  "shared DEGs" = "#263544",
  "Other fat body DEG" = "#F2F4F6"
)

expression_cluster_assignments <- full_heatmap_genes |>
  dplyr::mutate(
    expression_cluster_order = merged_cluster_index,
    expression_cluster = factor(merged_cluster_spans$expression_cluster[merged_cluster_match],
                                levels = cluster_levels),
    original_expression_cluster_order =
      row_cluster_spans$original_expression_cluster_order[cluster_index],
    original_expression_cluster =
      row_cluster_spans$original_expression_cluster[cluster_index],
    original_pheatmap_cluster_id = row_cluster_spans$cluster_id[cluster_index],
    within_cluster_order = heatmap_row_order -
      merged_cluster_spans$start[merged_cluster_match] + 1,
    original_within_cluster_order = heatmap_row_order -
      row_cluster_spans$start[cluster_index] + 1,
    DEG_context = dplyr::coalesce(DEG_context, "Other fat body DEG"),
    DEG_context = dplyr::if_else(
      DEG_context %in% deg_context_levels,
      DEG_context,
      "Other fat body DEG"
    ),
    DEG_context = factor(DEG_context, levels = deg_context_levels)
  ) |>
  dplyr::arrange(
    expression_cluster_order,
    DEG_context,
    min_padj,
    dplyr::desc(max_abs_log2FC),
    original_within_cluster_order
  ) |>
  dplyr::group_by(expression_cluster) |>
  dplyr::mutate(within_cluster_order = dplyr::row_number()) |>
  dplyr::ungroup() |>
  dplyr::select(heatmap_row_order, gene_id, expression_cluster_order,
                expression_cluster, original_expression_cluster_order,
                original_expression_cluster, original_pheatmap_cluster_id,
                within_cluster_order, original_within_cluster_order,
                Description, DEG_context, phase_class,
                n_deg_contrasts, comparisons, directions, max_abs_log2FC,
                min_padj, max_abs_sample_z, sample_with_max_abs_z,
                sample_deviation_direction)

write.csv(
  expression_cluster_assignments,
  file.path(out_dir, "full_heatmap_expression_cluster_assignments.csv"),
  row.names = FALSE
)

cluster_go_enrichment <- dplyr::bind_rows(lapply(cluster_levels, function(cl) {
  genes <- expression_cluster_assignments$gene_id[
    expression_cluster_assignments$expression_cluster == cl
  ]
  run_enrichment(
    genes = genes,
    universe = backgrounds[["All fat body DEGs in heatmap"]],
    term2gene = maps$term2gene_go,
    term2name = maps$term2name_go,
    database = "GO",
    background_label = "All fat body DEGs in heatmap"
  ) |>
    dplyr::mutate(
      expression_cluster = cl,
      n_cluster_genes = length(genes)
    )
}))

if (nrow(cluster_go_enrichment) > 0) {
  cluster_go_enrichment <- cluster_go_enrichment |>
    dplyr::left_join(
      maps$term2name_go |>
        dplyr::distinct(term, .keep_all = TRUE) |>
        dplyr::transmute(
          ID = term,
          ontology = dplyr::recode(
            ontology,
            BP = "Biological process",
            CC = "Cellular component",
            MF = "Molecular function",
            .default = NA_character_
          )
        ),
      by = "ID"
    ) |>
    dplyr::filter(!is.na(ontology)) |>
    dplyr::mutate(
      expression_cluster = factor(expression_cluster, levels = cluster_levels),
      ontology = factor(ontology, levels = names(go_ontology_colors))
    )
}

cluster_go_significant <- cluster_go_enrichment |>
  dplyr::filter(p.adjust <= params$alpha) |>
  dplyr::arrange(expression_cluster, ontology, p.adjust)

cluster_go_summary <- cluster_go_significant |>
  dplyr::count(expression_cluster, ontology, name = "n_significant_terms") |>
  tidyr::complete(expression_cluster = factor(cluster_levels,
                                             levels = cluster_levels),
                  ontology = factor(names(go_ontology_colors),
                                    levels = names(go_ontology_colors)),
                  fill = list(n_significant_terms = 0)) |>
  dplyr::arrange(expression_cluster, ontology)

write.csv(cluster_go_enrichment,
          file.path(out_dir,
                    "expression_cluster_GO_enrichment_all_terms.csv"),
          row.names = FALSE)
write.csv(cluster_go_significant,
          file.path(out_dir,
                    "expression_cluster_GO_enrichment_significant_terms.csv"),
          row.names = FALSE)
write.csv(cluster_go_summary,
          file.path(out_dir,
                    "expression_cluster_GO_enrichment_count_summary.csv"),
          row.names = FALSE)

cluster_gene_annotation <- expression_cluster_assignments |>
  dplyr::select(gene_id, DEG_context, phase_class, n_deg_contrasts,
                comparisons, directions, max_abs_log2FC, min_padj) |>
  dplyr::left_join(
    gff_annotation |> dplyr::select(gene_id, Description, gene_biotype),
    by = "gene_id"
  ) |>
  dplyr::mutate(
    Description = dplyr::coalesce(Description, "no GFF description"),
    gene_biotype = dplyr::coalesce(gene_biotype, "unknown")
  )

cluster_term_gene_long <- cluster_go_enrichment |>
  dplyr::filter(p.adjust <= params$alpha) |>
  dplyr::select(expression_cluster, database, background, ID, Description,
                GeneRatio, BgRatio, pvalue, p.adjust, qvalue, Count, geneID) |>
  tidyr::separate_rows(geneID, sep = "/") |>
  dplyr::rename(term_id = ID, term_name = Description, gene_id = geneID) |>
  dplyr::left_join(cluster_gene_annotation, by = "gene_id") |>
  dplyr::arrange(expression_cluster, p.adjust, term_id, gene_id)

cluster_term_gene_compact <- cluster_term_gene_long |>
  dplyr::group_by(expression_cluster, database, background, term_id, term_name,
                  GeneRatio, BgRatio, p.adjust, Count) |>
  dplyr::summarise(
    DEG_gene_IDs = paste(sort(unique(gene_id)), collapse = "; "),
    DEG_gene_descriptions = paste(sort(unique(Description)), collapse = "; "),
    DEG_contexts = paste(sort(unique(DEG_context)), collapse = "; "),
    .groups = "drop"
  ) |>
  dplyr::arrange(expression_cluster, p.adjust)

write.csv(cluster_term_gene_long,
          file.path(out_dir,
                    "expression_cluster_GO_term_gene_index_long.csv"),
          row.names = FALSE)
write.csv(cluster_term_gene_compact,
          file.path(out_dir,
                    "expression_cluster_GO_term_gene_index_compact.csv"),
          row.names = FALSE)

sample_group_ranges <- sample_metadata |>
  dplyr::group_by(Treatment_plot, Diet_plot, sample_group) |>
  dplyr::summarise(start = min(sample_index), end = max(sample_index),
                   .groups = "drop") |>
  dplyr::arrange(Treatment_plot, Diet_plot)
sample_gaps <- sample_group_ranges$end[-nrow(sample_group_ranges)] + 0.5
treatment_gaps <- sample_group_ranges$end[
  which(as.character(sample_group_ranges$Treatment_plot[-nrow(sample_group_ranges)]) !=
          as.character(sample_group_ranges$Treatment_plot[-1]))
] + 0.5
diet_gaps <- setdiff(sample_gaps, treatment_gaps)

annotation_fill_colors <- c(
  "Treatment: Infected" = "#8B0000",
  "Treatment: Control" = "#FAEBD7",
  "Diet: 33" = "#E6862E",
  "Diet: 50" = "#2F8F5B",
  "Diet: 83" = "#7B4FA3"
)

sample_x_limits <- c(-1.25, length(sample_order) + 0.5)

sample_annotation_long <- sample_metadata |>
  dplyr::transmute(
    sample_index,
    Treatment = as.character(Treatment),
    Diet = as.character(Diet)
  ) |>
  tidyr::pivot_longer(
    cols = c(Treatment, Diet),
    names_to = "track",
    values_to = "value"
  ) |>
  dplyr::mutate(
    track = factor(track, levels = c("Diet", "Treatment")),
    track_y = as.numeric(track),
    fill_key = paste(track, value, sep = ": ")
  )

sample_annotation_plot <- ggplot2::ggplot(
  sample_annotation_long,
  ggplot2::aes(x = sample_index, y = track_y, fill = fill_key)
) +
  ggplot2::geom_tile(width = 0.95, height = 0.9) +
  ggplot2::geom_vline(xintercept = diet_gaps, color = "black",
                      linewidth = 0.35) +
  ggplot2::geom_vline(xintercept = treatment_gaps, color = "black",
                      linewidth = 0.8) +
  ggplot2::scale_fill_manual(values = annotation_fill_colors, drop = FALSE) +
  ggplot2::scale_x_continuous(
    limits = sample_x_limits,
    breaks = sample_metadata$sample_index,
    labels = sample_metadata$label,
    expand = c(0, 0),
    position = "top"
  ) +
  ggplot2::scale_y_continuous(
    breaks = seq_len(2),
    labels = c("Diet", "Treatment"),
    limits = c(0.45, 2.55),
    expand = c(0, 0),
    sec.axis = ggplot2::dup_axis(
      breaks = seq_len(2),
      labels = c("Diet", "Treatment")
    )
  ) +
  ggplot2::labs(
    x = NULL,
    y = NULL,
    title = "A. Full fat body DEG expression landscape split by row clusters",
    fill = "Sample annotation"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank(),
    axis.text.y = ggplot2::element_text(face = "bold", size = 11),
    axis.text.y.right = ggplot2::element_text(face = "bold", size = 11),
    legend.position = "bottom",
    legend.title = ggplot2::element_text(face = "bold"),
    plot.title = ggplot2::element_text(face = "bold", size = 15),
    plot.margin = ggplot2::margin(3, 4, 0, 4)
  )

heatmap_z_plot <- full_heatmap_z[
  expression_cluster_assignments$gene_id,
  sample_order,
  drop = FALSE
]
heatmap_z_plot <- pmax(pmin(heatmap_z_plot, 3), -3)
heatmap_df <- as.data.frame(heatmap_z_plot, check.names = FALSE) |>
  tibble::rownames_to_column("gene_id") |>
  dplyr::left_join(
    expression_cluster_assignments |>
      dplyr::select(gene_id, expression_cluster, within_cluster_order,
                    DEG_context),
    by = "gene_id"
  ) |>
  tidyr::pivot_longer(
    cols = dplyr::all_of(sample_order),
    names_to = "sample_id",
    values_to = "z"
  ) |>
  dplyr::mutate(sample_index = match(sample_id, sample_order))

heatmap_colors <- grDevices::colorRampPalette(
  c("#237C8E", "#F7F3E8", "#C85A54")
)(101)

gene_context_strip_df <- expression_cluster_assignments |>
  dplyr::select(gene_id, expression_cluster, within_cluster_order,
                DEG_context) |>
  dplyr::mutate(
    DEG_context_plot = dplyr::case_when(
      DEG_context == "Diet in controls" ~
        "~ diet (controls)",
      DEG_context == "Diet in infected" ~
        "~ diet (infected)",
      DEG_context == "Infection within diet" ~
        "~ diet + infection",
      DEG_context %in% c("Infection + diet",
                         "Diet in controls + infected") ~
        "shared DEGs",
      TRUE ~ "Other fat body DEG"
    ),
    DEG_context_plot = factor(DEG_context_plot,
                              levels = deg_context_plot_levels)
  )
present_deg_context_levels <- deg_context_plot_levels[
  deg_context_plot_levels %in%
    as.character(gene_context_strip_df$DEG_context_plot)
]
present_deg_context_labels <- stringr::str_wrap(
  present_deg_context_levels,
  width = 26
)

cluster_heatmap_plot <- ggplot2::ggplot(
  heatmap_df,
  ggplot2::aes(x = sample_index, y = within_cluster_order, fill = z)
) +
  ggplot2::geom_segment(
    data = gene_context_strip_df,
    ggplot2::aes(
      x = -1.1,
      xend = 0.35,
      y = within_cluster_order,
      yend = within_cluster_order,
      color = DEG_context_plot
    ),
    inherit.aes = FALSE,
    linewidth = 1.05,
    lineend = "butt"
  ) +
  ggplot2::geom_raster() +
  ggplot2::geom_vline(xintercept = 0.5, color = "#9CA3AF",
                      linewidth = 0.4) +
  ggplot2::geom_vline(xintercept = diet_gaps, color = "black",
                      linewidth = 0.25) +
  ggplot2::geom_vline(xintercept = treatment_gaps, color = "black",
                      linewidth = 0.65) +
  ggplot2::facet_grid(expression_cluster ~ ., scales = "free_y",
                      switch = "y") +
  ggplot2::scale_fill_gradientn(
    colors = heatmap_colors,
    limits = c(-3, 3),
    oob = scales::squish,
    name = "Row z-score"
  ) +
  ggplot2::scale_color_manual(
    values = deg_context_plot_colors,
    breaks = present_deg_context_levels,
    labels = present_deg_context_labels,
    drop = TRUE,
    name = "DEG context"
  ) +
  ggplot2::scale_x_continuous(
    limits = sample_x_limits,
    breaks = sample_metadata$sample_index,
    labels = sample_metadata$label,
    expand = c(0, 0)
  ) +
  ggplot2::scale_y_reverse(expand = c(0, 0)) +
  ggplot2::labs(
    x = NULL,
    y = NULL,
    title = NULL,
    subtitle = NULL
  ) +
  ggplot2::theme_classic(base_size = 13) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, hjust = 1,
                                        vjust = 0.5, size = 13),
    axis.text.y = ggplot2::element_blank(),
    axis.ticks.y = ggplot2::element_blank(),
    strip.placement = "outside",
    strip.background.y = ggplot2::element_rect(fill = "#F3F4F6",
                                               color = "#9CA3AF"),
    strip.text.y.left = ggplot2::element_text(angle = 90, face = "bold",
                                              size = 13),
    panel.spacing.y = grid::unit(0.25, "lines"),
    legend.position = "right",
    legend.box = "vertical",
    legend.title = ggplot2::element_text(face = "bold"),
    legend.text = ggplot2::element_text(size = 9),
    plot.title = ggplot2::element_text(face = "bold"),
    plot.margin = ggplot2::margin(0, 4, 3, 4)
  ) +
  ggplot2::guides(
    fill = ggplot2::guide_colorbar(
      order = 1,
      title.position = "top",
      barheight = grid::unit(3.0, "cm"),
      barwidth = grid::unit(0.35, "cm")
    ),
    color = ggplot2::guide_legend(
      order = 2,
      title.position = "top",
      override.aes = list(linewidth = 3)
    )
  )

ggplot2::ggsave(
  file.path(out_dir, "expression_cluster_gene_level_heatmap.png"),
  cluster_heatmap_plot,
  width = 20,
  height = 12,
  dpi = 260
)

prepare_lollipop_df <- function(tab, ontologies, n_terms,
                                terms_per_ontology = FALSE) {
  group_vars <- if (terms_per_ontology) {
    c("expression_cluster", "ontology")
  } else {
    "expression_cluster"
  }

  tab |>
    dplyr::filter(p.adjust <= params$alpha, ontology %in% ontologies) |>
    dplyr::group_by(dplyr::across(dplyr::all_of(group_vars))) |>
    dplyr::slice_min(p.adjust, n = n_terms, with_ties = FALSE) |>
    dplyr::ungroup() |>
    dplyr::arrange(expression_cluster, ontology, p.adjust) |>
    dplyr::group_by(expression_cluster) |>
    dplyr::mutate(
      neg_log10_padj = -log10(pmax(p.adjust, .Machine$double.xmin)),
      term_label = stringr::str_wrap(Description, 46),
      term_plot_id = paste(term_label, dplyr::row_number(),
                           expression_cluster, sep = "___")
    ) |>
    dplyr::ungroup() |>
    dplyr::mutate(
      term_plot_id = factor(term_plot_id, levels = rev(unique(term_plot_id))),
      expression_cluster = factor(expression_cluster, levels = cluster_levels)
    )
}

clean_term_labels <- function(x) {
  labels <- sub("___.*$", "", x)
  dplyr::if_else(labels == "No significant GO terms", "", labels)
}

make_lollipop_plot <- function(plot_df, title, subtitle = NULL) {
  if (nrow(plot_df) == 0) {
    return(
      ggplot2::ggplot() +
        ggplot2::annotate("text", x = 0, y = 0,
                          label = "No significant GO terms",
                          size = 6, fontface = "bold") +
        ggplot2::theme_void() +
        ggplot2::labs(title = title, subtitle = subtitle)
      )
  }

  x_axis_max <- max(
    c(plot_df$neg_log10_padj, -log10(params$alpha)),
    na.rm = TRUE
  )
  x_axis_limit <- x_axis_max + max(0.8, x_axis_max * 0.18)
  label_space <- x_axis_max * 2.4
  x_breaks <- pretty(c(0, x_axis_limit), n = 5)
  x_breaks <- x_breaks[x_breaks >= 0 & x_breaks <= x_axis_limit]
  missing_label_x <- max(
    c(plot_df$neg_log10_padj, -log10(params$alpha)),
    na.rm = TRUE
  ) / 2
  missing_cluster_df <- tibble::tibble(
    expression_cluster = factor(
      setdiff(cluster_levels, as.character(unique(plot_df$expression_cluster))),
      levels = cluster_levels
    ),
    term_plot_id = "No significant GO terms",
    x = missing_label_x,
    label = "No significant GO terms"
  ) |>
    dplyr::filter(!is.na(expression_cluster))

  ggplot2::ggplot(
    plot_df,
    ggplot2::aes(x = neg_log10_padj, y = term_plot_id)
  ) +
    ggplot2::geom_vline(
      xintercept = -log10(params$alpha),
      linetype = "dashed",
      color = "#6B7280",
      linewidth = 0.35
    ) +
    ggplot2::geom_vline(
      xintercept = 0,
      color = "#9CA3AF",
      linewidth = 0.35
    ) +
    ggplot2::geom_text(
      data = plot_df,
      ggplot2::aes(
        x = -label_space * 0.98,
        y = term_plot_id,
        label = clean_term_labels(term_plot_id)
      ),
      inherit.aes = FALSE,
      hjust = 0,
      size = 5.2,
      lineheight = 0.84,
      color = "#111827"
    ) +
    ggplot2::geom_segment(
      ggplot2::aes(x = 0, xend = neg_log10_padj,
                   y = term_plot_id, yend = term_plot_id,
                   color = ontology),
      linewidth = 0.7,
      alpha = 0.75
    ) +
    ggplot2::geom_point(
      ggplot2::aes(size = Count, color = ontology),
      alpha = 0.95
    ) +
    ggplot2::geom_text(
      data = missing_cluster_df,
      ggplot2::aes(x = x, y = term_plot_id, label = label),
      inherit.aes = FALSE,
      size = 4,
      color = "#6B7280",
      fontface = "bold"
    ) +
    ggplot2::facet_wrap(~ expression_cluster, nrow = 1,
                        scales = "free_y", drop = FALSE) +
    ggplot2::scale_color_manual(values = go_ontology_colors, drop = FALSE) +
    ggplot2::scale_y_discrete(labels = function(x) rep("", length(x))) +
    ggplot2::scale_x_continuous(
      limits = c(-label_space, x_axis_limit),
      breaks = x_breaks,
      expand = c(0, 0)
    ) +
    ggplot2::scale_size_continuous(range = c(3.5, 9)) +
    ggplot2::labs(
      title = title,
      subtitle = subtitle,
      x = "-log10 adjusted p-value",
      y = NULL,
      color = "GO ontology",
      size = "Cluster genes in term"
    ) +
    ggplot2::theme_classic(base_size = 13) +
    ggplot2::theme(
      axis.text.y = ggplot2::element_blank(),
      axis.ticks.y = ggplot2::element_blank(),
      axis.line = ggplot2::element_blank(),
      axis.line.x.bottom = ggplot2::element_line(color = "black",
                                                 linewidth = 0.35),
      panel.border = ggplot2::element_blank(),
      strip.background = ggplot2::element_rect(fill = "#F3F4F6",
                                               color = "#9CA3AF"),
      strip.text = ggplot2::element_text(face = "bold", size = 12),
      legend.position = "bottom",
      plot.title = ggplot2::element_text(face = "bold")
    )
}

go_lollipop_df <- prepare_lollipop_df(
  cluster_go_significant,
  ontologies = names(go_ontology_colors),
  n_terms = 4,
  terms_per_ontology = TRUE
)

go_lollipop_plot <- make_lollipop_plot(
  go_lollipop_df,
  title = "B. GO terms by expression cluster",
  subtitle = "Foreground = genes in each expression cluster; background = all fat body DEGs in the heatmap"
)

gene_level_GO_figure <- sample_annotation_plot /
  cluster_heatmap_plot /
  go_lollipop_plot +
  patchwork::plot_layout(heights = c(0.95, 6.8, 5.6)) +
  patchwork::plot_annotation(
    title = "Full gene-level expression and GO enrichment by DEG cluster",
    subtitle = "All samples are retained; GO lollipops combine BP, CC, and MF terms in each cluster"
  )

ggplot2::ggsave(
  file.path(out_dir, "expression_cluster_gene_level_GO_lollipop.png"),
  gene_level_GO_figure,
  width = 21,
  height = 19,
  dpi = 220
)

knitr::kable(
  cluster_go_summary,
  caption = "Significant GO terms by expression cluster and ontology"
)
```

## Unmerged four-branch review {#unmerged-four-branch-review .tabset}

The four branches below are the direct `k = 4` cut of the current
main-chromosome gene-expression dendrogram. No branches are merged in this
section. This is the appropriate table to review before deciding whether the
publication figure should retain four clusters, merge selected branches, or use
a different cut of the tree.

Enrichment is recomputed independently for each original branch against all
fat-body DEGs in the heatmap. A zero therefore means that no term passed the
adjusted p-value threshold for that branch and background. It does not mean
that the branch contains no annotated genes.

### Coverage and enrichment

```{r unmerged-k4-coverage-enrichment}
original_branch_levels <- paste("Branch", seq_len(nrow(row_cluster_spans)))

go_gene_counts <- maps$term2gene_go |>
  dplyr::distinct(gene) |>
  dplyr::count(gene, name = "n_GO_annotations")
kegg_gene_counts <- maps$term2gene_kegg |>
  dplyr::distinct(gene) |>
  dplyr::count(gene, name = "n_KEGG_annotations")

unmerged_gene_table <- expression_cluster_assignments |>
  dplyr::mutate(
    original_branch = factor(
      paste("Branch", original_expression_cluster_order),
      levels = original_branch_levels
    ),
    proposed_merged_cluster = as.character(expression_cluster)
  ) |>
  dplyr::left_join(
    gff_annotation |> dplyr::select(gene_id, gene_biotype),
    by = "gene_id"
  ) |>
  dplyr::left_join(go_gene_counts, by = c("gene_id" = "gene")) |>
  dplyr::left_join(kegg_gene_counts, by = c("gene_id" = "gene")) |>
  dplyr::mutate(
    gene_biotype = dplyr::coalesce(gene_biotype, "unknown"),
    n_GO_annotations = dplyr::coalesce(n_GO_annotations, 0L),
    n_KEGG_annotations = dplyr::coalesce(n_KEGG_annotations, 0L)
  ) |>
  dplyr::arrange(original_expression_cluster_order,
                 original_within_cluster_order) |>
  dplyr::select(
    heatmap_row_order, original_branch, original_pheatmap_cluster_id,
    original_within_cluster_order, proposed_merged_cluster,
    gene_id, Description, gene_biotype, DEG_context, n_deg_contrasts,
    comparisons, directions, max_abs_log2FC, min_padj,
    max_abs_sample_z, sample_with_max_abs_z,
    sample_deviation_direction, n_GO_annotations, n_KEGG_annotations
  )

write.csv(
  unmerged_gene_table,
  file.path(out_dir, "unmerged_k4_full_gene_assignments.csv"),
  row.names = FALSE
)

unmerged_go_enrichment <- dplyr::bind_rows(lapply(
  original_branch_levels,
  function(branch) {
    genes <- unmerged_gene_table$gene_id[
      as.character(unmerged_gene_table$original_branch) == branch
    ]
    run_enrichment(
      genes = genes,
      universe = backgrounds[["All fat body DEGs in heatmap"]],
      term2gene = maps$term2gene_go,
      term2name = maps$term2name_go,
      database = "GO",
      background_label = "All fat body DEGs in heatmap"
    ) |>
      dplyr::mutate(original_branch = branch,
                    n_branch_genes = length(genes))
  }
)) |>
  dplyr::left_join(
    maps$term2name_go |>
      dplyr::distinct(term, .keep_all = TRUE) |>
      dplyr::transmute(
        ID = term,
        ontology = dplyr::recode(
          ontology,
          BP = "Biological process",
          CC = "Cellular component",
          MF = "Molecular function",
          .default = NA_character_
        )
      ),
    by = "ID"
  ) |>
  dplyr::filter(!is.na(ontology))

unmerged_kegg_enrichment <- dplyr::bind_rows(lapply(
  original_branch_levels,
  function(branch) {
    genes <- unmerged_gene_table$gene_id[
      as.character(unmerged_gene_table$original_branch) == branch
    ]
    run_enrichment(
      genes = genes,
      universe = backgrounds[["All fat body DEGs in heatmap"]],
      term2gene = maps$term2gene_kegg,
      term2name = maps$term2name_kegg,
      database = "KEGG_pathway",
      background_label = "All fat body DEGs in heatmap"
    ) |>
      dplyr::mutate(original_branch = branch,
                    n_branch_genes = length(genes))
  }
))

unmerged_go_significant <- unmerged_go_enrichment |>
  dplyr::filter(p.adjust <= params$alpha) |>
  dplyr::arrange(original_branch, ontology, p.adjust)
unmerged_kegg_significant <- unmerged_kegg_enrichment |>
  dplyr::filter(p.adjust <= params$alpha) |>
  dplyr::arrange(original_branch, p.adjust)

write.csv(
  unmerged_go_enrichment,
  file.path(out_dir, "unmerged_k4_GO_enrichment_all_terms.csv"),
  row.names = FALSE
)
write.csv(
  unmerged_go_significant,
  file.path(out_dir, "unmerged_k4_GO_enrichment_significant_terms.csv"),
  row.names = FALSE
)
write.csv(
  unmerged_kegg_enrichment,
  file.path(out_dir, "unmerged_k4_KEGG_enrichment_all_terms.csv"),
  row.names = FALSE
)
write.csv(
  unmerged_kegg_significant,
  file.path(out_dir, "unmerged_k4_KEGG_enrichment_significant_terms.csv"),
  row.names = FALSE
)

branch_coverage <- unmerged_gene_table |>
  dplyr::group_by(original_branch) |>
  dplyr::summarise(
    n_genes = dplyr::n(),
    n_with_GFF_description = sum(Description != "no GFF description"),
    n_with_GO = sum(n_GO_annotations > 0),
    percent_with_GO = round(100 * n_with_GO / n_genes, 1),
    n_with_KEGG = sum(n_KEGG_annotations > 0),
    percent_with_KEGG = round(100 * n_with_KEGG / n_genes, 1),
    .groups = "drop"
  )

branch_go_counts <- unmerged_go_significant |>
  dplyr::mutate(
    ontology_code = dplyr::recode(
      ontology,
      "Biological process" = "BP",
      "Cellular component" = "CC",
      "Molecular function" = "MF"
    )
  ) |>
  dplyr::count(original_branch, ontology_code, name = "n_terms") |>
  tidyr::complete(
    original_branch = original_branch_levels,
    ontology_code = c("BP", "CC", "MF"),
    fill = list(n_terms = 0)
  ) |>
  tidyr::pivot_wider(
    names_from = ontology_code,
    values_from = n_terms,
    names_prefix = "significant_GO_"
  )

branch_kegg_counts <- unmerged_kegg_significant |>
  dplyr::count(original_branch, name = "significant_KEGG_pathways") |>
  tidyr::complete(
    original_branch = original_branch_levels,
    fill = list(significant_KEGG_pathways = 0)
  )

unmerged_branch_summary <- branch_coverage |>
  dplyr::mutate(original_branch = as.character(original_branch)) |>
  dplyr::left_join(branch_go_counts, by = "original_branch") |>
  dplyr::left_join(branch_kegg_counts, by = "original_branch") |>
  dplyr::arrange(match(original_branch, original_branch_levels))

write.csv(
  unmerged_branch_summary,
  file.path(out_dir, "unmerged_k4_annotation_enrichment_summary.csv"),
  row.names = FALSE
)

knitr::kable(
  unmerged_branch_summary,
  caption = paste0(
    "Annotation coverage and significant enrichment for the four unmerged ",
    "branches; adjusted p-value <= ", params$alpha
  )
)
```

### Complete gene list

```{r unmerged-k4-complete-gene-list}
DT::datatable(
  unmerged_gene_table,
  rownames = FALSE,
  filter = "top",
  options = list(pageLength = 25, scrollX = TRUE),
  caption = paste0(
    "All ", nrow(unmerged_gene_table),
    " main-chromosome DEGs with their unmerged branch assignments"
  )
)
```

### Significant GO terms

```{r unmerged-k4-go-table}
if (nrow(unmerged_go_significant) > 0) {
  DT::datatable(
    unmerged_go_significant |>
      dplyr::select(original_branch, ontology, ID, Description,
                    GeneRatio, BgRatio, pvalue, p.adjust, qvalue,
                    Count, geneID),
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 20, scrollX = TRUE),
    caption = "Significant GO terms for each unmerged branch"
  )
} else {
  cat("No GO term passed the adjusted p-value threshold in any branch.")
}
```

### Significant KEGG pathways

```{r unmerged-k4-kegg-table}
if (nrow(unmerged_kegg_significant) > 0) {
  DT::datatable(
    unmerged_kegg_significant |>
      dplyr::select(original_branch, ID, Description, GeneRatio, BgRatio,
                    pvalue, p.adjust, qvalue, Count, geneID),
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 20, scrollX = TRUE),
    caption = "Significant KEGG pathways for each unmerged branch"
  )
} else {
  cat("No KEGG pathway passed the adjusted p-value threshold in any branch.")
}
```

## Combined GO and KEGG figure

The main figure below keeps the full gene-by-sample heatmap so every DEG in the
circled branch remains visible, then summarizes GO terms and KEGG pathways for
the same expression clusters. KEGG terms are eggNOG-linked `ko` reference
pathways, not insect-only organism pathways. The pathway-scope category is kept
in the exported KEGG tables, but panel C uses one brown color so the pathway
labels remain visually simple. The code also exports optional
per-sample and treatment-average cluster-score summaries as secondary diagnostic
files.

### DEG context annotation in panel A

The `DEG context` strip on the left of panel A is a contrast-family annotation.
It tells which type of DESeq2 contrast detected each gene as significant before
the heatmap was clustered. These labels are useful for reading the heatmap, but
they should not be interpreted as mutually exclusive biological mechanisms.

`~ diet (controls)` means the gene was significant in diet comparisons among
control animals only, for example control 33 vs 50, 83 vs 33, or 83 vs 50. This
is the closest label to a baseline diet effect without infection.

`~ diet (infected)` means the gene was significant in diet comparisons among
infected animals only. This asks whether infected animals on different diets
have different expression.

`~ diet + infection` means the gene was significant when infected and
control animals were compared within at least one diet. This is the most direct
infection-response label in this heatmap, but the comparison is diet-stratified,
so the label keeps both diet and infection in view.

`shared DEGs` means the gene was detected in more than one contrast family. This
combines genes that were diet-associated in both control and infected
backgrounds, and genes that were detected in both infection and diet contrast
families. It is an overlap label and does not by itself prove a formal
diet-by-infection interaction.

`Other fat body DEG` is a fallback category for genes in the full fat-body DEG
heatmap that do not fall cleanly into the simplified contrast families above. It
is retained in the code for completeness, but it may not appear in every figure.

```{r expression-cluster-treatment-summary, fig.width=18, fig.height=18}
cluster_sample_scores <- heatmap_df |>
  dplyr::group_by(expression_cluster, sample_id) |>
  dplyr::summarise(cluster_mean_z = mean(z, na.rm = TRUE),
                   .groups = "drop") |>
  dplyr::left_join(
    sample_metadata |>
      dplyr::select(label, Treatment, Diet, sample_group, outlier_status),
    by = c("sample_id" = "label")
  ) |>
  dplyr::mutate(
    expression_cluster = factor(expression_cluster, levels = cluster_levels),
    sample_group = factor(sample_group,
                          levels = unique(sample_metadata$sample_group)),
    sample_label = dplyr::if_else(
      outlier_status == "Discussed outlier",
      paste0(sample_id, " *"),
      sample_id
    ),
    sample_label = factor(
      sample_label,
      levels = rev(dplyr::if_else(
        sample_metadata$outlier_status == "Discussed outlier",
        paste0(sample_metadata$label, " *"),
        sample_metadata$label
      ))
    )
  )

treatment_cluster_summary <- cluster_sample_scores |>
  dplyr::group_by(expression_cluster, Treatment, Diet, sample_group) |>
  dplyr::summarise(
    n_samples = dplyr::n(),
    mean_z = mean(cluster_mean_z, na.rm = TRUE),
    sd_z = stats::sd(cluster_mean_z, na.rm = TRUE),
    se_z = sd_z / sqrt(n_samples),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    sample_group_plot = factor(sample_group,
                               levels = rev(unique(sample_metadata$sample_group))),
    tile_label = sprintf("%.2f", mean_z)
  )

write.csv(cluster_sample_scores,
          file.path(out_dir,
                    "expression_cluster_sample_scores.csv"),
          row.names = FALSE)
write.csv(treatment_cluster_summary,
          file.path(out_dir,
                    "expression_cluster_treatment_summary.csv"),
          row.names = FALSE)

sample_cluster_heatmap_plot <- ggplot2::ggplot(
  cluster_sample_scores,
  ggplot2::aes(x = expression_cluster, y = sample_label, fill = cluster_mean_z)
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.8) +
  ggplot2::facet_grid(sample_group ~ ., scales = "free_y",
                      space = "free_y", switch = "y") +
  ggplot2::scale_fill_gradientn(
    colors = heatmap_colors,
    limits = c(-2, 2),
    oob = scales::squish,
    name = "Sample cluster\nscore"
  ) +
  ggplot2::scale_x_discrete(position = "top") +
  ggplot2::labs(
    title = "A. Sample-level cluster expression heatmap",
    subtitle = "Each tile is one sample's mean score for a gene-expression cluster; * marks discussed outliers",
    x = "Expression cluster",
    y = "Sample grouped by diet/treatment"
  ) +
  ggplot2::theme_classic(base_size = 14) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(face = "bold", size = 12),
    axis.text.y = ggplot2::element_text(face = "bold", size = 8),
    axis.title = ggplot2::element_text(face = "bold"),
    strip.placement = "outside",
    strip.background.y = ggplot2::element_rect(fill = "#F3F4F6",
                                               color = "#9CA3AF"),
    strip.text.y.left = ggplot2::element_text(angle = 0, face = "bold",
                                              size = 11),
    panel.spacing.y = grid::unit(0.15, "lines"),
    legend.position = "right",
    plot.title = ggplot2::element_text(face = "bold")
  )

ggplot2::ggsave(
  file.path(out_dir, "expression_cluster_sample_level_heatmap.png"),
  sample_cluster_heatmap_plot,
  width = 10,
  height = 10,
  dpi = 300
)

treatment_cluster_heatmap_plot <- ggplot2::ggplot(
  treatment_cluster_summary,
  ggplot2::aes(x = expression_cluster, y = sample_group_plot, fill = mean_z)
) +
  ggplot2::geom_tile(color = "white", linewidth = 1.2) +
  ggplot2::geom_text(
    ggplot2::aes(label = tile_label),
    size = 5,
    fontface = "bold",
    color = "#111827"
  ) +
  ggplot2::scale_fill_gradientn(
    colors = heatmap_colors,
    limits = c(-2, 2),
    oob = scales::squish,
    name = "Mean row\nz-score"
  ) +
  ggplot2::scale_x_discrete(position = "top") +
  ggplot2::labs(
    title = "A. Treatment-by-cluster expression summary",
    subtitle = "Each tile is the mean cluster score across samples in that diet/treatment group",
    x = "Expression cluster",
    y = "Diet/treatment group"
  ) +
  ggplot2::theme_classic(base_size = 14) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(face = "bold", size = 12),
    axis.text.y = ggplot2::element_text(face = "bold", size = 12),
    axis.title = ggplot2::element_text(face = "bold"),
    legend.position = "right",
    plot.title = ggplot2::element_text(face = "bold")
  )

ggplot2::ggsave(
  file.path(out_dir, "expression_cluster_treatment_average_heatmap.png"),
  treatment_cluster_heatmap_plot,
  width = 10,
  height = 5.5,
  dpi = 300
)

cluster_kegg_enrichment <- dplyr::bind_rows(lapply(cluster_levels, function(cl) {
  genes <- expression_cluster_assignments$gene_id[
    expression_cluster_assignments$expression_cluster == cl
  ]
  run_enrichment(
    genes = genes,
    universe = backgrounds[["All fat body DEGs in heatmap"]],
    term2gene = maps$term2gene_kegg,
    term2name = maps$term2name_kegg,
    database = "KEGG_pathway",
    background_label = "All fat body DEGs in heatmap"
  ) |>
    dplyr::mutate(
      expression_cluster = cl,
      n_cluster_genes = length(genes)
    )
}))

if (nrow(cluster_kegg_enrichment) > 0) {
  cluster_kegg_enrichment <- cluster_kegg_enrichment |>
    dplyr::left_join(
      maps$term2name_kegg |>
        dplyr::distinct(term, .keep_all = TRUE) |>
        dplyr::transmute(ID = term, pathway_theme, pathway_scope),
      by = "ID"
    ) |>
    dplyr::mutate(
      pathway_theme = dplyr::coalesce(pathway_theme, "Other KEGG pathway"),
      pathway_scope = dplyr::coalesce(pathway_scope,
                                      "Unclassified KEGG reference label"),
      expression_cluster = factor(expression_cluster, levels = cluster_levels),
      pathway_scope = factor(pathway_scope, levels = names(kegg_scope_colors))
    )
}

cluster_kegg_significant <- cluster_kegg_enrichment |>
  dplyr::filter(p.adjust <= params$alpha) |>
  dplyr::arrange(expression_cluster, pathway_theme, p.adjust)

cluster_kegg_summary <- cluster_kegg_significant |>
  dplyr::count(expression_cluster, pathway_theme, pathway_scope,
               name = "n_significant_pathways") |>
  dplyr::arrange(expression_cluster, pathway_theme, pathway_scope)

write.csv(cluster_kegg_enrichment,
          file.path(out_dir,
                    "expression_cluster_KEGG_enrichment_all_terms.csv"),
          row.names = FALSE)
write.csv(cluster_kegg_significant,
          file.path(out_dir,
                    "expression_cluster_KEGG_enrichment_significant_terms.csv"),
          row.names = FALSE)
write.csv(cluster_kegg_summary,
          file.path(out_dir,
                    "expression_cluster_KEGG_enrichment_count_summary.csv"),
          row.names = FALSE)

cluster_kegg_term_gene_long <- cluster_kegg_enrichment |>
  dplyr::filter(p.adjust <= params$alpha) |>
  dplyr::select(expression_cluster, database, background, pathway_theme,
                pathway_scope, ID, Description, GeneRatio, BgRatio, pvalue,
                p.adjust, qvalue, Count, geneID) |>
  tidyr::separate_rows(geneID, sep = "/") |>
  dplyr::rename(term_id = ID, term_name = Description, gene_id = geneID) |>
  dplyr::left_join(cluster_gene_annotation, by = "gene_id") |>
  dplyr::arrange(expression_cluster, p.adjust, term_id, gene_id)

cluster_kegg_term_gene_compact <- cluster_kegg_term_gene_long |>
  dplyr::group_by(expression_cluster, database, background, pathway_theme,
                  pathway_scope, term_id, term_name, GeneRatio, BgRatio,
                  p.adjust, Count) |>
  dplyr::summarise(
    DEG_gene_IDs = paste(sort(unique(gene_id)), collapse = "; "),
    DEG_gene_descriptions = paste(sort(unique(Description)), collapse = "; "),
    DEG_contexts = paste(sort(unique(DEG_context)), collapse = "; "),
    .groups = "drop"
  ) |>
  dplyr::arrange(expression_cluster, p.adjust)

write.csv(cluster_kegg_term_gene_long,
          file.path(out_dir,
                    "expression_cluster_KEGG_term_gene_index_long.csv"),
          row.names = FALSE)
write.csv(cluster_kegg_term_gene_compact,
          file.path(out_dir,
                    "expression_cluster_KEGG_term_gene_index_compact.csv"),
          row.names = FALSE)

prepare_kegg_lollipop_df <- function(tab, n_terms) {
  tab |>
    dplyr::filter(p.adjust <= params$alpha) |>
    dplyr::group_by(expression_cluster) |>
    dplyr::slice_min(p.adjust, n = n_terms, with_ties = FALSE) |>
    dplyr::ungroup() |>
    dplyr::arrange(expression_cluster, pathway_theme, p.adjust) |>
    dplyr::group_by(expression_cluster) |>
    dplyr::mutate(
      neg_log10_padj = -log10(pmax(p.adjust, .Machine$double.xmin)),
      pathway_label = stringr::str_wrap(Description, 34),
      pathway_plot_id = paste(pathway_label, dplyr::row_number(),
                              expression_cluster, sep = "___")
    ) |>
    dplyr::ungroup() |>
    dplyr::mutate(
      pathway_plot_id = factor(pathway_plot_id,
                               levels = rev(unique(pathway_plot_id))),
      expression_cluster = factor(expression_cluster, levels = cluster_levels)
    )
}

clean_kegg_labels <- function(x) {
  labels <- sub("___.*$", "", x)
  dplyr::if_else(labels == "No significant KEGG pathways", "", labels)
}

make_kegg_lollipop_plot <- function(plot_df, title, subtitle = NULL) {
  if (nrow(plot_df) == 0) {
    return(
      ggplot2::ggplot() +
        ggplot2::annotate("text", x = 0, y = 0,
                          label = "No significant KEGG pathways",
                          size = 6, fontface = "bold") +
        ggplot2::theme_void() +
        ggplot2::labs(title = title, subtitle = subtitle)
      )
  }

  x_axis_max <- max(
    c(plot_df$neg_log10_padj, -log10(params$alpha)),
    na.rm = TRUE
  )
  x_axis_limit <- x_axis_max + max(0.8, x_axis_max * 0.18)
  label_space <- x_axis_max * 2.4
  x_breaks <- pretty(c(0, x_axis_limit), n = 5)
  x_breaks <- x_breaks[x_breaks >= 0 & x_breaks <= x_axis_limit]
  missing_label_x <- max(
    c(plot_df$neg_log10_padj, -log10(params$alpha)),
    na.rm = TRUE
  ) / 2
  missing_cluster_df <- tibble::tibble(
    expression_cluster = factor(
      setdiff(cluster_levels, as.character(unique(plot_df$expression_cluster))),
      levels = cluster_levels
    ),
    pathway_plot_id = "No significant KEGG pathways",
    x = missing_label_x,
    label = "No significant KEGG pathways"
  ) |>
    dplyr::filter(!is.na(expression_cluster))

  ggplot2::ggplot(
    plot_df,
    ggplot2::aes(x = neg_log10_padj, y = pathway_plot_id)
  ) +
    ggplot2::geom_vline(
      xintercept = -log10(params$alpha),
      linetype = "dashed",
      color = "#6B7280",
      linewidth = 0.35
    ) +
    ggplot2::geom_vline(
      xintercept = 0,
      color = "#9CA3AF",
      linewidth = 0.35
    ) +
    ggplot2::geom_text(
      data = plot_df,
      ggplot2::aes(
        x = -label_space * 0.98,
        y = pathway_plot_id,
        label = clean_kegg_labels(pathway_plot_id)
      ),
      inherit.aes = FALSE,
      hjust = 0,
      size = 5.2,
      lineheight = 0.84,
      color = "#111827"
    ) +
    ggplot2::geom_segment(
      ggplot2::aes(x = 0, xend = neg_log10_padj,
                   y = pathway_plot_id, yend = pathway_plot_id),
      color = kegg_lollipop_color,
      linewidth = 0.8,
      alpha = 0.78
    ) +
    ggplot2::geom_point(
      ggplot2::aes(size = Count),
      color = kegg_lollipop_color,
      alpha = 0.95
    ) +
    ggplot2::geom_text(
      data = missing_cluster_df,
      ggplot2::aes(x = x, y = pathway_plot_id, label = label),
      inherit.aes = FALSE,
      size = 4,
      color = "#6B7280",
      fontface = "bold"
    ) +
    ggplot2::facet_wrap(~ expression_cluster, nrow = 1,
                        scales = "free_y", drop = FALSE) +
    ggplot2::scale_y_discrete(labels = function(x) rep("", length(x))) +
    ggplot2::scale_x_continuous(
      limits = c(-label_space, x_axis_limit),
      breaks = x_breaks,
      expand = c(0, 0)
    ) +
    ggplot2::scale_size_continuous(range = c(4, 10)) +
    ggplot2::labs(
      title = title,
      subtitle = subtitle,
      x = "-log10 adjusted p-value",
      y = NULL,
      size = "Cluster genes in pathway"
    ) +
    ggplot2::theme_classic(base_size = 13) +
    ggplot2::theme(
      axis.text.y = ggplot2::element_blank(),
      axis.ticks.y = ggplot2::element_blank(),
      axis.line = ggplot2::element_blank(),
      axis.line.x.bottom = ggplot2::element_line(color = "black",
                                                 linewidth = 0.35),
      panel.border = ggplot2::element_blank(),
      strip.background = ggplot2::element_rect(fill = "#F3F4F6",
                                               color = "#9CA3AF"),
      strip.text = ggplot2::element_text(face = "bold", size = 12),
      legend.position = "bottom",
      legend.box = "vertical",
      plot.title = ggplot2::element_text(face = "bold")
    )
}

kegg_lollipop_df <- prepare_kegg_lollipop_df(
  cluster_kegg_significant,
  n_terms = 8
)
kegg_lollipop_plot <- make_kegg_lollipop_plot(
  kegg_lollipop_df,
  title = "C. KEGG pathways by expression cluster",
  subtitle = "Foreground = genes in each expression cluster; background = all fat body DEGs in the heatmap"
)

gene_level_KEGG_figure <- sample_annotation_plot /
  cluster_heatmap_plot /
  kegg_lollipop_plot +
  patchwork::plot_layout(heights = c(0.95, 6.8, 3.7)) +
  patchwork::plot_annotation(
    title = "Full gene-level expression and KEGG enrichment by expression cluster",
    subtitle = "Top panel keeps all genes in the circled heatmap branch; lower panel shows KEGG pathways enriched in each cluster"
  )

ggplot2::ggsave(
  file.path(out_dir,
            "expression_cluster_gene_level_KEGG_lollipop.png"),
  gene_level_KEGG_figure,
  width = 21,
  height = 15,
  dpi = 220
)

gene_level_GO_KEGG_figure <- sample_annotation_plot /
  cluster_heatmap_plot /
  go_lollipop_plot /
  kegg_lollipop_plot +
  patchwork::plot_layout(heights = c(0.95, 6.8, 5.6, 3.7)) +
  patchwork::plot_annotation(
    title = "Full gene-level expression and functional enrichment by DEG cluster"
  )

print(gene_level_GO_KEGG_figure)
ggplot2::ggsave(
  file.path(out_dir,
            "legacy_three_block_expression_cluster_gene_level_GO_KEGG_lollipop.png"),
  gene_level_GO_KEGG_figure,
  width = 21,
  height = 22,
  dpi = 220
)

knitr::kable(
  cluster_kegg_summary,
  caption = "Significant KEGG pathways by expression cluster"
)
```

## Composite with reference tissues

This companion composite keeps the same fat-body DEG rows as panel A above, but
adds VST expression columns for the available reference samples: the two
time-course FAT phase controls and the copied head/thorax phase-reference
samples. These reference samples are not used to call the DEGs in this figure;
they are shown only to ask whether the fat-body DEG clusters also have visible
expression structure in other tissues or in the independent FAT controls.

```{r expression-cluster-reference-composite, fig.width=24, fig.height=22}
timecourse_reference_metadata <- file.path(
  timecourse_fatbody_dir,
  "gregaria_FAT_phase_reference_metadata.csv"
)

timecourse_ref_meta <- if (file.exists(timecourse_reference_metadata)) {
  read.csv(timecourse_reference_metadata, check.names = FALSE) |>
    dplyr::mutate(
      count_file = file.path(timecourse_fatbody_dir, "readcounts", file_name),
      file_present = file.exists(count_file)
    )
} else {
  tibble::tibble()
}

head_thorax_ref_meta <- parse_head_thorax_reference_metadata(
  timecourse_head_thorax_dir
) |>
  dplyr::mutate(file_present = file.exists(count_file))

fatbody_ref_meta <- if (nrow(timecourse_ref_meta) > 0) {
  timecourse_ref_meta |>
    dplyr::mutate(
      tissue = "Fat body",
      source = "Time-course fat body phase control",
      display_group = dplyr::case_when(
        phase == "Gregarious" ~ "Fat body: gregarious/crowded control",
        phase == "Solitarious" ~ "Fat body: solitarious/isolated control",
        TRUE ~ "Fat body: reference control"
      ),
      sample_label = dplyr::case_when(
        phase == "Gregarious" ~ "FAT-G",
        phase == "Solitarious" ~ "FAT-S",
        TRUE ~ sample_id
      ),
      sample_block = "FAT controls",
      treatment_or_phase = phase,
      diet = "Reference"
    ) |>
    dplyr::select(
      sample_id, count_file, tissue, phase, rearing_condition, replicate,
      source, display_group, reference_role, sample_label, sample_block,
      treatment_or_phase, diet, file_present
    )
} else {
  tibble::tibble()
}

head_thorax_ref_meta <- head_thorax_ref_meta |>
  dplyr::mutate(
    treatment_or_phase = phase,
    diet = "Reference"
  ) |>
  dplyr::select(
    sample_id, count_file, tissue, phase, rearing_condition, replicate,
    source, display_group, reference_role, sample_label, sample_block,
    treatment_or_phase, diet, file_present
  )

reference_manifest <- dplyr::bind_rows(
  fatbody_ref_meta,
  head_thorax_ref_meta
)

reference_inputs <- reference_manifest |>
  dplyr::count(sample_block, tissue, treatment_or_phase, file_present,
               name = "n_samples")
knitr::kable(
  reference_inputs,
  caption = "Reference samples available for the expanded panel A"
)

reference_ready <- nrow(reference_manifest) > 0 &&
  all(reference_manifest$file_present)

if (reference_ready) {
  # Reuse the audited, expressed main-chromosome matrix built above.
  experiment_counts <- counts
  experiment_counts <- experiment_counts[, sample_order, drop = FALSE]

  experiment_reference_meta <- sample_metadata |>
    dplyr::transmute(
      sample_id = label,
      count_file = NA_character_,
      tissue = "Fat body",
      phase = "Diet-infection experiment",
      rearing_condition = "Semi-isolated experiment",
      replicate = NA_integer_,
      source = "Diet-infection fat body experiment",
      display_group = paste(Treatment, Diet, sep = "_"),
      reference_role = paste("Fat body", Treatment, "diet", Diet),
      sample_label = label,
      sample_block = "Experiment fat body",
      treatment_or_phase = as.character(Treatment),
      diet = as.character(Diet),
      file_present = TRUE
    )

  reference_manifest <- reference_manifest |>
    dplyr::mutate(
      sample_block = factor(
        sample_block,
        levels = c("FAT controls", "Head reference", "Thorax reference")
      ),
      tissue = factor(tissue, levels = c("Fat body", "Head", "Thorax")),
      treatment_or_phase = factor(
        treatment_or_phase,
        levels = c("Control", "Infected", "Gregarious", "Solitarious")
      )
    ) |>
    dplyr::arrange(sample_block, tissue, treatment_or_phase, replicate)

  combined_reference_meta <- dplyr::bind_rows(
    experiment_reference_meta,
    reference_manifest
  )

  reference_count_tables <- Map(
    read_featurecounts_gene_counts,
    reference_manifest$count_file,
    reference_manifest$sample_id
  )
  reference_counts_df <- Reduce(
    function(x, y) dplyr::full_join(x, y, by = "gene_id"),
    reference_count_tables
  )
  reference_counts <- as.matrix(reference_counts_df[, -1, drop = FALSE])
  rownames(reference_counts) <- reference_counts_df$gene_id
  reference_counts[is.na(reference_counts)] <- 0L
  storage.mode(reference_counts) <- "integer"

  shared_reference_genes <- Reduce(
    intersect,
    list(rownames(experiment_counts), rownames(reference_counts))
  )

  combined_counts <- cbind(
    experiment_counts[shared_reference_genes, , drop = FALSE],
    reference_counts[
      shared_reference_genes,
      reference_manifest$sample_id,
      drop = FALSE
    ]
  )

  combined_reference_meta <- combined_reference_meta[
    match(colnames(combined_counts), combined_reference_meta$sample_id),
  ]
  combined_reference_meta$sample_block <- factor(
    combined_reference_meta$sample_block,
    levels = c(
      "Experiment fat body",
      "FAT controls",
      "Head reference",
      "Thorax reference"
    )
  )
  combined_reference_meta$tissue <- factor(
    combined_reference_meta$tissue,
    levels = c("Fat body", "Head", "Thorax")
  )
  combined_reference_meta$treatment_or_phase <- factor(
    combined_reference_meta$treatment_or_phase,
    levels = c("Control", "Infected", "Gregarious", "Solitarious")
  )
  combined_reference_meta$diet <- factor(
    combined_reference_meta$diet,
    levels = c("33", "50", "83", "Reference")
  )
  rownames(combined_reference_meta) <- combined_reference_meta$sample_id

  keep_reference <- rowSums(combined_counts >= params$min_count) >= params$min_samples
  dds_reference <- DESeq2::DESeqDataSetFromMatrix(
    countData = combined_counts[keep_reference, , drop = FALSE],
    colData = as.data.frame(combined_reference_meta),
    design = ~ 1
  )
  vst_reference <- DESeq2::vst(dds_reference, blind = TRUE)
  vst_reference_mat <- SummarizedExperiment::assay(vst_reference)

  reference_gene_table <- expression_cluster_assignments |>
    dplyr::filter(gene_id %in% rownames(vst_reference_mat)) |>
    dplyr::arrange(expression_cluster_order, within_cluster_order) |>
    dplyr::group_by(expression_cluster) |>
    dplyr::mutate(reference_within_cluster_order = dplyr::row_number()) |>
    dplyr::ungroup()

  reference_sample_order <- combined_reference_meta$sample_id
  reference_heatmap_z <- vst_reference_mat[
    reference_gene_table$gene_id,
    reference_sample_order,
    drop = FALSE
  ]
  reference_heatmap_z <- t(scale(t(reference_heatmap_z)))
  reference_heatmap_z[is.na(reference_heatmap_z)] <- 0
  reference_heatmap_z <- pmax(pmin(reference_heatmap_z, 3), -3)

  reference_sample_metadata <- combined_reference_meta |>
    tibble::rownames_to_column("sample_id_check") |>
    dplyr::select(-sample_id_check) |>
    dplyr::mutate(
      sample_index = dplyr::row_number(),
      sample_group_key = paste(sample_block, treatment_or_phase, diet, sep = "|")
    )

  reference_sample_ranges <- reference_sample_metadata |>
    dplyr::group_by(sample_group_key, sample_block) |>
    dplyr::summarise(start = min(sample_index), end = max(sample_index),
                     .groups = "drop") |>
    dplyr::arrange(start)
  reference_group_gaps <- reference_sample_ranges$end[
    -nrow(reference_sample_ranges)
  ] + 0.5
  reference_block_gaps <- reference_sample_ranges$end[
    which(
      as.character(reference_sample_ranges$sample_block[-nrow(reference_sample_ranges)]) !=
        as.character(reference_sample_ranges$sample_block[-1])
    )
  ] + 0.5
  reference_group_gaps <- setdiff(reference_group_gaps, reference_block_gaps)

  reference_annotation_colors <- c(
    "Tissue: Fat body" = "#D9A441",
    "Tissue: Head" = "#009E9A",
    "Tissue: Thorax" = "#C51B7D",
    "Diet: 33" = "#E6862E",
    "Diet: 50" = "#2F8F5B",
    "Diet: 83" = "#7B4FA3",
    "Diet: Reference" = "#D1D5DB",
    "State: Control" = "#FAEBD7",
    "State: Infected" = "#8B0000",
    "State: Gregarious" = "#111827",
    "State: Solitarious" = "#6B7280"
  )

  reference_annotation_long <- reference_sample_metadata |>
    dplyr::transmute(
      sample_index,
      Tissue = as.character(tissue),
      Diet = as.character(diet),
      State = as.character(treatment_or_phase)
    ) |>
    tidyr::pivot_longer(
      cols = c(Tissue, Diet, State),
      names_to = "track",
      values_to = "value"
    ) |>
    dplyr::mutate(
      track = factor(track, levels = c("Tissue", "Diet", "State")),
      track_y = as.numeric(track),
      fill_key = paste(track, value, sep = ": ")
    )

  reference_x_limits <- c(-1.25, length(reference_sample_order) + 0.5)

  reference_sample_annotation_plot <- ggplot2::ggplot(
    reference_annotation_long,
    ggplot2::aes(x = sample_index, y = track_y, fill = fill_key)
  ) +
    ggplot2::geom_tile(width = 0.95, height = 0.9) +
    ggplot2::geom_vline(xintercept = reference_group_gaps,
                        color = "black", linewidth = 0.25) +
    ggplot2::geom_vline(xintercept = reference_block_gaps,
                        color = "black", linewidth = 0.8) +
    ggplot2::scale_fill_manual(values = reference_annotation_colors,
                               drop = FALSE) +
    ggplot2::scale_x_continuous(
      limits = reference_x_limits,
      breaks = reference_sample_metadata$sample_index,
      labels = reference_sample_metadata$sample_label,
      expand = c(0, 0),
      position = "top"
    ) +
    ggplot2::scale_y_continuous(
      breaks = seq_len(3),
      labels = c("Tissue", "Diet", "State"),
      limits = c(0.45, 3.55),
      expand = c(0, 0)
    ) +
    ggplot2::labs(x = NULL, y = NULL, fill = "Sample annotation") +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      panel.grid = ggplot2::element_blank(),
      axis.text.x = ggplot2::element_blank(),
      axis.ticks.x = ggplot2::element_blank(),
      axis.text.y = ggplot2::element_text(face = "bold", size = 9),
      legend.position = "bottom",
      legend.title = ggplot2::element_text(face = "bold"),
      legend.text = ggplot2::element_text(size = 8)
    ) +
    ggplot2::guides(
      fill = ggplot2::guide_legend(nrow = 2, byrow = TRUE)
    )

  reference_gene_context_strip_df <- reference_gene_table |>
    dplyr::select(gene_id, expression_cluster, reference_within_cluster_order,
                  DEG_context) |>
    dplyr::mutate(
      DEG_context_plot = dplyr::case_when(
        DEG_context == "Diet in controls" ~ "~ diet (controls)",
        DEG_context == "Diet in infected" ~ "~ diet (infected)",
        DEG_context == "Infection within diet" ~ "~ diet + infection",
        DEG_context %in% c("Infection + diet",
                           "Diet in controls + infected") ~ "shared DEGs",
        TRUE ~ "Other fat body DEG"
      ),
      DEG_context_plot = factor(DEG_context_plot,
                                levels = deg_context_plot_levels)
    )
  present_reference_deg_context_levels <- deg_context_plot_levels[
    deg_context_plot_levels %in%
      as.character(reference_gene_context_strip_df$DEG_context_plot)
  ]
  present_reference_deg_context_labels <- stringr::str_wrap(
    present_reference_deg_context_levels,
    width = 26
  )

  reference_heatmap_df <- as.data.frame(reference_heatmap_z,
                                        check.names = FALSE) |>
    tibble::rownames_to_column("gene_id") |>
    dplyr::left_join(
      reference_gene_table |>
        dplyr::select(gene_id, expression_cluster,
                      reference_within_cluster_order),
      by = "gene_id"
    ) |>
    tidyr::pivot_longer(
      cols = dplyr::all_of(reference_sample_order),
      names_to = "sample_id",
      values_to = "z"
    ) |>
    dplyr::mutate(sample_index = match(sample_id, reference_sample_order))

  reference_cluster_heatmap_plot <- ggplot2::ggplot(
    reference_heatmap_df,
    ggplot2::aes(x = sample_index, y = reference_within_cluster_order,
                 fill = z)
  ) +
    ggplot2::geom_segment(
      data = reference_gene_context_strip_df,
      ggplot2::aes(
        x = -1.1,
        xend = 0.35,
        y = reference_within_cluster_order,
        yend = reference_within_cluster_order,
        color = DEG_context_plot
      ),
      inherit.aes = FALSE,
      linewidth = 1.05,
      lineend = "butt"
    ) +
    ggplot2::geom_raster() +
    ggplot2::geom_vline(xintercept = 0.5, color = "#9CA3AF",
                        linewidth = 0.4) +
    ggplot2::geom_vline(xintercept = reference_group_gaps,
                        color = "black", linewidth = 0.25) +
    ggplot2::geom_vline(xintercept = reference_block_gaps,
                        color = "black", linewidth = 0.8) +
    ggplot2::facet_grid(expression_cluster ~ ., scales = "free_y",
                        switch = "y") +
    ggplot2::scale_fill_gradientn(
      colors = heatmap_colors,
      limits = c(-3, 3),
      oob = scales::squish,
      name = "Row z-score"
    ) +
    ggplot2::scale_color_manual(
      values = deg_context_plot_colors,
      breaks = present_reference_deg_context_levels,
      labels = present_reference_deg_context_labels,
      drop = TRUE,
      name = "DEG context"
    ) +
    ggplot2::scale_x_continuous(
      limits = reference_x_limits,
      breaks = reference_sample_metadata$sample_index,
      labels = reference_sample_metadata$sample_label,
      expand = c(0, 0)
    ) +
    ggplot2::scale_y_reverse(expand = c(0, 0)) +
    ggplot2::labs(
      x = NULL,
      y = NULL,
      title = "A. Fat body DEG expression landscape with reference tissues",
      subtitle = paste0(
        "Rows are ", nrow(reference_gene_table),
        " fat-body DEGs; head, thorax, and FAT references are VST context columns"
      )
    ) +
    ggplot2::theme_classic(base_size = 13) +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 90, hjust = 1,
                                          vjust = 0.5, size = 7),
      axis.text.y = ggplot2::element_blank(),
      axis.ticks.y = ggplot2::element_blank(),
      strip.placement = "outside",
      strip.background.y = ggplot2::element_rect(fill = "#F3F4F6",
                                                 color = "#9CA3AF"),
      strip.text.y.left = ggplot2::element_text(angle = 90, face = "bold",
                                                size = 13),
      panel.spacing.y = grid::unit(0.25, "lines"),
      legend.position = "right",
      legend.box = "vertical",
      legend.title = ggplot2::element_text(face = "bold"),
      legend.text = ggplot2::element_text(size = 9),
      plot.title = ggplot2::element_text(face = "bold")
    ) +
    ggplot2::guides(
      fill = ggplot2::guide_colorbar(
        order = 1,
        title.position = "top",
        barheight = grid::unit(3.0, "cm"),
        barwidth = grid::unit(0.35, "cm")
      ),
      color = ggplot2::guide_legend(
        order = 2,
        title.position = "top",
        override.aes = list(linewidth = 3)
      )
    )

  reference_gene_level_GO_KEGG_figure <- reference_sample_annotation_plot /
    reference_cluster_heatmap_plot /
    go_lollipop_plot /
    kegg_lollipop_plot +
    patchwork::plot_layout(heights = c(1.05, 6.8, 5.6, 3.7)) +
    patchwork::plot_annotation(
      title = "Fat body DEG expression clusters with reference tissues",
      subtitle = paste0(
        "DEGs were called only in the fat body diet-infection experiment; ",
        "reference columns show VST row-z-scores for context"
      )
    )

  print(reference_gene_level_GO_KEGG_figure)
  ggplot2::ggsave(
    file.path(out_dir,
              "expression_cluster_gene_level_GO_KEGG_lollipop_with_references.png"),
    reference_gene_level_GO_KEGG_figure,
    width = 24,
    height = 22,
    dpi = 220
  )
  grDevices::svg(
    file.path(out_dir,
              "expression_cluster_gene_level_GO_KEGG_lollipop_with_references.svg"),
    width = 24,
    height = 22,
    onefile = FALSE
  )
  print(reference_gene_level_GO_KEGG_figure)
  grDevices::dev.off()

  ggplot2::ggsave(
    file.path(out_dir,
              "expression_cluster_reference_panel_A_heatmap.png"),
    reference_sample_annotation_plot / reference_cluster_heatmap_plot +
      patchwork::plot_layout(heights = c(1.05, 6.8)),
    width = 24,
    height = 11,
    dpi = 260
  )

  write.csv(
    reference_gene_table,
    file.path(out_dir,
              "expression_cluster_reference_panel_A_genes.csv"),
    row.names = FALSE
  )
  write.csv(
    combined_reference_meta,
    file.path(out_dir,
              "expression_cluster_reference_panel_A_samples.csv"),
    row.names = FALSE
  )
} else {
  message("Reference composite skipped because one or more reference count files are missing.")
}
```

## Full heatmap and gene-level cladogram {#Cluster_cladogram}

This is the re-delimitation workspace. It contains every significant
main-chromosome DEG in the current all-samples analysis and does **not** draw any
provisional cluster boundaries. The dendrogram and heatmap are generated from
the same row-z-scored VST matrix, the same complete-linkage tree, and the same
leaf order. Thus, a branch seen in the tree aligns exactly with the genes on the
same horizontal band of the heatmap. The tree describes expression-profile
similarity; it is not a phylogeny.

```{r canonical-full-gene-tree-heatmap, fig.width=30, fig.height=24}
canonical_gene_dendro <- ggdendro::dendro_data(
  stats::as.dendrogram(row_tree),
  type = "rectangle"
)

canonical_gene_leaves <- ggdendro::label(canonical_gene_dendro) |>
  dplyr::rename(gene_id = label, leaf_x = x, leaf_y = y) |>
  dplyr::arrange(leaf_x) |>
  dplyr::mutate(canonical_gene_order = dplyr::row_number())

canonical_order_check <- data.frame(
  check = "Gene-level dendrogram leaf order matches the clustered heatmap order",
  value = identical(canonical_gene_leaves$gene_id, ordered_genes)
)
if (!canonical_order_check$value) {
  stop("Canonical dendrogram leaves do not match the clustered heatmap order.")
}

canonical_gene_table <- canonical_gene_leaves |>
  dplyr::select(canonical_gene_order, gene_id) |>
  dplyr::left_join(unmerged_gene_table, by = "gene_id") |>
  dplyr::arrange(canonical_gene_order)

write.csv(
  canonical_gene_table,
  file.path(out_dir, "full_heatmap_canonical_gene_order.csv"),
  row.names = FALSE
)
write.csv(
  canonical_order_check,
  file.path(out_dir, "full_heatmap_canonical_order_check.csv"),
  row.names = FALSE
)

canonical_gene_order <- canonical_gene_leaves$gene_id
canonical_heatmap_matrix <- full_heatmap_z_input[
  canonical_gene_order,
  sample_order,
  drop = FALSE
]
canonical_heatmap_matrix <- pmax(pmin(canonical_heatmap_matrix, 3), -3)

canonical_heatmap_df <- as.data.frame(
  canonical_heatmap_matrix,
  check.names = FALSE
) |>
  tibble::rownames_to_column("gene_id") |>
  dplyr::mutate(gene_y = match(gene_id, canonical_gene_order)) |>
  tidyr::pivot_longer(
    cols = dplyr::all_of(sample_order),
    names_to = "sample_id",
    values_to = "z"
  ) |>
  dplyr::mutate(sample_index = match(sample_id, sample_order))

n_canonical_genes <- length(canonical_gene_order)
canonical_y_limits <- c(n_canonical_genes + 0.5, 0.5)

canonical_tree_segments <- ggdendro::segment(canonical_gene_dendro) |>
  dplyr::transmute(
    tree_x = y,
    gene_y = x,
    tree_xend = yend,
    gene_yend = xend
  )

canonical_vertical_tree_plot <- ggplot2::ggplot(
  canonical_tree_segments,
  ggplot2::aes(
    x = tree_x,
    y = gene_y,
    xend = tree_xend,
    yend = gene_yend
  )
) +
  ggplot2::geom_segment(linewidth = 0.14, color = "#263241",
                        lineend = "square") +
  ggplot2::scale_x_reverse(expand = ggplot2::expansion(mult = c(0.01, 0.02))) +
  ggplot2::scale_y_reverse(limits = canonical_y_limits, expand = c(0, 0)) +
  ggplot2::labs(x = "Tree height", y = NULL) +
  ggplot2::theme_classic(base_size = 13) +
  ggplot2::theme(
    axis.text.y = ggplot2::element_blank(),
    axis.ticks.y = ggplot2::element_blank(),
    axis.line.y = ggplot2::element_blank(),
    plot.margin = ggplot2::margin(0, 0, 3, 3)
  )

canonical_heatmap_plot <- ggplot2::ggplot(
  canonical_heatmap_df,
  ggplot2::aes(x = sample_index, y = gene_y, fill = z)
) +
  ggplot2::geom_raster() +
  ggplot2::geom_vline(xintercept = diet_gaps, color = "black",
                      linewidth = 0.25) +
  ggplot2::geom_vline(xintercept = treatment_gaps, color = "black",
                      linewidth = 0.7) +
  ggplot2::scale_fill_gradientn(
    colors = heatmap_colors,
    limits = c(-3, 3),
    oob = scales::squish,
    name = "Row z-score"
  ) +
  ggplot2::scale_x_continuous(
    limits = c(0.5, length(sample_order) + 0.5),
    breaks = sample_metadata$sample_index,
    labels = sample_metadata$label,
    expand = c(0, 0)
  ) +
  ggplot2::scale_y_reverse(limits = canonical_y_limits, expand = c(0, 0)) +
  ggplot2::labs(x = NULL, y = NULL) +
  ggplot2::theme_classic(base_size = 13) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, hjust = 1,
                                        vjust = 0.5, size = 12),
    axis.text.y = ggplot2::element_blank(),
    axis.ticks.y = ggplot2::element_blank(),
    axis.line = ggplot2::element_blank(),
    legend.position = "right",
    legend.title = ggplot2::element_text(face = "bold"),
    plot.margin = ggplot2::margin(0, 4, 3, 0)
  ) +
  ggplot2::guides(
    fill = ggplot2::guide_colorbar(
      title.position = "top",
      barheight = grid::unit(4.5, "cm"),
      barwidth = grid::unit(0.45, "cm")
    )
  )

canonical_sample_annotation_plot <- sample_annotation_plot +
  ggplot2::scale_x_continuous(
    limits = c(0.5, length(sample_order) + 0.5),
    breaks = sample_metadata$sample_index,
    labels = sample_metadata$label,
    expand = c(0, 0),
    position = "top"
  ) +
  ggplot2::labs(
    title = paste0(
      "Full main-chromosome DEG heatmap with aligned gene dendrogram (",
      n_canonical_genes, " genes)"
    )
  ) +
  ggplot2::theme(legend.position = "none")

canonical_tree_column <- (
  patchwork::plot_spacer() /
    canonical_vertical_tree_plot
) + patchwork::plot_layout(heights = c(1.0, 12))
canonical_heatmap_column <- (
  canonical_sample_annotation_plot /
    canonical_heatmap_plot
) + patchwork::plot_layout(heights = c(1.0, 12))

canonical_tree_heatmap_figure <- (
  canonical_tree_column | canonical_heatmap_column
) + patchwork::plot_layout(widths = c(3.8, 22))

print(canonical_tree_heatmap_figure)
ggplot2::ggsave(
  file.path(out_dir, "full_canonical_gene_tree_heatmap.png"),
  canonical_tree_heatmap_figure,
  width = 30,
  height = 24,
  dpi = 260,
  limitsize = FALSE
)
ggplot2::ggsave(
  file.path(out_dir, "full_canonical_gene_tree_heatmap.pdf"),
  canonical_tree_heatmap_figure,
  width = 30,
  height = 24,
  limitsize = FALSE
)

knitr::kable(
  canonical_order_check,
  caption = "Required alignment check for the full dendrogram and heatmap"
)
```

### Full unpartitioned cladogram

The landscape version below shows the complete tree without cluster colors,
labels, or cut lines. It is intended for choosing biologically and visually
defensible branch boundaries before rerunning enrichment.

```{r full-gene-level-cladogram-unpartitioned, fig.width=28, fig.height=9}
full_gene_level_cladogram_plot <- ggplot2::ggplot() +
  ggplot2::geom_segment(
    data = ggdendro::segment(canonical_gene_dendro),
    ggplot2::aes(x = x, y = y, xend = xend, yend = yend),
    linewidth = 0.14,
    color = "#263241",
    lineend = "square"
  ) +
  ggplot2::labs(
    title = "Full unpartitioned gene-level cladogram",
    subtitle = paste0(
      n_canonical_genes,
      " main-chromosome DEG leaves; no cluster delimitation imposed"
    ),
    x = NULL,
    y = "Euclidean distance between gene expression profiles"
  ) +
  ggplot2::theme_classic(base_size = 14) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank(),
    plot.title = ggplot2::element_text(face = "bold")
  )

print(full_gene_level_cladogram_plot)
ggplot2::ggsave(
  file.path(out_dir, "full_gene_level_cladogram_unpartitioned.png"),
  full_gene_level_cladogram_plot,
  width = 28,
  height = 9,
  dpi = 300
)
ggplot2::ggsave(
  file.path(out_dir, "full_gene_level_cladogram_unpartitioned.pdf"),
  full_gene_level_cladogram_plot,
  width = 28,
  height = 9
)
```

### Proposed delimitations {.tabset}

The three tabs below apply candidate cuts at `k = 2`, `k = 3`, and `k = 4`
to the same complete-linkage tree. The dashed vertical line marks the cut height
on the dendrogram. The colored strip and horizontal boundaries identify the
resulting contiguous gene blocks on the heatmap. Cluster numbers follow the
top-to-bottom heatmap order and are only provisional labels.

```{r build-candidate-cut-figures, results='hide'}
candidate_cut_palette <- c(
  "Candidate cluster 1" = "#315C8A",
  "Candidate cluster 2" = "#C85A54",
  "Candidate cluster 3" = "#2F8F5B",
  "Candidate cluster 4" = "#D49A2A"
)

candidate_cut_height <- function(tree, k) {
  lower_index <- length(tree$height) - k + 1L
  mean(tree$height[c(lower_index, lower_index + 1L)])
}

build_candidate_cut_figure <- function(k) {
  raw_membership <- stats::cutree(row_tree, k = k)[canonical_gene_order]
  top_to_bottom_ids <- unique(raw_membership)
  display_membership <- match(raw_membership, top_to_bottom_ids)
  display_levels <- paste("Candidate cluster", seq_len(k))

  membership_df <- tibble::tibble(
    gene_id = canonical_gene_order,
    gene_y = seq_along(canonical_gene_order),
    candidate_cluster_number = display_membership,
    candidate_cluster = factor(
      paste("Candidate cluster", display_membership),
      levels = display_levels
    )
  )

  membership_runs <- rle(display_membership)
  if (length(membership_runs$lengths) != k) {
    stop("Candidate tree cut did not produce contiguous heatmap branches for k = ", k)
  }
  branch_ends <- cumsum(membership_runs$lengths)
  branch_starts <- c(1L, head(branch_ends, -1L) + 1L)
  branch_spans <- tibble::tibble(
    k = k,
    candidate_cluster_number = seq_len(k),
    candidate_cluster = factor(display_levels, levels = display_levels),
    start = branch_starts,
    end = branch_ends,
    midpoint = (branch_starts + branch_ends) / 2,
    n_genes = membership_runs$lengths
  )
  boundaries <- head(branch_ends, -1L) + 0.5
  cut_height <- candidate_cut_height(row_tree, k)

  candidate_tree_plot <- canonical_vertical_tree_plot +
    ggplot2::geom_vline(
      xintercept = cut_height,
      color = "#8B0000",
      linetype = "22",
      linewidth = 0.8
    ) +
    ggplot2::geom_hline(
      yintercept = boundaries,
      color = "#111827",
      linewidth = 0.35
    )

  candidate_strip_plot <- ggplot2::ggplot(
    membership_df,
    ggplot2::aes(x = 1, y = gene_y, fill = candidate_cluster)
  ) +
    ggplot2::geom_tile(width = 0.92, height = 1) +
    ggplot2::geom_hline(
      yintercept = boundaries,
      color = "#111827",
      linewidth = 0.8
    ) +
    ggplot2::geom_label(
      data = branch_spans,
      ggplot2::aes(
        x = 1,
        y = midpoint,
        label = paste0("C", candidate_cluster_number, "\n", n_genes, " genes")
      ),
      inherit.aes = FALSE,
      size = 4.2,
      fontface = "bold",
      color = "#111827",
      fill = "white",
      label.size = 0.2,
      label.padding = grid::unit(0.14, "lines")
    ) +
    ggplot2::scale_fill_manual(
      values = candidate_cut_palette[display_levels],
      drop = FALSE,
      guide = "none"
    ) +
    ggplot2::scale_x_continuous(
      limits = c(0.5, 1.5),
      breaks = 1,
      labels = paste0("k = ", k),
      position = "top",
      expand = c(0, 0)
    ) +
    ggplot2::scale_y_reverse(limits = canonical_y_limits, expand = c(0, 0)) +
    ggplot2::labs(x = NULL, y = NULL) +
    ggplot2::theme_classic(base_size = 13) +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(face = "bold", size = 13),
      axis.ticks = ggplot2::element_blank(),
      axis.text.y = ggplot2::element_blank(),
      axis.line = ggplot2::element_blank(),
      plot.margin = ggplot2::margin(0, 2, 3, 2)
    )

  candidate_heatmap_plot <- canonical_heatmap_plot +
    ggplot2::geom_hline(
      yintercept = boundaries,
      color = "#111827",
      linewidth = 0.8
    )

  candidate_annotation_plot <- canonical_sample_annotation_plot +
    ggplot2::labs(
      title = paste0(
        "Candidate k = ", k, " delimitation of ",
        n_canonical_genes, " main-chromosome DEGs"
      ),
      subtitle = paste0(
        "Dashed tree line = cut height ", round(cut_height, 2),
        "; heatmap boundaries follow the resulting branches"
      )
    )

  tree_column <- (
    patchwork::plot_spacer() /
      candidate_tree_plot
  ) + patchwork::plot_layout(heights = c(1.0, 12))
  strip_column <- (
    patchwork::plot_spacer() /
      candidate_strip_plot
  ) + patchwork::plot_layout(heights = c(1.0, 12))
  heatmap_column <- (
    candidate_annotation_plot /
      candidate_heatmap_plot
  ) + patchwork::plot_layout(heights = c(1.0, 12))

  figure <- (
    tree_column | strip_column | heatmap_column
  ) + patchwork::plot_layout(widths = c(3.8, 2.0, 22))

  list(
    figure = figure,
    membership = membership_df,
    spans = branch_spans,
    boundaries = boundaries,
    cut_height = cut_height
  )
}

candidate_cut_results <- stats::setNames(
  lapply(2:4, build_candidate_cut_figure),
  as.character(2:4)
)

candidate_assignment_table <- canonical_gene_table |>
  dplyr::select(canonical_gene_order, gene_id, Description, gene_biotype,
                DEG_context, comparisons, directions, max_abs_log2FC, min_padj)
for (k in 2:4) {
  candidate_assignment_table[[paste0("candidate_k", k)]] <-
    as.character(candidate_cut_results[[as.character(k)]]$membership$candidate_cluster)
}
candidate_cut_span_table <- dplyr::bind_rows(lapply(
  candidate_cut_results,
  function(x) x$spans
))

write.csv(
  candidate_assignment_table,
  file.path(out_dir, "candidate_k2_k3_k4_gene_assignments.csv"),
  row.names = FALSE
)
write.csv(
  candidate_cut_span_table,
  file.path(out_dir, "candidate_k2_k3_k4_branch_spans.csv"),
  row.names = FALSE
)

for (k in 2:4) {
  candidate_result <- candidate_cut_results[[as.character(k)]]
  ggplot2::ggsave(
    file.path(out_dir, paste0("candidate_k", k, "_gene_tree_heatmap.png")),
    candidate_result$figure,
    width = 32,
    height = 24,
    dpi = 260,
    limitsize = FALSE
  )
  ggplot2::ggsave(
    file.path(out_dir, paste0("candidate_k", k, "_gene_tree_heatmap.pdf")),
    candidate_result$figure,
    width = 32,
    height = 24,
    limitsize = FALSE
  )
}
```

#### Two clusters

```{r show-candidate-k2, echo=FALSE, fig.width=30, fig.height=22}
print(candidate_cut_results[["2"]]$figure)
```

#### Three clusters

```{r show-candidate-k3, echo=FALSE, fig.width=30, fig.height=22}
print(candidate_cut_results[["3"]]$figure)
```

#### Four clusters

```{r show-candidate-k4, echo=FALSE, fig.width=30, fig.height=22}
print(candidate_cut_results[["4"]]$figure)
```

### Candidate tree cuts

These diagnostics compare several possible horizontal cuts of the same tree.
The silhouette score summarizes expression cohesion and separation, while the
branch sizes reveal whether a solution is dominated by very small groups. They
are decision aids, not an automatic biological definition of a cluster.

```{r candidate-tree-cut-diagnostics}
candidate_k_diagnostics <- dplyr::bind_rows(lapply(2:10, function(k) {
  memberships <- stats::cutree(row_tree, k = k)
  branch_sizes <- sort(table(memberships), decreasing = TRUE)
  silhouette_values <- cluster::silhouette(memberships, row_distance)[, "sil_width"]
  tibble::tibble(
    k = k,
    mean_silhouette = mean(silhouette_values),
    median_silhouette = stats::median(silhouette_values),
    smallest_branch = min(branch_sizes),
    largest_branch = max(branch_sizes),
    branch_sizes = paste(branch_sizes, collapse = "; ")
  )
}))

write.csv(
  candidate_k_diagnostics,
  file.path(out_dir, "candidate_tree_cut_diagnostics.csv"),
  row.names = FALSE
)

knitr::kable(
  candidate_k_diagnostics,
  digits = 3,
  caption = "Candidate cuts of the full main-chromosome DEG expression tree"
)
```

## Primary two-cluster GO enrichment

The primary interpretation keeps every DEG in the statistically strongest
`k = 2` division. The narrow subbranches visible inside the two large blocks are
not removed: they are coherent parts of the expression tree and may represent
real specialized responses. Removing them solely because they look unusual
would make the enrichment dependent on visual judgement.

Three foregrounds are tested independently against the same expressed,
rRNA-filtered main-chromosome background:

1. all significant main-chromosome DEGs, to summarize the overall response;
2. candidate expression cluster 1;
3. candidate expression cluster 2.

The 71- and 45-gene subbranches from the `k = 4` view remain nested within these
two primary clusters. Their separate enrichment can be used later as a
sensitivity analysis without changing the primary gene sets.

```{r primary-k2-go-enrichment, fig.width=36, fig.height=24}
k2_membership <- candidate_cut_results[["2"]]$membership
primary_k2_set_levels <- c(
  "All main-chromosome DEGs",
  "Expression cluster 1",
  "Expression cluster 2"
)
primary_k2_gene_sets <- list(
  "All main-chromosome DEGs" = canonical_gene_order,
  "Expression cluster 1" = k2_membership$gene_id[
    k2_membership$candidate_cluster == "Candidate cluster 1"
  ],
  "Expression cluster 2" = k2_membership$gene_id[
    k2_membership$candidate_cluster == "Candidate cluster 2"
  ]
)

go_annotated_genes <- unique(maps$term2gene_go$gene)
primary_k2_set_summary <- dplyr::bind_rows(lapply(
  names(primary_k2_gene_sets),
  function(set_name) {
    genes <- primary_k2_gene_sets[[set_name]]
    tibble::tibble(
      foreground_set = set_name,
      n_genes = length(genes),
      n_GO_annotated_genes = length(intersect(genes, go_annotated_genes)),
      GO_annotation_percent = 100 * n_GO_annotated_genes / n_genes
    )
  }
))

primary_k2_go_enrichment <- dplyr::bind_rows(lapply(
  names(primary_k2_gene_sets),
  function(set_name) {
    run_enrichment(
      genes = primary_k2_gene_sets[[set_name]],
      universe = expressed_universe,
      term2gene = maps$term2gene_go,
      term2name = maps$term2name_go,
      database = "GO",
      background_label = "Expressed rRNA-filtered main-chromosome genes"
    ) |>
      dplyr::mutate(foreground_set = set_name)
  }
))

if (nrow(primary_k2_go_enrichment) > 0) {
  primary_k2_go_enrichment <- primary_k2_go_enrichment |>
    dplyr::left_join(
      maps$term2name_go |>
        dplyr::distinct(term, .keep_all = TRUE) |>
        dplyr::transmute(
          ID = term,
          ontology = dplyr::recode(
            ontology,
            BP = "Biological process",
            CC = "Cellular component",
            MF = "Molecular function",
            .default = NA_character_
          )
        ),
      by = "ID"
    ) |>
    dplyr::filter(!is.na(ontology)) |>
    dplyr::mutate(
      foreground_set = factor(foreground_set,
                              levels = primary_k2_set_levels),
      ontology = factor(ontology, levels = names(go_ontology_colors))
    )
}

primary_k2_go_significant <- primary_k2_go_enrichment |>
  dplyr::filter(p.adjust <= params$alpha) |>
  dplyr::arrange(foreground_set, ontology, p.adjust)

primary_k2_go_term_gene_index <- primary_k2_go_significant |>
  dplyr::select(foreground_set, ontology, ID, Description, GeneRatio, BgRatio,
                pvalue, p.adjust, qvalue, Count, geneID) |>
  tidyr::separate_rows(geneID, sep = "/") |>
  dplyr::rename(term_id = ID, term_name = Description, gene_id = geneID) |>
  dplyr::left_join(
    canonical_gene_table |>
      dplyr::select(gene_id, Description, gene_biotype, DEG_context,
                    comparisons, directions, max_abs_log2FC, min_padj),
    by = "gene_id"
  ) |>
  dplyr::arrange(foreground_set, ontology, p.adjust, term_id, gene_id)

write.csv(
  primary_k2_set_summary,
  file.path(out_dir, "primary_k2_GO_foreground_summary.csv"),
  row.names = FALSE
)
write.csv(
  primary_k2_go_enrichment,
  file.path(out_dir, "primary_k2_GO_enrichment_all_terms.csv"),
  row.names = FALSE
)
write.csv(
  primary_k2_go_significant,
  file.path(out_dir, "primary_k2_GO_enrichment_significant_terms.csv"),
  row.names = FALSE
)
write.csv(
  primary_k2_go_term_gene_index,
  file.path(out_dir, "primary_k2_GO_term_gene_index_long.csv"),
  row.names = FALSE
)

primary_k2_go_plot_df <- primary_k2_go_significant |>
  dplyr::group_by(foreground_set, ontology) |>
  dplyr::slice_min(p.adjust, n = 10, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::arrange(foreground_set, ontology, p.adjust) |>
  dplyr::mutate(
    neg_log10_padj = -log10(pmax(p.adjust, .Machine$double.xmin)),
    term_label = stringr::str_wrap(Description, 52),
    term_plot_id = paste(foreground_set, ontology, ID, term_label,
                         sep = "___"),
    term_plot_id = factor(term_plot_id,
                          levels = rev(unique(term_plot_id))),
    facet_panel = paste(foreground_set, ontology, sep = "\n")
  )

primary_k2_panel_levels <- unlist(lapply(
  names(go_ontology_colors),
  function(ontology_name) {
    paste(primary_k2_set_levels, ontology_name, sep = "\n")
  }
))
primary_k2_go_plot_df$facet_panel <- factor(
  primary_k2_go_plot_df$facet_panel,
  levels = primary_k2_panel_levels
)

clean_primary_go_label <- function(x) {
  sub("^.*___.*___.*___", "", x)
}

if (nrow(primary_k2_go_plot_df) > 0) {
  primary_k2_go_plot <- ggplot2::ggplot(
    primary_k2_go_plot_df,
    ggplot2::aes(x = neg_log10_padj, y = term_plot_id)
  ) +
    ggplot2::geom_vline(
      xintercept = -log10(params$alpha),
      linetype = "dashed",
      color = "#6B7280",
      linewidth = 0.4
    ) +
    ggplot2::geom_segment(
      ggplot2::aes(x = 0, xend = neg_log10_padj,
                   y = term_plot_id, yend = term_plot_id),
      linewidth = 0.65,
      color = "#7A7A7A"
    ) +
    ggplot2::geom_point(
      ggplot2::aes(size = Count, color = ontology),
      alpha = 0.95
    ) +
    ggplot2::facet_wrap(
      ~ facet_panel,
      ncol = 3,
      scales = "free_y"
    ) +
    ggplot2::scale_color_manual(values = go_ontology_colors, drop = FALSE) +
    ggplot2::scale_size_continuous(range = c(4.5, 13)) +
    ggplot2::scale_y_discrete(labels = clean_primary_go_label) +
    ggplot2::labs(
      title = "GO enrichment for all DEGs and the two primary expression clusters",
      subtitle = paste0(
        "All foregrounds use the same expressed main-chromosome background; ",
        "top 10 terms per ontology and foreground"
      ),
      x = "-log10 adjusted p-value",
      y = NULL,
      color = "GO ontology",
      size = "DEGs in term"
    ) +
    ggplot2::theme_bw(base_size = 14) +
    ggplot2::theme(
      axis.text.y = ggplot2::element_text(size = 12, color = "#111827"),
      axis.text.x = ggplot2::element_text(size = 12),
      strip.text = ggplot2::element_text(size = 14, face = "bold"),
      legend.position = "bottom",
      legend.title = ggplot2::element_text(face = "bold"),
      panel.grid.minor = ggplot2::element_blank(),
      plot.title = ggplot2::element_text(face = "bold", size = 18)
    )

  print(primary_k2_go_plot)
  ggplot2::ggsave(
    file.path(out_dir, "primary_k2_GO_enrichment_lollipop.png"),
    primary_k2_go_plot,
    width = 36,
    height = 24,
    dpi = 260,
    limitsize = FALSE
  )
  ggplot2::ggsave(
    file.path(out_dir, "primary_k2_GO_enrichment_lollipop.pdf"),
    primary_k2_go_plot,
    width = 36,
    height = 24,
    limitsize = FALSE
  )
}

knitr::kable(
  primary_k2_set_summary,
  digits = 1,
  caption = "Primary all-DEG and two-cluster GO foregrounds"
)
```

### Significant GO terms

```{r primary-k2-go-table}
DT::datatable(
  primary_k2_go_significant |>
    dplyr::select(foreground_set, ontology, ID, Description, GeneRatio,
                  BgRatio, pvalue, p.adjust, qvalue, Count,
                  foreground_genes_tested, background_genes_tested),
  rownames = FALSE,
  filter = "top",
  options = list(pageLength = 20, scrollX = TRUE),
  caption = "Significant GO terms for all DEGs and the two primary clusters"
)
```

### GO term-to-gene index

```{r primary-k2-go-reverse-index}
DT::datatable(
  primary_k2_go_term_gene_index,
  rownames = FALSE,
  filter = "top",
  options = list(pageLength = 20, scrollX = TRUE),
  caption = "Genes contributing to each significant primary GO term"
)
```

## Figure 3: two expression clusters {#Publication_figure}

The publication composite uses the two primary expression clusters. Panel A
aligns the 0-0.1% fungal-alignment zoom, treatment and diet annotations, and the
complete main-chromosome DEG heatmap by sample. Values above 0.1% are capped
only for display and retain their full percentage as enlarged white labels.
The row z-score and DEG-context guides sit in a horizontal legend strip below
panel A so the heatmap keeps its full width and the right side remains free for
an external graphical annotation. Panel A uses 80% of the exported canvas;
the remaining 20% is intentionally blank on the right for later figure assembly.
Panels B and C use only genes called differentially expressed for infected
versus control within at least one diet, then test cluster 1 and cluster 2
separately against the expressed, rRNA-filtered main-chromosome background. The
broader all-DEG GO
analysis remains available above as supporting context but is not repeated in
Figure 3. The primary KEGG panel uses the eukaryote-inclusive view: it retains
eukaryote/metazoan-compatible, conserved cross-domain, and broad KEGG reference
maps while excluding explicitly microbial/prokaryotic and disease-reference
labels. Figure 3 is exported with either a KEGG lollipop panel or the locust
pathway map. A strict insect-focused version remains available as a secondary
export, while all tested pathway scopes and genes remain in the audit tables.

```{r figure3-k2-build, fig.width=24, fig.height=24}
figure3_cluster_levels <- c("DEG cluster 1", "DEG cluster 2")

figure3_k2_assignments <- canonical_gene_table |>
  dplyr::left_join(
    k2_membership |>
      dplyr::select(gene_id, candidate_cluster),
    by = "gene_id"
  ) |>
  dplyr::mutate(
    expression_cluster = dplyr::recode(
      as.character(candidate_cluster),
      "Candidate cluster 1" = "DEG cluster 1",
      "Candidate cluster 2" = "DEG cluster 2"
    ),
    expression_cluster = factor(expression_cluster,
                                levels = figure3_cluster_levels),
    DEG_context = dplyr::coalesce(as.character(DEG_context),
                                  "Other fat body DEG"),
    DEG_context = factor(DEG_context, levels = deg_context_levels)
  ) |>
  dplyr::arrange(
    expression_cluster,
    DEG_context,
    min_padj,
    dplyr::desc(max_abs_log2FC),
    canonical_gene_order
  ) |>
  dplyr::group_by(expression_cluster) |>
  dplyr::mutate(within_cluster_order = dplyr::row_number()) |>
  dplyr::ungroup()

write.csv(
  figure3_k2_assignments,
  file.path(out_dir, "figure3_k2_gene_assignments.csv"),
  row.names = FALSE
)

figure3_within_diet_sets <- list(
  "Expression cluster 1" = figure3_k2_assignments$gene_id[
    figure3_k2_assignments$expression_cluster == "DEG cluster 1" &
      figure3_k2_assignments$DEG_context == "Infection within diet"
  ],
  "Expression cluster 2" = figure3_k2_assignments$gene_id[
    figure3_k2_assignments$expression_cluster == "DEG cluster 2" &
      figure3_k2_assignments$DEG_context == "Infection within diet"
  ]
)

figure3_foreground_summary <- tibble::tibble(
  foreground_set = names(figure3_within_diet_sets),
  n_cluster_genes = vapply(
    seq_along(figure3_within_diet_sets),
    function(i) {
      sum(as.character(figure3_k2_assignments$expression_cluster) ==
            figure3_cluster_levels[[i]])
    },
    integer(1)
  ),
  n_infection_within_diet_genes = vapply(
    figure3_within_diet_sets,
    length,
    integer(1)
  ),
  GO_annotated_genes = vapply(
    figure3_within_diet_sets,
    function(x) length(intersect(x, unique(maps$term2gene_go$gene))),
    integer(1)
  ),
  KEGG_annotated_genes = vapply(
    figure3_within_diet_sets,
    function(x) length(intersect(x, unique(maps$term2gene_kegg$gene))),
    integer(1)
  )
)
write.csv(
  figure3_foreground_summary,
  file.path(out_dir, "figure3_infection_within_diet_foreground_summary.csv"),
  row.names = FALSE
)

figure3_heatmap_matrix <- full_heatmap_z[
  figure3_k2_assignments$gene_id,
  sample_order,
  drop = FALSE
]
figure3_heatmap_matrix <- pmax(pmin(figure3_heatmap_matrix, 3), -3)
figure3_heatmap_df <- as.data.frame(
  figure3_heatmap_matrix,
  check.names = FALSE
) |>
  tibble::rownames_to_column("gene_id") |>
  dplyr::left_join(
    figure3_k2_assignments |>
      dplyr::select(gene_id, expression_cluster, within_cluster_order,
                    DEG_context),
    by = "gene_id"
  ) |>
  tidyr::pivot_longer(
    cols = dplyr::all_of(sample_order),
    names_to = "sample_id",
    values_to = "z"
  ) |>
  dplyr::mutate(sample_index = match(sample_id, sample_order))

figure3_context_strip <- figure3_k2_assignments |>
  dplyr::transmute(
    gene_id,
    expression_cluster,
    within_cluster_order,
    DEG_context_plot = dplyr::case_when(
      DEG_context == "Diet in controls" ~ "~ diet (controls)",
      DEG_context == "Diet in infected" ~ "~ diet (infected)",
      DEG_context == "Infection within diet" ~ "~ diet + infection",
      DEG_context %in% c("Infection + diet",
                         "Diet in controls + infected") ~ "shared DEGs",
      TRUE ~ "Other fat body DEG"
    ),
    DEG_context_plot = factor(DEG_context_plot,
                              levels = deg_context_plot_levels)
  )
figure3_present_contexts <- deg_context_plot_levels[
  deg_context_plot_levels %in% as.character(figure3_context_strip$DEG_context_plot)
]

figure3_sample_annotation_plot <- sample_annotation_plot +
  ggplot2::labs(title = NULL) +
  ggplot2::theme(
    legend.position = "none",
    axis.text.y = ggplot2::element_text(face = "bold", size = 15),
    axis.text.y.right = ggplot2::element_text(face = "bold", size = 15),
    plot.margin = ggplot2::margin(0, 4, 0, 4)
  )

figure3_mapping_summary <- readr::read_tsv(
  mapping_summary_file,
  show_col_types = FALSE
) |>
  dplyr::transmute(
    label = as.character(sample_id),
    pct_unique_fungus = as.numeric(pct_unique_fungus)
  )

figure3_fungal_percentage <- sample_metadata |>
  dplyr::select(
    label, sample_index, Treatment_plot, Diet_plot, sample_group
  ) |>
  dplyr::left_join(figure3_mapping_summary, by = "label")
if (anyNA(figure3_fungal_percentage$pct_unique_fungus)) {
  stop("Missing fungal-alignment percentage for one or more Figure 3 samples.")
}

figure3_percentage_limit <- 0.1
figure3_control_mean_percentage <- mean(
  figure3_fungal_percentage$pct_unique_fungus[
    figure3_fungal_percentage$Treatment_plot == "Control"
  ],
  na.rm = TRUE
)
figure3_fungal_percentage <- figure3_fungal_percentage |>
  dplyr::mutate(
    display_percentage = pmin(
      pct_unique_fungus,
      figure3_percentage_limit
    )
  )

figure3_fungal_percentage_plot <- ggplot2::ggplot(
  figure3_fungal_percentage,
  ggplot2::aes(
    x = sample_index,
    y = display_percentage,
    fill = Treatment_plot
  )
) +
  ggplot2::geom_col(
    width = 0.82,
    color = "#111827",
    linewidth = 0.25
  ) +
  ggplot2::geom_hline(
    yintercept = figure3_control_mean_percentage,
    color = "#B7791F",
    linewidth = 0.9,
    linetype = "dashed"
  ) +
  ggplot2::geom_vline(
    xintercept = diet_gaps,
    color = "black",
    linewidth = 0.25
  ) +
  ggplot2::geom_vline(
    xintercept = treatment_gaps,
    color = "black",
    linewidth = 0.7
  ) +
  ggplot2::geom_text(
    data = figure3_fungal_percentage |>
      dplyr::filter(pct_unique_fungus > figure3_percentage_limit),
    ggplot2::aes(
      y = figure3_percentage_limit * 0.94,
      label = paste0(
        scales::number(pct_unique_fungus, accuracy = 0.001),
        "%"
      )
    ),
    color = "white",
    angle = 90,
    hjust = 1,
    vjust = 0.5,
    size = 6.8,
    fontface = "bold",
    show.legend = FALSE
  ) +
  ggplot2::scale_fill_manual(
    values = c(Control = "#FAEBD7", Infected = "#8B0000"),
    guide = "none"
  ) +
  ggplot2::scale_x_continuous(
    limits = sample_x_limits,
    breaks = sample_metadata$sample_index,
    labels = sample_metadata$label,
    expand = c(0, 0)
  ) +
  ggplot2::scale_y_continuous(
    limits = c(0, figure3_percentage_limit),
    breaks = seq(0, figure3_percentage_limit, by = 0.025),
    labels = scales::label_number(suffix = "%", accuracy = 0.025),
    expand = c(0, 0)
  ) +
  ggplot2::labs(
    title = "A. Fungal alignment and main-chromosome DEG expression",
    subtitle = paste0(
      "Fungal alignment zoom: 0-0.1%; higher values are capped and labeled; ",
      "dashed line = control mean (",
      scales::number(
        figure3_control_mean_percentage,
        accuracy = 0.00001
      ),
      "%)"
    ),
    x = NULL,
    y = "Unique fungal\nalignments (%)"
  ) +
  ggplot2::theme_classic(base_size = 16) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank(),
    axis.text.y = ggplot2::element_text(size = 14),
    axis.title.y = ggplot2::element_text(size = 17),
    plot.title = ggplot2::element_text(face = "bold", size = 21),
    plot.subtitle = ggplot2::element_text(size = 14),
    plot.margin = ggplot2::margin(5, 4, 0, 4)
  )

figure3_panel_a <- ggplot2::ggplot(
  figure3_heatmap_df,
  ggplot2::aes(x = sample_index, y = within_cluster_order, fill = z)
) +
  ggplot2::geom_segment(
    data = figure3_context_strip,
    ggplot2::aes(
      x = -1.1,
      xend = 0.35,
      y = within_cluster_order,
      yend = within_cluster_order,
      color = DEG_context_plot
    ),
    inherit.aes = FALSE,
    linewidth = 1.15,
    lineend = "butt"
  ) +
  ggplot2::geom_raster() +
  ggplot2::geom_vline(xintercept = 0.5, color = "#9CA3AF",
                      linewidth = 0.4) +
  ggplot2::geom_vline(xintercept = diet_gaps, color = "black",
                      linewidth = 0.25) +
  ggplot2::geom_vline(xintercept = treatment_gaps, color = "black",
                      linewidth = 0.7) +
  ggplot2::facet_grid(expression_cluster ~ ., scales = "free_y",
                      space = "free_y", switch = "y") +
  ggplot2::scale_fill_gradientn(
    colors = heatmap_colors,
    limits = c(-3, 3),
    oob = scales::squish,
    name = "Row z-score"
  ) +
  ggplot2::scale_color_manual(
    values = deg_context_plot_colors,
    breaks = figure3_present_contexts,
    labels = stringr::str_wrap(figure3_present_contexts, 24),
    drop = TRUE,
    name = "DEG context"
  ) +
  ggplot2::scale_x_continuous(
    limits = sample_x_limits,
    breaks = sample_metadata$sample_index,
    labels = sample_metadata$label,
    expand = c(0, 0)
  ) +
  ggplot2::scale_y_reverse(expand = c(0, 0)) +
  ggplot2::labs(x = NULL, y = NULL) +
  ggplot2::theme_classic(base_size = 16) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, hjust = 1,
                                        vjust = 0.5, size = 17),
    axis.text.y = ggplot2::element_blank(),
    axis.ticks.y = ggplot2::element_blank(),
    strip.placement = "outside",
    strip.background.y = ggplot2::element_rect(fill = "#F3F4F6",
                                               color = "#111827"),
    strip.text.y.left = ggplot2::element_text(angle = 90, face = "bold",
                                              size = 18),
    panel.spacing.y = grid::unit(0.25, "lines"),
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.direction = "horizontal",
    legend.box.just = "center",
    legend.title = ggplot2::element_text(face = "bold", size = 17),
    legend.text = ggplot2::element_text(size = 15),
    legend.spacing.x = grid::unit(0.55, "cm"),
    legend.margin = ggplot2::margin(4, 4, 2, 4),
    plot.margin = ggplot2::margin(0, 4, 1, 4)
  ) +
  ggplot2::guides(
    fill = ggplot2::guide_colorbar(
      order = 1,
      title.position = "left",
      title.hjust = 0.5,
      direction = "horizontal",
      barheight = grid::unit(0.38, "cm"),
      barwidth = grid::unit(4.2, "cm")
    ),
    color = ggplot2::guide_legend(
      order = 2,
      title.position = "left",
      nrow = 1,
      byrow = TRUE,
      override.aes = list(linewidth = 4.2)
    )
  )

figure3_within_go_enrichment <- dplyr::bind_rows(lapply(
  names(figure3_within_diet_sets),
  function(set_name) {
    run_enrichment(
      genes = figure3_within_diet_sets[[set_name]],
      universe = expressed_universe,
      term2gene = maps$term2gene_go,
      term2name = maps$term2name_go,
      database = "GO",
      background_label = "Expressed rRNA-filtered main-chromosome genes"
    ) |>
      dplyr::mutate(foreground_set = set_name)
  }
))

figure3_within_go_enrichment <- figure3_within_go_enrichment |>
  dplyr::left_join(
    maps$term2name_go |>
      dplyr::distinct(term, .keep_all = TRUE) |>
      dplyr::transmute(
        ID = term,
        ontology = dplyr::recode(
          ontology,
          BP = "Biological process",
          CC = "Cellular component",
          MF = "Molecular function",
          .default = NA_character_
        )
      ),
    by = "ID"
  ) |>
  dplyr::filter(!is.na(ontology)) |>
  dplyr::mutate(
    ontology = factor(ontology, levels = names(go_ontology_colors))
  )

figure3_within_go_significant <- figure3_within_go_enrichment |>
  dplyr::filter(p.adjust <= params$alpha) |>
  dplyr::arrange(foreground_set, ontology, p.adjust)

figure3_within_go_term_gene_index <- figure3_within_go_significant |>
  dplyr::select(foreground_set, ontology, ID, Description, GeneRatio, BgRatio,
                pvalue, p.adjust, qvalue, Count, geneID) |>
  tidyr::separate_rows(geneID, sep = "/") |>
  dplyr::rename(term_id = ID, term_name = Description, gene_id = geneID) |>
  dplyr::left_join(
    canonical_gene_table |>
      dplyr::select(gene_id, Description, gene_biotype, DEG_context,
                    comparisons, directions, max_abs_log2FC, min_padj),
    by = "gene_id"
  ) |>
  dplyr::arrange(foreground_set, ontology, p.adjust, term_id, gene_id)

write.csv(
  figure3_within_go_enrichment,
  file.path(out_dir,
            "figure3_infection_within_diet_GO_enrichment_all_terms.csv"),
  row.names = FALSE
)
write.csv(
  figure3_within_go_significant,
  file.path(out_dir,
            "figure3_infection_within_diet_GO_enrichment_significant_terms.csv"),
  row.names = FALSE
)
write.csv(
  figure3_within_go_term_gene_index,
  file.path(out_dir,
            "figure3_infection_within_diet_GO_term_gene_index_long.csv"),
  row.names = FALSE
)

figure3_go_df <- figure3_within_go_significant |>
  dplyr::mutate(
    expression_cluster = dplyr::recode(
      as.character(foreground_set),
      "Expression cluster 1" = "DEG cluster 1",
      "Expression cluster 2" = "DEG cluster 2"
    ),
    expression_cluster = factor(expression_cluster,
                                levels = figure3_cluster_levels)
  ) |>
  dplyr::group_by(expression_cluster, ontology) |>
  dplyr::slice_min(p.adjust, n = 4, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::arrange(expression_cluster, ontology, p.adjust) |>
  dplyr::group_by(expression_cluster) |>
  dplyr::mutate(
    neg_log10_padj = -log10(pmax(p.adjust, .Machine$double.xmin)),
    term_label = stringr::str_wrap(Description, 52),
    term_plot_id = paste(term_label, dplyr::row_number(),
                         expression_cluster, sep = "___")
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    term_plot_id = factor(term_plot_id, levels = rev(unique(term_plot_id)))
  )

figure3_go_xmax <- max(
  c(figure3_go_df$neg_log10_padj, -log10(params$alpha)),
  na.rm = TRUE
)
figure3_go_label_space <- max(10, figure3_go_xmax * 3.3)
figure3_panel_b <- ggplot2::ggplot(
  figure3_go_df,
  ggplot2::aes(x = neg_log10_padj, y = term_plot_id)
) +
  ggplot2::geom_vline(
    xintercept = -log10(params$alpha),
    linetype = "dashed",
    color = "#6B7280",
    linewidth = 0.35
  ) +
  ggplot2::geom_text(
    ggplot2::aes(
      x = -figure3_go_label_space * 0.98,
      label = sub("___.*$", "", term_plot_id)
    ),
    hjust = 0,
    size = 8.2,
    lineheight = 0.95,
    color = "#111827"
  ) +
  ggplot2::geom_segment(
    ggplot2::aes(x = 0, xend = neg_log10_padj,
                 yend = term_plot_id, color = ontology),
    linewidth = 0.75,
    alpha = 0.8
  ) +
  ggplot2::geom_point(
    ggplot2::aes(size = Count, color = ontology),
    alpha = 0.95
  ) +
  ggplot2::facet_wrap(~ expression_cluster, nrow = 1,
                      scales = "free_y", drop = FALSE) +
  ggplot2::scale_color_manual(values = go_ontology_colors, drop = FALSE) +
  ggplot2::scale_size_continuous(range = c(5.5, 12)) +
  ggplot2::scale_y_discrete(labels = function(x) rep("", length(x))) +
  ggplot2::scale_x_continuous(
    limits = c(-figure3_go_label_space,
               figure3_go_xmax + max(0.8, figure3_go_xmax * 0.15)),
    expand = c(0, 0)
  ) +
  ggplot2::labs(
    title = "B. GO terms: infection response within diet",
    x = "-log10 adjusted p-value",
    y = NULL,
    color = "GO ontology",
    size = "Cluster genes in term"
  ) +
  ggplot2::theme_classic(base_size = 16) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(size = 15),
    axis.title.x = ggplot2::element_text(size = 17),
    axis.text.y = ggplot2::element_blank(),
    axis.ticks.y = ggplot2::element_blank(),
    axis.line = ggplot2::element_blank(),
    axis.line.x.bottom = ggplot2::element_line(color = "black",
                                               linewidth = 0.45),
    strip.background = ggplot2::element_rect(fill = "#F3F4F6",
                                             color = "#111827"),
    strip.text = ggplot2::element_text(face = "bold", size = 18),
    legend.position = "bottom",
    legend.title = ggplot2::element_text(face = "bold", size = 17),
    legend.text = ggplot2::element_text(size = 15),
    legend.key.height = grid::unit(0.6, "cm"),
    legend.key.width = grid::unit(0.9, "cm"),
    plot.title = ggplot2::element_text(face = "bold", size = 19)
  )

figure3_within_kegg_enrichment <- dplyr::bind_rows(lapply(
  names(figure3_within_diet_sets),
  function(set_name) {
    run_enrichment(
      genes = figure3_within_diet_sets[[set_name]],
      universe = expressed_universe,
      term2gene = maps$term2gene_kegg,
      term2name = maps$term2name_kegg,
      database = "KEGG_pathway",
      background_label = "Expressed rRNA-filtered main-chromosome genes"
    ) |>
      dplyr::mutate(foreground_set = set_name)
  }
))

if (nrow(figure3_within_kegg_enrichment) > 0) {
  figure3_within_kegg_enrichment <- figure3_within_kegg_enrichment |>
    dplyr::left_join(
      maps$term2name_kegg |>
        dplyr::distinct(term, .keep_all = TRUE) |>
        dplyr::transmute(ID = term, pathway_theme, pathway_scope),
      by = "ID"
    ) |>
    dplyr::mutate(
      pathway_theme = dplyr::coalesce(pathway_theme,
                                      "Other KEGG pathway"),
      pathway_scope = dplyr::coalesce(
        pathway_scope,
        "Unclassified KEGG reference label"
      )
    )
}

figure3_within_kegg_significant <- figure3_within_kegg_enrichment |>
  dplyr::filter(p.adjust <= params$alpha) |>
  dplyr::arrange(foreground_set, p.adjust)

figure3_within_kegg_term_gene_index <- figure3_within_kegg_significant |>
  dplyr::select(foreground_set, pathway_theme, pathway_scope, ID, Description,
                GeneRatio, BgRatio, pvalue, p.adjust, qvalue, Count, geneID) |>
  tidyr::separate_rows(geneID, sep = "/") |>
  dplyr::rename(term_id = ID, term_name = Description, gene_id = geneID) |>
  dplyr::left_join(
    canonical_gene_table |>
      dplyr::select(gene_id, Description, gene_biotype, DEG_context,
                    comparisons, directions, max_abs_log2FC, min_padj),
    by = "gene_id"
  ) |>
  dplyr::arrange(foreground_set, p.adjust, term_id, gene_id)

write.csv(
  figure3_within_kegg_enrichment,
  file.path(out_dir,
            "figure3_infection_within_diet_KEGG_enrichment_all_terms.csv"),
  row.names = FALSE
)
write.csv(
  figure3_within_kegg_significant,
  file.path(out_dir,
            "figure3_infection_within_diet_KEGG_enrichment_significant_terms.csv"),
  row.names = FALSE
)
write.csv(
  figure3_within_kegg_term_gene_index,
  file.path(out_dir,
            "figure3_infection_within_diet_KEGG_term_gene_index_long.csv"),
  row.names = FALSE
)

figure3_kegg_insect_ids <- c(
  "ko00981", # Insect hormone biosynthesis
  "ko04341", # Hedgehog signaling pathway - fly
  "ko04624"  # Toll and Imd signaling pathway
)

figure3_kegg_display_scopes_eukaryote <- c(
  "Eukaryote/metazoan-compatible",
  "Core metabolism or cross-domain",
  "Broad KEGG reference map"
)

figure3_kegg_insect_audit <- figure3_within_kegg_significant |>
  dplyr::mutate(
    insect_focused_pathway = ID %in% figure3_kegg_insect_ids
  ) |>
  dplyr::arrange(foreground_set, dplyr::desc(insect_focused_pathway),
                 p.adjust)
write.csv(
  figure3_kegg_insect_audit,
  file.path(out_dir, "figure3_KEGG_insect_pathway_eligibility.csv"),
  row.names = FALSE
)

figure3_kegg_scope_summary <- figure3_within_kegg_significant |>
  dplyr::mutate(
    explicit_non_eukaryote_label = stringr::str_detect(
      stringr::str_to_lower(Description),
      paste0(
        "microbial|antibiotic|two-component|bacterial|quorum|",
        "peptidoglycan|lipopolysaccharide"
      )
    ),
    displayed_in_insect_focused_figure = ID %in% figure3_kegg_insect_ids,
    displayed_in_eukaryote_inclusive_figure = pathway_scope %in%
      figure3_kegg_display_scopes_eukaryote &
      !explicit_non_eukaryote_label
  ) |>
  dplyr::group_by(foreground_set, pathway_scope) |>
  dplyr::summarise(
    n_significant_pathways = dplyr::n(),
    n_displayed_insect_focused = sum(displayed_in_insect_focused_figure),
    n_displayed_eukaryote_inclusive =
      sum(displayed_in_eukaryote_inclusive_figure),
    .groups = "drop"
  ) |>
  dplyr::arrange(foreground_set,
                 dplyr::desc(n_displayed_insect_focused),
                 pathway_scope)
write.csv(
  figure3_kegg_scope_summary,
  file.path(out_dir, "figure3_KEGG_pathway_scope_summary.csv"),
  row.names = FALSE
)

prepare_figure3_kegg_df <- function(display_scopes = NULL,
                                    include_ids = NULL,
                                    exclude_explicit_non_eukaryote = FALSE) {
  selected <- figure3_within_kegg_significant

  if (!is.null(display_scopes)) {
    selected <- selected |>
      dplyr::filter(pathway_scope %in% display_scopes)
  }

  if (!is.null(include_ids)) {
    selected <- selected |>
      dplyr::filter(ID %in% include_ids)
  }

  if (exclude_explicit_non_eukaryote) {
    selected <- selected |>
      dplyr::filter(
        !stringr::str_detect(
          stringr::str_to_lower(Description),
          paste0(
            "microbial|antibiotic|two-component|bacterial|quorum|",
            "peptidoglycan|lipopolysaccharide"
          )
        )
      )
  }

  selected |>
    dplyr::mutate(
      expression_cluster = dplyr::recode(
        foreground_set,
        "Expression cluster 1" = "DEG cluster 1",
        "Expression cluster 2" = "DEG cluster 2"
      ),
      expression_cluster = factor(expression_cluster,
                                  levels = figure3_cluster_levels)
    ) |>
    dplyr::group_by(expression_cluster) |>
    dplyr::slice_min(p.adjust, n = 8, with_ties = FALSE) |>
    dplyr::ungroup() |>
    dplyr::arrange(expression_cluster, pathway_theme, p.adjust) |>
    dplyr::group_by(expression_cluster) |>
    dplyr::mutate(
      neg_log10_padj = -log10(pmax(p.adjust, .Machine$double.xmin)),
      pathway_label = stringr::str_wrap(Description, 52),
      pathway_plot_id = paste(pathway_label, dplyr::row_number(),
                              expression_cluster, sep = "___")
    ) |>
    dplyr::ungroup() |>
    dplyr::mutate(
      pathway_plot_id = factor(pathway_plot_id,
                               levels = rev(unique(pathway_plot_id)))
    )
}

build_figure3_kegg_panel <- function(kegg_df, panel_title) {
  if (nrow(kegg_df) == 0) {
    return(
      ggplot2::ggplot() +
        ggplot2::annotate(
          "text", x = 0, y = 0,
          label = "No significant KEGG pathways in either primary cluster",
          size = 6.5,
          fontface = "bold"
        ) +
        ggplot2::theme_void() +
        ggplot2::labs(title = panel_title)
    )
  }

  kegg_xmax <- max(
    c(kegg_df$neg_log10_padj, -log10(params$alpha)),
    na.rm = TRUE
  )
  kegg_label_space <- max(9, kegg_xmax * 3.3)

  ggplot2::ggplot(
    kegg_df,
    ggplot2::aes(x = neg_log10_padj, y = pathway_plot_id)
  ) +
    ggplot2::geom_vline(
      xintercept = -log10(params$alpha),
      linetype = "dashed",
      color = "#6B7280",
      linewidth = 0.35
    ) +
    ggplot2::geom_text(
      ggplot2::aes(
        x = -kegg_label_space * 0.98,
        label = sub("___.*$", "", pathway_plot_id)
      ),
      hjust = 0,
      size = 8.2,
      lineheight = 0.95,
      color = "#111827"
    ) +
    ggplot2::geom_segment(
      ggplot2::aes(x = 0, xend = neg_log10_padj,
                   yend = pathway_plot_id),
      color = kegg_lollipop_color,
      linewidth = 0.8,
      alpha = 0.8
    ) +
    ggplot2::geom_point(
      ggplot2::aes(size = Count),
      color = kegg_lollipop_color,
      alpha = 0.95
    ) +
    ggplot2::facet_wrap(~ expression_cluster, nrow = 1,
                        scales = "free_y", drop = FALSE) +
    ggplot2::scale_size_continuous(range = c(5.5, 12)) +
    ggplot2::scale_y_discrete(labels = function(x) rep("", length(x))) +
    ggplot2::scale_x_continuous(
      limits = c(-kegg_label_space,
                 kegg_xmax + max(0.8, kegg_xmax * 0.15)),
      expand = c(0, 0)
    ) +
    ggplot2::labs(
      title = panel_title,
      x = "-log10 adjusted p-value",
      y = NULL,
      size = "Cluster genes in pathway"
    ) +
    ggplot2::theme_classic(base_size = 16) +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(size = 15),
      axis.title.x = ggplot2::element_text(size = 17),
      axis.text.y = ggplot2::element_blank(),
      axis.ticks.y = ggplot2::element_blank(),
      axis.line = ggplot2::element_blank(),
      axis.line.x.bottom = ggplot2::element_line(color = "black",
                                                 linewidth = 0.45),
      strip.background = ggplot2::element_rect(fill = "#F3F4F6",
                                               color = "#111827"),
      strip.text = ggplot2::element_text(face = "bold", size = 18),
      legend.position = "bottom",
      legend.title = ggplot2::element_text(face = "bold", size = 17),
      legend.text = ggplot2::element_text(size = 15),
      legend.key.height = grid::unit(0.6, "cm"),
      legend.key.width = grid::unit(0.9, "cm"),
      plot.title = ggplot2::element_text(face = "bold", size = 19)
    )
}

figure3_kegg_df <- prepare_figure3_kegg_df(
  include_ids = figure3_kegg_insect_ids
)
figure3_kegg_eukaryote_df <- prepare_figure3_kegg_df(
  figure3_kegg_display_scopes_eukaryote,
  exclude_explicit_non_eukaryote = TRUE
)

figure3_panel_c <- build_figure3_kegg_panel(
  figure3_kegg_df,
  "C. Insect KEGG pathways: infection response within diet"
)
figure3_panel_c_eukaryote <- build_figure3_kegg_panel(
  figure3_kegg_eukaryote_df,
  "C. KEGG pathways: eukaryote-inclusive view"
)

# Map pathway summaries around a locust. Connector positions are conceptual
# physiological anchors and do not imply measured sub-tissue localization.
figure3_pathway_anchor <- function(description) {
  tibble::tibble(
    anchor_x = dplyr::case_when(
      stringr::str_detect(description, "Hedgehog") ~ 0.34,
      stringr::str_detect(description, "Protein digestion") ~ 0.53,
      stringr::str_detect(description, "Toll and Imd") ~ 0.55,
      stringr::str_detect(description, "ABC transporters") ~ 0.68,
      stringr::str_detect(description, "cytochrome P450") ~ 0.59,
      stringr::str_detect(description, "Pentose") ~ 0.49,
      stringr::str_detect(description, "Ascorbate") ~ 0.53,
      stringr::str_detect(description, "Retinol") ~ 0.46,
      stringr::str_detect(description, "secondary metabolites") ~ 0.64,
      TRUE ~ 0.60
    ),
    anchor_y = dplyr::case_when(
      stringr::str_detect(description, "Hedgehog") ~ 0.58,
      stringr::str_detect(description, "Protein digestion") ~ 0.43,
      stringr::str_detect(description, "Toll and Imd") ~ 0.50,
      stringr::str_detect(description, "ABC transporters") ~ 0.47,
      stringr::str_detect(description, "cytochrome P450") ~ 0.43,
      stringr::str_detect(description, "Pentose") ~ 0.48,
      stringr::str_detect(description, "Ascorbate") ~ 0.45,
      stringr::str_detect(description, "Retinol") ~ 0.42,
      stringr::str_detect(description, "secondary metabolites") ~ 0.44,
      TRUE ~ 0.50
    )
  )
}

build_figure3_pathway_illustration <- function(kegg_df, cluster_level,
                                               connector_color) {
  selected <- kegg_df |>
    dplyr::filter(as.character(expression_cluster) == cluster_level) |>
    dplyr::arrange(pathway_theme, p.adjust) |>
    dplyr::mutate(pathway_order = dplyr::row_number())

  if (nrow(selected) == 0) {
    return(
      ggplot2::ggplot() +
        ggplot2::annotate(
          "text", x = 0.5, y = 0.5,
          label = "No significant KEGG pathways",
          size = 5.5,
          fontface = "bold"
        ) +
        ggplot2::theme_void() +
        ggplot2::labs(title = cluster_level)
    )
  }

  n_left <- ceiling(nrow(selected) / 2)
  selected <- selected |>
    dplyr::mutate(
      side = dplyr::if_else(pathway_order <= n_left, "left", "right")
    ) |>
    dplyr::group_by(side) |>
    dplyr::mutate(
      callout_y = if (dplyr::n() == 1) {
        0.50
      } else {
        seq(0.80, 0.20, length.out = dplyr::n())
      }
    ) |>
    dplyr::ungroup() |>
    dplyr::bind_cols(figure3_pathway_anchor(selected$Description)) |>
    dplyr::mutate(
      label_x = dplyr::if_else(side == "left", 0.015, 0.985),
      connector_x = dplyr::if_else(side == "left", 0.225, 0.775),
      text_hjust = dplyr::if_else(side == "left", 0, 1),
      callout_label = paste0(
        stringr::str_wrap(Description, width = 26),
        "\n-log10 adjusted P = ",
        scales::number(neg_log10_padj, accuracy = 0.01),
        " | ", Count, " genes"
      )
    )

  locust_grob <- grid::rasterGrob(
    png::readPNG(figure3_locust_path),
    interpolate = TRUE
  )

  ggplot2::ggplot() +
    ggplot2::annotation_custom(
      locust_grob,
      xmin = 0.23, xmax = 0.77,
      ymin = 0.25, ymax = 0.78
    ) +
    ggplot2::geom_segment(
      data = selected,
      ggplot2::aes(
        x = anchor_x, y = anchor_y,
        xend = connector_x, yend = callout_y
      ),
      color = connector_color,
      linewidth = 0.55,
      alpha = 0.8
    ) +
    ggplot2::geom_point(
      data = selected,
      ggplot2::aes(x = anchor_x, y = anchor_y),
      shape = 21,
      size = 3.6,
      stroke = 0.55,
      fill = "white",
      color = connector_color
    ) +
    ggplot2::geom_text(
      data = selected,
      ggplot2::aes(
        x = label_x, y = callout_y,
        label = callout_label,
        hjust = text_hjust
      ),
      size = 5.4,
      lineheight = 0.92,
      color = "#111827"
    ) +
    ggplot2::annotate(
      "rect",
      xmin = 0, xmax = 1, ymin = 0.92, ymax = 1,
      fill = "#F3F4F6",
      color = "#111827",
      linewidth = 0.45
    ) +
    ggplot2::annotate(
      "text",
      x = 0.5, y = 0.96,
      label = cluster_level,
      fontface = "bold",
      size = 6.5
    ) +
    ggplot2::coord_cartesian(
      xlim = c(0, 1), ylim = c(0, 1),
      expand = FALSE,
      clip = "off"
    ) +
    ggplot2::theme_void() +
    ggplot2::theme(
      panel.border = ggplot2::element_rect(
        color = "#111827", fill = NA, linewidth = 0.45
      ),
      plot.margin = ggplot2::margin(4, 8, 4, 8)
    )
}

figure3_panel_c_illustration <-
  (build_figure3_pathway_illustration(
    figure3_kegg_eukaryote_df,
    "DEG cluster 1",
    "#9A5A14"
  ) |
  build_figure3_pathway_illustration(
    figure3_kegg_eukaryote_df,
    "DEG cluster 2",
    "#2F6F72"
  )) +
  patchwork::plot_annotation(
    title = "C. KEGG pathways across locust physiology",
    subtitle = paste(
      "Values below pathways show -log10 adjusted P and cluster gene count;",
      "connector locations are conceptual, not measured sub-tissue localization"
    ),
    theme = ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold", size = 19),
      plot.subtitle = ggplot2::element_text(size = 14, color = "#4B5563")
    )
  )

figure3_panel_a_widths <- c(0.80, 0.20)

figure3_panel_a_heatmap_only_content <-
  figure3_sample_annotation_plot /
  figure3_panel_a +
  patchwork::plot_layout(heights = c(0.95, 6.8))

figure3_panel_a_heatmap_only <-
  (figure3_panel_a_heatmap_only_content | patchwork::plot_spacer()) +
  patchwork::plot_layout(widths = figure3_panel_a_widths)

figure3_panel_a_content <-
  figure3_fungal_percentage_plot /
  figure3_sample_annotation_plot /
  figure3_panel_a +
  patchwork::plot_layout(heights = c(2.4, 1.1, 8.0))

figure3_panel_a_composite <-
  (figure3_panel_a_content | patchwork::plot_spacer()) +
  patchwork::plot_layout(widths = figure3_panel_a_widths)

figure3_k2_composite <-
  figure3_panel_a_composite /
  figure3_panel_b /
  figure3_panel_c +
  patchwork::plot_layout(heights = c(11.5, 6.5, 3.5)) +
  patchwork::plot_annotation(
    title = paste(
      "Main-chromosome DEG expression programs and functional enrichment",
      "(insect-focused KEGG view)"
    ),
    theme = ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold", size = 20)
    )
  )

figure3_k2_composite_eukaryote_lollipop <-
  figure3_panel_a_composite /
  figure3_panel_b /
  figure3_panel_c_eukaryote +
  patchwork::plot_layout(heights = c(11.5, 6.5, 4.4)) +
  patchwork::plot_annotation(
    title = paste(
      "Main-chromosome DEG expression programs and functional enrichment",
      "(eukaryote-inclusive KEGG view)"
    ),
    theme = ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold", size = 20)
    )
  )

figure3_k2_composite_eukaryote <-
  figure3_panel_a_composite /
  figure3_panel_b /
  figure3_panel_c_illustration +
  patchwork::plot_layout(heights = c(11.5, 6.5, 7.2)) +
  patchwork::plot_annotation(
    title = paste(
      "Main-chromosome DEG expression programs and functional enrichment",
      "(eukaryote-inclusive KEGG pathway map)"
    ),
    theme = ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold", size = 20)
    )
  )

print(figure3_k2_composite_eukaryote)
print(figure3_k2_composite)
print(figure3_k2_composite_eukaryote_lollipop)

figure3_export_specs <- list(
  "figure3_fungal_percentage_track" = list(
    plot = figure3_fungal_percentage_plot, width = 24, height = 4
  ),
  "figure3_panel_A_two_cluster_heatmap" = list(
    plot = figure3_panel_a_composite, width = 24, height = 15.2
  ),
  "figure3_panel_A_two_cluster_heatmap_without_fungal_track" = list(
    plot = figure3_panel_a_heatmap_only, width = 24, height = 11
  ),
  "figure3_panel_B_two_cluster_GO" = list(
    plot = figure3_panel_b, width = 24, height = 9.5
  ),
  "figure3_panel_C_two_cluster_KEGG" = list(
    plot = figure3_panel_c_illustration, width = 24, height = 9.5
  ),
  "figure3_panel_C_two_cluster_KEGG_pathway_illustration" = list(
    plot = figure3_panel_c_illustration, width = 24, height = 9.5
  ),
  "figure3_panel_C_two_cluster_KEGG_insect_focused" = list(
    plot = figure3_panel_c, width = 24, height = 4
  ),
  "figure3_panel_C_two_cluster_KEGG_eukaryote_inclusive" = list(
    plot = figure3_panel_c_eukaryote, width = 24, height = 6.5
  ),
  "expression_cluster_gene_level_GO_KEGG_lollipop" = list(
    plot = figure3_k2_composite_eukaryote_lollipop, width = 24, height = 29.3
  ),
  "expression_cluster_gene_level_GO_KEGG_lollipop_insect_focused" = list(
    plot = figure3_k2_composite, width = 24, height = 27.3
  ),
  "expression_cluster_gene_level_GO_KEGG_lollipop_eukaryote_inclusive" = list(
    plot = figure3_k2_composite_eukaryote_lollipop, width = 24, height = 29.3
  ),
  "expression_cluster_gene_level_GO_KEGG_pathway_illustration" = list(
    plot = figure3_k2_composite_eukaryote, width = 24, height = 31.8
  ),
  "figure3_with_fungal_percentage_KEGG_lollipop" = list(
    plot = figure3_k2_composite_eukaryote_lollipop, width = 24, height = 29.3
  ),
  "figure3_with_fungal_percentage_KEGG_locust_map" = list(
    plot = figure3_k2_composite_eukaryote, width = 24, height = 31.8
  )
)

for (export_name in names(figure3_export_specs)) {
  spec <- figure3_export_specs[[export_name]]
  ggplot2::ggsave(
    file.path(out_dir, paste0(export_name, ".png")),
    spec$plot,
    width = spec$width,
    height = spec$height,
    dpi = 260,
    limitsize = FALSE
  )
  ggplot2::ggsave(
    file.path(out_dir, paste0(export_name, ".pdf")),
    spec$plot,
    width = spec$width,
    height = spec$height,
    limitsize = FALSE
  )
  if (requireNamespace("svglite", quietly = TRUE)) {
    ggplot2::ggsave(
      file.path(out_dir, paste0(export_name, ".svg")),
      spec$plot,
      width = spec$width,
      height = spec$height,
      limitsize = FALSE
    )
  }
}

knitr::kable(
  figure3_foreground_summary,
  caption = "Figure 3 infection-within-diet enrichment foregrounds"
)
knitr::kable(
  figure3_kegg_scope_summary,
  caption = paste(
    "KEGG pathway scopes retained in or excluded from the insect-focused",
    "and eukaryote-inclusive Figure 3 versions"
  )
)
```

```{r expression-cluster-go-table}
if (nrow(cluster_go_significant) > 0) {
  DT::datatable(
    cluster_go_significant |>
      dplyr::arrange(expression_cluster, ontology, p.adjust) |>
      dplyr::select(expression_cluster, ontology, ID, Description, GeneRatio,
                    BgRatio, pvalue, p.adjust, qvalue, Count, geneID),
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 20, scrollX = TRUE),
    caption = "Cluster-specific significant GO terms"
  )
}
```

```{r expression-cluster-go-reverse-index}
if (nrow(cluster_term_gene_compact) > 0) {
  DT::datatable(
    cluster_term_gene_compact,
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 20, scrollX = TRUE),
    caption = "Reverse index for expression-cluster GO terms"
  )
}
```

## GO enrichment

```{r go-enrichment-plot, fig.width=18, fig.height=14}
go_plot_df <- go_enrichment |>
  dplyr::filter(p.adjust <= params$alpha) |>
  dplyr::group_by(background, ontology) |>
  dplyr::slice_min(p.adjust, n = params$top_terms_per_panel,
                   with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    neg_log10_padj = -log10(pmax(p.adjust, .Machine$double.xmin)),
    term_label = stringr::str_wrap(Description, 58),
    term_label = factor(term_label, levels = rev(unique(term_label))),
    ontology = factor(ontology, levels = names(go_ontology_colors))
  )

if (nrow(go_plot_df) > 0) {
  go_plot <- ggplot2::ggplot(
    go_plot_df,
    ggplot2::aes(x = neg_log10_padj, y = term_label)
  ) +
    ggplot2::geom_vline(
      xintercept = -log10(params$alpha),
      linetype = "dashed",
      color = "#6B7280",
      linewidth = 0.35
    ) +
    ggplot2::geom_point(
      ggplot2::aes(size = Count, color = ontology),
      alpha = 0.9
    ) +
    ggplot2::facet_grid(ontology ~ background,
                        scales = "free_y", space = "free_y") +
    ggplot2::scale_color_manual(values = go_ontology_colors, drop = FALSE) +
    ggplot2::scale_size_continuous(range = c(3.5, 11)) +
    ggplot2::labs(
      title = "GO enrichment in the circled heatmap branch",
      subtitle = "Foreground is the upper branch from the all-samples full DEG heatmap",
      x = "-log10 adjusted p-value",
      y = NULL,
      color = "GO ontology",
      size = "Cluster genes in term"
    ) +
    ggplot2::theme_classic(base_size = 14) +
    ggplot2::theme(
      axis.text.y = ggplot2::element_text(size = 11),
      strip.text = ggplot2::element_text(face = "bold", size = 12),
      legend.position = "bottom",
      plot.title = ggplot2::element_text(face = "bold")
    )
  print(go_plot)
  ggplot2::ggsave(file.path(out_dir, "circled_cluster_GO_enrichment.png"),
                  go_plot, width = 18, height = 14, dpi = 300)
} else {
  cat("No GO terms passed the adjusted p-value cutoff for this cluster.")
}
```

```{r go-enrichment-table}
if (nrow(go_enrichment) > 0) {
  DT::datatable(
    go_enrichment |>
      dplyr::arrange(p.adjust) |>
      dplyr::select(background, ontology, ID, Description, GeneRatio, BgRatio,
                    pvalue, p.adjust, qvalue, Count, geneID),
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 20, scrollX = TRUE),
    caption = "GO enrichment for the circled heatmap branch"
  )
}
```

## KEGG pathway enrichment

```{r kegg-enrichment-plot, fig.width=18, fig.height=12}
kegg_plot_df <- kegg_enrichment |>
  dplyr::filter(p.adjust <= params$alpha) |>
  dplyr::group_by(background, pathway_theme) |>
  dplyr::slice_min(p.adjust, n = params$top_terms_per_panel,
                   with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    neg_log10_padj = -log10(pmax(p.adjust, .Machine$double.xmin)),
    term_label = stringr::str_wrap(Description, 55),
    term_label = factor(term_label, levels = rev(unique(term_label))),
    pathway_scope = factor(pathway_scope, levels = names(kegg_scope_colors))
  )

if (nrow(kegg_plot_df) > 0) {
  kegg_plot <- ggplot2::ggplot(
    kegg_plot_df,
    ggplot2::aes(x = neg_log10_padj, y = term_label)
  ) +
    ggplot2::geom_vline(
      xintercept = -log10(params$alpha),
      linetype = "dashed",
      color = "#6B7280",
      linewidth = 0.35
    ) +
    ggplot2::geom_point(
      ggplot2::aes(size = Count),
      color = kegg_lollipop_color,
      alpha = 0.9
    ) +
    ggplot2::facet_grid(pathway_theme ~ background,
                        scales = "free_y", space = "free_y") +
    ggplot2::scale_size_continuous(range = c(4, 12)) +
    ggplot2::labs(
      title = "KEGG pathway enrichment in the circled heatmap branch",
      subtitle = "All KEGG pathway points are shown in brown; pathway scope remains in the exported tables",
      x = "-log10 adjusted p-value",
      y = NULL,
      size = "Cluster genes in pathway"
    ) +
    ggplot2::theme_classic(base_size = 14) +
    ggplot2::theme(
      axis.text.y = ggplot2::element_text(size = 12),
      strip.text = ggplot2::element_text(face = "bold", size = 11),
      legend.position = "bottom",
      plot.title = ggplot2::element_text(face = "bold")
    )
  print(kegg_plot)
  ggplot2::ggsave(file.path(out_dir, "circled_cluster_KEGG_enrichment.png"),
                  kegg_plot, width = 18, height = 12, dpi = 300)
} else {
  cat("No KEGG pathways passed the adjusted p-value cutoff for this cluster.")
}
```

```{r kegg-enrichment-table}
if (nrow(kegg_enrichment) > 0) {
  DT::datatable(
    kegg_enrichment |>
      dplyr::arrange(p.adjust) |>
      dplyr::select(background, pathway_theme, pathway_scope, ID, Description,
                    GeneRatio, BgRatio, pvalue, p.adjust, qvalue, Count,
                    geneID),
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 20, scrollX = TRUE),
    caption = "KEGG pathway enrichment for the circled heatmap branch"
  )
}
```

## Reverse term-to-gene index

```{r reverse-term-gene-index}
gene_annotation <- circled_cluster |>
  dplyr::select(gene_id, DEG_context, phase_class, n_deg_contrasts,
                comparisons, directions, max_abs_log2FC, min_padj,
                max_abs_sample_z, sample_with_max_abs_z,
                sample_deviation_direction) |>
  dplyr::distinct(gene_id, .keep_all = TRUE) |>
  dplyr::left_join(
    gff_annotation |> dplyr::select(gene_id, Description, gene_biotype),
    by = "gene_id"
  ) |>
  dplyr::mutate(
    Description = dplyr::coalesce(Description, "no GFF description"),
    gene_biotype = dplyr::coalesce(gene_biotype, "unknown")
  )

term_gene_long <- dplyr::bind_rows(
  expand_term_genes(go_enrichment, gene_annotation) |>
    dplyr::mutate(database = "GO"),
  expand_term_genes(kegg_enrichment, gene_annotation) |>
    dplyr::mutate(database = "KEGG_pathway")
) |>
  dplyr::arrange(database, background, p.adjust, term_id, gene_id)

term_gene_compact <- term_gene_long |>
  dplyr::group_by(database, background, term_id, term_name, GeneRatio, BgRatio,
                  p.adjust, Count) |>
  dplyr::summarise(
    DEG_gene_IDs = paste(sort(unique(gene_id)), collapse = "; "),
    DEG_gene_descriptions = paste(sort(unique(Description)), collapse = "; "),
    DEG_contexts = paste(sort(unique(DEG_context)), collapse = "; "),
    strongest_deviation_samples = paste(
      sort(unique(as.character(sample_with_max_abs_z))),
      collapse = "; "
    ),
    .groups = "drop"
  ) |>
  dplyr::arrange(database, background, p.adjust)

write.csv(term_gene_long,
          file.path(out_dir,
                    "circled_cluster_enriched_term_gene_index_long.csv"),
          row.names = FALSE)
write.csv(term_gene_compact,
          file.path(out_dir,
                    "circled_cluster_enriched_term_gene_index_compact.csv"),
          row.names = FALSE)

if (nrow(term_gene_compact) > 0) {
  DT::datatable(
    term_gene_compact,
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 20, scrollX = TRUE),
    caption = "Reverse index: enriched terms and the circled-cluster genes contributing to them"
  )
} else {
  cat("No significant enriched terms were available for the reverse gene index.")
}
```

## Cluster gene table

```{r cluster-gene-table}
cluster_gene_display <- circled_cluster |>
  dplyr::left_join(
    gff_annotation |> dplyr::select(gene_id, GFF_Description = Description,
                                    gene_biotype),
    by = "gene_id"
  ) |>
  dplyr::mutate(
    GFF_Description = dplyr::coalesce(GFF_Description, "no GFF description"),
    gene_biotype = dplyr::coalesce(gene_biotype, "unknown")
  ) |>
  dplyr::select(
    heatmap_row_order, gene_id, GFF_Description, gene_biotype, DEG_context,
    phase_class, n_deg_contrasts, comparisons, directions, max_abs_log2FC,
    min_padj, max_abs_sample_z, sample_with_max_abs_z,
    sample_deviation_direction
  )

write.csv(cluster_gene_display,
          file.path(out_dir, "circled_cluster_gene_table_with_gff_descriptions.csv"),
          row.names = FALSE)

DT::datatable(
  cluster_gene_display,
  rownames = FALSE,
  filter = "top",
  options = list(pageLength = 25, scrollX = TRUE),
  caption = "Genes in the circled heatmap branch"
)
```
