---
title: "Fat body WGCNA: diet and infection traits"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  metadata_file: "data/metadata/mehreen_metadata_fixed.txt"
  counts_file: "data/raw_read_counts/master_counts_no_rRNA.csv"
  gff_file: "data/reference/GCF_023897955.1_iqSchGreg1.2_genomic.gff"
  latest_results_root: "output/rmd_runs"
  min_total_count: 10
  top_n_genes: 5000
  min_module_size: 40
  merge_cut_height: 0.25
  alpha: 0.05
  lfc_threshold: 1
  n_threads: 2
  output_tag: "fatbody-wgcna-diet-infection"
---

## Purpose

This page asks whether fat body genes form coexpression modules associated with
the two experimental traits: diet and infection status. WGCNA is used as a
module-discovery and prioritization layer, not as a replacement for DESeq2.

The network is inferred from all retained fat body samples using the
rRNA-filtered VST expression matrix. Module eigengenes are then correlated with
sample-level traits:

- infection status: `Control = 0`, `Infected = 1`;
- diet as an ordered numeric trait: `33`, `50`, `83`;
- diet membership indicators: `Diet_33`, `Diet_50`, and `Diet_83`.

DEG context is added only after module construction, so differential-expression
results do not define the coexpression modules. This follows the same logic as
the AGY WGCNA page: build modules from expression first, then ask which modules
are associated with traits, enriched for DEGs, or contain high-membership hub
genes.

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE)

required <- c("DESeq2", "WGCNA", "dynamicTreeCut", "dplyr", "tidyr",
              "stringr", "ggplot2", "scales", "patchwork", "DT")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

metadata_file <- file.path(project_dir, params$metadata_file)
counts_file <- file.path(project_dir, params$counts_file)
gff_file <- file.path(project_dir, params$gff_file)
results_root <- file.path(project_dir, params$latest_results_root)

condition_colors <- c(
  "33 Control" = "#F7D6A6",
  "33 Infected" = "#C75C1B",
  "50 Control" = "#CDE8D5",
  "50 Infected" = "#1F6F43",
  "83 Control" = "#D8C7E8",
  "83 Infected" = "#5B2E83"
)
treatment_colors <- c("Control" = "#FAEBD7", "Infected" = "#8B0000")
diet_colors <- c("33" = "#E6862E", "50" = "#2F8F5B", "83" = "#7B4FA3")
deg_context_colors <- c(
  "Infection within diets" = "#C51B7D",
  "Diet DEGs in controls" = "#D9A21B",
  "Diet DEGs in infected" = "#1B9E77",
  "Infection vs control" = "#377EB8"
)
```

## Helper Functions

```{r helpers}
latest_dir <- function(pattern) {
  dirs <- list.dirs(results_root, recursive = FALSE, full.names = TRUE)
  dirs <- dirs[grepl(pattern, basename(dirs))]
  if (length(dirs) == 0) {
    stop("No output folder matched pattern: ", pattern)
  }
  dirs[which.max(file.info(dirs)$mtime)]
}

save_png_svg <- function(plot, file_base, width, height, dpi = 300) {
  png_path <- file.path(out_dir, paste0(file_base, ".png"))
  svg_path <- file.path(out_dir, paste0(file_base, ".svg"))
  ggplot2::ggsave(png_path, plot, width = width, height = height,
                  dpi = dpi, limitsize = FALSE)
  grDevices::svg(svg_path, width = width, height = height, onefile = FALSE)
  on.exit(grDevices::dev.off(), add = TRUE)
  print(plot)
  invisible(c(png = png_path, svg = svg_path))
}

theme_wgcna <- function(base_size = 12) {
  ggplot2::theme_classic(base_size = base_size) +
    ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold"),
      strip.background = ggplot2::element_rect(fill = "#F3F4F6",
                                               color = "#111827"),
      strip.text = ggplot2::element_text(face = "bold"),
      legend.position = "bottom",
      legend.title = ggplot2::element_text(face = "bold")
    )
}

parse_gff_gene_annotation <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")

  gff |>
    dplyr::filter(type == "gene") |>
    dplyr::transmute(
      gene_id = dplyr::coalesce(
        stringr::str_match(attributes, "gene=([^;]+)")[, 2],
        stringr::str_match(attributes, "Name=([^;]+)")[, 2]
      ),
      Description = stringr::str_match(attributes, "description=([^;]+)")[, 2],
      gene_biotype = stringr::str_match(attributes, "gene_biotype=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(gene_id)) |>
    dplyr::mutate(
      Description = dplyr::if_else(
        is.na(Description) | Description == "",
        "no GFF description",
        utils::URLdecode(Description)
      ),
      gene_biotype = dplyr::coalesce(gene_biotype, "unknown")
    ) |>
    dplyr::distinct(gene_id, .keep_all = TRUE)
}

format_trait <- function(x) {
  dplyr::case_when(
    x == "Infected" ~ "Infection\nInfected vs control",
    x == "Diet_numeric" ~ "Diet\nnumeric 33-50-83",
    x == "Diet_33" ~ "Diet\n33",
    x == "Diet_50" ~ "Diet\n50",
    x == "Diet_83" ~ "Diet\n83",
    TRUE ~ x
  )
}
```

## Input Preparation

```{r input-preparation}
metadata <- read.delim(metadata_file, check.names = FALSE)
metadata <- metadata[, !is.na(names(metadata)) & nzchar(names(metadata)),
                     drop = FALSE]
metadata <- metadata |>
  dplyr::mutate(
    label = as.character(label),
    sample_id = paste0("mehreen_", label, "_MERGE"),
    Diet = factor(as.character(Diet), levels = c("33", "50", "83")),
    Treatment = factor(Treatment, levels = c("Control", "Infected")),
    Condition = factor(paste(Diet, Treatment),
                       levels = names(condition_colors))
  ) |>
  dplyr::arrange(Treatment, Diet, label)

raw_counts <- read.csv(counts_file, check.names = FALSE)
stopifnot("GeneID" %in% names(raw_counts))
count_matrix <- as.matrix(raw_counts[, setdiff(names(raw_counts), "GeneID"),
                                     drop = FALSE])
rownames(count_matrix) <- raw_counts$GeneID
storage.mode(count_matrix) <- "numeric"

sample_keep <- intersect(metadata$sample_id, colnames(count_matrix))
metadata <- metadata[match(sample_keep, metadata$sample_id), , drop = FALSE]
count_matrix <- count_matrix[, metadata$sample_id, drop = FALSE]
count_matrix <- round(count_matrix)
storage.mode(count_matrix) <- "integer"

gene_keep <- rowSums(count_matrix) >= params$min_total_count
count_matrix_filtered <- count_matrix[gene_keep, , drop = FALSE]

sample_summary <- metadata |>
  dplyr::count(Diet, Treatment, name = "n_samples")
knitr::kable(sample_summary,
             caption = "Samples used for fat body WGCNA; all samples are retained.")

write.csv(sample_summary,
          file.path(out_dir, "fatbody_wgcna_sample_summary.csv"),
          row.names = FALSE)
```

## VST Matrix And Network Gene Selection

```{r vst-and-network-genes}
dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = count_matrix_filtered,
  colData = as.data.frame(metadata),
  design = ~ Diet * Treatment
)
dds <- DESeq2::estimateSizeFactors(dds)
vst <- DESeq2::vst(dds, blind = FALSE)
vst_matrix <- SummarizedExperiment::assay(vst)

gene_variance <- apply(vst_matrix, 1, var, na.rm = TRUE)
gene_variance <- gene_variance[is.finite(gene_variance) & gene_variance > 0]
network_genes <- names(sort(gene_variance, decreasing = TRUE))[
  seq_len(min(params$top_n_genes, length(gene_variance)))
]

dat_expr <- t(vst_matrix[network_genes, metadata$sample_id, drop = FALSE])
good <- WGCNA::goodSamplesGenes(dat_expr, verbose = 0)
if (!good$allOK) {
  dat_expr <- dat_expr[good$goodSamples, good$goodGenes, drop = FALSE]
  metadata <- metadata[match(rownames(dat_expr), metadata$sample_id), ,
                       drop = FALSE]
  network_genes <- colnames(dat_expr)
}

write.csv(data.frame(gene_id = network_genes),
          file.path(out_dir, "fatbody_wgcna_input_genes.csv"),
          row.names = FALSE)
```

The network uses `r ncol(dat_expr)` high-variance genes across
`r nrow(dat_expr)` samples.

## Trait Coding

```{r trait-coding}
trait_table <- metadata |>
  dplyr::transmute(
    sample_id,
    Diet = as.character(Diet),
    Treatment = as.character(Treatment),
    Condition = as.character(Condition),
    Infected = ifelse(Treatment == "Infected", 1, 0),
    Diet_numeric = as.numeric(as.character(Diet)),
    Diet_33 = ifelse(Diet == "33", 1, 0),
    Diet_50 = ifelse(Diet == "50", 1, 0),
    Diet_83 = ifelse(Diet == "83", 1, 0)
  )
rownames(trait_table) <- trait_table$sample_id

numeric_traits <- trait_table[rownames(dat_expr),
                              c("Infected", "Diet_numeric",
                                "Diet_33", "Diet_50", "Diet_83"),
                              drop = FALSE]

write.csv(trait_table,
          file.path(out_dir, "fatbody_wgcna_trait_table.csv"),
          row.names = FALSE)

knitr::kable(trait_table,
             caption = "Sample-level traits used for module-trait correlations.")
```

## WGCNA Network Construction

```{r wgcna-network, results='hide'}
WGCNA::allowWGCNAThreads(nThreads = params$n_threads)
powers <- c(1:10, 12, 14, 16, 18, 20)

sft <- WGCNA::pickSoftThreshold(
  dat_expr,
  powerVector = powers,
  networkType = "signed",
  corFnc = "bicor",
  corOptions = "use = 'pairwise.complete.obs', maxPOutliers = 0.1",
  verbose = 0
)
fit_indices <- as.data.frame(sft$fitIndices)
power_hits <- fit_indices$Power[fit_indices$SFT.R.sq >= 0.80]
soft_power <- if (length(power_hits) > 0) {
  power_hits[1]
} else {
  fit_indices$Power[which.max(fit_indices$SFT.R.sq)]
}
if (!is.finite(soft_power) || length(soft_power) == 0) soft_power <- 6

net <- WGCNA::blockwiseModules(
  dat_expr,
  power = soft_power,
  networkType = "signed",
  TOMType = "signed",
  corType = "bicor",
  maxPOutliers = 0.1,
  minModuleSize = params$min_module_size,
  mergeCutHeight = params$merge_cut_height,
  numericLabels = FALSE,
  pamRespectsDendro = FALSE,
  saveTOMs = FALSE,
  maxBlockSize = ncol(dat_expr),
  verbose = 2
)

module_colors <- as.character(net$colors)
names(module_colors) <- colnames(dat_expr)
module_eigengenes <- WGCNA::moduleEigengenes(
  dat_expr,
  colors = module_colors
)$eigengenes
module_eigengenes <- WGCNA::orderMEs(module_eigengenes)

kme_matrix <- WGCNA::signedKME(
  dat_expr,
  module_eigengenes,
  outputColumnName = "kME",
  corFnc = "bicor",
  corOptions = "use = 'pairwise.complete.obs', maxPOutliers = 0.1"
)

connectivity <- tryCatch(
  WGCNA::intramodularConnectivity.fromExpr(
    dat_expr,
    colors = module_colors,
    power = soft_power,
    networkType = "signed",
    corFnc = "bicor",
    corOptions = "use = 'pairwise.complete.obs', maxPOutliers = 0.1"
  ),
  error = function(e) {
    data.frame(kTotal = NA_real_, kWithin = NA_real_, kOut = NA_real_,
               kDiff = NA_real_, row.names = names(module_colors))
  }
)

module_for_gene <- module_colors[colnames(dat_expr)]
kme_gene <- vapply(colnames(dat_expr), function(gene) {
  mod <- module_for_gene[[gene]]
  kme_col <- paste0("kME", mod)
  if (kme_col %in% colnames(kme_matrix)) {
    as.numeric(kme_matrix[gene, kme_col])
  } else {
    NA_real_
  }
}, numeric(1))

gene_annotation <- parse_gff_gene_annotation(gff_file)

module_assignments <- data.frame(
  gene_id = colnames(dat_expr),
  Module = module_for_gene,
  kME = kme_gene,
  kTotal = connectivity[colnames(dat_expr), "kTotal"],
  kWithin = connectivity[colnames(dat_expr), "kWithin"],
  stringsAsFactors = FALSE
) |>
  dplyr::group_by(Module) |>
  dplyr::mutate(
    Module_size = dplyr::n(),
    Hub_rank_within_module = rank(-abs(kME), ties.method = "first")
  ) |>
  dplyr::ungroup() |>
  dplyr::left_join(gene_annotation, by = "gene_id") |>
  dplyr::mutate(
    Description = dplyr::coalesce(Description, "no GFF description"),
    gene_biotype = dplyr::coalesce(gene_biotype, "unknown")
  )

module_sizes <- module_assignments |>
  dplyr::count(Module, name = "n_genes") |>
  dplyr::arrange(dplyr::desc(n_genes))

write.csv(fit_indices,
          file.path(out_dir, "fatbody_wgcna_soft_threshold_fit.csv"),
          row.names = FALSE)
write.csv(module_sizes,
          file.path(out_dir, "fatbody_wgcna_module_sizes.csv"),
          row.names = FALSE)
write.csv(module_assignments,
          file.path(out_dir, "fatbody_wgcna_module_assignments.csv"),
          row.names = FALSE)
saveRDS(net, file.path(out_dir, "fatbody_wgcna_blockwise_network.rds"))
```

Selected signed-network soft threshold power: **`r soft_power`**.

```{r soft-threshold-plot, fig.width=10, fig.height=4.5}
soft_threshold_plot <- ggplot2::ggplot(
  fit_indices,
  ggplot2::aes(x = Power, y = SFT.R.sq)
) +
  ggplot2::geom_hline(yintercept = 0.80, linetype = "dashed",
                      color = "#6B7280") +
  ggplot2::geom_line(color = "#374151") +
  ggplot2::geom_point(size = 2.4, color = "#111827") +
  ggplot2::geom_point(
    data = fit_indices[fit_indices$Power == soft_power, , drop = FALSE],
    color = "#C51B7D",
    size = 3.8
  ) +
  theme_wgcna(base_size = 12) +
  ggplot2::labs(
    title = "Soft-threshold selection",
    subtitle = "Dashed line marks scale-free topology R2 = 0.80",
    x = "Candidate power",
    y = "Scale-free topology fit"
  )
save_png_svg(soft_threshold_plot, "fatbody_wgcna_soft_threshold_fit",
             width = 10, height = 4.5)
soft_threshold_plot
```

## Module Sizes

```{r module-sizes, fig.width=9, fig.height=6}
module_size_plot <- module_sizes |>
  dplyr::mutate(Module = factor(Module, levels = rev(Module))) |>
  ggplot2::ggplot(ggplot2::aes(x = n_genes, y = Module, fill = Module)) +
  ggplot2::geom_col(width = 0.75, color = "white", linewidth = 0.2) +
  ggplot2::scale_fill_identity() +
  theme_wgcna(base_size = 12) +
  ggplot2::theme(legend.position = "none") +
  ggplot2::labs(
    title = "WGCNA module sizes",
    x = "Number of genes",
    y = "Module"
  )
save_png_svg(module_size_plot, "fatbody_wgcna_module_sizes",
             width = 9, height = 6)
module_size_plot
```

## Module-Trait Associations

```{r module-trait-correlations}
module_names <- sub("^ME", "", colnames(module_eigengenes))
module_trait <- dplyr::bind_rows(lapply(module_names, function(module) {
  me_col <- paste0("ME", module)
  dplyr::bind_rows(lapply(colnames(numeric_traits), function(trait) {
    x <- module_eigengenes[rownames(numeric_traits), me_col]
    y <- numeric_traits[[trait]]
    test <- suppressWarnings(stats::cor.test(x, y, method = "spearman",
                                             exact = FALSE))
    data.frame(
      Module = module,
      Trait = trait,
      Correlation = unname(test$estimate),
      P_value = test$p.value,
      stringsAsFactors = FALSE
    )
  }))
}))
module_trait$FDR <- p.adjust(module_trait$P_value, method = "BH")
module_trait <- module_trait |>
  dplyr::mutate(
    Trait_label = format_trait(Trait),
    Significance_label = dplyr::case_when(
      FDR < 0.001 ~ "***",
      FDR < 0.01 ~ "**",
      FDR < 0.05 ~ "*",
      TRUE ~ ""
    )
  ) |>
  dplyr::arrange(FDR, dplyr::desc(abs(Correlation)))

write.csv(module_trait,
          file.path(out_dir, "fatbody_wgcna_module_trait_correlations.csv"),
          row.names = FALSE)

module_trait_display <- module_trait |>
  dplyr::mutate(
    Module = factor(Module, levels = rev(unique(sub("^ME", "", colnames(module_eigengenes))))),
    Trait_label = factor(Trait_label, levels = format_trait(colnames(numeric_traits))),
    Tile_label = paste0(sprintf("%.2f", Correlation), "\n", Significance_label)
  )
```

```{r module-trait-heatmap, fig.width=10, fig.height=8}
module_trait_heatmap <- ggplot2::ggplot(
  module_trait_display,
  ggplot2::aes(x = Trait_label, y = Module, fill = Correlation)
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.35) +
  ggplot2::geom_text(ggplot2::aes(label = Tile_label), size = 3.0) +
  ggplot2::scale_fill_gradient2(
    low = "#1D4ED8",
    mid = "white",
    high = "#B91C1C",
    midpoint = 0,
    limits = c(-1, 1),
    name = "Spearman r"
  ) +
  theme_wgcna(base_size = 12) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 35, hjust = 1),
    panel.border = ggplot2::element_rect(color = "#111827",
                                         fill = NA,
                                         linewidth = 0.35)
  ) +
  ggplot2::labs(
    title = "Module eigengene associations with diet and infection traits",
    subtitle = "Asterisks mark BH-adjusted FDR thresholds: * < 0.05, ** < 0.01, *** < 0.001",
    x = "Sample trait",
    y = "WGCNA module"
  )
save_png_svg(module_trait_heatmap,
             "fatbody_wgcna_module_trait_heatmap",
             width = 10, height = 8)
module_trait_heatmap
```

## Eigengene Profiles For Top Trait-Associated Modules

```{r eigengene-profiles, fig.width=12, fig.height=7}
top_modules <- module_trait |>
  dplyr::filter(Module != "grey") |>
  dplyr::arrange(FDR, dplyr::desc(abs(Correlation))) |>
  dplyr::distinct(Module, .keep_all = TRUE) |>
  dplyr::slice_head(n = 6)
if (nrow(top_modules) == 0) {
  top_modules <- module_trait |>
    dplyr::filter(Module != "grey") |>
    dplyr::arrange(dplyr::desc(abs(Correlation))) |>
    dplyr::distinct(Module, .keep_all = TRUE) |>
    dplyr::slice_head(n = 6)
}

eigengene_long <- module_eigengenes |>
  as.data.frame(check.names = FALSE) |>
  tibble::rownames_to_column("sample_id") |>
  tidyr::pivot_longer(
    cols = tidyselect::starts_with("ME"),
    names_to = "ME",
    values_to = "Eigengene"
  ) |>
  dplyr::mutate(Module = sub("^ME", "", ME)) |>
  dplyr::filter(Module %in% top_modules$Module) |>
  dplyr::left_join(
    trait_table |>
      dplyr::select(sample_id, Diet, Treatment, Condition),
    by = "sample_id"
  ) |>
  dplyr::left_join(
    top_modules |>
      dplyr::select(Module, Trait, Correlation, FDR) |>
      dplyr::mutate(
        Module_label = paste0(
          "Module ", Module,
          "\nTop trait: ", format_trait(Trait),
          "\nr = ", sprintf("%.2f", Correlation),
          ", FDR = ", signif(FDR, 2)
        )
      ),
    by = "Module"
  ) |>
  dplyr::mutate(
    Diet = factor(Diet, levels = c("33", "50", "83")),
    Treatment = factor(Treatment, levels = c("Control", "Infected")),
    Condition = factor(Condition, levels = names(condition_colors)),
    Module_label = factor(Module_label, levels = unique(Module_label))
  )

eigengene_profile_plot <- ggplot2::ggplot(
  eigengene_long,
  ggplot2::aes(x = Diet, y = Eigengene, color = Treatment)
) +
  ggplot2::geom_boxplot(
    ggplot2::aes(group = interaction(Diet, Treatment)),
    width = 0.62,
    outlier.shape = NA,
    alpha = 0.25,
    linewidth = 0.35
  ) +
  ggplot2::geom_point(
    position = ggplot2::position_jitterdodge(
      jitter.width = 0.12,
      dodge.width = 0.68,
      seed = 17
    ),
    size = 2.2,
    alpha = 0.9
  ) +
  ggplot2::facet_wrap(~ Module_label, scales = "free_y", ncol = 3) +
  ggplot2::scale_color_manual(values = treatment_colors, drop = FALSE) +
  theme_wgcna(base_size = 12) +
  ggplot2::theme(
    strip.text = ggplot2::element_text(size = 9.5, face = "bold"),
    legend.position = "bottom"
  ) +
  ggplot2::labs(
    title = "Eigengene profiles for top diet/infection-associated modules",
    x = "Diet",
    y = "Module eigengene",
    color = "Treatment"
  )
save_png_svg(eigengene_profile_plot,
             "fatbody_wgcna_top_module_eigengene_profiles",
             width = 12, height = 7)
eigengene_profile_plot
```

## DEG Context Overlay After Module Construction

```{r deg-context-overlay}
scenario_dir <- latest_dir("^deg-sample-filter-scenarios_")
scenario_results_file <- file.path(scenario_dir, "all_deseq2_results.csv")
model_dir <- latest_dir("^all-samples-deseq2-interaction-model_")
model_coefficients_file <- file.path(model_dir, "all_model_coefficients.csv")

scenario_degs <- read.csv(scenario_results_file, check.names = FALSE) |>
  dplyr::filter(
    scenario == "All samples",
    significant,
    !is.na(gene_id),
    !is.na(padj),
    padj < params$alpha,
    abs(log2FoldChange) >= params$lfc_threshold
  )
model_degs <- read.csv(model_coefficients_file, check.names = FALSE) |>
  dplyr::filter(
    coefficient == "Treatment_Infected_vs_Control",
    significant,
    !is.na(gene_id),
    !is.na(padj),
    padj < params$alpha,
    abs(log2FoldChange) >= params$lfc_threshold
  )

deg_sets <- list(
  "Infection within diets" = scenario_degs |>
    dplyr::filter(comparison_type == "Infection within diet") |>
    dplyr::pull(gene_id) |>
    unique(),
  "Diet DEGs in controls" = scenario_degs |>
    dplyr::filter(comparison_type == "Diet contrasts among control") |>
    dplyr::pull(gene_id) |>
    unique(),
  "Diet DEGs in infected" = scenario_degs |>
    dplyr::filter(comparison_type == "Diet contrasts among infected") |>
    dplyr::pull(gene_id) |>
    unique(),
  "Infection vs control" = unique(model_degs$gene_id)
)
deg_sets <- deg_sets[lengths(deg_sets) > 0]

deg_context_long <- dplyr::bind_rows(lapply(names(deg_sets), function(context) {
  data.frame(DEG_context = context,
             gene_id = deg_sets[[context]],
             stringsAsFactors = FALSE)
})) |>
  dplyr::distinct(DEG_context, gene_id)

deg_context_summary <- deg_context_long |>
  dplyr::group_by(gene_id) |>
  dplyr::summarise(
    DEG_contexts = paste(sort(unique(DEG_context)), collapse = " | "),
    DEG_context_count = dplyr::n_distinct(DEG_context),
    .groups = "drop"
  )

module_assignments <- module_assignments |>
  dplyr::left_join(deg_context_summary, by = "gene_id") |>
  dplyr::mutate(
    DEG_contexts = dplyr::coalesce(DEG_contexts, "Not DEG"),
    DEG_context_count = dplyr::coalesce(DEG_context_count, 0L)
  )
write.csv(module_assignments,
          file.path(out_dir, "fatbody_wgcna_module_assignments_annotated.csv"),
          row.names = FALSE)

universe_genes <- unique(module_assignments$gene_id)
module_context_tests <- dplyr::bind_rows(lapply(unique(module_assignments$Module), function(mod) {
  module_genes <- module_assignments$gene_id[module_assignments$Module == mod]
  dplyr::bind_rows(lapply(names(deg_sets), function(context) {
    context_genes <- intersect(deg_sets[[context]], universe_genes)
    a <- length(intersect(module_genes, context_genes))
    b <- length(setdiff(module_genes, context_genes))
    c <- length(setdiff(context_genes, module_genes))
    d <- length(setdiff(universe_genes, union(module_genes, context_genes)))
    ft <- fisher.test(matrix(c(a, b, c, d), nrow = 2))
    data.frame(
      Module = mod,
      DEG_context = context,
      Module_genes = length(module_genes),
      Context_genes_in_universe = length(context_genes),
      Overlap_genes = a,
      Odds_ratio = unname(ft$estimate),
      P_value = ft$p.value,
      stringsAsFactors = FALSE
    )
  }))
}))
module_context_tests$FDR <- p.adjust(module_context_tests$P_value, method = "BH")
module_context_tests <- module_context_tests |>
  dplyr::arrange(FDR, dplyr::desc(Odds_ratio))
write.csv(module_context_tests,
          file.path(out_dir, "fatbody_wgcna_deg_context_enrichment.csv"),
          row.names = FALSE)

deg_enrichment_plot_data <- module_context_tests |>
  dplyr::filter(Module != "grey", Overlap_genes > 0) |>
  dplyr::group_by(DEG_context) |>
  dplyr::arrange(FDR, dplyr::desc(Odds_ratio), .by_group = TRUE) |>
  dplyr::slice_head(n = 8) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    DEG_context = factor(DEG_context, levels = names(deg_sets)),
    Module_label = paste0("Module ", Module),
    Module_label = factor(Module_label, levels = rev(unique(Module_label))),
    neg_log10_fdr = -log10(pmax(FDR, .Machine$double.xmin))
  )
```

```{r deg-context-enrichment-plot, fig.width=11, fig.height=8}
deg_context_enrichment_plot <- ggplot2::ggplot(
  deg_enrichment_plot_data,
  ggplot2::aes(x = neg_log10_fdr, y = Module_label,
               color = DEG_context, size = Overlap_genes)
) +
  ggplot2::geom_vline(xintercept = -log10(0.05), linetype = "dashed",
                      color = "#6B7280") +
  ggplot2::geom_point(alpha = 0.9) +
  ggplot2::facet_wrap(~ DEG_context, scales = "free_y", ncol = 2) +
  ggplot2::scale_color_manual(values = deg_context_colors[names(deg_sets)],
                              guide = "none") +
  ggplot2::scale_size_continuous(range = c(2.5, 8)) +
  theme_wgcna(base_size = 12) +
  ggplot2::theme(
    strip.text = ggplot2::element_text(size = 11, face = "bold"),
    axis.text.y = ggplot2::element_text(color = "#111827")
  ) +
  ggplot2::labs(
    title = "WGCNA module enrichment for DEG contexts",
    subtitle = "Fisher tests are performed after module construction; dashed line marks FDR = 0.05",
    x = "-log10 FDR",
    y = "WGCNA module",
    size = "DEGs in module"
  )
save_png_svg(deg_context_enrichment_plot,
             "fatbody_wgcna_deg_context_enrichment",
             width = 11, height = 8)
deg_context_enrichment_plot
```

## Hub Table

```{r hub-table, results='asis'}
hub_table <- module_assignments |>
  dplyr::filter(Module != "grey") |>
  dplyr::arrange(Module, Hub_rank_within_module) |>
  dplyr::mutate(
    abs_kME = abs(kME),
    Hub_rank_within_module = as.integer(Hub_rank_within_module)
  ) |>
  dplyr::select(
    Module,
    Module_size,
    Hub_rank_within_module,
    gene_id,
    Description,
    gene_biotype,
    kME,
    abs_kME,
    kWithin,
    kTotal,
    DEG_contexts,
    DEG_context_count
  )

write.csv(hub_table,
          file.path(out_dir, "fatbody_wgcna_hub_table.csv"),
          row.names = FALSE)

DT::datatable(
  hub_table,
  filter = "top",
  rownames = FALSE,
  extensions = "Buttons",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv")
  ),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left;",
    "Searchable WGCNA module-membership and hub table"
  )
)
```

## Output Manifest

```{r output-manifest}
run_manifest <- data.frame(
  item = c("metadata_file", "counts_file", "gff_file",
           "scenario_results_file", "model_coefficients_file",
           "samples_used", "genes_after_count_filter",
           "network_genes_used", "soft_threshold_power",
           "min_module_size", "merge_cut_height",
           "outliers_retained"),
  value = c(metadata_file, counts_file, gff_file,
            scenario_results_file, model_coefficients_file,
            nrow(metadata), nrow(count_matrix_filtered),
            ncol(dat_expr), soft_power,
            params$min_module_size, params$merge_cut_height,
            "yes; all samples in mehreen_metadata_fixed.txt are retained"),
  stringsAsFactors = FALSE
)
write.csv(run_manifest,
          file.path(out_dir, "fatbody_wgcna_run_manifest.csv"),
          row.names = FALSE)
knitr::kable(run_manifest, caption = "Fat body WGCNA run manifest")

exports <- list.files(out_dir, full.names = FALSE)
knitr::kable(tibble::tibble(file = exports),
             caption = paste("Files written to", out_dir))
```

## Session Information

```{r session-info}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
