---
title: "Origin audit of DEG-rich unplaced scaffolds"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  audit_review_dir: "output/runs/scaffold_origin_20260729_030308/local-analysis/scaffold-origin-review_20260730-234810"
  homology_review_dir: "output/runs/scaffold_homology_20260803_042409/local-analysis/homology-review_20260804-084526"
  taxonomy_lookup: "data/scaffold_origin/ncbi_taxonomy_lookup_20260804.tsv"
  output_tag: "scaffold-origin-audit"
---

## Purpose

::: {.walkthrough-grid}
::: {.walkthrough-step}
**Step 1 - Gather evidence**

Combine assembly placement, Kraken, BLASTn, DIAMOND, and cross-locust homology.
:::
::: {.walkthrough-step}
**Step 2 - Classify support**

Separate host-supported, ambiguous, and unresolved unplaced loci.
:::
::: {.walkthrough-step}
**Step 3 - Plan follow-up**

Preserve uncertain candidates for validation without mixing them into the primary host set.
:::
:::

This page asks whether the unplaced *Schistocerca gregaria* assembly scaffolds
carrying many differentially expressed genes (DEGs) have independent evidence
of locust origin. The audit itself does not relabel an unresolved scaffold as a
contaminant. Following this review, the primary analysis now uses
chromosome-assigned genes as its primary hypothesis family and preserves
unplaced genes as an explicit sensitivity analysis on the
[placed/unplaced analysis page](28-placed-unplaced-scaffold-analysis.html).

> **Evidence boundary:** this is a focused origin audit of the 948 genes on 59
> unresolved scaffolds that contribute to the infected diet 33 versus diet 50
> contrast. BLASTn and DIAMOND results are now included. Similarity to a
> non-metazoan sequence is treated as a review flag, not as proof that a gene or
> scaffold is microbial.

The evidence combines:

1. scaffold sequence and RefSeq gene-annotation metrics;
2. Kraken2 classification against PlusPF;
3. NCBI Foreign Contamination Screen (FCS) flags distributed with the assembly;
4. whole-scaffold alignments to three related *Schistocerca* assemblies;
5. BLASTn searches of gene regions against NCBI `core_nt`; and
6. DIAMOND searches of predicted proteins against NCBI `clustered_nr`.

Cross-species support uses only minimap2 alignments with mapping quality at
least 20 and divergence-derived nucleotide identity at least 70%. This avoids
counting numerous ambiguous repetitive alignments as independent evidence of
host conservation.

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c(
  "dplyr", "tidyr", "stringr", "readr", "ggplot2", "ggrepel",
  "scales", "DT", "knitr", "patchwork"
)
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

project_dir <- normalizePath(".", mustWork = TRUE)
review_dir <- file.path(project_dir, params$audit_review_dir)
homology_review_dir <- file.path(project_dir, params$homology_review_dir)
homology_extract_dir <- file.path(homology_review_dir, "extracted")
evidence_file <- file.path(
  review_dir, "corrected-evidence-v2", "scaffold_origin_evidence_mapq20.tsv"
)
homology_file <- file.path(
  review_dir, "corrected-evidence-v2",
  "cross_species_scaffold_homology_mapq20.tsv"
)
candidate_gene_file <- file.path(
  review_dir, "extracted", "00-input", "candidate_deg_genes.tsv"
)
validation_file <- file.path(review_dir, "transfer_validation.tsv")
homology_validation_file <- file.path(
  homology_review_dir, "transfer_validation.tsv"
)
homology_gene_file <- file.path(
  homology_extract_dir, "04-evidence", "gene_homology_summary.tsv"
)
homology_top_hit_file <- file.path(
  homology_extract_dir, "04-evidence", "top_tier_taxonomic_hits.tsv"
)
blastn_file <- file.path(
  homology_extract_dir, "02-blastn-core-nt", "gene_regions_vs_core_nt.tsv"
)
taxonomy_file <- file.path(project_dir, params$taxonomy_lookup)

run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(
  project_dir, "output", "rmd_runs",
  paste(params$output_tag, run_stamp, sep = "_")
)
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

status_levels <- c(
  "Strong cross-Schistocerca support for host origin",
  "Partial cross-Schistocerca support; likely host-associated",
  "Unresolved by Kraken/FCS"
)
status_labels <- c(
  "Strong host support",
  "Partial host support",
  "Unresolved"
)
status_colors <- c(
  "Strong host support" = "#2E7D32",
  "Partial host support" = "#D89000",
  "Unresolved" = "#9C3D3D"
)

homology_colors <- c(
  "External nucleotide support from Orthoptera/Arthropoda" = "#22543D",
  "External nucleotide support from other Metazoa" = "#5C8362",
  "Arthropod/orthopteran protein support only" = "#6F9C76",
  "Other-metazoan protein support only" = "#9DB89D",
  "Mixed external nucleotide evidence" = "#D7A144",
  "Mixed protein evidence" = "#E9C46A",
  "Conflicting nucleotide and protein evidence" = "#C98A40",
  "Non-metazoan nucleotide evidence only" = "#B77B5A",
  "Non-metazoan-only protein hits without nucleotide corroboration" = "#B76E49",
  "Concordant non-metazoan nucleotide and protein hits" = "#8E3B46",
  "No informative external evidence" = "#9A9A9A"
)
```

## Transferred files and validation

```{r transferred-files}
input_files <- data.frame(
  input = c(
    "Corrected scaffold evidence",
    "Corrected cross-species homology",
    "Candidate DEG genes",
    "Earlier transfer validation",
    "Homology transfer validation",
    "BLASTn raw hits",
    "DIAMOND/BLAST top-hit summary",
    "Gene homology summary",
    "NCBI taxonomy lookup"
  ),
  path = c(
    evidence_file, homology_file, candidate_gene_file, validation_file,
    homology_validation_file, blastn_file, homology_top_hit_file,
    homology_gene_file, taxonomy_file
  ),
  present = file.exists(c(
    evidence_file, homology_file, candidate_gene_file, validation_file,
    homology_validation_file, blastn_file, homology_top_hit_file,
    homology_gene_file, taxonomy_file
  ))
)

knitr::kable(input_files, caption = "Local inputs used for the origin audit")
if (!all(input_files$present)) {
  stop("One or more scaffold-audit inputs are missing.")
}

transfer_validation <- readr::read_tsv(validation_file, show_col_types = FALSE)
if (!all(transfer_validation$size_ok & transfer_validation$sha256_ok)) {
  stop("The earlier scaffold-audit transfer failed validation.")
}

homology_validation <- readr::read_tsv(
  homology_validation_file, show_col_types = FALSE
)
if (!all(homology_validation$size_ok & homology_validation$sha256_ok)) {
  stop("The BLASTn/DIAMOND transfer failed validation.")
}
```

All `r nrow(transfer_validation)` files from the earlier scaffold audit and all
`r nrow(homology_validation)` files from the BLASTn/DIAMOND transfer passed both
size and SHA-256 checks before analysis.

## BLASTn and DIAMOND origin audit

The homology search produced one important technical result that must be
corrected before biological interpretation. The raw BLASTn output contains
exact matches to taxid 7010, *S. gregaria*, even though that taxid was excluded
in the search command. If those self-hits are allowed to define the best bit
score, almost every external match falls outside the retained 5% score tier.

The code below therefore removes taxid 7010 **before** calculating the best
external score for each gene. It then retains external hits whose bit score is
at least 95% of the best remaining score. DIAMOND did not retain taxid 7010 and
is summarized from its transferred top-hit table.

```{r corrected-homology-load}
hit_fields <- c(
  "qseqid", "qlen", "sseqid", "staxids", "pident", "length",
  "qcovhsp", "evalue", "bitscore", "stitle"
)

classify_broad_group <- function(lineage, scientific_name) {
  value <- paste(lineage, scientific_name)
  dplyr::case_when(
    stringr::str_detect(value, "(^|;)Schistocerca(;| |$)") ~ "Schistocerca",
    stringr::str_detect(value, "(^|;)Acrididae(;|$)") ~ "Other Acrididae",
    stringr::str_detect(value, "(^|;)Orthoptera(;|$)") ~ "Other Orthoptera",
    stringr::str_detect(value, "(^|;)Arthropoda(;|$)") ~ "Other Arthropoda",
    stringr::str_detect(value, "(^|;)Metazoa(;|$)") ~ "Other Metazoa",
    stringr::str_detect(value, "(^|;)Fungi(;|$)") ~ "Fungi",
    stringr::str_detect(
      value, "(^|;)(Viridiplantae|Streptophyta)(;|$)"
    ) ~ "Plants",
    stringr::str_detect(value, "(^|;)Bacteria(;|$)") ~ "Bacteria",
    stringr::str_detect(value, "(^|;)Archaea(;|$)") ~ "Archaea",
    stringr::str_detect(value, "(^|;)Viruses(;|$)") ~ "Viruses",
    stringr::str_detect(value, "(^|;)Eukaryota(;|$)") ~ "Other Eukaryota",
    TRUE ~ "Other/unclassified"
  )
}

collapse_group_support <- function(groups) {
  metazoan <- groups %in% c(
    "Schistocerca", "Other Acrididae", "Other Orthoptera",
    "Other Arthropoda", "Other Metazoa"
  )
  non_metazoan <- groups %in% c(
    "Fungi", "Plants", "Bacteria", "Archaea", "Viruses",
    "Other Eukaryota"
  )
  if (any(metazoan) && any(non_metazoan)) {
    return("Mixed metazoan/non-metazoan")
  }
  if (any(groups %in% c(
    "Schistocerca", "Other Acrididae", "Other Orthoptera"
  ))) {
    return("Locust/orthopteran")
  }
  if (any(groups == "Other Arthropoda")) return("Other arthropod")
  if (any(groups == "Other Metazoa")) return("Other metazoan")
  if (any(non_metazoan)) return("Non-metazoan only")
  "No informative external hit"
}

blastn_raw <- readr::read_tsv(
  blastn_file,
  col_names = hit_fields,
  col_types = readr::cols(
    qseqid = readr::col_character(),
    qlen = readr::col_double(),
    sseqid = readr::col_character(),
    staxids = readr::col_character(),
    pident = readr::col_double(),
    length = readr::col_double(),
    qcovhsp = readr::col_double(),
    evalue = readr::col_double(),
    bitscore = readr::col_double(),
    stitle = readr::col_character()
  ),
  show_col_types = FALSE
) |>
  dplyr::mutate(
    self_hit_record = stringr::str_detect(
      paste0(";", staxids, ";"), ";7010;"
    )
  )

taxonomy_lookup <- readr::read_tsv(
  taxonomy_file,
  col_types = readr::cols(
    taxid = readr::col_integer(),
    scientific_name = readr::col_character(),
    rank = readr::col_character(),
    lineage = readr::col_character()
  ),
  show_col_types = FALSE
)

blastn_external <- blastn_raw |>
  tidyr::separate_longer_delim(staxids, delim = ";") |>
  dplyr::mutate(taxid = suppressWarnings(as.integer(staxids))) |>
  dplyr::filter(!is.na(taxid), taxid != 7010) |>
  dplyr::group_by(qseqid) |>
  dplyr::mutate(
    best_external_bitscore = max(bitscore),
    external_tier_cutoff = 0.95 * best_external_bitscore
  ) |>
  dplyr::filter(bitscore >= external_tier_cutoff) |>
  dplyr::ungroup() |>
  dplyr::left_join(taxonomy_lookup, by = "taxid") |>
  dplyr::mutate(
    broad_group = classify_broad_group(lineage, scientific_name),
    gene_id = stringr::str_extract(qseqid, "^[^|]+")
  )

nucleotide_gene <- blastn_external |>
  dplyr::group_by(gene_id) |>
  dplyr::summarise(
    nucleotide_support_corrected = collapse_group_support(
      unique(broad_group)
    ),
    nucleotide_groups = paste(sort(unique(broad_group)), collapse = ";"),
    nucleotide_taxa = paste(sort(unique(scientific_name)), collapse = ";"),
    nucleotide_best_evalue = min(evalue),
    nucleotide_max_identity = max(pident),
    nucleotide_max_query_coverage = max(qcovhsp),
    nucleotide_top_tier_records = dplyr::n(),
    .groups = "drop"
  )

homology_top_hits <- readr::read_tsv(
  homology_top_hit_file, show_col_types = FALSE
)
homology_gene <- readr::read_tsv(
  homology_gene_file, show_col_types = FALSE
) |>
  dplyr::rename(reported_nucleotide_support = nucleotide_support) |>
  dplyr::left_join(nucleotide_gene, by = "gene_id") |>
  dplyr::mutate(
    nucleotide_support_corrected = dplyr::coalesce(
      nucleotide_support_corrected, "No informative external hit"
    ),
    protein_evidence = dplyr::case_when(
      n_protein_queries == 0 ~ "lncRNA: no protein query",
      stringr::str_detect(protein_support, "Mixed/ambiguous") |
        (
          stringr::str_detect(protein_support, "Non-metazoan candidate") &
          stringr::str_detect(
            protein_support,
            "Locust/orthopteran|Broader arthropod|Other metazoan"
          )
        ) ~ "Mixed metazoan/non-metazoan",
      stringr::str_detect(
        protein_support, "Locust/orthopteran"
      ) ~ "Locust/orthopteran",
      stringr::str_detect(
        protein_support, "Broader arthropod"
      ) ~ "Other arthropod",
      stringr::str_detect(
        protein_support, "Other metazoan"
      ) ~ "Other metazoan",
      stringr::str_detect(
        protein_support, "Non-metazoan candidate"
      ) ~ "Non-metazoan only",
      TRUE ~ "No informative protein hit"
    ),
    integrated_evidence = dplyr::case_when(
      stringr::str_detect(
        nucleotide_support_corrected,
        "Locust/orthopteran|Other arthropod"
      ) ~ "External nucleotide support from Orthoptera/Arthropoda",
      stringr::str_detect(
        nucleotide_support_corrected, "Other metazoan"
      ) ~ "External nucleotide support from other Metazoa",
      nucleotide_support_corrected == "Mixed metazoan/non-metazoan" ~
        "Mixed external nucleotide evidence",
      nucleotide_support_corrected == "Non-metazoan only" &
        protein_evidence == "Non-metazoan only" ~
        "Concordant non-metazoan nucleotide and protein hits",
      nucleotide_support_corrected == "Non-metazoan only" &
        protein_evidence %in% c(
          "Locust/orthopteran", "Other arthropod", "Other metazoan",
          "Mixed metazoan/non-metazoan"
        ) ~ "Conflicting nucleotide and protein evidence",
      nucleotide_support_corrected == "Non-metazoan only" ~
        "Non-metazoan nucleotide evidence only",
      protein_evidence %in% c("Locust/orthopteran", "Other arthropod") ~
        "Arthropod/orthopteran protein support only",
      protein_evidence == "Other metazoan" ~
        "Other-metazoan protein support only",
      protein_evidence == "Mixed metazoan/non-metazoan" ~
        "Mixed protein evidence",
      protein_evidence == "Non-metazoan only" ~
        "Non-metazoan-only protein hits without nucleotide corroboration",
      TRUE ~ "No informative external evidence"
    )
  )

integrated_summary <- homology_gene |>
  dplyr::count(integrated_evidence, sort = TRUE, name = "genes") |>
  dplyr::mutate(
    percent = 100 * genes / sum(genes),
    integrated_evidence = factor(
      integrated_evidence,
      levels = rev(names(homology_colors))
    )
  )

readr::write_tsv(
  blastn_external,
  file.path(out_dir, "blastn_external_top_tier_hits.tsv")
)
readr::write_tsv(
  homology_gene,
  file.path(out_dir, "corrected_gene_homology_evidence.tsv")
)
readr::write_tsv(
  integrated_summary,
  file.path(out_dir, "corrected_homology_evidence_summary.tsv")
)
```

The raw BLASTn file contains `r scales::comma(nrow(blastn_raw))` hit records,
including `r scales::comma(sum(blastn_raw$self_hit_record))` records assigned to
*S. gregaria* itself. After removing taxid 7010 and re-ranking, only
`r scales::comma(nrow(nucleotide_gene))` of the 948 candidate genes have an
external nucleotide match passing the search thresholds. The other
`r scales::comma(948 - nrow(nucleotide_gene))` genes have no informative
external nucleotide match in this run.

### Taxonomic support by search

```{r homology-taxonomic-support, fig.width=14, fig.height=7}
blastn_gene_groups <- blastn_external |>
  dplyr::distinct(gene_id, broad_group) |>
  dplyr::mutate(search = "BLASTn: external nucleotide hits")

diamond_gene_groups <- homology_top_hits |>
  dplyr::filter(method == "DIAMOND nr_cluster_seq") |>
  dplyr::distinct(gene_id, broad_group) |>
  dplyr::mutate(search = "DIAMOND: predicted protein hits")

search_group_summary <- dplyr::bind_rows(
  blastn_gene_groups, diamond_gene_groups
) |>
  dplyr::count(search, broad_group, name = "genes")

readr::write_tsv(
  search_group_summary,
  file.path(out_dir, "homology_taxonomic_groups_by_search.tsv")
)

p_taxonomy <- ggplot2::ggplot(
  search_group_summary,
  ggplot2::aes(
    x = genes,
    y = stats::reorder(broad_group, genes),
    fill = search
  )
) +
  ggplot2::geom_col(width = 0.72, show.legend = FALSE) +
  ggplot2::geom_text(
    ggplot2::aes(label = scales::comma(genes)),
    hjust = -0.15,
    size = 4.2
  ) +
  ggplot2::facet_wrap(~ search, scales = "free_x", nrow = 1) +
  ggplot2::scale_fill_manual(values = c(
    "BLASTn: external nucleotide hits" = "#315C6B",
    "DIAMOND: predicted protein hits" = "#A8653E"
  )) +
  ggplot2::scale_x_continuous(
    labels = scales::label_comma(),
    expand = ggplot2::expansion(mult = c(0, 0.15))
  ) +
  ggplot2::labs(
    title = "External homology spans metazoan and non-metazoan groups",
    subtitle = "Counts are unique genes; a gene can contribute to more than one group",
    x = "Genes with a retained top-tier hit",
    y = NULL
  ) +
  ggplot2::theme_bw(base_size = 14) +
  ggplot2::theme(
    strip.text = ggplot2::element_text(face = "bold", size = 13),
    panel.grid.minor = ggplot2::element_blank()
  )

p_taxonomy
ggplot2::ggsave(
  file.path(out_dir, "homology_taxonomic_support_by_search.png"),
  p_taxonomy, width = 14, height = 7, dpi = 320
)
ggplot2::ggsave(
  file.path(out_dir, "homology_taxonomic_support_by_search.svg"),
  p_taxonomy, width = 14, height = 7
)
```

The DIAMOND result is dominated by non-metazoan top hits, but this does not
mean that those predicted proteins are microbial. NCBI `clustered_nr` is not a
balanced taxonomic sample, and many ancient eukaryotic proteins can have their
closest retained representative in fungi, plants, amoebozoans, or other
non-metazoan eukaryotes.

### Strength of the external nucleotide matches

```{r nucleotide-hit-strength, fig.width=11, fig.height=7}
nucleotide_plot_data <- nucleotide_gene |>
  dplyr::left_join(
    homology_gene |>
      dplyr::select(gene_id, integrated_evidence, max_abs_log2FoldChange),
    by = "gene_id",
    suffix = c("", ".joined")
  ) |>
  dplyr::mutate(
    review_flag = dplyr::if_else(
      integrated_evidence ==
        "Concordant non-metazoan nucleotide and protein hits",
      "Concordant non-metazoan top hits",
      "Other external match"
    )
  )

p_nucleotide_strength <- ggplot2::ggplot(
  nucleotide_plot_data,
  ggplot2::aes(
    x = nucleotide_max_query_coverage,
    y = nucleotide_max_identity,
    color = review_flag,
    size = max_abs_log2FoldChange
  )
) +
  ggplot2::geom_vline(xintercept = 50, linetype = "dashed", color = "grey55") +
  ggplot2::geom_hline(yintercept = 90, linetype = "dashed", color = "grey55") +
  ggplot2::geom_point(alpha = 0.78) +
  ggplot2::scale_color_manual(values = c(
    "Concordant non-metazoan top hits" = "#9D332F",
    "Other external match" = "#597C8D"
  )) +
  ggplot2::scale_size_continuous(range = c(2.5, 8)) +
  ggplot2::labs(
    title = "External nucleotide matches are generally divergent or partial",
    subtitle = "Dashed lines mark 90% identity and 50% query coverage for visual reference",
    x = "Maximum query coverage among top-tier external hits (%)",
    y = "Maximum nucleotide identity among top-tier external hits (%)",
    color = NULL,
    size = "Maximum absolute\nlog2 fold change"
  ) +
  ggplot2::theme_bw(base_size = 14) +
  ggplot2::theme(panel.grid.minor = ggplot2::element_blank())

p_nucleotide_strength
ggplot2::ggsave(
  file.path(out_dir, "external_nucleotide_hit_strength.png"),
  p_nucleotide_strength, width = 11, height = 7, dpi = 320
)
ggplot2::ggsave(
  file.path(out_dir, "external_nucleotide_hit_strength.svg"),
  p_nucleotide_strength, width = 11, height = 7
)
```

Among the
`r sum(homology_gene$integrated_evidence == "Concordant non-metazoan nucleotide and protein hits")`
genes with non-metazoan top hits in both searches, the median maximum nucleotide
identity is
`r round(median(homology_gene$nucleotide_max_identity[homology_gene$integrated_evidence == "Concordant non-metazoan nucleotide and protein hits"], na.rm = TRUE), 1)`%
and the median maximum query coverage is
`r round(median(homology_gene$nucleotide_max_query_coverage[homology_gene$integrated_evidence == "Concordant non-metazoan nucleotide and protein hits"], na.rm = TRUE), 1)`%.
None combines at least 90% identity with at least 50% query coverage. This is
not the near-identity signal expected for an obvious recently incorporated
microbial sequence.

### Integrated gene-level evidence

```{r integrated-homology-evidence, fig.width=13, fig.height=7.5}
p_integrated <- ggplot2::ggplot(
  integrated_summary,
  ggplot2::aes(
    x = genes,
    y = integrated_evidence,
    fill = integrated_evidence
  )
) +
  ggplot2::geom_col(width = 0.72, show.legend = FALSE) +
  ggplot2::geom_text(
    ggplot2::aes(
      label = paste0(
        scales::comma(genes), " (",
        scales::percent(percent / 100, accuracy = 0.1), ")"
      )
    ),
    hjust = -0.12,
    size = 4.1
  ) +
  ggplot2::scale_fill_manual(values = homology_colors, drop = FALSE) +
  ggplot2::scale_x_continuous(
    labels = scales::label_comma(),
    expand = ggplot2::expansion(mult = c(0, 0.25))
  ) +
  ggplot2::labs(
    title = "Integrated external homology evidence for 948 candidate genes",
    subtitle = "Protein-only non-metazoan hits remain weak provenance evidence",
    x = "Genes",
    y = NULL
  ) +
  ggplot2::theme_bw(base_size = 14) +
  ggplot2::theme(panel.grid.minor = ggplot2::element_blank())

p_integrated
ggplot2::ggsave(
  file.path(out_dir, "integrated_gene_homology_evidence.png"),
  p_integrated, width = 13, height = 7.5, dpi = 320
)
ggplot2::ggsave(
  file.path(out_dir, "integrated_gene_homology_evidence.svg"),
  p_integrated, width = 13, height = 7.5
)
```

Most flagged genes are in the protein-only non-metazoan or mixed-protein
classes. These classes describe the database hits; they do not establish the
organism of origin. Only
`r sum(homology_gene$integrated_evidence == "Concordant non-metazoan nucleotide and protein hits")`
genes have non-metazoan top hits in both the nucleotide and protein searches,
and their weak nucleotide identity and coverage still make them candidates for
review rather than confirmed contaminants.

### Searchable review list

```{r homology-candidate-table}
homology_review_candidates <- homology_gene |>
  dplyr::filter(
    integrated_evidence %in% c(
      "Concordant non-metazoan nucleotide and protein hits",
      "Conflicting nucleotide and protein evidence",
      "Non-metazoan-only protein hits without nucleotide corroboration"
    )
  ) |>
  dplyr::arrange(
    factor(
      integrated_evidence,
      levels = c(
        "Concordant non-metazoan nucleotide and protein hits",
        "Conflicting nucleotide and protein evidence",
        "Non-metazoan-only protein hits without nucleotide corroboration"
      )
    ),
    dplyr::desc(nucleotide_max_identity),
    dplyr::desc(max_abs_log2FoldChange)
  ) |>
  dplyr::transmute(
    gene_id,
    scaffold = seqid,
    description,
    gene_biotype,
    contrasts,
    max_abs_log2FoldChange = round(max_abs_log2FoldChange, 3),
    integrated_evidence,
    nucleotide_groups,
    nucleotide_taxa,
    nucleotide_max_identity = round(nucleotide_max_identity, 2),
    nucleotide_max_query_coverage = round(
      nucleotide_max_query_coverage, 2
    ),
    protein_support
  )

readr::write_tsv(
  homology_review_candidates,
  file.path(out_dir, "homology_review_candidate_genes.tsv")
)

DT::datatable(
  homology_review_candidates,
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    pageLength = 15,
    scrollX = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv")
  ),
  caption = paste(
    "Genes prioritized for review; categories are evidence labels,",
    "not automatic exclusion decisions"
  )
)
```

This table deliberately includes the broader protein-only class so individual
genes can be inspected, but the concordant class should be reviewed first. No
gene is removed from DESeq2 on the basis of this table.

### Scaffold-level concentration of review candidates

```{r homology-scaffold-table}
homology_scaffold_review <- homology_gene |>
  dplyr::group_by(scaffold = seqid) |>
  dplyr::summarise(
    candidate_DEG_genes = dplyr::n(),
    concordant_non_metazoan = sum(
      integrated_evidence ==
        "Concordant non-metazoan nucleotide and protein hits"
    ),
    non_metazoan_protein_only = sum(
      integrated_evidence ==
        "Non-metazoan-only protein hits without nucleotide corroboration"
    ),
    clear_metazoan_or_arthropod_support = sum(
      integrated_evidence %in% c(
        "External nucleotide support from Orthoptera/Arthropoda",
        "External nucleotide support from other Metazoa",
        "Arthropod/orthopteran protein support only",
        "Other-metazoan protein support only"
      )
    ),
    maximum_absolute_log2FC = max(max_abs_log2FoldChange, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    concordant_percent = 100 * concordant_non_metazoan /
      candidate_DEG_genes
  ) |>
  dplyr::arrange(
    dplyr::desc(concordant_non_metazoan),
    dplyr::desc(candidate_DEG_genes)
  )

readr::write_tsv(
  homology_scaffold_review,
  file.path(out_dir, "homology_review_by_scaffold.tsv")
)

DT::datatable(
  homology_scaffold_review |>
    dplyr::mutate(
      concordant_percent = round(concordant_percent, 1),
      maximum_absolute_log2FC = round(maximum_absolute_log2FC, 3)
    ),
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    pageLength = 15,
    scrollX = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv")
  ),
  caption = "Homology evidence summarized across all candidate genes on each scaffold"
)
```

Evidence from one conserved gene should not be used to label an entire
scaffold. Scaffolds with several concordant candidates are more useful targets
for whole-scaffold sequence and read-level follow-up than scaffolds flagged by
one protein alone.

## Evidence classification

```{r load-evidence}
evidence <- readr::read_tsv(evidence_file, show_col_types = FALSE) |>
  dplyr::mutate(
    evidence_class = factor(
      preliminary_evidence_status,
      levels = status_levels,
      labels = status_labels
    ),
    length_mb = length_bp / 1e6
  )

candidate_genes <- readr::read_tsv(candidate_gene_file, show_col_types = FALSE) |>
  dplyr::left_join(
    evidence |>
      dplyr::select(seqid = scaffold, evidence_class),
    by = "seqid"
  )

evidence_summary <- evidence |>
  dplyr::group_by(evidence_class) |>
  dplyr::summarise(
    scaffolds = dplyr::n(),
    scaffold_length_Mb = sum(length_bp) / 1e6,
    annotated_genes = sum(annotated_genes),
    candidate_DEG_genes = sum(candidate_deg_genes),
    median_gene_density_per_Mb = median(annotated_genes_per_mb),
    median_GC_percent = median(gc_percent),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    percent_candidate_DEGs = 100 * candidate_DEG_genes /
      sum(candidate_DEG_genes)
  )

readr::write_tsv(
  evidence_summary,
  file.path(out_dir, "scaffold_origin_evidence_summary.tsv")
)
knitr::kable(
  evidence_summary,
  digits = 2,
  caption = "Scaffold and DEG burden by independent origin-evidence class"
)
```

Of the `r nrow(evidence)` DEG-bearing unplaced scaffolds,
`r sum(evidence$evidence_class == "Strong host support")` have strong
cross-*Schistocerca* support and
`r sum(evidence$evidence_class == "Partial host support")` have partial support.
The remaining `r sum(evidence$evidence_class == "Unresolved")` are unclassified
by Kraken2, unflagged by the archived NCBI FCS report, and lack sufficient
high-confidence conservation in the three related locust assemblies.

Importantly, those unresolved scaffolds contain
`r scales::comma(sum(evidence$candidate_deg_genes[evidence$evidence_class == "Unresolved"]))`
of the `r scales::comma(sum(evidence$candidate_deg_genes))` candidate DEG genes
(`r scales::percent(sum(evidence$candidate_deg_genes[evidence$evidence_class == "Unresolved"]) / sum(evidence$candidate_deg_genes), accuracy = 0.1)`).
This concentration makes scaffold provenance a material sensitivity question,
but it is not proof of microbial origin.

## Scaffold structure and DEG burden

```{r scaffold-structure-plot, fig.width=14, fig.height=6.8}
top_unresolved <- evidence |>
  dplyr::filter(evidence_class == "Unresolved") |>
  dplyr::slice_max(candidate_deg_genes, n = 10, with_ties = FALSE)

p_density <- ggplot2::ggplot(
  evidence,
  ggplot2::aes(
    x = length_mb,
    y = annotated_genes_per_mb,
    color = evidence_class,
    size = candidate_deg_genes
  )
) +
  ggplot2::geom_point(alpha = 0.82) +
  ggrepel::geom_text_repel(
    data = top_unresolved,
    ggplot2::aes(label = scaffold),
    size = 3.7,
    max.overlaps = Inf,
    box.padding = 0.45,
    show.legend = FALSE
  ) +
  ggplot2::scale_x_log10(labels = scales::label_number(accuracy = 0.1)) +
  ggplot2::scale_y_log10(labels = scales::label_number(accuracy = 0.1)) +
  ggplot2::scale_color_manual(values = status_colors, drop = FALSE) +
  ggplot2::scale_size_continuous(range = c(2.5, 10)) +
  ggplot2::labs(
    title = "DEG-rich unresolved scaffolds are short and gene dense",
    x = "Scaffold length (Mb, log scale)",
    y = "Annotated genes per Mb (log scale)",
    color = "Origin evidence",
    size = "Candidate DEGs"
  ) +
  ggplot2::theme_bw(base_size = 14) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    legend.position = "right"
  )

p_burden <- ggplot2::ggplot(
  evidence_summary,
  ggplot2::aes(x = evidence_class, y = candidate_DEG_genes, fill = evidence_class)
) +
  ggplot2::geom_col(width = 0.68) +
  ggplot2::geom_text(
    ggplot2::aes(
      label = paste0(
        scales::comma(candidate_DEG_genes), "\n",
        scales::percent(percent_candidate_DEGs / 100, accuracy = 0.1)
      )
    ),
    vjust = -0.25,
    size = 4.5,
    fontface = "bold"
  ) +
  ggplot2::scale_fill_manual(values = status_colors, drop = FALSE) +
  ggplot2::scale_y_continuous(
    labels = scales::label_comma(),
    expand = ggplot2::expansion(mult = c(0, 0.15))
  ) +
  ggplot2::labs(
    title = "Candidate DEG burden by scaffold evidence",
    x = NULL,
    y = "Unique candidate DEG genes"
  ) +
  ggplot2::theme_bw(base_size = 14) +
  ggplot2::theme(
    legend.position = "none",
    axis.text.x = ggplot2::element_text(angle = 18, hjust = 1)
  )

combined_scaffold_plot <- p_density + p_burden +
  patchwork::plot_layout(widths = c(1.7, 1))
combined_scaffold_plot

ggplot2::ggsave(
  file.path(out_dir, "scaffold_origin_evidence_overview.png"),
  combined_scaffold_plot,
  width = 16,
  height = 7,
  dpi = 320
)
ggplot2::ggsave(
  file.path(out_dir, "scaffold_origin_evidence_overview.svg"),
  combined_scaffold_plot,
  width = 16,
  height = 7
)
```

The unresolved class has a median gene density of
`r round(evidence_summary$median_gene_density_per_Mb[evidence_summary$evidence_class == "Unresolved"], 1)`
annotated genes per Mb, compared with
`r round(evidence_summary$median_gene_density_per_Mb[evidence_summary$evidence_class == "Strong host support"], 1)`
for the strongly supported class. This unusual combination of short sequence,
high gene density, and weak cross-species support is compatible with several
alternatives: contamination, assembly fragmentation, collapsed or duplicated
gene-rich sequence, or genuinely lineage-specific locust sequence.

## Which DEG contrasts drive the unresolved signal?

```{r contrast-summary, fig.width=13, fig.height=7}
contrast_status <- candidate_genes |>
  dplyr::select(gene_id, seqid, evidence_class, contrasts) |>
  tidyr::separate_longer_delim(contrasts, delim = ";") |>
  dplyr::filter(!is.na(contrasts), contrasts != "") |>
  dplyr::distinct(gene_id, contrasts, evidence_class) |>
  dplyr::count(contrasts, evidence_class, name = "unique_DEG_genes") |>
  tidyr::complete(
    contrasts,
    evidence_class = factor(status_labels, levels = status_labels),
    fill = list(unique_DEG_genes = 0)
  ) |>
  dplyr::group_by(contrasts) |>
  dplyr::mutate(
    contrast_total = sum(unique_DEG_genes),
    percent = ifelse(contrast_total > 0, 100 * unique_DEG_genes / contrast_total, 0)
  ) |>
  dplyr::ungroup()

readr::write_tsv(
  contrast_status,
  file.path(out_dir, "contrast_by_scaffold_origin_evidence.tsv")
)

contrast_order <- contrast_status |>
  dplyr::group_by(contrasts) |>
  dplyr::summarise(total = max(contrast_total), .groups = "drop") |>
  dplyr::arrange(total) |>
  dplyr::pull(contrasts)

p_contrast <- contrast_status |>
  dplyr::mutate(contrasts = factor(contrasts, levels = contrast_order)) |>
  ggplot2::ggplot(
    ggplot2::aes(x = contrasts, y = unique_DEG_genes, fill = evidence_class)
  ) +
  ggplot2::geom_col(width = 0.72) +
  ggplot2::coord_flip() +
  ggplot2::scale_fill_manual(values = status_colors, drop = FALSE) +
  ggplot2::scale_y_continuous(labels = scales::label_comma()) +
  ggplot2::labs(
    title = "Origin evidence for unplaced-scaffold DEGs in each contrast",
    subtitle = "A gene may occur in more than one contrast",
    x = NULL,
    y = "Unique DEG genes",
    fill = "Origin evidence"
  ) +
  ggplot2::theme_bw(base_size = 14) +
  ggplot2::theme(panel.grid.minor = ggplot2::element_blank())

p_contrast
ggplot2::ggsave(
  file.path(out_dir, "contrast_by_scaffold_origin_evidence.png"),
  p_contrast,
  width = 13,
  height = 7,
  dpi = 320
)
ggplot2::ggsave(
  file.path(out_dir, "contrast_by_scaffold_origin_evidence.svg"),
  p_contrast,
  width = 13,
  height = 7
)
```

## High-priority unresolved scaffolds

```{r unresolved-table}
priority_unresolved <- evidence |>
  dplyr::filter(evidence_class == "Unresolved") |>
  dplyr::arrange(
    dplyr::desc(candidate_deg_genes),
    dplyr::desc(annotated_genes_per_mb)
  ) |>
  dplyr::transmute(
    scaffold,
    length_Mb = round(length_mb, 3),
    GC_percent = round(gc_percent, 2),
    annotated_genes,
    gene_density_per_Mb = round(annotated_genes_per_mb, 1),
    candidate_DEG_genes = candidate_deg_genes,
    DEG_contrasts = deg_contrasts,
    max_cross_species_coverage_percent =
      round(cross_species_max_coverage_percent, 2),
    cross_species_best_species,
    Kraken = kraken_taxon_name,
    NCBI_FCS_flagged = ncbi_fcs_flagged,
    recommended_next_test
  )

readr::write_tsv(
  priority_unresolved,
  file.path(out_dir, "high_priority_unresolved_scaffolds.tsv")
)

DT::datatable(
  priority_unresolved,
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    pageLength = 15,
    scrollX = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv")
  ),
  caption = "Unresolved scaffolds ranked by candidate DEG burden"
)
```

## Candidate genes for sensitivity analysis

```{r candidate-gene-table}
unresolved_genes <- candidate_genes |>
  dplyr::filter(evidence_class == "Unresolved") |>
  dplyr::arrange(seqid, gene_id) |>
  dplyr::select(
    gene_id, seqid, description, gene_biotype, mapping_support,
    n_contrasts, contrasts, min_padj, max_abs_log2FoldChange,
    competitive_host_retained_fraction, host_counts_not_retained,
    evidence_class
  )

readr::write_tsv(
  unresolved_genes,
  file.path(out_dir, "unresolved_scaffold_candidate_deg_genes.tsv")
)

biotype_summary <- unresolved_genes |>
  dplyr::count(gene_biotype, sort = TRUE, name = "candidate_DEG_genes") |>
  dplyr::mutate(
    percent = 100 * candidate_DEG_genes / sum(candidate_DEG_genes)
  )
readr::write_tsv(
  biotype_summary,
  file.path(out_dir, "unresolved_candidate_gene_biotypes.tsv")
)
knitr::kable(
  biotype_summary,
  digits = 2,
  caption = "Biotypes among unresolved-scaffold candidate DEG genes"
)

mapping_summary <- unresolved_genes |>
  dplyr::count(mapping_support, sort = TRUE, name = "candidate_DEG_genes")
readr::write_tsv(
  mapping_summary,
  file.path(out_dir, "unresolved_candidate_gene_mapping_support.tsv")
)
knitr::kable(
  mapping_summary,
  caption = "Host-only versus competitive host-fungus mapping support"
)
```

`r scales::comma(sum(unresolved_genes$mapping_support == "Significant in both mappings"))`
of the `r scales::comma(nrow(unresolved_genes))` unresolved-scaffold candidate
genes remained significant in both the host-only and competitive
host-*Metarhizium* mappings. Their median competitive-to-host-only retained
count fraction was
`r round(median(unresolved_genes$competitive_host_retained_fraction, na.rm = TRUE), 3)`.
Therefore, the signal is not explained by reads being preferentially reassigned
to the *M. robertsii* reference used in the competitive mapping. This result
does not exclude another taxon that was absent from the competitive reference.

The exported gene list is carried into the secondary unplaced-gene DESeq2 model
and the cluster 1 verification catalogue. Exclusion from the paper-facing
placed-host counts is a conservative reporting choice based on assembly
placement and uncertainty; it is not a claim that these genes are microbial.

## Interpretation and next tests

The combined audit provides four conclusions:

1. Some unplaced scaffolds are clearly or plausibly locust-derived because they
   are conserved in related *Schistocerca* assemblies.
2. Most of the unplaced-scaffold DEG burden is concentrated on 59 short,
   gene-dense scaffolds that the current Kraken2, FCS, and cross-species tests
   cannot resolve.
3. Competitive mapping against *Metarhizium robertsii* does not by itself
   establish the origin of these sequences, because absence from that single
   fungal reference does not distinguish locust-specific sequence from another
   microbe, assembly artifact, or divergent sequence.
4. The corrected BLASTn/DIAMOND analysis does not identify a near-identical
   non-metazoan sequence among the 948 genes. The 55 concordant non-metazoan
   candidates are divergent or partial at the nucleotide level and include
   many conserved cellular proteins. They warrant review but are not sufficient
   evidence for removal.

The most decisive follow-up is independent genomic evidence from the original
*S. gregaria* assembly project: genomic DNA depth, paired-read or long-read
support, and Hi-C linkage to chromosomes. For the highest-priority candidates,
the next sequence-level checks are whole-scaffold BLASTn, taxon-aware read
classification of reads mapping to each scaffold, GC and codon-usage comparison,
and examination of whether several neighboring genes support the same taxon.
Until those checks are complete, the reporting decision is
`placed_host_primary_unplaced_exploratory`, while the biological-origin status
of the unresolved genes remains `retain_pending_independent_validation`.

## Exported files

```{r exported-files}
exports <- list.files(out_dir, full.names = TRUE)
knitr::kable(
  data.frame(file = basename(exports), path = exports),
  caption = "Timestamped files produced by this page"
)
```

## Session information

```{r session-info}
sessionInfo()
```
