---
title: "Metarhizium burden and non-host taxonomic profiles"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  metadata_file: "data/metadata/mehreen_metadata_analysis_44_excluding_1044.txt"
  mapping_summary_file: "output/runs/host_pathogen_dual_corrected45_20260809-153146/05-mapping-comparison/competitive_mapping_summary.tsv"
  taxonomy_manifest_file: "output/runs/host_pathogen_dual_corrected45_20260809-153146/08-taxonomy/family_abundance_manifest.tsv"
  cluster_run_dir: ""
  output_tag: "metarhizium-read-burden"
---

## Purpose

::: {.walkthrough-grid}
::: {.walkthrough-step}
**Step 1 - Count fungal alignments**

Read the competitive STAR summary for each fat body library.
:::
::: {.walkthrough-step}
**Step 2 - Normalize signal**

Compare absolute fungal reads and their percentage of unique alignments.
:::
::: {.walkthrough-step}
**Step 3 - Interpret burden**

Use fungal RNA as supporting infection evidence, not as a direct biomass measurement.
:::
::: {.walkthrough-step}
**Step 4 - Compare non-host composition**

Use Kraken2 and Bracken to test whether retained non-host taxonomic family profiles differ among diets or treatments.
:::
:::

This page quantifies the RNA-seq evidence for *Metarhizium robertsii* in each
fat body library. It uses the competitive host-pathogen alignment, in which
each read is aligned against a combined *Schistocerca gregaria* and
*M. robertsii* reference. This avoids counting a read independently as both
host and fungus.

Two related measurements are reported:

1. **Fungal genome-mapped reads:** primary, uniquely mapped read alignments
   assigned by STAR anywhere on the *M. robertsii* genome.
2. **Unique fungal alignment percentage:** uniquely aligned reads assigned to the
   fungal reference, shown as a percentage of all unique alignments.

The first measurement is the closest available equivalent to a general
"mapped reads" plot because it is not restricted to annotated exons. The
percentage provides a library-normalized view of the same alignment evidence.
Neither measurement is a direct measurement of fungal biomass, and matched
qPCR validation was unavailable because no additional tissue remained.

## Setup and input validation

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c("dplyr", "ggplot2", "ggrepel", "scales", "patchwork", "DT",
              "knitr", "tidyr", "tibble", "stringr", "vegan")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

project_dir <- normalizePath(".", mustWork = TRUE)
metadata_file <- file.path(project_dir, params$metadata_file)
mapping_summary_file <- file.path(project_dir, params$mapping_summary_file)
taxonomy_manifest_file <- file.path(project_dir, params$taxonomy_manifest_file)

latest_run_dir <- function(prefix) {
  run_root <- file.path(project_dir, "output", "rmd_runs")
  candidates <- list.dirs(run_root, recursive = FALSE, full.names = TRUE)
  candidates <- candidates[grepl(paste0("^", prefix, "_"), basename(candidates))]
  if (length(candidates) == 0) {
    stop("No completed run found for: ", prefix)
  }
  candidates[[which.max(file.info(candidates)$mtime)]]
}

cluster_run_dir <- if (nzchar(params$cluster_run_dir)) {
  normalizePath(file.path(project_dir, params$cluster_run_dir), mustWork = TRUE)
} else {
  latest_run_dir("main-chromosome-cluster-enrichment")
}
cluster_matrix_file <- file.path(
  cluster_run_dir,
  "full_main_chromosome_deg_diagnostic_z_vst_matrix.csv.gz"
)
cluster_assignment_file <- file.path(
  cluster_run_dir,
  "figure3_k2_gene_assignments.csv"
)
mapping_mass_file <- file.path(
  latest_run_dir("count-qc-exploration"),
  "mapping_quality_and_individual_mass_45_samples.csv"
)

input_files <- c(
  metadata_file, mapping_summary_file,
  taxonomy_manifest_file, cluster_matrix_file, cluster_assignment_file,
  mapping_mass_file
)
if (!all(file.exists(input_files))) {
  stop("Missing input(s): ", paste(input_files[!file.exists(input_files)], collapse = ", "))
}

run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(
  project_dir, "output", "rmd_runs",
  paste(params$output_tag, run_stamp, sep = "_")
)
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

metadata <- read.delim(metadata_file, check.names = FALSE)
metadata <- metadata[, !is.na(names(metadata)) & nzchar(names(metadata)), drop = FALSE] |>
  dplyr::mutate(
    label = as.character(label),
    Treatment = factor(Treatment, levels = c("Control", "Infected")),
    Diet = factor(as.character(Diet), levels = c("33", "50", "83")),
    condition = factor(
      paste(Treatment, Diet),
      levels = c(
        "Control 33", "Control 50", "Control 83",
        "Infected 33", "Infected 50", "Infected 83"
      )
    )
  )

mapping_summary <- read.delim(mapping_summary_file, check.names = FALSE) |>
  dplyr::mutate(sample_id = as.character(sample_id))
missing_mapping <- setdiff(metadata$label, mapping_summary$sample_id)
if (length(missing_mapping) > 0) {
  stop("Missing competitive-mapping summaries for: ", paste(missing_mapping, collapse = ", "))
}
mapping_summary <- mapping_summary |>
  dplyr::filter(sample_id %in% metadata$label)
stopifnot(setequal(mapping_summary$sample_id, metadata$label))

fungal_burden <- mapping_summary |>
  dplyr::select(sample_id, unique_fungus, pct_unique_fungus) |>
  dplyr::rename(label = sample_id) |>
  dplyr::left_join(
    metadata |>
      dplyr::select(label, Treatment, Diet, condition),
    by = "label"
  ) |>
  dplyr::mutate(
    sample_name = as.character(label),
    condition = factor(condition, levels = levels(metadata$condition))
  ) |>
  dplyr::arrange(condition, as.numeric(label))

write.csv(
  fungal_burden,
  file.path(out_dir, "metarhizium_read_burden_by_sample.csv"),
  row.names = FALSE
)
```

## Fungal genome-mapped reads

This is the broad mapping view most comparable to the previously generated
"CDS alignment mapped reads" figure. Here, however, reads were aligned
competitively against the complete locust and fungal genomes. The bars count
primary read alignments that mapped uniquely anywhere on the fungal genome;
they are not restricted to exons or CDS features.

The values will not be identical to a direct CDS alignment because the
reference target, competition with the host genome, alignment algorithm, and
multimapping filters differ.

```{r fungal-genome-mapped-reads, fig.width=20, fig.height=8}
mapped_highlight_threshold <- stats::quantile(
  fungal_burden$unique_fungus,
  probs = 0.90,
  na.rm = TRUE
)

fungal_genome_plot <- ggplot2::ggplot(
  fungal_burden,
  ggplot2::aes(x = sample_name, y = unique_fungus)
) +
  ggplot2::geom_col(
    fill = "#4C78A8",
    color = "#17202A",
    linewidth = 0.35,
    width = 0.72
  ) +
  ggplot2::geom_text(
    data = fungal_burden |>
      dplyr::filter(unique_fungus >= mapped_highlight_threshold),
    ggplot2::aes(label = scales::comma(unique_fungus)),
    vjust = -0.35,
    size = 3.8,
    fontface = "bold"
  ) +
  ggplot2::facet_grid(
    cols = ggplot2::vars(condition),
    scales = "free_x",
    space = "free_x"
  ) +
  ggplot2::scale_y_continuous(
    labels = scales::label_comma(),
    expand = ggplot2::expansion(mult = c(0, 0.12))
  ) +
  ggplot2::labs(
    title = "Metarhizium robertsii genome-mapped reads",
    subtitle = "Unique primary reads from competitive host-pathogen STAR alignment",
    x = NULL,
    y = "Unique fungal read alignments"
  ) +
  ggplot2::theme_bw(base_size = 13) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_blank(),
    strip.background = ggplot2::element_rect(fill = "#E5E7EB", color = "#374151"),
    strip.text = ggplot2::element_text(face = "bold", size = 12),
    axis.text.x = ggplot2::element_text(angle = 58, hjust = 1, size = 9),
    plot.title = ggplot2::element_text(face = "bold", size = 18)
  )

print(fungal_genome_plot)
```

## Full and zoomed fungal read counts

The zoom retains every individual library. Bars above 7,000 reads are capped
at the upper boundary and labeled with their complete values, so their sample
positions remain visible. The dashed gold horizontal line gives the mean
fungal read count across all 20 control individuals (146.5 reads).

```{r fungal-genome-mapped-reads-zoom-7000, fig.width=20, fig.height=8}
read_zoom_limit <- 7000
control_mean_reads <- mean(
  fungal_burden$unique_fungus[fungal_burden$Treatment == "Control"],
  na.rm = TRUE
)

read_zoom_samples <- fungal_burden |>
  dplyr::mutate(
    display_value = pmin(unique_fungus, read_zoom_limit)
  )

fungal_genome_zoom_plot <- ggplot2::ggplot(
  read_zoom_samples,
  ggplot2::aes(x = sample_name, y = display_value)
) +
  ggplot2::geom_col(
    fill = "#4C78A8",
    color = "#17202A",
    linewidth = 0.35,
    width = 0.72
  ) +
  ggplot2::geom_hline(
    yintercept = control_mean_reads,
    color = "#B7791F",
    linewidth = 0.9,
    linetype = "dashed"
  ) +
  ggplot2::geom_text(
    data = read_zoom_samples |>
      dplyr::filter(unique_fungus > read_zoom_limit),
    ggplot2::aes(
      x = sample_name,
      y = read_zoom_limit * 0.94,
      label = paste0(scales::comma(unique_fungus), " (>7k)")
    ),
    inherit.aes = FALSE,
    angle = 90,
    hjust = 1,
    vjust = 0.5,
    size = 3.3,
    fontface = "bold",
    color = "white"
  ) +
  ggplot2::facet_grid(
    cols = ggplot2::vars(condition),
    scales = "free_x",
    space = "free_x"
  ) +
  ggplot2::scale_y_continuous(
    limits = c(0, read_zoom_limit),
    breaks = seq(0, read_zoom_limit, by = 1000),
    labels = scales::label_comma(),
    expand = ggplot2::expansion(mult = c(0, 0.01))
  ) +
  ggplot2::labs(
    title = "Zoomed Metarhizium robertsii alignment signal",
    subtitle = paste0(
      "All samples retained; bars above 7,000 are capped; dashed line = control mean (",
      round(control_mean_reads, 1), " reads)"
    ),
    x = NULL,
    y = "Unique fungal read alignments"
  ) +
  ggplot2::theme_bw(base_size = 13) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_blank(),
    strip.background = ggplot2::element_rect(fill = "#E5E7EB", color = "#374151"),
    strip.text = ggplot2::element_text(face = "bold", size = 12),
    axis.text.x = ggplot2::element_text(angle = 58, hjust = 1, size = 8.5),
    plot.title = ggplot2::element_text(face = "bold", size = 18)
  )

print(fungal_genome_zoom_plot)
```

```{r fungal-genome-full-and-zoom, fig.width=20, fig.height=16}
fungal_genome_full_zoom_plot <-
  (fungal_genome_plot / fungal_genome_zoom_plot) +
  patchwork::plot_annotation(tag_levels = "A")

print(fungal_genome_full_zoom_plot)
```

## Library-normalized fungal alignment signal

The percentage view controls for differences in total unique alignment depth.
It confirms whether samples with many fungal alignments also devote a larger
fraction of their sequenced library to the fungal reference.

```{r fungal-alignment-percentage, fig.width=20, fig.height=8}
fungal_percentage_plot <- ggplot2::ggplot(
  fungal_burden,
  ggplot2::aes(x = sample_name, y = pct_unique_fungus)
) +
  ggplot2::geom_col(
    fill = "#8B1A1A",
    color = "#17202A",
    linewidth = 0.35,
    width = 0.72
  ) +
  ggplot2::facet_grid(
    cols = ggplot2::vars(condition),
    scales = "free_x",
    space = "free_x"
  ) +
  ggplot2::scale_y_continuous(
    labels = scales::label_number(suffix = "%", accuracy = 0.01),
    expand = ggplot2::expansion(mult = c(0, 0.08))
  ) +
  ggplot2::labs(
    title = "Fraction of unique alignments assigned to M. robertsii",
    subtitle = "Unique fungal alignments divided by all unique competitive alignments",
    x = NULL,
    y = "Unique fungal alignments (%)"
  ) +
  ggplot2::theme_bw(base_size = 13) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_blank(),
    strip.background = ggplot2::element_rect(fill = "#E5E7EB", color = "#374151"),
    strip.text = ggplot2::element_text(face = "bold", size = 12),
    axis.text.x = ggplot2::element_text(angle = 58, hjust = 1, size = 9),
    plot.title = ggplot2::element_text(face = "bold", size = 18)
  )

print(fungal_percentage_plot)
```

## Full and zoomed fungal alignment percentages

The percentage zoom also retains every sample. Values exceeding 0.1% reach
the upper boundary and carry their complete percentages. The dashed gold line
is the mean across controls (0.00042% of unique competitive alignments), shown
as a common background reference.

```{r fungal-alignment-percentage-zoom-0-1, fig.width=20, fig.height=8}
percentage_zoom_limit <- 0.1
control_mean_percentage <- mean(
  fungal_burden$pct_unique_fungus[fungal_burden$Treatment == "Control"],
  na.rm = TRUE
)

percentage_zoom_samples <- fungal_burden |>
  dplyr::mutate(
    display_value = pmin(pct_unique_fungus, percentage_zoom_limit)
  )

fungal_percentage_zoom_plot <- ggplot2::ggplot(
  percentage_zoom_samples,
  ggplot2::aes(x = sample_name, y = display_value)
) +
  ggplot2::geom_col(
    fill = "#8B1A1A",
    color = "#17202A",
    linewidth = 0.35,
    width = 0.72
  ) +
  ggplot2::geom_hline(
    yintercept = control_mean_percentage,
    color = "#B7791F",
    linewidth = 0.9,
    linetype = "dashed"
  ) +
  ggplot2::geom_text(
    data = percentage_zoom_samples |>
      dplyr::filter(pct_unique_fungus > percentage_zoom_limit),
    ggplot2::aes(
      x = sample_name,
      y = percentage_zoom_limit * 0.94,
      label = paste0(
        scales::number(pct_unique_fungus, accuracy = 0.001),
        "% (>0.1%)"
      )
    ),
    inherit.aes = FALSE,
    angle = 90,
    hjust = 1,
    vjust = 0.5,
    size = 3.3,
    fontface = "bold",
    color = "white"
  ) +
  ggplot2::facet_grid(
    cols = ggplot2::vars(condition),
    scales = "free_x",
    space = "free_x"
  ) +
  ggplot2::scale_y_continuous(
    limits = c(0, percentage_zoom_limit),
    breaks = seq(0, percentage_zoom_limit, by = 0.025),
    labels = scales::label_number(suffix = "%", accuracy = 0.025),
    expand = ggplot2::expansion(mult = c(0, 0.01))
  ) +
  ggplot2::labs(
    title = "Zoomed fraction of alignments assigned to M. robertsii",
    subtitle = paste0(
      "All samples retained; values above 0.1% are capped; dashed line = control mean (",
      scales::number(control_mean_percentage, accuracy = 0.00001), "%)"
    ),
    x = NULL,
    y = "Unique fungal alignments (%)"
  ) +
  ggplot2::theme_bw(base_size = 13) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_blank(),
    strip.background = ggplot2::element_rect(fill = "#E5E7EB", color = "#374151"),
    strip.text = ggplot2::element_text(face = "bold", size = 12),
    axis.text.x = ggplot2::element_text(angle = 58, hjust = 1, size = 8.5),
    plot.title = ggplot2::element_text(face = "bold", size = 18)
  )

print(fungal_percentage_zoom_plot)
```

```{r fungal-percentage-full-and-zoom, fig.width=20, fig.height=16}
fungal_percentage_full_zoom_plot <-
  (fungal_percentage_plot / fungal_percentage_zoom_plot) +
  patchwork::plot_annotation(tag_levels = "A")

print(fungal_percentage_full_zoom_plot)
```

## Fungal percentage aligned above the host DEG heatmap

This figure aligns the 0-0.1% fungal-percentage zoom directly above the complete
two-cluster main-chromosome DEG heatmap. Columns are ordered first by treatment,
then diet, and finally sample number. Values above 0.1% are capped only for the
display and retain their complete percentage as a white label. The heatmap
contains all significant main-chromosome DEGs used in the current Figure 3, not
only genes annotated with GO or KEGG terms.

```{r fungal-percentage-above-deg-heatmap, fig.width=22, fig.height=16}
cluster_z <- read.csv(
  gzfile(cluster_matrix_file),
  check.names = FALSE,
  row.names = 1
)
cluster_z <- as.matrix(cluster_z)
storage.mode(cluster_z) <- "numeric"

cluster_assignments <- read.csv(
  cluster_assignment_file,
  check.names = FALSE,
  stringsAsFactors = FALSE
) |>
  dplyr::mutate(
    expression_cluster = factor(
      expression_cluster,
      levels = c("DEG cluster 1", "DEG cluster 2")
    ),
    DEG_context_plot = dplyr::case_when(
      DEG_context == "Diet in controls" ~ "~ diet (controls)",
      DEG_context == "Diet in infected" ~ "~ diet (infected)",
      DEG_context == "Infection within diet" ~ "~ diet + infection",
      DEG_context %in% c("Infection + diet", "Diet in controls + infected") ~
        "shared DEGs",
      TRUE ~ "Other fat body DEG"
    )
  ) |>
  dplyr::arrange(expression_cluster, within_cluster_order)

sample_order_table <- fungal_burden |>
  dplyr::arrange(condition, as.numeric(label)) |>
  dplyr::mutate(sample_index = dplyr::row_number())
sample_order <- sample_order_table$label

if (!setequal(sample_order, colnames(cluster_z))) {
  stop("Fungal-burden and DEG-heatmap samples do not match.")
}
if (!all(cluster_assignments$gene_id %in% rownames(cluster_z))) {
  stop("One or more Figure 3 genes are absent from the VST matrix.")
}

sample_groups <- sample_order_table |>
  dplyr::group_by(condition, Treatment, Diet) |>
  dplyr::summarise(
    start = min(sample_index),
    end = max(sample_index),
    .groups = "drop"
  ) |>
  dplyr::arrange(start)
sample_gaps <- utils::head(sample_groups$end, -1) + 0.5
treatment_gaps <- sample_groups$end[
  which(
    as.character(sample_groups$Treatment[-nrow(sample_groups)]) !=
      as.character(sample_groups$Treatment[-1])
  )
] + 0.5
diet_gaps <- setdiff(sample_gaps, treatment_gaps)
sample_x_limits <- c(-1.25, length(sample_order) + 0.5)

fungal_heatmap_percent <- sample_order_table |>
  dplyr::mutate(
    display_percentage = pmin(pct_unique_fungus, percentage_zoom_limit)
  )

fungal_percentage_track <- ggplot2::ggplot(
  fungal_heatmap_percent,
  ggplot2::aes(
    x = sample_index,
    y = display_percentage,
    fill = Treatment
  )
) +
  ggplot2::geom_col(width = 0.82, color = "#111827", linewidth = 0.25) +
  ggplot2::geom_hline(
    yintercept = control_mean_percentage,
    color = "#B7791F",
    linewidth = 0.9,
    linetype = "dashed"
  ) +
  ggplot2::geom_vline(
    xintercept = diet_gaps,
    color = "black",
    linewidth = 0.25
  ) +
  ggplot2::geom_vline(
    xintercept = treatment_gaps,
    color = "black",
    linewidth = 0.7
  ) +
  ggplot2::geom_text(
    data = fungal_heatmap_percent |>
      dplyr::filter(pct_unique_fungus > percentage_zoom_limit),
    ggplot2::aes(
      label = paste0(
        scales::number(pct_unique_fungus, accuracy = 0.001),
        "%"
      ),
      y = percentage_zoom_limit * 0.94
    ),
    color = "white",
    angle = 90,
    hjust = 1,
    vjust = 0.5,
    size = 5.8,
    fontface = "bold",
    show.legend = FALSE
  ) +
  ggplot2::scale_fill_manual(
    values = c(Control = "#FAEBD7", Infected = "#8B0000"),
    guide = "none"
  ) +
  ggplot2::scale_x_continuous(
    limits = sample_x_limits,
    breaks = sample_order_table$sample_index,
    labels = sample_order,
    expand = c(0, 0)
  ) +
  ggplot2::scale_y_continuous(
    limits = c(0, percentage_zoom_limit),
    breaks = seq(0, percentage_zoom_limit, by = 0.025),
    labels = scales::label_number(suffix = "%", accuracy = 0.025),
    expand = c(0, 0)
  ) +
  ggplot2::labs(
    title = "Fungal alignment percentage above host DEG expression",
    subtitle = paste0(
      "All samples retained; values above 0.1% are capped and labeled; ",
      "dashed line = control mean (",
      scales::number(control_mean_percentage, accuracy = 0.00001),
      "%)"
    ),
    x = NULL,
    y = "Unique fungal\nalignments (%)"
  ) +
  ggplot2::theme_classic(base_size = 13) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank(),
    plot.title = ggplot2::element_text(face = "bold", size = 18),
    plot.subtitle = ggplot2::element_text(size = 12),
    plot.margin = ggplot2::margin(5, 5, 0, 5)
  )

sample_annotation_long <- sample_order_table |>
  dplyr::transmute(
    sample_index,
    Treatment = as.character(Treatment),
    Diet = as.character(Diet)
  ) |>
  tidyr::pivot_longer(
    cols = c(Treatment, Diet),
    names_to = "track",
    values_to = "value"
  ) |>
  dplyr::mutate(
    track = factor(track, levels = c("Diet", "Treatment")),
    track_y = as.numeric(track),
    annotation_key = paste(track, value, sep = ": ")
  )

sample_annotation_colors <- c(
  "Treatment: Control" = "#FAEBD7",
  "Treatment: Infected" = "#8B0000",
  "Diet: 33" = "#E6862E",
  "Diet: 50" = "#2F8F5B",
  "Diet: 83" = "#7B4FA3"
)

fungal_heatmap_annotation <- ggplot2::ggplot(
  sample_annotation_long,
  ggplot2::aes(x = sample_index, y = track_y, fill = annotation_key)
) +
  ggplot2::geom_tile(width = 0.95, height = 0.9) +
  ggplot2::geom_vline(xintercept = diet_gaps, color = "black", linewidth = 0.25) +
  ggplot2::geom_vline(
    xintercept = treatment_gaps,
    color = "black",
    linewidth = 0.7
  ) +
  ggplot2::scale_fill_manual(values = sample_annotation_colors, guide = "none") +
  ggplot2::scale_x_continuous(
    limits = sample_x_limits,
    expand = c(0, 0)
  ) +
  ggplot2::scale_y_continuous(
    breaks = seq_len(2),
    labels = c("Diet", "Treatment"),
    limits = c(0.45, 2.55),
    expand = c(0, 0)
  ) +
  ggplot2::labs(x = NULL, y = NULL) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank(),
    axis.text.y = ggplot2::element_text(face = "bold", size = 11),
    plot.margin = ggplot2::margin(0, 5, 0, 5)
  )

heatmap_matrix <- cluster_z[
  cluster_assignments$gene_id,
  sample_order,
  drop = FALSE
]
heatmap_matrix <- pmax(pmin(heatmap_matrix, 3), -3)
heatmap_df <- as.data.frame(heatmap_matrix, check.names = FALSE) |>
  tibble::rownames_to_column("gene_id") |>
  dplyr::left_join(
    cluster_assignments |>
      dplyr::select(
        gene_id, expression_cluster, within_cluster_order, DEG_context_plot
      ),
    by = "gene_id"
  ) |>
  tidyr::pivot_longer(
    cols = dplyr::all_of(sample_order),
    names_to = "sample_id",
    values_to = "z"
  ) |>
  dplyr::mutate(sample_index = match(sample_id, sample_order))

context_levels <- c(
  "~ diet (controls)", "~ diet (infected)", "~ diet + infection",
  "shared DEGs", "Other fat body DEG"
)
context_colors <- c(
  "~ diet (controls)" = "#DDE3EA",
  "~ diet (infected)" = "#AEBBC8",
  "~ diet + infection" = "#65788C",
  "shared DEGs" = "#263544",
  "Other fat body DEG" = "#F2F4F6"
)
cluster_assignments$DEG_context_plot <- factor(
  cluster_assignments$DEG_context_plot,
  levels = context_levels
)
present_contexts <- context_levels[
  context_levels %in% as.character(cluster_assignments$DEG_context_plot)
]

fungal_deg_heatmap <- ggplot2::ggplot(
  heatmap_df,
  ggplot2::aes(x = sample_index, y = within_cluster_order, fill = z)
) +
  ggplot2::geom_segment(
    data = cluster_assignments,
    ggplot2::aes(
      x = -1.1,
      xend = 0.35,
      y = within_cluster_order,
      yend = within_cluster_order,
      color = DEG_context_plot
    ),
    inherit.aes = FALSE,
    linewidth = 1.15,
    lineend = "butt"
  ) +
  ggplot2::geom_raster() +
  ggplot2::geom_vline(xintercept = 0.5, color = "#9CA3AF", linewidth = 0.4) +
  ggplot2::geom_vline(xintercept = diet_gaps, color = "black", linewidth = 0.25) +
  ggplot2::geom_vline(
    xintercept = treatment_gaps,
    color = "black",
    linewidth = 0.7
  ) +
  ggplot2::facet_grid(
    expression_cluster ~ .,
    scales = "free_y",
    space = "free_y",
    switch = "y"
  ) +
  ggplot2::scale_fill_gradientn(
    colors = grDevices::colorRampPalette(
      c("#237C8E", "#F7F3E8", "#C85A54")
    )(101),
    limits = c(-3, 3),
    oob = scales::squish,
    name = "Row z-score"
  ) +
  ggplot2::scale_color_manual(
    values = context_colors,
    breaks = present_contexts,
    labels = stringr::str_wrap(present_contexts, 24),
    drop = TRUE,
    name = "DEG context"
  ) +
  ggplot2::scale_x_continuous(
    limits = sample_x_limits,
    breaks = sample_order_table$sample_index,
    labels = sample_order,
    expand = c(0, 0)
  ) +
  ggplot2::scale_y_reverse(expand = c(0, 0)) +
  ggplot2::labs(x = NULL, y = NULL) +
  ggplot2::theme_classic(base_size = 13) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(
      angle = 90,
      hjust = 1,
      vjust = 0.5,
      size = 12
    ),
    axis.text.y = ggplot2::element_blank(),
    axis.ticks.y = ggplot2::element_blank(),
    strip.placement = "outside",
    strip.background.y = ggplot2::element_rect(
      fill = "#F3F4F6",
      color = "#111827"
    ),
    strip.text.y.left = ggplot2::element_text(
      angle = 90,
      face = "bold",
      size = 14
    ),
    panel.spacing.y = grid::unit(0.2, "lines"),
    legend.position = "right",
    legend.box = "vertical",
    legend.title = ggplot2::element_text(face = "bold"),
    plot.margin = ggplot2::margin(0, 5, 5, 5)
  )

fungal_percentage_deg_heatmap <-
  fungal_percentage_track /
  fungal_heatmap_annotation /
  fungal_deg_heatmap +
  patchwork::plot_layout(heights = c(2.2, 0.65, 8.5))

print(fungal_percentage_deg_heatmap)

write.csv(
  fungal_heatmap_percent |>
    dplyr::select(
      sample_index, label, Treatment, Diet,
      pct_unique_fungus, display_percentage
    ),
  file.path(out_dir, "fungal_percentage_heatmap_sample_order.csv"),
  row.names = FALSE
)
```

## Kraken2 and Bracken family composition

Competitive alignment measures reads assigned to the focal fungal reference.
Kraken2 and Bracken provide a broader, complementary view of the non-host
taxonomic signal. The main composition analysis uses reads that did not align
to the host-only reference. This enriches microbial signal while retaining
fungal reads; results from all trimmed and competitive-unmapped reads remain in
the transferred taxonomy directory as sensitivity branches.

These profiles should not be described automatically as contamination. They can
contain the challenge fungus, environmental or diet-associated organisms,
reagent background, and taxonomic assignments caused by incomplete reference
databases. Metazoan and plant families are therefore excluded from the
composition display, including host-like assignments such as Hominidae. The
retained profiles comprise fungi, bacteria, archaea, viruses, and other
microbial eukaryotes. Relative abundance is calculated among Bracken-estimated
reads assigned to these retained non-host taxonomic families, not as a
percentage of the entire RNA-seq library.

```{r bracken-family-composition-prepare}
read_kraken_family_lineages <- function(path) {
  fields <- strsplit(readLines(path, warn = FALSE), "\t", fixed = TRUE)
  stack <- character()
  result <- vector("list", length(fields))
  n_result <- 0L

  for (entry in fields) {
    if (length(entry) < 6) next
    rank <- trimws(entry[[4]])
    taxon_id <- trimws(entry[[5]])
    raw_name <- entry[[6]]
    taxon_name <- trimws(raw_name)
    indent <- nchar(raw_name) - nchar(sub("^[[:space:]]+", "", raw_name))
    depth <- as.integer(indent / 2)

    if (length(stack) > depth) stack <- stack[seq_len(depth)]
    stack[depth + 1L] <- taxon_name

    if (rank == "F") {
      n_result <- n_result + 1L
      result[[n_result]] <- data.frame(
        taxonomy_id = taxon_id,
        family = taxon_name,
        lineage = paste(stack, collapse = "; "),
        stringsAsFactors = FALSE
      )
    }
  }

  if (n_result == 0L) {
    return(data.frame(
      taxonomy_id = character(), family = character(), lineage = character()
    ))
  }
  dplyr::bind_rows(result[seq_len(n_result)])
}

classify_family_lineage <- function(lineage) {
  dplyr::case_when(
    grepl("(^|; )Metazoa(;|$)", lineage) ~ "Metazoa",
    grepl("Viridiplantae|Streptophyta", lineage) ~ "Plants",
    grepl("(^|; )Fungi(;|$)", lineage) ~ "Fungi",
    grepl("(^|; )Bacteria(;|$)", lineage) ~ "Bacteria",
    grepl("(^|; )Archaea(;|$)", lineage) ~ "Archaea",
    grepl("Viruses|Duplodnaviria|Riboviria|Monodnaviria", lineage) ~ "Viruses",
    grepl("(^|; )Eukaryota(;|$)", lineage) ~ "Other microbial eukaryotes",
    TRUE ~ "Other"
  )
}

taxonomy_manifest <- read.delim(
  taxonomy_manifest_file,
  check.names = FALSE,
  stringsAsFactors = FALSE
) |>
  dplyr::mutate(sample_id = as.character(sample_id)) |>
  dplyr::filter(source == "host_unmapped", sample_id %in% metadata$label) |>
  dplyr::arrange(as.numeric(sample_id))

if (nrow(taxonomy_manifest) != nrow(metadata)) {
  stop(
    "Expected one host-unmapped taxonomy entry per retained sample; found ",
    nrow(taxonomy_manifest), " for ", nrow(metadata), " samples."
  )
}

resolve_manifest_path <- function(path) {
  if (file.exists(path)) return(normalizePath(path))
  candidate <- file.path(project_dir, path)
  if (!file.exists(candidate)) stop("Missing taxonomy file: ", path)
  normalizePath(candidate)
}

taxonomy_manifest <- taxonomy_manifest |>
  dplyr::mutate(
    abundance_file = vapply(abundance_file, resolve_manifest_path, character(1)),
    kraken_report = vapply(kraken_report, resolve_manifest_path, character(1))
  )

family_lineages <- dplyr::bind_rows(lapply(
  taxonomy_manifest$kraken_report,
  read_kraken_family_lineages
)) |>
  dplyr::distinct(taxonomy_id, .keep_all = TRUE) |>
  dplyr::mutate(taxonomic_group = classify_family_lineage(lineage))

bracken_family <- dplyr::bind_rows(lapply(
  seq_len(nrow(taxonomy_manifest)),
  function(i) {
    read.delim(
      taxonomy_manifest$abundance_file[[i]],
      check.names = FALSE,
      stringsAsFactors = FALSE
    ) |>
      dplyr::transmute(
        label = taxonomy_manifest$sample_id[[i]],
        taxonomy_id = as.character(taxonomy_id),
        family = as.character(name),
        estimated_reads = as.numeric(new_est_reads)
      )
  }
)) |>
  dplyr::left_join(
    family_lineages |>
      dplyr::select(taxonomy_id, lineage, taxonomic_group),
    by = "taxonomy_id"
  ) |>
  dplyr::mutate(
    taxonomic_group = dplyr::coalesce(taxonomic_group, "Other")
  ) |>
  dplyr::filter(
    estimated_reads > 0,
    taxonomic_group %in% c(
      "Fungi", "Bacteria", "Archaea", "Viruses",
      "Other microbial eukaryotes"
    )
  ) |>
  dplyr::left_join(
    metadata |>
      dplyr::select(label, Treatment, Diet, condition),
    by = "label"
  )

sample_microbe_totals <- bracken_family |>
  dplyr::group_by(label) |>
  dplyr::summarise(microbial_family_reads = sum(estimated_reads), .groups = "drop")

mapping_mass <- read.csv(mapping_mass_file, check.names = FALSE) |>
  dplyr::mutate(label = as.character(label)) |>
  dplyr::filter(label %in% metadata$label)
stopifnot(setequal(mapping_mass$label, metadata$label))

microbial_burden <- sample_microbe_totals |>
  dplyr::left_join(
    mapping_mass |>
      dplyr::select(label, input_reads, mass_g),
    by = "label"
  ) |>
  dplyr::left_join(
    metadata |>
      dplyr::select(label, Diet, Treatment, condition),
    by = "label"
  ) |>
  dplyr::mutate(
    microbial_reads_per_million =
      microbial_family_reads / input_reads * 1e6,
    log10_microbial_rpm = log10(microbial_reads_per_million + 1)
  )

bracken_family <- bracken_family |>
  dplyr::left_join(sample_microbe_totals, by = "label") |>
  dplyr::mutate(
    relative_abundance = estimated_reads / microbial_family_reads
  )

top_family_levels <- bracken_family |>
  dplyr::group_by(family) |>
  dplyr::summarise(total_reads = sum(estimated_reads), .groups = "drop") |>
  dplyr::slice_max(total_reads, n = 12, with_ties = FALSE) |>
  dplyr::arrange(dplyr::desc(total_reads)) |>
  dplyr::pull(family)

sample_order_microbes <- metadata |>
  dplyr::arrange(condition, as.numeric(label)) |>
  dplyr::pull(label)

bracken_display <- bracken_family |>
  dplyr::mutate(
    display_family = dplyr::if_else(
      family %in% top_family_levels,
      family,
      "Other microbial families"
    )
  ) |>
  dplyr::group_by(
    label, Treatment, Diet, condition, microbial_family_reads, display_family
  ) |>
  dplyr::summarise(estimated_reads = sum(estimated_reads), .groups = "drop") |>
  dplyr::group_by(label) |>
  dplyr::mutate(
    relative_abundance = estimated_reads / sum(estimated_reads)
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    label = factor(label, levels = sample_order_microbes),
    display_family = factor(
      display_family,
      levels = c(top_family_levels, "Other microbial families")
    )
  )

write.csv(
  bracken_family,
  file.path(out_dir, "bracken_host_unmapped_family_abundance_long.csv"),
  row.names = FALSE
)
write.csv(
  bracken_display,
  file.path(out_dir, "bracken_host_unmapped_family_display_table.csv"),
  row.names = FALSE
)
```

```{r bracken-family-composition-figure, fig.width=22, fig.height=18}
family_palette <- setNames(
  grDevices::hcl.colors(length(levels(bracken_display$display_family)), "Dark 3"),
  levels(bracken_display$display_family)
)
family_palette["Other microbial families"] <- "#B8BDC5"
if ("Clavicipitaceae" %in% names(family_palette)) {
  family_palette["Clavicipitaceae"] <- "#39B54A"
}

family_absolute_plot <- ggplot2::ggplot(
  bracken_display,
  ggplot2::aes(x = label, y = estimated_reads, fill = display_family)
) +
  ggplot2::geom_col(width = 0.82) +
  ggplot2::facet_grid(
    cols = ggplot2::vars(condition),
    scales = "free_x",
    space = "free_x"
  ) +
  ggplot2::scale_fill_manual(values = family_palette, drop = FALSE) +
  ggplot2::scale_y_continuous(
    labels = scales::label_comma(),
    expand = ggplot2::expansion(mult = c(0, 0.05))
  ) +
  ggplot2::labs(
    title = "A. Bracken-estimated non-host family reads",
    subtitle = paste(
      "Host-unmapped reads; fungi, bacteria, archaea, viruses, and other",
      "microbial eukaryotes retained"
    ),
    x = NULL,
    y = "Estimated family-assigned reads",
    fill = "Taxonomic family"
  ) +
  ggplot2::theme_bw(base_size = 13) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_blank(),
    strip.text = ggplot2::element_text(face = "bold", size = 12),
    axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, size = 8),
    legend.position = "right"
  )

family_relative_plot <- ggplot2::ggplot(
  bracken_display,
  ggplot2::aes(x = label, y = relative_abundance, fill = display_family)
) +
  ggplot2::geom_col(width = 0.82) +
  ggplot2::facet_grid(
    cols = ggplot2::vars(condition),
    scales = "free_x",
    space = "free_x"
  ) +
  ggplot2::scale_fill_manual(values = family_palette, drop = FALSE) +
  ggplot2::scale_y_continuous(
    labels = scales::label_percent(accuracy = 1),
    expand = c(0, 0)
  ) +
  ggplot2::labs(
    title = "B. Relative non-host family composition",
    subtitle = "Percentage of retained Bracken-estimated non-host family reads",
    x = NULL,
    y = "Relative abundance",
    fill = "Taxonomic family"
  ) +
  ggplot2::theme_bw(base_size = 13) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    strip.text = ggplot2::element_text(face = "bold", size = 12),
    axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, size = 8),
    legend.position = "right"
  )

family_matrix <- xtabs(
  estimated_reads ~ label + family,
  data = bracken_family
)
family_matrix <- family_matrix[metadata$label, , drop = FALSE]
family_relative_matrix <- vegan::decostand(family_matrix, method = "total")
family_bray <- vegan::vegdist(family_relative_matrix, method = "bray")
family_pcoa <- stats::cmdscale(family_bray, eig = TRUE, k = 2)
positive_eigenvalues <- family_pcoa$eig[family_pcoa$eig > 0]
family_pcoa_variance <- 100 * positive_eigenvalues / sum(positive_eigenvalues)

family_pcoa_scores <- data.frame(
  label = rownames(family_pcoa$points),
  PCoA1 = family_pcoa$points[, 1],
  PCoA2 = family_pcoa$points[, 2],
  stringsAsFactors = FALSE
) |>
  dplyr::left_join(
    metadata |>
      dplyr::select(label, Treatment, Diet, condition),
    by = "label"
  )

family_centroids <- family_pcoa_scores |>
  dplyr::group_by(Diet, Treatment) |>
  dplyr::summarise(
    PCoA1 = mean(PCoA1),
    PCoA2 = mean(PCoA2),
    .groups = "drop"
  )

family_pcoa_plot <- ggplot2::ggplot(
  family_pcoa_scores,
  ggplot2::aes(x = PCoA1, y = PCoA2, color = Diet, shape = Treatment)
) +
  ggplot2::geom_point(size = 3.1, alpha = 0.82) +
  ggplot2::geom_line(
    data = family_centroids,
    ggplot2::aes(group = Diet),
    linewidth = 0.9,
    show.legend = FALSE
  ) +
  ggplot2::geom_point(
    data = family_centroids,
    size = 5,
    stroke = 1.2,
    fill = "white",
    show.legend = FALSE
  ) +
  ggplot2::scale_color_manual(
    values = c("33" = "#E6862E", "50" = "#2F8F5B", "83" = "#7B4FA3")
  ) +
  ggplot2::scale_shape_manual(values = c(Control = 16, Infected = 17)) +
  ggplot2::labs(
    title = "C. Bray-Curtis ordination of non-host family profiles",
    subtitle = "Open points are diet-by-treatment centroids; lines connect control and infected centroids",
    x = paste0("PCoA1 (", round(family_pcoa_variance[[1]], 1), "%)"),
    y = paste0("PCoA2 (", round(family_pcoa_variance[[2]], 1), "%)"),
    color = "Diet",
    shape = "Treatment"
  ) +
  ggplot2::theme_classic(base_size = 14) +
  ggplot2::theme(legend.position = "bottom")

bracken_family_composition_figure <-
  family_absolute_plot /
  family_relative_plot /
  family_pcoa_plot +
  patchwork::plot_layout(heights = c(1, 1, 0.8), guides = "collect") &
  ggplot2::theme(legend.position = "bottom")

print(bracken_family_composition_figure)

ggplot2::ggsave(
  file.path(out_dir, "figureS2_bracken_host_unmapped_family_composition.png"),
  bracken_family_composition_figure,
  width = 22, height = 18, dpi = 300, limitsize = FALSE
)
ggplot2::ggsave(
  file.path(out_dir, "figureS2_bracken_host_unmapped_family_composition.pdf"),
  bracken_family_composition_figure,
  width = 22, height = 18, limitsize = FALSE
)
if (requireNamespace("svglite", quietly = TRUE)) {
  ggplot2::ggsave(
    file.path(out_dir, "figureS2_bracken_host_unmapped_family_composition.svg"),
    bracken_family_composition_figure,
    width = 22, height = 18, limitsize = FALSE
  )
}

write.csv(
  family_pcoa_scores,
  file.path(out_dir, "bracken_host_unmapped_family_pcoa_scores.csv"),
  row.names = FALSE
)
```

### Figure S2 large-label narrow version

This alternative export keeps the same data and ordination but uses a taller,
narrower canvas for presentation and manuscript assembly. Sample labels are
shown only for points whose distance from the overall PCoA median exceeds the
median distance plus three median absolute deviations. This reproducible rule
identifies visibly displaced profiles without labelling every library.

```{r bracken-family-composition-figure-large-text, fig.width=16, fig.height=24}
pcoa_median <- c(
  PCoA1 = stats::median(family_pcoa_scores$PCoA1),
  PCoA2 = stats::median(family_pcoa_scores$PCoA2)
)

family_pcoa_scores_labeled <- family_pcoa_scores |>
  dplyr::mutate(
    distance_from_pcoa_median = sqrt(
      (PCoA1 - pcoa_median[["PCoA1"]])^2 +
        (PCoA2 - pcoa_median[["PCoA2"]])^2
    )
  )

pcoa_label_cutoff <- stats::median(
  family_pcoa_scores_labeled$distance_from_pcoa_median
) + 3 * stats::mad(family_pcoa_scores_labeled$distance_from_pcoa_median)

family_pcoa_offset_labels <- family_pcoa_scores_labeled |>
  dplyr::filter(distance_from_pcoa_median > pcoa_label_cutoff)

large_text_family_palette <- family_palette
large_text_family_colors <- c(
  "Plasmodiidae" = "#E8899B",
  "Vibrionaceae" = "#E89A62",
  "Streptomycetaceae" = "#D8B23F",
  "Clavicipitaceae" = "#5DBB72",
  "Cryptococcaceae" = "#91B94A",
  "Enterobacteriaceae" = "#1EA66F",
  "Dictyosteliaceae" = "#35B7A6",
  "Sarcocystidae" = "#42AFC2",
  "Baculoviridae" = "#4C9ED9",
  "Lysobacteraceae" = "#6D91C7",
  "Bacillaceae" = "#9A82D4",
  "Saccharomycetaceae" = "#C77BCB",
  "Other microbial families" = "#BFC5CC"
)
large_text_family_palette[
  intersect(names(large_text_family_palette), names(large_text_family_colors))
] <- large_text_family_colors[
  intersect(names(large_text_family_palette), names(large_text_family_colors))
]

large_text_theme <- ggplot2::theme(
  plot.title = ggplot2::element_text(size = 22, face = "bold"),
  axis.title = ggplot2::element_text(size = 25, face = "bold"),
  axis.text.y = ggplot2::element_text(size = 15),
  axis.text.x = ggplot2::element_text(size = 12),
  strip.text = ggplot2::element_text(size = 17, face = "bold"),
  legend.title = ggplot2::element_text(size = 18, face = "bold"),
  legend.text = ggplot2::element_text(size = 16),
  legend.key.height = grid::unit(0.65, "cm"),
  legend.key.width = grid::unit(0.65, "cm")
)

family_absolute_plot_large <- family_absolute_plot +
  ggplot2::labs(subtitle = NULL) +
  ggplot2::scale_fill_manual(values = large_text_family_palette, drop = FALSE) +
  large_text_theme

family_relative_plot_large <- family_relative_plot +
  ggplot2::labs(subtitle = NULL) +
  ggplot2::scale_fill_manual(values = large_text_family_palette, drop = FALSE) +
  large_text_theme

family_pcoa_plot_large <- ggplot2::ggplot(
  family_pcoa_scores_labeled,
  ggplot2::aes(x = PCoA1, y = PCoA2, color = Diet, shape = Treatment)
) +
  ggplot2::geom_point(size = 5.2, alpha = 0.9, stroke = 1.1) +
  ggplot2::geom_line(
    data = family_centroids,
    ggplot2::aes(group = Diet),
    linewidth = 1.35,
    show.legend = FALSE
  ) +
  ggplot2::geom_point(
    data = family_centroids,
    size = 7.2,
    stroke = 1.6,
    fill = "white",
    show.legend = FALSE
  ) +
  ggrepel::geom_label_repel(
    data = family_pcoa_offset_labels,
    ggplot2::aes(label = label),
    size = 5.3,
    fontface = "bold",
    color = "#222222",
    fill = grDevices::adjustcolor("white", alpha.f = 0.92),
    box.padding = 0.55,
    point.padding = 0.35,
    min.segment.length = 0,
    segment.color = "#555555",
    seed = 83,
    show.legend = FALSE
  ) +
  ggplot2::scale_color_manual(
    values = c("33" = "#E28A3B", "50" = "#369D67", "83" = "#8060AE")
  ) +
  ggplot2::scale_shape_manual(values = c(Control = 16, Infected = 17)) +
  ggplot2::labs(
    title = "C. Bray-Curtis ordination of non-host family profiles",
    subtitle = NULL,
    x = paste0("PCoA1 (", round(family_pcoa_variance[[1]], 1), "%)"),
    y = paste0("PCoA2 (", round(family_pcoa_variance[[2]], 1), "%)"),
    color = "Diet",
    shape = "Treatment"
  ) +
  ggplot2::theme_classic(base_size = 18) +
  large_text_theme +
  ggplot2::theme(legend.position = "bottom")

bracken_family_composition_figure_large_text <-
  family_absolute_plot_large /
  family_relative_plot_large /
  family_pcoa_plot_large +
  patchwork::plot_layout(heights = c(1.05, 1.05, 1.15), guides = "collect") &
  ggplot2::theme(
    legend.position = "bottom",
    legend.box = "vertical",
    legend.box.just = "center",
    legend.title = ggplot2::element_text(size = 18, face = "bold"),
    legend.text = ggplot2::element_text(size = 16),
    legend.key.height = grid::unit(0.65, "cm"),
    legend.key.width = grid::unit(0.65, "cm")
  )

print(bracken_family_composition_figure_large_text)

large_text_stem <- file.path(
  out_dir,
  "figureS2_bracken_host_unmapped_family_composition_large_text_narrow"
)
ggplot2::ggsave(
  paste0(large_text_stem, ".png"),
  bracken_family_composition_figure_large_text,
  width = 16, height = 24, dpi = 300, limitsize = FALSE
)
ggplot2::ggsave(
  paste0(large_text_stem, ".pdf"),
  bracken_family_composition_figure_large_text,
  width = 16, height = 24, limitsize = FALSE
)
if (requireNamespace("svglite", quietly = TRUE)) {
  ggplot2::ggsave(
    paste0(large_text_stem, ".svg"),
    bracken_family_composition_figure_large_text,
    width = 16, height = 24, limitsize = FALSE
  )
}

write.csv(
  family_pcoa_offset_labels,
  file.path(out_dir, "bracken_host_unmapped_family_pcoa_offset_labels.csv"),
  row.names = FALSE
)
```

### Composition tests

PERMANOVA tests whether the multivariate family composition differs with diet,
infection treatment, or their interaction. The dispersion test is reported
alongside it because PERMANOVA can be influenced by unequal within-group
variability. These tests describe the taxonomic signal in the RNA-seq reads;
they are not a dedicated microbiome survey and do not establish organismal
colonization.

```{r bracken-family-composition-tests}
family_test_metadata <- metadata |>
  dplyr::filter(label %in% rownames(family_relative_matrix)) |>
  dplyr::arrange(match(label, rownames(family_relative_matrix))) |>
  dplyr::mutate(
    Diet = factor(Diet, levels = c("33", "50", "83")),
    Treatment = factor(Treatment, levels = c("Control", "Infected")),
    condition = interaction(Diet, Treatment, drop = TRUE)
  )

set.seed(20260810)
family_permanova <- vegan::adonis2(
  family_bray ~ Diet * Treatment,
  data = family_test_metadata,
  permutations = 999,
  by = "terms"
)
family_permanova_table <- as.data.frame(family_permanova) |>
  tibble::rownames_to_column("term") |>
  dplyr::rename(
    df = Df,
    sum_of_squares = SumOfSqs,
    R2 = R2,
    F = F,
    p_value = `Pr(>F)`
  )

family_dispersion <- vegan::betadisper(
  family_bray,
  group = family_test_metadata$condition
)
set.seed(20260810)
family_dispersion_test <- vegan::permutest(family_dispersion, permutations = 999)
family_dispersion_table <- data.frame(
  test = "PERMDISP among diet-by-treatment groups",
  F = unname(family_dispersion_test$tab[1, "F"]),
  p_value = unname(family_dispersion_test$tab[1, "Pr(>F)"])
)

burden_interaction_model <- stats::lm(
  log10_microbial_rpm ~ Diet * Treatment,
  data = microbial_burden
)
burden_interaction_drop <- stats::drop1(
  burden_interaction_model,
  test = "F"
)
burden_additive_model <- stats::lm(
  log10_microbial_rpm ~ Diet + Treatment,
  data = microbial_burden
)
burden_additive_drop <- stats::drop1(burden_additive_model, test = "F")

burden_design_tests <- data.frame(
  term = c("Diet x Infection", "Diet", "Infection"),
  model = c(
    "Full factorial model",
    "Additive model after unsupported interaction",
    "Additive model after unsupported interaction"
  ),
  F = c(
    burden_interaction_drop["Diet:Treatment", "F value"],
    burden_additive_drop["Diet", "F value"],
    burden_additive_drop["Treatment", "F value"]
  ),
  p_value = c(
    burden_interaction_drop["Diet:Treatment", "Pr(>F)"],
    burden_additive_drop["Diet", "Pr(>F)"],
    burden_additive_drop["Treatment", "Pr(>F)"]
  )
)

burden_mass_interaction_model <- stats::lm(
  log10_microbial_rpm ~ mass_g * Treatment + Diet,
  data = microbial_burden
)
burden_mass_interaction_drop <- stats::drop1(
  burden_mass_interaction_model,
  test = "F"
)
burden_mass_additive_model <- stats::lm(
  log10_microbial_rpm ~ mass_g + Treatment + Diet,
  data = microbial_burden
)
burden_mass_additive_drop <- stats::drop1(
  burden_mass_additive_model,
  test = "F"
)
burden_mass_tests <- data.frame(
  term = c("Mass x Infection", "Mass"),
  model = c(
    "Mass-by-infection model adjusted for diet",
    "Additive model after unsupported mass-by-infection interaction"
  ),
  F = c(
    burden_mass_interaction_drop["mass_g:Treatment", "F value"],
    burden_mass_additive_drop["mass_g", "F value"]
  ),
  p_value = c(
    burden_mass_interaction_drop["mass_g:Treatment", "Pr(>F)"],
    burden_mass_additive_drop["mass_g", "Pr(>F)"]
  )
)

burden_group_summary <- microbial_burden |>
  dplyr::group_by(Treatment) |>
  dplyr::summarise(
    samples = dplyr::n(),
    median_microbial_reads_per_million =
      stats::median(microbial_reads_per_million),
    mean_microbial_reads_per_million = mean(microbial_reads_per_million),
    IQR_microbial_reads_per_million =
      stats::IQR(microbial_reads_per_million),
    .groups = "drop"
  )

write.csv(
  family_permanova_table,
  file.path(out_dir, "bracken_host_unmapped_family_permanova.csv"),
  row.names = FALSE
)
write.csv(
  family_dispersion_table,
  file.path(out_dir, "bracken_host_unmapped_family_dispersion_test.csv"),
  row.names = FALSE
)
write.csv(
  microbial_burden,
  file.path(out_dir, "bracken_host_unmapped_burden_by_sample.csv"),
  row.names = FALSE
)
write.csv(
  burden_design_tests,
  file.path(out_dir, "bracken_host_unmapped_burden_design_tests.csv"),
  row.names = FALSE
)
write.csv(
  burden_mass_tests,
  file.path(out_dir, "bracken_host_unmapped_burden_mass_tests.csv"),
  row.names = FALSE
)
write.csv(
  burden_group_summary,
  file.path(out_dir, "bracken_host_unmapped_burden_group_summary.csv"),
  row.names = FALSE
)

knitr::kable(
  family_permanova_table |>
    dplyr::mutate(
      dplyr::across(c(sum_of_squares, R2, F, p_value), ~ round(.x, 4))
    ),
  caption = "PERMANOVA of host-unmapped Bracken family composition"
)
knitr::kable(
  family_dispersion_table |>
    dplyr::mutate(dplyr::across(c(F, p_value), ~ round(.x, 4))),
  caption = "Dispersion check for the family-composition PERMANOVA"
)
knitr::kable(
  burden_design_tests,
  digits = 4,
  caption = paste(
    "Diet and infection tests for Bracken-estimated microbial family reads",
    "per million input reads"
  )
)
knitr::kable(
  burden_mass_tests,
  digits = 4,
  caption = "Body-mass tests for normalized non-host taxonomic burden"
)
```

### Results and interpretation

The same broad family-level assignments recur across most libraries, although
the absolute number of host-unmapped reads assigned by Bracken varies markedly
among individuals. Diet did not explain a supported fraction of the relative
family composition (`r sprintf("R2 = %.3f, P = %.3f", family_permanova_table$R2[family_permanova_table$term == "Diet"], family_permanova_table$p_value[family_permanova_table$term == "Diet"])`).
In contrast, control and infected libraries differed in their multivariate
taxonomic profiles (`r sprintf("R2 = %.3f, P = %.3f", family_permanova_table$R2[family_permanova_table$term == "Treatment"], family_permanova_table$p_value[family_permanova_table$term == "Treatment"])`),
while the diet-by-infection interaction was not supported
(`r sprintf("P = %.3f", family_permanova_table$p_value[family_permanova_table$term == "Diet:Treatment"])`).
The dispersion test was not significant
(`r sprintf("P = %.3f", family_dispersion_table$p_value)`), indicating that the
treatment result was not explained simply by unequal within-group spread.

The normalized abundance of retained microbial-family assignments was also
higher in infected libraries (median
`r scales::comma(round(burden_group_summary$median_microbial_reads_per_million[burden_group_summary$Treatment == "Infected"]))`
reads per million input reads) than controls (median
`r scales::comma(round(burden_group_summary$median_microbial_reads_per_million[burden_group_summary$Treatment == "Control"]))`).
The diet-by-infection interaction was unsupported
(`r sprintf("P = %.3f", burden_design_tests$p_value[burden_design_tests$term == "Diet x Infection"])`),
whereas the infection effect was supported after accounting for diet
(`r sprintf("P = %.4f", burden_design_tests$p_value[burden_design_tests$term == "Infection"])`).
Recorded mass did not explain normalized non-host burden in the additive model
(`r sprintf("P = %.3f", burden_mass_tests$p_value[burden_mass_tests$term == "Mass"])`),
and the mass-by-infection interaction remained just above the prespecified
threshold (`r sprintf("P = %.3f", burden_mass_tests$p_value[burden_mass_tests$term == "Mass x Infection"])`).

Thus, the profiles do not support a broad diet-associated restructuring of the
taxonomic signal, but neither should they be interpreted as completely invariant
among treatments. Because these assignments come from total-RNA libraries rather
than a dedicated microbiome assay, they are best treated as supporting
metatranscriptomic evidence. Individual family calls can reflect the challenge
fungus, other microorganisms, residual host sequence, or database ambiguity and
do not by themselves demonstrate viable colonization.

## Combined export

```{r combined-export, fig.width=20, fig.height=16}
combined_plot <- fungal_genome_plot / fungal_percentage_plot +
  patchwork::plot_annotation(
    title = "RNA-seq evidence for Metarhizium robertsii across fat body samples",
    tag_levels = "A"
  )

print(combined_plot)

export_specs <- list(
  "metarhizium_genome_mapped_reads" = list(
    plot = fungal_genome_plot, width = 20, height = 8
  ),
  "metarhizium_genome_mapped_reads_zoom_7000_all_samples" = list(
    plot = fungal_genome_zoom_plot, width = 20, height = 8
  ),
  "metarhizium_genome_mapped_reads_full_and_zoom" = list(
    plot = fungal_genome_full_zoom_plot, width = 20, height = 16
  ),
  "metarhizium_unique_alignment_percentage" = list(
    plot = fungal_percentage_plot, width = 20, height = 8
  ),
  "metarhizium_unique_alignment_percentage_zoom_0_1_all_samples" = list(
    plot = fungal_percentage_zoom_plot, width = 20, height = 8
  ),
  "metarhizium_unique_alignment_percentage_full_and_zoom" = list(
    plot = fungal_percentage_full_zoom_plot, width = 20, height = 16
  ),
  "metarhizium_percentage_above_main_chromosome_deg_heatmap" = list(
    plot = fungal_percentage_deg_heatmap, width = 22, height = 16
  ),
  "metarhizium_alignment_burden_combined" = list(
    plot = combined_plot, width = 20, height = 16
  )
)

for (export_name in names(export_specs)) {
  spec <- export_specs[[export_name]]
  ggplot2::ggsave(
    file.path(out_dir, paste0(export_name, ".png")),
    spec$plot,
    width = spec$width,
    height = spec$height,
    dpi = 300,
    limitsize = FALSE
  )
  ggplot2::ggsave(
    file.path(out_dir, paste0(export_name, ".pdf")),
    spec$plot,
    width = spec$width,
    height = spec$height,
    limitsize = FALSE
  )
  if (requireNamespace("svglite", quietly = TRUE)) {
    ggplot2::ggsave(
      file.path(out_dir, paste0(export_name, ".svg")),
      spec$plot,
      width = spec$width,
      height = spec$height,
      limitsize = FALSE
    )
  }
}
```

## Sample-level values

```{r sample-table}
DT::datatable(
  fungal_burden |>
    dplyr::arrange(dplyr::desc(unique_fungus)) |>
    dplyr::select(
      label, Treatment, Diet, unique_fungus, pct_unique_fungus
    ),
  rownames = FALSE,
  filter = "top",
  options = list(pageLength = 15, scrollX = TRUE),
  caption = "Competitive-mapping evidence for M. robertsii by sample"
)
```

## Interpretation

The fungal signal is strongly heterogeneous among infected individuals rather
than uniformly high after challenge. Samples 1029, 1058, 1025, and 1057 have
the highest whole-genome fungal alignment counts and fungal alignment
percentages. Controls and most infected samples have low counts. Low non-zero
values should be treated as alignment background or weak fungal evidence,
whereas agreement between absolute and library-normalized alignment signal
provides stronger RNA-seq evidence of fungal burden.
