---
title: "Curated target genes by sequence placement"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  workbook_file: "data/external/curated_target_genes/2026-07-26_curated-hypoth_DEGs_v14.xlsx"
  placement_file: "output/runs/host_pathogen_dual_20260727_000952/local-analysis/host-mapping-audit_20260729-163928/unified-model/audit-summary/host_gene_sequence_placement.tsv"
  latest_results_root: "output/rmd_runs"
  results_pattern: "^placed-unplaced-scaffold-analysis_"
  homology_results_pattern: "^scaffold-origin-audit_"
  alpha: 0.05
  lfc_threshold: 1
  output_tag: "curated-target-gene-figure"
---

## Purpose

This page follows a fixed set of immune and protein-anabolism genes through two
matched views: genes on the main *Schistocerca gregaria* chromosomes and the
complete all-scaffold set used in the original curated figure. The biological
categories come from the supplied workbook, while every differential-expression
status is recalculated from the current competitive-host DESeq2 results.

::: {.walkthrough-grid}
::: {.walkthrough-step}
**Step 1 - Match genes**

Join the curated IDs to the GFF annotation and official assembly placement.
:::
::: {.walkthrough-step}
**Step 2 - Recalculate status**

Apply the same DEG thresholds to every target and contrast.
:::
::: {.walkthrough-step}
**Step 3 - Compare scopes**

Read the main-chromosome and complete all-scaffold panels side by side.
:::
:::

A target is shown as `N.S.` when it does not meet both `padj < alpha` and
`|log2FC| >= lfc_threshold`. Keeping the target set fixed makes changes in the
category percentages directly comparable across contrasts and placement
classes.

The label below each facet reports the number of **curated target genes in the
denominator**, not the number of DEGs. Differential-expression status is
recalculated from the current transcript-plus-exon model, so a target can move
into or out of significance even though restricting the target list to main
chromosomes can only reduce its size.

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required_packages <- c(
  "readxl", "readr", "dplyr", "tidyr", "tibble", "ggplot2",
  "patchwork", "cowplot", "scales", "DT", "knitr"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages) > 0) {
  stop(
    "Install the following R packages before running this page: ",
    paste(missing_packages, collapse = ", ")
  )
}

suppressPackageStartupMessages({
  library(readxl)
  library(readr)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(patchwork)
})

page_defaults <- list(
  workbook_file = paste0(
    "data/external/curated_target_genes/",
    "2026-07-26_curated-hypoth_DEGs_v14.xlsx"
  ),
  placement_file = paste0(
    "output/runs/host_pathogen_dual_20260727_000952/local-analysis/",
    "host-mapping-audit_20260729-163928/unified-model/audit-summary/",
    "host_gene_sequence_placement.tsv"
  ),
  latest_results_root = "output/rmd_runs",
  results_pattern = "^placed-unplaced-scaffold-analysis_",
  homology_results_pattern = "^scaffold-origin-audit_",
  alpha = 0.05,
  lfc_threshold = 1,
  output_tag = "curated-target-gene-figure"
)

if (!exists("params", inherits = TRUE) || !is.list(params)) {
  page_params <- page_defaults
} else {
  page_params <- utils::modifyList(page_defaults, params)
}

path_ancestors <- function(path) {
  path <- normalizePath(path, mustWork = FALSE)
  ancestors <- path
  repeat {
    parent <- dirname(path)
    if (identical(parent, path)) break
    ancestors <- c(ancestors, parent)
    path <- parent
  }
  ancestors
}

is_project_root <- function(path) {
  file.exists(file.path(path, "analysis", "30-curated-target-gene-figure.Rmd")) &&
    dir.exists(file.path(path, "data")) &&
    dir.exists(file.path(path, "output"))
}

find_project_root <- function() {
  starts <- c(getwd())
  explicit_root <- Sys.getenv("GREGARIA_DIET_PROJECT_DIR", unset = "")
  if (nzchar(explicit_root)) starts <- c(explicit_root, starts)

  input_file <- tryCatch(
    knitr::current_input(dir = TRUE),
    error = function(e) ""
  )
  if (nzchar(input_file)) starts <- c(dirname(input_file), starts)

  candidates <- unique(unlist(lapply(starts, path_ancestors)))
  hits <- candidates[vapply(candidates, is_project_root, logical(1))]
  if (length(hits) == 0) {
    stop("Could not locate the gregaria-diet-infection-interaction project root.")
  }
  hits[[1]]
}

project_dir <- find_project_root()
resolve_path <- function(path) {
  if (grepl("^/", path)) path else file.path(project_dir, path)
}

latest_matching_dir <- function(root, pattern) {
  root <- resolve_path(root)
  candidates <- list.dirs(root, recursive = FALSE, full.names = TRUE)
  candidates <- candidates[grepl(pattern, basename(candidates))]
  if (length(candidates) == 0) {
    stop("No result folder matched ", pattern, " under ", root)
  }
  sort(candidates)[[length(candidates)]]
}

make_run_dir <- function(root, tag) {
  stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
  base <- file.path(resolve_path(root), paste0(tag, "_", stamp))
  candidate <- base
  suffix <- 1L
  while (dir.exists(candidate)) {
    suffix <- suffix + 1L
    candidate <- paste0(base, "_", suffix)
  }
  dir.create(candidate, recursive = TRUE, showWarnings = FALSE)
  candidate
}

workbook_file <- resolve_path(page_params$workbook_file)
placement_file <- resolve_path(page_params$placement_file)
source_results_dir <- latest_matching_dir(
  page_params$latest_results_root,
  page_params$results_pattern
)
homology_results_dir <- latest_matching_dir(
  page_params$latest_results_root,
  page_params$homology_results_pattern
)
results_file <- file.path(
  source_results_dir,
  "placed_unplaced_all_deseq2_results.csv"
)
homology_evidence_file <- file.path(
  homology_results_dir,
  "corrected_gene_homology_evidence.tsv"
)
out_dir <- make_run_dir(page_params$latest_results_root, page_params$output_tag)

required_files <- c(
  workbook_file, placement_file, results_file, homology_evidence_file
)
missing_files <- required_files[!file.exists(required_files)]
if (length(missing_files) > 0) {
  stop("Missing required input files: ", paste(missing_files, collapse = ", "))
}

alpha <- as.numeric(page_params$alpha)
lfc_threshold <- as.numeric(page_params$lfc_threshold)

write_csv_out <- function(x, filename) {
  readr::write_csv(x, file.path(out_dir, filename), na = "")
}

save_figure <- function(plot, stem, width, height) {
  ggplot2::ggsave(
    file.path(out_dir, paste0(stem, ".png")), plot,
    width = width, height = height, units = "in", dpi = 400,
    bg = "white"
  )
  ggplot2::ggsave(
    file.path(out_dir, paste0(stem, ".svg")), plot,
    width = width, height = height, units = "in", bg = "white"
  )
  ggplot2::ggsave(
    file.path(out_dir, paste0(stem, ".pdf")), plot,
    width = width, height = height, units = "in",
    device = grDevices::cairo_pdf, bg = "white"
  )
}
```

## Inputs and contrast direction

The workbook contains 163 unique target genes divided between immune and
protein-anabolism categories. Its values came from the earlier all-sequence DEG
catalogue. Here, the workbook supplies only gene membership and biological
categories; the current log2 fold changes and adjusted p-values come from the
latest placed/unplaced sensitivity run.

Within each diet, positive values mean higher expression in infected locusts.
For across-diet comparisons, the sign is oriented to the higher-protein diet:
50 versus 33, 83 versus 33, and 83 versus 50. Thus, `Upregulated` always means
higher expression in the first condition printed below the bar.

```{r read-inputs}
curated_targets <- dplyr::bind_rows(
  readxl::read_excel(workbook_file, sheet = "Immune") |>
    dplyr::mutate(source_sheet = "Immune"),
  readxl::read_excel(workbook_file, sheet = "Protein-Anabolism") |>
    dplyr::mutate(source_sheet = "Protein-Anabolism")
) |>
  dplyr::rename(
    umbrella = `Umbrella Category`,
    subcategory = `Sub-Category`,
    gene_id = `Gene ID`,
    description_workbook = `Gene Description`
  )

if (anyDuplicated(curated_targets$gene_id)) {
  stop("The curated workbook contains duplicated gene IDs.")
}

placement <- readr::read_tsv(placement_file, show_col_types = FALSE)
all_results <- readr::read_csv(results_file, show_col_types = FALSE) |>
  dplyr::filter(.data$mapping_branch == "Competitive host")
homology_evidence <- readr::read_tsv(
  homology_evidence_file,
  show_col_types = FALSE
)

if (anyDuplicated(homology_evidence$gene_id)) {
  stop("The scaffold-origin homology audit contains duplicated gene IDs.")
}

contrast_map <- tibble::tribble(
  ~contrast_id, ~contrast_label, ~figure_scope, ~x_label, ~x_order,
  ~desired_sign, ~positive_condition, ~negative_condition,
  ~workbook_lfc_col, ~workbook_padj_col, ~show_in_figure,
  "infection_diet33", "Diet 33: infected vs control",
  "Within-diet infection response", "33", 1, 1,
  "Infected diet 33", "Control diet 33",
  "Diet 33 log2FC", "Diet 33 padj", TRUE,
  "infection_diet50", "Diet 50: infected vs control",
  "Within-diet infection response", "50", 2, 1,
  "Infected diet 50", "Control diet 50",
  "Diet 50 log2FC", "Diet 50 padj", TRUE,
  "infection_diet83", "Diet 83: infected vs control",
  "Within-diet infection response", "83", 3, 1,
  "Infected diet 83", "Control diet 83",
  "Diet 83 log2FC", "Diet 83 padj", TRUE,
  "infected_diet33_vs50", "Infected: diet 50 vs 33",
  "Diet comparison among infected", "50\nvs\n33", 1, -1,
  "Infected diet 50", "Infected diet 33",
  "inf33 vs inf50_log2FC", "inf33 vs inf50_padj", TRUE,
  "infected_diet83_vs33", "Infected: diet 83 vs 33",
  "Diet comparison among infected", "83\nvs\n33", 2, 1,
  "Infected diet 83", "Infected diet 33",
  "inf83 vs inf33_log2FC", "inf83 vs inf33_padj", TRUE,
  "infected_diet83_vs50", "Infected: diet 83 vs 50",
  "Diet comparison among infected", "83\nvs\n50", 3, 1,
  "Infected diet 83", "Infected diet 50",
  "inf83 vs inf50_log2FC", "inf83 vs inf50_padj", TRUE,
  "control_diet33_vs50", "Control: diet 50 vs 33",
  "Diet comparison among controls", "50\nvs\n33", 1, -1,
  "Control diet 50", "Control diet 33",
  "cont33 vs cont50_log2FC", "cont33 vs cont50_padj", FALSE,
  "control_diet83_vs33", "Control: diet 83 vs 33",
  "Diet comparison among controls", "83\nvs\n33", 2, 1,
  "Control diet 83", "Control diet 33",
  "cont83 vs cont33_log2FC", "cont83 vs cont33_padj", FALSE,
  "control_diet83_vs50", "Control: diet 83 vs 50",
  "Diet comparison among controls", "83\nvs\n50", 3, 1,
  "Control diet 83", "Control diet 50",
  "cont83 vs cont50_log2FC", "cont83 vs cont50_padj", FALSE
)

target_placement <- curated_targets |>
  dplyr::left_join(
    placement |>
      dplyr::select(
        gene_id, seqid, description_gff = description, gene_biotype,
        sequence_name, sequence_role, assigned_molecule, location_type,
        sequence_class
      ),
    by = "gene_id"
  )

if (any(is.na(target_placement$sequence_class))) {
  stop(
    "Placement was not found for: ",
    paste(target_placement$gene_id[is.na(target_placement$sequence_class)],
          collapse = ", ")
  )
}

current_target_results <- all_results |>
  dplyr::filter(.data$contrast_id %in% contrast_map$contrast_id) |>
  dplyr::select(
    gene_id, analysis_scope, contrast_id, baseMean, log2FoldChange,
    lfcSE, stat, pvalue, padj
  )

duplicate_result_rows <- current_target_results |>
  dplyr::count(.data$gene_id, .data$contrast_id) |>
  dplyr::filter(.data$n > 1)
if (nrow(duplicate_result_rows) > 0) {
  stop("More than one competitive-host result was found per gene and contrast.")
}

target_analysis <- target_placement |>
  dplyr::select(
    source_sheet, umbrella, subcategory, gene_id, description_workbook,
    description_gff, gene_biotype, seqid, sequence_name, sequence_role,
    assigned_molecule, location_type, sequence_class
  ) |>
  tidyr::crossing(contrast_map) |>
  dplyr::left_join(
    current_target_results,
    by = c("gene_id", "contrast_id")
  ) |>
  dplyr::mutate(
    significant = !is.na(.data$padj) &
      .data$padj < alpha &
      abs(.data$log2FoldChange) >= lfc_threshold,
    oriented_log2FC = .data$log2FoldChange * .data$desired_sign,
    status = dplyr::case_when(
      !.data$significant ~ "N.S.",
      .data$oriented_log2FC > 0 ~ "Upregulated",
      TRUE ~ "Downregulated"
    ),
    higher_expression_condition = dplyr::case_when(
      !.data$significant ~ "Not significant",
      .data$oriented_log2FC > 0 ~ .data$positive_condition,
      TRUE ~ .data$negative_condition
    ),
    primary_figure = .data$sequence_class == "Nuclear chromosome"
  )

original_status_long <- dplyr::bind_rows(lapply(
  seq_len(nrow(contrast_map)),
  function(i) {
    original_value <- as.character(
      curated_targets[[contrast_map$workbook_lfc_col[[i]]]]
    )
    numeric_value <- suppressWarnings(as.numeric(original_value))
    tibble::tibble(
      gene_id = curated_targets$gene_id,
      contrast_id = contrast_map$contrast_id[[i]],
      original_value = original_value,
      original_oriented_log2FC = numeric_value * contrast_map$desired_sign[[i]],
      original_status = dplyr::case_when(
        is.na(numeric_value) ~ "N.S.",
        numeric_value * contrast_map$desired_sign[[i]] > 0 ~ "Upregulated",
        numeric_value * contrast_map$desired_sign[[i]] < 0 ~ "Downregulated",
        TRUE ~ "N.S."
      )
    )
  }
))

target_analysis <- target_analysis |>
  dplyr::left_join(
    original_status_long,
    by = c("gene_id", "contrast_id")
  )

if (nrow(target_analysis) != dplyr::n_distinct(target_analysis$gene_id) *
    dplyr::n_distinct(target_analysis$contrast_id)) {
  stop("The curated audit is not one row per gene and contrast.")
}

gene_audit <- target_analysis |>
  dplyr::group_by(
    .data$source_sheet, .data$umbrella, .data$subcategory,
    .data$gene_id, .data$description_workbook, .data$description_gff,
    .data$gene_biotype, .data$seqid, .data$sequence_name,
    .data$sequence_role, .data$sequence_class
  ) |>
  dplyr::summarise(
    original_significant_contrasts = sum(.data$original_status != "N.S."),
    current_significant_contrasts = sum(.data$status != "N.S."),
    newly_significant_contrasts = sum(
      .data$original_status == "N.S." & .data$status != "N.S."
    ),
    no_longer_significant_contrasts = sum(
      .data$original_status != "N.S." & .data$status == "N.S."
    ),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    figure_scope = dplyr::if_else(
      .data$sequence_class == "Nuclear chromosome",
      "Main chromosome",
      "Unplaced scaffold"
    )
  )

summarize_reconciliation <- function(data, sequence_scope) {
  data |>
    dplyr::filter(.data$show_in_figure) |>
    dplyr::group_by(
      .data$source_sheet, .data$umbrella, .data$figure_scope
    ) |>
    dplyr::summarise(
      sequence_scope = sequence_scope,
      target_genes = dplyr::n_distinct(.data$gene_id),
      gene_contrast_cells = dplyr::n(),
      original_significant_cells = sum(.data$original_status != "N.S."),
      current_significant_cells = sum(.data$status != "N.S."),
      newly_significant_cells = sum(
        .data$original_status == "N.S." & .data$status != "N.S."
      ),
      no_longer_significant_cells = sum(
        .data$original_status != "N.S." & .data$status == "N.S."
      ),
      net_change = .data$current_significant_cells -
        .data$original_significant_cells,
      .groups = "drop"
    ) |>
    dplyr::relocate(.data$sequence_scope)
}

count_reconciliation <- dplyr::bind_rows(
  summarize_reconciliation(target_analysis, "All curated targets"),
  summarize_reconciliation(
    dplyr::filter(target_analysis, .data$primary_figure),
    "Main chromosome"
  ),
  summarize_reconciliation(
    dplyr::filter(target_analysis, !.data$primary_figure),
    "Unplaced scaffold"
  )
)

contrast_audit <- target_analysis |>
  dplyr::select(
    .data$source_sheet, .data$umbrella, .data$subcategory,
    .data$gene_id, .data$description_workbook, .data$description_gff,
    .data$gene_biotype, .data$seqid, .data$sequence_class,
    .data$contrast_id, .data$contrast_label, .data$figure_scope,
    .data$positive_condition, .data$negative_condition,
    .data$baseMean, .data$log2FoldChange, .data$oriented_log2FC,
    .data$padj, .data$status, .data$higher_expression_condition,
    .data$original_status, .data$original_value,
    .data$show_in_figure
  )

write_csv_out(gene_audit, "figure4_curated_gene_audit.csv")
write_csv_out(count_reconciliation, "figure4_count_reconciliation.csv")
write_csv_out(contrast_audit, "figure4_contrast_audit.csv")
write_csv_out(
  dplyr::filter(contrast_audit, .data$sequence_class == "Nuclear chromosome"),
  "figure4_main_chromosome_contrast_audit.csv"
)
write_csv_out(
  dplyr::filter(contrast_audit, .data$sequence_class == "Unplaced scaffold"),
  "figure4_unplaced_scaffold_contrast_audit.csv"
)
```

## Effect of the chromosome restriction

```{r placement-summary}
target_gene_summary <- target_analysis |>
  dplyr::group_by(
    .data$source_sheet, .data$umbrella, .data$gene_id,
    .data$description_workbook, .data$description_gff,
    .data$gene_biotype, .data$seqid, .data$sequence_class
  ) |>
  dplyr::summarise(
    currently_significant = any(.data$significant),
    n_current_significant_contrasts = sum(.data$significant),
    .groups = "drop"
  )

category_impact <- target_gene_summary |>
  dplyr::group_by(.data$source_sheet, .data$umbrella) |>
  dplyr::summarise(
    original_targets = dplyr::n(),
    main_chromosome_targets = sum(.data$sequence_class == "Nuclear chromosome"),
    unplaced_targets = sum(.data$sequence_class == "Unplaced scaffold"),
    retained_percent = 100 * .data$main_chromosome_targets / .data$original_targets,
    main_chromosome_targets_significant_now = sum(
      .data$sequence_class == "Nuclear chromosome" &
        .data$currently_significant
    ),
    .groups = "drop"
  )

total_targets <- nrow(target_gene_summary)
main_targets <- sum(target_gene_summary$sequence_class == "Nuclear chromosome")
unplaced_targets_n <- sum(target_gene_summary$sequence_class == "Unplaced scaffold")
main_targets_significant <- sum(
  target_gene_summary$sequence_class == "Nuclear chromosome" &
    target_gene_summary$currently_significant
)

protein_row <- category_impact |>
  dplyr::filter(.data$umbrella == "Protein Synthesis & Translation")

write_csv_out(category_impact, "curated_target_placement_summary.csv")
write_csv_out(target_analysis, "curated_target_gene_reanalysis_long.csv")
```

Of the `r total_targets` curated genes, **`r main_targets` are on main nuclear
chromosomes** and enter the revised primary figure. The remaining
**`r unplaced_targets_n` genes are on unplaced scaffolds**. Among the retained
targets, `r main_targets_significant` are currently significant in at least one
of the nine reconstructed contrasts.

There is no double counting: the workbook contains `r total_targets` unique gene
IDs, and the audit contains exactly one row per gene and contrast. For the six
contrasts displayed in Figure 4, the main-chromosome subset contains exactly
`r main_targets * 6` gene-by-contrast cells.

The largest change is Protein Synthesis & Translation: only
`r protein_row$main_chromosome_targets` of its
`r protein_row$original_targets` targets are retained, while
`r protein_row$unplaced_targets` are moved to the audit set. This is why the
across-diet protein-synthesis pattern in the older figure cannot be carried
forward as a main-chromosome result.

```{r target-placement-table}
knitr::kable(
  category_impact |>
    dplyr::mutate(
      retained_percent = scales::percent(.data$retained_percent / 100,
                                         accuracy = 0.1)
    ),
  col.names = c(
    "Group", "Curated category", "Original targets", "Main chromosome",
    "Unplaced", "Retained", "Main-chromosome targets significant now"
  ),
  align = c("l", "l", "r", "r", "r", "r", "r")
)
```

## DEG count reconciliation

This table separates three quantities that can otherwise be confused. Target
genes are the curated denominator; significant cells count one gene in one
contrast; newly significant and no-longer-significant cells describe changes
caused by recalculating the model. A strict placement filter reduces the target
list from 163 to 99 genes. The recalculated main-chromosome Figure 4 has fewer
significant gene-by-contrast cells overall than the earlier workbook, although
some individual categories gain one or two calls while others lose calls.

[Download the gene-level Figure 4 audit workbook](assets/downloads/figure4_curated_DEG_audit.xlsx).

```{r count-reconciliation-table}
knitr::kable(
  count_reconciliation |>
    dplyr::arrange(
      factor(
        .data$sequence_scope,
        levels = c(
          "All curated targets", "Main chromosome", "Unplaced scaffold"
        )
      ),
      .data$source_sheet, .data$umbrella, .data$figure_scope
    ),
  col.names = c(
    "Sequence scope", "Group", "Category", "Comparison group",
    "Target genes", "Gene-contrast cells", "Earlier significant",
    "Current significant", "Newly significant", "No longer significant",
    "Net change"
  ),
  align = c("l", "l", "l", "l", rep("r", 7))
)
```

```{r target-placement-plot, fig.width=12, fig.height=7.5}
category_order <- c(
  "Innate Immune Recognition & Signaling",
  "Humoral Immunity",
  "Cellular Immunity",
  "Detoxification",
  "Protein Synthesis & Translation",
  "Nutrient Sensing & TOR/Insulin Signaling",
  "Nitrogen & Amino Acid Metabolism",
  "Storage Hexamerins & Reserves"
)

category_labels <- c(
  "Innate Immune Recognition & Signaling" =
    "Innate immune recognition & signaling",
  "Humoral Immunity" = "Humoral immunity",
  "Cellular Immunity" = "Cellular immunity",
  "Detoxification" = "Detoxification",
  "Protein Synthesis & Translation" = "Protein synthesis & translation",
  "Nutrient Sensing & TOR/Insulin Signaling" =
    "Nutrient sensing & TOR/insulin signaling",
  "Nitrogen & Amino Acid Metabolism" =
    "Nitrogen & amino acid metabolism",
  "Storage Hexamerins & Reserves" = "Storage hexamerins & reserves"
)

placement_plot_data <- target_gene_summary |>
  dplyr::mutate(
    placement_for_figure = dplyr::if_else(
      .data$sequence_class == "Nuclear chromosome",
      "Main chromosome - retained",
      "Unplaced scaffold - audit only"
    )
  ) |>
  dplyr::count(
    .data$source_sheet, .data$umbrella, .data$placement_for_figure,
    name = "n_genes"
  ) |>
  dplyr::group_by(.data$source_sheet, .data$umbrella) |>
  tidyr::complete(
    placement_for_figure = c(
      "Main chromosome - retained", "Unplaced scaffold - audit only"
    ),
    fill = list(n_genes = 0)
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    umbrella = factor(.data$umbrella, levels = rev(category_order)),
    source_sheet = dplyr::recode(
      .data$source_sheet,
      "Protein-Anabolism" = "Protein anabolism"
    )
  )

placement_impact_plot <- ggplot(
  placement_plot_data,
  aes(x = n_genes, y = umbrella, fill = placement_for_figure)
) +
  geom_col(width = 0.72, color = "white", linewidth = 0.4) +
  geom_text(
    aes(label = ifelse(n_genes > 0, n_genes, "")),
    position = position_stack(vjust = 0.5),
    color = "white", fontface = "bold", size = 4.2
  ) +
  facet_grid(source_sheet ~ ., scales = "free_y", space = "free_y") +
  scale_fill_manual(
    values = c(
      "Main chromosome - retained" = "#2F7D5B",
      "Unplaced scaffold - audit only" = "#B85C4B"
    ),
    breaks = c(
      "Main chromosome - retained", "Unplaced scaffold - audit only"
    )
  ) +
  scale_y_discrete(labels = category_labels) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Curated target genes retained for primary Figure 4",
    subtitle = paste0(
      main_targets, " of ", total_targets,
      " targets are on main chromosomes; unplaced targets remain in the audit"
    ),
    x = "Number of curated target genes",
    y = NULL,
    fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12),
    axis.text.y = element_text(size = 12, color = "black"),
    axis.text.x = element_text(size = 11, color = "black"),
    strip.text.y = element_text(face = "bold", size = 13, angle = 0),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.text = element_text(size = 12),
    plot.margin = margin(12, 18, 12, 12)
  )

placement_impact_plot
save_figure(placement_impact_plot, "curated_target_placement_impact", 12, 7.5)
```

## Curated target figures {#Curated_target_figures}

```{r build-revised-figure4, fig.width=15, fig.height=11.5, results='hide', fig.show='hide'}
figure_contrasts <- contrast_map |>
  dplyr::filter(.data$show_in_figure)

figure_gene_data <- target_analysis |>
  dplyr::filter(
    .data$primary_figure,
    .data$show_in_figure
  )

figure_counts <- figure_gene_data |>
  dplyr::count(
    .data$source_sheet, .data$umbrella, .data$figure_scope,
    .data$x_label, .data$x_order, .data$status,
    name = "n_genes"
  ) |>
  dplyr::group_by(
    .data$source_sheet, .data$umbrella, .data$figure_scope,
    .data$x_label, .data$x_order
  ) |>
  dplyr::mutate(percentage = 100 * .data$n_genes / sum(.data$n_genes)) |>
  dplyr::ungroup()

category_n <- target_gene_summary |>
  dplyr::filter(.data$sequence_class == "Nuclear chromosome") |>
  dplyr::count(.data$source_sheet, .data$umbrella, name = "n_targets")

format_test_label <- function(p) {
  if (is.na(p) || p >= 0.05) return("N.S.D.")
  if (p < 0.0001) return("p < 0.0001")
  if (p < 0.001) return("p < 0.001")
  paste0("p = ", formatC(p, digits = 3, format = "f"))
}

calculate_figure_tests <- function(count_data) {
  set.seed(20260805)
  test_rows <- list()
  row_i <- 0L
  for (sheet_name in unique(count_data$source_sheet)) {
    for (category_name in unique(
      count_data$umbrella[count_data$source_sheet == sheet_name]
    )) {
      for (scope_name in unique(count_data$figure_scope)) {
        test_df <- count_data |>
          dplyr::filter(
            .data$source_sheet == sheet_name,
            .data$umbrella == category_name,
            .data$figure_scope == scope_name
          ) |>
          tidyr::complete(
            x_order = 1:3,
            status = c("Upregulated", "N.S.", "Downregulated"),
            fill = list(n_genes = 0)
          ) |>
          dplyr::arrange(.data$x_order, match(
            .data$status,
            c("Upregulated", "N.S.", "Downregulated")
          ))

        count_matrix <- matrix(test_df$n_genes, nrow = 3, byrow = TRUE)
        count_matrix <- count_matrix[, colSums(count_matrix) > 0, drop = FALSE]
        p_value <- NA_real_
        if (ncol(count_matrix) >= 2 && nrow(count_matrix) >= 2) {
          p_value <- suppressWarnings(stats::chisq.test(
            count_matrix,
            simulate.p.value = any(count_matrix < 5),
            B = 20000
          )$p.value)
        }

        row_i <- row_i + 1L
        test_rows[[row_i]] <- tibble::tibble(
          source_sheet = sheet_name,
          umbrella = category_name,
          figure_scope = scope_name,
          p_value = p_value,
          p_label = format_test_label(p_value)
        )
      }
    }
  }
  dplyr::bind_rows(test_rows)
}

figure_tests <- calculate_figure_tests(figure_counts)

figure_status_counts <- figure_counts |>
  dplyr::left_join(category_n, by = c("source_sheet", "umbrella")) |>
  dplyr::left_join(
    figure_tests,
    by = c("source_sheet", "umbrella", "figure_scope")
  )

write_csv_out(figure_status_counts, "figure4_main_chromosome_status_counts.csv")
write_csv_out(figure_tests, "figure4_exploratory_chisquare_tests.csv")

all_scaffold_counts <- target_analysis |>
  dplyr::filter(.data$show_in_figure) |>
  dplyr::count(
    .data$source_sheet, .data$umbrella, .data$figure_scope,
    .data$x_label, .data$x_order, .data$status,
    name = "n_genes"
  ) |>
  dplyr::group_by(
    .data$source_sheet, .data$umbrella, .data$figure_scope,
    .data$x_label, .data$x_order
  ) |>
  dplyr::mutate(percentage = 100 * .data$n_genes / sum(.data$n_genes)) |>
  dplyr::ungroup()

all_scaffold_category_n <- target_gene_summary |>
  dplyr::count(.data$source_sheet, .data$umbrella, name = "n_targets")

all_scaffold_tests <- calculate_figure_tests(all_scaffold_counts)
all_scaffold_status_counts <- all_scaffold_counts |>
  dplyr::left_join(
    all_scaffold_category_n,
    by = c("source_sheet", "umbrella")
  ) |>
  dplyr::left_join(
    all_scaffold_tests,
    by = c("source_sheet", "umbrella", "figure_scope")
  )

write_csv_out(
  all_scaffold_status_counts,
  "figure4_all_scaffolds_status_counts.csv"
)
write_csv_out(
  all_scaffold_tests,
  "figure4_all_scaffolds_exploratory_chisquare_tests.csv"
)

unplaced_counts <- target_analysis |>
  dplyr::filter(
    .data$sequence_class == "Unplaced scaffold",
    .data$show_in_figure
  ) |>
  dplyr::count(
    .data$source_sheet, .data$umbrella, .data$figure_scope,
    .data$x_label, .data$x_order, .data$status,
    name = "n_genes"
  ) |>
  dplyr::group_by(
    .data$source_sheet, .data$umbrella, .data$figure_scope,
    .data$x_label, .data$x_order
  ) |>
  dplyr::mutate(percentage = 100 * .data$n_genes / sum(.data$n_genes)) |>
  dplyr::ungroup()

unplaced_category_n <- target_gene_summary |>
  dplyr::filter(.data$sequence_class == "Unplaced scaffold") |>
  dplyr::count(.data$source_sheet, .data$umbrella, name = "n_targets")

unplaced_tests <- calculate_figure_tests(unplaced_counts)
unplaced_status_counts <- unplaced_counts |>
  dplyr::left_join(
    unplaced_category_n,
    by = c("source_sheet", "umbrella")
  ) |>
  dplyr::left_join(
    unplaced_tests,
    by = c("source_sheet", "umbrella", "figure_scope")
  )

write_csv_out(
  unplaced_status_counts,
  "figure4_unplaced_scaffold_status_counts.csv"
)
write_csv_out(
  unplaced_tests,
  "figure4_unplaced_exploratory_chisquare_tests.csv"
)

# Use the same three review classes defined on the scaffold-origin audit page.
# These classes flag provenance uncertainty; they do not establish contamination.
homology_review_classes <- c(
  "Concordant non-metazoan nucleotide and protein hits",
  "Conflicting nucleotide and protein evidence",
  "Non-metazoan-only protein hits without nucleotide corroboration"
)

homology_review_targets <- homology_evidence |>
  dplyr::filter(.data$integrated_evidence %in% homology_review_classes) |>
  dplyr::transmute(
    gene_id,
    homology_evidence = .data$integrated_evidence,
    evidence_tier = dplyr::case_when(
      .data$integrated_evidence ==
        "Concordant non-metazoan nucleotide and protein hits" ~
        "Concordant nucleotide and protein flag",
      .data$integrated_evidence ==
        "Conflicting nucleotide and protein evidence" ~
        "Conflicting nucleotide and protein evidence",
      TRUE ~ "Protein-only non-metazoan flag"
    ),
    nucleotide_groups,
    nucleotide_taxa,
    nucleotide_max_identity,
    nucleotide_max_query_coverage,
    protein_support
  )

homology_flag_analysis <- target_analysis |>
  dplyr::inner_join(homology_review_targets, by = "gene_id") |>
  dplyr::filter(.data$show_in_figure)

homology_flag_gene_audit <- gene_audit |>
  dplyr::inner_join(homology_review_targets, by = "gene_id") |>
  dplyr::arrange(
    factor(
      .data$evidence_tier,
      levels = c(
        "Concordant nucleotide and protein flag",
        "Conflicting nucleotide and protein evidence",
        "Protein-only non-metazoan flag"
      )
    ),
    .data$source_sheet,
    .data$umbrella,
    .data$gene_id
  )

homology_flag_counts <- homology_flag_analysis |>
  dplyr::count(
    .data$source_sheet, .data$umbrella, .data$figure_scope,
    .data$x_label, .data$x_order, .data$status,
    name = "n_genes"
  ) |>
  dplyr::group_by(
    .data$source_sheet, .data$umbrella, .data$figure_scope,
    .data$x_label, .data$x_order
  ) |>
  dplyr::mutate(percentage = 100 * .data$n_genes / sum(.data$n_genes)) |>
  dplyr::ungroup()

homology_flag_category_n <- homology_flag_gene_audit |>
  dplyr::count(.data$source_sheet, .data$umbrella, name = "n_targets")

homology_flag_tests <- calculate_figure_tests(homology_flag_counts)
homology_flag_status_counts <- homology_flag_counts |>
  dplyr::left_join(
    homology_flag_category_n,
    by = c("source_sheet", "umbrella")
  ) |>
  dplyr::left_join(
    homology_flag_tests,
    by = c("source_sheet", "umbrella", "figure_scope")
  )

write_csv_out(
  homology_flag_gene_audit,
  "figure4_non_grasshopper_homology_flag_gene_audit.csv"
)
write_csv_out(
  homology_flag_analysis |>
    dplyr::select(
      .data$gene_id, .data$description_gff, .data$source_sheet,
      .data$umbrella, .data$subcategory, .data$seqid,
      .data$homology_evidence, .data$evidence_tier,
      .data$nucleotide_groups, .data$nucleotide_taxa,
      .data$nucleotide_max_identity,
      .data$nucleotide_max_query_coverage,
      .data$protein_support, .data$contrast_label,
      .data$oriented_log2FC, .data$padj, .data$status,
      .data$higher_expression_condition
    ),
  "figure4_non_grasshopper_homology_flag_contrasts.csv"
)
write_csv_out(
  homology_flag_status_counts,
  "figure4_non_grasshopper_homology_flag_status_counts.csv"
)

category_display <- c(
  "Innate Immune Recognition & Signaling" =
    "Innate immune\nrecognition &\nsignaling",
  "Humoral Immunity" = "Humoral\nimmunity",
  "Cellular Immunity" = "Cellular\nimmunity",
  "Detoxification" = "Detoxification",
  "Protein Synthesis & Translation" =
    "Protein synthesis\n& translation",
  "Nutrient Sensing & TOR/Insulin Signaling" =
    "Nutrient sensing\n& TOR/insulin\nsignaling",
  "Nitrogen & Amino Acid Metabolism" =
    "Nitrogen & amino acid\nmetabolism",
  "Storage Hexamerins & Reserves" =
    "Storage hexamerins\n& reserves"
)

status_colors <- c(
  "Downregulated" = "#3B59B6",
  "N.S." = "#CCCCCC",
  "Upregulated" = "#9E1B22"
)
diet_colors <- c("33" = "#C8641A", "50" = "#1B6B3A", "83" = "#5A2A82")

build_figure4_panel <- function(
    status_count_data, sheet_name, scope_name, tag,
    show_y = TRUE, show_legend = FALSE) {
  panel_data <- status_count_data |>
    dplyr::filter(
      .data$source_sheet == sheet_name,
      .data$figure_scope == scope_name
    ) |>
    dplyr::mutate(
      umbrella = factor(
        .data$umbrella,
        levels = category_order[category_order %in% unique(.data$umbrella)]
      ),
      status = factor(
        .data$status,
        levels = c("Downregulated", "N.S.", "Upregulated")
      ),
      x_label = factor(
        .data$x_label,
        levels = unique(.data$x_label[order(.data$x_order)])
      ),
      outline = if (scope_name == "Within-diet infection response") {
        unname(diet_colors[as.character(.data$x_label)])
      } else {
        "#252525"
      }
    )

  panel_annotations <- panel_data |>
    dplyr::distinct(
      .data$umbrella, .data$p_label, .data$n_targets
    ) |>
    dplyr::mutate(
      x_label = factor(
        levels(panel_data$x_label)[[2]],
        levels = levels(panel_data$x_label)
      )
    )

  panel_title <- if (scope_name == "Within-diet infection response") {
    "Within-diet comparisons\n(Infected vs control)"
  } else {
    "Across-diet comparisons among infected\n(Higher protein vs lower protein)"
  }

  group_title <- dplyr::recode(
    sheet_name,
    "Immune" = "Immune-related",
    "Protein-Anabolism" = "Protein anabolism"
  )
  y_title <- if (show_y) {
    paste0(group_title, "\n\nPercentage of curated target genes (%)")
  } else {
    NULL
  }

  ggplot(panel_data, aes(x = x_label, y = percentage, fill = status)) +
    geom_col(
      aes(color = outline),
      width = 0.66, linewidth = 0.65
    ) +
    geom_text(
      data = panel_annotations,
      aes(x = x_label, y = 108, label = p_label),
      inherit.aes = FALSE,
      size = 3.3, fontface = "bold"
    ) +
    geom_text(
      data = panel_annotations,
      aes(x = x_label, y = -13, label = paste0("targets = ", n_targets)),
      inherit.aes = FALSE,
      size = 3.2
    ) +
    facet_wrap(
      ~ umbrella,
      nrow = 1,
      labeller = as_labeller(category_display)
    ) +
    scale_fill_manual(values = status_colors, drop = FALSE) +
    scale_color_identity() +
    scale_y_continuous(
      breaks = c(0, 50, 100),
      limits = c(-17, 113),
      expand = expansion(mult = c(0, 0))
    ) +
    coord_cartesian(clip = "off") +
    labs(
      tag = tag,
      title = panel_title,
      x = NULL,
      y = y_title,
      fill = NULL
    ) +
    theme_classic(base_size = 11) +
    theme(
      plot.tag = element_text(face = "bold", size = 17),
      plot.tag.position = c(0.005, 1.02),
      plot.title = element_text(
        face = "bold", size = 13, hjust = 0.5,
        margin = margin(b = 10)
      ),
      axis.title.y = element_text(size = 12, margin = margin(r = 7)),
      axis.text.y = if (show_y) {
        element_text(size = 10, color = "black")
      } else {
        element_blank()
      },
      axis.ticks.y = if (show_y) element_line() else element_blank(),
      axis.line.y = if (show_y) element_line() else element_blank(),
      axis.text.x = element_text(size = 10.5, color = "black", lineheight = 0.9),
      axis.ticks.x = element_blank(),
      strip.background = element_blank(),
      strip.text = element_text(
        face = "bold", size = 10.5, lineheight = 0.9,
        margin = margin(b = 7)
      ),
      panel.spacing.x = grid::unit(0.35, "lines"),
      legend.position = if (show_legend) "bottom" else "none",
      legend.text = element_text(size = 11),
      legend.key.size = grid::unit(4.5, "mm"),
      plot.margin = margin(12, 8, 30, 8)
    )
}

p_a <- build_figure4_panel(
  figure_status_counts, "Immune", "Within-diet infection response", "A",
  TRUE, FALSE
)
p_b <- build_figure4_panel(
  figure_status_counts, "Immune", "Diet comparison among infected", "B", FALSE
)
p_c <- build_figure4_panel(
  figure_status_counts, "Protein-Anabolism",
  "Within-diet infection response", "C", TRUE
)
p_d <- build_figure4_panel(
  figure_status_counts, "Protein-Anabolism",
  "Diet comparison among infected", "D", FALSE
)

legend_data <- tibble::tibble(
  status = factor(
    c("Downregulated", "N.S.", "Upregulated"),
    levels = c("Downregulated", "N.S.", "Upregulated")
  ),
  x = seq_len(3),
  y = 1
)
legend_plot <- ggplot(legend_data, aes(x = x, y = y, fill = status)) +
  geom_point(shape = 22, size = 5, color = NA) +
  scale_fill_manual(values = status_colors, drop = FALSE) +
  guides(fill = guide_legend(title = NULL, nrow = 1)) +
  theme_void(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 12),
    legend.key.size = grid::unit(5, "mm")
  )
shared_status_legend <- cowplot::get_legend(legend_plot)

figure4_main_chromosome <- (
  (p_a | p_b) /
  (p_c | p_d) /
  patchwork::wrap_elements(full = shared_status_legend)
) +
  patchwork::plot_layout(
    heights = c(1, 1, 0.075)
  ) +
  patchwork::plot_annotation(
    title = "Curated immune and protein-anabolism targets on main chromosomes",
    subtitle = paste0(
      "Competitive-host DESeq2; ", main_targets,
      " retained targets; padj < ", alpha,
      " and |log2FC| >= ", lfc_threshold
    ),
    theme = theme(
      plot.title = element_text(face = "bold", size = 20, hjust = 0.5),
      plot.subtitle = element_text(size = 12.5, hjust = 0.5),
      plot.margin = margin(10, 10, 5, 10)
    )
  )

figure4_main_chromosome
save_figure(
  figure4_main_chromosome,
  "figure4_curated_targets_main_chromosomes",
  15,
  11.5
)

all_a <- build_figure4_panel(
  all_scaffold_status_counts, "Immune", "Within-diet infection response", "A",
  TRUE, FALSE
)
all_b <- build_figure4_panel(
  all_scaffold_status_counts, "Immune", "Diet comparison among infected", "B",
  FALSE
)
all_c <- build_figure4_panel(
  all_scaffold_status_counts, "Protein-Anabolism",
  "Within-diet infection response", "C", TRUE
)
all_d <- build_figure4_panel(
  all_scaffold_status_counts, "Protein-Anabolism",
  "Diet comparison among infected", "D", FALSE
)

figure4_all_scaffolds <- (
  (all_a | all_b) /
  (all_c | all_d) /
  patchwork::wrap_elements(full = shared_status_legend)
) +
  patchwork::plot_layout(heights = c(1, 1, 0.075)) +
  patchwork::plot_annotation(
    title = "Curated immune and protein-anabolism targets across all scaffolds",
    subtitle = paste0(
      "Competitive-host DESeq2; ", total_targets,
      " original curated targets; padj < ", alpha,
      " and |log2FC| >= ", lfc_threshold
    ),
    theme = theme(
      plot.title = element_text(face = "bold", size = 20, hjust = 0.5),
      plot.subtitle = element_text(size = 12.5, hjust = 0.5),
      plot.margin = margin(10, 10, 5, 10)
    )
  )

save_figure(
  figure4_all_scaffolds,
  "figure4_curated_targets_all_scaffolds",
  15,
  11.5
)

u_a <- build_figure4_panel(
  unplaced_status_counts, "Immune", "Within-diet infection response", "A",
  TRUE, FALSE
)
u_b <- build_figure4_panel(
  unplaced_status_counts, "Immune", "Diet comparison among infected", "B",
  FALSE
)
u_c <- build_figure4_panel(
  unplaced_status_counts, "Protein-Anabolism",
  "Within-diet infection response", "C", TRUE
)
u_d <- build_figure4_panel(
  unplaced_status_counts, "Protein-Anabolism",
  "Diet comparison among infected", "D", FALSE
)

figure4_unplaced_scaffolds <- (
  (u_a | u_b) /
  (u_c | u_d) /
  patchwork::wrap_elements(full = shared_status_legend)
) +
  patchwork::plot_layout(heights = c(1, 1, 0.075)) +
  patchwork::plot_annotation(
    title = "Curated immune and protein-anabolism targets on unplaced scaffolds",
    subtitle = paste0(
      "Competitive-host DESeq2; ", unplaced_targets_n,
      " audit targets; padj < ", alpha,
      " and |log2FC| >= ", lfc_threshold
    ),
    theme = theme(
      plot.title = element_text(face = "bold", size = 20, hjust = 0.5),
      plot.subtitle = element_text(size = 12.5, hjust = 0.5),
      plot.margin = margin(10, 10, 5, 10)
    )
  )

save_figure(
  figure4_unplaced_scaffolds,
  "figure4_curated_targets_unplaced_scaffolds",
  15,
  11.5
)

flag_a <- build_figure4_panel(
  homology_flag_status_counts, "Immune",
  "Within-diet infection response", "A", TRUE, FALSE
)
flag_b <- build_figure4_panel(
  homology_flag_status_counts, "Immune",
  "Diet comparison among infected", "B", FALSE
)
flag_c <- build_figure4_panel(
  homology_flag_status_counts, "Protein-Anabolism",
  "Within-diet infection response", "C", TRUE
)
flag_d <- build_figure4_panel(
  homology_flag_status_counts, "Protein-Anabolism",
  "Diet comparison among infected", "D", FALSE
)

homology_flag_target_n <- dplyr::n_distinct(homology_flag_analysis$gene_id)
homology_concordant_n <- sum(
  homology_flag_gene_audit$evidence_tier ==
    "Concordant nucleotide and protein flag"
)

figure4_homology_flags <- (
  (flag_a | flag_b) /
  (flag_c | flag_d) /
  patchwork::wrap_elements(full = shared_status_legend)
) +
  patchwork::plot_layout(heights = c(1, 1, 0.075)) +
  patchwork::plot_annotation(
    title = "Curated targets with non-grasshopper homology review flags",
    subtitle = paste0(
      homology_flag_target_n,
      " unplaced-scaffold targets; only ", homology_concordant_n,
      " have concordant non-metazoan nucleotide and protein hits"
    ),
    theme = theme(
      plot.title = element_text(face = "bold", size = 20, hjust = 0.5),
      plot.subtitle = element_text(size = 12.5, hjust = 0.5),
      plot.margin = margin(10, 10, 5, 10)
    )
  )

save_figure(
  figure4_homology_flags,
  "figure4_curated_targets_non_grasshopper_homology_flags",
  15,
  11.5
)
```

### Figure 4 views {.tabset}

#### Main chromosomes

```{r display-figure4-main, echo=FALSE, fig.width=15, fig.height=11.5}
figure4_main_chromosome
```

#### All scaffolds

```{r display-figure4-all-scaffolds, echo=FALSE, fig.width=15, fig.height=11.5}
figure4_all_scaffolds
```

#### Homology review flags

```{r display-figure4-homology-flags, echo=FALSE, fig.width=15, fig.height=11.5}
figure4_homology_flags
```

The first composite is the conservative host-chromosome view used for the main
paper interpretation. The second restores all 163 genes from the original
curated workbook, including the 64 targets on unplaced scaffolds, so its target
denominators are directly comparable with the earlier figure. Both use the same
current DESeq2 model, contrasts, thresholds, colors, and panel geometry. The
left panels summarize infection responses within each diet; the right panels
compare infected locusts across diets. The exploratory chi-square labels
summarize category composition and do not replace the gene-level DESeq2 tests.

The third composite isolates the `r homology_flag_target_n` curated targets that
meet the scaffold-origin page's predefined homology-review rule. Of these,
`r homology_concordant_n` have non-metazoan top hits in both BLASTn and DIAMOND;
the remainder have conflicting evidence or protein-only support. These are
**provenance-review flags, not confirmed non-locust genes**. Conserved proteins,
incomplete reference representation, and short or divergent alignments can all
produce a non-grasshopper top hit.

### Genes behind the homology-flagged figure

```{r homology-flagged-curated-table, echo=FALSE}
homology_flag_table <- homology_flag_analysis |>
  dplyr::select(
    .data$gene_id, description = .data$description_gff,
    functional_group = .data$umbrella, .data$subcategory,
    scaffold = .data$seqid, .data$evidence_tier,
    .data$nucleotide_groups, .data$nucleotide_taxa,
    .data$contrast_label, log2FC = .data$oriented_log2FC,
    adjusted_p_value = .data$padj, .data$status,
    higher_in = .data$higher_expression_condition
  ) |>
  dplyr::mutate(
    log2FC = round(.data$log2FC, 3),
    adjusted_p_value = signif(.data$adjusted_p_value, 3)
  ) |>
  dplyr::arrange(
    factor(
      .data$evidence_tier,
      levels = c(
        "Concordant nucleotide and protein flag",
        "Conflicting nucleotide and protein evidence",
        "Protein-only non-metazoan flag"
      )
    ),
    .data$functional_group,
    .data$gene_id,
    .data$contrast_label
  )

DT::datatable(
  homology_flag_table,
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    pageLength = 15,
    scrollX = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv")
  ),
  caption = paste0(
    "Curated Figure 4 targets carrying scaffold-origin homology review flags. ",
    "Positive log2FC follows the condition order printed in the contrast."
  )
) |>
  DT::formatStyle(
    "status",
    color = DT::styleEqual(
      c("Downregulated", "N.S.", "Upregulated"),
      c("#244A9B", "#555555", "#8B1018")
    ),
    fontWeight = "bold"
  )
```

## Comparison with the earlier workbook

```{r old-new-comparison}
status_transition <- target_analysis |>
  dplyr::filter(.data$primary_figure) |>
  dplyr::count(
    .data$original_status,
    current_status = .data$status,
    name = "gene_contrast_cells"
  ) |>
  dplyr::arrange(.data$original_status, .data$current_status)

changed_cells <- target_analysis |>
  dplyr::filter(
    .data$primary_figure,
    .data$original_status != .data$status
  )

write_csv_out(status_transition, "figure4_old_to_current_status_transitions.csv")
write_csv_out(changed_cells, "figure4_changed_placed_target_cells.csv")

knitr::kable(
  status_transition,
  col.names = c("Workbook status", "Current status", "Gene-contrast cells"),
  align = c("l", "l", "r")
)
```

There are `r nrow(changed_cells)` placed-gene contrast cells whose status differs
between the supplied workbook and the current competitive-host analysis. This
comparison separates two effects: removal of unplaced targets changes the
category denominator, while remapping and re-fitting DESeq2 can change the
status of targets that remain on main chromosomes.

## Unplaced target details

```{r unplaced-targets}
unplaced_followup <- target_analysis |>
  dplyr::filter(.data$sequence_class == "Unplaced scaffold") |>
  dplyr::group_by(
    .data$source_sheet, .data$umbrella, .data$subcategory,
    .data$gene_id, .data$description_workbook, .data$description_gff,
    .data$gene_biotype, .data$seqid
  ) |>
  dplyr::summarise(
    significant_in_current_unplaced_analysis = any(.data$significant),
    n_significant_contrasts = sum(.data$significant),
    significant_contrasts = paste(
      .data$contrast_label[.data$significant],
      collapse = "; "
    ),
    higher_expression_conditions = paste(
      .data$higher_expression_condition[.data$significant],
      collapse = "; "
    ),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    significant_contrasts = dplyr::if_else(
      .data$significant_contrasts == "",
      "None at current thresholds",
      .data$significant_contrasts
    ),
    higher_expression_conditions = dplyr::if_else(
      .data$higher_expression_conditions == "",
      "Not significant",
      .data$higher_expression_conditions
    )
  ) |>
  dplyr::arrange(.data$source_sheet, .data$umbrella, .data$gene_id)

unplaced_33_vs_50 <- target_analysis |>
  dplyr::filter(
    .data$sequence_class == "Unplaced scaffold",
    .data$contrast_id == "infected_diet33_vs50",
    .data$significant
  ) |>
  dplyr::distinct(.data$gene_id) |>
  nrow()

write_csv_out(unplaced_followup, "curated_unplaced_targets_for_followup.csv")
```

Of the `r unplaced_targets_n` excluded curated targets,
`r unplaced_33_vs_50` are significant in the current unplaced-scaffold analysis
for infected diet 33 versus infected diet 50. These genes are therefore a
specific follow-up set for the scaffold-origin work, not evidence to include in
the primary host-chromosome Figure 4.

```{r unplaced-searchable-table}
DT::datatable(
  unplaced_followup,
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv")
  )
)
```

## Searchable retained target table

```{r retained-searchable-table}
retained_target_table <- target_analysis |>
  dplyr::filter(.data$primary_figure) |>
  dplyr::select(
    source_sheet, umbrella, subcategory, gene_id,
    description_gff, gene_biotype, assigned_molecule,
    contrast_label, log2FoldChange, padj, status,
    higher_expression_condition
  ) |>
  dplyr::arrange(.data$source_sheet, .data$umbrella, .data$gene_id)

DT::datatable(
  retained_target_table,
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv")
  )
)
```

## Exported files

```{r export-manifest}
manifest <- tibble::tibble(
  item = c(
    "Frozen curated workbook",
    "Official placement table",
    "Current DESeq2 result table",
    "Corrected BLASTn/DIAMOND homology evidence",
    "Output directory"
  ),
  path = c(
    workbook_file,
    placement_file,
    results_file,
    homology_evidence_file,
    out_dir
  )
)
write_csv_out(manifest, "figure4_input_manifest.csv")

knitr::kable(manifest, col.names = c("Item", "Path"), align = c("l", "l"))
```

Each knit creates a new timestamped output folder. Matched composites are
exported as `figure4_curated_targets_main_chromosomes` and
`figure4_curated_targets_all_scaffolds` in PNG, SVG, and PDF formats. The
standalone homology-review composite is exported as
`figure4_curated_targets_non_grasshopper_homology_flags`. The unplaced-only
composite, placement-impact figure, and all gene-level audit tables are exported
beside them for sensitivity review.

## Session information

```{r session-info}
sessionInfo()
```
