---
title: "All samples + all scaffolds: searchable DEG catalogue"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  placement_file: "output/runs/host_pathogen_dual_20260727_000952/local-analysis/host-mapping-audit_20260729-163928/unified-model/audit-summary/host_gene_sequence_placement.tsv"
  eggnog_file: "data/GO_Annotations/GCF_023897955.1_iqSchGreg1.2_Arthopoda_one2one.emapper.annotations"
  gff_file: "data/reference/GCF_023897955.1_iqSchGreg1.2_genomic.gff"
  latest_results_root: "output/rmd_runs"
  alpha: 0.05
  lfc_threshold: 1
  all_results_file: "output/runs/host_pathogen_dual_20260727_000952/local-analysis/host-mapping-audit_20260729-163928/unified-model/competitive-host/host_deseq2_all_contrasts.csv"
  output_tag: "all-scaffolds-all-deg-catalogue"
---

## Purpose

This page is a searchable catalogue of all significant DEGs from the
all-samples retained, all-scaffold analysis. It uses the fresh joint
competitive-host DESeq2 model and keeps every expression-filtered non-rRNA host
locus, including genes on nuclear chromosomes and unplaced scaffolds. The
official placement class is retained as a searchable column rather than used as
a filter.

The table shows the gene ID, RefSeq/GFF description, contrast, log2 fold
change, and which side of the contrast has higher expression. Positive log2
fold change is colored red; negative log2 fold change is colored blue.

The `Description` column is taken directly from the gene-level `description=`
attribute in the RefSeq/GFF annotation from the same *S. gregaria* assembly.
eggNOG is still used only for the GO, KEGG, and PFAM annotation columns.

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c("dplyr", "tidyr", "stringr", "DT", "knitr", "ggplot2",
              "scales", "readr", "tibble")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

placement_file <- file.path(project_dir, params$placement_file)
eggnog_file <- file.path(project_dir, params$eggnog_file)
gff_file <- file.path(project_dir, params$gff_file)
results_root <- file.path(project_dir, params$latest_results_root)

direction_colors <- c("Upregulated" = "#E41A1C",
                      "Downregulated" = "#1E88E5")
cluster_colors <- c("DEG cluster 1" = "#315C8A",
                    "DEG cluster 2" = "#C85A54",
                    "DEG cluster 3+4" = "#2F8F5B")
```

## Helper functions

```{r helpers}
latest_dir <- function(pattern) {
  dirs <- list.dirs(results_root, recursive = FALSE, full.names = TRUE)
  dirs <- dirs[grepl(pattern, basename(dirs))]
  if (length(dirs) == 0) {
    stop("No output folder matched pattern: ", pattern)
  }
  dirs[which.max(file.info(dirs)$mtime)]
}

read_emapper <- function(path) {
  cols <- c(
    "query", "seed_ortholog", "evalue", "score", "eggNOG_OGs",
    "max_annot_lvl", "COG_category", "eggNOG_Description", "eggNOG_name",
    "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
    "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC", "CAZy",
    "BiGG_Reaction", "PFAMs"
  )
  tab <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", fill = TRUE)
  names(tab) <- cols[seq_len(ncol(tab))]
  tab |>
    dplyr::mutate(protein_id = stringr::str_extract(query, "XP_[0-9]+[.][0-9]+"))
}

parse_cds_gene_bridge <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  gff |>
    dplyr::filter(type == "CDS") |>
    dplyr::transmute(
      protein_id = stringr::str_match(attributes, "protein_id=([^;]+)")[, 2],
      gene_id = stringr::str_match(attributes, "gene=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(protein_id), !is.na(gene_id)) |>
    dplyr::distinct()
}

parse_gff_gene_annotation <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  gff |>
    dplyr::filter(type == "gene") |>
    dplyr::transmute(
      gene_id = dplyr::coalesce(
        stringr::str_match(attributes, "gene=([^;]+)")[, 2],
        stringr::str_match(attributes, "Name=([^;]+)")[, 2]
      ),
      gff_description = stringr::str_match(attributes, "description=([^;]+)")[, 2],
      gene_biotype = stringr::str_match(attributes, "gene_biotype=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(gene_id)) |>
    dplyr::distinct(gene_id, .keep_all = TRUE)
}

first_nonempty <- function(x, fallback = "unannotated") {
  x <- x[!is.na(x) & x != "" & x != "-"]
  if (length(x) == 0) fallback else x[1]
}

make_condition_label <- function(comparison_type, numerator, denominator, diet,
                                 side = c("numerator", "denominator")) {
  side <- match.arg(side)
  value <- if (side == "numerator") numerator else denominator
  dplyr::case_when(
    comparison_type == "Infection within diet" ~
      paste(value, "diet", diet),
    grepl("control", comparison_type, ignore.case = TRUE) ~
      paste("Control diet", value),
    grepl("infected", comparison_type, ignore.case = TRUE) ~
      paste("Infected diet", value),
    TRUE ~ value
  )
}

save_png_svg <- function(plot, file_base, width, height, dpi = 300) {
  png_path <- file.path(out_dir, paste0(file_base, ".png"))
  svg_path <- file.path(out_dir, paste0(file_base, ".svg"))

  ggplot2::ggsave(png_path, plot, width = width, height = height,
                  dpi = dpi, limitsize = FALSE)

  grDevices::svg(svg_path, width = width, height = height, onefile = FALSE)
  on.exit(grDevices::dev.off(), add = TRUE)
  print(plot)

  invisible(c(png = png_path, svg = svg_path))
}

slugify <- function(x) {
  x |>
    stringr::str_to_lower() |>
    stringr::str_replace_all("[^a-z0-9]+", "_") |>
    stringr::str_replace_all("^_|_$", "")
}
```

## Input files

```{r input-files}
scenario_results_file <- file.path(project_dir, params$all_results_file)

input_files <- data.frame(
  input = c("Joint all-scaffold DESeq2 table", "Official gene placement",
            "eggNOG functional annotation", "GFF gene descriptions"),
  path = c(scenario_results_file, placement_file, eggnog_file, gff_file),
  present = c(file.exists(scenario_results_file),
              file.exists(placement_file), file.exists(eggnog_file),
              file.exists(gff_file))
)

knitr::kable(input_files, caption = "Input files used for the DEG catalogue")
if (!all(input_files$present)) {
  stop("One or more required inputs are missing.")
}
```

## Annotation and DEG catalogue

```{r build-catalogue}
placement <- readr::read_tsv(placement_file, show_col_types = FALSE) |>
  dplyr::distinct(gene_id, .keep_all = TRUE)
all_results <- read.csv(scenario_results_file, check.names = FALSE)
gene_universe <- unique(all_results$gene_id[!is.na(all_results$gene_id)])

gff_annotation <- parse_gff_gene_annotation(gff_file) |>
  dplyr::filter(gene_id %in% gene_universe)

eggnog_annotation <- read_emapper(eggnog_file) |>
  dplyr::left_join(parse_cds_gene_bridge(gff_file), by = "protein_id") |>
  dplyr::filter(!is.na(gene_id), gene_id %in% gene_universe) |>
  dplyr::group_by(gene_id) |>
  dplyr::summarise(
    GOs = first_nonempty(GOs, ""),
    KEGG_ko = first_nonempty(KEGG_ko, ""),
    KEGG_Pathway = first_nonempty(KEGG_Pathway, ""),
    PFAMs = first_nonempty(PFAMs, ""),
    .groups = "drop"
  )

annotation <- data.frame(gene_id = gene_universe) |>
  dplyr::left_join(gff_annotation, by = "gene_id") |>
  dplyr::left_join(eggnog_annotation, by = "gene_id") |>
  dplyr::mutate(
    Description = dplyr::coalesce(gff_description, "no GFF description"),
    gene_biotype = dplyr::coalesce(gene_biotype, "unknown")
  ) |>
  dplyr::select(
    gene_id, Description, gene_biotype,
    GOs, KEGG_ko, KEGG_Pathway, PFAMs
  )

contrast_map <- tibble::tribble(
  ~contrast_id, ~comparison_type, ~comparison,
  "infection_diet33", "Infection within diet", "Diet 33: infected vs control",
  "infection_diet50", "Infection within diet", "Diet 50: infected vs control",
  "infection_diet83", "Infection within diet", "Diet 83: infected vs control",
  "control_diet33_vs50", "Diet contrasts among control", "Control 33 vs 50",
  "control_diet83_vs33", "Diet contrasts among control", "Control 83 vs 33",
  "control_diet83_vs50", "Diet contrasts among control", "Control 83 vs 50",
  "infected_diet33_vs50", "Diet contrasts among infected", "Infected 33 vs 50",
  "infected_diet83_vs33", "Diet contrasts among infected", "Infected 83 vs 33",
  "infected_diet83_vs50", "Diet contrasts among infected", "Infected 83 vs 50",
  "infection_average_all_diets", "Overall infection", "Average infected vs control",
  "interaction_diet33_vs50", "Diet by infection interaction", "Infection response: 33 vs 50",
  "interaction_diet83_vs50", "Diet by infection interaction", "Infection response: 83 vs 50",
  "interaction_diet83_vs33", "Diet by infection interaction", "Infection response: 83 vs 33"
)

deg_raw <- all_results |>
  dplyr::filter(
    mapping_branch == "competitive-host",
    gene_id %in% gene_universe,
    significant,
    !is.na(gene_id),
    !is.na(padj),
    padj < params$alpha,
    abs(log2FoldChange) >= params$lfc_threshold
  ) |>
  dplyr::inner_join(contrast_map, by = "contrast_id") |>
  dplyr::left_join(
    placement |>
      dplyr::select(gene_id, seqid, sequence_class, assigned_molecule,
                    location_type),
    by = "gene_id"
  ) |>
  # The searchable table uses the exact current GFF annotation joined below.
  dplyr::select(-dplyr::any_of(c("description", "gene_biotype")))

deg_catalogue <- deg_raw |>
  dplyr::mutate(
    numerator_condition = numerator,
    denominator_condition = denominator,
    higher_in = dplyr::if_else(log2FoldChange > 0,
                               numerator_condition, denominator_condition),
    lower_in = dplyr::if_else(log2FoldChange > 0,
                              denominator_condition, numerator_condition),
    signed_direction = dplyr::if_else(log2FoldChange > 0,
                                      "Upregulated", "Downregulated")
  ) |>
  dplyr::left_join(annotation, by = "gene_id") |>
  dplyr::mutate(
    Description = dplyr::coalesce(Description, "no GFF description")
  ) |>
  dplyr::select(
    gene_id, Description,
    gene_biotype, seqid, sequence_class, assigned_molecule, location_type,
    comparison_type, comparison, numerator_condition, denominator_condition,
    higher_in, lower_in, signed_direction, log2FoldChange, lfcSE, baseMean,
    stat, pvalue, padj, GOs, KEGG_ko, KEGG_Pathway, PFAMs
  ) |>
  dplyr::arrange(comparison_type, comparison, padj,
                 dplyr::desc(abs(log2FoldChange)))

write.csv(deg_catalogue, file.path(out_dir, "all_samples_deg_catalogue.csv"),
          row.names = FALSE)

catalogue_summary <- deg_catalogue |>
  dplyr::count(comparison_type, comparison, signed_direction,
               name = "n_DEGs") |>
  dplyr::arrange(comparison_type, comparison, signed_direction)

comparison_type_order <- c("Diet contrasts among control",
                           "Diet contrasts among infected",
                           "Infection within diet",
                           "Overall infection",
                           "Diet by infection interaction")
comparison_order <- catalogue_summary |>
  dplyr::distinct(comparison_type, comparison) |>
  dplyr::arrange(factor(comparison_type, levels = comparison_type_order),
                 comparison) |>
  dplyr::pull(comparison)

catalogue_summary <- catalogue_summary |>
  dplyr::mutate(
    comparison_type = factor(comparison_type,
                             levels = comparison_type_order),
    comparison = factor(comparison, levels = comparison_order),
    signed_direction = factor(signed_direction,
                              levels = c("Downregulated", "Upregulated"))
  )

write.csv(catalogue_summary,
          file.path(out_dir, "all_samples_deg_catalogue_summary.csv"),
          row.names = FALSE)

knitr::kable(
  catalogue_summary,
  caption = "Significant DEG counts in the all-samples retained catalogue"
)
```

```{r deg-summary-plot, fig.width=12, fig.height=7}
p_summary <- ggplot2::ggplot(
  catalogue_summary,
  ggplot2::aes(x = comparison, y = n_DEGs, fill = signed_direction)
) +
  ggplot2::geom_col(position = "dodge", width = 0.75) +
  ggplot2::facet_wrap(~ comparison_type, scales = "free_x") +
  ggplot2::scale_fill_manual(values = direction_colors) +
  ggplot2::labs(
    title = "All-samples retained DEG catalogue",
    subtitle = "Counts use padj < alpha and absolute log2FC >= lfc_threshold",
    x = NULL,
    y = "Number of significant DEGs",
    fill = NULL
  ) +
  ggplot2::theme_classic(base_size = 12) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 35, hjust = 1),
    legend.position = "bottom",
    plot.title = ggplot2::element_text(face = "bold")
  )

save_png_svg(p_summary, "all_samples_deg_catalogue_summary",
             width = 12, height = 7, dpi = 300)
p_summary
```

## DEG context zooms and expression-cluster proportions

The full DEG catalogue plot is useful for comparing all contrasts at once, but
the large infected-diet contrast compresses the smaller contexts. The exported
zoom panels below repeat the same counts one context at a time, with an
independent y-axis. The proportion plots then ask which expression cluster
contains the genes from each DEG contrast.

```{r deg-summary-context-exports, fig.width=13, fig.height=8}
make_context_count_plot <- function(context_label) {
  context_df <- catalogue_summary |>
    dplyr::filter(comparison_type == context_label)

  ggplot2::ggplot(
    context_df,
    ggplot2::aes(x = comparison, y = n_DEGs, fill = signed_direction)
  ) +
    ggplot2::geom_col(position = "dodge", width = 0.75) +
    ggplot2::geom_text(
      ggplot2::aes(label = n_DEGs),
      position = ggplot2::position_dodge(width = 0.75),
      vjust = -0.3,
      size = 3.8,
      fontface = "bold"
    ) +
    ggplot2::scale_fill_manual(values = direction_colors, drop = FALSE) +
    ggplot2::scale_y_continuous(
      expand = ggplot2::expansion(mult = c(0, 0.12))
    ) +
    ggplot2::labs(
      title = "All-samples retained DEG catalogue",
      subtitle = paste0(as.character(context_label),
                        " - independent y-axis zoom"),
      x = NULL,
      y = "Number of significant DEGs",
      fill = NULL
    ) +
    ggplot2::theme_classic(base_size = 13) +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 35, hjust = 1),
      legend.position = "bottom",
      plot.title = ggplot2::element_text(face = "bold")
    )
}

context_count_plots <- lapply(levels(catalogue_summary$comparison_type),
                              make_context_count_plot)
names(context_count_plots) <- levels(catalogue_summary$comparison_type)

for (context_label in names(context_count_plots)) {
  n_comparisons <- dplyr::n_distinct(
    catalogue_summary$comparison[catalogue_summary$comparison_type == context_label]
  )
  save_png_svg(
    context_count_plots[[context_label]],
    paste0("all_samples_deg_catalogue_summary_zoom_",
           slugify(context_label)),
    width = max(6.5, 2.3 * n_comparisons),
    height = 5.2,
    dpi = 300
  )
}

cluster_assignment_file <- NA_character_

cluster_export_note <- data.frame(
  input = "Expression-cluster assignments (not imposed across placement classes)",
  path = cluster_assignment_file,
  present = file.exists(cluster_assignment_file)
)
knitr::kable(cluster_export_note,
             caption = "Cluster assignment file used for proportion plots")

if (file.exists(cluster_assignment_file)) {
  cluster_levels <- names(cluster_colors)
  cluster_assignments <- read.csv(cluster_assignment_file,
                                  check.names = FALSE) |>
    dplyr::transmute(
      gene_id,
      expression_cluster = factor(expression_cluster,
                                  levels = cluster_levels)
    ) |>
    dplyr::filter(!is.na(expression_cluster)) |>
    dplyr::distinct()

  deg_cluster_counts <- deg_catalogue |>
    dplyr::inner_join(cluster_assignments, by = "gene_id") |>
    dplyr::mutate(
      comparison_type = factor(comparison_type,
                               levels = comparison_type_order),
      comparison = factor(comparison, levels = comparison_order),
      signed_direction = factor(signed_direction,
                                levels = c("Downregulated", "Upregulated"))
    ) |>
    dplyr::distinct(comparison_type, comparison, signed_direction,
                    gene_id, expression_cluster) |>
    dplyr::count(comparison_type, comparison, signed_direction,
                 expression_cluster, name = "n_genes") |>
    dplyr::group_by(comparison_type, comparison, signed_direction) |>
    tidyr::complete(
      expression_cluster = factor(cluster_levels, levels = cluster_levels),
      fill = list(n_genes = 0)
    ) |>
    dplyr::mutate(
      total_DEGs = sum(n_genes),
      proportion = dplyr::if_else(total_DEGs > 0, n_genes / total_DEGs, 0)
    ) |>
    dplyr::ungroup() |>
    dplyr::arrange(comparison_type, comparison, signed_direction,
                   expression_cluster)

  write.csv(
    deg_cluster_counts,
    file.path(out_dir, "all_samples_deg_catalogue_cluster_proportions.csv"),
    row.names = FALSE
  )

  p_cluster_proportion <- ggplot2::ggplot(
    deg_cluster_counts,
    ggplot2::aes(x = comparison, y = proportion, fill = expression_cluster)
  ) +
    ggplot2::geom_col(width = 0.78, color = "white", linewidth = 0.4) +
    ggplot2::facet_grid(
      signed_direction ~ comparison_type,
      scales = "free_x",
      space = "free_x"
    ) +
    ggplot2::scale_fill_manual(values = cluster_colors, drop = FALSE) +
    ggplot2::scale_y_continuous(
      labels = scales::percent_format(accuracy = 1),
      limits = c(0, 1),
      expand = c(0, 0)
    ) +
    ggplot2::labs(
      title = "All-samples DEG catalogue by expression cluster",
      subtitle = "Each bar sums to 100% of significant DEGs for one contrast and direction",
      x = NULL,
      y = "Proportion of significant DEGs",
      fill = "Expression cluster"
    ) +
    ggplot2::theme_classic(base_size = 13) +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 35, hjust = 1),
      strip.background = ggplot2::element_rect(fill = "#F3F4F6",
                                               color = "#111827"),
      strip.text = ggplot2::element_text(face = "bold"),
      legend.position = "bottom",
      plot.title = ggplot2::element_text(face = "bold")
    )

  save_png_svg(
    p_cluster_proportion,
    "all_samples_deg_catalogue_cluster_proportions",
    width = 13,
    height = 8,
    dpi = 300
  )

  make_context_cluster_plot <- function(context_label) {
    context_df <- deg_cluster_counts |>
      dplyr::filter(comparison_type == context_label)

    ggplot2::ggplot(
      context_df,
      ggplot2::aes(x = comparison, y = proportion, fill = expression_cluster)
    ) +
      ggplot2::geom_col(width = 0.78, color = "white", linewidth = 0.4) +
      ggplot2::facet_wrap(~ signed_direction, nrow = 1) +
      ggplot2::scale_fill_manual(values = cluster_colors, drop = FALSE) +
      ggplot2::scale_y_continuous(
        labels = scales::percent_format(accuracy = 1),
        limits = c(0, 1),
        expand = c(0, 0)
      ) +
      ggplot2::labs(
        title = "Expression-cluster proportions by DEG context",
        subtitle = paste0(as.character(context_label),
                          " - independent context zoom"),
        x = NULL,
        y = "Proportion of significant DEGs",
        fill = "Expression cluster"
      ) +
      ggplot2::theme_classic(base_size = 13) +
      ggplot2::theme(
        axis.text.x = ggplot2::element_text(angle = 35, hjust = 1),
        strip.background = ggplot2::element_rect(fill = "#F3F4F6",
                                                 color = "#111827"),
        strip.text = ggplot2::element_text(face = "bold"),
        legend.position = "bottom",
        plot.title = ggplot2::element_text(face = "bold")
      )
  }

  context_cluster_plots <- lapply(levels(catalogue_summary$comparison_type),
                                  make_context_cluster_plot)
  names(context_cluster_plots) <- levels(catalogue_summary$comparison_type)

  for (context_label in names(context_cluster_plots)) {
    n_comparisons <- dplyr::n_distinct(
      catalogue_summary$comparison[catalogue_summary$comparison_type == context_label]
    )
    save_png_svg(
      context_cluster_plots[[context_label]],
      paste0("all_samples_deg_catalogue_cluster_proportions_zoom_",
             slugify(context_label)),
      width = max(7, 2.5 * n_comparisons),
      height = 5.4,
      dpi = 300
    )
  }

  knitr::kable(
    deg_cluster_counts,
    caption = "Significant DEG proportions assigned to expression clusters"
  )
  p_cluster_proportion
} else {
  message("Cluster proportion plots skipped because the cluster assignment file was not found.")
}
```

## Searchable DEG catalogue

Use the search boxes to find a gene ID, annotation term, contrast, or condition.
The `higher_in` column is the plain-language interpretation of the log2 fold
change sign.

```{r searchable-deg-table}
display_table <- deg_catalogue |>
  dplyr::mutate(
    log2FoldChange = round(log2FoldChange, 3),
    lfcSE = round(lfcSE, 3),
    baseMean = round(baseMean, 2),
    stat = round(stat, 3),
    pvalue = signif(pvalue, 3),
    padj = signif(padj, 3)
  ) |>
  dplyr::select(
    gene_id, Description, gene_biotype, comparison_type, comparison,
    higher_in, lower_in, signed_direction, log2FoldChange, padj, baseMean,
    numerator_condition, denominator_condition, GOs, KEGG_ko, KEGG_Pathway,
    PFAMs
  )

DT::datatable(
  display_table,
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv")
  )
) |>
  DT::formatStyle(
    "log2FoldChange",
    color = DT::styleInterval(0, c("#1E88E5", "#E41A1C")),
    fontWeight = "bold"
  ) |>
  DT::formatStyle(
    "signed_direction",
    color = DT::styleEqual(names(direction_colors), direction_colors),
    fontWeight = "bold"
  )
```

## Exported files

```{r exported-files}
exports <- list.files(out_dir, full.names = FALSE)
knitr::kable(tibble::tibble(file = exports),
             caption = paste("Files written to", out_dir))
```

## Session information

```{r session-info}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
