---
title: "Exon-only versus transcript-plus-exon counting"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  exon_counts_file: "output/rmd_runs/exon-only-primary-excluding1044_20260809-233420/host_counts_no_rrna.tsv"
  transcript_exon_counts_file: "output/rmd_runs/transcript-exon-primary-excluding1044_20260809-233420/host_counts_no_rrna.tsv"
  metadata_file: "data/metadata/mehreen_metadata_analysis_44_excluding_1044.txt"
  placement_file: "output/runs/host_pathogen_dual_20260727_000952/local-analysis/host-mapping-audit_20260729-163928/unified-model/audit-summary/host_gene_sequence_placement.tsv"
  rrna_file: "data/excluded_loci/gregaria_rrna_list.txt"
  min_count: 10
  min_samples: 3
  alpha: 0.05
  lfc_threshold: 1
  output_tag: "count-definition-sensitivity"
---

## Purpose

::: {.walkthrough-grid}
::: {.walkthrough-step}
**Step 1 - Hold the experiment constant**

Retain the 44-sample analysis cohort, the competitive host alignment, the same chromosome
scope, and the same rRNA exclusion list.
:::
::: {.walkthrough-step}
**Step 2 - Change only counting mode**

Compare featureCounts using `-t exon -g gene_id` with
`-t transcript,exon -g gene_id`.
:::
::: {.walkthrough-step}
**Step 3 - Compare DEG calls**

Run identical DESeq2 models and report shared and counting-mode-specific DEGs.
:::
:::

The transcript-plus-exon definition can assign reads that overlap the span of
annotated transcript features, including intronic sequence. This makes the
additional signal **consistent with unspliced or nascent pre-mRNA**, but it does
not prove that every added read is pre-mRNA. Annotation structure, overlapping
features, and retained introns can also contribute.

This page is a sensitivity analysis. The manuscript analysis continues to use
transcript-plus-exon counts, while this comparison shows how strongly that
choice changes the tested genes and DEG calls.

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c("DESeq2", "dplyr", "tidyr", "tibble", "ggplot2", "scales",
              "patchwork", "DT")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(
  project_dir, "output", "rmd_runs",
  paste(params$output_tag, run_stamp, sep = "_")
)
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

input_paths <- c(
  exon_counts = file.path(project_dir, params$exon_counts_file),
  transcript_exon_counts = file.path(project_dir, params$transcript_exon_counts_file),
  metadata = file.path(project_dir, params$metadata_file),
  placement = file.path(project_dir, params$placement_file),
  rrna = file.path(project_dir, params$rrna_file)
)
if (!all(file.exists(input_paths))) {
  stop("Missing input(s): ", paste(input_paths[!file.exists(input_paths)], collapse = ", "))
}

mode_colors <- c(
  "Exon only" = "#4C78A8",
  "Transcript + exon" = "#B85C38"
)
membership_colors <- c(
  "Exon only" = "#4C78A8",
  "Shared" = "#6B7280",
  "Transcript + exon only" = "#B85C38"
)

theme_report <- function(base_size = 15) {
  ggplot2::theme_classic(base_size = base_size) +
    ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold", size = base_size + 3),
      plot.subtitle = ggplot2::element_text(size = base_size),
      strip.text = ggplot2::element_text(face = "bold", size = base_size),
      axis.text = ggplot2::element_text(size = base_size - 1),
      axis.title = ggplot2::element_text(face = "bold", size = base_size),
      legend.position = "bottom",
      legend.title = ggplot2::element_text(face = "bold")
    )
}

save_png_svg <- function(plot, stem, width, height) {
  ggplot2::ggsave(
    file.path(out_dir, paste0(stem, ".png")), plot,
    width = width, height = height, dpi = 300, limitsize = FALSE
  )
  if (requireNamespace("svglite", quietly = TRUE)) {
    ggplot2::ggsave(
      file.path(out_dir, paste0(stem, ".svg")), plot,
      width = width, height = height, limitsize = FALSE
    )
  }
}
```

## Step 1: prepare matched matrices

```{r prepare-inputs}
read_count_matrix <- function(path) {
  tab <- read.delim(path, check.names = FALSE)
  if (!"gene_id" %in% names(tab)) stop("Count matrix lacks gene_id: ", path)
  if (anyDuplicated(tab$gene_id)) stop("Duplicated gene IDs in: ", path)
  mat <- as.matrix(tab[, setdiff(names(tab), "gene_id"), drop = FALSE])
  storage.mode(mat) <- "integer"
  rownames(mat) <- as.character(tab$gene_id)
  mat
}

raw_counts <- list(
  "Exon only" = read_count_matrix(input_paths[["exon_counts"]]),
  "Transcript + exon" = read_count_matrix(input_paths[["transcript_exon_counts"]])
)

metadata <- read.delim(input_paths[["metadata"]], check.names = FALSE)
metadata <- metadata[, !is.na(names(metadata)) & nzchar(names(metadata)), drop = FALSE]
metadata$label <- as.character(metadata$label)
metadata$Diet <- factor(metadata$Diet, levels = c("33", "50", "83"))
metadata$Treatment <- factor(metadata$Treatment, levels = c("Control", "Infected"))
metadata$group <- factor(
  paste0("diet", metadata$Diet, "_", tolower(as.character(metadata$Treatment))),
  levels = c(
    "diet33_control", "diet50_control", "diet83_control",
    "diet33_infected", "diet50_infected", "diet83_infected"
  )
)

sample_ids <- metadata$label
if (!all(vapply(raw_counts, function(x) setequal(colnames(x), sample_ids), logical(1)))) {
  stop("Count matrices and metadata do not contain the same sample IDs.")
}
raw_counts <- lapply(raw_counts, function(x) x[, sample_ids, drop = FALSE])
rownames(metadata) <- metadata$label

rrna_ids <- trimws(readLines(input_paths[["rrna"]], warn = FALSE))
rrna_ids <- rrna_ids[nzchar(rrna_ids) & !startsWith(rrna_ids, "#")]
placement <- read.delim(input_paths[["placement"]], check.names = FALSE) |>
  dplyr::distinct(.data$gene_id, .keep_all = TRUE)
nuclear_ids <- unique(placement$gene_id[placement$sequence_class == "Nuclear chromosome"])

rrna_filtered_counts <- lapply(raw_counts, function(x) {
  x[!rownames(x) %in% rrna_ids, , drop = FALSE]
})
nuclear_counts <- lapply(rrna_filtered_counts, function(x) {
  x[rownames(x) %in% nuclear_ids, , drop = FALSE]
})

input_audit <- dplyr::bind_rows(lapply(names(raw_counts), function(mode) {
  raw <- raw_counts[[mode]]
  no_rrna <- rrna_filtered_counts[[mode]]
  nuclear <- nuclear_counts[[mode]]
  keep <- rowSums(nuclear >= params$min_count) >= params$min_samples
  tibble::tibble(
    counting_mode = mode,
    samples = ncol(raw),
    input_gene_rows = nrow(raw),
    listed_rRNA_rows_removed = sum(rownames(raw) %in% rrna_ids),
    genes_after_rRNA_removal = nrow(no_rrna),
    nuclear_chromosome_genes = nrow(nuclear),
    rRNA_rows_remaining = sum(rownames(nuclear) %in% rrna_ids),
    median_assigned_counts_per_sample = median(colSums(nuclear)),
    genes_entering_DESeq2 = sum(keep)
  )
}))

write.csv(input_audit, file.path(out_dir, "count_definition_input_audit.csv"),
          row.names = FALSE)

DT::datatable(
  input_audit,
  rownames = FALSE,
  options = list(pageLength = 5, scrollX = TRUE),
  caption = "Matched count-definition and rRNA audit"
)
```

Both matrices must report zero in `rRNA_rows_remaining`. This is a direct check
against `gregaria_rrna_list`, rather than an inference from the featureCounts
labels.

```{r input-summary-figure, fig.width=13, fig.height=5.5}
p_reads <- input_audit |>
  dplyr::mutate(
    counting_mode = factor(.data$counting_mode, levels = names(mode_colors)),
    median_millions = .data$median_assigned_counts_per_sample / 1e6
  ) |>
  ggplot2::ggplot(ggplot2::aes(
    x = .data$counting_mode, y = .data$median_millions,
    fill = .data$counting_mode
  )) +
  ggplot2::geom_col(width = 0.62, color = "#111827") +
  ggplot2::geom_text(
    ggplot2::aes(label = sprintf("%.2f M", .data$median_millions)),
    vjust = -0.35, size = 5.2, fontface = "bold"
  ) +
  ggplot2::scale_fill_manual(values = mode_colors, guide = "none") +
  ggplot2::scale_y_continuous(
    expand = ggplot2::expansion(mult = c(0, 0.14))
  ) +
  ggplot2::labs(
    title = "A. Assigned nuclear-chromosome counts",
    subtitle = "Median per sample after rRNA removal",
    x = NULL, y = "Assigned counts (millions)"
  ) +
  theme_report()

p_tested <- input_audit |>
  dplyr::mutate(counting_mode = factor(.data$counting_mode,
                                        levels = names(mode_colors))) |>
  ggplot2::ggplot(ggplot2::aes(
    x = .data$counting_mode, y = .data$genes_entering_DESeq2,
    fill = .data$counting_mode
  )) +
  ggplot2::geom_col(width = 0.62, color = "#111827") +
  ggplot2::geom_text(
    ggplot2::aes(label = scales::comma(.data$genes_entering_DESeq2)),
    vjust = -0.35, size = 5.2, fontface = "bold"
  ) +
  ggplot2::scale_fill_manual(values = mode_colors, guide = "none") +
  ggplot2::scale_y_continuous(
    labels = scales::label_comma(),
    expand = ggplot2::expansion(mult = c(0, 0.14))
  ) +
  ggplot2::labs(
    title = "B. Genes entering DESeq2",
    subtitle = paste0("Count >= ", params$min_count, " in >= ",
                      params$min_samples, " samples"),
    x = NULL, y = "Genes"
  ) +
  theme_report()

p_input_summary <- p_reads | p_tested
save_png_svg(p_input_summary, "count_definition_input_summary", 13, 5.5)
p_input_summary
```

## Step 2: run identical DESeq2 contrasts

```{r deseq2-comparison}
pair_contrast <- function(id, family, numerator, denominator,
                          numerator_label, denominator_label) {
  list(
    id = id,
    family = family,
    numerator = numerator_label,
    denominator = denominator_label,
    weights = stats::setNames(c(1, -1), c(numerator, denominator))
  )
}

contrast_definitions <- list(
  pair_contrast("infection_diet33", "Infection within diet", "diet33_infected", "diet33_control", "Infected diet 33", "Control diet 33"),
  pair_contrast("infection_diet50", "Infection within diet", "diet50_infected", "diet50_control", "Infected diet 50", "Control diet 50"),
  pair_contrast("infection_diet83", "Infection within diet", "diet83_infected", "diet83_control", "Infected diet 83", "Control diet 83"),
  pair_contrast("control_diet33_vs50", "Diet among controls", "diet33_control", "diet50_control", "Control diet 33", "Control diet 50"),
  pair_contrast("control_diet83_vs50", "Diet among controls", "diet83_control", "diet50_control", "Control diet 83", "Control diet 50"),
  pair_contrast("control_diet83_vs33", "Diet among controls", "diet83_control", "diet33_control", "Control diet 83", "Control diet 33"),
  pair_contrast("infected_diet33_vs50", "Diet among infected", "diet33_infected", "diet50_infected", "Infected diet 33", "Infected diet 50"),
  pair_contrast("infected_diet83_vs50", "Diet among infected", "diet83_infected", "diet50_infected", "Infected diet 83", "Infected diet 50"),
  pair_contrast("infected_diet83_vs33", "Diet among infected", "diet83_infected", "diet33_infected", "Infected diet 83", "Infected diet 33"),
  list(
    id = "infection_average_all_diets",
    family = "Average infection response",
    numerator = "Average infected",
    denominator = "Average control",
    weights = c(
      diet33_infected = 1 / 3, diet50_infected = 1 / 3,
      diet83_infected = 1 / 3, diet33_control = -1 / 3,
      diet50_control = -1 / 3, diet83_control = -1 / 3
    )
  ),
  list(
    id = "interaction_diet33_vs50",
    family = "Diet by infection interaction",
    numerator = "Infection response diet 33",
    denominator = "Infection response diet 50",
    weights = c(
      diet33_infected = 1, diet33_control = -1,
      diet50_infected = -1, diet50_control = 1
    )
  ),
  list(
    id = "interaction_diet83_vs50",
    family = "Diet by infection interaction",
    numerator = "Infection response diet 83",
    denominator = "Infection response diet 50",
    weights = c(
      diet83_infected = 1, diet83_control = -1,
      diet50_infected = -1, diet50_control = 1
    )
  ),
  list(
    id = "interaction_diet83_vs33",
    family = "Diet by infection interaction",
    numerator = "Infection response diet 83",
    denominator = "Infection response diet 33",
    weights = c(
      diet83_infected = 1, diet83_control = -1,
      diet33_infected = -1, diet33_control = 1
    )
  )
)

fit_count_mode <- function(counts, counting_mode) {
  keep <- rowSums(counts >= params$min_count) >= params$min_samples
  analysis_counts <- counts[keep, , drop = FALSE]
  dds <- DESeq2::DESeqDataSetFromMatrix(
    countData = analysis_counts,
    colData = as.data.frame(metadata[sample_ids, , drop = FALSE]),
    design = ~ 0 + group
  )
  dds <- DESeq2::DESeq(
    dds, minReplicatesForReplace = Inf, quiet = TRUE
  )
  coefficient_names <- DESeq2::resultsNames(dds)

  results <- dplyr::bind_rows(lapply(contrast_definitions, function(definition) {
    contrast_vector <- stats::setNames(
      rep(0, length(coefficient_names)), coefficient_names
    )
    requested <- paste0("group", names(definition$weights))
    if (!all(requested %in% coefficient_names)) {
      stop("Missing coefficient(s) for ", definition$id)
    }
    contrast_vector[requested] <- unname(definition$weights)
    as.data.frame(DESeq2::results(
      dds,
      contrast = contrast_vector,
      alpha = params$alpha,
      cooksCutoff = FALSE
    )) |>
      tibble::rownames_to_column("gene_id") |>
      tibble::as_tibble() |>
      dplyr::mutate(
        counting_mode = counting_mode,
        contrast_id = definition$id,
        contrast_family = definition$family,
        numerator = definition$numerator,
        denominator = definition$denominator,
        significant = !is.na(.data$padj) &
          .data$padj < params$alpha &
          abs(.data$log2FoldChange) >= params$lfc_threshold,
        direction = dplyr::case_when(
          !.data$significant ~ "Not significant",
          .data$log2FoldChange > 0 ~ "Higher in numerator",
          TRUE ~ "Higher in denominator"
        )
      )
  }))
  list(dds = dds, results = results)
}

mode_fits <- lapply(names(nuclear_counts), function(mode) {
  fit_count_mode(nuclear_counts[[mode]], mode)
})
names(mode_fits) <- names(nuclear_counts)
all_results <- dplyr::bind_rows(lapply(mode_fits, `[[`, "results"))

deg_counts <- all_results |>
  dplyr::filter(.data$significant) |>
  dplyr::count(
    .data$counting_mode, .data$contrast_family, .data$contrast_id,
    .data$direction, name = "n_DEGs"
  ) |>
  tidyr::complete(
    counting_mode = names(mode_colors),
    tidyr::nesting(contrast_family, contrast_id),
    direction = c("Higher in numerator", "Higher in denominator"),
    fill = list(n_DEGs = 0)
  ) |>
  dplyr::group_by(.data$counting_mode, .data$contrast_family, .data$contrast_id) |>
  dplyr::mutate(total_DEGs = sum(.data$n_DEGs)) |>
  dplyr::ungroup()

write.csv(all_results, file.path(out_dir, "count_definition_all_deseq2_results.csv"),
          row.names = FALSE)
write.csv(deg_counts, file.path(out_dir, "count_definition_deg_counts.csv"),
          row.names = FALSE)
```

## Step 3: compare DEG numbers

```{r deg-count-figure, fig.width=18, fig.height=10}
contrast_labels <- c(
  infection_diet33 = "Diet 33: infected vs control",
  infection_diet50 = "Diet 50: infected vs control",
  infection_diet83 = "Diet 83: infected vs control",
  control_diet33_vs50 = "Controls: diet 33 vs 50",
  control_diet83_vs50 = "Controls: diet 83 vs 50",
  control_diet83_vs33 = "Controls: diet 83 vs 33",
  infected_diet33_vs50 = "Infected: diet 33 vs 50",
  infected_diet83_vs50 = "Infected: diet 83 vs 50",
  infected_diet83_vs33 = "Infected: diet 83 vs 33",
  infection_average_all_diets = "Average infected vs control",
  interaction_diet33_vs50 = "Interaction: diet 33 vs 50",
  interaction_diet83_vs50 = "Interaction: diet 83 vs 50",
  interaction_diet83_vs33 = "Interaction: diet 83 vs 33"
)

deg_totals <- deg_counts |>
  dplyr::distinct(
    .data$counting_mode, .data$contrast_family,
    .data$contrast_id, .data$total_DEGs
  ) |>
  dplyr::mutate(
    counting_mode = factor(.data$counting_mode, levels = names(mode_colors)),
    contrast_label = factor(
      unname(contrast_labels[.data$contrast_id]),
      levels = rev(unname(contrast_labels))
    )
  )

p_deg_counts <- ggplot2::ggplot(
  deg_totals,
  ggplot2::aes(
    x = .data$total_DEGs, y = .data$contrast_label,
    fill = .data$counting_mode
  )
) +
  ggplot2::geom_col(
    position = ggplot2::position_dodge(width = 0.76),
    width = 0.68, color = "#111827", linewidth = 0.25
  ) +
  ggplot2::geom_text(
    ggplot2::aes(label = scales::comma(.data$total_DEGs)),
    position = ggplot2::position_dodge(width = 0.76),
    hjust = -0.12, size = 4.5, fontface = "bold"
  ) +
  ggplot2::scale_fill_manual(values = mode_colors) +
  ggplot2::scale_x_continuous(
    labels = scales::label_comma(),
    expand = ggplot2::expansion(mult = c(0, 0.16))
  ) +
  ggplot2::labs(
    title = "DEG totals under exon-only and transcript-plus-exon counting",
    subtitle = paste0(
      "The 44-sample analysis cohort; nuclear chromosomes; rRNA removed; padj < ",
      params$alpha, "; |log2FC| >= ", params$lfc_threshold
    ),
    x = "Significant genes", y = NULL, fill = "Counting mode"
  ) +
  theme_report(16) +
  ggplot2::theme(
    axis.text.y = ggplot2::element_text(face = "bold", size = 14)
  )

save_png_svg(p_deg_counts, "count_definition_deg_totals", 18, 10)
p_deg_counts
```

```{r deg-overlap}
significant_long <- all_results |>
  dplyr::filter(.data$significant) |>
  dplyr::distinct(.data$counting_mode, .data$contrast_id, .data$gene_id)

overlap_membership <- dplyr::bind_rows(lapply(names(contrast_labels), function(id) {
  exon_ids <- significant_long$gene_id[
    significant_long$contrast_id == id &
      significant_long$counting_mode == "Exon only"
  ]
  transcript_ids <- significant_long$gene_id[
    significant_long$contrast_id == id &
      significant_long$counting_mode == "Transcript + exon"
  ]
  union_ids <- union(exon_ids, transcript_ids)
  tibble::tibble(
    contrast_id = id,
    gene_id = union_ids,
    exon_only_DEG = union_ids %in% exon_ids,
    transcript_exon_DEG = union_ids %in% transcript_ids,
    membership = dplyr::case_when(
      exon_only_DEG & transcript_exon_DEG ~ "Shared",
      exon_only_DEG ~ "Exon only",
      TRUE ~ "Transcript + exon only"
    )
  )
}))

overlap_summary <- overlap_membership |>
  dplyr::count(.data$contrast_id, .data$membership, name = "n_genes") |>
  tidyr::complete(
    contrast_id = names(contrast_labels),
    membership = names(membership_colors),
    fill = list(n_genes = 0)
  ) |>
  dplyr::mutate(
    membership = factor(.data$membership, levels = names(membership_colors)),
    contrast_label = factor(
      unname(contrast_labels[.data$contrast_id]),
      levels = rev(unname(contrast_labels))
    )
  )

write.csv(overlap_membership,
          file.path(out_dir, "count_definition_deg_membership.csv"),
          row.names = FALSE)
write.csv(overlap_summary,
          file.path(out_dir, "count_definition_deg_overlap_summary.csv"),
          row.names = FALSE)
```

```{r deg-overlap-figure, fig.width=18, fig.height=8.5}
p_overlap <- ggplot2::ggplot(
  overlap_summary,
  ggplot2::aes(
    x = .data$n_genes, y = .data$contrast_label,
    fill = .data$membership
  )
) +
  ggplot2::geom_col(width = 0.72, color = "white", linewidth = 0.2) +
  ggplot2::scale_fill_manual(values = membership_colors) +
  ggplot2::scale_x_continuous(labels = scales::label_comma()) +
  ggplot2::labs(
    title = "How counting mode changes DEG membership",
    subtitle = "Shared and mode-specific calls are reported independently for each contrast",
    x = "Genes in the union of both DEG sets", y = NULL,
    fill = "DEG membership"
  ) +
  theme_report(16) +
  ggplot2::theme(axis.text.y = ggplot2::element_text(face = "bold", size = 14))

save_png_svg(p_overlap, "count_definition_deg_membership", 18, 8.5)
p_overlap
```

## Results table

```{r searchable-results-table}
comparison_table <- deg_totals |>
  dplyr::select(
    .data$contrast_family, .data$contrast_id,
    contrast = .data$contrast_label,
    .data$counting_mode, .data$total_DEGs
  ) |>
  tidyr::pivot_wider(
    names_from = .data$counting_mode,
    values_from = .data$total_DEGs
  ) |>
  dplyr::rename(
    exon_only_DEGs = `Exon only`,
    transcript_exon_DEGs = `Transcript + exon`
  ) |>
  dplyr::left_join(
    overlap_summary |>
      dplyr::mutate(membership = as.character(.data$membership)) |>
      tidyr::pivot_wider(
        id_cols = .data$contrast_id,
        names_from = .data$membership,
        values_from = .data$n_genes,
        values_fill = 0
      ) |>
      dplyr::rename(
        exon_only_unique = `Exon only`,
        shared_DEGs = Shared,
        transcript_exon_unique = `Transcript + exon only`
      ),
    by = "contrast_id"
  ) |>
  dplyr::arrange(.data$contrast_family, .data$contrast_id)

write.csv(comparison_table,
          file.path(out_dir, "count_definition_contrast_comparison.csv"),
          row.names = FALSE)

DT::datatable(
  comparison_table,
  rownames = FALSE,
  filter = "top",
  options = list(pageLength = 20, scrollX = TRUE),
  caption = "DEG totals and overlap by counting mode"
)
```

## Step 4: omnibus diet-by-infection interaction

The pairwise difference-of-differences contrasts above ask which two diets have
different infection responses. The omnibus likelihood-ratio test (LRT) asks the
broader question first: does allowing any diet-specific infection response
improve the model relative to additive diet and infection effects?

The two nested models are:

```r
full    = ~ Diet + Treatment + Diet:Treatment
reduced = ~ Diet + Treatment
```

Unlike a one-coefficient Wald test, the omnibus LRT has no single log2 fold
change. We therefore report FDR-significant genes and, for the main-chromosome
analysis, summarize the largest absolute pairwise interaction log2 fold change
as an accompanying effect-size diagnostic.

```{r omnibus-interaction-lrt}
fit_omnibus_interaction <- function(counts, counting_mode, sequence_scope) {
  keep <- rowSums(counts >= params$min_count) >= params$min_samples
  analysis_counts <- counts[keep, , drop = FALSE]

  interaction_metadata <- metadata[sample_ids, , drop = FALSE]
  interaction_metadata$Diet <- stats::relevel(
    factor(interaction_metadata$Diet), ref = "50"
  )
  interaction_metadata$Treatment <- stats::relevel(
    factor(interaction_metadata$Treatment), ref = "Control"
  )

  dds <- DESeq2::DESeqDataSetFromMatrix(
    countData = analysis_counts,
    colData = as.data.frame(interaction_metadata),
    design = ~ Diet + Treatment + Diet:Treatment
  )
  dds <- DESeq2::DESeq(
    dds,
    test = "LRT",
    reduced = ~ Diet + Treatment,
    minReplicatesForReplace = Inf,
    quiet = TRUE
  )

  result <- as.data.frame(DESeq2::results(
    dds,
    alpha = params$alpha,
    cooksCutoff = FALSE
  )) |>
    tibble::rownames_to_column("gene_id") |>
    tibble::as_tibble() |>
    dplyr::mutate(
      counting_mode = counting_mode,
      sequence_scope = sequence_scope,
      omnibus_significant = !is.na(.data$padj) & .data$padj < params$alpha
    )

  list(dds = dds, results = result, genes_entering_model = sum(keep))
}

lrt_inputs <- list(
  "Exon only" = list(
    "Main chromosomes" = nuclear_counts[["Exon only"]],
    "All scaffolds" = rrna_filtered_counts[["Exon only"]]
  ),
  "Transcript + exon" = list(
    "Main chromosomes" = nuclear_counts[["Transcript + exon"]],
    "All scaffolds" = rrna_filtered_counts[["Transcript + exon"]]
  )
)

lrt_fits <- list()
for (mode in names(lrt_inputs)) {
  for (scope in names(lrt_inputs[[mode]])) {
    key <- paste(mode, scope, sep = "__")
    lrt_fits[[key]] <- fit_omnibus_interaction(
      lrt_inputs[[mode]][[scope]], mode, scope
    )
  }
}

lrt_results <- dplyr::bind_rows(lapply(lrt_fits, `[[`, "results"))

safe_max_abs <- function(x) {
  x <- abs(x[is.finite(x)])
  if (length(x) == 0) NA_real_ else max(x)
}

pairwise_interaction_effects <- all_results |>
  dplyr::filter(.data$contrast_family == "Diet by infection interaction") |>
  dplyr::group_by(.data$counting_mode, .data$gene_id) |>
  dplyr::summarise(
    max_abs_pairwise_interaction_log2FC = safe_max_abs(.data$log2FoldChange),
    .groups = "drop"
  )

lrt_results <- lrt_results |>
  dplyr::left_join(
    pairwise_interaction_effects,
    by = c("counting_mode", "gene_id")
  ) |>
  dplyr::left_join(
    placement |>
      dplyr::select(.data$gene_id, .data$sequence_class),
    by = "gene_id"
  )

lrt_summary <- dplyr::bind_rows(lapply(names(lrt_fits), function(key) {
  fit <- lrt_fits[[key]]
  result <- fit$results |>
    dplyr::left_join(
      pairwise_interaction_effects,
      by = c("counting_mode", "gene_id")
    )
  tibble::tibble(
    counting_mode = unique(result$counting_mode),
    sequence_scope = unique(result$sequence_scope),
    genes_entering_model = fit$genes_entering_model,
    omnibus_FDR_lt_0_05 = sum(result$omnibus_significant),
    omnibus_FDR_and_max_abs_pairwise_log2FC_ge_1 = if (
      unique(result$sequence_scope) == "Main chromosomes"
    ) {
      sum(
        result$omnibus_significant &
          !is.na(result$max_abs_pairwise_interaction_log2FC) &
          result$max_abs_pairwise_interaction_log2FC >= params$lfc_threshold
      )
    } else {
      NA_integer_
    }
  )
}))

lrt_placement_summary <- lrt_results |>
  dplyr::filter(
    .data$sequence_scope == "All scaffolds",
    .data$omnibus_significant
  ) |>
  dplyr::mutate(
    sequence_class = dplyr::coalesce(.data$sequence_class, "Unresolved")
  ) |>
  dplyr::count(
    .data$counting_mode, .data$sequence_class,
    name = "omnibus_FDR_lt_0_05"
  ) |>
  dplyr::arrange(.data$counting_mode, dplyr::desc(.data$omnibus_FDR_lt_0_05))

write.csv(
  lrt_results,
  file.path(out_dir, "count_definition_omnibus_interaction_LRT_results.csv"),
  row.names = FALSE
)
write.csv(
  lrt_summary,
  file.path(out_dir, "count_definition_omnibus_interaction_LRT_summary.csv"),
  row.names = FALSE
)
write.csv(
  lrt_placement_summary,
  file.path(out_dir, "count_definition_omnibus_interaction_placement_summary.csv"),
  row.names = FALSE
)

DT::datatable(
  lrt_summary,
  rownames = FALSE,
  options = list(pageLength = 8, scrollX = TRUE),
  caption = "Omnibus diet-by-infection LRT under both count definitions"
)
```

```{r omnibus-interaction-lrt-figure, fig.width=12, fig.height=6.5}
p_lrt <- lrt_summary |>
  dplyr::mutate(
    counting_mode = factor(.data$counting_mode, levels = names(mode_colors)),
    sequence_scope = factor(
      .data$sequence_scope,
      levels = c("Main chromosomes", "All scaffolds")
    )
  ) |>
  ggplot2::ggplot(ggplot2::aes(
    x = .data$sequence_scope,
    y = .data$omnibus_FDR_lt_0_05,
    fill = .data$counting_mode
  )) +
  ggplot2::geom_col(
    position = ggplot2::position_dodge(width = 0.76),
    width = 0.68,
    color = "#111827"
  ) +
  ggplot2::geom_text(
    ggplot2::aes(label = scales::comma(.data$omnibus_FDR_lt_0_05)),
    position = ggplot2::position_dodge(width = 0.76),
    vjust = -0.35,
    size = 5,
    fontface = "bold"
  ) +
  ggplot2::scale_fill_manual(values = mode_colors) +
  ggplot2::scale_y_continuous(
    labels = scales::label_comma(),
    expand = ggplot2::expansion(mult = c(0, 0.16))
  ) +
  ggplot2::labs(
    title = "Formal diet-by-infection interaction is count-definition sensitive",
    subtitle = "Omnibus LRT, Benjamini-Hochberg FDR < 0.05; 44-library analysis cohort",
    x = NULL,
    y = "Genes with an omnibus interaction",
    fill = "Counting mode"
  ) +
  theme_report(16)

save_png_svg(p_lrt, "count_definition_omnibus_interaction_LRT", 12, 6.5)
p_lrt
```

The LRT is interpreted before the pairwise interaction contrasts. A small
omnibus set supports a broad shared infection response with selective
diet-dependent divergence. Differences between counting modes are reported as
sensitivity rather than treated as evidence that one definition is universally
correct.

## Interpretation

The difference between modes should be described as a **count-definition
sensitivity**, not solely as a pre-mRNA effect. Transcript-plus-exon counting
captures substantially more signal and allows more genes to pass the expression
filter. The overlap plot then distinguishes genes supported by both definitions
from genes whose statistical call depends on including transcript-span signal.

The rRNA audit above applies before DESeq2 in both branches. A non-zero
`rRNA_rows_remaining` value would stop the biological interpretation and require
the exclusion list or count-matrix provenance to be corrected.

## Reproducibility

```{r session-info}
writeLines(capture.output(sessionInfo()), file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
