---
title: "Exon-only curated target genes"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  workbook_file: "data/external/curated_target_genes/2026-08-06_curated-hypoth_DEGs_v21.xlsx"
  metadata_file: "data/metadata/mehreen_metadata_analysis_44_excluding_1044.txt"
  frozen_gene_audit_file: "output/rmd_runs/curated-target-gene-figure_20260806-021426/figure4_curated_gene_audit.csv"
  placement_file: "output/runs/host_pathogen_dual_20260727_000952/local-analysis/host-mapping-audit_20260729-163928/unified-model/audit-summary/host_gene_sequence_placement.tsv"
  results_file: "output/rmd_runs/exon-only-deseq2-analysis_20260809-234532/placed_unplaced_all_deseq2_results.csv"
  homology_results_pattern: "^scaffold-origin-audit_"
  alpha: 0.05
  lfc_threshold: 1
  output_tag: "exon-only-curated-target-genes"
---

## Purpose

This page follows a fixed set of immune and protein-anabolism genes through two
matched views: genes on the main *Schistocerca gregaria* chromosomes and the
complete all-scaffold set used in the original curated figure. The biological
categories come from the supplied workbook, while every differential-expression
status is recalculated from the current competitive-host DESeq2 results.

::: {.walkthrough-grid}
::: {.walkthrough-step}
**Step 1 - Match genes**

Join the curated IDs to the GFF annotation and official assembly placement.
:::
::: {.walkthrough-step}
**Step 2 - Recalculate status**

Apply the same DEG thresholds to every target and contrast.
:::
::: {.walkthrough-step}
**Step 3 - Compare scopes**

Read the main-chromosome and complete all-scaffold panels side by side.
:::
:::

A target is shown as `N.S.` when it does not meet both `padj < alpha` and
`|log2FC| >= lfc_threshold`. Keeping the target set fixed makes changes in the
category percentages directly comparable across contrasts and placement
classes.

The label below each facet reports the number of **curated target genes in the
denominator**, not the number of DEGs. Differential-expression status is
recalculated from the current exon-only model, so a target can move
into or out of significance even though restricting the target list to main
chromosomes can only reduce its size.

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required_packages <- c(
  "readxl", "readr", "dplyr", "tidyr", "tibble", "ggplot2",
  "patchwork", "cowplot", "scales", "DT", "knitr"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages) > 0) {
  stop(
    "Install the following R packages before running this page: ",
    paste(missing_packages, collapse = ", ")
  )
}

suppressPackageStartupMessages({
  library(readxl)
  library(readr)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(patchwork)
})

page_defaults <- list(
  workbook_file = paste0(
    "data/external/curated_target_genes/",
    "2026-08-06_curated-hypoth_DEGs_v21.xlsx"
  ),
  metadata_file = "data/metadata/mehreen_metadata_analysis_44_excluding_1044.txt",
  frozen_gene_audit_file = paste0(
    "output/rmd_runs/curated-target-gene-figure_20260806-021426/",
    "figure4_curated_gene_audit.csv"
  ),
  placement_file = paste0(
    "output/runs/host_pathogen_dual_20260727_000952/local-analysis/",
    "host-mapping-audit_20260729-163928/unified-model/audit-summary/",
    "host_gene_sequence_placement.tsv"
  ),
  results_file = paste0(
    "output/rmd_runs/exon-only-deseq2-analysis_20260809-234532/",
    "placed_unplaced_all_deseq2_results.csv"
  ),
  latest_results_root = "output/rmd_runs",
  homology_results_pattern = "^scaffold-origin-audit_",
  alpha = 0.05,
  lfc_threshold = 1,
  output_tag = "exon-only-curated-target-genes"
)

if (!exists("params", inherits = TRUE) || !is.list(params)) {
  page_params <- page_defaults
} else {
  page_params <- utils::modifyList(page_defaults, params)
}

path_ancestors <- function(path) {
  path <- normalizePath(path, mustWork = FALSE)
  ancestors <- path
  repeat {
    parent <- dirname(path)
    if (identical(parent, path)) break
    ancestors <- c(ancestors, parent)
    path <- parent
  }
  ancestors
}

is_project_root <- function(path) {
  file.exists(file.path(path, "analysis", "30-curated-target-gene-figure.Rmd")) &&
    dir.exists(file.path(path, "data")) &&
    dir.exists(file.path(path, "output"))
}

find_project_root <- function() {
  starts <- c(getwd())
  explicit_root <- Sys.getenv("GREGARIA_DIET_PROJECT_DIR", unset = "")
  if (nzchar(explicit_root)) starts <- c(explicit_root, starts)

  input_file <- tryCatch(
    knitr::current_input(dir = TRUE),
    error = function(e) ""
  )
  if (nzchar(input_file)) starts <- c(dirname(input_file), starts)

  candidates <- unique(unlist(lapply(starts, path_ancestors)))
  hits <- candidates[vapply(candidates, is_project_root, logical(1))]
  if (length(hits) == 0) {
    stop("Could not locate the gregaria-diet-infection-interaction project root.")
  }
  hits[[1]]
}

project_dir <- find_project_root()
resolve_path <- function(path) {
  if (grepl("^/", path)) path else file.path(project_dir, path)
}

latest_matching_dir <- function(root, pattern) {
  root <- resolve_path(root)
  candidates <- list.dirs(root, recursive = FALSE, full.names = TRUE)
  candidates <- candidates[grepl(pattern, basename(candidates))]
  if (length(candidates) == 0) {
    stop("No result folder matched ", pattern, " under ", root)
  }
  sort(candidates)[[length(candidates)]]
}

make_run_dir <- function(root, tag) {
  stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
  base <- file.path(resolve_path(root), paste0(tag, "_", stamp))
  candidate <- base
  suffix <- 1L
  while (dir.exists(candidate)) {
    suffix <- suffix + 1L
    candidate <- paste0(base, "_", suffix)
  }
  dir.create(candidate, recursive = TRUE, showWarnings = FALSE)
  candidate
}

workbook_file <- resolve_path(page_params$workbook_file)
metadata_file <- resolve_path(page_params$metadata_file)
frozen_gene_audit_file <- resolve_path(page_params$frozen_gene_audit_file)
placement_file <- resolve_path(page_params$placement_file)
results_file <- resolve_path(page_params$results_file)
source_results_dir <- dirname(results_file)
homology_results_dir <- latest_matching_dir(
  page_params$latest_results_root,
  page_params$homology_results_pattern
)
homology_evidence_file <- file.path(
  homology_results_dir,
  "corrected_gene_homology_evidence.tsv"
)
out_dir <- make_run_dir(page_params$latest_results_root, page_params$output_tag)

required_files <- c(
  metadata_file, placement_file, results_file, homology_evidence_file
)
missing_files <- required_files[!file.exists(required_files)]
if (length(missing_files) > 0) {
  stop("Missing required input files: ", paste(missing_files, collapse = ", "))
}
if (!file.exists(workbook_file) && !file.exists(frozen_gene_audit_file)) {
  stop(
    "Neither the curated workbook nor its frozen gene-level audit is available: ",
    workbook_file, "; ", frozen_gene_audit_file
  )
}

alpha <- as.numeric(page_params$alpha)
lfc_threshold <- as.numeric(page_params$lfc_threshold)

metadata_audit <- read.delim(metadata_file, check.names = FALSE)
metadata_audit <- metadata_audit[
  , !is.na(names(metadata_audit)) & nzchar(names(metadata_audit)),
  drop = FALSE
]
expected_assignments <- data.frame(
  label = "1024",
  Diet = "33",
  Treatment = "Infected"
)
observed_assignments <- metadata_audit[
  match(expected_assignments$label, as.character(metadata_audit$label)),
  c("label", "Diet", "Treatment"),
  drop = FALSE
]
observed_assignments[] <- lapply(observed_assignments, as.character)
rownames(observed_assignments) <- NULL
if (nrow(metadata_audit) != 44L || "1044" %in% metadata_audit$label ||
    anyNA(observed_assignments$label) ||
    !identical(observed_assignments, expected_assignments)) {
  stop("Figure 4 requires the frozen 44-sample cohort with 1044 excluded.")
}

write_csv_out <- function(x, filename) {
  readr::write_csv(x, file.path(out_dir, filename), na = "")
}

save_figure <- function(plot, stem, width, height) {
  ggplot2::ggsave(
    file.path(out_dir, paste0(stem, ".png")), plot,
    width = width, height = height, units = "in", dpi = 400,
    bg = "white"
  )
  ggplot2::ggsave(
    file.path(out_dir, paste0(stem, ".svg")), plot,
    width = width, height = height, units = "in", bg = "white"
  )
  ggplot2::ggsave(
    file.path(out_dir, paste0(stem, ".pdf")), plot,
    width = width, height = height, units = "in",
    device = grDevices::cairo_pdf, bg = "white"
  )
}
```

## Inputs and contrast direction

The v21 workbook supplies the final curated gene membership and biological
categories. Its stored fold changes are retained only for reconciliation; the
figure itself recalculates every status from the current exon-only,
competitive-host DESeq2 analysis. The main paper view then retains only genes
assigned to nuclear chromosomes by the official assembly placement table.

Within each diet, positive values mean higher expression in infected locusts.
For across-diet comparisons, the sign is oriented to the higher-protein diet:
50 versus 33, 83 versus 33, and 83 versus 50. Thus, `Upregulated` always means
higher expression in the first condition printed below the bar.

```{r read-inputs}
if (file.exists(workbook_file)) {
  curated_targets <- dplyr::bind_rows(
    readxl::read_excel(workbook_file, sheet = "Immune") |>
      dplyr::mutate(source_sheet = "Immune"),
    readxl::read_excel(workbook_file, sheet = "Protein-Anabolism") |>
      dplyr::mutate(source_sheet = "Protein-Anabolism")
  ) |>
    dplyr::rename(
      umbrella = `Umbrella Category`,
      subcategory = `Sub-Category`,
      gene_id = `Gene ID`,
      description_workbook = `Gene Description`
    )
} else {
  # The frozen audit preserves the exact v21 gene membership and categories.
  # Fold changes are always recalculated from the current DESeq2 tables, so the
  # missing workbook affects only historical status reconciliation.
  curated_targets <- readr::read_csv(
    frozen_gene_audit_file,
    show_col_types = FALSE
  ) |>
    dplyr::distinct(
      .data$source_sheet, .data$umbrella, .data$subcategory,
      .data$gene_id, .data$description_workbook
    )

  historical_columns <- c(
    "Diet 33 log2FC", "Diet 33 padj",
    "Diet 50 log2FC", "Diet 50 padj",
    "Diet 83 log2FC", "Diet 83 padj",
    "inf33 vs inf50_log2FC", "inf33 vs inf50_padj",
    "inf83 vs inf33_log2FC", "inf83 vs inf33_padj",
    "inf83 vs inf50_log2FC", "inf83 vs inf50_padj",
    "cont33 vs cont50_log2FC", "cont33 vs cont50_padj",
    "cont83 vs cont33_log2FC", "cont83 vs cont33_padj",
    "cont83 vs cont50_log2FC", "cont83 vs cont50_padj"
  )
  for (column in historical_columns) curated_targets[[column]] <- NA_character_
}

if (anyDuplicated(curated_targets$gene_id)) {
  stop("The curated workbook contains duplicated gene IDs.")
}

placement <- readr::read_tsv(placement_file, show_col_types = FALSE)
all_results <- readr::read_csv(results_file, show_col_types = FALSE) |>
  dplyr::filter(.data$mapping_branch == "Competitive host")
homology_evidence <- readr::read_tsv(
  homology_evidence_file,
  show_col_types = FALSE
)

if (anyDuplicated(homology_evidence$gene_id)) {
  stop("The scaffold-origin homology audit contains duplicated gene IDs.")
}

contrast_map <- tibble::tribble(
  ~contrast_id, ~contrast_label, ~figure_scope, ~x_label, ~x_order,
  ~desired_sign, ~positive_condition, ~negative_condition,
  ~workbook_lfc_col, ~workbook_padj_col, ~show_in_figure,
  "infection_diet33", "Diet 33: infected vs control",
  "Within-diet infection response", "33", 1, 1,
  "Infected diet 33", "Control diet 33",
  "Diet 33 log2FC", "Diet 33 padj", TRUE,
  "infection_diet50", "Diet 50: infected vs control",
  "Within-diet infection response", "50", 2, 1,
  "Infected diet 50", "Control diet 50",
  "Diet 50 log2FC", "Diet 50 padj", TRUE,
  "infection_diet83", "Diet 83: infected vs control",
  "Within-diet infection response", "83", 3, 1,
  "Infected diet 83", "Control diet 83",
  "Diet 83 log2FC", "Diet 83 padj", TRUE,
  "infected_diet33_vs50", "Infected: diet 50 vs 33",
  "Diet comparison among infected", "50\nvs\n33", 1, -1,
  "Infected diet 50", "Infected diet 33",
  "inf33 vs inf50_log2FC", "inf33 vs inf50_padj", TRUE,
  "infected_diet83_vs33", "Infected: diet 83 vs 33",
  "Diet comparison among infected", "83\nvs\n33", 2, 1,
  "Infected diet 83", "Infected diet 33",
  "inf83 vs inf33_log2FC", "inf83 vs inf33_padj", TRUE,
  "infected_diet83_vs50", "Infected: diet 83 vs 50",
  "Diet comparison among infected", "83\nvs\n50", 3, 1,
  "Infected diet 83", "Infected diet 50",
  "inf83 vs inf50_log2FC", "inf83 vs inf50_padj", TRUE,
  "control_diet33_vs50", "Control: diet 50 vs 33",
  "Diet comparison among controls", "50\nvs\n33", 1, -1,
  "Control diet 50", "Control diet 33",
  "cont33 vs cont50_log2FC", "cont33 vs cont50_padj", FALSE,
  "control_diet83_vs33", "Control: diet 83 vs 33",
  "Diet comparison among controls", "83\nvs\n33", 2, 1,
  "Control diet 83", "Control diet 33",
  "cont83 vs cont33_log2FC", "cont83 vs cont33_padj", FALSE,
  "control_diet83_vs50", "Control: diet 83 vs 50",
  "Diet comparison among controls", "83\nvs\n50", 3, 1,
  "Control diet 83", "Control diet 50",
  "cont83 vs cont50_log2FC", "cont83 vs cont50_padj", FALSE
)

target_placement <- curated_targets |>
  dplyr::left_join(
    placement |>
      dplyr::select(
        gene_id, seqid, description_gff = description, gene_biotype,
        sequence_name, sequence_role, assigned_molecule, location_type,
        sequence_class
      ),
    by = "gene_id"
  )

if (any(is.na(target_placement$sequence_class))) {
  stop(
    "Placement was not found for: ",
    paste(target_placement$gene_id[is.na(target_placement$sequence_class)],
          collapse = ", ")
  )
}

current_target_results <- all_results |>
  dplyr::filter(.data$contrast_id %in% contrast_map$contrast_id) |>
  dplyr::select(
    gene_id, analysis_scope, contrast_id, baseMean, log2FoldChange,
    lfcSE, stat, pvalue, padj
  )

duplicate_result_rows <- current_target_results |>
  dplyr::count(.data$gene_id, .data$contrast_id) |>
  dplyr::filter(.data$n > 1)
if (nrow(duplicate_result_rows) > 0) {
  stop("More than one competitive-host result was found per gene and contrast.")
}

target_analysis <- target_placement |>
  dplyr::select(
    source_sheet, umbrella, subcategory, gene_id, description_workbook,
    description_gff, gene_biotype, seqid, sequence_name, sequence_role,
    assigned_molecule, location_type, sequence_class
  ) |>
  tidyr::crossing(contrast_map) |>
  dplyr::left_join(
    current_target_results,
    by = c("gene_id", "contrast_id")
  ) |>
  dplyr::mutate(
    significant = !is.na(.data$padj) &
      .data$padj < alpha &
      abs(.data$log2FoldChange) >= lfc_threshold,
    oriented_log2FC = .data$log2FoldChange * .data$desired_sign,
    status = dplyr::case_when(
      !.data$significant ~ "N.S.",
      .data$oriented_log2FC > 0 ~ "Upregulated",
      TRUE ~ "Downregulated"
    ),
    higher_expression_condition = dplyr::case_when(
      !.data$significant ~ "Not significant",
      .data$oriented_log2FC > 0 ~ .data$positive_condition,
      TRUE ~ .data$negative_condition
    ),
    primary_figure = .data$sequence_class == "Nuclear chromosome"
  )

original_status_long <- dplyr::bind_rows(lapply(
  seq_len(nrow(contrast_map)),
  function(i) {
    original_value <- as.character(
      curated_targets[[contrast_map$workbook_lfc_col[[i]]]]
    )
    numeric_value <- suppressWarnings(as.numeric(original_value))
    tibble::tibble(
      gene_id = curated_targets$gene_id,
      contrast_id = contrast_map$contrast_id[[i]],
      original_value = original_value,
      original_oriented_log2FC = numeric_value * contrast_map$desired_sign[[i]],
      original_status = dplyr::case_when(
        is.na(numeric_value) ~ "N.S.",
        numeric_value * contrast_map$desired_sign[[i]] > 0 ~ "Upregulated",
        numeric_value * contrast_map$desired_sign[[i]] < 0 ~ "Downregulated",
        TRUE ~ "N.S."
      )
    )
  }
))

target_analysis <- target_analysis |>
  dplyr::left_join(
    original_status_long,
    by = c("gene_id", "contrast_id")
  )

if (nrow(target_analysis) != dplyr::n_distinct(target_analysis$gene_id) *
    dplyr::n_distinct(target_analysis$contrast_id)) {
  stop("The curated audit is not one row per gene and contrast.")
}

gene_audit <- target_analysis |>
  dplyr::group_by(
    .data$source_sheet, .data$umbrella, .data$subcategory,
    .data$gene_id, .data$description_workbook, .data$description_gff,
    .data$gene_biotype, .data$seqid, .data$sequence_name,
    .data$sequence_role, .data$sequence_class
  ) |>
  dplyr::summarise(
    original_significant_contrasts = sum(.data$original_status != "N.S."),
    current_significant_contrasts = sum(.data$status != "N.S."),
    newly_significant_contrasts = sum(
      .data$original_status == "N.S." & .data$status != "N.S."
    ),
    no_longer_significant_contrasts = sum(
      .data$original_status != "N.S." & .data$status == "N.S."
    ),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    figure_scope = dplyr::if_else(
      .data$sequence_class == "Nuclear chromosome",
      "Main chromosome",
      "Unplaced scaffold"
    )
  )

summarize_reconciliation <- function(data, sequence_scope) {
  data |>
    dplyr::filter(.data$show_in_figure) |>
    dplyr::group_by(
      .data$source_sheet, .data$umbrella, .data$figure_scope
    ) |>
    dplyr::summarise(
      sequence_scope = sequence_scope,
      target_genes = dplyr::n_distinct(.data$gene_id),
      gene_contrast_cells = dplyr::n(),
      original_significant_cells = sum(.data$original_status != "N.S."),
      current_significant_cells = sum(.data$status != "N.S."),
      newly_significant_cells = sum(
        .data$original_status == "N.S." & .data$status != "N.S."
      ),
      no_longer_significant_cells = sum(
        .data$original_status != "N.S." & .data$status == "N.S."
      ),
      net_change = .data$current_significant_cells -
        .data$original_significant_cells,
      .groups = "drop"
    ) |>
    dplyr::relocate(.data$sequence_scope)
}

count_reconciliation <- dplyr::bind_rows(
  summarize_reconciliation(target_analysis, "All curated targets"),
  summarize_reconciliation(
    dplyr::filter(target_analysis, .data$primary_figure),
    "Main chromosome"
  ),
  summarize_reconciliation(
    dplyr::filter(target_analysis, !.data$primary_figure),
    "Unplaced scaffold"
  )
)

contrast_audit <- target_analysis |>
  dplyr::select(
    .data$source_sheet, .data$umbrella, .data$subcategory,
    .data$gene_id, .data$description_workbook, .data$description_gff,
    .data$gene_biotype, .data$seqid, .data$sequence_class,
    .data$contrast_id, .data$contrast_label, .data$figure_scope,
    .data$positive_condition, .data$negative_condition,
    .data$baseMean, .data$log2FoldChange, .data$oriented_log2FC,
    .data$padj, .data$status, .data$higher_expression_condition,
    .data$original_status, .data$original_value,
    .data$show_in_figure
  )

write_csv_out(gene_audit, "figure4_curated_gene_audit.csv")
write_csv_out(count_reconciliation, "figure4_count_reconciliation.csv")
write_csv_out(contrast_audit, "figure4_contrast_audit.csv")
write_csv_out(
  dplyr::filter(contrast_audit, .data$sequence_class == "Nuclear chromosome"),
  "figure4_main_chromosome_contrast_audit.csv"
)
write_csv_out(
  dplyr::filter(contrast_audit, .data$sequence_class == "Unplaced scaffold"),
  "figure4_unplaced_scaffold_contrast_audit.csv"
)
```

## Effect of the chromosome restriction

```{r placement-summary}
target_gene_summary <- target_analysis |>
  dplyr::group_by(
    .data$source_sheet, .data$umbrella, .data$gene_id,
    .data$description_workbook, .data$description_gff,
    .data$gene_biotype, .data$seqid, .data$sequence_class
  ) |>
  dplyr::summarise(
    currently_significant = any(.data$significant),
    n_current_significant_contrasts = sum(.data$significant),
    .groups = "drop"
  )

category_impact <- target_gene_summary |>
  dplyr::group_by(.data$source_sheet, .data$umbrella) |>
  dplyr::summarise(
    original_targets = dplyr::n(),
    main_chromosome_targets = sum(.data$sequence_class == "Nuclear chromosome"),
    unplaced_targets = sum(.data$sequence_class == "Unplaced scaffold"),
    retained_percent = 100 * .data$main_chromosome_targets / .data$original_targets,
    main_chromosome_targets_significant_now = sum(
      .data$sequence_class == "Nuclear chromosome" &
        .data$currently_significant
    ),
    .groups = "drop"
  )

total_targets <- nrow(target_gene_summary)
main_targets <- sum(target_gene_summary$sequence_class == "Nuclear chromosome")
unplaced_targets_n <- sum(target_gene_summary$sequence_class == "Unplaced scaffold")
main_targets_significant <- sum(
  target_gene_summary$sequence_class == "Nuclear chromosome" &
    target_gene_summary$currently_significant
)

placement_interpretation <- if (unplaced_targets_n == 0) {
  paste0(
    "All targets in the supplied v21 workbook pass the main-chromosome ",
    "restriction. Differences from the earlier Figure 4 therefore reflect ",
    "the corrected sample assignments and the current count definition, ",
    "rather than removal of curated genes by sequence placement."
  )
} else {
  paste0(
    main_targets, " of ", total_targets,
    " curated genes pass the main-chromosome restriction; ",
    unplaced_targets_n,
    " remain available in the all-scaffold audit but are excluded from the ",
    "primary manuscript figure."
  )
}

write_csv_out(category_impact, "curated_target_placement_summary.csv")
write_csv_out(target_analysis, "curated_target_gene_reanalysis_long.csv")
```

Of the `r total_targets` curated genes, **`r main_targets` are on main nuclear
chromosomes** and enter the revised primary figure. The remaining
**`r unplaced_targets_n` genes are on unplaced scaffolds**. Among the retained
targets, `r main_targets_significant` are currently significant in at least one
of the nine reconstructed contrasts.

There is no double counting: the workbook contains `r total_targets` unique gene
IDs, and the audit contains exactly one row per gene and contrast. For the six
contrasts displayed in Figure 4, the main-chromosome subset contains exactly
`r main_targets * 6` gene-by-contrast cells.

`r placement_interpretation`

```{r target-placement-table}
knitr::kable(
  category_impact |>
    dplyr::mutate(
      retained_percent = scales::percent(.data$retained_percent / 100,
                                         accuracy = 0.1)
    ),
  col.names = c(
    "Group", "Curated category", "Original targets", "Main chromosome",
    "Unplaced", "Retained", "Main-chromosome targets significant now"
  ),
  align = c("l", "l", "r", "r", "r", "r", "r")
)
```

## DEG count reconciliation

This table separates three quantities that can otherwise be confused. Target
genes are the curated denominator; significant cells count one gene in one
contrast; newly significant and no-longer-significant cells describe changes
caused by recalculating the model. The counts above provide the current v21
target totals and the exact effect of the chromosome restriction without
hard-coding values from an older workbook version.

```{r export-exon-only-audit-download, include=FALSE}
audit_download_dir <- file.path(project_dir, "docs", "assets", "downloads")
dir.create(audit_download_dir, recursive = TRUE, showWarnings = FALSE)
readr::write_csv(
  gene_audit,
  file.path(
    audit_download_dir,
    "figure4_curated_DEG_audit_exon_only.csv"
  )
)
```

[Download the exon-only gene-level Figure 4 audit](assets/downloads/figure4_curated_DEG_audit_exon_only.csv).

```{r count-reconciliation-table}
knitr::kable(
  count_reconciliation |>
    dplyr::arrange(
      factor(
        .data$sequence_scope,
        levels = c(
          "All curated targets", "Main chromosome", "Unplaced scaffold"
        )
      ),
      .data$source_sheet, .data$umbrella, .data$figure_scope
    ),
  col.names = c(
    "Sequence scope", "Group", "Category", "Comparison group",
    "Target genes", "Gene-contrast cells", "Earlier significant",
    "Current significant", "Newly significant", "No longer significant",
    "Net change"
  ),
  align = c("l", "l", "l", "l", rep("r", 7))
)
```

```{r target-placement-plot, fig.width=12, fig.height=7.5}
category_order <- c(
  "Innate Immune Recognition & Signaling",
  "Humoral Immunity",
  "Cellular Immunity",
  "Detoxification",
  "Protein Synthesis & Translation",
  "Nutrient Sensing & TOR/Insulin Signaling",
  "Nitrogen, Amino Acid Transport & Metabolism",
  "Nitrogen & Amino Acid Metabolism",
  "Storage Hexamerins & Reserves",
  "Total Immune",
  "Total Protein Anabolism"
)

category_labels <- c(
  "Innate Immune Recognition & Signaling" =
    "Innate immune recognition & signaling",
  "Humoral Immunity" = "Humoral immunity",
  "Cellular Immunity" = "Cellular immunity",
  "Detoxification" = "Detoxification",
  "Protein Synthesis & Translation" = "Protein synthesis & translation",
  "Nutrient Sensing & TOR/Insulin Signaling" =
    "Nutrient sensing & TOR/insulin signaling",
  "Nitrogen, Amino Acid Transport & Metabolism" =
    "Nitrogen, amino acid transport & metabolism",
  "Nitrogen & Amino Acid Metabolism" =
    "Nitrogen & amino acid metabolism",
  "Storage Hexamerins & Reserves" = "Storage hexamerins & reserves"
)

placement_plot_data <- target_gene_summary |>
  dplyr::mutate(
    placement_for_figure = dplyr::if_else(
      .data$sequence_class == "Nuclear chromosome",
      "Main chromosome - retained",
      "Unplaced scaffold - audit only"
    )
  ) |>
  dplyr::count(
    .data$source_sheet, .data$umbrella, .data$placement_for_figure,
    name = "n_genes"
  ) |>
  dplyr::group_by(.data$source_sheet, .data$umbrella) |>
  tidyr::complete(
    placement_for_figure = c(
      "Main chromosome - retained", "Unplaced scaffold - audit only"
    ),
    fill = list(n_genes = 0)
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    umbrella = factor(.data$umbrella, levels = rev(category_order)),
    source_sheet = dplyr::recode(
      .data$source_sheet,
      "Protein-Anabolism" = "Protein anabolism"
    )
  )

placement_impact_plot <- ggplot(
  placement_plot_data,
  aes(x = n_genes, y = umbrella, fill = placement_for_figure)
) +
  geom_col(width = 0.72, color = "white", linewidth = 0.4) +
  geom_text(
    aes(label = ifelse(n_genes > 0, n_genes, "")),
    position = position_stack(vjust = 0.5),
    color = "white", fontface = "bold", size = 4.2
  ) +
  facet_grid(source_sheet ~ ., scales = "free_y", space = "free_y") +
  scale_fill_manual(
    values = c(
      "Main chromosome - retained" = "#2F7D5B",
      "Unplaced scaffold - audit only" = "#B85C4B"
    ),
    breaks = c(
      "Main chromosome - retained", "Unplaced scaffold - audit only"
    )
  ) +
  scale_y_discrete(labels = category_labels) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Curated target genes retained for primary Figure 4",
    subtitle = paste0(
      main_targets, " of ", total_targets,
      " targets are on main chromosomes; unplaced targets remain in the audit"
    ),
    x = "Number of curated target genes",
    y = NULL,
    fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12),
    axis.text.y = element_text(size = 12, color = "black"),
    axis.text.x = element_text(size = 11, color = "black"),
    strip.text.y = element_text(face = "bold", size = 13, angle = 0),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.text = element_text(size = 12),
    plot.margin = margin(12, 18, 12, 12)
  )

placement_impact_plot
save_figure(placement_impact_plot, "curated_target_placement_impact", 12, 7.5)
```

## Curated target figures {#Curated_target_figures}

The curated workbook contains candidate genes selected because they were
significant in at least one differential-expression contrast. The figures below
therefore summarize the direction and frequency of responses within this
preselected set. They are descriptive visualizations rather than independent
tests of whether a functional category is enriched for differential expression.

```{r build-revised-figure4, fig.width=15, fig.height=11.5, results='hide', fig.show='hide'}
figure_contrasts <- contrast_map |>
  dplyr::filter(.data$show_in_figure)

figure_gene_data <- target_analysis |>
  dplyr::filter(
    .data$primary_figure,
    .data$show_in_figure
  )

figure_counts <- figure_gene_data |>
  dplyr::count(
    .data$source_sheet, .data$umbrella, .data$figure_scope,
    .data$x_label, .data$x_order, .data$status,
    name = "n_genes"
  ) |>
  dplyr::group_by(
    .data$source_sheet, .data$umbrella, .data$figure_scope,
    .data$x_label, .data$x_order
  ) |>
  dplyr::mutate(percentage = 100 * .data$n_genes / sum(.data$n_genes)) |>
  dplyr::ungroup()

category_n <- target_gene_summary |>
  dplyr::filter(.data$sequence_class == "Nuclear chromosome") |>
  dplyr::count(.data$source_sheet, .data$umbrella, name = "n_targets")

format_test_label <- function(p) {
  if (is.na(p) || p >= 0.05) return("N.S.D.")
  if (p < 0.0001) return("p < 0.0001")
  if (p < 0.001) return("p < 0.001")
  paste0("p = ", formatC(p, digits = 3, format = "f"))
}

calculate_figure_tests <- function(count_data) {
  empty_result <- tibble::tibble(
    source_sheet = character(),
    umbrella = character(),
    figure_scope = character(),
    p_value = double(),
    p_label = character()
  )
  if (nrow(count_data) == 0) return(empty_result)

  set.seed(20260805)
  test_rows <- list()
  row_i <- 0L
  for (sheet_name in unique(count_data$source_sheet)) {
    for (category_name in unique(
      count_data$umbrella[count_data$source_sheet == sheet_name]
    )) {
      for (scope_name in unique(count_data$figure_scope)) {
        test_df <- count_data |>
          dplyr::filter(
            .data$source_sheet == sheet_name,
            .data$umbrella == category_name,
            .data$figure_scope == scope_name
          ) |>
          tidyr::complete(
            x_order = 1:3,
            status = c("Upregulated", "N.S.", "Downregulated"),
            fill = list(n_genes = 0)
          ) |>
          dplyr::arrange(.data$x_order, match(
            .data$status,
            c("Upregulated", "N.S.", "Downregulated")
          ))

        count_matrix <- matrix(test_df$n_genes, nrow = 3, byrow = TRUE)
        count_matrix <- count_matrix[, colSums(count_matrix) > 0, drop = FALSE]
        p_value <- NA_real_
        if (ncol(count_matrix) >= 2 && nrow(count_matrix) >= 2) {
          p_value <- suppressWarnings(stats::chisq.test(
            count_matrix,
            simulate.p.value = any(count_matrix < 5),
            B = 20000
          )$p.value)
        }

        row_i <- row_i + 1L
        test_rows[[row_i]] <- tibble::tibble(
          source_sheet = sheet_name,
          umbrella = category_name,
          figure_scope = scope_name,
          p_value = p_value,
          p_label = format_test_label(p_value)
        )
      }
    }
  }
  if (length(test_rows) == 0) empty_result else dplyr::bind_rows(test_rows)
}

figure_tests <- calculate_figure_tests(figure_counts)

figure_status_counts <- figure_counts |>
  dplyr::left_join(category_n, by = c("source_sheet", "umbrella")) |>
  dplyr::left_join(
    figure_tests,
    by = c("source_sheet", "umbrella", "figure_scope")
  )

write_csv_out(figure_status_counts, "figure4_main_chromosome_status_counts.csv")
write_csv_out(figure_tests, "figure4_exploratory_chisquare_tests.csv")

build_total_status_counts <- function(gene_data, gene_summary) {
  total_counts <- gene_data |>
    dplyr::mutate(
      umbrella = dplyr::recode(
        .data$source_sheet,
        "Immune" = "Total Immune",
        "Protein-Anabolism" = "Total Protein Anabolism"
      )
    ) |>
    dplyr::count(
      .data$source_sheet, .data$umbrella, .data$figure_scope,
      .data$x_label, .data$x_order, .data$status,
      name = "n_genes"
    ) |>
    dplyr::group_by(
      .data$source_sheet, .data$umbrella, .data$figure_scope,
      .data$x_label, .data$x_order
    ) |>
    dplyr::mutate(percentage = 100 * .data$n_genes / sum(.data$n_genes)) |>
    dplyr::ungroup()

  total_n <- gene_summary |>
    dplyr::mutate(
      umbrella = dplyr::recode(
        .data$source_sheet,
        "Immune" = "Total Immune",
        "Protein-Anabolism" = "Total Protein Anabolism"
      )
    ) |>
    dplyr::count(.data$source_sheet, .data$umbrella, name = "n_targets")

  total_tests <- calculate_figure_tests(total_counts)
  total_counts |>
    dplyr::left_join(total_n, by = c("source_sheet", "umbrella")) |>
    dplyr::left_join(
      total_tests,
      by = c("source_sheet", "umbrella", "figure_scope")
    )
}

figure_total_status_counts <- build_total_status_counts(
  figure_gene_data,
  dplyr::filter(
    target_gene_summary,
    .data$sequence_class == "Nuclear chromosome"
  )
)
write_csv_out(
  figure_total_status_counts,
  "figure4_main_chromosome_total_status_counts.csv"
)

all_scaffold_counts <- target_analysis |>
  dplyr::filter(.data$show_in_figure) |>
  dplyr::count(
    .data$source_sheet, .data$umbrella, .data$figure_scope,
    .data$x_label, .data$x_order, .data$status,
    name = "n_genes"
  ) |>
  dplyr::group_by(
    .data$source_sheet, .data$umbrella, .data$figure_scope,
    .data$x_label, .data$x_order
  ) |>
  dplyr::mutate(percentage = 100 * .data$n_genes / sum(.data$n_genes)) |>
  dplyr::ungroup()

all_scaffold_category_n <- target_gene_summary |>
  dplyr::count(.data$source_sheet, .data$umbrella, name = "n_targets")

all_scaffold_tests <- calculate_figure_tests(all_scaffold_counts)
all_scaffold_status_counts <- all_scaffold_counts |>
  dplyr::left_join(
    all_scaffold_category_n,
    by = c("source_sheet", "umbrella")
  ) |>
  dplyr::left_join(
    all_scaffold_tests,
    by = c("source_sheet", "umbrella", "figure_scope")
  )

write_csv_out(
  all_scaffold_status_counts,
  "figure4_all_scaffolds_status_counts.csv"
)
write_csv_out(
  all_scaffold_tests,
  "figure4_all_scaffolds_exploratory_chisquare_tests.csv"
)

all_scaffold_total_status_counts <- build_total_status_counts(
  dplyr::filter(target_analysis, .data$show_in_figure),
  target_gene_summary
)
write_csv_out(
  all_scaffold_total_status_counts,
  "figure4_all_scaffolds_total_status_counts.csv"
)

unplaced_counts <- target_analysis |>
  dplyr::filter(
    .data$sequence_class == "Unplaced scaffold",
    .data$show_in_figure
  ) |>
  dplyr::count(
    .data$source_sheet, .data$umbrella, .data$figure_scope,
    .data$x_label, .data$x_order, .data$status,
    name = "n_genes"
  ) |>
  dplyr::group_by(
    .data$source_sheet, .data$umbrella, .data$figure_scope,
    .data$x_label, .data$x_order
  ) |>
  dplyr::mutate(percentage = 100 * .data$n_genes / sum(.data$n_genes)) |>
  dplyr::ungroup()

unplaced_category_n <- target_gene_summary |>
  dplyr::filter(.data$sequence_class == "Unplaced scaffold") |>
  dplyr::count(.data$source_sheet, .data$umbrella, name = "n_targets")

unplaced_tests <- calculate_figure_tests(unplaced_counts)
unplaced_status_counts <- unplaced_counts |>
  dplyr::left_join(
    unplaced_category_n,
    by = c("source_sheet", "umbrella")
  ) |>
  dplyr::left_join(
    unplaced_tests,
    by = c("source_sheet", "umbrella", "figure_scope")
  )

write_csv_out(
  unplaced_status_counts,
  "figure4_unplaced_scaffold_status_counts.csv"
)
write_csv_out(
  unplaced_tests,
  "figure4_unplaced_exploratory_chisquare_tests.csv"
)

# Use the same three review classes defined on the scaffold-origin audit page.
# These classes flag provenance uncertainty; they do not establish contamination.
homology_review_classes <- c(
  "Concordant non-metazoan nucleotide and protein hits",
  "Conflicting nucleotide and protein evidence",
  "Non-metazoan-only protein hits without nucleotide corroboration"
)

homology_review_targets <- homology_evidence |>
  dplyr::filter(.data$integrated_evidence %in% homology_review_classes) |>
  dplyr::transmute(
    gene_id,
    homology_evidence = .data$integrated_evidence,
    evidence_tier = dplyr::case_when(
      .data$integrated_evidence ==
        "Concordant non-metazoan nucleotide and protein hits" ~
        "Concordant nucleotide and protein flag",
      .data$integrated_evidence ==
        "Conflicting nucleotide and protein evidence" ~
        "Conflicting nucleotide and protein evidence",
      TRUE ~ "Protein-only non-metazoan flag"
    ),
    nucleotide_groups,
    nucleotide_taxa,
    nucleotide_max_identity,
    nucleotide_max_query_coverage,
    protein_support
  )

homology_flag_analysis <- target_analysis |>
  dplyr::inner_join(homology_review_targets, by = "gene_id") |>
  dplyr::filter(.data$show_in_figure)

homology_flag_gene_audit <- gene_audit |>
  dplyr::inner_join(homology_review_targets, by = "gene_id") |>
  dplyr::arrange(
    factor(
      .data$evidence_tier,
      levels = c(
        "Concordant nucleotide and protein flag",
        "Conflicting nucleotide and protein evidence",
        "Protein-only non-metazoan flag"
      )
    ),
    .data$source_sheet,
    .data$umbrella,
    .data$gene_id
  )

homology_flag_counts <- homology_flag_analysis |>
  dplyr::count(
    .data$source_sheet, .data$umbrella, .data$figure_scope,
    .data$x_label, .data$x_order, .data$status,
    name = "n_genes"
  ) |>
  dplyr::group_by(
    .data$source_sheet, .data$umbrella, .data$figure_scope,
    .data$x_label, .data$x_order
  ) |>
  dplyr::mutate(percentage = 100 * .data$n_genes / sum(.data$n_genes)) |>
  dplyr::ungroup()

homology_flag_category_n <- homology_flag_gene_audit |>
  dplyr::count(.data$source_sheet, .data$umbrella, name = "n_targets")

homology_flag_tests <- calculate_figure_tests(homology_flag_counts)
homology_flag_status_counts <- homology_flag_counts |>
  dplyr::left_join(
    homology_flag_category_n,
    by = c("source_sheet", "umbrella")
  ) |>
  dplyr::left_join(
    homology_flag_tests,
    by = c("source_sheet", "umbrella", "figure_scope")
  ) |>
  dplyr::mutate(
    # This review subset was selected from one DEG contrast, so a composition
    # test across the same contrasts would not be an independent inference.
    p_label = ""
  )

write_csv_out(
  homology_flag_gene_audit,
  "figure4_non_grasshopper_homology_flag_gene_audit.csv"
)
write_csv_out(
  homology_flag_analysis |>
    dplyr::select(
      .data$gene_id, .data$description_gff, .data$source_sheet,
      .data$umbrella, .data$subcategory, .data$seqid,
      .data$homology_evidence, .data$evidence_tier,
      .data$nucleotide_groups, .data$nucleotide_taxa,
      .data$nucleotide_max_identity,
      .data$nucleotide_max_query_coverage,
      .data$protein_support, .data$contrast_label,
      .data$oriented_log2FC, .data$padj, .data$status,
      .data$higher_expression_condition
    ),
  "figure4_non_grasshopper_homology_flag_contrasts.csv"
)
write_csv_out(
  homology_flag_status_counts,
  "figure4_non_grasshopper_homology_flag_status_counts.csv"
)

category_display <- c(
  "Innate Immune Recognition & Signaling" =
    "Innate immune\nrecognition &\nsignaling",
  "Humoral Immunity" = "Humoral\nimmunity",
  "Cellular Immunity" = "Cellular\nimmunity",
  "Detoxification" = "Detoxification",
  "Protein Synthesis & Translation" =
    "Protein synthesis\n& translation",
  "Nutrient Sensing & TOR/Insulin Signaling" =
    "Nutrient sensing\n& TOR/insulin\nsignaling",
  "Nitrogen, Amino Acid Transport & Metabolism" =
    "Nitrogen, amino acid\ntransport & metabolism",
  "Nitrogen & Amino Acid Metabolism" =
    "Nitrogen & amino acid\nmetabolism",
  "Storage Hexamerins & Reserves" =
    "Storage hexamerins\n& reserves",
  "Total Immune" = "Total\nimmune",
  "Total Protein Anabolism" = "Total protein\nanabolism"
)

status_colors <- c(
  "Downregulated" = "#3B59B6",
  "N.S." = "#CCCCCC",
  "Upregulated" = "#9E1B22"
)
diet_colors <- c("33" = "#C8641A", "50" = "#1B6B3A", "83" = "#5A2A82")

build_figure4_panel <- function(
    status_count_data, sheet_name, scope_name, tag,
    show_y = TRUE, show_legend = FALSE, show_tests = TRUE) {
  panel_data <- status_count_data |>
    dplyr::filter(
      .data$source_sheet == sheet_name,
      .data$figure_scope == scope_name
    ) |>
    dplyr::mutate(
      umbrella = factor(
        .data$umbrella,
        levels = category_order[category_order %in% unique(.data$umbrella)]
      ),
      status = factor(
        .data$status,
        levels = c("Downregulated", "N.S.", "Upregulated")
      ),
      x_label = factor(
        .data$x_label,
        levels = unique(.data$x_label[order(.data$x_order)])
      ),
      outline = if (scope_name == "Within-diet infection response") {
        unname(diet_colors[as.character(.data$x_label)])
      } else {
        "#252525"
      }
    )

  panel_annotations <- panel_data |>
    dplyr::distinct(
      .data$umbrella, .data$p_label, .data$n_targets
    ) |>
    dplyr::mutate(
      p_label = if (show_tests) .data$p_label else "",
      x_label = factor(
        levels(panel_data$x_label)[[2]],
        levels = levels(panel_data$x_label)
      )
    )

  panel_title <- if (scope_name == "Within-diet infection response") {
    "Within-diet comparisons\n(Infected vs control)"
  } else {
    "Across-diet comparisons among infected\n(Higher protein vs lower protein)"
  }

  group_title <- dplyr::recode(
    sheet_name,
    "Immune" = "Immune-related",
    "Protein-Anabolism" = "Protein anabolism"
  )
  y_title <- if (show_y) {
    paste0(group_title, "\n\nPercentage of curated target genes (%)")
  } else {
    NULL
  }

  ggplot(panel_data, aes(x = x_label, y = percentage, fill = status)) +
    geom_col(
      aes(color = outline),
      width = 0.66, linewidth = 0.65
    ) +
    geom_text(
      data = panel_annotations,
      aes(x = x_label, y = 108, label = p_label),
      inherit.aes = FALSE,
      size = 3.3, fontface = "bold"
    ) +
    geom_text(
      data = panel_annotations,
      aes(x = x_label, y = -21, label = paste0("targets = ", n_targets)),
      inherit.aes = FALSE,
      size = 3.2
    ) +
    facet_wrap(
      ~ umbrella,
      nrow = 1,
      labeller = as_labeller(category_display)
    ) +
    scale_fill_manual(values = status_colors, drop = FALSE) +
    scale_color_identity() +
    scale_y_continuous(
      breaks = c(0, 50, 100),
      limits = c(-25, 113),
      expand = expansion(mult = c(0, 0))
    ) +
    coord_cartesian(clip = "off") +
    labs(
      tag = tag,
      title = panel_title,
      x = NULL,
      y = y_title,
      fill = NULL
    ) +
    theme_classic(base_size = 11) +
    theme(
      plot.tag = element_text(face = "bold", size = 17),
      plot.tag.position = c(0.005, 1.02),
      plot.title = element_text(
        face = "bold", size = 13, hjust = 0.5,
        margin = margin(b = 10)
      ),
      axis.title.y = element_text(size = 12, margin = margin(r = 7)),
      axis.text.y = if (show_y) {
        element_text(size = 10, color = "black")
      } else {
        element_blank()
      },
      axis.ticks.y = if (show_y) element_line() else element_blank(),
      axis.line.y = if (show_y) element_line() else element_blank(),
      axis.text.x = element_text(size = 10.5, color = "black", lineheight = 0.9),
      axis.ticks.x = element_blank(),
      strip.background = element_blank(),
      strip.text = element_text(
        face = "bold", size = 10.5, lineheight = 0.9,
        margin = margin(b = 7)
      ),
      panel.spacing.x = grid::unit(0.35, "lines"),
      legend.position = if (show_legend) "bottom" else "none",
      legend.text = element_text(size = 11),
      legend.key.size = grid::unit(4.5, "mm"),
      plot.margin = margin(12, 10, 40, 10)
    )
}

build_figure4_summary_panel <- function(
    status_count_data, scope_name, tag, show_y = TRUE, show_tests = TRUE) {
  panel_data <- status_count_data |>
    dplyr::filter(.data$figure_scope == scope_name) |>
    dplyr::mutate(
      umbrella = factor(
        .data$umbrella,
        levels = c("Total Immune", "Total Protein Anabolism")
      ),
      status = factor(
        .data$status,
        levels = c("Downregulated", "N.S.", "Upregulated")
      ),
      x_label = factor(
        .data$x_label,
        levels = unique(.data$x_label[order(.data$x_order)])
      ),
      outline = if (scope_name == "Within-diet infection response") {
        unname(diet_colors[as.character(.data$x_label)])
      } else {
        "#252525"
      }
    )

  panel_annotations <- panel_data |>
    dplyr::distinct(.data$umbrella, .data$p_label, .data$n_targets) |>
    dplyr::mutate(
      p_label = if (show_tests) .data$p_label else "",
      x_label = factor(
        levels(panel_data$x_label)[[2]],
        levels = levels(panel_data$x_label)
      )
    )

  panel_title <- if (scope_name == "Within-diet infection response") {
    "Within-diet comparisons\n(Infected vs control)"
  } else {
    "Across-diet comparisons among infected\n(Higher protein vs lower protein)"
  }

  ggplot(panel_data, aes(x = x_label, y = percentage, fill = status)) +
    geom_col(aes(color = outline), width = 0.66, linewidth = 0.65) +
    geom_text(
      data = panel_annotations,
      aes(x = x_label, y = 108, label = p_label),
      inherit.aes = FALSE,
      size = 3.3,
      fontface = "bold"
    ) +
    geom_text(
      data = panel_annotations,
      aes(x = x_label, y = -21, label = paste0("targets = ", n_targets)),
      inherit.aes = FALSE,
      size = 3.2
    ) +
    facet_wrap(
      ~ umbrella,
      nrow = 1,
      labeller = as_labeller(category_display)
    ) +
    scale_fill_manual(values = status_colors, drop = FALSE) +
    scale_color_identity() +
    scale_y_continuous(
      breaks = c(0, 50, 100),
      limits = c(-25, 113),
      expand = expansion(mult = c(0, 0))
    ) +
    coord_cartesian(clip = "off") +
    labs(
      tag = tag,
      title = panel_title,
      x = NULL,
      y = if (show_y) "Summary totals\n\nPercentage of curated target genes (%)" else NULL,
      fill = NULL
    ) +
    theme_classic(base_size = 11) +
    theme(
      plot.tag = element_text(face = "bold", size = 17),
      plot.tag.position = c(0.005, 1.02),
      plot.title = element_text(
        face = "bold", size = 13, hjust = 0.5,
        margin = margin(b = 10)
      ),
      axis.title.y = element_text(size = 12, margin = margin(r = 7)),
      axis.text.y = if (show_y) {
        element_text(size = 10, color = "black")
      } else {
        element_blank()
      },
      axis.ticks.y = if (show_y) element_line() else element_blank(),
      axis.line.y = if (show_y) element_line() else element_blank(),
      axis.text.x = element_text(size = 10.5, color = "black", lineheight = 0.9),
      axis.ticks.x = element_blank(),
      strip.background = element_blank(),
      strip.text = element_text(
        face = "bold", size = 10.5, lineheight = 0.9,
        margin = margin(b = 7)
      ),
      panel.spacing.x = grid::unit(0.35, "lines"),
      legend.position = "none",
      plot.margin = margin(12, 10, 40, 10)
    )
}

p_a <- build_figure4_panel(
  figure_status_counts, "Immune", "Within-diet infection response", "A",
  TRUE, FALSE
)
p_b <- build_figure4_panel(
  figure_status_counts, "Immune", "Diet comparison among infected", "B", FALSE
)
p_c <- build_figure4_panel(
  figure_status_counts, "Protein-Anabolism",
  "Within-diet infection response", "C", TRUE
)
p_d <- build_figure4_panel(
  figure_status_counts, "Protein-Anabolism",
  "Diet comparison among infected", "D", FALSE
)
p_e <- build_figure4_summary_panel(
  figure_total_status_counts,
  "Within-diet infection response", "E", TRUE
)
p_f <- build_figure4_summary_panel(
  figure_total_status_counts,
  "Diet comparison among infected", "F", FALSE
)

legend_data <- tibble::tibble(
  status = factor(
    c("Downregulated", "N.S.", "Upregulated"),
    levels = c("Downregulated", "N.S.", "Upregulated")
  ),
  x = seq_len(3),
  y = 1
)
legend_plot <- ggplot(legend_data, aes(x = x, y = y, fill = status)) +
  geom_point(shape = 22, size = 5, color = NA) +
  scale_fill_manual(values = status_colors, drop = FALSE) +
  guides(fill = guide_legend(title = NULL, nrow = 1)) +
  theme_void(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 12),
    legend.key.size = grid::unit(5, "mm")
  )
shared_status_legend <- cowplot::get_legend(legend_plot)

figure4_main_chromosome <- (
  (p_a | p_b) /
  (p_c | p_d) /
  (p_e | p_f) /
  patchwork::wrap_elements(full = shared_status_legend)
) +
  patchwork::plot_layout(
    heights = c(1, 1, 0.88, 0.07)
  ) +
  patchwork::plot_annotation(
    title = "Curated DEGs relating to immune and protein-anabolism pathways",
    subtitle = paste0(
      "Main nuclear chromosomes; competitive-host exon-only DESeq2; ",
      main_targets, " retained targets; padj < ", alpha,
      " and |log2FC| >= ", lfc_threshold
    ),
    theme = theme(
      plot.title = element_text(face = "bold", size = 20, hjust = 0.5),
      plot.subtitle = element_text(size = 12.5, hjust = 0.5),
      plot.margin = margin(10, 10, 5, 10)
    )
  )

save_figure(
  figure4_main_chromosome,
  "figure4_curated_targets_main_chromosomes",
  15,
  14.2
)

# The workbook was assembled from genes significant in at least one contrast.
# This manuscript version therefore keeps the status bars descriptive and
# removes category-level p-values that would reuse the selection criterion.
desc_a <- build_figure4_panel(
  figure_status_counts, "Immune", "Within-diet infection response", "A",
  TRUE, FALSE, FALSE
)
desc_b <- build_figure4_panel(
  figure_status_counts, "Immune", "Diet comparison among infected", "B",
  FALSE, FALSE, FALSE
)
desc_c <- build_figure4_panel(
  figure_status_counts, "Protein-Anabolism",
  "Within-diet infection response", "C", TRUE, FALSE, FALSE
)
desc_d <- build_figure4_panel(
  figure_status_counts, "Protein-Anabolism",
  "Diet comparison among infected", "D", FALSE, FALSE, FALSE
)
desc_e <- build_figure4_summary_panel(
  figure_total_status_counts,
  "Within-diet infection response", "E", TRUE, FALSE
)
desc_f <- build_figure4_summary_panel(
  figure_total_status_counts,
  "Diet comparison among infected", "F", FALSE, FALSE
)

figure4_main_chromosome_descriptive <- (
  (desc_a | desc_b) /
  (desc_c | desc_d) /
  (desc_e | desc_f) /
  patchwork::wrap_elements(full = shared_status_legend)
) +
  patchwork::plot_layout(heights = c(1, 1, 0.88, 0.07)) +
  patchwork::plot_annotation(
    title = "Curated infection-responsive immune and protein-anabolism genes",
    subtitle = paste0(
      "Descriptive main-chromosome view; ", main_targets,
      " retained targets; exon-only DESeq2"
    ),
    theme = theme(
      plot.title = element_text(face = "bold", size = 20, hjust = 0.5),
      plot.subtitle = element_text(size = 12.5, hjust = 0.5),
      plot.margin = margin(10, 10, 5, 10)
    )
  )

save_figure(
  figure4_main_chromosome_descriptive,
  "figure4_curated_targets_main_chromosomes_descriptive",
  15,
  14.2
)

all_a <- build_figure4_panel(
  all_scaffold_status_counts, "Immune", "Within-diet infection response", "A",
  TRUE, FALSE
)
all_b <- build_figure4_panel(
  all_scaffold_status_counts, "Immune", "Diet comparison among infected", "B",
  FALSE
)
all_c <- build_figure4_panel(
  all_scaffold_status_counts, "Protein-Anabolism",
  "Within-diet infection response", "C", TRUE
)
all_d <- build_figure4_panel(
  all_scaffold_status_counts, "Protein-Anabolism",
  "Diet comparison among infected", "D", FALSE
)
all_e <- build_figure4_summary_panel(
  all_scaffold_total_status_counts,
  "Within-diet infection response", "E", TRUE
)
all_f <- build_figure4_summary_panel(
  all_scaffold_total_status_counts,
  "Diet comparison among infected", "F", FALSE
)

figure4_all_scaffolds <- (
  (all_a | all_b) /
  (all_c | all_d) /
  (all_e | all_f) /
  patchwork::wrap_elements(full = shared_status_legend)
) +
  patchwork::plot_layout(heights = c(1, 1, 0.88, 0.07)) +
  patchwork::plot_annotation(
    title = "Curated DEGs relating to immune and protein-anabolism pathways",
    subtitle = paste0(
      "All scaffolds; competitive-host exon-only DESeq2; ",
      total_targets, " curated targets; padj < ", alpha,
      " and |log2FC| >= ", lfc_threshold
    ),
    theme = theme(
      plot.title = element_text(face = "bold", size = 20, hjust = 0.5),
      plot.subtitle = element_text(size = 12.5, hjust = 0.5),
      plot.margin = margin(10, 10, 5, 10)
    )
  )

save_figure(
  figure4_all_scaffolds,
  "figure4_curated_targets_all_scaffolds",
  15,
  14.2
)

all_desc_a <- build_figure4_panel(
  all_scaffold_status_counts, "Immune", "Within-diet infection response", "A",
  TRUE, FALSE, FALSE
)
all_desc_b <- build_figure4_panel(
  all_scaffold_status_counts, "Immune", "Diet comparison among infected", "B",
  FALSE, FALSE, FALSE
)
all_desc_c <- build_figure4_panel(
  all_scaffold_status_counts, "Protein-Anabolism",
  "Within-diet infection response", "C", TRUE, FALSE, FALSE
)
all_desc_d <- build_figure4_panel(
  all_scaffold_status_counts, "Protein-Anabolism",
  "Diet comparison among infected", "D", FALSE, FALSE, FALSE
)
all_desc_e <- build_figure4_summary_panel(
  all_scaffold_total_status_counts,
  "Within-diet infection response", "E", TRUE, FALSE
)
all_desc_f <- build_figure4_summary_panel(
  all_scaffold_total_status_counts,
  "Diet comparison among infected", "F", FALSE, FALSE
)

figure4_all_scaffolds_descriptive <- (
  (all_desc_a | all_desc_b) /
  (all_desc_c | all_desc_d) /
  (all_desc_e | all_desc_f) /
  patchwork::wrap_elements(full = shared_status_legend)
) +
  patchwork::plot_layout(heights = c(1, 1, 0.88, 0.07)) +
  patchwork::plot_annotation(
    title = "Curated infection-responsive immune and protein-anabolism genes",
    subtitle = paste0(
      "Descriptive all-scaffold view; ", total_targets,
      " retained targets; exon-only DESeq2"
    ),
    theme = theme(
      plot.title = element_text(face = "bold", size = 20, hjust = 0.5),
      plot.subtitle = element_text(size = 12.5, hjust = 0.5),
      plot.margin = margin(10, 10, 5, 10)
    )
  )

save_figure(
  figure4_all_scaffolds_descriptive,
  "figure4_curated_targets_all_scaffolds_descriptive",
  15,
  14.2
)

if (unplaced_targets_n > 0) {
  u_a <- build_figure4_panel(
    unplaced_status_counts, "Immune", "Within-diet infection response", "A",
    TRUE, FALSE
  )
  u_b <- build_figure4_panel(
    unplaced_status_counts, "Immune", "Diet comparison among infected", "B",
    FALSE
  )
  u_c <- build_figure4_panel(
    unplaced_status_counts, "Protein-Anabolism",
    "Within-diet infection response", "C", TRUE
  )
  u_d <- build_figure4_panel(
    unplaced_status_counts, "Protein-Anabolism",
    "Diet comparison among infected", "D", FALSE
  )

  figure4_unplaced_scaffolds <- (
    (u_a | u_b) /
    (u_c | u_d) /
    patchwork::wrap_elements(full = shared_status_legend)
  ) +
    patchwork::plot_layout(heights = c(1, 1, 0.075)) +
    patchwork::plot_annotation(
      title = "Curated immune and protein-anabolism targets on unplaced scaffolds",
      subtitle = paste0(
        "Competitive-host DESeq2; ", unplaced_targets_n,
        " audit targets; padj < ", alpha,
        " and |log2FC| >= ", lfc_threshold
      ),
      theme = theme(
        plot.title = element_text(face = "bold", size = 20, hjust = 0.5),
        plot.subtitle = element_text(size = 12.5, hjust = 0.5),
        plot.margin = margin(10, 10, 5, 10)
      )
    )
} else {
  figure4_unplaced_scaffolds <- ggplot() +
    annotate(
      "text", x = 0.5, y = 0.5,
      label = "No genes in the supplied v21 target list are on unplaced scaffolds.",
      size = 6
    ) +
    xlim(0, 1) + ylim(0, 1) + theme_void() +
    labs(title = "Unplaced-scaffold curated-target audit") +
    theme(plot.title = element_text(face = "bold", size = 20, hjust = 0.5))
}

save_figure(
  figure4_unplaced_scaffolds,
  "figure4_curated_targets_unplaced_scaffolds",
  15,
  11.5
)

homology_flag_target_n <- dplyr::n_distinct(homology_flag_analysis$gene_id)
homology_concordant_n <- sum(
  homology_flag_gene_audit$evidence_tier ==
    "Concordant nucleotide and protein flag"
)

if (homology_flag_target_n > 0) {
  flag_a <- build_figure4_panel(
    homology_flag_status_counts, "Immune",
    "Within-diet infection response", "A", TRUE, FALSE
  )
  flag_b <- build_figure4_panel(
    homology_flag_status_counts, "Immune",
    "Diet comparison among infected", "B", FALSE
  )
  flag_c <- build_figure4_panel(
    homology_flag_status_counts, "Protein-Anabolism",
    "Within-diet infection response", "C", TRUE
  )
  flag_d <- build_figure4_panel(
    homology_flag_status_counts, "Protein-Anabolism",
    "Diet comparison among infected", "D", FALSE
  )

  figure4_homology_flags <- (
    (flag_a | flag_b) /
    (flag_c | flag_d) /
    patchwork::wrap_elements(full = shared_status_legend)
  ) +
    patchwork::plot_layout(heights = c(1, 1, 0.075)) +
    patchwork::plot_annotation(
      title = "Curated targets with non-grasshopper homology review flags",
      subtitle = paste0(
        homology_flag_target_n,
        " targets selected from unresolved infected diet 33 vs 50 DEGs; only ",
        homology_concordant_n,
        " have concordant non-metazoan nucleotide and protein hits"
      ),
      theme = theme(
        plot.title = element_text(face = "bold", size = 20, hjust = 0.5),
        plot.subtitle = element_text(size = 12.5, hjust = 0.5),
        plot.margin = margin(10, 10, 5, 10)
      )
    )
} else {
  figure4_homology_flags <- ggplot() +
    annotate(
      "text", x = 0.5, y = 0.5,
      label = "No genes in the supplied v21 target list meet the homology-review flag.",
      size = 6
    ) +
    xlim(0, 1) + ylim(0, 1) + theme_void() +
    labs(title = "Curated-target homology audit") +
    theme(plot.title = element_text(face = "bold", size = 20, hjust = 0.5))
}

save_figure(
  figure4_homology_flags,
  "figure4_curated_targets_non_grasshopper_homology_flags",
  15,
  11.5
)
```

### Figure 4 views {.tabset}

#### Main chromosomes

```{r display-figure4-main, echo=FALSE, fig.width=15, fig.height=14.2}
figure4_main_chromosome_descriptive
```

#### All scaffolds

```{r display-figure4-all-scaffolds, echo=FALSE, fig.width=15, fig.height=14.2}
figure4_all_scaffolds_descriptive
```

#### Homology review flags

```{r display-figure4-homology-flags, echo=FALSE, fig.width=15, fig.height=11.5}
figure4_homology_flags
```

The first composite is the conservative host-chromosome view used for the main
paper interpretation. The second restores all `r total_targets` genes from the
v21 curated workbook, including the `r unplaced_targets_n` targets on unplaced
scaffolds, so its target denominators remain comparable with the complete
curated list. Both use the same
current DESeq2 model, contrasts, thresholds, colors, and panel geometry. The
left panels summarize infection responses within each diet; the right panels
compare infected locusts across diets. The exploratory chi-square labels
summarize category composition and do not replace the gene-level DESeq2 tests.

The third composite isolates the `r homology_flag_target_n` curated targets that
meet the scaffold-origin page's predefined homology-review rule. Of these,
`r homology_concordant_n` have non-metazoan top hits in both BLASTn and DIAMOND;
the remainder have conflicting evidence or protein-only support. These are
**provenance-review flags, not confirmed non-locust genes**. Conserved proteins,
incomplete reference representation, and short or divergent alignments can all
produce a non-grasshopper top hit.

All `r homology_flag_target_n` targets are significant for infected diet 50
versus infected diet 33
because the scaffold-origin homology audit was itself constructed from the
unresolved genes driving that contrast. That column documents the selection
rule; it is not an independent confirmation. For this reason, the audit-only
composite does not display the exploratory chi-square labels used for the fixed
main and all-scaffold target sets.

### Genes behind the homology-flagged figure

```{r homology-flagged-curated-table, echo=FALSE}
homology_flag_table <- homology_flag_analysis |>
  dplyr::select(
    .data$gene_id, description = .data$description_gff,
    functional_group = .data$umbrella, .data$subcategory,
    scaffold = .data$seqid, .data$evidence_tier,
    .data$nucleotide_groups, .data$nucleotide_taxa,
    .data$contrast_label, log2FC = .data$oriented_log2FC,
    adjusted_p_value = .data$padj, .data$status,
    higher_in = .data$higher_expression_condition
  ) |>
  dplyr::mutate(
    log2FC = round(.data$log2FC, 3),
    adjusted_p_value = signif(.data$adjusted_p_value, 3)
  ) |>
  dplyr::arrange(
    factor(
      .data$evidence_tier,
      levels = c(
        "Concordant nucleotide and protein flag",
        "Conflicting nucleotide and protein evidence",
        "Protein-only non-metazoan flag"
      )
    ),
    .data$functional_group,
    .data$gene_id,
    .data$contrast_label
  )

DT::datatable(
  homology_flag_table,
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    pageLength = 15,
    scrollX = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv")
  ),
  caption = paste0(
    "Curated Figure 4 targets carrying scaffold-origin homology review flags. ",
    "Positive log2FC follows the condition order printed in the contrast."
  )
) |>
  DT::formatStyle(
    "status",
    color = DT::styleEqual(
      c("Downregulated", "N.S.", "Upregulated"),
      c("#244A9B", "#555555", "#8B1018")
    ),
    fontWeight = "bold"
  )
```

## Comparison with the earlier workbook

```{r old-new-comparison}
status_transition <- target_analysis |>
  dplyr::filter(.data$primary_figure) |>
  dplyr::count(
    .data$original_status,
    current_status = .data$status,
    name = "gene_contrast_cells"
  ) |>
  dplyr::arrange(.data$original_status, .data$current_status)

changed_cells <- target_analysis |>
  dplyr::filter(
    .data$primary_figure,
    .data$original_status != .data$status
  )

write_csv_out(status_transition, "figure4_old_to_current_status_transitions.csv")
write_csv_out(changed_cells, "figure4_changed_placed_target_cells.csv")

knitr::kable(
  status_transition,
  col.names = c("Workbook status", "Current status", "Gene-contrast cells"),
  align = c("l", "l", "r")
)
```

There are `r nrow(changed_cells)` placed-gene contrast cells whose status differs
between the supplied workbook and the current competitive-host analysis. This
comparison separates two effects: removal of unplaced targets changes the
category denominator, while remapping and re-fitting DESeq2 can change the
status of targets that remain on main chromosomes.

## Unplaced target details

```{r unplaced-targets}
unplaced_followup <- target_analysis |>
  dplyr::filter(.data$sequence_class == "Unplaced scaffold") |>
  dplyr::group_by(
    .data$source_sheet, .data$umbrella, .data$subcategory,
    .data$gene_id, .data$description_workbook, .data$description_gff,
    .data$gene_biotype, .data$seqid
  ) |>
  dplyr::summarise(
    significant_in_current_unplaced_analysis = any(.data$significant),
    n_significant_contrasts = sum(.data$significant),
    significant_contrasts = paste(
      .data$contrast_label[.data$significant],
      collapse = "; "
    ),
    higher_expression_conditions = paste(
      .data$higher_expression_condition[.data$significant],
      collapse = "; "
    ),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    significant_contrasts = dplyr::if_else(
      .data$significant_contrasts == "",
      "None at current thresholds",
      .data$significant_contrasts
    ),
    higher_expression_conditions = dplyr::if_else(
      .data$higher_expression_conditions == "",
      "Not significant",
      .data$higher_expression_conditions
    )
  ) |>
  dplyr::arrange(.data$source_sheet, .data$umbrella, .data$gene_id)

unplaced_33_vs_50 <- target_analysis |>
  dplyr::filter(
    .data$sequence_class == "Unplaced scaffold",
    .data$contrast_id == "infected_diet33_vs50",
    .data$significant
  ) |>
  dplyr::distinct(.data$gene_id) |>
  nrow()

write_csv_out(unplaced_followup, "curated_unplaced_targets_for_followup.csv")
```

Of the `r unplaced_targets_n` excluded curated targets,
`r unplaced_33_vs_50` are significant in the current unplaced-scaffold analysis
for infected diet 33 versus infected diet 50. These genes are therefore a
specific follow-up set for the scaffold-origin work, not evidence to include in
the primary host-chromosome Figure 4.

```{r unplaced-searchable-table}
DT::datatable(
  unplaced_followup,
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv")
  )
)
```

## Searchable retained target table

```{r retained-searchable-table}
retained_target_table <- target_analysis |>
  dplyr::filter(.data$primary_figure) |>
  dplyr::select(
    source_sheet, umbrella, subcategory, gene_id,
    description_gff, gene_biotype, assigned_molecule,
    contrast_label, log2FoldChange, padj, status,
    higher_expression_condition
  ) |>
  dplyr::arrange(.data$source_sheet, .data$umbrella, .data$gene_id)

DT::datatable(
  retained_target_table,
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv")
  )
)
```

## Exported files

```{r export-manifest}
manifest <- tibble::tibble(
  item = c(
    "Curated v21 workbook",
    "Frozen 44-sample metadata excluding 1044",
    "Official placement table",
    "Current DESeq2 result table",
    "Corrected BLASTn/DIAMOND homology evidence",
    "Output directory"
  ),
  path = c(
    workbook_file,
    metadata_file,
    placement_file,
    results_file,
    homology_evidence_file,
    out_dir
  )
)
write_csv_out(manifest, "figure4_input_manifest.csv")

knitr::kable(manifest, col.names = c("Item", "Path"), align = c("l", "l"))
```

Each knit creates a new timestamped output folder. Matched composites are
exported as `figure4_curated_targets_main_chromosomes` and
`figure4_curated_targets_all_scaffolds` in PNG, SVG, and PDF formats. The
standalone homology-review composite is exported as
`figure4_curated_targets_non_grasshopper_homology_flags`. The unplaced-only
composite, placement-impact figure, and all gene-level audit tables are exported
beside them for sensitivity review.

## Session information

```{r session-info}
sessionInfo()
```
