---
title: "Exon-only publication figure exports"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  metadata_file: "data/metadata/mehreen_metadata_analysis_44_excluding_1044.txt"
  counts_file: "output/rmd_runs/exon-only-primary-excluding1044_20260809-233420/host_counts_no_rrna.tsv"
  placement_file: "output/runs/host_pathogen_dual_20260727_000952/local-analysis/host-mapping-audit_20260729-163928/unified-model/audit-summary/host_gene_sequence_placement.tsv"
  rrna_file: "data/excluded_loci/gregaria_rrna_list.txt"
  gff_file: "data/reference/GCF_023897955.1_iqSchGreg1.2_genomic.gff"
  eggnog_file: "data/GO_Annotations/GCF_023897955.1_iqSchGreg1.2_Arthopoda_one2one.emapper.annotations"
  latest_results_root: "output/rmd_runs"
  primary_results_pattern: "^exon-only-deseq2-analysis_"
  cluster_results_pattern: "^exon-only-cluster-enrichment_"
  curated_results_pattern: "^exon-only-curated-target-genes_"
  min_total_count: 10
  min_count: 10
  min_samples: 3
  alpha: 0.05
  lfc_threshold: 1
  output_tag: "exon-only-publication-figures"
---

<style>
img {
  max-width: 100%;
  height: auto;
}
</style>

## Purpose

This page rebuilds the established publication figure set using only genes
classified as `Nuclear chromosome` by the official assembly-derived placement
table. The 44-library analysis cohort excludes QC outlier 1044, annotated rRNA loci are excluded,
and expression counts come from the fresh competitive host-pathogen mapping.
Unplaced-scaffold results remain on the audit page and are not mixed into these
primary paper figures.

Every knit writes to a new folder under `output/rmd_runs`, so previous figure
drafts remain available for comparison.

::: {.walkthrough-grid}
::: {.walkthrough-step}
**Step 1 - Verify inputs**

Confirm the count definition, sample set, placement table, and source reports.
:::
::: {.walkthrough-step}
**Step 2 - Assemble panels**

Copy or rebuild each panel from its analysis page without changing the statistics.
:::
::: {.walkthrough-step}
**Step 3 - Export**

Write publication-ready PNG, SVG, PDF, and source-table files together.
:::
:::

## Setup

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

publication_default_params <- list(
  metadata_file = "data/metadata/mehreen_metadata_analysis_44_excluding_1044.txt",
  counts_file = paste0(
    "output/rmd_runs/exon-only-primary-excluding1044_20260809-233420/",
    "host_counts_no_rrna.tsv"
  ),
  placement_file = paste0(
    "output/runs/host_pathogen_dual_20260727_000952/local-analysis/",
    "host-mapping-audit_20260729-163928/unified-model/audit-summary/",
    "host_gene_sequence_placement.tsv"
  ),
  rrna_file = "data/excluded_loci/gregaria_rrna_list.txt",
  gff_file = "data/reference/GCF_023897955.1_iqSchGreg1.2_genomic.gff",
  eggnog_file = paste0(
    "data/GO_Annotations/",
    "GCF_023897955.1_iqSchGreg1.2_Arthopoda_one2one.emapper.annotations"
  ),
  latest_results_root = "output/rmd_runs",
  primary_results_pattern = "^exon-only-deseq2-analysis_",
  cluster_results_pattern = "^exon-only-cluster-enrichment_",
  curated_results_pattern = "^exon-only-curated-target-genes_",
  min_total_count = 10,
  min_count = 10,
  min_samples = 3,
  alpha = 0.05,
  lfc_threshold = 1,
  output_tag = "exon-only-publication-figures"
)

# R Markdown supplies params while knitting. These defaults make Run All Chunks
# behave the same way in an interactive RStudio session.
if (!exists("params", inherits = TRUE) || !is.list(params)) {
  publication_params <- publication_default_params
} else {
  publication_params <- utils::modifyList(publication_default_params, params)
}

path_ancestors <- function(path) {
  path <- normalizePath(path, mustWork = FALSE)
  ancestors <- path
  repeat {
    parent <- dirname(path)
    if (identical(parent, path)) break
    ancestors <- c(ancestors, parent)
    path <- parent
  }
  ancestors
}

is_publication_project <- function(path) {
  file.exists(file.path(path, "analysis", "08-publication-figures.Rmd")) &&
    dir.exists(file.path(path, "data")) &&
    dir.exists(file.path(path, "output"))
}

find_publication_project <- function() {
  starts <- character()

  explicit_root <- Sys.getenv("GREGARIA_DIET_PROJECT_DIR", unset = "")
  if (nzchar(explicit_root)) starts <- c(starts, explicit_root)

  input_file <- tryCatch(knitr::current_input(dir = TRUE),
                         error = function(e) "")
  if (length(input_file) == 1 && nzchar(input_file)) {
    starts <- c(starts, dirname(input_file))
  }

  if (requireNamespace("rstudioapi", quietly = TRUE) &&
      rstudioapi::isAvailable()) {
    active_file <- tryCatch(
      rstudioapi::getSourceEditorContext()$path,
      error = function(e) ""
    )
    if (length(active_file) == 1 && nzchar(active_file)) {
      starts <- c(starts, dirname(active_file))
    }
  }

  starts <- unique(c(starts, getwd()))
  candidates <- unique(unlist(lapply(starts, path_ancestors), use.names = FALSE))
  matches <- candidates[vapply(candidates, is_publication_project, logical(1))]

  if (length(matches) == 0) {
    stop(
      "Could not locate gregaria-diet-infection-interaction. ",
      "Open this Rmd from that project, set the RStudio working directory to ",
      "the repository root, or set GREGARIA_DIET_PROJECT_DIR explicitly."
    )
  }

  normalizePath(matches[1], mustWork = TRUE)
}

required <- c("DESeq2", "SummarizedExperiment", "dplyr", "tidyr", "tibble", "stringr",
              "ggplot2", "ggrepel", "patchwork", "scales",
              "ggVennDiagram", "UpSetR", "ggplotify", "png", "grid")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

library(ggplot2)
library(patchwork)

project_dir <- find_publication_project()
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(project_dir, "output", "rmd_runs",
                     paste(publication_params$output_tag, run_stamp, sep = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

results_root <- file.path(project_dir, publication_params$latest_results_root)
metadata_file <- file.path(project_dir, publication_params$metadata_file)
counts_file <- file.path(project_dir, publication_params$counts_file)
placement_file <- file.path(project_dir, publication_params$placement_file)
rrna_file <- file.path(project_dir, publication_params$rrna_file)
gff_file <- file.path(project_dir, publication_params$gff_file)
eggnog_file <- file.path(project_dir, publication_params$eggnog_file)

diet_colors <- c("33" = "#E6862E", "50" = "#2F8F5B", "83" = "#7B4FA3")
treatment_colors <- c("Control" = "#FAEBD7", "Infected" = "#8B0000")
condition_colors <- c(
  "33 Control" = "#F7D6A6",
  "33 Infected" = "#C75C1B",
  "50 Control" = "#CDE8D5",
  "50 Infected" = "#1F6F43",
  "83 Control" = "#D8C7E8",
  "83 Infected" = "#5B2E83"
)

direction_biotype_colors <- c(
  "Upregulated - Coding genes" = "#B91C1C",
  "Upregulated - Long non-coding RNA" = "#F87171",
  "Upregulated - Other / unknown" = "#FCA5A5",
  "Downregulated - Coding genes" = "#1D4ED8",
  "Downregulated - Long non-coding RNA" = "#60A5FA",
  "Downregulated - Other / unknown" = "#BFDBFE"
)

physiology_theme_colors <- c(
  "Phase-related changes" = "#C51B7D",
  "Antimicrobial peptides" = "#D7191C",
  "Immune-related" = "#F58220",
  "Glycogen synthesis / storage" = "#FFD92F",
  "Lipid synthesis / metabolism" = "#1B9E77",
  "Protein synthesis / storage" = "#377EB8",
  "Detoxification / stress" = "#984EA3",
  "Other / unannotated genes" = "#BDBDBD"
)

theme_pub <- function(base_size = 12) {
  ggplot2::theme_classic(base_size = base_size) +
    ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold"),
      strip.background = ggplot2::element_rect(fill = "#F3F4F6",
                                               color = "#111827"),
      strip.text = ggplot2::element_text(face = "bold"),
      legend.position = "bottom",
      legend.title = ggplot2::element_text(face = "bold")
    )
}
```

## Helper functions

```{r helper-functions}
latest_dir <- function(pattern) {
  dirs <- list.dirs(results_root, recursive = FALSE, full.names = TRUE)
  dirs <- dirs[grepl(pattern, basename(dirs))]
  if (length(dirs) == 0) return(NA_character_)
  dirs[which.max(file.info(dirs)$mtime)]
}

save_png_svg <- function(plot, file_base, width, height, dpi = 300) {
  png_path <- file.path(out_dir, paste0(file_base, ".png"))
  svg_path <- file.path(out_dir, paste0(file_base, ".svg"))

  ggplot2::ggsave(png_path, plot, width = width, height = height,
                  dpi = dpi, limitsize = FALSE)

  grDevices::svg(svg_path, width = width, height = height, onefile = FALSE)
  on.exit(grDevices::dev.off(), add = TRUE)
  print(plot)

  invisible(c(png = png_path, svg = svg_path))
}

parse_gff_gene_annotation <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")

  gff |>
    dplyr::filter(type == "gene") |>
    dplyr::transmute(
      gene_id = dplyr::coalesce(
        stringr::str_match(attributes, "gene=([^;]+)")[, 2],
        stringr::str_match(attributes, "Name=([^;]+)")[, 2]
      ),
      Description = stringr::str_match(attributes, "description=([^;]+)")[, 2],
      gene_biotype = stringr::str_match(attributes, "gene_biotype=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(gene_id)) |>
    dplyr::mutate(
      Description = dplyr::if_else(
        is.na(Description) | Description == "",
        "no GFF description",
        utils::URLdecode(Description)
      ),
      gene_biotype = dplyr::coalesce(gene_biotype, "unknown")
    ) |>
    dplyr::distinct(gene_id, .keep_all = TRUE)
}

parse_gff_gene_coordinates <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")

  sequence_regions <- gff |>
    dplyr::filter(type == "region") |>
    dplyr::transmute(
      seqid,
      seq_start = start,
      seq_end = end,
      chromosome = stringr::str_match(attributes, "chromosome=([^;]+)")[, 2],
      region_name = stringr::str_match(attributes, "Name=([^;]+)")[, 2],
      genome = stringr::str_match(attributes, "genome=([^;]+)")[, 2]
    ) |>
    dplyr::mutate(
      chromosome = dplyr::coalesce(chromosome, region_name, seqid),
      chromosome = utils::URLdecode(chromosome),
      chromosome_class = dplyr::if_else(
        chromosome %in% c(as.character(1:11), "X"),
        "Named chromosome",
        "Unplaced scaffold"
      )
    ) |>
    dplyr::distinct(seqid, .keep_all = TRUE)

  gene_ranges <- gff |>
    dplyr::filter(type == "gene") |>
    dplyr::transmute(
      seqid,
      gene_id = dplyr::coalesce(
        stringr::str_match(attributes, "gene=([^;]+)")[, 2],
        stringr::str_match(attributes, "Name=([^;]+)")[, 2]
      ),
      start,
      end,
      midpoint = (start + end) / 2,
      strand
    ) |>
    dplyr::filter(!is.na(gene_id)) |>
    dplyr::distinct(gene_id, .keep_all = TRUE) |>
    dplyr::left_join(sequence_regions, by = "seqid")

  list(gene_ranges = gene_ranges, sequence_regions = sequence_regions)
}

first_nonempty <- function(x, fallback = "") {
  x <- x[!is.na(x) & x != "" & x != "-"]
  if (length(x) == 0) fallback else x[1]
}

read_emapper <- function(path) {
  cols <- c(
    "query", "seed_ortholog", "evalue", "score", "eggNOG_OGs",
    "max_annot_lvl", "COG_category", "eggNOG_Description",
    "eggNOG_name", "GOs", "EC", "KEGG_ko", "KEGG_Pathway",
    "KEGG_Module", "KEGG_Reaction", "KEGG_rclass", "BRITE",
    "KEGG_TC", "CAZy", "BiGG_Reaction", "PFAMs"
  )
  tab <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", fill = TRUE)
  names(tab) <- cols[seq_len(ncol(tab))]
  tab |>
    dplyr::mutate(protein_id = stringr::str_extract(query, "XP_[0-9]+[.][0-9]+"))
}

parse_cds_gene_bridge <- function(path) {
  gff <- read.delim(path, comment.char = "#", header = FALSE, sep = "\t",
                    quote = "", stringsAsFactors = FALSE)
  names(gff)[1:9] <- c("seqid", "source", "type", "start", "end", "score",
                       "strand", "phase", "attributes")
  gff |>
    dplyr::filter(type == "CDS") |>
    dplyr::transmute(
      protein_id = stringr::str_match(attributes, "protein_id=([^;]+)")[, 2],
      gene_id = stringr::str_match(attributes, "gene=([^;]+)")[, 2]
    ) |>
    dplyr::filter(!is.na(protein_id), !is.na(gene_id)) |>
    dplyr::distinct()
}

build_gene_annotation <- function(gff_path, eggnog_path) {
  gff_annotation <- parse_gff_gene_annotation(gff_path)

  eggnog_annotation <- read_emapper(eggnog_path) |>
    dplyr::left_join(parse_cds_gene_bridge(gff_path), by = "protein_id") |>
    dplyr::filter(!is.na(gene_id)) |>
    dplyr::group_by(gene_id) |>
    dplyr::summarise(
      eggNOG_Description = first_nonempty(eggNOG_Description),
      eggNOG_name = first_nonempty(eggNOG_name),
      GOs = first_nonempty(GOs),
      KEGG_ko = first_nonempty(KEGG_ko),
      KEGG_Pathway = first_nonempty(KEGG_Pathway),
      PFAMs = first_nonempty(PFAMs),
      .groups = "drop"
    )

  gff_annotation |>
    dplyr::full_join(eggnog_annotation, by = "gene_id") |>
    dplyr::mutate(
      Description = dplyr::coalesce(Description, "no GFF description"),
      gene_biotype = dplyr::coalesce(gene_biotype, "unknown"),
      annotation_text = stringr::str_squish(paste(
        Description, eggNOG_Description, eggNOG_name, PFAMs,
        KEGG_Pathway, sep = " | "
      ))
    )
}

classify_physiology_theme <- function(annotation_text) {
  text <- stringr::str_to_lower(dplyr::coalesce(annotation_text, ""))
  dplyr::case_when(
    stringr::str_detect(
      text,
      "antimicrobial|defensin|attacin|cecropin|diptericin|gloverin|metchnikowin|thaumatin|termicin|lysozyme"
    ) ~ "Antimicrobial peptides",
    stringr::str_detect(
      text,
      "toll|imd|relish|myd88|cactus|dorsal|nf-kappa|nf kappa|jak|stat"
    ) ~ "Immune-related",
    stringr::str_detect(
      text,
      "immune|peptidoglycan|pgrp|gram-negative|gram negative|beta-1,3-glucan|glucan binding|phenoloxidase|prophenoloxidase|melanization|lectin|hemocyanin|complement|serpin"
    ) ~ "Immune-related",
    stringr::str_detect(
      text,
      "glycogen|glycogenin|branching enzyme|starch"
    ) ~ "Glycogen synthesis / storage",
    stringr::str_detect(
      text,
      "lipid|fatty acid|fatty-acid|acyl-coa|acyl coa|acetyl-coa carboxylase|elongase|desaturase|glycerolipid|phospholipid|triacylglycerol|diacylglycerol|sterol|lipase"
    ) ~ "Lipid synthesis / metabolism",
    stringr::str_detect(
      text,
      "protein synthesis|translation|ribosom|elongation factor|initiation factor|aminoacyl|trna synthetase|vitellogenin|hexamerin|arylphorin|storage protein"
    ) ~ "Protein synthesis / storage",
    stringr::str_detect(
      text,
      "detox|cytochrome p450|\\bcyp[0-9a-z]|glutathione|gst|carboxylesterase|esterase|udp-glucuronosyltransferase|\\bugt\\b|abc transporter|oxidoreductase|peroxidase|catalase|superoxide|heat shock|stress"
    ) ~ "Detoxification / stress",
    stringr::str_detect(text, "no gff description|^\\s*$") ~
      "Other / unannotated genes",
    TRUE ~ "Other / unannotated genes"
  )
}

classify_gene_class <- function(gene_biotype) {
  dplyr::case_when(
    gene_biotype == "protein_coding" ~ "Coding genes",
    gene_biotype == "lncRNA" ~ "Long non-coding RNA",
    TRUE ~ "Other / unknown"
  )
}

make_condition_label <- function(comparison_type, numerator, denominator, diet,
                                 side = c("numerator", "denominator")) {
  side <- match.arg(side)
  value <- if (side == "numerator") numerator else denominator

  dplyr::case_when(
    comparison_type == "Infection within diet" ~ paste(value, "diet", diet),
    grepl("control", comparison_type, ignore.case = TRUE) ~
      paste("Control diet", value),
    grepl("infected", comparison_type, ignore.case = TRUE) ~
      paste("Infected diet", value),
    TRUE ~ value
  )
}

make_condition_hull <- function(df) {
  if (nrow(df) < 3) return(df[0, ])
  df[chull(df$PC1, df$PC2), ]
}

copy_publication_asset <- function(source, file_base) {
  if (!file.exists(source)) return(NA_character_)
  target <- file.path(out_dir, paste0(file_base, ".", tools::file_ext(source)))
  file.copy(source, target, overwrite = TRUE)
  target
}
```

## Current main-chromosome figure inputs

Every figure below is regenerated from the same main-chromosome contract. The
input table records the exact count matrix, official placement source, and
current DESeq2 or cluster output used for each export.

### Input data

```{r input-data}
placed_dir <- latest_dir(publication_params$primary_results_pattern)
primary_results_file <- file.path(
  placed_dir,
  "placed_unplaced_all_deseq2_results.csv"
)
head_phase_file <- file.path(
  project_dir,
  "data/external/phase_reference/gregaria_head_phase_degs.csv"
)
thorax_phase_file <- file.path(
  project_dir,
  "data/external/phase_reference/gregaria_thorax_phase_degs.csv"
)
circled_cluster_dir <- latest_dir(publication_params$cluster_results_pattern)
figure3_composite_source <- file.path(
  circled_cluster_dir,
  "expression_cluster_gene_level_GO_KEGG_pathway_illustration.png"
)
figure3_lollipop_source <- file.path(
  circled_cluster_dir,
  "expression_cluster_gene_level_GO_KEGG_lollipop.png"
)
figure3_insect_source <- file.path(
  circled_cluster_dir,
  "expression_cluster_gene_level_GO_KEGG_lollipop_insect_focused.png"
)
figure3_reference_source <- file.path(
  circled_cluster_dir,
  "expression_cluster_gene_level_GO_KEGG_lollipop_with_references.png"
)
curated_target_dir <- latest_dir(publication_params$curated_results_pattern)
figure4_curated_source <- file.path(
  curated_target_dir,
  "figure4_curated_targets_main_chromosomes.png"
)
figure4_placement_source <- file.path(
  curated_target_dir,
  "curated_target_placement_impact.png"
)
figure4_all_scaffolds_source <- file.path(
  curated_target_dir,
  "figure4_curated_targets_all_scaffolds.png"
)
figure4_homology_flags_source <- file.path(
  curated_target_dir,
  "figure4_curated_targets_non_grasshopper_homology_flags.png"
)

base_inputs <- c(
  metadata_file, counts_file, placement_file, rrna_file, gff_file, eggnog_file,
  primary_results_file, head_phase_file, thorax_phase_file,
  figure3_composite_source, figure3_lollipop_source, figure3_insect_source,
  figure3_reference_source, figure4_curated_source, figure4_all_scaffolds_source,
  figure4_homology_flags_source, figure4_placement_source
)
if (!all(file.exists(base_inputs))) {
  stop("Missing publication input(s): ",
       paste(base_inputs[!file.exists(base_inputs)], collapse = ", "))
}

placement <- read.delim(placement_file, check.names = FALSE) |>
  dplyr::distinct(gene_id, .keep_all = TRUE)
rrna_ids <- trimws(readLines(rrna_file, warn = FALSE))
rrna_ids <- rrna_ids[nzchar(rrna_ids) & !startsWith(rrna_ids, "#")]
main_chromosome_ids <- placement$gene_id[
  placement$sequence_class == "Nuclear chromosome" &
    !placement$gene_id %in% rrna_ids
]

counts_table <- read.delim(counts_file, check.names = FALSE)
main_chromosome_counts <- as.matrix(
  counts_table[
    counts_table$gene_id %in% main_chromosome_ids,
    setdiff(names(counts_table), "gene_id"),
    drop = FALSE
  ]
)
rownames(main_chromosome_counts) <- counts_table$gene_id[
  counts_table$gene_id %in% main_chromosome_ids
]
storage.mode(main_chromosome_counts) <- "integer"
main_chromosome_counts <- main_chromosome_counts[
  rowSums(main_chromosome_counts >= publication_params$min_count) >=
    publication_params$min_samples,
  ,
  drop = FALSE
]

metadata_for_models <- read.delim(metadata_file, check.names = FALSE)
metadata_for_models <- metadata_for_models[
  , !is.na(names(metadata_for_models)) & nzchar(names(metadata_for_models)),
  drop = FALSE
]
metadata_for_models$label <- as.character(metadata_for_models$label)
metadata_for_models$Diet <- factor(metadata_for_models$Diet,
                                   levels = c("33", "50", "83"))
metadata_for_models$Treatment <- factor(
  metadata_for_models$Treatment,
  levels = c("Control", "Infected")
)
metadata_for_models <- metadata_for_models[
  match(colnames(main_chromosome_counts), metadata_for_models$label),
  ,
  drop = FALSE
]
if (anyNA(metadata_for_models$label)) {
  stop("Fresh competitive counts contain samples missing from metadata.")
}
rownames(metadata_for_models) <- metadata_for_models$label
metadata_for_models$group <- factor(
  paste0("diet", metadata_for_models$Diet, "_",
         tolower(as.character(metadata_for_models$Treatment))),
  levels = c(
    "diet33_control", "diet50_control", "diet83_control",
    "diet33_infected", "diet50_infected", "diet83_infected"
  )
)

direct_contrasts <- tibble::tribble(
  ~contrast_id, ~comparison_type, ~comparison, ~numerator, ~denominator,
  ~diet, ~numerator_group, ~denominator_group,
  "infection_diet33", "Infection within diet", "Diet 33: infected vs control", "Infected", "Control", "33", "diet33_infected", "diet33_control",
  "infection_diet50", "Infection within diet", "Diet 50: infected vs control", "Infected", "Control", "50", "diet50_infected", "diet50_control",
  "infection_diet83", "Infection within diet", "Diet 83: infected vs control", "Infected", "Control", "83", "diet83_infected", "diet83_control",
  "control_diet33_vs50", "Diet contrasts among control", "Control 33 vs 50", "33", "50", NA_character_, "diet33_control", "diet50_control",
  "control_diet83_vs33", "Diet contrasts among control", "Control 83 vs 33", "83", "33", NA_character_, "diet83_control", "diet33_control",
  "control_diet83_vs50", "Diet contrasts among control", "Control 83 vs 50", "83", "50", NA_character_, "diet83_control", "diet50_control",
  "infected_diet33_vs50", "Diet contrasts among infected", "Infected 33 vs 50", "33", "50", NA_character_, "diet33_infected", "diet50_infected",
  "infected_diet83_vs33", "Diet contrasts among infected", "Infected 83 vs 33", "83", "33", NA_character_, "diet83_infected", "diet33_infected",
  "infected_diet83_vs50", "Diet contrasts among infected", "Infected 83 vs 50", "83", "50", NA_character_, "diet83_infected", "diet50_infected"
)

primary_results <- read.csv(primary_results_file, check.names = FALSE) |>
  dplyr::filter(
    mapping_branch == "Competitive host",
    analysis_scope == "Placed/host-assigned",
    sequence_class == "Nuclear chromosome"
  )

all_sample_results <- primary_results |>
  dplyr::inner_join(
    direct_contrasts |>
      dplyr::select(contrast_id, comparison_type, comparison,
                    numerator_simple = numerator,
                    denominator_simple = denominator, diet),
    by = "contrast_id"
  ) |>
  dplyr::transmute(
    scenario = "All samples",
    comparison_type, comparison,
    numerator = numerator_simple,
    denominator = denominator_simple,
    diet,
    gene_id, baseMean, log2FoldChange, lfcSE, stat, pvalue, padj,
    significant = significant & !is.na(padj) &
      padj < publication_params$alpha &
        abs(log2FoldChange) >= publication_params$lfc_threshold,
    direction = dplyr::if_else(log2FoldChange > 0,
                               "Upregulated", "Downregulated")
  )

fit_direct_scenario <- function(excluded_samples, scenario_name) {
  keep_samples <- setdiff(colnames(main_chromosome_counts), excluded_samples)
  scenario_counts <- main_chromosome_counts[, keep_samples, drop = FALSE]
  scenario_counts <- scenario_counts[
    rowSums(scenario_counts >= publication_params$min_count) >=
      publication_params$min_samples,
    ,
    drop = FALSE
  ]
  scenario_meta <- metadata_for_models[keep_samples, , drop = FALSE]
  dds <- DESeq2::DESeqDataSetFromMatrix(
    countData = scenario_counts,
    colData = data.frame(group = scenario_meta$group,
                         row.names = scenario_meta$label),
    design = ~ 0 + group
  )
  dds <- DESeq2::DESeq(dds, minReplicatesForReplace = Inf, quiet = TRUE)

  dplyr::bind_rows(lapply(seq_len(nrow(direct_contrasts)), function(i) {
    spec <- direct_contrasts[i, ]
    result <- as.data.frame(DESeq2::results(
      dds,
      contrast = c("group", spec$numerator_group, spec$denominator_group),
      alpha = publication_params$alpha,
      cooksCutoff = FALSE
    )) |>
      tibble::rownames_to_column("gene_id")
    result |>
      dplyr::transmute(
        scenario = scenario_name,
        comparison_type = spec$comparison_type,
        comparison = spec$comparison,
        numerator = spec$numerator,
        denominator = spec$denominator,
        diet = spec$diet,
        gene_id, baseMean, log2FoldChange, lfcSE, stat, pvalue, padj,
        significant = !is.na(padj) & padj < publication_params$alpha &
          abs(log2FoldChange) >= publication_params$lfc_threshold,
        direction = dplyr::if_else(log2FoldChange > 0,
                                   "Upregulated", "Downregulated")
      )
  }))
}

# The current analysis retains every QC-passed library; only 1044 was removed
# upstream for technical mapping failure.
scenario_results <- all_sample_results
scenario_results_file <- file.path(
  out_dir,
  "main_chromosome_sample_filter_scenario_results.csv"
)
write.csv(scenario_results, scenario_results_file, row.names = FALSE)

scenario_sample_counts <- metadata_for_models |>
  dplyr::count(Diet, Treatment, name = "n_samples") |>
  dplyr::mutate(scenario = "All QC-passed samples")
scenario_sample_counts_file <- file.path(
  out_dir,
  "main_chromosome_sample_counts_by_scenario.csv"
)
write.csv(scenario_sample_counts, scenario_sample_counts_file,
          row.names = FALSE)

model_coefficients <- primary_results |>
  dplyr::filter(contrast_id == "infection_average_all_diets") |>
  dplyr::mutate(coefficient = "Treatment_Infected_vs_Control")
model_coefficients_file <- file.path(
  out_dir,
  "main_chromosome_average_infection_results.csv"
)
write.csv(model_coefficients, model_coefficients_file, row.names = FALSE)

inputs <- data.frame(
  input = c("metadata", "fresh competitive host counts",
            "official main-chromosome placement", "rRNA exclusion list",
            "GFF annotation", "eggNOG annotation",
            "main-chromosome DESeq2 scenario table",
            "sample-filter scenario counts",
            "main-chromosome average infection results", "Head phase DEGs",
            "Thorax phase DEGs", "Figure 3 composite source"),
  path = c(metadata_file, counts_file, placement_file, rrna_file,
           gff_file, eggnog_file,
           scenario_results_file, scenario_sample_counts_file,
           model_coefficients_file,
           head_phase_file, thorax_phase_file, figure3_composite_source),
  present = c(file.exists(metadata_file), file.exists(counts_file),
              file.exists(placement_file), file.exists(rrna_file),
              file.exists(gff_file), file.exists(eggnog_file),
              file.exists(scenario_results_file),
              file.exists(scenario_sample_counts_file),
              file.exists(model_coefficients_file),
              file.exists(head_phase_file), file.exists(thorax_phase_file),
              file.exists(figure3_composite_source))
)

knitr::kable(inputs, caption = "Inputs used to assemble publication figures")
if (!all(inputs$present)) {
  stop("One or more publication figure inputs are missing.")
}
```

## Standalone experimental-design asset

The design summary remains available as a separate export but is not included
in the numbered Figure 1 composite.

```{r figure-1-panel-a, fig.width=7.2, fig.height=4.6}
metadata <- read.delim(metadata_file, check.names = FALSE)
metadata <- metadata[, !is.na(names(metadata)) & nzchar(names(metadata)),
                     drop = FALSE]
metadata$label <- as.character(metadata$label)
metadata$Diet <- factor(metadata$Diet, levels = c("33", "50", "83"))
metadata$Treatment <- factor(metadata$Treatment,
                             levels = c("Control", "Infected"))

design_counts <- metadata |>
  dplyr::count(Diet, Treatment, name = "n_samples") |>
  tidyr::complete(Diet, Treatment, fill = list(n_samples = 0)) |>
  dplyr::mutate(
    Condition = paste(Diet, Treatment),
    label = paste0("Diet ", Diet, "\n", Treatment, "\nn = ", n_samples),
    text_color = dplyr::if_else(Treatment == "Infected", "white", "#111827")
  )

p_design <- ggplot2::ggplot(
  design_counts,
  ggplot2::aes(x = Treatment, y = Diet, fill = Condition)
) +
  ggplot2::geom_tile(width = 0.84, height = 0.72, color = "#111827",
                     linewidth = 0.45) +
  ggplot2::geom_text(
    ggplot2::aes(label = label, color = text_color),
    lineheight = 0.88,
    size = 3.6,
    fontface = "bold"
  ) +
  ggplot2::annotate(
    "label",
    x = 1.5, y = 4.18,
    label = paste(
      "Female S. gregaria fat body RNA-seq",
      "96 h after control or Metarhizium challenge",
      "all analysis-cohort samples retained; rRNA loci removed bioinformatically",
      sep = "\n"
    ),
    label.size = 0.25,
    fill = "white",
    color = "#111827",
    size = 3.5,
    fontface = "bold",
    lineheight = 0.96
  ) +
  ggplot2::scale_fill_manual(values = condition_colors, drop = FALSE) +
  ggplot2::scale_color_identity() +
  ggplot2::scale_y_discrete(limits = rev(levels(metadata$Diet))) +
  ggplot2::coord_cartesian(clip = "off", ylim = c(0.55, 4.45)) +
  theme_pub(base_size = 12) +
  ggplot2::theme(
    axis.title = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_text(face = "bold", size = 12),
    axis.text.y = ggplot2::element_text(face = "bold", size = 12),
    axis.ticks = ggplot2::element_blank(),
    legend.position = "none",
    plot.margin = ggplot2::margin(8, 8, 8, 8)
  ) +
  ggplot2::labs(title = "Experimental design")

save_png_svg(p_design, "experimental_design_standalone",
             width = 7.2, height = 4.6)
p_design
```

## Figure 1: fat body transcriptome PCA

Figure 1 shows the VST PCA from all expressed, chromosome-assigned,
rRNA-filtered genes. Pale hulls show the spread of each diet-by-treatment
group. Sample color identifies diet, shape identifies treatment, and the hollow
centroids are connected within each diet to show the control-to-infected shift.

```{r figure-1-panel-b, fig.width=7.2, fig.height=5.2}
counts <- main_chromosome_counts
count_labels_all <- colnames(counts)
keep_samples <- count_labels_all %in% metadata$label
counts <- counts[, keep_samples, drop = FALSE]
count_labels <- count_labels_all[keep_samples]
metadata_ordered <- metadata[match(count_labels, metadata$label), ]
stopifnot(identical(metadata_ordered$label, count_labels))
rownames(metadata_ordered) <- metadata_ordered$label
colnames(counts) <- count_labels

keep_vst <- rowSums(counts) >= publication_params$min_total_count
dds_vst <- DESeq2::DESeqDataSetFromMatrix(
  countData = counts[keep_vst, , drop = FALSE],
  colData = metadata_ordered,
  design = ~ Diet + Treatment
)
vsd <- DESeq2::vst(dds_vst, blind = TRUE)
vst_mat <- SummarizedExperiment::assay(vsd)
nonzero_variance <- apply(vst_mat, 1, var) > 0
pca <- prcomp(t(vst_mat[nonzero_variance, , drop = FALSE]), scale. = FALSE)
percent_var <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)

pca_scores <- data.frame(
  label = rownames(pca$x),
  PC1 = pca$x[, "PC1"],
  PC2 = pca$x[, "PC2"],
  Diet = metadata_ordered$Diet,
  Treatment = metadata_ordered$Treatment,
  check.names = FALSE
)
pca_scores$Condition <- factor(
  paste(pca_scores$Diet, pca_scores$Treatment),
  levels = names(condition_colors)
)

pca_centroids <- aggregate(cbind(PC1, PC2) ~ Diet + Treatment + Condition,
                           pca_scores, mean)
pca_centroids$centroid_label <- paste(
  pca_centroids$Diet,
  pca_centroids$Treatment
)
pca_hulls <- do.call(rbind, lapply(split(pca_scores, pca_scores$Condition),
                                   make_condition_hull))
centroid_shift <- merge(
  pca_centroids[pca_centroids$Treatment == "Control",
                c("Diet", "PC1", "PC2")],
  pca_centroids[pca_centroids$Treatment == "Infected",
                c("Diet", "PC1", "PC2")],
  by = "Diet",
  suffixes = c("_control", "_infected")
)

p_pca <- ggplot2::ggplot() +
  ggplot2::geom_polygon(
    data = pca_hulls,
    ggplot2::aes(x = PC1, y = PC2, fill = Condition, group = Condition),
    alpha = 0.18,
    color = NA
  ) +
  ggplot2::geom_point(
    data = pca_scores,
    ggplot2::aes(x = PC1, y = PC2, color = Diet, shape = Treatment),
    size = 5,
    stroke = 0.2,
    alpha = 0.9
  ) +
  ggplot2::geom_segment(
    data = centroid_shift,
    ggplot2::aes(x = PC1_control, y = PC2_control,
                 xend = PC1_infected, yend = PC2_infected,
                 color = Diet),
    linewidth = 0.8,
    arrow = grid::arrow(length = grid::unit(0.16, "cm"), type = "closed")
  ) +
  ggplot2::geom_point(
    data = pca_centroids,
    ggplot2::aes(x = PC1, y = PC2, color = Diet),
    shape = 21,
    fill = "white",
    stroke = 1.2,
    size = 5
  ) +
  ggrepel::geom_label_repel(
    data = pca_centroids,
    ggplot2::aes(
      x = PC1, y = PC2, label = centroid_label,
      color = Diet
    ),
    fill = "white",
    label.size = 0.25,
    label.padding = grid::unit(0.18, "lines"),
    box.padding = 0.45,
    point.padding = 0.55,
    min.segment.length = 0,
    segment.color = "#4B5563",
    segment.size = 0.3,
    size = 5,
    show.legend = FALSE,
    seed = 20260807
  ) +
  ggplot2::scale_fill_manual(values = condition_colors, drop = FALSE) +
  ggplot2::scale_color_manual(values = diet_colors, drop = FALSE) +
  ggplot2::scale_shape_manual(values = c("Control" = 16, "Infected" = 17)) +
  theme_pub(base_size = 12) +
  ggplot2::theme(
    legend.position = "bottom",
    legend.box = "vertical",
    legend.text = ggplot2::element_text(size = 11.5),
    legend.title = ggplot2::element_text(size = 12, face = "bold"),
    axis.title = ggplot2::element_text(size = 18),
    axis.text = ggplot2::element_text(size = 14, color = "#111827"),
    panel.grid = ggplot2::element_blank()
  ) +
  ggplot2::guides(
    fill = ggplot2::guide_legend(
      nrow = 2, title.position = "top", order = 3,
      override.aes = list(alpha = 0.22, color = NA)
    ),
    color = ggplot2::guide_legend(
      nrow = 1, title.position = "top", order = 2,
      override.aes = list(shape = 21, fill = "white", size = 4.2)
    ),
    shape = ggplot2::guide_legend(
      nrow = 1, title.position = "top", order = 1,
      override.aes = list(color = "#111827", size = 3.8)
    )
  ) +
  ggplot2::labs(
    title = "VST PCA with diet-by-treatment centroids",
    x = paste0("PC1: ", percent_var[1], "% variance"),
    y = paste0("PC2: ", percent_var[2], "% variance"),
    fill = "Condition polygon",
    color = "Diet",
    shape = "Treatment"
  )

write.csv(pca_scores, file.path(out_dir, "figure1_pca_scores.csv"),
          row.names = FALSE)
write.csv(pca_centroids, file.path(out_dir, "figure1_pca_centroids.csv"),
          row.names = FALSE)
write.csv(centroid_shift, file.path(out_dir,
                                    "figure1_pca_centroid_shifts.csv"),
          row.names = FALSE)
save_png_svg(p_pca, "figure1_fatbody_pca_centroid_shifts",
             width = 10.2, height = 7.2)
p_pca
```

## Figure 2: DEG burden, context overlap, and phase-reference overlap

Panel A reports the complete directional DEG burden for every primary
contrast. Panel B compares infection-response DEGs called independently in
diets 33, 50, and 83, making shared and diet-specific responses directly
visible. Venn membership ignores direction. Panel C compares the
infection-within-diet DEG union with independent phase DEGs detected between
solitarious and gregarious locusts in head and thorax. Every set in these panels
is restricted to genes on the official nuclear chromosomes. The broader
four-context Venn is retained as a standalone alternative export.

### A. Directional DEG burden

```{r figure-1-panel-c, fig.width=11.2, fig.height=8.4}
gene_biotypes <- parse_gff_gene_annotation(gff_file)

scenario_degs_for_barplot <- read.csv(scenario_results_file, check.names = FALSE) |>
  dplyr::filter(
    scenario == "All samples",
    significant,
    !is.na(gene_id),
    !is.na(padj),
    padj < publication_params$alpha,
    abs(log2FoldChange) >= publication_params$lfc_threshold
  )

model_degs_for_barplot <- read.csv(model_coefficients_file,
                                   check.names = FALSE) |>
  dplyr::filter(
    coefficient == "Treatment_Infected_vs_Control",
    significant,
    !is.na(gene_id),
    !is.na(padj),
    padj < publication_params$alpha,
    abs(log2FoldChange) >= publication_params$lfc_threshold
  )

deg_raw <- dplyr::bind_rows(
  scenario_degs_for_barplot |>
    dplyr::transmute(
      DEG_context = dplyr::case_when(
        comparison_type == "Infection within diet" ~
          "Infection within diets",
        comparison_type == "Diet contrasts among control" ~
          "Diet DEGs in controls",
        comparison_type == "Diet contrasts among infected" ~
          "Diet DEGs in infected",
        TRUE ~ comparison_type
      ),
      comparison_type,
      comparison,
      numerator,
      denominator,
      diet = as.character(diet),
      gene_id,
      log2FoldChange
    ),
  model_degs_for_barplot |>
    dplyr::transmute(
      DEG_context = "Infection vs control",
      comparison_type = "Infection vs control",
      comparison = "All diets: infected vs control",
      numerator = "Infected",
      denominator = "Control",
      diet = NA_character_,
      gene_id,
      log2FoldChange
    )
)

context_order <- c(
  "Infection vs control",
  "Infection within diets",
  "Diet DEGs in controls",
  "Diet DEGs in infected"
)

contrast_order <- c(
  "All diets\ninfected vs control",
  "Diet 83\ninfected vs control",
  "Diet 50\ninfected vs control",
  "Diet 33\ninfected vs control",
  "Control\n83 vs 50",
  "Control\n83 vs 33",
  "Control\n33 vs 50",
  "Infected\n83 vs 50",
  "Infected\n83 vs 33",
  "Infected\n33 vs 50"
)

deg_biotype_counts <- deg_raw |>
  dplyr::mutate(
    signed_direction = dplyr::if_else(log2FoldChange > 0,
                                      "Upregulated", "Downregulated"),
    numerator_condition = make_condition_label(
      comparison_type, numerator, denominator, diet, "numerator"
    ),
    denominator_condition = make_condition_label(
      comparison_type, numerator, denominator, diet, "denominator"
    ),
    higher_in = dplyr::if_else(log2FoldChange > 0,
                               numerator_condition, denominator_condition),
    contrast_label = dplyr::case_when(
      DEG_context == "Infection within diets" ~
        paste0("Diet ", diet, "\ninfected vs control"),
      DEG_context == "Diet DEGs in controls" ~
        paste0("Control\n", numerator, " vs ", denominator),
      DEG_context == "Diet DEGs in infected" ~
        paste0("Infected\n", numerator, " vs ", denominator),
      DEG_context == "Infection vs control" ~
        "All diets\ninfected vs control",
      TRUE ~ comparison
    )
  ) |>
  dplyr::left_join(gene_biotypes, by = "gene_id") |>
  dplyr::mutate(
    Gene_class = classify_gene_class(gene_biotype),
    DEG_context = factor(DEG_context, levels = context_order),
    contrast_label = factor(contrast_label, levels = contrast_order),
    signed_direction = factor(signed_direction,
                              levels = c("Downregulated", "Upregulated")),
    direction_class = paste(signed_direction, Gene_class, sep = " - ")
  ) |>
  dplyr::distinct(DEG_context, comparison, contrast_label, signed_direction,
                  gene_id, Gene_class, .keep_all = TRUE) |>
  dplyr::count(DEG_context, comparison, contrast_label,
               signed_direction, Gene_class,
               direction_class, name = "n_DEGs") |>
  dplyr::group_by(DEG_context, comparison, contrast_label, signed_direction) |>
  tidyr::complete(
    Gene_class = c("Coding genes", "Long non-coding RNA", "Other / unknown"),
    fill = list(n_DEGs = 0)
  ) |>
  dplyr::ungroup() |>
  dplyr::filter(!is.na(DEG_context), !is.na(contrast_label)) |>
  dplyr::mutate(
    direction_class = paste(signed_direction, Gene_class, sep = " - "),
    signed_n = dplyr::if_else(signed_direction == "Upregulated",
                              n_DEGs, -n_DEGs)
  )

deg_total_labels <- deg_biotype_counts |>
  dplyr::group_by(DEG_context, comparison, contrast_label, signed_direction) |>
  dplyr::summarise(total_DEGs = sum(n_DEGs), .groups = "drop") |>
  dplyr::mutate(
    signed_total = dplyr::if_else(signed_direction == "Upregulated",
                                  total_DEGs, -total_DEGs)
  )

context_max <- deg_total_labels |>
  dplyr::group_by(DEG_context) |>
  dplyr::summarise(max_total = max(abs(signed_total), 1),
                   .groups = "drop")

deg_total_labels <- deg_total_labels |>
  dplyr::left_join(context_max, by = "DEG_context") |>
  dplyr::filter(total_DEGs > 0) |>
  dplyr::mutate(
    label_x = signed_total + dplyr::if_else(
      signed_total > 0,
      max_total * 0.025,
      -max_total * 0.025
    ),
    hjust = dplyr::if_else(signed_total > 0, 0, 1)
  )

p_deg_biotype <- ggplot2::ggplot(
  deg_biotype_counts,
  ggplot2::aes(x = signed_n, y = contrast_label, fill = direction_class)
) +
  ggplot2::geom_vline(xintercept = 0, linewidth = 0.35, color = "#111827") +
  ggplot2::geom_col(width = 0.74, color = "white", linewidth = 0.2) +
  ggplot2::geom_text(
    data = deg_total_labels,
    ggplot2::aes(x = label_x, y = contrast_label, label = total_DEGs,
                 hjust = hjust),
    inherit.aes = FALSE,
    size = 6,
    fontface = "bold"
  ) +
  ggplot2::facet_wrap(~ DEG_context, ncol = 2, scales = "free") +
  ggplot2::scale_fill_manual(values = direction_biotype_colors, drop = FALSE) +
  ggplot2::scale_x_continuous(
    labels = abs,
    expand = ggplot2::expansion(mult = c(0.22, 0.22))
  ) +
  theme_pub(base_size = 16) +
  ggplot2::theme(
    axis.title.y = ggplot2::element_blank(),
    axis.text.y = ggplot2::element_text(size = 15.5, face = "bold"),
    axis.text.x = ggplot2::element_text(size = 13.5),
    axis.title.x = ggplot2::element_text(size = 15, face = "bold"),
    strip.text = ggplot2::element_text(size = 16, face = "bold"),
    panel.spacing = grid::unit(0.55, "lines"),
    legend.position = "bottom",
    legend.title = ggplot2::element_text(size = 13.5, face = "bold"),
    legend.text = ggplot2::element_text(size = 12),
    plot.margin = ggplot2::margin(6, 28, 6, 6)
  ) +
  ggplot2::guides(
    fill = ggplot2::guide_legend(nrow = 3, title.position = "top",
                                 byrow = TRUE)
  ) +
  ggplot2::labs(
    x = "Number of significant DEGs",
    fill = "Direction and gene class"
  )

write.csv(deg_biotype_counts,
          file.path(out_dir, "figure2_panel_A_deg_biotype_counts.csv"),
          row.names = FALSE)
write.csv(deg_raw,
          file.path(out_dir, "figure2_panel_A_exact_contrast_deg_gene_table.csv"),
          row.names = FALSE)
save_png_svg(p_deg_biotype, "figure2_panel_A_deg_biotype_burden",
             width = 11.2, height = 8.4)
p_deg_biotype
```

### B-C. DEG-context and phase-reference Venn overlaps

The Venn diagram is used in the primary composite. The ten-set exact-contrast
UpSet plot is retained as a standalone alternative export.

```{r figure-2-overlap-composition, fig.width=14, fig.height=13, results='asis'}
scenario_degs <- read.csv(scenario_results_file, check.names = FALSE) |>
  dplyr::filter(
    scenario == "All samples",
    significant,
    !is.na(gene_id),
    !is.na(padj),
    padj < publication_params$alpha,
    abs(log2FoldChange) >= publication_params$lfc_threshold
  )

model_degs <- read.csv(model_coefficients_file, check.names = FALSE) |>
  dplyr::filter(
    coefficient == "Treatment_Infected_vs_Control",
    significant,
    !is.na(gene_id),
    !is.na(padj),
    padj < publication_params$alpha,
    abs(log2FoldChange) >= publication_params$lfc_threshold
  )

exact_contrast_order <- c(
  "Diet 33: infected vs control",
  "Diet 50: infected vs control",
  "Diet 83: infected vs control",
  "Control: diet 33 vs 50",
  "Control: diet 83 vs 33",
  "Control: diet 83 vs 50",
  "Infected: diet 33 vs 50",
  "Infected: diet 83 vs 33",
  "Infected: diet 83 vs 50",
  "All diets: infected vs control"
)

exact_deg_long <- deg_raw |>
  dplyr::mutate(
    exact_contrast = dplyr::case_when(
      DEG_context == "Infection within diets" ~
        paste0("Diet ", diet, ": infected vs control"),
      DEG_context == "Diet DEGs in controls" ~
        paste0("Control: diet ", numerator, " vs ", denominator),
      DEG_context == "Diet DEGs in infected" ~
        paste0("Infected: diet ", numerator, " vs ", denominator),
      DEG_context == "Infection vs control" ~
        "All diets: infected vs control",
      TRUE ~ comparison
    )
  ) |>
  dplyr::filter(exact_contrast %in% exact_contrast_order) |>
  dplyr::distinct(exact_contrast, gene_id)

exact_deg_sets <- lapply(exact_contrast_order, function(contrast_name) {
  exact_deg_long$gene_id[exact_deg_long$exact_contrast == contrast_name]
})
names(exact_deg_sets) <- exact_contrast_order
exact_deg_sets <- exact_deg_sets[lengths(exact_deg_sets) > 0]

exact_overlap_genes <- sort(unique(unlist(exact_deg_sets,
                                          use.names = FALSE)))
exact_overlap_membership <- data.frame(gene_id = exact_overlap_genes)
for (set_name in names(exact_deg_sets)) {
  exact_overlap_membership[[set_name]] <-
    exact_overlap_membership$gene_id %in% exact_deg_sets[[set_name]]
}
write.csv(
  exact_overlap_membership,
  file.path(out_dir, "figure2_panel_B_exact_contrast_upset_membership.csv"),
  row.names = FALSE
)

upset_input <- UpSetR::fromList(exact_deg_sets)
upset_object <- UpSetR::upset(
  upset_input,
  sets = rev(names(exact_deg_sets)),
  keep.order = TRUE,
  nsets = length(exact_deg_sets),
  nintersects = 30,
  order.by = c("freq", "degree"),
  decreasing = c(TRUE, FALSE),
  mb.ratio = c(0.62, 0.38),
  main.bar.color = "#8B0000",
  sets.bar.color = "#D9A21B",
  matrix.color = "#374151",
  shade.color = "#E5E7EB",
  shade.alpha = 0.55,
  point.size = 3.0,
  line.size = 0.9,
  text.scale = c(1.45, 1.25, 1.05, 1.15, 1.05, 1.15),
  mainbar.y.label = "Genes in exact intersections",
  sets.x.label = "DEGs per exact contrast"
)
upset_grob <- grid::grid.grabExpr(
  print(upset_object),
  wrap = TRUE,
  wrap.grobs = TRUE
)

p_upset <- ggplotify::as.ggplot(upset_grob)

p_upset_heading <- ggplot2::ggplot() +
  ggplot2::annotate(
    "text", x = 0, y = 0.72,
    label = "Alternative exact-contrast DEG overlap",
    hjust = 0,
    fontface = "bold",
    size = 5.2
  ) +
  ggplot2::annotate(
    "text", x = 0, y = 0.20,
    label = "The 30 largest intersections are shown; set membership ignores direction",
    hjust = 0,
    size = 3.7
  ) +
  ggplot2::xlim(0, 1) +
  ggplot2::ylim(0, 1) +
  ggplot2::theme_void() +
  ggplot2::theme(plot.margin = ggplot2::margin(2, 8, 0, 8))

p_upset_panel <- p_upset_heading / p_upset +
  patchwork::plot_layout(heights = c(0.10, 0.90))

save_png_svg(
  p_upset_panel,
  "figure2_panel_B_exact_contrast_upset",
  width = 14.5,
  height = 8.6
)

broad_deg_sets <- list(
  "Infection within diets" = scenario_degs |>
    dplyr::filter(comparison_type == "Infection within diet") |>
    dplyr::pull(gene_id) |>
    unique(),
  "Diet DEGs in controls" = scenario_degs |>
    dplyr::filter(comparison_type == "Diet contrasts among control") |>
    dplyr::pull(gene_id) |>
    unique(),
  "Diet DEGs in infected" = scenario_degs |>
    dplyr::filter(comparison_type == "Diet contrasts among infected") |>
    dplyr::pull(gene_id) |>
    unique(),
  "Infection vs control" = unique(model_degs$gene_id)
)
broad_deg_sets <- broad_deg_sets[lengths(broad_deg_sets) > 0]

infection_diet_sets <- list(
  "Diet 33" = scenario_degs |>
    dplyr::filter(
      comparison_type == "Infection within diet",
      as.character(diet) == "33"
    ) |>
    dplyr::pull(gene_id) |>
    unique(),
  "Diet 50" = scenario_degs |>
    dplyr::filter(
      comparison_type == "Infection within diet",
      as.character(diet) == "50"
    ) |>
    dplyr::pull(gene_id) |>
    unique(),
  "Diet 83" = scenario_degs |>
    dplyr::filter(
      comparison_type == "Infection within diet",
      as.character(diet) == "83"
    ) |>
    dplyr::pull(gene_id) |>
    unique()
)
infection_diet_sets <- infection_diet_sets[lengths(infection_diet_sets) > 0]

infection_diet_genes <- sort(unique(unlist(infection_diet_sets,
                                            use.names = FALSE)))
infection_diet_membership <- data.frame(gene_id = infection_diet_genes)
for (set_name in names(infection_diet_sets)) {
  infection_diet_membership[[set_name]] <-
    infection_diet_membership$gene_id %in% infection_diet_sets[[set_name]]
}
infection_diet_membership$intersection <- apply(
  infection_diet_membership[names(infection_diet_sets)],
  1,
  function(x) paste(names(infection_diet_sets)[x], collapse = " + ")
)
write.csv(
  infection_diet_membership,
  file.path(out_dir, "figure2_infection_diet_venn_membership.csv"),
  row.names = FALSE
)

infection_diet_summary <- infection_diet_membership |>
  dplyr::count(intersection, name = "n_genes") |>
  dplyr::arrange(dplyr::desc(n_genes), intersection)
write.csv(
  infection_diet_summary,
  file.path(out_dir, "figure2_infection_diet_venn_summary.csv"),
  row.names = FALSE
)

all_overlap_genes <- sort(unique(unlist(broad_deg_sets, use.names = FALSE)))
overlap_membership <- data.frame(gene_id = all_overlap_genes)
for (set_name in names(broad_deg_sets)) {
  overlap_membership[[set_name]] <- overlap_membership$gene_id %in%
    broad_deg_sets[[set_name]]
}
overlap_membership$intersection <- apply(
  overlap_membership[names(broad_deg_sets)],
  1,
  function(x) paste(names(broad_deg_sets)[x], collapse = " + ")
)
write.csv(overlap_membership,
          file.path(out_dir, "figure2_deg_family_overlap_membership.csv"),
          row.names = FALSE)

overlap_summary <- overlap_membership |>
  dplyr::count(intersection, name = "n_genes") |>
  dplyr::arrange(dplyr::desc(n_genes), intersection)
write.csv(overlap_summary,
          file.path(out_dir, "figure2_deg_family_overlap_summary.csv"),
          row.names = FALSE)

head_phase_degs <- read.csv(head_phase_file, check.names = FALSE) |>
  dplyr::filter(
    significant,
    gene_id %in% main_chromosome_ids,
    !is.na(gene_id),
    !is.na(padj),
    padj < publication_params$alpha,
    abs(log2FoldChange) >= publication_params$lfc_threshold
  )
thorax_phase_degs <- read.csv(thorax_phase_file, check.names = FALSE) |>
  dplyr::filter(
    significant,
    gene_id %in% main_chromosome_ids,
    !is.na(gene_id),
    !is.na(padj),
    padj < publication_params$alpha,
    abs(log2FoldChange) >= publication_params$lfc_threshold
  )
head_phase_genes <- unique(head_phase_degs$gene_id)
thorax_phase_genes <- unique(thorax_phase_degs$gene_id)
phase_genes <- intersect(head_phase_genes, thorax_phase_genes)

phase_overlap_sets <- list(
  "Infection within diets" = broad_deg_sets[["Infection within diets"]],
  "Head phase DEGs" = head_phase_genes,
  "Thorax phase DEGs" = thorax_phase_genes
)

# Verify the exact gene sets displayed anywhere in Figure 2, including panel A.
figure2_gene_sets_for_rrna_audit <- c(
  list("Panel A: directional DEG burden" = unique(deg_raw$gene_id)),
  stats::setNames(exact_deg_sets, paste0("Exact contrast: ", names(exact_deg_sets))),
  stats::setNames(broad_deg_sets, paste0("Broad context: ", names(broad_deg_sets))),
  stats::setNames(infection_diet_sets,
                  paste0("Infection Venn: ", names(infection_diet_sets))),
  stats::setNames(phase_overlap_sets,
                  paste0("Phase Venn: ", names(phase_overlap_sets)))
)
figure2_rrna_audit <- dplyr::bind_rows(lapply(
  names(figure2_gene_sets_for_rrna_audit),
  function(set_name) {
    ids <- unique(figure2_gene_sets_for_rrna_audit[[set_name]])
    tibble::tibble(
      figure2_set = set_name,
      genes_plotted = length(ids),
      listed_rRNA_genes = sum(ids %in% rrna_ids)
    )
  }
))
write.csv(
  figure2_rrna_audit,
  file.path(out_dir, "figure2_rrna_exclusion_audit.csv"),
  row.names = FALSE
)
if (any(figure2_rrna_audit$listed_rRNA_genes > 0)) {
  stop("Figure 2 rRNA audit failed: at least one plotted set contains a listed rRNA gene.")
}

phase_overlap_genes <- sort(unique(unlist(phase_overlap_sets,
                                          use.names = FALSE)))
phase_overlap_membership <- data.frame(gene_id = phase_overlap_genes)
for (set_name in names(phase_overlap_sets)) {
  phase_overlap_membership[[set_name]] <-
    phase_overlap_membership$gene_id %in% phase_overlap_sets[[set_name]]
}
phase_overlap_membership$intersection <- apply(
  phase_overlap_membership[names(phase_overlap_sets)],
  1,
  function(x) paste(names(phase_overlap_sets)[x], collapse = " + ")
)
write.csv(phase_overlap_membership,
          file.path(out_dir, "figure2_infection_phase_overlap_membership.csv"),
          row.names = FALSE)

phase_overlap_summary <- phase_overlap_membership |>
  dplyr::count(intersection, name = "n_genes") |>
  dplyr::arrange(dplyr::desc(n_genes), intersection)
write.csv(phase_overlap_summary,
          file.path(out_dir, "figure2_infection_phase_overlap_summary.csv"),
          row.names = FALSE)

broad_deg_plot_sets <- broad_deg_sets
names(broad_deg_plot_sets) <- c(
  "Infection\nwithin diet",
  "Diet\ncontrols",
  "Diet\ninfected",
  "Infect.\nvs ctrl"
)

p_broad_context_venn <- ggVennDiagram::ggVennDiagram(
  broad_deg_plot_sets,
  label = "count",
  label_alpha = 0,
  edge_size = 0.75,
  set_size = 7.2,
  label_size = 7.2
) +
  ggplot2::scale_fill_gradient(low = "#FFF7D1", high = "#D9A21B") +
  ggplot2::theme_void(base_size = 12) +
  ggplot2::theme(
    legend.position = "none",
    plot.background = ggplot2::element_rect(fill = "white", color = NA),
    panel.background = ggplot2::element_rect(fill = "white", color = NA),
    plot.margin = ggplot2::margin(8, 24, 8, 24)
  ) +
  ggplot2::labs(
    fill = "Genes"
  )

save_png_svg(p_broad_context_venn, "figure2_broad_context_overlap_venn",
             width = 10.5, height = 9.0)

p_venn <- ggVennDiagram::ggVennDiagram(
  infection_diet_sets,
  label = "count",
  label_alpha = 0,
  edge_size = 0.75,
  set_size = 7.2,
  label_size = 7.2
) +
  ggplot2::scale_fill_gradient(low = "#EFF8F2", high = "#64B987") +
  ggplot2::theme_void(base_size = 12) +
  ggplot2::theme(
    legend.position = "none",
    plot.background = ggplot2::element_rect(fill = "white", color = NA),
    panel.background = ggplot2::element_rect(fill = "white", color = NA),
    plot.margin = ggplot2::margin(8, 24, 8, 24)
  ) +
  ggplot2::labs(fill = "Genes")

save_png_svg(p_venn, "figure2_infection_degs_by_diet_venn",
             width = 10.5, height = 9.0)

phase_overlap_plot_sets <- phase_overlap_sets
names(phase_overlap_plot_sets) <- c(
  "Infection\nDEGs",
  "Head\nphase",
  "Thorax\nphase"
)

p_phase_venn <- ggVennDiagram::ggVennDiagram(
  phase_overlap_plot_sets,
  label = "count",
  label_alpha = 0,
  edge_size = 0.75,
  set_size = 7.2,
  label_size = 7.2
) +
  ggplot2::scale_fill_gradient(low = "#FFF7D1", high = "#D9A21B") +
  ggplot2::theme_void(base_size = 12) +
  ggplot2::theme(
    legend.position = "none",
    plot.background = ggplot2::element_rect(fill = "white", color = NA),
    panel.background = ggplot2::element_rect(fill = "white", color = NA),
    plot.margin = ggplot2::margin(8, 24, 8, 24)
  ) +
  ggplot2::labs(
    fill = "Genes"
  )

save_png_svg(p_phase_venn, "figure2_infection_phase_overlap_venn",
             width = 10.5, height = 9.0)

gene_annotation <- build_gene_annotation(gff_file, eggnog_file)
deg_family_long <- dplyr::bind_rows(lapply(names(broad_deg_sets), function(set_name) {
  data.frame(
    DEG_family = set_name,
    gene_id = broad_deg_sets[[set_name]],
    stringsAsFactors = FALSE
  )
}))
deg_family_long$DEG_family <- factor(
  deg_family_long$DEG_family,
  levels = names(broad_deg_sets)
)

candidate_theme_table <- deg_family_long |>
  dplyr::left_join(gene_annotation, by = "gene_id") |>
  dplyr::mutate(
    annotation_text = dplyr::coalesce(annotation_text, ""),
    base_physiology_theme = classify_physiology_theme(annotation_text),
    phase_related = gene_id %in% phase_genes,
    phase_reference_source = dplyr::case_when(
      gene_id %in% head_phase_genes & gene_id %in% thorax_phase_genes ~
        "Head + thorax phase DEG",
      gene_id %in% head_phase_genes ~ "Head phase DEG",
      gene_id %in% thorax_phase_genes ~ "Thorax phase DEG",
      TRUE ~ "Not phase-related"
    ),
    physiology_theme = dplyr::case_when(
      base_physiology_theme != "Other / unannotated genes" ~
        base_physiology_theme,
      phase_related ~ "Phase-related changes",
      TRUE ~ base_physiology_theme
    ),
    physiology_theme = factor(physiology_theme,
                              levels = names(physiology_theme_colors))
  )

write.csv(candidate_theme_table,
          file.path(out_dir, "figure2_deg_family_candidate_theme_gene_table.csv"),
          row.names = FALSE)

composition_counts <- candidate_theme_table |>
  dplyr::distinct(DEG_family, gene_id, physiology_theme) |>
  dplyr::count(DEG_family, physiology_theme, name = "n_genes") |>
  dplyr::group_by(DEG_family) |>
  tidyr::complete(
    physiology_theme = factor(names(physiology_theme_colors),
                              levels = names(physiology_theme_colors)),
    fill = list(n_genes = 0)
  ) |>
  dplyr::mutate(
    total_genes = sum(n_genes),
    proportion = dplyr::if_else(total_genes > 0, n_genes / total_genes, 0)
  ) |>
  dplyr::ungroup()

write.csv(composition_counts,
          file.path(out_dir, "figure2_deg_family_candidate_theme_counts.csv"),
          row.names = FALSE)

candidate_theme_levels <- names(physiology_theme_colors)
candidate_composition_counts <- composition_counts |>
  dplyr::filter(physiology_theme %in% candidate_theme_levels) |>
  dplyr::group_by(DEG_family) |>
  dplyr::mutate(
    candidate_genes = sum(n_genes),
    candidate_proportion = dplyr::if_else(candidate_genes > 0,
                                          n_genes / candidate_genes, 0)
  ) |>
  dplyr::ungroup()

write.csv(candidate_composition_counts,
          file.path(out_dir,
                    "figure2_deg_family_candidate_theme_counts_all_categories.csv"),
          row.names = FALSE)

family_label_levels <- candidate_composition_counts |>
  dplyr::distinct(DEG_family, total_genes, candidate_genes) |>
  dplyr::arrange(DEG_family) |>
  dplyr::mutate(
    family_label = paste0(
      DEG_family,
      "\nall DEG n = ", total_genes
    )
  ) |>
  dplyr::pull(family_label)

pie_counts <- candidate_composition_counts |>
  dplyr::filter(n_genes > 0) |>
  dplyr::mutate(
    family_label = paste0(
      DEG_family,
      "\nall DEG n = ", total_genes
    ),
    family_label = factor(family_label, levels = family_label_levels)
  )

p_composition <- ggplot2::ggplot(
  pie_counts,
  ggplot2::aes(x = "", y = candidate_proportion,
               fill = physiology_theme)
) +
  ggplot2::geom_col(width = 1, color = "white", linewidth = 0.4) +
  ggplot2::coord_polar(theta = "y") +
  ggplot2::facet_wrap(~ family_label, nrow = 1) +
  ggplot2::scale_fill_manual(
    values = physiology_theme_colors[candidate_theme_levels],
    breaks = candidate_theme_levels,
    drop = FALSE
  ) +
  ggplot2::theme_void(base_size = 13) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold", size = 16),
    plot.subtitle = ggplot2::element_text(size = 11),
    strip.background = ggplot2::element_rect(fill = "#F3F4F6",
                                             color = "#111827"),
    strip.text = ggplot2::element_text(size = 10.5, face = "bold",
                                       lineheight = 0.95),
    legend.position = "bottom",
    legend.text = ggplot2::element_text(size = 9),
    legend.title = ggplot2::element_text(face = "bold"),
    plot.margin = ggplot2::margin(4, 8, 8, 8)
  ) +
  ggplot2::guides(
    fill = ggplot2::guide_legend(nrow = 3, byrow = TRUE,
                                 title.position = "top")
  ) +
  ggplot2::labs(
    title = "Standalone DEG physiology and phase-reference composition",
    subtitle = "Immune and physiology annotations are prioritized; phase-related marks remaining DEGs common to head+thorax phase references",
    fill = "Gene category"
  )

save_png_svg(p_composition,
             "figure2_panel_C_candidate_physiology_phase_composition",
             width = 12, height = 5.6)

# Reuse the complete standalone PNG canvases in the composite. ggVennDiagram
# places set labels beyond the circle bounds, which patchwork otherwise clips
# when it resizes the live ggplot objects.
p_venn_composite <- patchwork::wrap_elements(
  full = grid::rasterGrob(
    png::readPNG(file.path(out_dir, "figure2_infection_degs_by_diet_venn.png")),
    interpolate = TRUE
  )
)
p_phase_venn_composite <- patchwork::wrap_elements(
  full = grid::rasterGrob(
    png::readPNG(file.path(out_dir, "figure2_infection_phase_overlap_venn.png")),
    interpolate = TRUE
  )
)

figure2 <- p_deg_biotype / (p_venn_composite | p_phase_venn_composite) +
  patchwork::plot_layout(heights = c(0.40, 0.60))

save_png_svg(figure2, "figure2_deg_burden_context_phase_venns",
             width = 20.5, height = 19.0)
figure2

figure2_exports <- data.frame(
  figure = c("Figure 2 composite PNG", "Figure 2 composite SVG",
             "Figure 2A DEG burden PNG",
             "Figure 2B infection DEGs by diet Venn PNG",
             "Figure 2B infection DEGs by diet Venn SVG",
             "Figure 2B infection-by-diet membership table",
             "Figure 2C infection-phase Venn PNG",
             "Figure 2C infection-phase Venn SVG",
             "Alternative broad-context Venn PNG",
             "Alternative broad-context Venn SVG",
             "Alternative broad-context membership table",
             "Alternative exact-contrast UpSet PNG",
             "Alternative exact-contrast UpSet SVG",
             "Alternative exact membership table",
             "Standalone physiology composition PNG",
             "Standalone candidate category gene table",
             "Figure 2 rRNA exclusion audit"),
  path = c(
    file.path(out_dir, "figure2_deg_burden_context_phase_venns.png"),
    file.path(out_dir, "figure2_deg_burden_context_phase_venns.svg"),
    file.path(out_dir, "figure2_panel_A_deg_biotype_burden.png"),
    file.path(out_dir, "figure2_infection_degs_by_diet_venn.png"),
    file.path(out_dir, "figure2_infection_degs_by_diet_venn.svg"),
    file.path(out_dir, "figure2_infection_diet_venn_membership.csv"),
    file.path(out_dir, "figure2_infection_phase_overlap_venn.png"),
    file.path(out_dir, "figure2_infection_phase_overlap_venn.svg"),
    file.path(out_dir, "figure2_broad_context_overlap_venn.png"),
    file.path(out_dir, "figure2_broad_context_overlap_venn.svg"),
    file.path(out_dir, "figure2_deg_family_overlap_membership.csv"),
    file.path(out_dir, "figure2_panel_B_exact_contrast_upset.png"),
    file.path(out_dir, "figure2_panel_B_exact_contrast_upset.svg"),
    file.path(out_dir,
              "figure2_panel_B_exact_contrast_upset_membership.csv"),
    file.path(out_dir,
              "figure2_panel_C_candidate_physiology_phase_composition.png"),
    file.path(out_dir, "figure2_deg_family_candidate_theme_gene_table.csv"),
    file.path(out_dir, "figure2_rrna_exclusion_audit.csv")
  ),
  present = file.exists(c(
    file.path(out_dir, "figure2_deg_burden_context_phase_venns.png"),
    file.path(out_dir, "figure2_deg_burden_context_phase_venns.svg"),
    file.path(out_dir, "figure2_panel_A_deg_biotype_burden.png"),
    file.path(out_dir, "figure2_infection_degs_by_diet_venn.png"),
    file.path(out_dir, "figure2_infection_degs_by_diet_venn.svg"),
    file.path(out_dir, "figure2_infection_diet_venn_membership.csv"),
    file.path(out_dir, "figure2_infection_phase_overlap_venn.png"),
    file.path(out_dir, "figure2_infection_phase_overlap_venn.svg"),
    file.path(out_dir, "figure2_broad_context_overlap_venn.png"),
    file.path(out_dir, "figure2_broad_context_overlap_venn.svg"),
    file.path(out_dir, "figure2_deg_family_overlap_membership.csv"),
    file.path(out_dir, "figure2_panel_B_exact_contrast_upset.png"),
    file.path(out_dir, "figure2_panel_B_exact_contrast_upset.svg"),
    file.path(out_dir,
              "figure2_panel_B_exact_contrast_upset_membership.csv"),
    file.path(out_dir,
              "figure2_panel_C_candidate_physiology_phase_composition.png"),
    file.path(out_dir, "figure2_deg_family_candidate_theme_gene_table.csv"),
    file.path(out_dir, "figure2_rrna_exclusion_audit.csv")
  ))
)
knitr::kable(figure2_exports, caption = "Figure 2 and standalone overlap exports")
```

## Figure 3: expression clusters and enrichment

Figure 3 uses the two primary chromosome-only expression clusters. Panel A
aligns the 0-0.1% fungal-alignment zoom above the treatment/diet annotations and
the complete gene-level main-chromosome DEG heatmap. Values exceeding 0.1% are
capped only for display and retain their full percentage as enlarged white
labels. Panels B and C use only genes differentially expressed for infected
versus control within at least one diet, with cluster 1 and cluster 2 tested
separately. The overall all-DEG enrichment remains on the analysis page and is
not repeated here.
The primary KEGG panel is insect/arthropod-focused and retains only pathways
whose KEGG definition explicitly concerns insects or a fly model. A separate
eukaryote-inclusive lollipop version preserves conserved cross-domain and broad
reference pathways after excluding explicitly microbial/prokaryotic and
disease-reference labels. The broader pathways are also available around a
locust illustration; connector positions are conceptual and do not represent
measured sub-tissue localization. The complete pathway-scope and pathway-gene
audits remain available.

```{r figure-3-composite, echo=FALSE, fig.width=16, fig.height=12}
figure3_asset_specs <- tibble::tribble(
  ~figure, ~source_base, ~target_base,
  "Composite: eukaryote-inclusive KEGG pathway illustration",
  tools::file_path_sans_ext(figure3_composite_source),
  "figure3_expression_cluster_gene_level_GO_KEGG_pathway_illustration",
  "Composite: eukaryote-inclusive KEGG lollipops",
  tools::file_path_sans_ext(figure3_lollipop_source),
  "figure3_expression_cluster_gene_level_GO_KEGG_lollipop_eukaryote_inclusive",
  "Composite without panel titles: KEGG pathway illustration",
  file.path(
    circled_cluster_dir,
    "expression_cluster_gene_level_GO_KEGG_pathway_illustration_title_free"
  ),
  "figure3_expression_cluster_gene_level_GO_KEGG_pathway_illustration_title_free",
  "Composite without panel titles: eukaryote-inclusive KEGG lollipops",
  file.path(
    circled_cluster_dir,
    "expression_cluster_gene_level_GO_KEGG_lollipop_eukaryote_inclusive_title_free"
  ),
  "figure3_expression_cluster_gene_level_GO_KEGG_lollipop_eukaryote_inclusive_title_free",
  "Composite: insect-focused KEGG lollipops",
  tools::file_path_sans_ext(figure3_insect_source),
  "figure3_expression_cluster_gene_level_GO_KEGG_lollipop_insect_focused",
  "Panel C: insect/arthropod-focused KEGG lollipops",
  file.path(circled_cluster_dir,
            "figure3_panel_C_two_cluster_KEGG_insect_focused"),
  "figure3_panel_C_two_cluster_KEGG_insect_arthropod_focused",
  "Composite: reference-context companion",
  tools::file_path_sans_ext(figure3_reference_source),
  "figure3_expression_cluster_gene_level_GO_KEGG_lollipop_with_references",
  "Panel A: fungal percentage and two-cluster heatmap",
  file.path(circled_cluster_dir, "figure3_panel_A_two_cluster_heatmap"),
  "figure3_panel_A_fungal_percentage_two_cluster_heatmap",
  "Panel A without title: fungal percentage and two-cluster heatmap",
  file.path(
    circled_cluster_dir,
    "figure3_panel_A_two_cluster_heatmap_title_free"
  ),
  "figure3_panel_A_fungal_percentage_two_cluster_heatmap_title_free",
  "Panel B: two-cluster GO enrichment",
  file.path(circled_cluster_dir, "figure3_panel_B_two_cluster_GO"),
  "figure3_panel_B_two_cluster_GO",
  "Panel B without title: two-cluster GO enrichment",
  file.path(circled_cluster_dir, "figure3_panel_B_two_cluster_GO_title_free"),
  "figure3_panel_B_two_cluster_GO_title_free",
  "Panel C: eukaryote-inclusive KEGG pathway illustration",
  file.path(circled_cluster_dir,
            "figure3_panel_C_two_cluster_KEGG_pathway_illustration"),
  "figure3_panel_C_two_cluster_KEGG_pathway_illustration",
  "Panel C without title: KEGG pathway illustration",
  file.path(
    circled_cluster_dir,
    "figure3_panel_C_two_cluster_KEGG_pathway_illustration_title_free"
  ),
  "figure3_panel_C_two_cluster_KEGG_pathway_illustration_title_free",
  "Panel C alternative: eukaryote-inclusive KEGG lollipops",
  file.path(circled_cluster_dir,
            "figure3_panel_C_two_cluster_KEGG_eukaryote_inclusive"),
  "figure3_panel_C_two_cluster_KEGG_eukaryote_inclusive",
  "Panel C without title: eukaryote-inclusive KEGG lollipops",
  file.path(
    circled_cluster_dir,
    "figure3_panel_C_two_cluster_KEGG_eukaryote_inclusive_title_free"
  ),
  "figure3_panel_C_two_cluster_KEGG_eukaryote_inclusive_title_free"
)

figure3_exports <- tidyr::crossing(
  asset_index = seq_len(nrow(figure3_asset_specs)),
  format = c("png", "svg")
) |>
  dplyr::mutate(
    figure = figure3_asset_specs$figure[asset_index],
    source = paste0(figure3_asset_specs$source_base[asset_index], ".", format),
    target_base = figure3_asset_specs$target_base[asset_index]
  )

missing_figure3_assets <- figure3_exports$source[
  !file.exists(figure3_exports$source)
]
if (length(missing_figure3_assets) > 0) {
  stop("Missing Figure 3 publication asset(s): ",
       paste(missing_figure3_assets, collapse = ", "))
}

figure3_exports$copied_path <- mapply(
  copy_publication_asset,
  source = figure3_exports$source,
  file_base = figure3_exports$target_base,
  USE.NAMES = FALSE
)
figure3_exports <- figure3_exports |>
  dplyr::mutate(
    source = basename(source),
    copied_to = basename(copied_path),
    present = file.exists(copied_path)
  ) |>
  dplyr::select(figure, format, source, copied_to, present)

figure3_png <- file.path(
  out_dir,
  "figure3_expression_cluster_gene_level_GO_KEGG_pathway_illustration.png"
)
figure3_lollipop_png <- file.path(
  out_dir,
  "figure3_expression_cluster_gene_level_GO_KEGG_lollipop_eukaryote_inclusive.png"
)
figure3_insect_png <- file.path(
  out_dir,
  "figure3_expression_cluster_gene_level_GO_KEGG_lollipop_insect_focused.png"
)
knitr::kable(figure3_exports, caption = "Figure 3 copied exports")

```

### Figure 3 versions {.tabset}

#### Insect/arthropod-focused

```{r figure-3-insect-display, echo=FALSE, fig.width=16, fig.height=17}
if (file.exists(figure3_insect_png)) {
  grid::grid.newpage()
  grid::grid.raster(png::readPNG(figure3_insect_png))
}
```

#### Eukaryote-inclusive

```{r figure-3-kegg-lollipop-display, echo=FALSE, fig.width=16, fig.height=18}
if (file.exists(figure3_lollipop_png)) {
  grid::grid.newpage()
  grid::grid.raster(png::readPNG(figure3_lollipop_png))
}
```

#### Eukaryote-inclusive locust map

```{r figure-3-locust-map-display, echo=FALSE, fig.width=16, fig.height=19}
if (file.exists(figure3_png)) {
  grid::grid.newpage()
  grid::grid.raster(png::readPNG(figure3_png))
}
```

## Figure 4: curated immune and protein-anabolism targets

Figure 4 preserves the curated biological categories and recalculates every
DEG status from the corrected 44-sample competitive-host exon-only analysis.
The supplied v21 workbook contains 111 unique targets (77 immune and 34
protein-anabolism genes), and all 111 map to main nuclear chromosomes. Each
target therefore stays in the denominator for every contrast and is shown as
`N.S.` when it does not meet the current adjusted-P-value and fold-change
thresholds. The matched all-scaffold view is retained as a reproducibility
check and has the same target denominator. Placement details remain searchable on the
[exon-only curated target-gene analysis page](45-exon-only-curated-target-genes.html).
A separate audit view isolates the curated targets carrying the scaffold-origin
page's non-metazoan homology review flags; those flags do not prove that the
genes are contaminants.

```{r figure-4-curated-targets, echo=FALSE, fig.width=15, fig.height=11.5}
figure4_png <- copy_publication_asset(
  figure4_curated_source,
  "figure4_curated_targets_main_chromosomes"
)
figure4_placement_png <- copy_publication_asset(
  figure4_placement_source,
  "figure4_curated_target_placement_impact"
)
figure4_all_scaffolds_png <- copy_publication_asset(
  figure4_all_scaffolds_source,
  "figure4_curated_targets_all_scaffolds"
)
figure4_homology_flags_png <- copy_publication_asset(
  figure4_homology_flags_source,
  "figure4_curated_targets_non_grasshopper_homology_flags"
)

figure4_exports <- data.frame(
  figure = c(
    "Figure 4 main-chromosome curated targets",
    "Figure 4 all-scaffold curated targets",
    "Figure 4 homology-review targets",
    "Figure 4 placement-impact audit"
  ),
  source = basename(c(
    figure4_curated_source, figure4_all_scaffolds_source,
    figure4_homology_flags_source, figure4_placement_source
  )),
  copied_to = basename(c(
    figure4_png, figure4_all_scaffolds_png, figure4_homology_flags_png,
    figure4_placement_png
  )),
  present = file.exists(c(
    figure4_png, figure4_all_scaffolds_png, figure4_homology_flags_png,
    figure4_placement_png
  ))
)
knitr::kable(figure4_exports, caption = "Figure 4 copied exports")
```

### Figure 4 views {.tabset}

#### Main chromosomes

```{r figure-4-main-display, echo=FALSE, fig.width=15, fig.height=11.5}
if (file.exists(figure4_png)) {
  grid::grid.newpage()
  grid::grid.raster(png::readPNG(figure4_png))
}
```

#### All scaffolds

```{r figure-4-all-scaffolds-display, echo=FALSE, fig.width=15, fig.height=11.5}
if (file.exists(figure4_all_scaffolds_png)) {
  grid::grid.newpage()
  grid::grid.raster(png::readPNG(figure4_all_scaffolds_png))
}
```

#### Homology review flags

```{r figure-4-homology-flags-display, echo=FALSE, fig.width=15, fig.height=11.5}
if (file.exists(figure4_homology_flags_png)) {
  grid::grid.newpage()
  grid::grid.raster(png::readPNG(figure4_homology_flags_png))
}
```

#### Placement impact

```{r figure-4-placement-display, echo=FALSE, fig.width=12, fig.height=7.5}
if (file.exists(figure4_placement_png)) {
  grid::grid.newpage()
  grid::grid.raster(png::readPNG(figure4_placement_png))
}
```

<!-- The historical PCA-based removal of 1007, 1036, 1037, and 1039 is retained
below as non-evaluated provenance. These QC-passed samples remain in every
current biological analysis; the visible historical reports are in the archive. -->

```{r archived-pca-sample-removal, eval=FALSE, include=FALSE}
s1_scenario_results <- read.csv(scenario_results_file, check.names = FALSE)
s1_sample_counts <- read.csv(scenario_sample_counts_file, check.names = FALSE)

s1_scenario_levels <- c("All samples", "Current outliers removed",
                        "Expanded outlier set")
s1_scenario_levels <- s1_scenario_levels[
  s1_scenario_levels %in% unique(s1_scenario_results$scenario)
]
s1_direction_levels <- c("Downregulated", "Upregulated")
s1_direction_colors <- c("Downregulated" = "#1D4ED8",
                         "Upregulated" = "#B91C1C")
s1_sensitivity_colors <- c(
  "Retained after outlier removal" = "#B7791F",
  "Lost after outlier removal" = "#B91C1C",
  "Gained after outlier removal" = "#2563EB"
)

format_comparison_label <- function(x) {
  dplyr::case_when(
    x == "Diet 33: infected vs control" ~ "Diet 33\ninfected vs control",
    x == "Diet 50: infected vs control" ~ "Diet 50\ninfected vs control",
    x == "Diet 83: infected vs control" ~ "Diet 83\ninfected vs control",
    x == "Control 33 vs 50" ~ "Control\n33 vs 50",
    x == "Control 83 vs 33" ~ "Control\n83 vs 33",
    x == "Control 83 vs 50" ~ "Control\n83 vs 50",
    x == "Infected 33 vs 50" ~ "Infected\n33 vs 50",
    x == "Infected 83 vs 33" ~ "Infected\n83 vs 33",
    x == "Infected 83 vs 50" ~ "Infected\n83 vs 50",
    TRUE ~ x
  )
}

s1_comparison_levels <- c(
  "Diet 33: infected vs control",
  "Diet 50: infected vs control",
  "Diet 83: infected vs control",
  "Control 33 vs 50",
  "Control 83 vs 33",
  "Control 83 vs 50",
  "Infected 33 vs 50",
  "Infected 83 vs 33",
  "Infected 83 vs 50"
)
s1_comparison_levels <- s1_comparison_levels[
  s1_comparison_levels %in% unique(s1_scenario_results$comparison)
]

s1_sample_counts <- s1_sample_counts |>
  dplyr::mutate(
    scenario = factor(scenario, levels = s1_scenario_levels),
    Diet = factor(Diet, levels = c("33", "50", "83")),
    Treatment = factor(Treatment, levels = c("Control", "Infected"))
  )

p_s1_samples <- ggplot2::ggplot(
  s1_sample_counts,
  ggplot2::aes(x = Diet, y = n_samples, fill = Treatment)
) +
  ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.75),
                    width = 0.66, color = "#111827", linewidth = 0.25) +
  ggplot2::geom_text(
    ggplot2::aes(label = n_samples),
    position = ggplot2::position_dodge(width = 0.75),
    vjust = -0.25,
    size = 3.6
  ) +
  ggplot2::facet_wrap(~ scenario, nrow = 1) +
  ggplot2::scale_fill_manual(values = treatment_colors, drop = FALSE) +
  ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0, 0.15))) +
  theme_pub(base_size = 12) +
  ggplot2::theme(
    strip.text = ggplot2::element_text(size = 11, face = "bold"),
    axis.text = ggplot2::element_text(color = "#111827"),
    legend.position = "bottom"
  ) +
  ggplot2::labs(
    title = "A. Samples retained per scenario",
    x = "Diet",
    y = "Number of samples",
    fill = "Treatment"
  )

s1_sig_results <- s1_scenario_results |>
  dplyr::filter(
    scenario %in% s1_scenario_levels,
    comparison %in% s1_comparison_levels,
    significant,
    !is.na(gene_id),
    !is.na(padj),
    padj < publication_params$alpha,
    abs(log2FoldChange) >= publication_params$lfc_threshold,
    direction %in% s1_direction_levels
  ) |>
  dplyr::mutate(
    scenario = factor(scenario, levels = s1_scenario_levels),
    comparison = factor(comparison, levels = s1_comparison_levels),
    comparison_label = factor(format_comparison_label(as.character(comparison)),
                              levels = rev(format_comparison_label(s1_comparison_levels))),
    direction = factor(direction, levels = s1_direction_levels)
  )

s1_deg_counts <- s1_sig_results |>
  dplyr::count(scenario, comparison_type, comparison_label, direction,
               name = "n_genes") |>
  tidyr::complete(
    scenario,
    tidyr::nesting(comparison_type, comparison_label),
    direction,
    fill = list(n_genes = 0)
  ) |>
  dplyr::mutate(
    signed_n_genes = dplyr::if_else(direction == "Downregulated",
                                    -n_genes, n_genes)
  )

write.csv(s1_sample_counts,
          file.path(out_dir, "figureS1_sample_counts_by_scenario.csv"),
          row.names = FALSE)
write.csv(s1_deg_counts,
          file.path(out_dir, "figureS1_deg_counts_by_scenario.csv"),
          row.names = FALSE)

p_s1_burden <- ggplot2::ggplot(
  s1_deg_counts,
  ggplot2::aes(x = signed_n_genes, y = comparison_label, fill = direction)
) +
  ggplot2::geom_vline(xintercept = 0, color = "#111827", linewidth = 0.3) +
  ggplot2::geom_col(width = 0.72, color = "white", linewidth = 0.2) +
  ggplot2::facet_wrap(~ scenario, nrow = 1) +
  ggplot2::scale_fill_manual(values = s1_direction_colors, drop = FALSE) +
  ggplot2::scale_x_continuous(labels = abs) +
  theme_pub(base_size = 12) +
  ggplot2::theme(
    axis.text.y = ggplot2::element_text(size = 10, color = "#111827"),
    strip.text = ggplot2::element_text(size = 10.5, face = "bold"),
    legend.position = "bottom"
  ) +
  ggplot2::labs(
    title = "B. Direction-aware DEG burden across filtering scenarios",
    x = "Number of significant DEGs",
    y = NULL,
    fill = "Direction"
  )

s1_compare_scenarios <- c("All samples", "Current outliers removed")
if (all(s1_compare_scenarios %in% unique(s1_sig_results$scenario))) {
  s1_retained_lost_gained <- do.call(rbind, lapply(s1_comparison_levels,
                                                   function(this_comparison) {
    all_genes <- unique(s1_sig_results$gene_id[
      s1_sig_results$scenario == "All samples" &
        s1_sig_results$comparison == this_comparison
    ])
    no_outlier_genes <- unique(s1_sig_results$gene_id[
      s1_sig_results$scenario == "Current outliers removed" &
        s1_sig_results$comparison == this_comparison
    ])
    data.frame(
      comparison = this_comparison,
      sensitivity_class = c("Retained after outlier removal",
                            "Lost after outlier removal",
                            "Gained after outlier removal"),
      n_genes = c(
        length(intersect(all_genes, no_outlier_genes)),
        length(setdiff(all_genes, no_outlier_genes)),
        length(setdiff(no_outlier_genes, all_genes))
      ),
      stringsAsFactors = FALSE
    )
  }))
} else {
  s1_retained_lost_gained <- data.frame(
    comparison = character(),
    sensitivity_class = character(),
    n_genes = integer()
  )
}

s1_retained_lost_gained <- s1_retained_lost_gained |>
  dplyr::mutate(
    comparison = factor(comparison, levels = s1_comparison_levels),
    comparison_label = factor(format_comparison_label(as.character(comparison)),
                              levels = rev(format_comparison_label(s1_comparison_levels))),
    sensitivity_class = factor(sensitivity_class,
                               levels = names(s1_sensitivity_colors))
  )

write.csv(s1_retained_lost_gained,
          file.path(out_dir, "figureS1_retained_lost_gained_degs.csv"),
          row.names = FALSE)

p_s1_retained <- ggplot2::ggplot(
  s1_retained_lost_gained,
  ggplot2::aes(x = n_genes, y = comparison_label, fill = sensitivity_class)
) +
  ggplot2::geom_col(width = 0.72, color = "white", linewidth = 0.2) +
  ggplot2::scale_fill_manual(values = s1_sensitivity_colors, drop = FALSE) +
  theme_pub(base_size = 12) +
  ggplot2::theme(
    axis.text.y = ggplot2::element_text(size = 10, color = "#111827"),
    legend.position = "bottom"
  ) +
  ggplot2::labs(
    title = "C. DEG calls retained, lost, or gained after current outlier removal",
    x = "Number of significant DEGs",
    y = NULL,
    fill = "Sensitivity class"
  )

figure_s1 <- p_s1_samples / p_s1_burden / p_s1_retained +
  patchwork::plot_layout(heights = c(0.22, 0.48, 0.30)) +
  patchwork::plot_annotation(
    title = "Figure S1. Outlier-removal sensitivity analysis",
    subtitle = "All comparisons use the rRNA-filtered count matrix and the same DEG thresholds as the main figures",
    theme = ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold", size = 18),
      plot.subtitle = ggplot2::element_text(size = 12)
    )
  )

save_png_svg(figure_s1, "figureS1_outlier_removal_sensitivity",
             width = 17.5, height = 15)
figure_s1

figure_s1_exports <- data.frame(
  figure = c("Figure S1 composite PNG", "Figure S1 composite SVG",
             "Figure S1 sample counts", "Figure S1 DEG counts",
             "Figure S1 retained/lost/gained counts"),
  path = c(
    file.path(out_dir, "figureS1_outlier_removal_sensitivity.png"),
    file.path(out_dir, "figureS1_outlier_removal_sensitivity.svg"),
    file.path(out_dir, "figureS1_sample_counts_by_scenario.csv"),
    file.path(out_dir, "figureS1_deg_counts_by_scenario.csv"),
    file.path(out_dir, "figureS1_retained_lost_gained_degs.csv")
  ),
  present = file.exists(c(
    file.path(out_dir, "figureS1_outlier_removal_sensitivity.png"),
    file.path(out_dir, "figureS1_outlier_removal_sensitivity.svg"),
    file.path(out_dir, "figureS1_sample_counts_by_scenario.csv"),
    file.path(out_dir, "figureS1_deg_counts_by_scenario.csv"),
    file.path(out_dir, "figureS1_retained_lost_gained_degs.csv")
  ))
)
knitr::kable(figure_s1_exports,
             caption = "Figure S1 outlier-sensitivity exports")
```

## Figure S1: mapping-quality basis for excluding 1044

Figure S1 preserves all 45 sequenced libraries for the technical QC comparison,
while all biological models use the 44-library cohort after excluding 1044.
Panel A shows unique mapping and panel B shows unique plus accepted
multimapping. The factorial diet, infection, and interaction tests are printed
on the panels. Samples `1007`, `1036`, `1037`, and `1039` remain included.

```{r figure-s1-mapping-quality, echo=FALSE, out.width="82%"}
qc_dir <- latest_dir("^count-qc-exploration_")
s1_mapping_stem <- "figureS3_mapping_quality_exclusion_1044"
s1_mapping_sources <- file.path(
  qc_dir,
  paste0(s1_mapping_stem, c(".png", ".svg", ".pdf"))
)
if (!all(file.exists(s1_mapping_sources))) {
  stop("Missing Figure S1 mapping-QC source exports in: ", qc_dir)
}
s1_mapping_targets <- file.path(
  out_dir,
  sub("^figureS3_", "figureS1_", basename(s1_mapping_sources))
)
file.copy(s1_mapping_sources, s1_mapping_targets, overwrite = TRUE)
s1_mapping_png <- s1_mapping_sources[grepl("[.]png$", s1_mapping_sources)]
grid::grid.newpage()
grid::grid.raster(png::readPNG(s1_mapping_png))
```

```{r figure-s1-mapping-export-manifest, echo=FALSE}
knitr::kable(
  tibble::tibble(
    file = basename(s1_mapping_targets),
    present = file.exists(s1_mapping_targets)
  ),
  caption = "Figure S1 mapping-quality exports"
)
```

## Figure S2: non-host taxonomic profiles

Figure S2 summarizes Kraken2/Bracken family assignments among reads that did
not align to the host-only reference. Metazoan and plant families are excluded;
the retained profiles comprise fungi, bacteria, archaea, viruses, and other
microbial eukaryotes. Panels A and B show absolute Bracken-estimated reads and
their relative composition among retained non-host taxonomic families; panel C
shows Bray-Curtis ordination. This is supporting metatranscriptomic evidence
rather than a dedicated microbiome assay.

```{r figure-s2-bracken, echo=FALSE, out.width="96%"}
taxonomy_dir <- latest_dir("^metarhizium-read-burden_")
s2_taxonomy_files <- c(
  "figureS2_bracken_host_unmapped_family_composition.png",
  "figureS2_bracken_host_unmapped_family_composition.svg",
  "figureS2_bracken_host_unmapped_family_composition.pdf",
  "figureS2_bracken_host_unmapped_family_composition_large_text_narrow.png",
  "figureS2_bracken_host_unmapped_family_composition_large_text_narrow.svg",
  "figureS2_bracken_host_unmapped_family_composition_large_text_narrow.pdf",
  "bracken_host_unmapped_family_abundance_long.csv",
  "bracken_host_unmapped_family_display_table.csv",
  "bracken_host_unmapped_family_pcoa_scores.csv",
  "bracken_host_unmapped_family_pcoa_offset_labels.csv",
  "bracken_host_unmapped_family_permanova.csv",
  "bracken_host_unmapped_family_dispersion_test.csv",
  "bracken_host_unmapped_burden_by_sample.csv",
  "bracken_host_unmapped_burden_design_tests.csv",
  "bracken_host_unmapped_burden_mass_tests.csv",
  "bracken_host_unmapped_burden_group_summary.csv"
)
s2_taxonomy_sources <- file.path(taxonomy_dir, s2_taxonomy_files)
if (!all(file.exists(s2_taxonomy_sources))) {
  stop("Missing Figure S2 Kraken/Bracken source exports in: ", taxonomy_dir)
}
s2_taxonomy_targets <- file.path(out_dir, basename(s2_taxonomy_sources))
file.copy(s2_taxonomy_sources, s2_taxonomy_targets, overwrite = TRUE)
grid::grid.newpage()
grid::grid.raster(png::readPNG(s2_taxonomy_sources[[4]]))
```

```{r figure-s2-bracken-export-manifest, echo=FALSE}
knitr::kable(
  tibble::tibble(
    file = basename(s2_taxonomy_targets),
    present = file.exists(s2_taxonomy_targets)
  ),
  caption = "Figure S2 Kraken2/Bracken taxonomic-profile exports"
)
```

<!-- Chromosomal DEG localization remains an unnumbered candidate supplementary
figure until the composition of Figure S3 and later supplements is decided. -->

```{r candidate-deg-genome-localization, eval=FALSE, include=FALSE}
s2_scenario_degs <- read.csv(scenario_results_file, check.names = FALSE) |>
  dplyr::filter(
    scenario == "All samples",
    significant,
    !is.na(gene_id),
    !is.na(padj),
    padj < publication_params$alpha,
    abs(log2FoldChange) >= publication_params$lfc_threshold
  )

s2_model_degs <- read.csv(model_coefficients_file, check.names = FALSE) |>
  dplyr::filter(
    coefficient == "Treatment_Infected_vs_Control",
    significant,
    !is.na(gene_id),
    !is.na(padj),
    padj < publication_params$alpha,
    abs(log2FoldChange) >= publication_params$lfc_threshold
  )

s2_deg_sets <- list(
  "Infection within diets" = s2_scenario_degs |>
    dplyr::filter(comparison_type == "Infection within diet") |>
    dplyr::pull(gene_id) |>
    unique(),
  "Diet DEGs in controls" = s2_scenario_degs |>
    dplyr::filter(comparison_type == "Diet contrasts among control") |>
    dplyr::pull(gene_id) |>
    unique(),
  "Diet DEGs in infected" = s2_scenario_degs |>
    dplyr::filter(comparison_type == "Diet contrasts among infected") |>
    dplyr::pull(gene_id) |>
    unique(),
  "Infection vs control" = unique(s2_model_degs$gene_id)
)
s2_deg_sets <- s2_deg_sets[lengths(s2_deg_sets) > 0]
s2_context_levels <- names(s2_deg_sets)
s2_context_colors <- c(
  "Infection within diets" = "#C51B7D",
  "Diet DEGs in controls" = "#D9A21B",
  "Diet DEGs in infected" = "#1B9E77",
  "Infection vs control" = "#377EB8"
)

s2_deg_long <- dplyr::bind_rows(lapply(s2_context_levels, function(context) {
  data.frame(
    DEG_context = context,
    gene_id = s2_deg_sets[[context]],
    stringsAsFactors = FALSE
  )
})) |>
  dplyr::distinct(DEG_context, gene_id) |>
  dplyr::mutate(DEG_context = factor(DEG_context,
                                     levels = s2_context_levels))

gff_coordinates <- parse_gff_gene_coordinates(gff_file)
chromosome_order <- c(as.character(1:11), "X")

chromosome_regions <- gff_coordinates$sequence_regions |>
  dplyr::filter(chromosome %in% chromosome_order) |>
  dplyr::mutate(
    chromosome = factor(chromosome, levels = rev(chromosome_order)),
    length_mb = seq_end / 1e6
  )

annotated_gene_counts <- gff_coordinates$gene_ranges |>
  dplyr::filter(chromosome %in% chromosome_order) |>
  dplyr::count(chromosome, name = "annotated_genes")

s2_deg_coordinates <- s2_deg_long |>
  dplyr::left_join(gff_coordinates$gene_ranges, by = "gene_id") |>
  dplyr::mutate(
    chromosome = as.character(chromosome),
    chromosome_class = dplyr::coalesce(chromosome_class, "Missing from GFF"),
    position_mb = midpoint / 1e6
  )

write.csv(s2_deg_coordinates,
          file.path(out_dir, "figureS4_deg_coordinates_by_context.csv"),
          row.names = FALSE)

s2_unplaced_summary <- s2_deg_coordinates |>
  dplyr::filter(!(chromosome %in% chromosome_order)) |>
  dplyr::count(DEG_context, chromosome_class, name = "n_degs") |>
  dplyr::arrange(DEG_context, chromosome_class)

write.csv(s2_unplaced_summary,
          file.path(out_dir, "figureS4_unplaced_or_missing_deg_summary.csv"),
          row.names = FALSE)

s2_coordinate_class_counts <- s2_deg_coordinates |>
  dplyr::mutate(
    coordinate_class = dplyr::if_else(
      chromosome_class == "Named chromosome",
      "Named chromosomes 1-11/X",
      chromosome_class
    )
  ) |>
  dplyr::count(DEG_context, coordinate_class, name = "n_degs") |>
  dplyr::group_by(DEG_context) |>
  dplyr::mutate(proportion = n_degs / sum(n_degs)) |>
  dplyr::ungroup()

write.csv(s2_coordinate_class_counts,
          file.path(out_dir, "figureS4_coordinate_class_counts.csv"),
          row.names = FALSE)

s2_chromosome_degs <- s2_deg_coordinates |>
  dplyr::filter(chromosome %in% chromosome_order, !is.na(position_mb)) |>
  dplyr::mutate(
    chromosome = factor(chromosome, levels = rev(chromosome_order)),
    DEG_context = factor(DEG_context, levels = s2_context_levels)
  )

s2_chromosome_counts <- s2_chromosome_degs |>
  dplyr::count(DEG_context, chromosome, name = "deg_genes") |>
  tidyr::complete(
    DEG_context,
    chromosome = factor(chromosome_order, levels = rev(chromosome_order)),
    fill = list(deg_genes = 0)
  ) |>
  dplyr::mutate(chromosome = as.character(chromosome)) |>
  dplyr::left_join(annotated_gene_counts, by = "chromosome") |>
  dplyr::mutate(
    degs_per_1000_genes = 1000 * deg_genes / annotated_genes,
    chromosome = factor(chromosome, levels = rev(chromosome_order)),
    DEG_context = factor(DEG_context, levels = s2_context_levels)
  )

write.csv(s2_chromosome_counts,
          file.path(out_dir, "figureS4_chromosome_deg_density.csv"),
          row.names = FALSE)

p_s2_locations <- ggplot2::ggplot() +
  ggplot2::geom_segment(
    data = chromosome_regions,
    ggplot2::aes(x = 0, xend = length_mb,
                 y = chromosome, yend = chromosome),
    linewidth = 4.8,
    color = "#E5E7EB",
    lineend = "round"
  ) +
  ggplot2::geom_point(
    data = s2_chromosome_degs,
    ggplot2::aes(x = position_mb, y = chromosome, color = DEG_context),
    size = 0.55,
    alpha = 0.45,
    position = ggplot2::position_jitter(width = 0, height = 0.15, seed = 17)
  ) +
  ggplot2::facet_wrap(~ DEG_context, ncol = 2) +
  ggplot2::scale_color_manual(values = s2_context_colors, guide = "none") +
  ggplot2::scale_x_continuous(
    expand = ggplot2::expansion(mult = c(0.01, 0.04))
  ) +
  theme_pub(base_size = 12) +
  ggplot2::theme(
    panel.grid.major.y = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank(),
    strip.text = ggplot2::element_text(size = 12, face = "bold"),
    axis.text.y = ggplot2::element_text(size = 10, color = "#111827"),
    legend.position = "none"
  ) +
  ggplot2::labs(
    title = "A. Chromosomal positions of DEGs by context",
    subtitle = "Each point is one DEG-context membership; genes shared by contexts appear in each relevant panel",
    x = "Genomic position (Mb)",
    y = "Chromosome"
  )

p_s2_coordinate_class <- ggplot2::ggplot(
  s2_chromosome_counts,
  ggplot2::aes(x = chromosome, y = DEG_context,
               fill = degs_per_1000_genes)
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.45) +
  ggplot2::scale_fill_gradient(low = "#FFF7D1", high = "#8B0000") +
  theme_pub(base_size = 12) +
  ggplot2::theme(
    axis.text.y = ggplot2::element_text(size = 10, color = "#111827"),
    axis.text.x = ggplot2::element_text(face = "bold"),
    legend.position = "bottom"
  ) +
  ggplot2::labs(
    title = "C. DEG density by chromosome and context",
    x = "Chromosome",
    y = NULL,
    fill = "DEGs per 1,000\nannotated genes"
  )

figure_s2 <- p_s2_locations / p_s2_coordinate_class +
  patchwork::plot_layout(heights = c(0.78, 0.22)) +
  patchwork::plot_annotation(
    title = "Figure S4. Chromosomal localization of DEG contexts",
    subtitle = "Primary paper DEG sets contain only official nuclear-chromosome genes",
    theme = ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold", size = 18),
      plot.subtitle = ggplot2::element_text(size = 12)
    )
  )

save_png_svg(figure_s2, "figureS4_deg_chromosome_localization",
             width = 16, height = 10.5)
figure_s2

figure_s2_exports <- data.frame(
  figure = c("Figure S4 composite PNG", "Figure S4 composite SVG",
             "Figure S4 DEG coordinates", "Chromosome density table",
             "Figure S4 coordinate class counts",
             "Figure S4 unplaced/missing summary"),
  path = c(
    file.path(out_dir, "figureS4_deg_chromosome_localization.png"),
    file.path(out_dir, "figureS4_deg_chromosome_localization.svg"),
    file.path(out_dir, "figureS4_deg_coordinates_by_context.csv"),
    file.path(out_dir, "figureS4_chromosome_deg_density.csv"),
    file.path(out_dir, "figureS4_coordinate_class_counts.csv"),
    file.path(out_dir, "figureS4_unplaced_or_missing_deg_summary.csv")
  ),
  present = file.exists(c(
    file.path(out_dir, "figureS4_deg_chromosome_localization.png"),
    file.path(out_dir, "figureS4_deg_chromosome_localization.svg"),
    file.path(out_dir, "figureS4_deg_coordinates_by_context.csv"),
    file.path(out_dir, "figureS4_chromosome_deg_density.csv"),
    file.path(out_dir, "figureS4_coordinate_class_counts.csv"),
    file.path(out_dir, "figureS4_unplaced_or_missing_deg_summary.csv")
  ))
)
knitr::kable(figure_s2_exports,
             caption = "Figure S4 chromosome-localization exports")
knitr::kable(s2_unplaced_summary,
             caption = "Audit check: excluded unplaced or missing DEG memberships (expected zero)")
```

<!-- Historical source-copy block retained for provenance; Figure S1 above is current. -->

```{r historical-figure-s3-mapping-quality, eval=FALSE, include=FALSE}
qc_dir <- latest_dir("^count-qc-exploration_")
s3_stem <- "figureS3_mapping_quality_exclusion_1044"
s3_sources <- file.path(qc_dir, paste0(s3_stem, c(".png", ".svg", ".pdf")))
if (!all(file.exists(s3_sources))) {
  stop("Missing Figure S3 mapping-QC exports in: ", qc_dir)
}
s3_targets <- file.path(out_dir, basename(s3_sources))
file.copy(s3_sources, s3_targets, overwrite = TRUE)
s3_png_source <- s3_sources[grepl("[.]png$", s3_sources)]
grid::grid.newpage()
grid::grid.raster(png::readPNG(s3_png_source))
```

```{r historical-figure-s3-export-manifest, eval=FALSE, include=FALSE}
knitr::kable(
  tibble::tibble(file = basename(s3_targets), present = file.exists(s3_targets)),
  caption = "Figure S3 mapping-quality exports"
)
```

## Exported files

```{r exported-files}
exports <- list.files(out_dir, full.names = FALSE)
knitr::kable(tibble::tibble(file = exports),
             caption = paste("Files written to", out_dir))
```

## Session information

```{r session-info}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
