---
title: "Statistical analysis reference"
author: "Emily Baker and Maeva A. Techer"
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
params:
  metadata_file: "data/metadata/mehreen_metadata_analysis_44_excluding_1044.txt"
  counts_file: "output/rmd_runs/transcript-exon-primary-excluding1044_20260809-233420/host_counts_no_rrna.tsv"
  placement_file: "output/runs/host_pathogen_dual_20260727_000952/local-analysis/host-mapping-audit_20260729-163928/unified-model/audit-summary/host_gene_sequence_placement.tsv"
  rrna_file: "data/excluded_loci/gregaria_rrna_list.txt"
  head_phase_file: "data/external/phase_reference/gregaria_head_phase_degs.csv"
  thorax_phase_file: "data/external/phase_reference/gregaria_thorax_phase_degs.csv"
  alpha: 0.05
  lfc_threshold: 1
  min_count: 10
  min_samples: 3
  output_tag: "paper-statistical-reference"
---

## Purpose

This page is the statistical map for the paper. It brings the model formulas,
decision thresholds, current results, and biological interpretation into one
place without replacing the full analysis pages. The aim is to make every
reported statement easy to trace from an input table to an R model and then to
a cautious conclusion.

::: {.walkthrough-grid}
::: {.walkthrough-step}
**Step 1 - Define the question**

Identify the response variable, experimental terms, and analysis population.
:::
::: {.walkthrough-step}
**Step 2 - Fit the model**

Use the model appropriate for proportions, mass, counts, or compositions.
:::
::: {.walkthrough-step}
**Step 3 - Report the evidence**

Pair an effect size or test statistic with its uncertainty and P value.
:::
::: {.walkthrough-step}
**Step 4 - Interpret carefully**

State what the result supports and what the design cannot establish.
:::
:::

## Setup and frozen inputs

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

required <- c("dplyr", "readr", "tibble", "tidyr", "knitr")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) {
  stop("Install required packages before knitting: ", paste(missing, collapse = ", "))
}

project_dir <- normalizePath(".", mustWork = TRUE)
run_stamp <- format(Sys.time(), "%Y%m%d-%H%M%S")
out_dir <- file.path(
  project_dir, "output", "rmd_runs",
  paste(params$output_tag, run_stamp, sep = "_")
)
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

latest_dir <- function(prefix) {
  root <- file.path(project_dir, "output", "rmd_runs")
  candidates <- list.dirs(root, recursive = FALSE, full.names = TRUE)
  candidates <- candidates[startsWith(basename(candidates), paste0(prefix, "_"))]
  if (length(candidates) == 0) stop("No output directory found for: ", prefix)
  candidates[order(file.info(candidates)$mtime, decreasing = TRUE)][1]
}

read_required_csv <- function(path) {
  if (!file.exists(path)) stop("Missing required result table: ", path)
  read.csv(path, check.names = FALSE)
}

format_p <- function(x) {
  ifelse(
    is.na(x), "NA",
    ifelse(x < 0.001, formatC(x, format = "e", digits = 2),
           formatC(x, format = "f", digits = 3))
  )
}

pretty_table <- function(x, caption = NULL, digits = 3) {
  table <- knitr::kable(
    x, format = "html", caption = caption, digits = digits,
    row.names = FALSE, escape = TRUE
  )
  if (requireNamespace("kableExtra", quietly = TRUE)) {
    table <- kableExtra::kable_styling(
      table,
      bootstrap_options = c("striped", "hover", "condensed", "responsive"),
      full_width = FALSE,
      position = "left"
    )
  }
  table
}

qc_dir <- latest_dir("count-qc-exploration")
interaction_dir <- latest_dir("main-chromosome-all-samples-deseq2-interaction-model")
diet_dir <- latest_dir("main-chromosome-all-samples-diet-pairwise-deg_All")
infection_dir <- latest_dir("main-chromosome-all-samples-infection-within-diet-deg")
publication_dir <- latest_dir("publication-figures")
enrichment_dir <- latest_dir("go-term-enrichment")
taxonomy_dir <- latest_dir("metarhizium-read-burden")

source_manifest <- data.frame(
  analysis = c(
    "Mapping and body mass", "Full DESeq2 interaction", "Diet contrasts",
    "Infection within diet", "Publication contrasts and overlap",
    "GO and KEGG enrichment", "Non-host taxonomic profiles"
  ),
  source_page = c(
    "analysis/03-count-qc-exploration.Rmd",
    "analysis/14-all-samples-deseq2-interaction-model.Rmd",
    "analysis/15-all-samples-diet-pairwise-deg.Rmd",
    "analysis/16-all-samples-infection-within-diet-deg.Rmd",
    "analysis/08-publication-figures.Rmd",
    "analysis/10-go-term-enrichment.Rmd",
    "analysis/29-metarhizium-read-burden.Rmd"
  ),
  result_directory = c(
    qc_dir, interaction_dir, diet_dir, infection_dir,
    publication_dir, enrichment_dir, taxonomy_dir
  )
)
write.csv(source_manifest, file.path(out_dir, "statistical_source_manifest.csv"),
          row.names = FALSE)
pretty_table(source_manifest, "Canonical analysis pages and result directories")
```

Every knit creates a new audit folder. The page reads the newest completed
result directory for each canonical analysis, but it records those exact paths
in `statistical_source_manifest.csv` so the reported values remain traceable.

## Analysis scope and decision rules

```{r analysis-scope}
input_audit <- read_required_csv(
  file.path(interaction_dir, "main_chromosome_input_audit.csv")
)

analysis_rules <- data.frame(
  item = c(
    "Analysis cohort", "Primary count definition", "Mapping branch",
    "Gene placement", "rRNA handling", "Expression filter",
    "DEG criterion", "Multiple testing", "Ordination"
  ),
  current_choice = c(
    paste0(input_audit$samples_retained, " QC-passed libraries; sample 1044 excluded"),
    "featureCounts -t transcript,exon -g gene_id",
    "Competitive host-pathogen alignment; primary alignments only",
    "Genes assigned to the 12 nuclear chromosomes",
    "All identifiers on the project rRNA exclusion list removed before modelling",
    paste0("Count >= ", params$min_count, " in at least ", params$min_samples, " libraries"),
    paste0("BH-adjusted P < ", params$alpha,
           " and absolute log2 fold-change >= ", params$lfc_threshold),
    "Benjamini-Hochberg correction within each contrast",
    "VST for PCA and heatmaps only; raw counts enter DESeq2"
  )
)
pretty_table(analysis_rules, "Current paper analysis conventions")
```

The count matrix contains `r format(input_audit$raw_count_rows, big.mark = ",")`
non-rRNA gene identifiers. Restricting it to official nuclear chromosomes leaves
`r format(input_audit$main_chromosome_non_rRNA_rows, big.mark = ",")` loci before
expression filtering and `r format(input_audit$genes_entering_DESeq2, big.mark = ",")`
genes entering the primary DESeq2 model.

### What is a test and what is a display?

```{r evidence-types}
evidence_types <- data.frame(
  output = c(
    "Mapping-quality models", "Body-mass model", "DESeq2 contrasts",
    "GO/KEGG over-representation", "PERMANOVA and PERMDISP",
    "Chromosome distribution", "PCA and heatmaps", "Venn/UpSet diagrams",
    "Curated candidate-gene panels"
  ),
  role = c(
    rep("Inferential", 6), rep("Descriptive", 3)
  ),
  interpretation = c(
    "Tests diet, infection, and diet-by-infection effects on mapping yield",
    "Tests diet, infection, and their interaction on recorded mass",
    "Tests one gene at a time while controlling the false-discovery rate",
    "Tests whether annotated terms are over-represented in a DEG foreground",
    "Tests taxonomic-composition location and dispersion",
    "Tests whether DEGs are distributed in proportion to the tested background",
    "Shows dominant expression structure; no P value is assigned",
    "Shows set membership and overlap; overlap alone is not causal evidence",
    "Connects statistical results to hypotheses; selection is intentionally curated"
  )
)
pretty_table(evidence_types, "Statistical and descriptive evidence used in the paper")
```

## Mapping quality

### Question

Did diet, infection treatment, or their interaction affect the percentage of
reads mapping to the locust reference? Mapping percentages are bounded between
0 and 100, so the proportions were logit-transformed before fitting Gaussian
linear models.

```{r mapping-models}
mapping_all <- read_required_csv(
  file.path(qc_dir, "competitive_mapping_quality_all_45_samples.csv")
)
mapping_44 <- mapping_all |>
  dplyr::filter(label != "1044") |>
  dplyr::mutate(
    Diet = factor(Diet, levels = c("33", "50", "83")),
    Treatment = factor(Treatment, levels = c("Control", "Infected"))
  )

run_mapping_model <- function(data, metric) {
  proportion <- pmin(pmax(data[[metric]] / 100, 1e-6), 1 - 1e-6)
  data$mapping_logit <- qlogis(proportion)

  factorial <- stats::lm(mapping_logit ~ Diet * Treatment, data = data)
  interaction_test <- stats::drop1(
    factorial, scope = ~ Diet:Treatment, test = "F"
  )

  additive <- stats::lm(mapping_logit ~ Diet + Treatment, data = data)
  main_tests <- stats::drop1(additive, test = "F")

  data.frame(
    mapping_definition = metric,
    interaction_p = interaction_test["Diet:Treatment", "Pr(>F)"],
    diet_p = main_tests["Diet", "Pr(>F)"],
    infection_p = main_tests["Treatment", "Pr(>F)"]
  )
}

mapping_recomputed <- dplyr::bind_rows(
  run_mapping_model(mapping_44, "unique_pct"),
  run_mapping_model(mapping_44, "unique_plus_multimapped_pct")
) |>
  dplyr::mutate(
    mapping_definition = c("Unique mapping", "Unique + multimapping")
  )

mapping_exported <- read_required_csv(
  file.path(qc_dir, "competitive_mapping_design_tests.csv")
) |>
  dplyr::filter(scenario == "Final 44-sample analysis cohort") |>
  dplyr::select(mapping_definition, interaction_p, diet_p, infection_p)

stopifnot(isTRUE(all.equal(
  mapping_recomputed[, names(mapping_exported)], mapping_exported,
  tolerance = 1e-10, check.attributes = FALSE
)))

mapping_report <- mapping_exported |>
  dplyr::mutate(
    `Diet x infection P` = format_p(interaction_p),
    `Diet P` = format_p(diet_p),
    `Infection P` = format_p(infection_p)
  ) |>
  dplyr::select(`Mapping definition` = mapping_definition,
                `Diet x infection P`, `Diet P`, `Infection P`)
pretty_table(mapping_report, "Mapping-quality factorial tests in the 44-library cohort")
```

**Result and interpretation.** Infection treatment was associated with lower
mapping yield for both uniquely mapped reads (P = 0.028) and unique plus
accepted multimapping reads (P = 0.046). Neither diet nor the interaction was
supported. This points to a treatment-associated difference in the fraction of
reads assigned to the locust genome, but mapping yield is a technical and
biological mixture and is not itself a measure of fungal load.

## Body mass and mapping yield

### Question

Did diet or infection affect recorded individual mass, and could body mass
explain mapping differences?

```{r mass-models}
mass_data <- read_required_csv(
  file.path(qc_dir, "individual_mass_analysis_44_samples.csv")
) |>
  dplyr::mutate(
    Diet = factor(Diet, levels = c("33", "50", "83")),
    Treatment = factor(Treatment, levels = c("Control", "Infected"))
  )

mass_factorial <- stats::lm(mass_g ~ Diet * Treatment, data = mass_data)
mass_interaction <- stats::drop1(
  mass_factorial, scope = ~ Diet:Treatment, test = "F"
)
mass_additive <- stats::lm(mass_g ~ Diet + Treatment, data = mass_data)
mass_main <- stats::drop1(mass_additive, test = "F")
mass_ci <- stats::confint(mass_additive, "TreatmentInfected", level = 0.95)

mass_tests <- data.frame(
  term = c("Diet x Infection", "Diet", "Infection"),
  F = c(
    mass_interaction["Diet:Treatment", "F value"],
    mass_main["Diet", "F value"],
    mass_main["Treatment", "F value"]
  ),
  p_value = c(
    mass_interaction["Diet:Treatment", "Pr(>F)"],
    mass_main["Diet", "Pr(>F)"],
    mass_main["Treatment", "Pr(>F)"]
  )
)

mass_exported <- read_required_csv(
  file.path(qc_dir, "individual_mass_diet_infection_tests.csv")
)
stopifnot(isTRUE(all.equal(
  mass_tests$p_value, mass_exported$p_value,
  tolerance = 1e-10, check.attributes = FALSE
)))

mass_effect <- read_required_csv(
  file.path(qc_dir, "individual_mass_infection_effect.csv")
)
mass_report <- mass_tests |>
  dplyr::mutate(P = format_p(p_value)) |>
  dplyr::select(Term = term, F, P)
pretty_table(mass_report, "Factorial tests for recorded individual mass")
pretty_table(
  mass_effect |>
    dplyr::transmute(
      Comparison = comparison,
      `Difference (g)` = estimate_g,
      `95% CI` = sprintf("%.3f to %.3f", lower_ci_g, upper_ci_g),
      P = format_p(p_value)
    ),
  "Adjusted infection-treatment difference in mass"
)
```

```{r mapping-mass-models}
run_mapping_mass_test <- function(data, metric) {
  spearman <- stats::cor.test(
    data$mass_g, data[[metric]], method = "spearman", exact = FALSE
  )
  data$mapping_logit <- qlogis(
    pmin(pmax(data[[metric]] / 100, 1e-6), 1 - 1e-6)
  )
  adjusted <- stats::lm(
    mapping_logit ~ mass_g + Diet + Treatment, data = data
  )
  adjusted_test <- stats::drop1(adjusted, test = "F")
  data.frame(
    mapping_definition = metric,
    spearman_rho = unname(spearman$estimate),
    spearman_p = spearman$p.value,
    adjusted_mass_p = adjusted_test["mass_g", "Pr(>F)"]
  )
}

mapping_mass_recomputed <- dplyr::bind_rows(
  run_mapping_mass_test(mass_data, "unique_pct"),
  run_mapping_mass_test(mass_data, "unique_plus_multimapped_pct")
) |>
  dplyr::mutate(
    mapping_definition = c("Unique mapping", "Unique + multimapping")
  )

mapping_mass_exported <- read_required_csv(
  file.path(qc_dir, "mapping_quality_mass_association_tests.csv")
) |>
  dplyr::filter(scenario == "Final 44-sample analysis cohort") |>
  dplyr::select(mapping_definition, spearman_rho, spearman_p,
                adjusted_mass_p)

stopifnot(isTRUE(all.equal(
  mapping_mass_recomputed[, names(mapping_mass_exported)],
  mapping_mass_exported,
  tolerance = 1e-10, check.attributes = FALSE
)))

mapping_mass_report <- mapping_mass_exported |>
  dplyr::mutate(
    `Spearman rho` = round(spearman_rho, 3),
    `Spearman P` = format_p(spearman_p),
    `Adjusted mass P` = format_p(adjusted_mass_p)
  ) |>
  dplyr::select(`Mapping definition` = mapping_definition,
                `Spearman rho`, `Spearman P`, `Adjusted mass P`)
pretty_table(mapping_mass_report, "Association between mass and mapping yield")
```

**Result and interpretation.** Infection was associated with a mean mass
difference of -0.189 g after accounting for diet (95% CI -0.368 to -0.010 g;
P = 0.039). Diet and the diet-by-infection interaction were not supported.
Mass was not significantly associated with either mapping definition, with or
without adjustment for diet and infection. The lower host mapping in infected
libraries therefore was not explained by recorded body mass in this cohort.

## Differential expression

### Why raw counts are the model input

DESeq2 does normalize before testing. Raw integer counts are supplied to the
negative-binomial model, and DESeq2 estimates sample-specific size factors by
its median-of-ratios method. Variance-stabilized counts are used only for PCA
and heatmaps because they are designed for distances and visualization, not for
the count likelihood used by differential-expression tests.

### Full factorial model

The full model asks whether the infection response changes with diet:

```{r deseq2-interaction-code, eval=FALSE}
dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = counts[keep_genes, ],
  colData = metadata,
  design = ~ Diet + Treatment + Diet:Treatment
)
dds <- DESeq2::DESeq(
  dds, minReplicatesForReplace = Inf, quiet = TRUE
)

# Example: diet 83 changes the infection response relative to diet 50.
res_interaction_83 <- DESeq2::results(
  dds,
  name = "Diet83.TreatmentInfected",
  alpha = 0.05,
  cooksCutoff = FALSE
)
```

The code above is executed in
[Main-chromosome interaction model](14-all-samples-deseq2-interaction-model.html).
With diet 50 and controls as reference levels, the main treatment coefficient
is the infected-versus-control response in diet 50. Interaction coefficients
measure how that response differs in diets 33 or 83.

```{r interaction-results}
interaction_counts <- read_required_csv(
  file.path(interaction_dir, "significant_gene_counts.csv")
)
pretty_table(
  interaction_counts,
  "Significant coefficients in the full diet-by-infection model"
)
```

**Result and interpretation.** The infection response in diet 33 did not have
significant interaction coefficients relative to diet 50 at the current DEG
threshold. Diet 83 had 38 significant interaction coefficients relative to
diet 50. These coefficients provide the direct gene-level test of response
heterogeneity; differences in the number of DEGs called separately within each
diet are descriptive and should not be interpreted as interaction tests by
themselves.

### Infection within each diet and average infection response

The within-diet contrasts use a no-intercept group model so every requested
comparison is an explicit difference between two observed conditions.

```{r deseq2-direct-contrast-code, eval=FALSE}
metadata$group <- factor(paste0(
  "diet", metadata$Diet, "_", tolower(metadata$Treatment)
))

dds_group <- DESeq2::DESeqDataSetFromMatrix(
  countData = counts[keep_genes, ],
  colData = metadata,
  design = ~ 0 + group
)
dds_group <- DESeq2::DESeq(
  dds_group, minReplicatesForReplace = Inf, quiet = TRUE
)

# Positive log2FC means higher expression in infected diet-50 locusts.
res_diet50 <- DESeq2::results(
  dds_group,
  contrast = c("group", "diet50_infected", "diet50_control"),
  alpha = 0.05,
  cooksCutoff = FALSE
)

is_deg <- !is.na(res_diet50$padj) &
  res_diet50$padj < 0.05 & abs(res_diet50$log2FoldChange) >= 1
```

The group model and within-diet contrasts are executed in
[Infection within diet](16-all-samples-infection-within-diet-deg.html). The
average infected-versus-control contrast across diets and all Figure 2
contrasts are assembled in
[Publication figure exports](08-publication-figures.html).

```{r infection-deg-results}
infection_summary <- read_required_csv(
  file.path(infection_dir, "infection_within_diet_summary.csv")
)
infection_direction <- read_required_csv(
  file.path(infection_dir, "infection_within_diet_direction_counts.csv")
)
infection_biotype <- read_required_csv(
  file.path(infection_dir, "infection_deg_burden_by_biotype_counts.csv")
)
average_infection <- read_required_csv(
  file.path(publication_dir, "main_chromosome_average_infection_results.csv")
)

average_sig <- average_infection |>
  dplyr::filter(!is.na(padj), padj < params$alpha,
                abs(log2FoldChange) >= params$lfc_threshold)
average_summary <- data.frame(
  contrast = "All diets: infected vs control",
  total_DEGs = nrow(average_sig),
  higher_in_infected = sum(average_sig$log2FoldChange > 0),
  higher_in_control = sum(average_sig$log2FoldChange < 0)
)

within_summary <- infection_direction |>
  tidyr::pivot_wider(
    names_from = direction, values_from = n_significant_genes,
    values_fill = 0
  ) |>
  dplyr::left_join(infection_summary, by = "diet") |>
  dplyr::transmute(
    contrast = paste0("Diet ", diet, ": infected vs control"),
    total_DEGs = n_significant_genes,
    higher_in_infected = `Higher in infected`,
    higher_in_control = `Higher in control`
  )

deg_summary <- dplyr::bind_rows(average_summary, within_summary)
pretty_table(deg_summary, "Primary infection-response DEG counts")

lncrna_summary <- infection_biotype |>
  dplyr::filter(Gene_class == "Long non-coding RNA") |>
  dplyr::group_by(diet) |>
  dplyr::summarise(lncRNA_DEGs = sum(n_significant_genes), .groups = "drop")
pretty_table(lncrna_summary, "Long non-coding RNA DEGs within each diet")
```

**Result and interpretation.** The average infection contrast contained 1,696
DEGs (907 higher in infected locusts and 789 higher in controls). Infection
altered 1,036 genes in diet 33, 2,000 in diet 50, and 1,013 in diet 83. The
balanced diet therefore had the largest within-diet DEG burden, but the formal
interaction model remains the appropriate test of whether diet modifies the
infection response.

### Diet contrasts

```{r deseq2-diet-code, eval=FALSE}
# All-sample adjusted diet model.
dds_diet <- DESeq2::DESeqDataSetFromMatrix(
  countData = counts[keep_genes, ],
  colData = metadata,
  design = ~ Treatment + Diet
)
dds_diet <- DESeq2::DESeq(dds_diet, quiet = TRUE)
res_33_vs_50 <- DESeq2::results(
  dds_diet, contrast = c("Diet", "33", "50"),
  alpha = 0.05, cooksCutoff = FALSE
)

# Within controls or within infected animals, use design = ~ Diet after
# subsetting to the relevant treatment.
```

```{r diet-results}
diet_summary <- read_required_csv(
  file.path(diet_dir, "diet_pairwise_summary.csv")
)
pretty_table(
  diet_summary,
  "Diet DEGs after adjustment for infection status in the all-sample model"
)

exact_contrasts <- read_required_csv(
  file.path(publication_dir, "figure2_panel_A_exact_contrast_deg_gene_table.csv")
)
exact_contrast_summary <- exact_contrasts |>
  dplyr::mutate(
    direction = dplyr::if_else(
      log2FoldChange > 0,
      paste0("Higher in ", numerator),
      paste0("Higher in ", denominator)
    )
  ) |>
  dplyr::count(DEG_context, comparison, direction, name = "n_DEGs")
pretty_table(
  exact_contrast_summary,
  "Directional DEG counts for every exact comparison in Figure 2"
)
```

The adjusted all-sample diet contrasts found fewer DEGs than the infection
contrasts. The exact control-only and infected-only comparisons remain
important because they distinguish baseline diet differences from diet
differences expressed during infection.

## Chromosomal distribution of infection DEGs

### Question

Were infection-responsive genes concentrated on one chromosome, or distributed
throughout the placed host genome? The expected count for each chromosome is
based on the number of genes from that chromosome that actually entered the
DESeq2 tests, not on chromosome length.

```{r chromosome-distribution-test}
placement <- readr::read_tsv(
  file.path(project_dir, params$placement_file), show_col_types = FALSE
) |>
  dplyr::distinct(gene_id, .keep_all = TRUE)

within_all <- read_required_csv(
  file.path(infection_dir, "all_infection_within_diet_results.csv")
)
sig_gene_ids <- unique(c(
  average_infection$gene_id[
    !is.na(average_infection$padj) &
      average_infection$padj < params$alpha &
      abs(average_infection$log2FoldChange) >= params$lfc_threshold
  ],
  within_all$gene_id[
    !is.na(within_all$padj) &
      within_all$padj < params$alpha &
      abs(within_all$log2FoldChange) >= params$lfc_threshold
  ]
))
testable_gene_ids <- unique(c(
  average_infection$gene_id[!is.na(average_infection$padj)],
  within_all$gene_id[!is.na(within_all$padj)]
))

chromosome_background <- placement |>
  dplyr::filter(
    gene_id %in% testable_gene_ids,
    sequence_class == "Nuclear chromosome"
  ) |>
  dplyr::count(assigned_molecule, name = "testable_genes")

chromosome_degs <- placement |>
  dplyr::filter(
    gene_id %in% sig_gene_ids,
    sequence_class == "Nuclear chromosome"
  ) |>
  dplyr::count(assigned_molecule, name = "DEGs")

chromosome_table <- chromosome_background |>
  dplyr::left_join(chromosome_degs, by = "assigned_molecule") |>
  dplyr::mutate(
    DEGs = dplyr::coalesce(DEGs, 0L),
    non_DEGs = testable_genes - DEGs,
    DEG_percent = 100 * DEGs / testable_genes
  )

chromosome_matrix <- rbind(
  DEG = chromosome_table$DEGs,
  non_DEG = chromosome_table$non_DEGs
)
chromosome_test <- suppressWarnings(stats::chisq.test(chromosome_matrix))
cramers_v <- sqrt(
  unname(chromosome_test$statistic) / sum(chromosome_matrix)
)

chromosome_test_summary <- data.frame(
  unique_infection_DEGs = length(sig_gene_ids),
  testable_genes = length(testable_gene_ids),
  chromosomes_with_DEGs = sum(chromosome_table$DEGs > 0),
  chi_square = unname(chromosome_test$statistic),
  df = unname(chromosome_test$parameter),
  p_value = chromosome_test$p.value,
  cramers_v = cramers_v
)
write.csv(chromosome_table,
          file.path(out_dir, "infection_deg_chromosome_counts.csv"),
          row.names = FALSE)
write.csv(chromosome_test_summary,
          file.path(out_dir, "infection_deg_chromosome_test.csv"),
          row.names = FALSE)

pretty_table(chromosome_test_summary,
             "Chromosome-distribution test for infection-responsive genes")
```

**Result and interpretation.** Across the average and within-diet infection
contrasts, `r format(length(sig_gene_ids), big.mark = ",")` unique DEGs occurred
on all `r sum(chromosome_table$DEGs > 0)` nuclear chromosomes. Their distribution
differed from the expression-tested background
($\chi^2_{`r unname(chromosome_test$parameter)`}$ =
`r round(unname(chromosome_test$statistic), 2)`, P =
`r format_p(chromosome_test$p.value)`), but the effect was weak (Cramer's V =
`r round(cramers_v, 3)`). This supports a genome-wide response with modest
chromosomal heterogeneity, rather than a response confined to one chromosome.

## GO and KEGG enrichment

### Test and background

For each DEG set, an over-representation test asks whether a GO term or KEGG
reference pathway contains more DEGs than expected from the annotated genes in
the expression-tested background. Upregulated and downregulated genes are
combined in the main enrichment views. Enrichment does not test whether every
gene in a term changes in the same direction.

```{r enrichment-code, eval=FALSE}
run_enrichment <- function(genes, term2gene, term2name, universe) {
  genes <- unique(genes[genes %in% universe])
  if (length(genes) < 5) return(NULL)

  clusterProfiler::enricher(
    gene = genes,
    universe = universe,
    TERM2GENE = term2gene,
    TERM2NAME = term2name[, c("term", "name")],
    pvalueCutoff = 0.1,
    pAdjustMethod = "BH",
    qvalueCutoff = 0.2
  )
}
```

This exact function is executed in
[GO and KEGG enrichment](10-go-term-enrichment.html). The broader export retains
terms under its exploratory configuration; the paper-facing summary below
applies the stricter requirement of BH-adjusted P < 0.05.

```{r enrichment-results}
go_enrichment <- read_required_csv(
  file.path(enrichment_dir, "GO_enrichment_all_sets.csv")
) |>
  dplyr::filter(p.adjust < params$alpha)
kegg_enrichment <- read_required_csv(
  file.path(enrichment_dir, "KEGG_pathway_enrichment_all_sets.csv")
) |>
  dplyr::filter(p.adjust < params$alpha)

enrichment_summary <- dplyr::bind_rows(
  go_enrichment |>
    dplyr::count(set, name = "significant_terms") |>
    dplyr::mutate(database = "GO"),
  kegg_enrichment |>
    dplyr::count(set, name = "significant_terms") |>
    dplyr::mutate(database = "KEGG reference pathway")
) |>
  dplyr::select(database, DEG_set = set, significant_terms)
pretty_table(enrichment_summary,
             "Enriched terms with BH-adjusted P below 0.05")

top_enrichment <- dplyr::bind_rows(
  go_enrichment |>
    dplyr::mutate(database = "GO"),
  kegg_enrichment |>
    dplyr::mutate(database = "KEGG")
) |>
  dplyr::group_by(database, set) |>
  dplyr::slice_min(p.adjust, n = 3, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::transmute(
    Database = database, `DEG set` = set, Term = Description,
    `DEGs in term` = Count, `Adjusted P` = format_p(p.adjust)
  )
pretty_table(top_enrichment,
             "Three strongest adjusted enrichment results per DEG set")
```

GO terms provide ontology-based summaries of biological process, cellular
component, and molecular function. KEGG entries here are KO reference maps
derived from eggNOG annotations, not direct *S. gregaria*-specific pathway
measurements. The pathway-scope labels and arthropod-focused figure variants
should therefore be used when making organism-level claims.

## Phase-reference overlap

### Question

Could semi-isolation during the infection protocol have contributed a broader
phase-associated expression signal? Infection DEGs were compared with
independent solitarious-versus-gregarious DEG sets from head and thorax. The
most conservative phase-reference set is the intersection shared by both
reference tissues.

```{r phase-overlap-code}
phase_membership <- read_required_csv(
  file.path(publication_dir, "figure2_infection_phase_overlap_membership.csv")
)

infection_phase_genes <- phase_membership$gene_id[
  phase_membership$`Infection within diets`
]
head_phase_genes <- phase_membership$gene_id[phase_membership$`Head phase DEGs`]
thorax_phase_genes <- phase_membership$gene_id[
  phase_membership$`Thorax phase DEGs`
]
shared_phase_genes <- intersect(head_phase_genes, thorax_phase_genes)
infection_shared_phase <- intersect(infection_phase_genes, shared_phase_genes)

phase_overlap_summary <- data.frame(
  set = c(
    "Infection within diets", "Head phase", "Thorax phase",
    "Phase shared by head and thorax",
    "Infection also shared by head and thorax phase references"
  ),
  genes = c(
    length(unique(infection_phase_genes)),
    length(unique(head_phase_genes)),
    length(unique(thorax_phase_genes)),
    length(shared_phase_genes),
    length(infection_shared_phase)
  )
)
pretty_table(phase_overlap_summary, "Phase-reference overlap audit")
```

**Result and interpretation.** The conservative phase-reference set contained
`r length(shared_phase_genes)` genes shared between head and thorax, and
`r length(infection_shared_phase)` of these also occurred among the fat-body
infection DEGs. This overlap shows that part of the infection-associated signal
uses genes responsive to density phase in two independent tissues. It identifies
a plausible shared physiological programme and a potential source of
confounding, but it cannot establish that the experimental fat bodies underwent
a phase transition or separate infection from semi-isolation causally.

No overlap P value is assigned here. The phase sets come from different tissues
and an independent experiment with limited reference replication; presenting
the observed membership is more defensible than imposing a universe that would
make a formal overlap test appear more precise than the design allows.

### Which infection-responsive genes overlap the phase references?

The Venn counts come from explicit gene identifiers. The table below joins
those identifiers to the official host annotation, the within-diet infection
results, and the independent head and thorax phase results. Positive infection
log2 fold-changes mean higher expression in infected fat bodies. The phase
direction labels retain the direction reported by the independent reference
analysis.

```{r phase-overlap-gene-catalogue}
head_phase_results <- read_required_csv(
  file.path(project_dir, params$head_phase_file)
) |>
  dplyr::filter(significant) |>
  dplyr::transmute(
    gene_id,
    head_phase_log2FC = log2FoldChange,
    head_phase_padj = padj,
    head_phase_direction = direction
  )

thorax_phase_results <- read_required_csv(
  file.path(project_dir, params$thorax_phase_file)
) |>
  dplyr::filter(significant) |>
  dplyr::transmute(
    gene_id,
    thorax_phase_log2FC = log2FoldChange,
    thorax_phase_padj = padj,
    thorax_phase_direction = direction
  )

infection_sig_by_diet <- within_all |>
  dplyr::filter(
    !is.na(padj), padj < params$alpha,
    abs(log2FoldChange) >= params$lfc_threshold
  ) |>
  dplyr::mutate(
    infection_direction = dplyr::if_else(
      log2FoldChange > 0, "Higher in infected", "Higher in control"
    ),
    diet_result = paste0(
      "Diet ", diet, ": ", infection_direction,
      " (log2FC ", sprintf("%.2f", log2FoldChange), ")"
    )
  )

infection_gene_summary <- infection_sig_by_diet |>
  dplyr::arrange(gene_id, as.numeric(diet)) |>
  dplyr::group_by(gene_id) |>
  dplyr::summarise(
    infection_DEG_diets = paste(diet, collapse = ", "),
    infection_results = paste(diet_result, collapse = "; "),
    min_infection_padj = min(padj),
    max_abs_infection_log2FC = max(abs(log2FoldChange)),
    .groups = "drop"
  )

phase_gene_catalogue <- phase_membership |>
  dplyr::filter(`Infection within diets`) |>
  dplyr::mutate(
    phase_overlap_category = dplyr::case_when(
      `Head phase DEGs` & `Thorax phase DEGs` ~
        "Shared head + thorax phase overlap",
      `Head phase DEGs` ~ "Head phase overlap only",
      `Thorax phase DEGs` ~ "Thorax phase overlap only",
      TRUE ~ "No head/thorax phase overlap"
    )
  ) |>
  dplyr::select(
    gene_id, phase_overlap_category,
    head_phase_DEG = `Head phase DEGs`,
    thorax_phase_DEG = `Thorax phase DEGs`
  ) |>
  dplyr::left_join(
    placement |>
      dplyr::select(
        gene_id, description, gene_biotype,
        chromosome = assigned_molecule
      ),
    by = "gene_id"
  ) |>
  dplyr::left_join(infection_gene_summary, by = "gene_id") |>
  dplyr::left_join(head_phase_results, by = "gene_id") |>
  dplyr::left_join(thorax_phase_results, by = "gene_id") |>
  dplyr::arrange(
    factor(
      phase_overlap_category,
      levels = c(
        "Shared head + thorax phase overlap",
        "Head phase overlap only", "Thorax phase overlap only",
        "No head/thorax phase overlap"
      )
    ),
    min_infection_padj
  )

write.csv(
  phase_gene_catalogue,
  file.path(out_dir, "infection_phase_overlap_annotated_gene_catalogue.csv"),
  row.names = FALSE
)

phase_category_audit <- phase_gene_catalogue |>
  dplyr::count(phase_overlap_category, name = "infection_DEGs") |>
  dplyr::mutate(
    percent_of_infection_DEGs = 100 * infection_DEGs / sum(infection_DEGs)
  )
pretty_table(
  phase_category_audit,
  "Infection DEGs assigned to each phase-reference category"
)

phase_overlap_gene_table <- phase_gene_catalogue |>
  dplyr::filter(phase_overlap_category != "No head/thorax phase overlap") |>
  dplyr::select(
    gene_id, description, gene_biotype, chromosome,
    phase_overlap_category, infection_DEG_diets, infection_results,
    head_phase_direction, head_phase_log2FC,
    thorax_phase_direction, thorax_phase_log2FC,
    min_infection_padj
  )

if (requireNamespace("DT", quietly = TRUE)) {
  DT::datatable(
    phase_overlap_gene_table,
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 15, scrollX = TRUE),
    caption = "Infection-responsive genes with head and/or thorax phase-reference overlap"
  )
} else {
  pretty_table(
    utils::head(phase_overlap_gene_table, 25),
    "First 25 infection-responsive genes with phase-reference overlap"
  )
}
```

The exported catalogue contains all 2,581 infection-responsive genes, including
the 2,117 genes with no phase-reference overlap. The searchable table displayed
here focuses on the 464 genes supported by at least one phase-reference tissue.

## Non-host taxonomic profiles

### Questions

1. Does the relative composition of non-host taxonomic families differ with
   diet, infection treatment, or their interaction?
2. Are apparent compositional differences explained by unequal dispersion?
3. Does total normalized non-host read burden differ across the design?

```{r taxonomy-code, eval=FALSE}
set.seed(20260810)
family_permanova <- vegan::adonis2(
  family_bray ~ Diet * Treatment,
  data = family_test_metadata,
  permutations = 999,
  by = "terms"
)

family_dispersion <- vegan::betadisper(
  family_bray,
  group = family_test_metadata$condition
)
set.seed(20260810)
family_dispersion_test <- vegan::permutest(
  family_dispersion, permutations = 999
)

burden_factorial <- lm(log10_microbial_rpm ~ Diet * Treatment,
                       data = microbial_burden)
burden_interaction <- drop1(burden_factorial, test = "F")
burden_additive <- lm(log10_microbial_rpm ~ Diet + Treatment,
                      data = microbial_burden)
burden_main <- drop1(burden_additive, test = "F")
```

The complete construction of the Bracken family matrix and Bray-Curtis distance
object is executed in
[Fungal burden and taxonomic profiles](29-metarhizium-read-burden.html).

```{r taxonomy-results}
permanova <- read_required_csv(
  file.path(taxonomy_dir, "bracken_host_unmapped_family_permanova.csv")
)
permdisp <- read_required_csv(
  file.path(taxonomy_dir, "bracken_host_unmapped_family_dispersion_test.csv")
)
burden_tests <- read_required_csv(
  file.path(taxonomy_dir, "bracken_host_unmapped_burden_design_tests.csv")
)
burden_groups <- read_required_csv(
  file.path(taxonomy_dir, "bracken_host_unmapped_burden_group_summary.csv")
)

pretty_table(
  permanova |>
    dplyr::filter(term %in% c("Diet", "Treatment", "Diet:Treatment")) |>
    dplyr::transmute(Term = term, df, R2, F, P = format_p(p_value)),
  "PERMANOVA of non-host family composition"
)
pretty_table(
  permdisp |>
    dplyr::transmute(Test = test, F, P = format_p(p_value)),
  "Dispersion test accompanying PERMANOVA"
)
pretty_table(
  burden_tests |>
    dplyr::transmute(Term = term, Model = model, F, P = format_p(p_value)),
  "Linear-model tests of normalized non-host read burden"
)
pretty_table(
  burden_groups,
  "Normalized non-host read burden by infection treatment"
)
```

**Result and interpretation.** Infection treatment explained 11.5% of the
Bray-Curtis variation in non-host family composition (PERMANOVA P = 0.001),
whereas diet and the interaction were unsupported. Group dispersion did not
differ (PERMDISP P = 0.625), reducing concern that the PERMANOVA result was
driven only by unequal variability. Normalized non-host burden was also higher
in infected libraries (P < 0.001), with no significant diet-by-infection
interaction. These RNA-seq classifications support treatment-associated
non-host signal, but they are not a dedicated microbiome survey and do not by
themselves establish viable colonization.

## Paper-ready statistical summary

```{r paper-summary}
paper_summary <- data.frame(
  topic = c(
    "Unique host mapping", "Unique plus multimapping", "Recorded mass",
    "Average infection response", "Within-diet infection response",
    "Diet-by-infection interaction", "Chromosome distribution",
    "Non-host family composition", "Non-host burden", "Phase-reference overlap"
  ),
  result = c(
    "Infection P = 0.028; diet and interaction unsupported",
    "Infection P = 0.046; diet and interaction unsupported",
    "Infected - control = -0.189 g (95% CI -0.368 to -0.010), P = 0.039",
    "1,696 DEGs: 907 higher in infected and 789 higher in controls",
    "1,036, 2,000, and 1,013 DEGs in diets 33, 50, and 83",
    "0 significant diet-33 interaction genes and 38 diet-83 interaction genes relative to diet 50",
    paste0("All 12 chromosomes; chi-square P ", format_p(chromosome_test$p.value),
           ", Cramer's V = ", round(cramers_v, 3)),
    "Treatment R2 = 0.115, PERMANOVA P = 0.001; PERMDISP P = 0.625",
    "Infection P < 0.001; diet and interaction unsupported",
    paste0(length(infection_shared_phase),
           " infection DEGs shared with the head-and-thorax phase intersection")
  ),
  reporting_caution = c(
    "Mapping percentage is not fungal load",
    "Multimapping reads were not assigned to genes by featureCounts",
    "Mass was recorded before dissection and is not a fat-body-specific trait",
    "This is the average response across diets",
    "DEG-count differences are not formal interaction tests",
    "Interaction coefficients test response modification directly",
    "The significant deviation is weak in effect size",
    "Taxonomic assignment from RNA-seq is not a microbiome assay",
    "Non-host reads can reflect several biological and technical sources",
    "Cross-study overlap is descriptive and cannot prove phase transition"
  )
)
write.csv(paper_summary, file.path(out_dir, "paper_statistical_results_summary.csv"),
          row.names = FALSE)
pretty_table(paper_summary, "Statistical statements and their interpretation limits")
```

## Reproducibility notes

- All DESeq2 pages retain full gene-level tables, including non-significant
  genes, so thresholds and rankings can be audited without reconstructing the
  model from a figure.
- Each contrast receives its own Benjamini-Hochberg correction. Adjusted P
  values should not be compared as if they arose from one combined family of
  tests across all contrasts.
- The exon-only branch is a sensitivity analysis with the same sample cohort,
  placement rule, rRNA exclusion, filters, models, and DEG thresholds.
- PCA, heatmaps, Venn diagrams, and curated biological themes help explain the
  data but do not replace the corresponding model tests.

```{r session-info}
writeLines(capture.output(sessionInfo()), file.path(out_dir, "sessionInfo.txt"))
sessionInfo()
```
